### 2017-10-21 1-preprocessing.Rmd
begin notebooks

---
title: "Preprocessing"
author: "Florian Privé"
date: "October 20, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
                      fig.asp = 0.7, error = TRUE)
options(width = 85)
```

## Data

### QC and imputation

This was done in another paper TODO(add ref).

### Add population from external files

```{r}
file.copy(paste0("../paper1-packages/backingfiles/celiacQC", c(".bk", ".rds")),
          paste0("backingfiles/celiacQC", c(".bk", ".rds")))

library(bigsnpr)
celiac <- snp_attach("backingfiles/celiacQC.rds")

# Get population
pop.files <- list.files(path = "../thesis-celiac/Dubois2010_data/", 
                        pattern = "cluster_*", full.names = TRUE)
pop <- snp_getSampleInfos(celiac, pop.files)[[1]]
pop.names <- c("Netherlands", "Italy", "UK1", "UK2", "Finland")
celiac$fam$family.ID <- pop.names[pop]
celiac <- snp_save(celiac)
# Read modified file to check
celiac <- snp_attach("backingfiles/celiacQC.rds")
rle(celiac$fam$family.ID)
```

### Filter samples based on population structure

```{r}
# Compute PCA
obj.svd <- snp_autoSVD(celiac$genotypes, 
                       celiac$map$chromosome,
                       celiac$map$physical.pos, 
                       ncores = nb_cores())

library(ggplot2)
plot(obj.svd, type = "scores", scores = 1:2) + 
  aes(color = celiac$fam$family.ID) + 
  labs(color = "Population")
```

```{r}
# Which are UK controls?
ind.UKcontrols <- which(startsWith(celiac$fam$family.ID, "UK") & 
                          celiac$fam$affection == 1)
obj.svd$u[-ind.UKcontrols, ] <- NA
```

```{r}
# Compute Mahalanobis distance on PCs
distM <- rep(NA, nrow(celiac$genotypes))
distM[ind.UKcontrols] <- robust::covRob(obj.svd$u[ind.UKcontrols, ])$dist
THR <- 30
# Plot UK controls, coloring with Mahalanobis distance
plotlist <- lapply(list(1:2, 3:4, 5:6, 7:8), function(scores) {
  plot(obj.svd, type = "scores", scores = scores, coeff = 0.7) + 
    aes(color = (distM > THR)) + 
    labs(color = "Outlier?")
}) 
cowplot::plot_grid(plotlist = plotlist, ncol = 2, align = "hv")
```

```{r}
# New dataset with only non-outliers UK controls
length(ind.keep <- which(distM < THR))
celiac2 <- snp_attach(subset(celiac, ind.row = ind.keep))
obj.svd2 <- snp_autoSVD(celiac2$genotypes, 
                        celiac2$map$chromosome,
                        celiac2$map$physical.pos,
                        ncores = nb_cores())
plot(obj.svd2)
plot(obj.svd2, type = "scores", scores = 1:2) 
# Save this new PCA
saveRDS(obj.svd2, "backingfiles/PCA2.rds")
```

### 2017-10-21 2-newfiles.Rmd
begin notebooks

---
title: "New files"
author: "Florian Privé"
date: "20 octobre 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
                      fig.asp = 0.7, error = TRUE)
options(width = 85)
```

### Files for ttrees

```{r}
library(bigsnpr)
celiac2 <- snp_attach("backingfiles/celiacQC_sub1.rds")
G <- celiac2$genotypes
```

```{r}
# jdb
file.jdb <- "backingfiles/ttrees.jdb"
cat(rep(";", 5), sep = "\n", file = file.jdb)
big_apply(G, a.FUN = function(X, ind) {
  df <- cbind(paste0("ind_", ind), rep(3, length(ind)), as.data.frame(X[ind, ]))
  data.table::fwrite(df, file = file.jdb, sep = " ", append = TRUE,
                     quote = FALSE, col.names = FALSE)
  NULL
}, a.combine = "c", ind = rows_along(G), block.size = 1e3)
```

```{r}
# bloc
bigsnpr:::write.table2(
  bigstatsr:::CutBySize(ncol(G), block.size = 10)[, 1:2] - 1,
  file = file.bloc <- sub("\\.jdb$", ".bloc", file.jdb)
)
readLines(file.bloc, 5)
```

### File with recessive/dominant information

```{Rcpp}
// [[Rcpp::depends(bigstatsr, BH)]]
#include <bigstatsr/BMCodeAcc.h>

// [[Rcpp::export]]
void tripleBM(Environment BM, Environment BM2) {

  XPtr<FBM> xpMat = BM["address"];
  int n = xpMat->nrow();
  int m = xpMat->ncol();
  SubBMCode256Acc macc(xpMat, seq_len(n)-1, seq_len(m)-1, BM["code256"]);

  XPtr<FBM> xpMat2 = BM2["address"];
  BMAcc<unsigned char> macc2(xpMat2);

  int i, j, j2;

  for (j = j2 = 0; j < m; j++, j2 += 3) {
    for (i = 0; i < n; i++) {
      macc2(i, j2)   = macc(i, j);
      macc2(i, j2+1) = macc(i, j) >= 0.5;
      macc2(i, j2+2) = macc(i, j) >  1.5;
    }
  }
}
```

```{r}
snp_triple <- function(x) {

  G0 <- x$genotypes
  G <- G0$copy(code = round(G0$code256))

  newfile <- bigsnpr:::getNewFile(x, "tripled")
  G2 <- FBM.code256(nrow(G), 3 * ncol(G), 
                    code = G$code256,
                    backingfile = newfile,
                    save = TRUE)

  tripleBM(G, G2)

  paste0(newfile, ".rds")
}
```

```{r}
snp_triple(celiac2)
```

```{r}
# List all the files with their size in GB
round(sapply(list.files("backingfiles", full.names = TRUE), file.size) / 1024^3, 1)
```

### 2017-10-21 save-scripts/7-ttrees-final.R
mv scripts

@@ -8,11 +8,8 @@ ttrees <- function(TTree, file.base, pheno.all, ind.train, ind.test,
   tmpfile <- tempfile()
   write(c(rep("", 5), pheno.all), ncolumns = 1, tmpfile)
   # https://stackoverflow.com/a/7846550/6103040
-  writeLines(
-    system(sprintf("awk 'FNR==NR{a[NR]=$1;next}{$2=a[FNR]}1' %s %s",
-                   tmpfile, file0.jdb), intern = TRUE),
-    file.jdb <- paste0(tmpfile, ".jdb")
-  )
+  system(sprintf("awk 'FNR==NR{a[NR]=$1;next}{$2=a[FNR]}1' %s %s > %s",
+                 tmpfile, file0.jdb, file.jdb <- paste0(tmpfile, ".jdb")))
   # Write indices of learning and validation sets
   file.learn <- paste0(tmpfile, "_learn.txt")
   file.val   <- paste0(tmpfile, "_val.txt")
@@ -32,15 +29,16 @@ ttrees <- function(TTree, file.base, pheno.all, ind.train, ind.test,
     ))
   )[3]
   
-  preds <- read.table(sprintf("%s_k1000_m3_t%d_ic5_nmin2000_0000.roc", 
-                              file.jdb, n.trees), 
-                      header = FALSE)
+  file.roc <- sprintf("%s_k1000_m3_t%d_ic5_nmin2000_0000.roc",
+                      file.jdb, n.trees)
+  file.vim <- sub("\\.roc$", ".vim", file.roc)
+  preds <- read.table(file.roc, header = FALSE)
   
   tibble(
-    method   = "T-trees",
+    method   = "T-Trees",
     eval     = list(preds[match(ind.test - 1, preds[[1]]), 2:3]),
     timing   = timing,
-    nb.preds = NA_integer_
+    nb.preds = sum(read.table(file.vim)$V2 != 0)
   )
 }
 

### 2017-10-21 save-scripts/9-final-simus.R

@@ -12,7 +12,7 @@ params.grid <- expand.grid(
   stringsAsFactors = FALSE
 ) 
 
-ind.HLA <- subset(LD.wiki34, ID == "hild12")
+
 
 for (i in rows_along(results)) {
   

### 2017-10-21 3-simus.Rmd
save

---
title: "Simulations"
author: "Florian Privé"
date: "October 20, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
                      fig.asp = 0.7, error = TRUE)
options(width = 85)
```

## Methods' functions

```{r, message=FALSE}
library(tidyverse)
```

Each method's function returns a tibble (data frame) with 4 columns:

1. the name of the method,
2. the predictive scores and true phenotypes for the test set, as a [list-column](http://r4ds.had.co.nz/many-models.html),
3. the timing of the main computations (in seconds),
4. the number of SNPs used for the prediction.

### PRS

```{r}
PRS <- function(G, infos.chr, infos.pos, 
                pheno.all, covar.all, 
                ind.train, ind.test) {

  # GWAS
  gwas.train <- big_univLogReg(
    G, pheno.all[ind.train], ind.train = ind.train, 
    covar.train = covar.all[ind.train, , drop = FALSE], 
    ncores = nb_cores()
  )
  gwas.train.gc <- snp_gc(gwas.train)

  timing <- system.time({

    # Clumping on the test set
    ind.keep <- snp_clumping(G, infos.chr = infos.chr,
                             ind.row = ind.test,
                             thr.r2 = 0.2, 
                             S = abs(gwas.train.gc$score),
                             size = 500,
                             is.size.in.bp = TRUE,
                             infos.pos = infos.pos,
                             ncores = nb_cores())

    # PRS
    thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
    lpS <- -predict(gwas.train.gc)
    nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
    prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
                   ind.test = ind.test,
                   ind.keep = ind.keep,
                   lpS.keep = lpS[ind.keep], 
                   thr.list = thrs)
  })[3]

  ind.best <- which.max(apply(prs, 2, AUC, target = pheno.all[ind.test]))

  methods <- c("PRS-all", "PRS-stringent", "PRS-max")
  indices <- c(1:2, ind.best)

  lapply(1:3, function(i) {
    k <- indices[i]
    tibble(
      method = methods[i],
      eval = list(cbind(prs[, k], pheno.all[ind.test])),
      timing = timing,
      nb.preds = nb.pred[k]
    )
  }) %>%
    bind_rows()
}
```

### Logistic regression

```{r}
logit.CMSA <- function(G, pheno.all, covar.all, ind.train, ind.test, method) {

  timing <- system.time({

    cmsa.logit <- big_CMSA(FUN = big_spLogReg, feval = AUC,
                           X = G, y.train = pheno.all[ind.train], 
                           ind.train = ind.train, 
                           covar.train = covar.all[ind.train, , drop = FALSE],
                           alpha = 0.5, dfmax = 20e3, 
                           ncores = nb_cores())

    preds <- predict(cmsa.logit, X = G, ind.row = ind.test, 
                     covar.row = covar.all[ind.test, , drop = FALSE])
  })[3]

  tibble(
    method   = method, 
    eval     = list(cbind(preds, pheno.all[ind.test])),
    timing   = timing,
    nb.preds = sum(cmsa.logit != 0)
  )
}
```

### T-Trees

```{r}
TTree <- "../../TTree-source/TTree"

ttrees <- function(TTree, file.base, pheno.all, ind.train, ind.test,
                   n.trees = 100) {

  file0.jdb  <- paste0(file.base, ".jdb")
  file0.bloc <- paste0(file.base, ".bloc")

  # Write jdb file with new pheno
  tmpfile <- tempfile()
  write(c(rep("", 5), pheno.all), ncolumns = 1, tmpfile)
  # https://stackoverflow.com/a/7846550/6103040
  system(sprintf("awk 'FNR==NR{a[NR]=$1;next}{$2=a[FNR]}1' %s %s > %s",
                 tmpfile, file0.jdb, file.jdb <- paste0(tmpfile, ".jdb")))
  # Write indices of learning and validation sets
  file.learn <- paste0(tmpfile, "_learn.txt")
  file.val   <- paste0(tmpfile, "_val.txt")
  cat(ind.train - 1, file = file.learn, sep = "\t")
  cat(ind.test - 1,  file = file.val,   sep = "\t")

  timing <- system.time(
    system(glue::glue(
      "{TTree}",
      " -j {file.jdb}",
      " -m 3",
      " -b {file0.bloc}",
      " -l {file.learn}",
      " -v {file.val}", 
      " -t {n.trees} -k 1000 -c 5 -n 2000",
      " -x -s"
    ))
  )[3]

  file.roc <- sprintf("%s_k1000_m3_t%d_ic5_nmin2000_0000.roc",
                      file.jdb, n.trees)
  file.vim <- sub("\\.roc$", ".vim", file.roc)
  preds <- read.table(file.roc, header = FALSE)

  tibble(
    method   = "T-Trees",
    eval     = list(preds[match(ind.test - 1, preds[[1]]), 2:3]),
    timing   = timing,
    nb.preds = sum(read.table(file.vim)$V2 != 0)
  )
}
```

## Simulations

### Data

```{r}
library(bigsnpr)
celiac2 <- snp_attach("backingfiles/celiacQC_sub1.rds")
CHR <- celiac2$map$chromosome
POS <- celiac2$map$physical.pos
(G <- celiac2$genotypes)
(G2 <- big_attach("backingfiles/celiacQC_sub1_tripled1.rds"))

covar.all <- readRDS("backingfiles/PCA2.rds")$u
```

### Simulation of phenotypes

```{r}
get_pheno <- function(
  G,                                        ## matrix of genotypes
  h2,                                       ## heritability 
  M,                                        ## nbs of causal variants
  ind.possible = cols_along(G),             ## indices of possible causal variants
  effects.dist = c("gaussian", "laplace"),  ## distribution of effects 
  model = c("simple", "fancy"),             ## model for simulation
  K = 0.3                                   ## prevalence
) {

  effects.dist  <- match.arg(effects.dist)
  model <- match.arg(model)

  set <- sample(ind.possible, size = M)
  effects <- `if`(effects.dist == "gaussian", 
                  rnorm(M, sd = sqrt(h2 / M)),
                  rmutil::rlaplace(M, s = sqrt(h2 / (2*M))))

  if (model == "simple") {
    # only linear
    y.simu <- scale(G[, set]) %*% effects
  } else {
    sets <- split(1:M, sample(rep_len(1:3, M)))
    # linear
    ind1 <- sets[[1]]
    y.simu <- scale(G[, set[ind1]]) %*% effects[ind1]
    # recessive / dominant
    ind2 <- sets[[2]]
    y.simu <- y.simu + scale(G[, set[ind2]] > 0.5) %*% effects[ind2]
    # interactions
    ind3 <- matrix(sets[[3]], ncol = 2)
    y.simu <- y.simu + scale(G[, set[ind3[, 1]]] * G[, set[ind3[, 2]]]) %*% 
      effects[ind3[, 1]] * sqrt(2)
  }

  y.simu <- y.simu / sd(y.simu) * sqrt(h2)
  stopifnot(all.equal(drop(var(y.simu)), h2))
  y.simu <- y.simu + rnorm(nrow(G), sd = sqrt(1 - h2))
  pheno <- as.numeric(y.simu > qnorm(1 - K))
}
```

### Simulations

```{r}
n <- nrow(G)
n.train <- 6e3
ind.HLA <- snp_indLRLDR(CHR, POS, subset(LD.wiki34, ID == "hild12"))
```

```{r}
params.grid <- expand.grid(
  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
  par.dist   = c("gaussian", "laplace"), 
  par.h2     = c(0.5, 0.8), 
  par.model  = c("simple", "fancy"),
  num.simu   = 1:100,
  stringsAsFactors = FALSE
) 
```

```{r}
for (i in rows_along(params.grid)) {

  res.file <- paste0("results1/simu_", i, ".rds")
  if (file.exists(res.file)) next

  params <- params.grid[i, ]
  par.causal <- params[["par.causal"]][[1]]

  # Simulate phenotypes
  pheno.all <- get_pheno(
    G,    
    h2 = params[["par.h2"]], 
    M = as.integer(par.causal[1]), 
    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
    effects.dist = params[["par.dist"]], 
    model = params[["par.model"]], 
    K = 0.3                 
  )

  # Split in training/test sets
  ind.train <- sort(sample(n, size = n.train))
  ind.test <- setdiff(1:n, ind.train)

  # Get results from all methods
  res <- bind_rows(
    PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test),
    logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple"),
    logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple"),
    ttrees(TTree, "ttrees", pheno.all, ind.train, ind.test, n.trees = 100)
  )
  print(res)
  saveRDS(res, file = res.file)
}
```

### 2017-10-21 get-results.R

@@ -1,0 +1,10 @@
+tmp1 <- PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test)
+map_dbl(tmp1$eval, ~AUC(.x[, 1], .x[, 2]))
+
+tmp2 <- logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple")
+
+
+tmp3 <- logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple")
+
+
+tmp4 <- ttrees(TTree, "ttrees", pheno.all, ind.train, ind.test, n.trees = 10)

### 2017-10-21 3-simus.Rmd
okay?

@@
     logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple"),
-    ttrees(TTree, "ttrees", pheno.all, ind.train, ind.test, n.trees = 100)
+    ttrees(TTree, "backingfiles/ttrees", pheno.all, ind.train, ind.test, n.trees = 100)
   )
   print(res)
-  saveRDS(res, file = res.file)
+  params[["res"]] <- list(res)
+  saveRDS(unnest(param1, res), file = res.file)
 }

### 2017-10-21 get-results.R

@@ -7,4 +7,13 @@ tmp2 <- logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple"
 tmp3 <- logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple")
 
 
-tmp4 <- ttrees(TTree, "ttrees", pheno.all, ind.train, ind.test, n.trees = 10)
+tmp4 <- ttrees(TTree, "backingfiles/ttrees", pheno.all, ind.train, ind.test, n.trees = 10)
+
+tmp <- bind_rows(tmp1, tmp2, tmp3, tmp4)
+tmp$AUC <- map_dbl(tmp$eval, ~AUC(.x[, 1], .x[, 2]))
+tmp
+
+param1 <- params.grid[1, ]
+param1$res <- list(tmp)
+test <- unnest(param1, res)
+bind_cols(param1[rep(1, 6), ], tmp)

### 2017-10-24 3-simus.Rmd
save

--- 2017-10-21 3-simus.Rmd
@@
 knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
-                      fig.asp = 0.7, error = TRUE)
+                      fig.asp = 0.7)
 options(width = 85)
@@
 covar.all <- readRDS("backingfiles/PCA2.rds")$u
+
+n <- nrow(G)
+n.train <- 6e3
+ind.HLA <- snp_indLRLDR(CHR, POS, subset(LD.wiki34, ID == "hild12"))
 ```
@@
 
-### Simulations
+### First simulations (with T-Trees)
 
 ```{r}
-n <- nrow(G)
-n.train <- 6e3
-ind.HLA <- snp_indLRLDR(CHR, POS, subset(LD.wiki34, ID == "hild12"))
-```
-
-```{r}
-params.grid <- expand.grid(
+params.grid1 <- expand.grid(
   par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
   par.dist   = c("gaussian", "laplace"), 
-  par.h2     = c(0.5, 0.8), 
+  par.h2     = 0.8, 
   par.model  = c("simple", "fancy"),
-  num.simu   = 1:100,
+  num.simu   = 1:5,
   stringsAsFactors = FALSE
@@
 ```{r}
-for (i in rows_along(params.grid)) {
+if (!dir.exists("results1")) dir.create("results1")
+
+for (i in rows_along(params.grid1)) {
 
@@
 
-  params <- params.grid[i, ]
+  params <- params.grid1[i, ]
   par.causal <- params[["par.causal"]][[1]]
@@
     logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple"),
-    ttrees(TTree, "ttrees", pheno.all, ind.train, ind.test, n.trees = 100)
+    ttrees(TTree, "backingfiles/ttrees", pheno.all, ind.train, ind.test, n.trees = 100)
   )
   print(res)
-  saveRDS(res, file = res.file)
+  params[["res"]] <- list(res)
+  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
 }
 ```
+
+### Second simulations (without T-Trees)
+
+```{r}
+params.grid2 <- expand.grid(
+  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
+  par.dist   = c("gaussian", "laplace"), 
+  par.h2     = c(0.5, 0.8), 
+  par.model  = c("simple", "fancy"),
+  num.simu   = 1:100,
+  stringsAsFactors = FALSE
+) 
+```
+
+```{r}
+if (!dir.exists("results2")) dir.create("results2")
+
+for (i in rows_along(params.grid2)) {
+
+  res.file <- paste0("results2/simu_", i, ".rds")
+  if (file.exists(res.file)) next
+
+  params <- params.grid2[i, ]
+  par.causal <- params[["par.causal"]][[1]]
+
+  # Simulate phenotypes
+  pheno.all <- get_pheno(
+    G,    
+    h2 = params[["par.h2"]], 
+    M = as.integer(par.causal[1]), 
+    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
+    effects.dist = params[["par.dist"]], 
+    model = params[["par.model"]], 
+    K = 0.3                 
+  )
+
+  # Split in training/test sets
+  ind.train <- sort(sample(n, size = n.train))
+  ind.test <- setdiff(1:n, ind.train)
+
+  # Get results from all methods
+  res <- bind_rows(
+    PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test),
+    logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple"),
+    logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple")
+  )
+  print(res)
+  params[["res"]] <- list(res)
+  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
+}
+```

### 2017-10-24 paper/draft.tex
save

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english, 12pt]{article}
\usepackage{times}
%\usepackage{algorithm2e}
\usepackage{url}
\usepackage{bbm}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
\usepackage{rotating}
\usepackage{color}
\usepackage{graphicx}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{bbm}

\linenumbers
\doublespacing
%\usepackage[authoryear]{natbib}
\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}

%Pour les rajouts
\usepackage{color}
\definecolor{trustcolor}{rgb}{0,0,1}

\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{graphicx}
\graphicspath{{../figures/}}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Bold symbol macro for standard LaTeX users
%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}

%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

\usepackage{babel}
\makeatother

\begin{document}

\title{Predicting complex diseases: performance and robustness}
\author{Florian Privé\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}

\date{~ }
\maketitle

\noindent$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Intégrative (C3BI), Institut Pasteur, Paris, France \\
\noindent$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.

\noindent$^\ast$To whom correspondence should be addressed.

\newpage
\abstract{
\textbf{Motivation:} \\
\textbf{Results:} \\
\textbf{Availability:}
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:}
}

\newpage
\section{Introduction}

Polygenic Risk Scores (PRSs) combine the information contained in many single-nucleotide polymorphisms (SNPs) in a score that should reflect the risk of developing diseases. PRSs have proven to be useful for different applications, such as .... 
Personalized medicine is one of the major applications of PRSs. Personalized medicine will use PRSs in screening campains in order to identify people at higher risk for a given disease. ...This could lead to prevention for these identified people...
Personalized medicine with also use PRSs for helping in the diagnostic of disease, either confirming a diagnostic or even early diagnostic. Yet, diagnostic based on PRSs won't be possible unless these PRSs shows a high discrimitative power between cases and controls [REF].
Many methods have been developed in order to maximize the predictive power of genotypes for diseases, or more generally phenotypes.
A commonly used technique, called P+T (which stands for ``Pruning + Thresholding'') or genetic profiling, is used to derive a PRS from results of Genome-Wide Association Study (GWAS). This technique only use summary statistics which make it very ..suitable.. and also very fast.
Linear Mixed-Models are also widely-used in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height [REF]. Yet, these models are not ..constructed.. for predicting a binary trait such as a disease and have proven to ..fail.. at such in another comparative study [REF GAD]. Moreover, these methods and their derivations are often computationally demanding and won't be usable for largest cohorts to data [REFS].
Statistical learning methods such as logistic regression, Support Vector Machine (SVM) or random forests were also used to derive PRSs for complex human disease by jointly estimating SNP effects.

We recently developed two R packages, bigstatsr and bigsnpr, for efficient management and analysis of large-scale genome-wide data. This include efficient functions for computing penalized linear and logistic regressions on huge datasets. In this paper, we present a comprehensive analysis of the P+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has given exceptionally good results in its corresponding paper [REF BOTTA]. Note that SVM is expected to give similar results to logistic regression [REF GAD], and therefore isn't added to the comparison.
For the P+T model, we compare different threshold of inclusion of SNPs, as it has been a burning question for some years.
For the logistic regression, we include two novel approaches. First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for ``choosing'' the optimal threshold of inclusion of SNPs. We also show how easy it is to tweak the method in order to capture not only linear effects, but also recessive/dominant effects.
Finally, we quickly discard T-Trees as being a suitable method for predicting disease status based on genotypes.
In order to make our comparison as comprehensive as possible, we compare different architectures of disease (number, size and location of causal effects and heritability) with different model of generating liability ..scores.., one with only linear effects, and one which combine linear, recessive, dominant and interaction effects.

We show that penalized logistic regression consistently performs better than the P+T method whereas predictive performance of the P+T method is very sensitive to the threshold of inclusion of SNPs [MONTRER UN GRAPH EN FCT D'INCLUSION].

\section{Methods}

\subsection{Simulations}

For simulations, we use real genotypes of European individuals from a case/control celiac disease cohort [REF]. 
This dataset has been quality-controled and imputed in another study [REF MOI].
To get rid of the structure induced by the celiac disease, we keep only controls from this cohort. Moreover, to get rid of population structure, we keep only people from UK and further subset these British people based on deviation of Robust Mahalanobis distance on Principal Components (see ..SupMat..), leaving 7102 individuals genotypes on ..28x,xxx.. SNPs.

To simulate phenotypes, we use a the Liability Threshold Model (LTM). We vary different parameters including the number of causal variants and their location (30, 300 or 3000 anywhere in the genome and 30 in the HLA region).

\subsection{Compared methods}

\section{Results}

\newpage
\section*{Acknowledgements}

Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d?avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).

\vspace*{-12pt}

\bibliographystyle{natbib}
\bibliography{document}

\newpage
\section*{Supplementary Data}

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}

\end{document}

### 2017-10-24 results1.R

@@ -1,0 +1,44 @@
+
+library(tidyverse)
+
+myggplot <- function(..., coeff = 1) {
+  bigstatsr:::MY_THEME(ggplot(...), coeff = coeff)
+} 
+
+results <- list.files("../thesis/paper2-PRS/results1/", full.names = TRUE) %>%
+  map_dfr(~readRDS(.x)) %>%
+  as_tibble() %>%
+  mutate(
+    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
+                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
+    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2]))
+  )
+  
+print(results, n = Inf)
+
+results %>%
+  filter(method == "T-Trees") %>%
+  pull(timing)
+
+results %>%
+  group_by_at(c(vars(starts_with("par")), "method")) %>%
+  summarise_at(c("timing", "nb.preds", "AUC"), mean)
+
+results %>%
+  filter(!grepl("PRS", method)) %>%
+  myggplot() +
+  geom_boxplot(aes(method, timing, color = par.dist, fill = par.dist), alpha = 0.3) + 
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  facet_grid(par.model ~ par.causal) +
+  scale_y_continuous(breaks = 0:10 * 2000, minor_breaks = NULL) + 
+  theme(strip.text.x = element_text(size = rel(2)),
+        strip.text.y = element_text(size = rel(2))) 
+
+results %>%
+  myggplot() +
+  geom_boxplot(aes(method, AUC, color = par.dist, fill = par.dist), alpha = 0.3) + 
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  facet_grid(par.model ~ par.causal) +
+  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = NULL) + 
+  theme(strip.text.x = element_text(size = rel(2)),
+        strip.text.y = element_text(size = rel(2))) 

### 2017-10-25 paper/draft.tex
save draft

@@
 
+\usepackage{dsfont}
 \usepackage{adjustbox}
@@
 \textbf{Results:} \\
-\textbf{Availability:}
+\textbf{Availability:} \\
 \textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
@@
 Polygenic Risk Scores (PRSs) combine the information contained in many single-nucleotide polymorphisms (SNPs) in a score that should reflect the risk of developing diseases.
-PRSs have proven to be useful for different applications, such as ....
+PRSs have proven to be useful for different applications, such as finding a common genetic contribution between two diseases [@Purcell2009].
 Personalized medicine is one of the major applications of PRSs.
@@
 Moreover, to get rid of population structure, we keep only people from UK and further subset these British people based on deviation of Robust Mahalanobis distance on Principal Components (see ..SupMat..), leaving 7102 individuals genotypes on ..28x,xxx..
-SNPs.
+SNPs [ADD DIM TO RMD] with minimal population structure (see ..SupMat..) because this....
 
-To simulate phenotypes, we use a the Liability Threshold Model (LTM).
-We vary different parameters including the number of causal variants and their location (30, 300 or 3000 anywhere in the genome and 30 in the HLA region).
+To simulate phenotypes, we use a the Liability Threshold Model (LTM) with a prevalence of 30\%.
+We vary different parameters: the number of causal variants and their location (30, 300 or 3000 anywhere in the genome and 30 only in the HLA region), the heritability (0.5 or 0.8), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part..
+of the phenotypes (one with only linear effects, and one which combine linear, recessive, dominant and interaction effects).
+This simulation is broader yet similar to the one used in [REF STEPHENS].
+
+For the ``simple model'', we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights ($w_j \neq 0$ if and only if SNP $j$ is causal), $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs).
+\\
+For the ``fancy model'', we separate the causal SNPs in three equal sets $E_\text{causal}^{(1)}$, $E_\text{causal}^{(2)}$ 
+and $E_\text{causal}^{(3)}$ ($E_\text{causal}^{(3)}$ is further separated in two equal sets, $E_\text{causal}^{(3.1)}$ and $E_\text{causal}^{(3.2)}$) 
+and we compute $$y_i = \underbrace{\sum_{j\in E_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in E_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{recessive/dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|E_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} \cdot G_{i,j_2}}}_\text{interaction}$$ where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $E_\text{causal}^{(q)}=\left\{e_k^{(q)}, k \in \left\{1, \ldots, \left|E_\text{causal}^{(q)}\right|\right\}\right\}$.
+Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still display a marginal effect.
+\\
+For both models, the $y_i$ are then standardized such that they have a variance equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$ where $\epsilon \sim N(0, 1 - h^2)$.
+
+This results in 32 combinations of parameters and we run 100 simulations for each of these combinations (5 times more than in [REF STEPHENS]), resulting of a total of 3200 simulations.
+For each of these 3200 simulations, we randomly re-sample a training set of 6000 individuals, with the corresponding test set composed of the remaining individuals.
+
+\subsection{Predictive performance measures}
 
 \subsection{Compared methods}
+
+In this study, we compare three different types of methods: the P+T method, the T-Trees method and a penalized logistic regression.
+
+The P+T (Pruning + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWASs) that are commonly called summary statistics.
+For the P+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part).
+GWAS was performed using logistic regressions, with 10 first Principal Components (PCs) as covariates in the models.
+The SNPs are first clumped (P) so that there remains only loci that are weakly correlated with each other. 
+Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined). 
+Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$S_i = \sum_{j \in E_\text{clumping}} \mathds{1}\{p_j < T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS and $T$ is a threshold to be determined.
+In this study, we report scores for a clumping threshold at $r^2 > 0.2$ with regions of 500kb.
+[threshold in results?]
+[DIRE CA??:] All these steps are implemented in R packages bigstatsr and bigsnpr [REF MOI].
+
+T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests [REF 2001, BReiman] that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data [REF BOTTA].
+In a previous study on WTCCC data [REF], it achieved too-good-to-be-true results.
+Since this method is significantly different from the other tested here, we decided to include it in this comparison in order to assess whether or not it could give superior predictive performance.
+We use the same parameters as reported in Table 4 of the T-Trees paper, yet, we use only 100 trees (instead of 1000) because it showed only a very subtle increase of AUC while taking way too much time (e.g.
+AUC of 81.5\% instead of 81\%, data not shown).
+
+The last method we compare is penalized logistic regression.
 

### 2017-10-25 paper/methods.Rmd
save draft

---
# title: "Untitled"
# author: "Florian Privé"
# date: "2 février 2017"
output:
  # html_document: default
  word_document: default
bibliography: ../../articles/library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center',
                      out.width = 600, out.height = 480)
```

# Methods

## Models in statistical learning

####TODO: Define n, m, $x_i$ and $y_i$

In statistical learning, most basic model consists in the minimization of a function of the following form:
$$L(\beta_0,\beta) + \lambda P_\alpha(\beta)$$
where $L$ is called the "loss function" and $P_{\alpha}$ is the regularization term (usually, the intercept term is not penalized). 
Let us denote by $s_i=S_{\beta_0,\beta}(x_i)=\beta_0+x_i^T\beta$ the score function associated with the $i$-th individual and weights $\beta_0$ and $\beta$. 
Standard loss functions are:

- Gaussian loss (or mean squared error): $L_\text{gaussian}(\beta_0,\beta)=\frac{1}{n}\sum_{i=1}^n(y_i-s_i)^2$,
- Logistic loss: $L_\text{logistic}(\beta_0,\beta)=\frac{1}{n}\sum_{i=1}^n \log(1+e^{-y_i s_i})$,
- Hinge loss (SVM): $L_\text{hinge}(\beta_0,\beta)=\frac{1}{n}\sum_{i=1}^n \max(0, 1-y_i s_i)$.

```{r, fig.cap=""}
knitr::include_graphics("Images/loss-functions.png", dpi = 500)
```

Different regularizations can be used to prevent overfitting:

- L2-regularization (ridge, @Hoerl1970): $P_{0}(\beta)=\frac{1}{2}\|\beta\|_2^2$. The ridge penalty shrink coefficients and is ideal if there are many predictors drawn from a Gaussian distribution. 
- L1-regularization (lasso, @Tibshirani1996): $P_{1}(\beta)=\|\beta\|_1$. The Lasso penalty will force some of the coefficients to be exactly zero. This can be used as a means of variable selection, leading to sparse (and therefore more interpretable) models.
- L1- and L2-regularization (elastic-net, @Zou2005): $P_{\alpha}(\beta)=(1-\alpha)P_{0}(\beta) + \alpha P_{1}(\beta)$. This is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation, or any situation where there are many correlated predictors (@Friedman2010).

## State-of-the-art polygenic models

Two models are well used in the litterature:

## Hyper-parameter tuning

For tuning hyper-parameters, we used 10-fold cross-validation.

```{r, fig.cap="Illustration of an inner 10-fold cross-validation."}
knitr::include_graphics("https://sebastianraschka.com/images/faq/evaluate-a-model/k-fold.png")
```

# References

### 2017-10-26 paper/draft.tex
par on logit

@@
 \graphicspath{{../figures/}}
+\DeclareMathOperator*{\argmin}{\arg\!\min}
 
@@
 The last method we compare is penalized logistic regression.
+We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is the genotypes and covariables (e.g.
+PCs), $y$ is disease status we want to predict and $\lambda$ is a regularization parameter that need to be determined.
+Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (ridge, @Hoerl1970) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$),
+ the L1-regularization (lasso, @Tibshirani1996) forces some of the coefficients to be exactly zero and can be used as a means of variable selection, leading to sparse (and therefore more interpretable) models  (corresponds to $\alpha = 1$),
+the L1- and L2-regularization (elastic-net, @Zou2005) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation, or any situation where there are many correlated predictors (corresponds to $0 < \alpha < 1$) (@Friedman2010).
+In this study, we always use $\alpha = 0.5$, without trying to tune the value of this hyper-parameter $\alpha$ (e.g.
+by grid-search and cross-validation).
+
+To fit this penalized logistic regression, we use a very efficient algorithms based on (@Friedman2010, ...2012 and biglasso...), also implemented in our package bigstatsr.
+This type of algorithms build predictions for many values of $\lambda$ (typically 100).
+To get an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA).
+First, this method separates the training set in $K$ folds (e.g.
+10).
+Secondly, in turn, each fold is considered as an inner validation set and the others ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes a criterion (e.g.
+AUC) is determined and the vector of coefficients corresponding to the previous vector of scores is chosen.
+Finally, the $K$ resulting vectors of coefficients are then combined into one vector (e.g.
+using the geometric median).
+Because of L1-regularization, this vector of coefficients is typically very sparse and can be used to make a PRS based on a linear combination of allele counts.
 

### 2017-10-26 4-results.Rmd
save

---
title: "Results"
author: "Florian Privé"
date: "October 26, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", fig.asp = 0.7)
options(width = 110)
```

## Useful functions

```{r, message=FALSE}
library(tidyverse)
```

```{r}
myggplot <- function(..., coeff = 1) {
  bigstatsr:::MY_THEME(ggplot(...), coeff = coeff)
} 
```

```{r}
plot_results <- function(results, y, ylab = y) {

  dist <- "Distribution\nof effects"

  myggplot(results) +
    geom_boxplot(aes_string("method", y, color = "par.dist", 
                            fill = "par.dist"), alpha = 0.3) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    facet_grid(par.model ~ par.causal) +
    theme(strip.text.x = element_text(size = rel(2)),
          strip.text.y = element_text(size = rel(2))) +
    labs(x = "Method", y = ylab, fill = dist, color = dist)
}
```

## Results with T-Trees

```{r}
top10 <- 1:110
top20 <- 1:220

# Put all results in a single tibble
results <- list.files("results1", full.names = TRUE) %>%
  map_dfr(~readRDS(.x)) %>%
  as_tibble() %>%
  mutate(
    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2])),
    percCases10 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[top10], 2])),
    percCases20 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[top20], 2]))
  )
```

```{r}
results %>%
  filter(method %in% c("T-Trees", "logit-simple")) %>%
  group_by_at(c(vars(starts_with("par")), "method")) %>%
  summarise_at(c("timing", "nb.preds", "AUC", "percCases10", "percCases20"), mean) %>%
  print(n = Inf)
```

```{r, fig.width=16, out.width="95%", dev="svg"}
ttrees_vs_logit <- filter(results, method %in% c("T-Trees", "logit-simple"))

p_list <- list(
  plot_results(ttrees_vs_logit, "timing", "Timing (in seconds)") +
    scale_y_continuous(breaks = 0:10 * 2000, minor_breaks = NULL),
  plot_results(ttrees_vs_logit, "nb.preds", "Number of predictors (log-scale)") +
    scale_y_log10(breaks = c(10^(0:7), 3 * 10^(0:7)), minor_breaks = NULL,
                  labels = scales::comma_format()),
  plot_results(ttrees_vs_logit, "AUC") +
    scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10),
  plot_results(ttrees_vs_logit, "percCases10", "Percentage of cases in top 10%") +
    scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10)
)

lapply(p_list, function(p) p + theme(legend.position = "none")) %>%
  cowplot::plot_grid(plotlist = ., ncol = 2, align = "hv", scale = 0.9,
                     labels = LETTERS[1:4], label_size = 25) %>%
  cowplot::plot_grid(cowplot::get_legend(p_list[[1]]),
                     rel_widths = c(1, 0.15))
```

```{r}
ggsave("figures/ttrees.pdf", scale = 1/90, width = 1580, height = 1070)
```

## Results without T-Trees

### 2017-10-26 paper/draft.tex
save

@@
 \DeclareMathOperator*{\argmin}{\arg\!\min}
+
+\let\tabbeg\tabular
+\let\tabend\endtabular
+\renewenvironment{tabular}{\begin{adjustbox}{max width=\textwidth}\tabbeg}{\tabend\end{adjustbox}}
 
@@
 
-\subsection{Simulations}
+\subsection{Simulations} \label{sec:simus}
 
@@
 \subsection{Predictive performance measures}
+
+In this study, we use two different measures of predictive accuracy. 
+First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC). In the case of our study, the AUC is the probability that a PRS of a case is greater than one of a control.
+This measure indicates how well we can distinguish between cases and controls using PRSs.
+As a second measure, we report the percentage of cases in the 10\% largest PRSs [UTILISER 20\%?].
+This measure indicates how well we can identify people at higher risk from a higher PRS in screening campaigns.
+
+Note that we also report the timing of the main computations and the number of SNPs used for the prediction.
 
@@
 
+\subsection{Reproduciblity}
+
+All the code used in this paper along with results,  such as execution times and figures,  are available as HTML R notebooks in the Supplementary Materials.
+
 \section{Results}
+
+\subsection{Overview and naming}
+
+\subsection{T-Trees}
+
+First, we ran only 5 simulations for each combination of parameters (see section \ref{sec:simus}), excepted for the heritability that we fixed at $0.8$.
+The goal was to quickly discard the T-Trees method as being competitive with the other methods. 
+For example, compared to our penalized logistic regression, T-Trees perform worse for both predictive measures, the AUC and the percentage of cases in highest scores.
+Moreover, T-Trees takes longer to run and makes more complex predictive models because it uses more SNPs in the models and have non-linear effects (Figure \ref{fig:ttrees}).
+So, we decided to discard this method for the rest of the simulations.
+
+\subsection{Main simulations}
+
+For the P+T method, we report three different scores of prediction: one including all the SNPs (after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximize the AUC for these two thresholds and a sequence of 100 values from $10^{-0.1}$ to $10^{-100}$ equally spaced on the log-log-scale (Table \ref{tab:thr}).
+Note that, normally, the optimal threshold would have to be learned on the training set only.
+We consider this reported maximum AUC as an upper bound of the AUC for the P+T method.
+
+[Also return p-values?]
+
+[TIMING: PRS -> 0; logit -> 500-600 with small increase when more causal variants, logit-triple x3]
+
+[AUC: decrease with M, PRS-all never good, logit-triple better for fancy and almost as good for simple]
+
+[\% Cases: Linear with AUC, Attention perc cases is == 30%]
+
+\section{Discussion}
 
@@
 
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{ttrees}}
+\caption{Results of T-Trees vs penalized logistic regression.
+\textbf{A.} Timing (in seconds).
+\textbf{B.} Number of predictors of the model.
+\textbf{C.} AUC.
+\textbf{D.} Percentage of cases in the 10\% largest scores.}\label{fig:ttrees}
+\end{figure}
+
+% latex table generated in R 3.4.1 by xtable 1.8-2 package
+% Thu Oct 26 16:46:26 2017
+\begin{table}[ht]
+\centering
+\begin{tabular}{ccccccccccccccc}
+  \hline
+1.00e+00 & 7.22e-01 & 5.87e-01 & 4.20e-01 & 2.43e-01 & 1.00e-01 & 2.35e-02 & 2.21e-03 & 4.69e-05 & 8.81e-08 & 3.18e-12 & 1.83e-19 & 2.89e-31 & 1.70e-50 & 7.71e-82 \\ 
+  5.00e-08 & 7.05e-01 & 5.65e-01 & 3.95e-01 & 2.20e-01 & 8.47e-02 & 1.79e-02 & 1.42e-03 & 2.28e-05 & 2.73e-08 & 4.69e-13 & 8.08e-21 & 1.80e-33 & 4.30e-54 & 1.06e-87 \\ 
+  7.94e-01 & 6.87e-01 & 5.42e-01 & 3.69e-01 & 1.97e-01 & 7.08e-02 & 1.34e-02 & 8.83e-04 & 1.05e-05 & 7.74e-09 & 6.03e-14 & 2.86e-22 & 7.73e-36 & 5.97e-58 & 5.49e-94 \\ 
+  7.81e-01 & 6.69e-01 & 5.19e-01 & 3.43e-01 & 1.75e-01 & 5.85e-02 & 9.79e-03 & 5.31e-04 & 4.61e-06 & 2.01e-09 & 6.69e-15 & 7.92e-24 & 2.24e-38 & 4.37e-62 & 1.00e-100 \\ 
+  7.67e-01 & 6.50e-01 & 4.95e-01 & 3.18e-01 & 1.54e-01 & 4.76e-02 & 7.01e-03 & 3.08e-04 & 1.90e-06 & 4.72e-10 & 6.32e-16 & 1.70e-25 & 4.26e-41 & 1.61e-66 &  \\ 
+  7.53e-01 & 6.30e-01 & 4.70e-01 & 2.93e-01 & 1.35e-01 & 3.82e-02 & 4.90e-03 & 1.72e-04 & 7.31e-07 & 1.00e-10 & 5.04e-17 & 2.75e-27 & 5.16e-44 & 2.83e-71 &  \\ 
+  7.38e-01 & 6.09e-01 & 4.46e-01 & 2.68e-01 & 1.17e-01 & 3.02e-02 & 3.33e-03 & 9.18e-05 & 2.63e-07 & 1.89e-11 & 3.35e-18 & 3.31e-29 & 3.84e-47 & 2.26e-76 &  \\ 
+   \hline
+\end{tabular}
+\caption{The 102 thresholds used in the P+T method for this study.}
+\label{tab:thr}
+\end{table}
+
 \end{document}

### 2017-10-26 results2.R

@@ -10,6 +10,7 @@ top20 <- 1:220
 
 results <- list.files("../thesis/paper2-PRS/results1/", full.names = TRUE) %>%
   map_dfr(~readRDS(.x)) %>%
+  filter(method != "T-Trees") %>%
   as_tibble() %>%
   mutate(
     par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
@@ -21,13 +22,10 @@ results <- list.files("../thesis/paper2-PRS/results1/", full.names = TRUE) %>%
 
 print(results, n = Inf)
 
-results %>%
-  filter(method == "T-Trees") %>%
-  pull(timing)
 
 results %>%
   group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise_at(c("timing", "nb.preds", "AUC", "percCases"), mean)
+  summarise_at(c("timing", "nb.preds", "AUC", "percCases10"), mean)
 
 plot_results <- function(results, y, ylab = y) {
   
@@ -44,41 +42,41 @@ plot_results <- function(results, y, ylab = y) {
 }
 
 results %>%
-  filter(!grepl("PRS", method)) %>%
+  filter(!grepl("PRS", method), par.h2 == 0.8) %>%
   plot_results("timing", "Timing") +
-  scale_y_continuous(breaks = 0:10 * 2000, minor_breaks = NULL) 
+  scale_y_continuous(breaks = 0:50 * 100, minor_breaks = NULL) 
 ggsave("timing.pdf", scale = 1/90, width = 1200, height = 900)
 
-plot_results(results, "AUC") +
+results %>%
+  filter(par.h2 == 0.8) %>%
+  plot_results("AUC") +
   scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10)
 ggsave("AUC.pdf", scale = 1/90, width = 1200, height = 900)
 
 ggsave("perc-cases.pdf", scale = 1/90, width = 1200, height = 900)
 
-plot_results(results, "percCases20", "Percentage of cases in top 20%") +
+results %>%
+  filter(par.h2 == 0.8) %>%
+  plot_results("percCases20", "Percentage of cases in top 20%") +
   scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10)
 
-plot_results(results, "nb.preds", "Number of predictors") +
+results %>%
[…]

### 2017-10-26 paper/draft.tex
save

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \newpage
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Methods}
@@
 For each of these 3200 simulations, we randomly re-sample a training set of 6000 individuals, with the corresponding test set composed of the remaining individuals.
+
+[MONTRER UN MANHATTAN PLOT POUR JUSTIFIER 30 IN HLA]
 
@@
 
+[ALSO TRIPLE LOGIT]
+
 To fit this penalized logistic regression, we use a very efficient algorithms based on (@Friedman2010, ...2012 and biglasso...), also implemented in our package bigstatsr.
@@
 All the code used in this paper along with results,  such as execution times and figures,  are available as HTML R notebooks in the Supplementary Materials.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-[TIMING: PRS -> 0; logit -> 500-600 with small increase when more causal variants, logit-triple x3]
+First, we show that AUC is highly correlated with the percentage of cases in either the 10\% or 20\% largest PRSs (Figure \ref{fig:AUC-corr}).
+So, we report only predictive performance in terms of AUC in following results.
 
-[AUC: decrease with M, PRS-all never good, logit-triple better for fancy and almost as good for simple]
+As for the comparison of our methods with the P+T method, we report AUCs greater than 95\%, which is not obtained by the upper bound of the P+T method in any scenario.
+Our penalized regression method consistently provides better predictive performance than the P+T method.
+Our method is parallelized and requires only less than 10 minutes to run (with 6 physical cores).
+Moreover, it results in sparse models (...).
+The P+T runs in only a few seconds because we don't consider the GWAS computation as part of the execution time of the method because this is generally executed by others and then the resulting summary statistics are used to construct a PRS.
 
-[\% Cases: Linear with AUC, Attention perc cases is == 30%]
+As for the comparison between our two penalized logistic regression, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects.
+Yet, the ``triple'' solution takes approximately 3 more times to run.
+
+As for general results, we report a decrease of AUC when the number of causal variants $M$ is increasing.
+
+[TROUVER DES AUTRES: \% of decrease for h2 0.8 -> 0.5? Pas bon d'utliser tous les SNPs]
+
+[REPLICATE INDIVIDUALS TO SHOW IF CAN BE USED ON LARGER DATA]
 
 \section{Discussion}
+
+In this comparative study, we proposed other methods as a replacement to the widely used P+T method. 
+We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really good performance as reported in previous studies [REF GAD + AUTRE?], better than with the P+T method.
+We are well aware that the P+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions.
+Yet, in the case of prediction of disease to be used for clinical interventions, best possible methods should be used, even if it has to be used by large consortia only.
+
+One does not simply identify high-risk individuals (Figure \ref{fig:meme}).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d?avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).
+Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).
+We are also grateful to Félix Balazard for useful discussions about T-Trees.
 
@@
 \bibliography{document}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{ttrees}}
-\caption{Results of T-Trees vs penalized logistic regression.
-\textbf{A.} Timing (in seconds).
-\textbf{B.} Number of predictors of the model.
-\textbf{C.} AUC.
-\textbf{D.} Percentage of cases in the 10\% largest scores.}\label{fig:ttrees}
-\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{ttrees}}
+\caption{Results of T-Trees vs penalized logistic regression.
+\textbf{A.} Timing (in seconds).
+\textbf{B.} Number of predictors of the model.
+\textbf{C.} AUC.
+\textbf{D.} Percentage of cases in the 10\% largest scores.}\label{fig:ttrees}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{AUC-corr}}
+\caption{Percentage of cases in the 2 highest deciles of PRSs as a function of AUC.}\label{fig:AUC-corr}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{meme}}
+\caption{Meme to use on Twitter when promoting the preprint?}\label{fig:meme}
+\end{figure}
+
 \end{document}

### 2017-10-27 4-results.Rmd
save

@@
 ## Results without T-Trees
+
+### Correlation between predictive performance measures
+
+```{r, fig.width=14, out.width="90%", dev="svg"}
+cowplot::plot_grid(
+  results %>%
+    filter(par.h2 == 0.8) %>%
+    myggplot(aes(AUC, percCases10, color = par.dist)) +
+    geom_point() +
+    geom_smooth(method = "lm") +
+    theme(legend.position = c(0.8, 0.2)) +
+    labs(y = "Percentage of cases in top 10%", 
+         color = "Distribution\nof effects"),
+  results %>%
+    filter(par.h2 == 0.8) %>%
+    myggplot(aes(AUC, percCases20, color = par.dist)) +
+    geom_point() +
+    geom_smooth(method = "lm") +
+    theme(legend.position = c(0.8, 0.2)) +
+    labs(y = "Percentage of cases in top 20%", 
+         color = "Distribution\nof effects"),
+  results %>%
+    filter(par.h2 == 0.8) %>%
+    myggplot(aes(AUC, percCases10, color = par.model)) +
+    geom_point() +
+    geom_smooth(method = "lm") +
+    theme(legend.position = c(0.8, 0.2)) +
+    scale_colour_brewer(palette = "Set1") +
+    labs(y = "Percentage of cases in top 10%", 
+         color = "Model"),
+  results %>%
+    filter(par.h2 == 0.8) %>%
+    myggplot(aes(AUC, percCases20, color = par.model)) +
+    geom_point() +
+    geom_smooth(method = "lm") +
+    theme(legend.position = c(0.8, 0.2)) +
+    scale_colour_brewer(palette = "Set1") +
+    labs(y = "Percentage of cases in top 20%", 
+         color = "Model"),
+  labels = LETTERS[1:4], align = "hv", label_size = 25, scale = 0.95
+)
+```
+
+```{r}
+ggsave("figures/AUC-corr.pdf", scale = 1/90, width = 1300, height = 950)
+```

### 2017-10-27 paper/draft.tex
save

@@
 
-\title{Predicting complex diseases: performance and robustness}
-\author{Florian Privé\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
+\title{Predicting complex diseases: performance, robustness and ..utility..(impact?)}
+\author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
 
@@
 
-\noindent$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
-\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Intégrative (C3BI), Institut Pasteur, Paris, France \\
+\noindent$^{\text{\sf 1}}$Universit\'e Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
+\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France \\
 \noindent$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.
@@
 Polygenic Risk Scores (PRSs) combine the information contained in many single-nucleotide polymorphisms (SNPs) in a score that should reflect the risk of developing diseases.
-PRSs have proven to be useful for different applications, such as finding a common genetic contribution between two diseases [@Purcell2009].
-Personalized medicine is one of the major applications of PRSs.
-Personalized medicine will use PRSs in screening campains in order to identify people at higher risk for a given disease.
-...This could lead to prevention for these identified people...
+PRSs have proven to be useful for different applications, such as finding a common genetic contribution between two diseases \cite[]{purcell2009common}.
+Personalized medicine is another major application of PRSs.
+Personalized medicine will use PRSs in screening campains in order to identify people at higher risk for a given disease and ...
+could lead to prevention for these identified people...
+[REFS?].
 Personalized medicine with also use PRSs for helping in the diagnostic of disease, either confirming a diagnostic or even early diagnostic.
-Yet, diagnostic based on PRSs won't be possible unless these PRSs shows a high discrimitative power between cases and controls [REF].
-Many methods have been developed in order to maximize the predictive power of genotypes for diseases, or more generally phenotypes.
-A commonly used technique, called P+T (which stands for ``Pruning + Thresholding'') or genetic profiling, is used to derive a PRS from results of Genome-Wide Association Study (GWAS).
+Yet, diagnostic based on PRSs won't be possible unless these PRSs shows a high discrimitative power between cases and controls.
+For example, for screening high-risk individuals and for presymtomatic diagnosis of the general population, it is suggested that the AUC must be  greater than respectively 75\% and 99\% \cite[]{janssens2007impact}.
+
+Many methods have been developed in order to maximize the predictive power of genotypes for diseases, or phenotypes in general.
+A commonly used technique, called P+T (which stands for ``Pruning + Thresholding'') or genetic profiling ..or just PRS.., is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,golan2014effective,purcell2009common,schizophrenia2014biological}.
 This technique only use summary statistics which make it very ..suitable..
 and also very fast.
-Linear Mixed-Models are also widely-used in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height [REF].
+Linear Mixed-Models are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}.
 Yet, these models are not ..constructed..
-for predicting a binary trait such as a disease and have proven to ..fail..
-at such in another comparative study [REF GAD].
-Moreover, these methods and their derivations are often computationally demanding and won't be usable for largest cohorts to data [REFS].
-Statistical learning methods such as logistic regression, Support Vector Machine (SVM) or random forests were also used to derive PRSs for complex human disease by jointly estimating SNP effects.
+for predicting a binary trait such as a disease status and have proven to ..fail..
+at such in another comparative study \cite[]{abraham2013performance}.
+Moreover, these methods and their ..derivations..
+are often computationally demanding, both in terms of memory and time requirements, which makes them unlikely to be used for prediction on large cohorts \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
+Statistical learning methods have also been used to derive PRSs for complex human disease by jointly estimating SNP effects.
+Such methods include logistic regression, Support Vector Machine (SVM) or random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2013walk,botta2014exploiting}.
 
-We recently developed two R packages, bigstatsr and bigsnpr, for efficient management and analysis of large-scale genome-wide data.
-This include efficient functions for computing penalized linear and logistic regressions on huge datasets.
-In this paper, we present a comprehensive analysis of the P+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has given exceptionally good results in its corresponding paper [REF BOTTA].
-Note that SVM is expected to give similar results to logistic regression [REF GAD], and therefore isn't added to the comparison.
-For the P+T model, we compare different threshold of inclusion of SNPs, as it has been a burning question for some years.
+We recently developed two R packages, bigstatsr and bigsnpr, for efficient management and analysis of large-scale genome-wide data \cite[]{prive2017efficient}.
+Package bigstatsr includes efficient functions for computing penalized linear and logistic regressions on huge datasets.
+In this paper, we present a comprehensive comparative study of the P+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in \cite{botta2014exploiting}.
+Note that SVM is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp} and therefore isn't added to the comparison.
+For the P+T model, we compare different threshold of inclusion of SNPs, since choosing this threshold has been a burning question for some years [REFS].
 For the logistic regression, we include two novel approaches. 
 First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for ``choosing'' the optimal threshold of inclusion of SNPs. 
-We also show how easy it is to tweak the method in order to capture not only linear effects, but also recessive/dominant effects.
-Finally, we quickly discard T-Trees as being a suitable method for predicting disease status based on genotypes.
+We also show how easy it is to tweak the method in order to capture not only linear effects, but also recessive and dominant effects.
+
 In order to make our comparison as comprehensive as possible, we compare different architectures of disease (number, size and location of causal effects and heritability) with different model of generating liability ..scores.., one with only linear effects, and one which combine linear, recessive, dominant and interaction effects.
+We quickly discard T-Trees as being a suitable method for predicting disease status based on genotypes.
+Then, we show that penalized logistic regression consistently performs better than the P+T method whereas predictive performance of the P+T method is very sensitive to the threshold of inclusion of SNPs [MONTRER UN GRAPH EN FCT D'INCLUSION].
+Such comparative study... [PARLER QUE LES COMPARAISONS ONT DEJA ETE FAITES] [REFS]
 
-We show that penalized logistic regression consistently performs better than the P+T method whereas predictive performance of the P+T method is very sensitive to the threshold of inclusion of SNPs [MONTRER UN GRAPH EN FCT D'INCLUSION].
+[PARLER DU PROBLEME DE POP?]
+[PARLER DE L'ASPESCT CLINIQUE PLUS EN DETAIL?]
 
@@
 
-\subsection{Simulations} \label{sec:simus}
+\subsection{Genotype data}
 
-For simulations, we use real genotypes of European individuals from a case/control celiac disease cohort [REF]. 
-This dataset has been quality-controled and imputed in another study [REF MOI].
-To get rid of the structure induced by the celiac disease, we keep only controls from this cohort.
-Moreover, to get rid of population structure, we keep only people from UK and further subset these British people based on deviation of Robust Mahalanobis distance on Principal Components (see ..SupMat..), leaving 7102 individuals genotypes on ..28x,xxx..
+For simulations, we use real genotypes of European individuals from a case/control celiac disease cohort \cite[]{dubois2010multiple}. 
+This dataset has been quality-controled and imputed in another study \cite[](prive2017efficient).
+To get rid of the structure induced by the celiac disease status, we keep only controls from this cohort.
+Moreover, to get rid of population structure, we keep only people from UK and further subset these British people based on deviation of robust Mahalanobis distance on Principal Components (see ..SupMat..), leaving 7102 individuals genotyped on ..28x,xxx..
 SNPs [ADD DIM TO RMD] with minimal population structure (see ..SupMat..) because this....
+
+\subsection{Simulations of phenotypes} \label{sec:simus}
 
@@
 PCs), $y$ is disease status we want to predict and $\lambda$ is a regularization parameter that need to be determined.
-Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (ridge, @Hoerl1970) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$),
- the L1-regularization (lasso, @Tibshirani1996) forces some of the coefficients to be exactly zero and can be used as a means of variable selection, leading to sparse (and therefore more interpretable) models  (corresponds to $\alpha = 1$),
-the L1- and L2-regularization (elastic-net, @Zou2005) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation, or any situation where there are many correlated predictors (corresponds to $0 < \alpha < 1$) (@Friedman2010).
+Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (ridge, \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$),
+ the L1-regularization (lasso, \cite{tibshirani1996regression}) forces some of the coefficients to be exactly zero and can be used as a means of variable selection, leading to sparse (and therefore more interpretable) models  (corresponds to $\alpha = 1$),
+the L1- and L2-regularization (elastic-net, \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation, or any situation where there are many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
 In this study, we always use $\alpha = 0.5$, without trying to tune the value of this hyper-parameter $\alpha$ (e.g.
@@
 
-To fit this penalized logistic regression, we use a very efficient algorithms based on (@Friedman2010, ...2012 and biglasso...), also implemented in our package bigstatsr.
+To fit this penalized logistic regression, we use a very efficient algorithms based on \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient}, also implemented in our package bigstatsr.
 This type of algorithms build predictions for many values of $\lambda$ (typically 100).
@@
 
-First, we show that AUC is highly correlated with the percentage of cases in either the 10\% or 20\% largest PRSs (Figure \ref{fig:AUC-corr}).
+First, we show that AUC is highly correlated ($r^2 > 95\%$) with the percentage of cases in either the 10\% or 20\% largest PRSs (Figure \ref{fig:AUC-corr}).
 So, we report only predictive performance in terms of AUC in following results.
@@
 In this comparative study, we proposed other methods as a replacement to the widely used P+T method. 
-We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really good performance as reported in previous studies [REF GAD + AUTRE?], better than with the P+T method.
+We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really good performance as reported in previous studies [REF GAD + AUTRE?] \cite[]{abraham2013performance}, better than with the P+T method.
 We are well aware that the P+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions.
@@
 Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).
-We are also grateful to Félix Balazard for useful discussions about T-Trees.
+We are also grateful to F\'elix Balazard for useful discussions about T-Trees.
 
@@
 \bibliographystyle{natbib}
-\bibliography{document}
+\bibliography{refs}
 

### 2017-10-30 paper/draft.tex
typo
save

@@
 
-Polygenic Risk Scores (PRSs) combine the information contained in many single-nucleotide polymorphisms (SNPs) in a score that should reflect the risk of developing diseases.
+Polygenic Risk Scores (PRSs) combine the information contained in many single-nucleotide polymorphisms (SNPs) in a score that aims at reflecting the risk of developing a disease for which this score has been created.
 PRSs have proven to be useful for different applications, such as finding a common genetic contribution between two diseases \cite[]{purcell2009common}.
 Personalized medicine is another major application of PRSs.
-Personalized medicine will use PRSs in screening campains in order to identify people at higher risk for a given disease and ...
-could lead to prevention for these identified people...
-[REFS?].
-Personalized medicine with also use PRSs for helping in the diagnostic of disease, either confirming a diagnostic or even early diagnostic.
-Yet, diagnostic based on PRSs won't be possible unless these PRSs shows a high discrimitative power between cases and controls.
-For example, for screening high-risk individuals and for presymtomatic diagnosis of the general population, it is suggested that the AUC must be  greater than respectively 75\% and 99\% \cite[]{janssens2007impact}.
+Personalized medicine will use PRSs in screening campaigns in order to identify people at higher risk for a given disease.
+As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}.
+Moreover, screening based on sequencing seems to make individuals positively change their health behavior while not causing patient anxiety nor depression \cite[]{vassy2017impact}.
+Yet, for PRSs to be used for helping in the diagnostic of disease, either confirming a diagnostic or even early diagnostic, PRSs would have to show a high discriminative power between cases and controls.
+For example, for screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be  greater than respectively 75\% and 99\% \cite[]{janssens2007impact}.
 
 Many methods have been developed in order to maximize the predictive power of genotypes for diseases, or phenotypes in general.
-A commonly used technique, called P+T (which stands for ``Pruning + Thresholding'') or genetic profiling ..or just PRS.., is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,golan2014effective,purcell2009common,schizophrenia2014biological}.
+A commonly used technique, called P+T (which stands for ``Pruning + Thresholding'') or genetic profiling ..or even just PRS.., is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
 This technique only use summary statistics which make it very ..suitable..
 and also very fast.
-Linear Mixed-Models are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}.
+Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}.
 Yet, these models are not ..constructed..
@@
 Moreover, these methods and their ..derivations..
-are often computationally demanding, both in terms of memory and time requirements, which makes them unlikely to be used for prediction on large cohorts \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
+are often computationally demanding, both in terms of memory and time ..requirements.., which makes them unlikely to be used for prediction on large cohorts \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
 Statistical learning methods have also been used to derive PRSs for complex human disease by jointly estimating SNP effects.
-Such methods include logistic regression, Support Vector Machine (SVM) or random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2013walk,botta2014exploiting}.
+Such methods include logistic regression, Support Vector Machine (SVM) or random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2013walk,botta2014exploiting,wei2009disease}.
 
@@
 Note that SVM is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp} and therefore isn't added to the comparison.
-For the P+T model, we compare different threshold of inclusion of SNPs, since choosing this threshold has been a burning question for some years [REFS].
+For the P+T model, we compare different threshold of inclusion of SNPs.
 For the logistic regression, we include two novel approaches. 
-First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for ``choosing'' the optimal threshold of inclusion of SNPs. 
-We also show how easy it is to tweak the method in order to capture not only linear effects, but also recessive and dominant effects.
+First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for choosing the amount of regularization used, which directly affects the number of SNPs included in the model.
+We also show how easy it is to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.
 
-In order to make our comparison as comprehensive as possible, we compare different architectures of disease (number, size and location of causal effects and heritability) with different model of generating liability ..scores.., one with only linear effects, and one which combine linear, recessive, dominant and interaction effects.
-We quickly discard T-Trees as being a suitable method for predicting disease status based on genotypes.
-Then, we show that penalized logistic regression consistently performs better than the P+T method whereas predictive performance of the P+T method is very sensitive to the threshold of inclusion of SNPs [MONTRER UN GRAPH EN FCT D'INCLUSION].
-Such comparative study... [PARLER QUE LES COMPARAISONS ONT DEJA ETE FAITES] [REFS]
-
-[PARLER DU PROBLEME DE POP?]
-[PARLER DE L'ASPESCT CLINIQUE PLUS EN DETAIL?]
+There have already been comparative studies about predicting phenotypes based on genotypes for human data. 
+\citep{abraham2012sparsnp} showed that penalized logistic regression and penalized SVM performed better than standard P+T model and LMMs for a wide-range of human diseases.
+\cite{zhou2013polygenic} used simulations on real genotypes of Australian individuals to show that a hybrid of linear mixed models and sparse regression models perform better in a range of sparse to polygenic disease architectures.
+\cite{spiliopoulou2015genomic} investigated the effect of relatedness in prediction performance.
+\cite{ware2017heterogeneity} evaluated best practices for using the P+T method and found out, for example, that the optimal inclusion threshold is trait-dependent. 
+In this study, we particularly extend the work of \cite{abraham2012sparsnp} by using a simulation framework similar but richer than the one used in \cite{zhou2013polygenic}.
+In order to make our comparison as comprehensive as possible, we compare different architectures of disease (number, size and location of causal effects and heritability) with different model of generating liability ..scores.., one with only linear effects, and one which combine linear, dominant and interaction effects.
+We quickly discard T-Trees as being the best method for predicting disease status based on genotypes.
+Then, we show that penalized logistic regression consistently performs better than the P+T method whereas predictive performance of the P+T method is very sensitive to the threshold of inclusion of SNPs, depending of the architecture of disease, as shown in \cite{ware2017heterogeneity}.
 
@@
 For simulations, we use real genotypes of European individuals from a case/control celiac disease cohort \cite[]{dubois2010multiple}. 
-This dataset has been quality-controled and imputed in another study \cite[](prive2017efficient).
+This dataset has been quality-controled and imputed in another study \cite[]{prive2017efficient}.
 To get rid of the structure induced by the celiac disease status, we keep only controls from this cohort.
@@
 We vary different parameters: the number of causal variants and their location (30, 300 or 3000 anywhere in the genome and 30 only in the HLA region), the heritability (0.5 or 0.8), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part..
-of the phenotypes (one with only linear effects, and one which combine linear, recessive, dominant and interaction effects).
-This simulation is broader yet similar to the one used in [REF STEPHENS].
-
+of the phenotypes (one ``simple model'' with only linear effects, and one ``fancy model'' which combine linear, dominant and interaction effects).
 For the ``simple model'', we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights ($w_j \neq 0$ if and only if SNP $j$ is causal), $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs).
-\\
 For the ``fancy model'', we separate the causal SNPs in three equal sets $E_\text{causal}^{(1)}$, $E_\text{causal}^{(2)}$ 
-and $E_\text{causal}^{(3)}$ ($E_\text{causal}^{(3)}$ is further separated in two equal sets, $E_\text{causal}^{(3.1)}$ and $E_\text{causal}^{(3.2)}$) 
-and we compute $$y_i = \underbrace{\sum_{j\in E_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in E_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{recessive/dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|E_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} \cdot G_{i,j_2}}}_\text{interaction}$$ where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $E_\text{causal}^{(q)}=\left\{e_k^{(q)}, k \in \left\{1, \ldots, \left|E_\text{causal}^{(q)}\right|\right\}\right\}$.
+and $E_\text{causal}^{(3)}$ ($E_\text{causal}^{(3)}$ is further separated in two equal sets, $E_\text{causal}^{(3.1)}$ and $E_\text{causal}^{(3.2)}$).
+We then compute $$y_i = \underbrace{\sum_{j\in E_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in E_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|E_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$ where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $E_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|E_\text{causal}^{(q)}\right|\right\}\right\}$.
 Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still display a marginal effect.
-\\
 For both models, the $y_i$ are then standardized such that they have a variance equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$ where $\epsilon \sim N(0, 1 - h^2)$.
-
-This results in 32 combinations of parameters and we run 100 simulations for each of these combinations (5 times more than in [REF STEPHENS]), resulting of a total of 3200 simulations.
+This results in 32 combinations of parameters and we run 100 simulations for each of these combinations (5 times more than in \cite{zhou2013polygenic}), resulting of a total of 3200 simulations.
 For each of these 3200 simulations, we randomly re-sample a training set of 6000 individuals, with the corresponding test set composed of the remaining individuals.
-
-[MONTRER UN MANHATTAN PLOT POUR JUSTIFIER 30 IN HLA]
 
@@
 In this study, we use two different measures of predictive accuracy. 
-First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC). In the case of our study, the AUC is the probability that a PRS of a case is greater than one of a control.
+First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction}.
+In the case of our study, the AUC is the probability that a PRS of a case is greater than one of a control.
 This measure indicates how well we can distinguish between cases and controls using PRSs.
-As a second measure, we report the percentage of cases in the 10\% largest PRSs [UTILISER 20\%?].
-This measure indicates how well we can identify people at higher risk from a higher PRS in screening campaigns.
+As a second measure, we report the percentage of cases in the 10\% and 20\% largest PRSs. This measure indicates how well we can identify high-risk people from a higher PRS in screening campaigns.
 
-Note that we also report the timing of the main computations and the number of SNPs used for the prediction.
+Note that we also report the timing of the main computations and the number of SNPs used in the predictions.
 
@@
 
-The P+T (Pruning + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWASs) that are commonly called summary statistics.
+The P+T (Pruning + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWASs), called summary statistics.
 For the P+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
-GWAS was performed using logistic regressions, with 10 first Principal Components (PCs) as covariates in the models.
 The SNPs are first clumped (P) so that there remains only loci that are weakly correlated with each other. 
-Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined). 
-Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$S_i = \sum_{j \in E_\text{clumping}} \mathds{1}\{p_j < T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS and $T$ is a threshold to be determined.
+Thresholding (T) consists in removing SNPs that are under a certain level of significance (p-value threshold to be determined). 
+Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$S_i(T) = \sum_{j \in E_\text{clumping}} \mathds{1}\{p_j < T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS and $T$ is a threshold on p-values to be determined.
 In this study, we report scores for a clumping threshold at $r^2 > 0.2$ with regions of 500kb.
-[threshold in results?]
-[DIRE CA??:] All these steps are implemented in R packages bigstatsr and bigsnpr [REF MOI].
+[thresholds in results?] All these steps are implemented in R packages bigstatsr and bigsnpr [DIRE CA??
+REF MOI].
+We call this method ``PRS'' in the results.
 
-T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests [REF 2001, BReiman] that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data [REF BOTTA].
-In a previous study on WTCCC data [REF], it achieved too-good-to-be-true results.
-Since this method is significantly different from the other tested here, we decided to include it in this comparison in order to assess whether or not it could give superior predictive performance.
+T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}.
+Since this method is significantly different from the other tested here and achieved surprisingly (too-good-to-be-true) results, we decided to include it in this comparison in order to assess whether or not it could give superior predictive performance.
 We use the same parameters as reported in Table 4 of the T-Trees paper, yet, we use only 100 trees (instead of 1000) because it showed only a very subtle increase of AUC while taking way too much time (e.g.
 AUC of 81.5\% instead of 81\%, data not shown).
+We call this method ``T-Trees'' in the results.
 
@@
 We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is the genotypes and covariables (e.g.
-PCs), $y$ is disease status we want to predict and $\lambda$ is a regularization parameter that need to be determined.
-Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (ridge, \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$),
+Principal Components), $y$ is disease status we want to predict and $\lambda$ is a regularization parameter that need to be determined.
+Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (ridge, \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation),
  the L1-regularization (lasso, \cite{tibshirani1996regression}) forces some of the coefficients to be exactly zero and can be used as a means of variable selection, leading to sparse (and therefore more interpretable) models  (corresponds to $\alpha = 1$),
-the L1- and L2-regularization (elastic-net, \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation, or any situation where there are many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
-In this study, we always use $\alpha = 0.5$, without trying to tune the value of this hyper-parameter $\alpha$ (e.g.
-by grid-search and cross-validation).
-
-[ALSO TRIPLE LOGIT]
+the L1- and L2-regularization (elastic-net, \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation where there are many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
+In this study, we always use $\alpha = 0.5$, without trying to tune (e.g.
+by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.
 
@@
 Because of L1-regularization, this vector of coefficients is typically very sparse and can be used to make a PRS based on a linear combination of allele counts.
+We call this method ``logit-simple'' in the results.
+
+In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct another dataset with, for each SNP variable, two more variables coding for recessive and dominant effects.
+This results in a dataset with 3 times more variables than the initial one, on which we can apply the standard method, described previously.
+We call this method ``logit-triple'' in the results.
 
@@
 
-All the code used in this paper along with results,  such as execution times and figures,  are available as HTML R notebooks in the Supplementary Materials.
+All the code used in this paper along with results, such as figures, are available as HTML R notebooks in the Supplementary Materials.
 
@@
 \section{Results}
+
+[MONTRER UN MANHATTAN PLOT POUR JUSTIFIER 30 IN HLA]
+
+[MONTRER UN GRAPH EN FCT D'INCLUSION]
 
@@
 One does not simply identify high-risk individuals (Figure \ref{fig:meme}).
+
+[PARLER DU PROBLEME DE POP? + future paper]
 

### 2017-10-30 3-simus.Rmd
save

@@
 ```{r}
+paste(sprintf("%.1e", 10^(-c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100))))), collapse = ", ")
 PRS <- function(G, infos.chr, infos.pos, 
@@
   par.model  = c("simple", "fancy"),
-  num.simu   = 1:100,
+  num.simu   = 1:50,
   stringsAsFactors = FALSE

### 2017-10-30 paper/draft.tex
save

@@
 In this study, we report scores for a clumping threshold at $r^2 > 0.2$ with regions of 500kb.
-[thresholds in results?] All these steps are implemented in R packages bigstatsr and bigsnpr [DIRE CA??
-REF MOI].
-We call this method ``PRS'' in the results.
+
+For the P+T method, we report three different scores of prediction: one including all the SNPs (after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximize the AUC for these two thresholds and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
+Note that, normally, the optimal threshold would have to be learned on the training set only.
+We consider this reported maximum AUC as an upper bound of the AUC for the P+T method.
+We call these three reported scores respectively ``PRS-all'', ``PRS-stringent'' and ``PRS-max'' in the results.
 
@@
 
-To fit this penalized logistic regression, we use a very efficient algorithms based on \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient}, also implemented in our package bigstatsr.
+To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient}, also implemented in our package bigstatsr \cite[]{prive2017efficient}.
 This type of algorithms build predictions for many values of $\lambda$ (typically 100).
@@
 AUC) is determined and the vector of coefficients corresponding to the previous vector of scores is chosen.
-Finally, the $K$ resulting vectors of coefficients are then combined into one vector (e.g.
+Finally, the $K$ resulting vectors of coefficients are combined into one vector (e.g.
 using the geometric median).
@@
 In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct another dataset with, for each SNP variable, two more variables coding for recessive and dominant effects.
-This results in a dataset with 3 times more variables than the initial one, on which we can apply the standard method, described previously.
+This results in a dataset with 3 times more variables than the initial one, on which we can apply the standard logistic regression with the CMSA procedure, described previously.
 We call this method ``logit-triple'' in the results.
@@
 
-First, we ran only 5 simulations for each combination of parameters (see section \ref{sec:simus}), excepted for the heritability that we fixed at $0.8$.
+First, we ran only 5 simulations for each combination of parameters), excepted for the heritability that we fixed at $0.8$ (see section \ref{sec:simus}.
 The goal was to quickly discard the T-Trees method as being competitive with the other methods. 
-For example, compared to our penalized logistic regression, T-Trees perform worse for both predictive measures, the AUC and the percentage of cases in highest scores.
+For example, compared to our penalized logistic regression, T-Trees perform worse for both predictive measures, the AUC and the percentage of cases in highest scores (Figure \ref{fig:ttrees}).
 Moreover, T-Trees takes longer to run and makes more complex predictive models because it uses more SNPs in the models and have non-linear effects (Figure \ref{fig:ttrees}).
@@
 
-For the P+T method, we report three different scores of prediction: one including all the SNPs (after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximize the AUC for these two thresholds and a sequence of 100 values from $10^{-0.1}$ to $10^{-100}$ equally spaced on the log-log-scale (Table \ref{tab:thr}).
-Note that, normally, the optimal threshold would have to be learned on the training set only.
-We consider this reported maximum AUC as an upper bound of the AUC for the P+T method.
-
 [Also return p-values?]
@@
 First, we show that AUC is highly correlated ($r^2 > 95\%$) with the percentage of cases in either the 10\% or 20\% largest PRSs (Figure \ref{fig:AUC-corr}).
-So, we report only predictive performance in terms of AUC in following results.
+Therefore, AUC can be a good measure at determining whether a score can identify high-risk individuals or not, so we report only predictive performance in terms of AUC in following results.
 
-As for the comparison of our methods with the P+T method, we report AUCs greater than 95\%, which is not obtained by the upper bound of the P+T method in any scenario.
-Our penalized regression method consistently provides better predictive performance than the P+T method.
+Using the penalized logistic regressions, we report AUCs greater than 95\%, which is not obtained by the upper bound of the P+T method in any scenario.
+Our penalized regression methods consistently provides better predictive performance than the P+T method for all disease architectures tested.
 Our method is parallelized and requires only less than 10 minutes to run (with 6 physical cores).
-Moreover, it results in sparse models (...).
+Moreover, it results in relatively sparse models with no more than 14,000 predictors entering the models even for the linear model of 3000 causal SNPs.
 The P+T runs in only a few seconds because we don't consider the GWAS computation as part of the execution time of the method because this is generally executed by others and then the resulting summary statistics are used to construct a PRS.
@@
 In this comparative study, we proposed other methods as a replacement to the widely used P+T method. 
-We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really good performance as reported in previous studies [REF GAD + AUTRE?] \cite[]{abraham2013performance}, better than with the P+T method.
+We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really good performance, as reported in previous studies \cite[]{abraham2013performance}, better than with the P+T method.
 We are well aware that the P+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions.

### 2017-10-30 reg-paths.R

@@ -1,0 +1,57 @@
+# Simulate some phenotypes
+pheno.all <- get_pheno(G, h2 = 0.8, M = 30,
+                       ind.possible = ind.HLA,
+                       effects.dist = "gaussian", 
+                       model = "simple")
+
+# Split in training/test sets
+ind.train <- sort(sample(n, size = n.train))
+ind.test <- setdiff(1:n, ind.train)
+
+# GWAS
+gwas.train <- big_univLogReg(
+  G, pheno.all[ind.train], ind.train = ind.train, 
+  covar.train = covar.all[ind.train, , drop = FALSE], 
+  ncores = nb_cores()
+)
+gwas.train.gc <- snp_gc(gwas.train)
+
+# Clumping on the test set
+ind.keep <- snp_clumping(G, infos.chr = CHR,
+                         ind.row = ind.test,
+                         thr.r2 = 0.2, 
+                         S = abs(gwas.train.gc$score),
+                         size = 500,
+                         is.size.in.bp = TRUE,
+                         infos.pos = POS,
+                         ncores = nb_cores())
+
+# PRS
+thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
+lpS <- -predict(gwas.train.gc)
+nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
+prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
+               ind.test = ind.test,
+               ind.keep = ind.keep,
+               lpS.keep = lpS[ind.keep], 
+               thr.list = thrs)
+
+aucs <- apply(prs, 2, AUC, target = pheno.all[ind.test])
+
+cmsa.logit <- big_spLogReg(X = G, y01.train = pheno.all[ind.train], 
+                           ind.train = ind.train, 
+                           covar.train = covar.all[ind.train, , drop = FALSE],
+                           alpha = 0.5, dfmax = 20e3)
+preds <- predict(cmsa.logit, X = G, ind.row = ind.test, 
+                 covar.row = covar.all[ind.test, , drop = FALSE])
+aucs2 <- apply(preds, 2, AUC, target = pheno.all[ind.test])
+
+library(Matrix)
+bind_rows(
+  data.frame(nb = nb.pred, AUC = aucs, method = "PRS"),
[…]

### 2017-10-30 results1.R

@@ -27,7 +27,7 @@ results %>%
 
 results %>%
   group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise_at(c("timing", "nb.preds", "AUC", "percCases"), mean)
+  summarise_at(c("timing", "nb.preds", "AUC", "percCases10"), mean)
 
 plot_results <- function(results, y, ylab = y) {
   

### 2017-10-30 results2.R

@@ -41,17 +41,13 @@ plot_results <- function(results, y, ylab = y) {
     labs(x = "Method", y = ylab, fill = dist, color = dist)
 }
 
-results %>%
-  filter(!grepl("PRS", method), par.h2 == 0.8) %>%
-  plot_results("timing", "Timing") +
-  scale_y_continuous(breaks = 0:50 * 100, minor_breaks = NULL) 
-ggsave("timing.pdf", scale = 1/90, width = 1200, height = 900)
 
 results %>%
   filter(par.h2 == 0.8) %>%
   plot_results("AUC") +
   scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10)
-ggsave("AUC.pdf", scale = 1/90, width = 1200, height = 900)
+ggsave("../thesis/paper2-PRS/figures/AUC.pdf", 
+       scale = 1/90, width = 1200, height = 900)
 
 ggsave("perc-cases.pdf", scale = 1/90, width = 1200, height = 900)
 
@@ -67,16 +63,24 @@ results %>%
                 labels = scales::comma_format())
 ggsave("nb-preds.pdf", scale = 1/90, width = 1200, height = 900)
 
-cowplot::plot_grid(
-  results %>%
-    filter(par.h2 == 0.8) %>%
-    myggplot(aes(AUC, percCases10, color = par.dist)) +
-    geom_point() +
-    geom_smooth(method = "lm"),
-  results %>%
-    filter(par.h2 == 0.8) %>%
-    myggplot(aes(AUC, percCases20, color = par.dist)) +
-    geom_point() +
-    geom_smooth(method = "lm"),
-  ncol = 1
-)
+results %>%
+  filter(grepl("logit", method)) %>%
+  group_by(method, par.causal) %>%
+  summarise_at("nb.preds", list("Min" = min, "Max" = max))
+
+
+#### Correlation between 
+
+
+ggsave("../thesis/paper2-PRS/figures/AUC-corr.pdf", 
+       scale = 1/90, width = 1300, height = 950)
+
+summary(lm(AUC ~ percCases10, data = results))
+summary(lm(AUC ~ percCases20, data = results))
+
+results %>%
+  group_by_at(c(vars(starts_with("par")), "num.simu")) %>%
+  mutate(AUC_rel = AUC / AUC[method == "logit-simple"]) %>%
[…]

### 2017-10-31 paper/draft.tex
small changes

@@
 Personalized medicine is another major application of PRSs.
-Personalized medicine will use PRSs in screening campaigns in order to identify people at higher risk for a given disease.
+Personalized medicine will use PRSs in screening campaigns in order to identify high-risk individuals for a given disease.
 As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}.
 Moreover, screening based on sequencing seems to make individuals positively change their health behavior while not causing patient anxiety nor depression \cite[]{vassy2017impact}.
-Yet, for PRSs to be used for helping in the diagnostic of disease, either confirming a diagnostic or even early diagnostic, PRSs would have to show a high discriminative power between cases and controls.
+Yet, for PRSs to be used for helping in the diagnosis of disease, either confirming a diagnostic or even early diagnostic, PRSs would have to show a high discriminative power between cases and controls.
 For example, for screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be  greater than respectively 75\% and 99\% \cite[]{janssens2007impact}.
@@
 Many methods have been developed in order to maximize the predictive power of genotypes for diseases, or phenotypes in general.
-A commonly used technique, called P+T (which stands for ``Pruning + Thresholding'') or genetic profiling ..or even just PRS.., is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
-This technique only use summary statistics which make it very ..suitable..
+A commonly used technique, called P+T (which stands for ``Pruning + Thresholding'') -- or genetic profiling ..or even just PRS..
+-- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
+This technique only use summary statistics which makes it very ..suitable..
 and also very fast.
@@
 First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for choosing the amount of regularization used, which directly affects the number of SNPs included in the model.
-We also show how easy it is to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.
+We also show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.
 
 There have already been comparative studies about predicting phenotypes based on genotypes for human data. 
-\citep{abraham2012sparsnp} showed that penalized logistic regression and penalized SVM performed better than standard P+T model and LMMs for a wide-range of human diseases.
-\cite{zhou2013polygenic} used simulations on real genotypes of Australian individuals to show that a hybrid of linear mixed models and sparse regression models perform better in a range of sparse to polygenic disease architectures.
+\citep{abraham2013performance} showed that penalized logistic regression and penalized SVM performed better than the standard P+T method and LMMs, for a wide-range of human diseases.
+\cite{zhou2013polygenic} used simulations on real genotypes of Australian individuals to show that a hybrid of linear mixed models and sparse regression models performed better in a range of sparse to polygenic disease architectures.
 \cite{spiliopoulou2015genomic} investigated the effect of relatedness in prediction performance.
 \cite{ware2017heterogeneity} evaluated best practices for using the P+T method and found out, for example, that the optimal inclusion threshold is trait-dependent. 
-In this study, we particularly extend the work of \cite{abraham2012sparsnp} by using a simulation framework similar but richer than the one used in \cite{zhou2013polygenic}.
-In order to make our comparison as comprehensive as possible, we compare different architectures of disease (number, size and location of causal effects and heritability) with different model of generating liability ..scores.., one with only linear effects, and one which combine linear, dominant and interaction effects.
-We quickly discard T-Trees as being the best method for predicting disease status based on genotypes.
+In this study, we particularly extend the work of \cite{abraham2013performance} by using a simulation framework similar but ..richer..
+than the one used in \cite{zhou2013polygenic}.
+In order to make our comparison as comprehensive as possible, we compare different architectures of disease (number, size and location of causal effects and heritability) with different model of generating liability ..scores.., one with only linear effects, and one which combines linear, dominant and interaction effects.
+First, we quickly discard T-Trees as being the best method for predicting disease status based on genotypes.
 Then, we show that penalized logistic regression consistently performs better than the P+T method whereas predictive performance of the P+T method is very sensitive to the threshold of inclusion of SNPs, depending of the architecture of disease, as shown in \cite{ware2017heterogeneity}.
@@
 For simulations, we use real genotypes of European individuals from a case/control celiac disease cohort \cite[]{dubois2010multiple}. 
-This dataset has been quality-controled and imputed in another study \cite[]{prive2017efficient}.
+Details of quality control and imputation for this dataset are available in another study \cite[]{prive2017efficient}.
 To get rid of the structure induced by the celiac disease status, we keep only controls from this cohort.
-Moreover, to get rid of population structure, we keep only people from UK and further subset these British people based on deviation of robust Mahalanobis distance on Principal Components (see ..SupMat..), leaving 7102 individuals genotyped on ..28x,xxx..
-SNPs [ADD DIM TO RMD] with minimal population structure (see ..SupMat..) because this....
+Moreover, to get rid of population structure, we keep only people from UK and further subset these British people based on deviation of robust Mahalanobis distance on Principal Components (see ..SupMat..).
+This results in 7102 individuals genotyped on ..28x,xxx..
+SNPs [ADD DIM TO RMD] with minimal population structure (see ..SupMat..).
+We decided to get rid of population structure because it can affect the predictive performance of methods, which we will investigate in another study.
 
@@
 To simulate phenotypes, we use a the Liability Threshold Model (LTM) with a prevalence of 30\%.
-We vary different parameters: the number of causal variants and their location (30, 300 or 3000 anywhere in the genome and 30 only in the HLA region), the heritability (0.5 or 0.8), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part..
+We vary different parameters: the number of causal variants and their location (30, 300 or 3000 anywhere in the genome or 30 only in the HLA region), the heritability (0.5 or 0.8), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part..
 of the phenotypes (one ``simple model'' with only linear effects, and one ``fancy model'' which combine linear, dominant and interaction effects).
@@
 This results in 32 combinations of parameters and we run 100 simulations for each of these combinations (5 times more than in \cite{zhou2013polygenic}), resulting of a total of 3200 simulations.
-For each of these 3200 simulations, we randomly re-sample a training set of 6000 individuals, with the corresponding test set composed of the remaining individuals.
+For each of these 3200 simulations, we randomly re-sample a training set of 6000 individuals, with the corresponding test set composed of the remaining individuals on which we evaluate the different methods.
 
@@
 First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction}.
-In the case of our study, the AUC is the probability that a PRS of a case is greater than one of a control.
+In the case of our study, the AUC is the probability that a PRS of a case is greater than the PRS of a control.
 This measure indicates how well we can distinguish between cases and controls using PRSs.
-As a second measure, we report the percentage of cases in the 10\% and 20\% largest PRSs. This measure indicates how well we can identify high-risk people from a higher PRS in screening campaigns.
+As a second measure, we report the percentage of cases in the 10\% and 20\% largest PRSs.
+This measure indicates how well we can identify high-risk individuals from a higher PRS in screening campaigns.
 
@@
 
-For the P+T method, we report three different scores of prediction: one including all the SNPs (after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximize the AUC for these two thresholds and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
+For the P+T method, we report three different scores of prediction: one including all the SNPs (after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximizes the AUC for these two thresholds and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
 Note that, normally, the optimal threshold would have to be learned on the training set only.
@@
 T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}.
-Since this method is significantly different from the other tested here and achieved surprisingly (too-good-to-be-true) results, we decided to include it in this comparison in order to assess whether or not it could give superior predictive performance.
-We use the same parameters as reported in Table 4 of the T-Trees paper, yet, we use only 100 trees (instead of 1000) because it showed only a very subtle increase of AUC while taking way too much time (e.g.
-AUC of 81.5\% instead of 81\%, data not shown).
-We call this method ``T-Trees'' in the results.
+Since this method is significantly different from the other tested here and achieved surprisingly results in the past, we decided to include T-Trees in this comparison in order to assess whether or not it could give superior predictive performance.
+We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}.
+Yet, we use only 100 trees (instead of 1000) because it showed only a very subtle increase of AUC while taking way too much time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown).
+%We call this method ``T-Trees'' in the results.
 
 The last method we compare is penalized logistic regression.
-We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is the genotypes and covariables (e.g.
-Principal Components), $y$ is disease status we want to predict and $\lambda$ is a regularization parameter that need to be determined.
+We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that need to be determined.
 Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (ridge, \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation),
@@
 the L1- and L2-regularization (elastic-net, \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation where there are many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
-In this study, we always use $\alpha = 0.5$, without trying to tune (e.g.
-by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.
+In this study, we always use the elastic-net regularization with $\alpha = 0.5$, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.
 
 To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient}, also implemented in our package bigstatsr \cite[]{prive2017efficient}.
-This type of algorithms build predictions for many values of $\lambda$ (typically 100).
+This type of algorithm builds predictions for many values of $\lambda$ (typically a ``regularization path'' of 100 values).
 To get an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA).
-First, this method separates the training set in $K$ folds (e.g.
-10).
-Secondly, in turn, each fold is considered as an inner validation set and the others ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes a criterion (e.g.
-AUC) is determined and the vector of coefficients corresponding to the previous vector of scores is chosen.
-Finally, the $K$ resulting vectors of coefficients are combined into one vector (e.g.
-using the geometric median).
-Because of L1-regularization, this vector of coefficients is typically very sparse and can be used to make a PRS based on a linear combination of allele counts.
+First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds).
+Secondly, in turn, each fold is considered as an inner validation set and the others ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes a criterion (e.g.\ AUC) is determined and the vector of coefficients corresponding to the previous vector of scores is chosen.
+Finally, the $K$ resulting vectors of coefficients are combined into one vector (e.g.\ using the geometric median).
+Because of L1-regularization, this vector of coefficients is typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts.
 We call this method ``logit-simple'' in the results.
@@
 
-First, we ran only 5 simulations for each combination of parameters), excepted for the heritability that we fixed at $0.8$ (see section \ref{sec:simus}.
+First, we ran only 5 simulations for each combination of parameters), excepted for the heritability that we fixed at $0.8$ (see section \ref{sec:simus}).
 The goal was to quickly discard the T-Trees method as being competitive with the other methods. 
 For example, compared to our penalized logistic regression, T-Trees perform worse for both predictive measures, the AUC and the percentage of cases in highest scores (Figure \ref{fig:ttrees}).
-Moreover, T-Trees takes longer to run and makes more complex predictive models because it uses more SNPs in the models and have non-linear effects (Figure \ref{fig:ttrees}).
+Moreover, T-Trees takes longer to run and makes more complex predictive models because it uses more SNPs in the models and has non-linear effects (Figure \ref{fig:ttrees}).
 So, we decided to discard this method for the rest of the simulations.
@@
 
-[Also return p-values?]
+[Also return p-values?]\\
 
 First, we show that AUC is highly correlated ($r^2 > 95\%$) with the percentage of cases in either the 10\% or 20\% largest PRSs (Figure \ref{fig:AUC-corr}).
-Therefore, AUC can be a good measure at determining whether a score can identify high-risk individuals or not, so we report only predictive performance in terms of AUC in following results.
+Therefore, AUC can be a good measure at determining whether a score can identify high-risk individuals or not, so we report only predictive performance in terms of AUC in the following results.
 
 Using the penalized logistic regressions, we report AUCs greater than 95\%, which is not obtained by the upper bound of the P+T method in any scenario.
-Our penalized regression methods consistently provides better predictive performance than the P+T method for all disease architectures tested.
-Our method is parallelized and requires only less than 10 minutes to run (with 6 physical cores).
-Moreover, it results in relatively sparse models with no more than 14,000 predictors entering the models even for the linear model of 3000 causal SNPs.
-The P+T runs in only a few seconds because we don't consider the GWAS computation as part of the execution time of the method because this is generally executed by others and then the resulting summary statistics are used to construct a PRS.
+Our penalized regression methods consistently provides better predictive performance than the P+T method for all disease architectures tested [FIGURE].
+Our method is parallelized and requires only less than 10 minutes to run (with 6 physical cores) [FIGURE].
+Moreover, it results in relatively sparse models with no more than 14,000 predictors entering the models even for the linear model of 3000 causal SNPs [TABLE ?FIGURE?].
+P+T runs in only a few seconds because we don't consider the GWAS computation as part of the execution time of the method because this is generally executed by others (e.g.\ consortia) and then only the resulting summary statistics are used to construct a PRS.
 
-As for the comparison between our two penalized logistic regression, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects.
+As for the comparison between our two penalized logistic regression, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects [FIGURE].
 Yet, the ``triple'' solution takes approximately 3 more times to run.
 
-As for general results, we report a decrease of AUC when the number of causal variants $M$ is increasing.
+As for general results, we report a decrease of AUC when the number of causal variants $M$ is increasing [BATEAU? quoi d'autre?].
 
@@
 In this comparative study, we proposed other methods as a replacement to the widely used P+T method. 
-We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really good performance, as reported in previous studies \cite[]{abraham2013performance}, better than with the P+T method.
+We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really good performance, as reported in a previous study \cite[]{abraham2013performance}, higher than with the P+T method.
 We are well aware that the P+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions.
-Yet, in the case of prediction of disease to be used for clinical interventions, best possible methods should be used, even if it has to be used by large consortia only.
+Yet, in the case of prediction of disease to be used for clinical interventions, best possible methods should be used, even if it has to be used by large consortia only [EVEN LARGE CONSORTIA DON'T HAVE ALL DATA?].
 

### 2017-11-02 paper/draft.tex
save

@@
 
-Using the penalized logistic regressions, we report AUCs greater than 95\%, which is not obtained by the upper bound of the P+T method in any scenario.
-Our penalized regression methods consistently provides better predictive performance than the P+T method for all disease architectures tested [FIGURE].
-Our method is parallelized and requires only less than 10 minutes to run (with 6 physical cores) [FIGURE].
+Using the penalized logistic regressions, we report AUCs greater than 95\%, which is not obtained by the upper bound of the P+T method in any scenario (Figure \ref{fig:AUC}).
+Our penalized regression methods consistently provides better predictive performance than the P+T method for all disease architectures tested [ou pas, FIGURE].
+Our method is parallelized and requires only less than 10 minutes to run (with 6 physical cores) [TABLE ?FIGURE?].
 Moreover, it results in relatively sparse models with no more than 14,000 predictors entering the models even for the linear model of 3000 causal SNPs [TABLE ?FIGURE?].
@@
 
-As for the comparison between our two penalized logistic regression, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects [FIGURE].
+As for the comparison between our two penalized logistic regression, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects (Figure \ref{fig:AUC-rel}).
 Yet, the ``triple'' solution takes approximately 3 more times to run.
@@
 [PARLER DU PROBLEME DE POP? + future paper]
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{AUC}}
+\caption{AUC...}\label{fig:AUC}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{AUC-rel}}
+\caption{Relative AUC...}\label{fig:AUC-rel}
+\end{figure}
 

### 2017-11-04 1-preprocessing.Rmd
save

@@
 
-## Data
-
 ### QC and imputation
 
-This was done in another paper TODO(add ref).
-
-### Add population from external files
+This was done in [another paper](https://doi.org/10.1101/190926).
 
@@
 library(bigsnpr)
-celiac <- snp_attach("backingfiles/celiacQC.rds")
+str(celiac <- snp_attach("backingfiles/celiacQC.rds"), max.level = 1)
+```
 
+### Add population from external files
+
+```{r}
 # Get population
@@
 
-```{r}
+```{r, warning=FALSE}
 # Compute Mahalanobis distance on PCs
@@
 # Plot UK controls, coloring with Mahalanobis distance
-plotlist <- lapply(list(1:2, 3:4, 5:6, 7:8), function(scores) {
+library(magrittr)
+p_list <- lapply(list(1:2, 3:4, 5:6, 7:8), function(scores) {
   plot(obj.svd, type = "scores", scores = scores, coeff = 0.7) + 
     aes(color = (distM > THR)) + 
-    labs(color = "Outlier?")
+    labs(color = "Outlier?") +
+    scale_color_manual(breaks = c(TRUE, FALSE), values = scales::hue_pal()(2))
 }) 
-cowplot::plot_grid(plotlist = plotlist, ncol = 2, align = "hv")
+lapply(p_list, function(p) p + theme(legend.position = "none")) %>%
+  cowplot::plot_grid(plotlist = ., ncol = 2, align = "hv", scale = 0.95) %>%
+  cowplot::plot_grid(cowplot::get_legend(p_list[[1]]),
+                     rel_widths = c(1, 0.15))
+```
+
+```{r}
+ggsave("figures/outliers-pop.pdf", scale = 1/90, width = 870, height = 650)
 ```
@@
 plot(obj.svd2)
-plot(obj.svd2, type = "scores", scores = 1:2) 
+plot(obj.svd2, type = "scores") 
 # Save this new PCA
@@
 ```
+
+```{r}
+ggsave("figures/no-pop.pdf", scale = 1/80, width = 620, height = 570)
+```

### 2017-11-04 2-newfiles.Rmd
save
okay?

@@
   newfile <- bigsnpr:::getNewFile(x, "tripled")
-  G2 <- FBM.code256(nrow(G), 3 * ncol(G), 
-                    code = G$code256,
-                    backingfile = newfile,
-                    save = TRUE)
+  G2 <- FBM.code256(nrow(G), 3 * ncol(G), code = bigsnpr:::CODE_012, 
+                    backingfile = newfile, save = TRUE)
 
@@
 ```{r}
+snp_triple(celiac)
 snp_triple(celiac2)

### 2017-11-04 5-regularization-paths.Rmd
save

---
title: "Analysis of celiac disease"
author: "Florian Privé"
date: "November 3, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
                      fig.asp = 0.7, error = TRUE)
options(width = 85)
```

```{r}
library(ggplot2)
myggplot <- function(..., coeff = 1) {
  bigstatsr:::MY_THEME(ggplot(...), coeff = coeff)
} 
```

## Data

```{r}
# Read file, see "Preprocessing" notebook
celiac <- snp_attach("backingfiles/celiacQC.rds")
G <- celiac$genotypes
CHR <- celiac$map$chromosome
POS <- celiac$map$physical.pos
NCORES <- nb_cores()
y <- celiac$fam$affection - 1
```

```{r}
# Compute PCA
obj.svd <- snp_autoSVD(G, CHR, POS, ncores = NCORES)
```

## Predictive models

```{r}
# Divide in training/test sets
dim(G)
ind.train <- sort(sample(nrow(G), size = 12e3))
ind.test <- setdiff(rows_along(G), ind.train)
```

### PRS

```{r}
# GWAS
gwas.train <- big_univLogReg(
  G, y[ind.train], ind.train = ind.train, 
  covar.train = obj.svd$u[ind.train, , drop = FALSE], 
  ncores = NCORES
)
gwas.train.gc <- snp_gc(gwas.train)
```

```{r}
labels <- 1:22; labels[c(11, 13, 15, 17, 18, 20, 21)] <- ""
cowplot::plot_grid(
  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels),
  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels) +
    coord_cartesian(ylim = c(0, 30)) + 
    geom_hline(yintercept = -log10(5e-8), color = "red", linetype = 3),
  align = "hv", ncol = 1, labels = LETTERS[1:2], label_size = 25, scale = 0.95
)
```

```{r}
ggsave("figures/celiac-man.png", scale = 1/90, width = 1400, height = 1000)
```

```{r}
# Clumping on the test set
ind.keep <- snp_clumping(G, infos.chr = CHR,
                         ind.row = ind.test,
                         thr.r2 = 0.2, 
                         S = abs(gwas.train.gc$score),
                         size = 500,
                         is.size.in.bp = TRUE,
                         infos.pos = POS,
                         ncores = NCORES)
```

```{r}
# PRS
thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
lpS <- -predict(gwas.train.gc)
prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
               ind.test = ind.test,
               ind.keep = ind.keep,
               lpS.keep = lpS[ind.keep], 
               thr.list = thrs)
nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
aucs <- apply(prs, 2, AUC, target = y[ind.test])
```

### Logistic regression

```{r}
logit <- big_spLogReg(X = G, y01.train = y[ind.train], 
                      ind.train = ind.train, 
                      covar.train = obj.svd$u[ind.train, ],
                      alpha = 0.5, dfmax = 20e3)
preds1 <- predict(logit, X = G, ind.row = ind.test, 
                  covar.row = obj.svd$u[ind.test, ])
library(Matrix)
nb.pred1 <- colSums(logit$beta != 0)
aucs1 <- apply(preds1, 2, AUC, target = y[ind.test])
```

Cross-Model Selection and Averaging:

```{r}
system.time({
  cmsa.logit <- big_CMSA(FUN = big_spLogReg, feval = AUC,
                         X = G, y.train = y[ind.train], 
                         ind.train = ind.train, 
                         covar.train = obj.svd$u[ind.train, ],
                         alpha = 0.5, dfmax = 20e3, 
                         ncores = NCORES)

  preds2 <- predict(cmsa.logit, X = G, ind.row = ind.test, 
                    covar.row = obj.svd$u[ind.test, ])
})
(auc.cmsa <- AUC(preds2, y[ind.test]))
```

### Logistic regression with feature engineering

```{r}
G2 <- big_attach("backingfiles/celiacQC_tripled1.rds")
logit2 <- big_spLogReg(X = G2, y01.train = y[ind.train], 
                       ind.train = ind.train, 
                       covar.train = obj.svd$u[ind.train, ],
                       alpha = 0.5, dfmax = 20e3)
preds3 <- predict(logit2, X = G2, ind.row = ind.test, 
                  covar.row = obj.svd$u[ind.test, ])
nb.pred2 <- colSums(logit2$beta != 0)
aucs2 <- apply(preds3, 2, AUC, target = y[ind.test])
```

Cross-Model Selection and Averaging:

```{r}
system.time({
  cmsa.logit2 <- big_CMSA(FUN = big_spLogReg, feval = AUC,
                          X = G2, y.train = y[ind.train], 
                          ind.train = ind.train, 
                          covar.train = obj.svd$u[ind.train, ],
                          alpha = 0.5, dfmax = 20e3, 
                          ncores = NCORES)

  preds4 <- predict(cmsa.logit2, X = G2, ind.row = ind.test, 
                    covar.row = obj.svd$u[ind.test, ])
})
(auc.cmsa2 <- AUC(preds4, y[ind.test]))
```

### Regularization paths

```{r}
h <- round(c(max(aucs), auc.cmsa, auc.cmsa2), 4)
p <- myggplot(rbind(
  cbind.data.frame(nb = nb.pred,      AUC = aucs,      Method = "PRS"),
  cbind.data.frame(nb = nb.pred1[-1], AUC = aucs1[-1], Method = "logit-simple"),
  cbind.data.frame(nb = nb.pred2[-1], AUC = aucs2[-1], Method = "logit-triple")
)) +
  geom_point(aes(nb, AUC, color = Method)) +
  scale_x_log10(breaks = 10^(0:6)) +
  labs(x = "Number of predictors") +
  geom_hline(yintercept = h, color = scales::hue_pal()(3), linetype = 2) +
  theme(legend.position = c(0.85, 0.5))
p + 
  scale_y_continuous(breaks = sort(c(ggplot_build(p)$layout$panel_ranges[[1]]$y.major_source, h)))
```

```{r}
ggsave("figures/celiac-regpath.pdf", scale = 1/90, width = 1000, height = 700)
```

### 2017-11-04 results2.R

@@ -8,9 +8,8 @@ myggplot <- function(..., coeff = 1) {
 top10 <- 1:110
 top20 <- 1:220
 
-results <- list.files("../thesis/paper2-PRS/results1/", full.names = TRUE) %>%
+results <- list.files("../thesis/paper2-PRS/results2/", full.names = TRUE) %>%
   map_dfr(~readRDS(.x)) %>%
-  filter(method != "T-Trees") %>%
   as_tibble() %>%
   mutate(
     par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
@@ -43,7 +42,7 @@ plot_results <- function(results, y, ylab = y) {
 
 
 results %>%
-  filter(par.h2 == 0.8) %>%
+  filter(par.h2 == 0.5) %>%
   plot_results("AUC") +
   scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10)
 ggsave("../thesis/paper2-PRS/figures/AUC.pdf", 
@@ -83,4 +82,6 @@ results %>%
   mutate(AUC_rel = AUC / AUC[method == "logit-simple"]) %>%
   # select(-eval, -nb.preds, -percCases10, -percCases20) %>%
   plot_results(y = "AUC_rel", ylab = "Relative AUC / 'logit-simple'")
+ggsave("../thesis/paper2-PRS/figures/AUC-rel.pdf", 
+       scale = 1/90, width = 1200, height = 900)
 

### 2017-11-17 3-simus.Rmd
save

@@
 
-```{r}
+```{r, eval=TRUE}
 if (!dir.exists("results2")) dir.create("results2")
@@
 ```
+
+```{r, eval=FALSE}
+if (!dir.exists("results3")) dir.create("results3")
+
+G6 <- big_copy(G, ind.col = which(CHR == 6))
+ind.HLA6 <- snp_indLRLDR(CHR[CHR == 6], POS[CHR == 6], subset(LD.wiki34, ID == "hild12"))
+
+for (i in rows_along(params.grid2)) {
+
+  res.file <- paste0("results3/simu_", i, ".rds")
+  if (file.exists(res.file)) next
+
+  params <- params.grid2[i, ]
+  par.causal <- params[["par.causal"]][[1]]
+
+  # Simulate phenotypes
+  pheno.all <- get_pheno(
+    G6,    
+    h2 = params[["par.h2"]], 
+    M = as.integer(par.causal[1]), 
+    ind.possible = `if`(par.causal[2] == "all", cols_along(G6), ind.HLA6),
+    effects.dist = params[["par.dist"]], 
+    model = params[["par.model"]], 
+    K = 0.3                 
+  )
+
+  # Split in training/test sets
+  ind.train <- sort(sample(n, size = n.train))
+  ind.test <- setdiff(1:n, ind.train)
+
+  # Get results from all methods
+  res <- bind_rows(
+    PRS(G6, CHR[CHR == 6], POS[CHR == 6], pheno.all, covar.all, ind.train, ind.test),
+    logit.CMSA(G6, pheno.all, covar.all, ind.train, ind.test, "logit-simple")
+  )
+  print(res)
+  params[["res"]] <- list(res)
+  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
+}
+```

### 2017-11-17 tmp-results.Rmd
save

---
title: "Results"
author: "Florian Privé"
date: "November 17, 2017"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", fig.asp = 0.7)
options(width = 110)
```

## Useful functions

```{r, message=FALSE}
library(tidyverse)
```

```{r}
myggplot <- function(..., coeff = 1) {
  bigstatsr:::MY_THEME(ggplot(...), coeff = coeff)
} 
```

```{r}
plot_results <- function(results, y, ylab = y) {

  dist <- "Distribution\nof effects"

  myggplot(results) +
    geom_boxplot(aes_string("method", y, color = "par.dist", 
                            fill = "par.dist"), alpha = 0.3) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    facet_grid(par.model ~ par.causal) +
    theme(strip.text.x = element_text(size = rel(2)),
          strip.text.y = element_text(size = rel(2))) +
    labs(x = "Method", y = ylab, fill = dist, color = dist)
}
```

## Results with T-Trees

```{r}
top10 <- 1:110
top20 <- 1:220

# Put all results in a single tibble
results <- list.files("results1", full.names = TRUE) %>%
  map_dfr(~readRDS(.x)) %>%
  as_tibble() %>%
  mutate(
    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2])),
    percCases10 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[top10], 2])),
    percCases20 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[top20], 2]))
  )
```

```{r}
results %>%
  filter(method %in% c("T-Trees", "logit-simple")) %>%
  group_by_at(c(vars(starts_with("par")), "method")) %>%
  summarise_at(c("timing", "nb.preds", "AUC", "percCases10", "percCases20"), mean) %>%
  print(n = Inf)
```

```{r, fig.width=16, out.width="95%", dev="svg"}
ttrees_vs_logit <- filter(results, method %in% c("T-Trees", "logit-simple"))

p_list <- list(
  plot_results(ttrees_vs_logit, "timing", "Timing (in seconds)") +
    scale_y_continuous(breaks = 0:10 * 2000, minor_breaks = NULL),
  plot_results(ttrees_vs_logit, "nb.preds", "Number of predictors (log-scale)") +
    scale_y_log10(breaks = c(10^(0:7), 3 * 10^(0:7)), minor_breaks = NULL,
                  labels = scales::comma_format()),
  plot_results(ttrees_vs_logit, "AUC") +
    scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10),
  plot_results(ttrees_vs_logit, "percCases10", "Percentage of cases in top 10%") +
    scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10)
)

lapply(p_list, function(p) p + theme(legend.position = "none")) %>%
  cowplot::plot_grid(plotlist = ., ncol = 2, align = "hv", scale = 0.9,
                     labels = LETTERS[1:4], label_size = 25) %>%
  cowplot::plot_grid(cowplot::get_legend(p_list[[1]]),
                     rel_widths = c(1, 0.15))
```

```{r}
ggsave("figures/ttrees.pdf", scale = 1/90, width = 1580, height = 1070)
```

## Results without T-Trees

### Correlation between predictive performance measures

```{r}
top10 <- 1:110
top20 <- 1:220

# Put all results in a single tibble
results2 <- list.files("results2", full.names = TRUE) %>%
  map_dfr(~readRDS(.x)) %>%
  as_tibble() %>%
  mutate(
    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2])),
    percCases10 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[top10], 2])),
    percCases20 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[top20], 2]))
  )
```

```{r, fig.width=14, out.width="90%", dev="svg"}
H2 <- 0.8
cowplot::plot_grid(
  results2 %>%
    filter(par.h2 == H2) %>%
    myggplot(aes(AUC, percCases10, color = par.dist)) +
    geom_point() +
    geom_smooth(method = "lm") +
    theme(legend.position = c(0.8, 0.2)) +
    labs(y = "Percentage of cases in top 10%", 
         color = "Distribution\nof effects"),
  results2 %>%
    filter(par.h2 == H2) %>%
    myggplot(aes(AUC, percCases20, color = par.dist)) +
    geom_point() +
    geom_smooth(method = "lm") +
    theme(legend.position = c(0.8, 0.2)) +
    labs(y = "Percentage of cases in top 20%", 
         color = "Distribution\nof effects"),
  results2 %>%
    filter(par.h2 == H2) %>%
    myggplot(aes(AUC, percCases10, color = par.model)) +
    geom_point() +
    geom_smooth(method = "lm") +
    theme(legend.position = c(0.8, 0.2)) +
    scale_colour_brewer(palette = "Set1") +
    labs(y = "Percentage of cases in top 10%", 
         color = "Model"),
  results2 %>%
    filter(par.h2 == H2) %>%
    myggplot(aes(AUC, percCases20, color = par.model)) +
    geom_point() +
    geom_smooth(method = "lm") +
    theme(legend.position = c(0.8, 0.2)) +
    scale_colour_brewer(palette = "Set1") +
    labs(y = "Percentage of cases in top 20%", 
         color = "Model"),
  labels = LETTERS[1:4], align = "hv", label_size = 25, scale = 0.95
)
```

```{r}
ggsave("figures/AUC-corr.pdf", scale = 1/90, width = 1300, height = 950)
```

### Results of AUC

```{r, fig.cap="All AUC results for h2=0.8 and all chromosomes"}
results2 %>%
  filter(par.h2 == 0.8) %>%
  plot_results("AUC") +
  geom_hline(yintercept = 0.94, color = "blue", linetype = 3) +
  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10)
```

```{r, fig.cap="All relative AUC results for h2=0.8 and all chromosomes"}
results2 %>%
  filter(par.h2 == 0.8) %>%
  group_by_at(c(vars(starts_with("par")), "num.simu")) %>%
  mutate(AUC_rel = AUC / AUC[method == "logit-simple"]) %>%
  plot_results(y = "AUC_rel", ylab = "Relative AUC / 'logit-simple'") +
  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10)
```

```{r, fig.cap="All results for h2=0.5 and all chromosomes"}
results2 %>%
  filter(par.h2 == 0.5) %>%
  plot_results("AUC") +
  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10)
```

```{r}
# represent h2=0.8 as a function of h2=0.5
results2 %>%
  select(starts_with("par"), method, AUC) %>%
  group_by(par.causal, par.dist, par.model, method, par.h2) %>%
  summarise(AUC = mean(AUC)) %>%
  spread(par.h2, AUC) %>%
  myggplot(aes(`0.5`, `0.8`, color = method)) +
  geom_smooth(size = 2, alpha = 0.2) +
  geom_point(size = 3)

results2 %>%
  select(starts_with("par"), method, AUC) %>%
  group_by(par.causal, par.dist, par.model, method, par.h2) %>%
  summarise(AUC = mean(AUC)) %>%
  spread(par.h2, AUC) %>%
  with(cor(`0.5`, `0.8`))
```

```{r, fig.cap="Main results: logit simple vs PRS max"}
results2 %>%
  filter(par.dist == "laplace", par.h2 == 0.8, par.model == "simple",
         method %in% c("logit-simple", "PRS-max")) %>%
  group_by(par.causal, method) %>%
  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, 1e4)) %>%
  myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
  geom_hline(yintercept = 0.5, linetype = 2) +
  geom_bar(stat = "identity", alpha = 0.3, position=position_dodge()) +
  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
                position=position_dodge(width=0.9), color = "black", width = 0.2) +
  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = 0:10 / 10 + 0.05)
```

```{r, fig.cap="Main results: all PRS"}
results2 %>%
  filter(par.dist == "laplace", par.h2 == 0.8, par.model == "simple",
         grepl("PRS", method)) %>%
  group_by(par.causal, method) %>%
  summarise(AUC_mean = mean(AUC),
            AUC_boot = boot(AUC, 1e4)) %>%
  myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
  geom_hline(yintercept = 0.5, linetype = 2) +
  geom_bar(stat = "identity", alpha = 0.3, position=position_dodge()) +
  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
                position=position_dodge(width=0.9), color = "black", width = 0.2) +
  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = 0:10 / 10 + 0.05)
```

### Results for chromosome 6

```{r}
# Put all results in a single tibble
results3 <- list.files("results3", full.names = TRUE) %>%
  map_dfr(~readRDS(.x)) %>%
  as_tibble() %>%
  mutate(
    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2]))
  )
```

```{r, fig.cap="All AUC results for h2=0.8 and chromosome 6"}
results3 %>%
  filter(par.h2 == 0.8) %>%
  plot_results("AUC") +
  geom_hline(yintercept = 0.94, color = "blue", linetype = 3) +
  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10)
```

```{r, fig.cap="All relative AUC results for h2=0.8 and chromosome 6"}
results3 %>%
  filter(par.h2 == 0.8) %>%
  group_by_at(c(vars(starts_with("par")), "num.simu")) %>%
  mutate(AUC_rel = AUC / AUC[method == "logit-simple"]) %>%
  plot_results(y = "AUC_rel", ylab = "Relative AUC / 'logit-simple'") +
  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10)
```

```{r, fig.cap="Main results: logit simple vs PRS max for chromosome 6"}
results3 %>%
  filter(par.dist == "laplace", par.h2 == 0.8, par.model == "simple",
         method %in% c("logit-simple", "PRS-max")) %>%
  group_by(par.causal, method) %>%
  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, 1e4)) %>%
  myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
  geom_hline(yintercept = 0.5, linetype = 2) +
  geom_bar(stat = "identity", alpha = 0.3, position=position_dodge()) +
  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
                position=position_dodge(width=0.9), color = "black", width = 0.2) +
  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = 0:10 / 10 + 0.05)
```

```{r, fig.cap="Main results: all PRS for chromosome 6"}
results3 %>%
  filter(par.dist == "laplace", par.h2 == 0.8, par.model == "simple",
         grepl("PRS", method)) %>%
  group_by(par.causal, method) %>%
  summarise(AUC_mean = mean(AUC),
            AUC_boot = boot(AUC, 1e4)) %>%
  myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
  geom_hline(yintercept = 0.5, linetype = 2) +
  geom_bar(stat = "identity", alpha = 0.3, position=position_dodge()) +
  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
                position=position_dodge(width=0.9), color = "black", width = 0.2) +
  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = 0:10 / 10 + 0.05)
```

### Prediction on Celiac

```{r}
knitr::include_graphics("figures/celiac-man.png")
```

```{r}
knitr::include_graphics("figures/celiac-regpath.png")
```

### Misc

```{r}
results2 %>%
  filter(
    # method %in% c("logit-simple", "PRS-max"),
    grepl("all", par.causal)
  ) %>%
  mutate(M = readr::parse_number(par.causal)) %>%
  group_by(M, par.dist, par.h2, par.model, method) %>%
  summarise(AUC = mean(AUC)) %>%
  # filter(par.h2 == 0.8, par.dist == "gaussian", par.model == "simple") %>%
  filter(par.h2 == 0.8) %>%
  myggplot(aes(M, AUC, color = method)) +
  geom_line() + 
  geom_point(size = 2) +
  facet_grid(par.dist ~ par.model) +
  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10) +
  scale_x_sqrt(breaks = c(30, 300, 1200, 3000)) +
  labs(x = "Number of causal SNPs (sqrt-scale)")
```

```{r}
results3 %>%
  filter(
    # method %in% c("logit-simple", "PRS-max"),
    grepl("all", par.causal)
  ) %>%
  mutate(M = readr::parse_number(par.causal)) %>%
  group_by(M, par.dist, par.h2, par.model, method) %>%
  summarise(AUC = mean(AUC)) %>%
  # filter(par.h2 == 0.8, par.dist == "gaussian", par.model == "simple") %>%
  filter(par.h2 == 0.8) %>%
  myggplot(aes(M, AUC, color = method)) +
  geom_line() + 
  geom_point(size = 2) +
  facet_grid(par.dist ~ par.model) +
  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10) +
  scale_x_sqrt(breaks = c(30, 300, 1200, 3000)) +
  labs(x = "Number of causal SNPs (sqrt-scale)")
```

```{r}
bind_rows(
  bind_cols(results2, simu = rep("all", nrow(results2))),
  bind_cols(results3, simu = rep("chr6", nrow(results3)))
) %>%
  filter(par.h2 == 0.8, par.dist == "gaussian", method != "logit-triple") %>%
  myggplot() +
  geom_boxplot(aes(method, AUC, fill = simu, color = simu), alpha = 0.3) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(par.model ~ par.causal) +
  theme(strip.text.x = element_text(size = rel(2)),
        strip.text.y = element_text(size = rel(2))) +
  scale_colour_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  geom_hline(yintercept = 0.94, color = "blue", linetype = 2)
```

```{r, fig.cap="Size of effects GWAS vs logistic"}
knitr::include_graphics("effects.png")
```

```{r}
knitr::include_graphics("preds-density2.png")
```

```{r}
knitr::include_graphics("preds-density3.png")
```

```{r, fig.cap="Percent of controls (errors) in 199 (homoz bad in test)."}
knitr::include_graphics("perc_cases2.png")
```

```{r, fig.cap="Projection on test set of score of Gad, train on same dataset"}
knitr::include_graphics("gad-pred.png")
```

### 2017-11-17 oracle.R

@@ -1,0 +1,84 @@
+get_oracle <- function(
+  G,                                        ## matrix of genotypes
+  h2,                                       ## heritability 
+  M,                                        ## nbs of causal variants
+  ind.possible = cols_along(G),             ## indices of possible causal variants
+  effects.dist = c("gaussian", "laplace"),  ## distribution of effects 
+  model = c("simple", "fancy"),             ## model for simulation
+  K = 0.3                                   ## prevalence
+) {
+  
+  effects.dist  <- match.arg(effects.dist)
+  model <- match.arg(model)
+  
+  set <- sample(ind.possible, size = M)
+  effects <- `if`(effects.dist == "gaussian", 
+                  rnorm(M, sd = sqrt(h2 / M)),
+                  rmutil::rlaplace(M, s = sqrt(h2 / (2*M))))
+  
+  if (model == "simple") {
+    # only linear
+    y.simu <- scale(G[, set]) %*% effects
+  } else {
+    sets <- split(1:M, sample(rep_len(1:3, M)))
+    # linear
+    ind1 <- sets[[1]]
+    y.simu <- scale(G[, set[ind1]]) %*% effects[ind1]
+    # recessive / dominant
+    ind2 <- sets[[2]]
+    y.simu <- y.simu + scale(G[, set[ind2]] > 0.5) %*% effects[ind2]
+    # interactions
+    ind3 <- matrix(sets[[3]], ncol = 2)
+    y.simu <- y.simu + scale(G[, set[ind3[, 1]]] * G[, set[ind3[, 2]]]) %*% 
+      effects[ind3[, 1]] * sqrt(2)
+  }
+  
+  y.simu <- y.simu / sd(y.simu) * sqrt(h2)
+  stopifnot(all.equal(drop(var(y.simu)), h2))
+  S <- y.simu
+  y.simu <- S + rnorm(nrow(G), sd = sqrt(1 - h2))
+  pheno <- as.numeric(y.simu > qnorm(1 - K))
+  AUC(S, pheno)
+}
+
+get_oracle(G, 0.8, 30, effects.dist = "gaussian", model = "simple")
+get_oracle(G, 0.8, 3000, effects.dist = "gaussian", model = "simple")
+
+
+params.grid2 <- expand.grid(
+  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
+  par.dist   = "gaussian", 
[…]

### 2017-11-17 prs-gad.R

@@ -1,0 +1,29 @@
+library(bigsnpr)
+
+plink <- download_plink()
+celiac <- "../thesis-celiac/Dubois2010_data/FinnuncorrNLITUK3hap550"
+
+system(glue::glue(
+  "{plink} --score grs.txt --bfile {celiac}"
+))
+
+
+d <- read.table("plink.profile", header=TRUE)
+intercept <- -0.757226
+d$GRS <- d$SCORE * d$CNT + intercept
+# write.table(d[, c("FID", "IID", "GRS")],
+#             file="profile.txt", row.names=FALSE, col.names=FALSE)
+
+
+
+ind.test500 <- ind.test[celiac$fam$family.ID[ind.test] != "UK1"]
+preds500 <- preds2[celiac$fam$family.ID[ind.test] != "UK1"]
+
+preds_gad <- d$GRS[match(celiac$fam$sample.ID[ind.test500], d$IID)]
+
+qplot(preds500, preds_gad) +
+  geom_smooth(method = "lm", color = "blue") +
+  labs(x = "My pred (AUC=88.6%)", y = "Gad pred (AUC=89.3%)")
+
+AUC(preds500, y[ind.test500])
+AUC(preds_gad, y[ind.test500])

### 2017-11-17 results2.R

@@ -1,87 +1,0 @@
-
-library(tidyverse)
-
-myggplot <- function(..., coeff = 1) {
-  bigstatsr:::MY_THEME(ggplot(...), coeff = coeff)
-} 
-
-top10 <- 1:110
-top20 <- 1:220
-
-results <- list.files("../thesis/paper2-PRS/results2/", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  as_tibble() %>%
-  mutate(
-    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
-                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
-    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2])),
-    percCases10 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[top10], 2])),
-    percCases20 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[top20], 2]))
-  )
-
-print(results, n = Inf)
-
-
-results %>%
-  group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise_at(c("timing", "nb.preds", "AUC", "percCases10"), mean)
-
-plot_results <- function(results, y, ylab = y) {
-  
-  dist <- "Distribution\nof effects"
-  
-  myggplot(results) +
-    geom_boxplot(aes_string("method", y, color = "par.dist", 
-                            fill = "par.dist"), alpha = 0.3) + 
-    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-    facet_grid(par.model ~ par.causal) +
-    theme(strip.text.x = element_text(size = rel(2)),
-          strip.text.y = element_text(size = rel(2))) +
-    labs(x = "Method", y = ylab, fill = dist, color = dist)
-}
-
-
-results %>%
-  filter(par.h2 == 0.5) %>%
-  plot_results("AUC") +
-  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10)
-ggsave("../thesis/paper2-PRS/figures/AUC.pdf", 
-       scale = 1/90, width = 1200, height = 900)
-
-ggsave("perc-cases.pdf", scale = 1/90, width = 1200, height = 900)
-
-results %>%
-  filter(par.h2 == 0.8) %>%
[…]

### 2017-11-17 results3.R

@@ -1,0 +1,14 @@
+# Put all results in a single tibble
+results3 <- list.files("results3", full.names = TRUE) %>%
+  map_dfr(~readRDS(.x)) %>%
+  as_tibble() %>%
+  mutate(
+    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
+                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
+    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2]))
+  )
+
+results3 %>%
+  filter(par.h2 == 0.8) %>%
+  plot_results("AUC") +
+  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10)

### 2017-11-17 tmp-results.Rmd
tmp results to discuss

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(fig.align = "center", out.width = "70%", fig.asp = 0.7)
+knitr::opts_chunk$set(fig.align = "center", out.width = "70%", fig.asp = 0.7,
+                      dev = "svg")
 options(width = 110)
@@
 
+```{r}
+boot <- function(x, n = 1e5, f = mean) {
+  sd(replicate(n, f(sample(x, replace = TRUE))))
+}
+```
+
 ## Results with T-Trees
@@
 
-```{r, fig.width=16, out.width="95%", dev="svg"}
+```{r, fig.width=16, out.width="95%", dev="svg", fig.cap="Results of T-Trees vs penalized logistic regression.
+**A.** Timing (in seconds).
+**B.** Number of predictors of the model.
+**C.** AUC.
+**D.** Percentage of cases in the 10% largest scores."}
 ttrees_vs_logit <- filter(results, method %in% c("T-Trees", "logit-simple"))
@@
 
-```{r, fig.width=14, out.width="90%", dev="svg"}
+```{r, fig.width=14, out.width="90%", dev="svg", fig.cap="Percentage of cases in the 2 highest deciles of PRSs as a function of AUC."}
 H2 <- 0.8
@@
 
-```{r}
+```{r, fig.cap="Results of AUC for all combination of parameters and methods when h2=0.5 and h2=0.8"}
 # represent h2=0.8 as a function of h2=0.5
@@
   group_by(par.causal, method) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, 1e4)) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, 1e5, mean)) %>%
   myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
@@
   summarise(AUC_mean = mean(AUC),
-            AUC_boot = boot(AUC, 1e4)) %>%
+            AUC_boot = boot(AUC, 1e5, mean)) %>%
   myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
@@
 
-```{r}
+```{r, fig.cap="Manhanttan plot for Celiac"}
 knitr::include_graphics("figures/celiac-man.png")
@@
 
-```{r}
+```{r, fig.cap="Regularization paths for the methods. For LR, line in the result given by CMSA."}
 knitr::include_graphics("figures/celiac-regpath.png")
@@
 
-```{r}
+```{r, fig.cap="AUC as function of M. For all chromosomes."}
 results2 %>%
@@
 
-```{r}
+```{r, fig.cap="AUC as function of M. For chromosome 6."}
 results3 %>%
@@
 
-```{r}
+```{r, fig.cap="AUCs for h2=0.8, dist=gaussian, comparing all chromosomes and chromosome 6."}
 bind_rows(
@@
 
-```{r, fig.cap="Size of effects GWAS vs logistic"}
+```{r, fig.cap="Size of effects GWAS vs logistic, for Celiac"}
 knitr::include_graphics("effects.png")
@@
 
-```{r}
+```{r, fig.cap="Density of scores from logistic regression by pop."}
 knitr::include_graphics("preds-density2.png")
@@
 
-```{r}
+```{r, fig.cap="Density of scores from logistic regression by pop and genotype."}
 knitr::include_graphics("preds-density3.png")

### 2017-11-20 3-simus.Rmd
save

@@
 
-```{r}
+```{r, eval=FALSE}
 if (!dir.exists("results1")) dir.create("results1")
@@
   )
-  print(res)
   params[["res"]] <- list(res)
@@
 
-### Second simulations (without T-Trees)
+### Second simulations (without T-Trees, with logit-triple)
 
@@
   par.model  = c("simple", "fancy"),
-  num.simu   = 1:50,
+  num.simu   = 1:20,
   stringsAsFactors = FALSE
@@
 
-```{r, eval=TRUE}
+```{r, eval=FALSE}
 if (!dir.exists("results2")) dir.create("results2")
@@
 
+### Third simulations (without logit-triple)
+
+```{r}
+params.grid3 <- expand.grid(
+  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
+  par.dist   = c("gaussian", "laplace"), 
+  par.h2     = c(0.5, 0.8), 
+  par.model  = c("simple"),
+  num.simu   = 21:100,
+  stringsAsFactors = FALSE
+) 
+```
+
 ```{r, eval=FALSE}
@@
 
-G6 <- big_copy(G, ind.col = which(CHR == 6))
-ind.HLA6 <- snp_indLRLDR(CHR[CHR == 6], POS[CHR == 6], subset(LD.wiki34, ID == "hild12"))
-
-for (i in rows_along(params.grid2)) {
+for (i in rows_along(params.grid3)) {
 
@@
 
-  params <- params.grid2[i, ]
+  params <- params.grid3[i, ]
+  par.causal <- params[["par.causal"]][[1]]
+
+  # Simulate phenotypes
+  pheno.all <- get_pheno(
+    G,    
+    h2 = params[["par.h2"]], 
+    M = as.integer(par.causal[1]), 
+    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
+    effects.dist = params[["par.dist"]], 
+    model = params[["par.model"]], 
+    K = 0.3                 
+  )
+
+  # Split in training/test sets
+  ind.train <- sort(sample(n, size = n.train))
+  ind.test <- setdiff(1:n, ind.train)
+
+  # Get results from all methods
+  res <- bind_rows(
+    PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test),
+    logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple")
+  )
+  params[["res"]] <- list(res)
+  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
+}
+```
+
+### Fourth simulations (dataset with only chromosome 6)
+
+```{r}
+params.grid4 <- expand.grid(
+  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
+  par.dist   = c("gaussian", "laplace"), 
+  par.h2     = c(0.5, 0.8), 
+  par.model  = c("simple"),
+  num.simu   = 1:100,
+  stringsAsFactors = FALSE
+) 
+```
+
+```{r, eval=FALSE}
+if (!dir.exists("results4")) dir.create("results4")
+
+G6 <- big_copy(G, ind.col = which(CHR == 6))
+ind.HLA6 <- snp_indLRLDR(CHR[CHR == 6], POS[CHR == 6], 
+                         subset(LD.wiki34, ID == "hild12"))
+
+for (i in rows_along(params.grid4)) {
+
+  res.file <- paste0("results4/simu_", i, ".rds")
+  if (file.exists(res.file)) next
+
+  params <- params.grid4[i, ]
   par.causal <- params[["par.causal"]][[1]]
@@
 ```
+
+### Fifth simulations (varying training size)
+
+```{r}
+params.grid5 <- expand.grid(
+  n.train    = 1:5 * 1000,
+  par.causal = list(c(300, "all")), 
+  par.dist   = c("gaussian", "laplace"), 
+  par.h2     = c(0.5, 0.8), 
+  par.model  = c("simple"),
+  num.simu   = 1:100,
+  stringsAsFactors = FALSE
+) 
+```
+
+```{r, eval=TRUE}
+if (!dir.exists("results5")) dir.create("results5")
+
+for (i in rows_along(params.grid5)) {
+
+  res.file <- paste0("results5/simu_", i, ".rds")
+  if (file.exists(res.file)) next
+
+  params <- params.grid5[i, ]
+  par.causal <- params[["par.causal"]][[1]]
+
+  # Simulate phenotypes
+  pheno.all <- get_pheno(
+    G,    
+    h2 = params[["par.h2"]], 
+    M = as.integer(par.causal[1]), 
+    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
+    effects.dist = params[["par.dist"]], 
+    model = params[["par.model"]], 
+    K = 0.3                 
+  )
+
+  # Split in training/test sets
+  ind.train <- sort(sample(n, size = params[["n.train"]]))
+  ind.test <- setdiff(1:n, ind.train)
+
+  # Get results from all methods
+  res <- bind_rows(
+    PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test),
+    logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple")
+  )
+  params[["res"]] <- list(res)
+  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
+}
+```

### 2017-11-20 paper/draft.tex
small modifs

@@
 Moreover, screening based on sequencing seems to make individuals positively change their health behavior while not causing patient anxiety nor depression \cite[]{vassy2017impact}.
-Yet, for PRSs to be used for helping in the diagnosis of disease, either confirming a diagnostic or even early diagnostic, PRSs would have to show a high discriminative power between cases and controls.
-For example, for screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be  greater than respectively 75\% and 99\% \cite[]{janssens2007impact}.
+Yet, PRSs would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases.
+For example, for respectively screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than respectively 75\% and 99\% \cite[]{janssens2007impact}.
 
-Many methods have been developed in order to maximize the predictive power of genotypes for diseases, or phenotypes in general.
-A commonly used technique, called P+T (which stands for ``Pruning + Thresholding'') -- or genetic profiling ..or even just PRS..
+Many methods have been developed in order to maximize the predictive power of genotypes for diseases, or more generally for phenotypes.
+A commonly used technique, called P+T (which stands for ``Pruning and Thresholding'') -- or genetic profiling ..or even just PRS..
 -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
@@
 for predicting a binary trait such as a disease status and have proven to ..fail..
-at such in another comparative study \cite[]{abraham2013performance}.
+at such task in another comparative study \cite[]{abraham2013performance}.
 Moreover, these methods and their ..derivations..
@@
 Statistical learning methods have also been used to derive PRSs for complex human disease by jointly estimating SNP effects.
-Such methods include logistic regression, Support Vector Machine (SVM) or random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2013walk,botta2014exploiting,wei2009disease}.
+Such methods include logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2013walk,botta2014exploiting,wei2009disease}.
 
@@
 Package bigstatsr includes efficient functions for computing penalized linear and logistic regressions on huge datasets.
-In this paper, we present a comprehensive comparative study of the P+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in \cite{botta2014exploiting}.
-Note that SVM is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp} and therefore isn't added to the comparison.
-For the P+T model, we compare different threshold of inclusion of SNPs.
+In this paper, we present a comprehensive comparative study of the P+T method, our penalized logistic regression and finally the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in \cite{botta2014exploiting}.
+Note that the SVM model is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp} and therefore isn't added to the comparison.
+For the P+T model, we compare different thresholds of inclusion of SNPs.
 For the logistic regression, we include two novel approaches. 
@@
 
-There have already been comparative studies about predicting phenotypes based on genotypes for human data. 
+There have already been comparative studies about predicting human binary phenotypes based on genotype data. 
 \citep{abraham2013performance} showed that penalized logistic regression and penalized SVM performed better than the standard P+T method and LMMs, for a wide-range of human diseases.

### 2017-11-23 3-simus.Rmd
save

@@
 
-### First simulations (with T-Trees)
+### Simulations n°1 (with T-Trees)
 
@@
 
-### Second simulations (without T-Trees, with logit-triple)
+### Simulations n°2 (without T-Trees, with logit-triple)
 
@@
 
-### Third simulations (without logit-triple)
+### Simulations n°3 (without logit-triple)
 
@@
 
-### Fourth simulations (dataset with only chromosome 6)
+### Simulations n°4 (dataset with only chromosome 6)
 
@@
 
-```{r, eval=FALSE}
+```{r, eval=TRUE}
 if (!dir.exists("results4")) dir.create("results4")
@@
 
-### Fifth simulations (varying training size)
+### Simulations n°5 (varying training size)
 

### 2017-11-23 paper/draft.tex
save

@@
 
-For simulations, we use real genotypes of European individuals from a case/control celiac disease cohort \cite[]{dubois2010multiple}. 
+We use real genotypes of European individuals from a case/control celiac disease cohort \cite[]{dubois2010multiple}. 
 Details of quality control and imputation for this dataset are available in another study \cite[]{prive2017efficient}.
-To get rid of the structure induced by the celiac disease status, we keep only controls from this cohort.
+For simulations, in order to get rid of the structure induced by the celiac disease status, we keep only controls from this cohort.
 Moreover, to get rid of population structure, we keep only people from UK and further subset these British people based on deviation of robust Mahalanobis distance on Principal Components (see ..SupMat..).
@@
 \newpage
+\begin{table}[htbp]
+\caption{Summary of all simulations. When there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.}
+\vspace*{0.5em}
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
+\hline
+Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & Number of & \multirow{2}{*}{Methods} \\ 
+simulation & & training set & (number and location) & of effects & & model & simulations & \\
+\hline
+\hline
+\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{4}{*}{0.8} & \multirow{2.5}{*}{simple} & \multirow{4}{*}{5} & PRS \\
+& & & 30 in all & & & & & logit-simple \\
+& & & 300 in all & \multirow{1.5}{*}{Laplace} & & \multirow{1.5}{*}{fancy} & & logit-triple \\
+& & & 3000 in all & & & & & T-Trees \\
+\hline
+\multirow{3}{*}{2} & \multirow{3}{*}{-} & \multirow{3}{*}{-} & \multirow{3}{*}{-} & \multirow{3}{*}{-} & \multirow{2}{*}{0.5} & \multirow{3}{*}{-} & \multirow{3}{*}{20} & PRS \\ 
+& & & & & \multirow{2.2}{*}{0.8} & & & logit-simple \\
+& & & & & & & & logit-triple \\
+\hline
+\multirow{2}{*}{3} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{simple} & \multirow{2}{*}{80} & PRS \\ 
+& & & & & & & & logit-simple \\
+\hline
+\multirow{2}{*}{4} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{100} & \multirow{2}{*}{-} \\
+& & & & & & & & \\
+\hline
+\multirow{5}{*}{5} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{100} & \multirow{5}{*}{-} \\
+& & 2000 & & & & & & \\
+& & 3000 & & & & & & \\
+& & 4000 & & & & & & \\
+& & 5000 & & & & & & \\
+\hline
+\end{tabular}
+\label{}
+\end{table}
+
+\newpage
 \begin{figure}[h]

### 2017-11-24 paper/draft.tex
save

@@
 \usepackage{dsfont}
+\usepackage[warn]{textcomp}
 \usepackage{adjustbox}
@@
 
-We use real genotypes of European individuals from a case/control celiac disease cohort \cite[]{dubois2010multiple}. 
-Details of quality control and imputation for this dataset are available in another study \cite[]{prive2017efficient}.
-For simulations, in order to get rid of the structure induced by the celiac disease status, we keep only controls from this cohort.
-Moreover, to get rid of population structure, we keep only people from UK and further subset these British people based on deviation of robust Mahalanobis distance on Principal Components (see ..SupMat..).
-This results in 7102 individuals genotyped on ..28x,xxx..
-SNPs [ADD DIM TO RMD] with minimal population structure (see ..SupMat..).
+We use real genotypes of European individuals from a case/control celiac disease cohort \cite[]{dubois2010multiple}. Composition of this dataset is presented in table \ref{tab:celiac-data}.
+Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
+For simulations, in order to remove the structure induced by the celiac disease status, we first keep only controls from this cohort.
+Secondly, in order to remove population structure, we keep only people from the UK.
+Thirdly, we further subset these British people based on deviation of robust Mahalanobis distance on principal components.
+This 3-step filtering results in 7102 individuals with minimal population structure (see ..SupMat..).
 We decided to get rid of population structure because it can affect the predictive performance of methods, which we will investigate in another study.
@@
 
-To simulate phenotypes, we use a the Liability Threshold Model (LTM) with a prevalence of 30\%.
-We vary different parameters: the number of causal variants and their location (30, 300 or 3000 anywhere in the genome or 30 only in the HLA region), the heritability (0.5 or 0.8), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part..
-of the phenotypes (one ``simple model'' with only linear effects, and one ``fancy model'' which combine linear, dominant and interaction effects).
-For the ``simple model'', we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights ($w_j \neq 0$ if and only if SNP $j$ is causal), $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs).
+To simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\%.
+We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only the chromosome 6), the size of the training set (from 1000 to 6000 by 1000, and the corresponding test set is composed of the remaining individuals), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part..
+of the phenotypes (one ``simple model'' with only linear effects, and one ``fancy model'' which combines linear, dominant and interaction effects).
+For the ``simple model'', we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights (generated with a Gaussian or a Laplace distribution), $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs).
 For the ``fancy model'', we separate the causal SNPs in three equal sets $E_\text{causal}^{(1)}$, $E_\text{causal}^{(2)}$ 
@@
 For both models, the $y_i$ are then standardized such that they have a variance equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$ where $\epsilon \sim N(0, 1 - h^2)$.
-This results in 32 combinations of parameters and we run 100 simulations for each of these combinations (5 times more than in \cite{zhou2013polygenic}), resulting of a total of 3200 simulations.
-For each of these 3200 simulations, we randomly re-sample a training set of 6000 individuals, with the corresponding test set composed of the remaining individuals on which we evaluate the different methods.
+
+We make 5 different simulation procedures, summarized in table \ref{tab:simus}. 
+..Simulation..
+\textnumero1 uses the whole dataset (the 22 chromosomes), a training set of size 6000 and an heritability of 80\%.
+It compares all methods (see section \ref{sec:methods}) on all other combination of parameters in 5 simulations.
+The goal of the first simulation is to quickly discard the T-Trees method.
+For simulation \textnumero2, we make also vary the heritability (50\% or 80\%) but we don't run the T-Trees method anymore.
+The goal of the second simulation is to compare all remaining methods, especially to see how well the ``logit-triple'' method works.
+We run 20 simulations.
+For simulation \textnumero3, we continue with 80 more simulations (to reach 100 simulations, i.e.\ 5 times more than in \cite{zhou2013polygenic}), but discarding the ``logit-triple'' method and the ``fancy'' model.
+Simulation \textnumero4 consists in 100 simulations of remaining methods and combinations of parameters on a dataset composed of only the chromosome 6.
+Reducing the number of SNPs aims at increasing the ..polygenicity..
+of our simulated models and to ..virtually..
+increase the sample size [REF?
+DUDBRIDGE?].
+Finally, simulation \textnumero5 reuses the whole dataset but varying the size of the training set in order to assess how it affects predictive performance of the PRS and the ``logit-simple'' methods.
+100 simulations are run with only 300 causal SNPs anywhere On the genome.
 
@@
 In this study, we use two different measures of predictive accuracy. 
-First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction}.
+First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction,hanley1982meaning,lusted1971signal}.
 In the case of our study, the AUC is the probability that a PRS of a case is greater than the PRS of a control.
@@
 This measure indicates how well we can identify high-risk individuals from a higher PRS in screening campaigns.
+[COMMENT ON PREVALENCE?]
 
@@
 
-\subsection{Compared methods}
+\subsection{Compared methods} \label{sec:methods}
 
@@
 \end{tabular}
-\label{}
+\label{tab:simus}
 \end{table}
@@
 \setcounter{table}{0}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{|l|cccc|r|}
+\hline
+Population & UK & Finland & Netherlands & Italy & Total\\
+\hline
+Cases & 2569 & 637 & 795 & 495 & 4496\\
+Controls & 7492 & 1799 & 828 & 540 & 10659\\
+\hline
+Total & 10061 & 2436 & 1623 & 1035 & 15155\\
+\hline
+\end{tabular}
+\caption{Number of individuals by population and disease status for the celiac disease cohort (after quality control, genotyped on 281,122 ..mutual.. SNPs).}
+\label{tab:celiac-data}
+\end{table}
 

### 2017-11-27 results1-2.R
save

@@ -1,0 +1,32 @@
+results_ttrees <- results1 %>%
+  filter(method %in% c("logit-simple", "T-Trees")) %>%
+  group_by_at(c(vars(starts_with("par")), "method")) %>%
+  summarise(
+    AUC_mean = mean(AUC), 
+    AUC_boot = boot(AUC, 1e4, mean), 
+    N = n(),
+    nb_preds_mean = mean(nb.preds),
+    timing_mean = mean(timing)
+  ) %>%
+  print(n = Inf)
+
+
+results_ttrees 
+rle(results_ttrees$N)
+
+myggplot(results_ttrees, aes(par.causal, AUC_mean, fill = method)) +
+  geom_hline(yintercept = 0.5, linetype = 2) +
+  geom_bar(stat = "identity", alpha = 0.5, position = "dodge") +
+  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
+                position=position_dodge(width = 0.9), color = "black", width = 0.2) +
+  facet_grid(par.model + par.h2 ~ par.dist) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1),
+        strip.text.x = element_text(size = rel(2)),
+        strip.text.y = element_text(size = rel(2))) +
+  scale_y_continuous(limits = c(0.5, 1), minor_breaks = 0:20 / 20,
+                     # https://stackoverflow.com/a/10365218/6103040
+                     oob = scales::rescale_none) + 
+  scale_fill_brewer(type = "qual", palette = 2) +
+  labs(x = "Causal SNPs (number and location)", y = "Mean of 5 AUCs", fill = "Method")
+
+ggsave("figures/ttrees.pdf", scale = 1/90, width = 560, height = 480)

### 2017-11-27 results2.R

@@ -1,0 +1,24 @@
+results_logit <- results2 %>%
+  filter(grepl("logit", method)) %>%
+  group_by_at(c(vars(starts_with("par")), "method")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, 1e4, mean), N = n()) 
+
+results_logit 
+rle(results_logit$N)
+
+myggplot(results_logit, aes(par.causal, AUC_mean, fill = method)) +
+  geom_hline(yintercept = 0.5, linetype = 2) +
+  geom_bar(stat = "identity", alpha = 0.5, position = "dodge") +
+  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
+                position=position_dodge(width = 0.9), color = "black", width = 0.2) +
+  facet_grid(par.model + par.h2 ~ par.dist) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1),
+        strip.text.x = element_text(size = rel(2)),
+        strip.text.y = element_text(size = rel(2))) +
+  scale_y_continuous(limits = c(0.5, 1), minor_breaks = 0:20 / 20,
+                     oob = scales::rescale_none) +
+  scale_fill_brewer(type = "qual", palette = 2) +
+  labs(x = "Causal SNPs (number and location)", y = "Mean of 20 AUCs",
+       fill = "Method")
+
+ggsave("figures/logit-triple.pdf", scale = 1/90, width = 600, height = 640)

### 2017-11-27 results3-2.R

@@ -1,0 +1,18 @@
+results3 <- list.files("results3", full.names = TRUE) %>%
+  map_dfr(~readRDS(.x)) %>%
+  as_tibble() %>%
+  mutate(
+    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
+                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
+    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2])),
+    percCases10 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[top10], 2])),
+    percCases20 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[top20], 2]))
+  ) %>%
+  bind_rows(filter(results2, method != "logit-triple"))
+
+
+results3 %>%
+  group_by_at(c(vars(starts_with("par")), "method")) %>%
+  summarise(N = n()) %>%
+  pull(N)
+

### 2017-11-27 results5.R

@@ -1,0 +1,40 @@
+results5 <- list.files("results5", full.names = TRUE) %>%
+  map_dfr(~readRDS(.x)) %>%
+  as_tibble() %>%
+  mutate(
+    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
+                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
+    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2]))
+  )
+
+pryr::object_size(results5)
+
+results5_3 <- results3 %>%
+  select(-eval) %>%
+  filter(par.model == "simple", par.causal == "300 in all") %>%
+  cbind(n.train = 6000, .) %>%
+  bind_rows(select(results5, -eval)) %>%
+  as_tibble()
+
+results5_3 %>%
+  group_by_at(c(vars(starts_with("par")), "method", "n.train")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, 1e4, mean), N = n()) %>%
+  print(n = Inf)
+
+tmp <- .Last.value
+
+myggplot(tmp, aes(n.train, AUC_mean, color = method)) +
+  geom_hline(yintercept = 0.5, linetype = 2) +
+  geom_point(size = 3) +
+  geom_line(size = 1.5) +
+  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, 
+                    ymax = AUC_mean + 2 * AUC_boot), 
+                size = 1.5, width = 0) +
+  facet_grid(par.model + par.h2 ~ par.causal + par.dist) +
+  theme(strip.text.x = element_text(size = rel(2)),
+        strip.text.y = element_text(size = rel(2))) +
+  scale_x_continuous(breaks = 1:6 * 1000, minor_breaks = NULL) + 
+  scale_y_continuous(minor_breaks = 0:20 / 20) +
+  scale_fill_brewer(type = "qual", palette = 2) +
+  labs(x = "Causal SNPs (number and location)", y = "Mean of 100 AUCs",
+       fill = "Method")

### 2017-11-27 paper/draft.tex
save

@@
 To simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\%.
-We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only the chromosome 6), the size of the training set (from 1000 to 6000 by 1000, and the corresponding test set is composed of the remaining individuals), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part..
+We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only the chromosome 6), the size of the training set (from 1000 to 6000 by 1000, and the corresponding test set is composed of the remaining individuals), the number of causal variants and their location (30, 300 or 3000 anywhere on the dataset or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part..
 of the phenotypes (one ``simple model'' with only linear effects, and one ``fancy model'' which combines linear, dominant and interaction effects).
@@
 The goal of the second simulation is to compare all remaining methods, especially to see how well the ``logit-triple'' method works.
-We run 20 simulations.
+We run 20 simulations for each combination of parameters.
 For simulation \textnumero3, we continue with 80 more simulations (to reach 100 simulations, i.e.\ 5 times more than in \cite{zhou2013polygenic}), but discarding the ``logit-triple'' method and the ``fancy'' model.
@@
 Reducing the number of SNPs aims at increasing the ..polygenicity..
-of our simulated models and to ..virtually..
-increase the sample size [REF?
+of our simulated models and at ..virtually..
+increasing the sample size [REF?
 DUDBRIDGE?].
-Finally, simulation \textnumero5 reuses the whole dataset but varying the size of the training set in order to assess how it affects predictive performance of the PRS and the ``logit-simple'' methods.
-100 simulations are run with only 300 causal SNPs anywhere On the genome.
+Finally, simulation \textnumero5 reuses the whole dataset but varying the size of the training set in order to assess how the number of samples affects predictive performance of the PRS and the ``logit-simple'' methods.
+100 simulations are run using 300 causal SNPs anywhere on the genome.
 
@@
 For the P+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
-The SNPs are first clumped (P) so that there remains only loci that are weakly correlated with each other. 
-Thresholding (T) consists in removing SNPs that are under a certain level of significance (p-value threshold to be determined). 
+The SNPs are first clumped (P) so that there remain only loci that are weakly correlated with each other. 
+Then, thresholding (T) consists in removing SNPs that are within a certain level of significance (p-value threshold to be determined). 
 Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$S_i(T) = \sum_{j \in E_\text{clumping}} \mathds{1}\{p_j < T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS and $T$ is a threshold on p-values to be determined.
-In this study, we report scores for a clumping threshold at $r^2 > 0.2$ with regions of 500kb.
+In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.
 
-For the P+T method, we report three different scores of prediction: one including all the SNPs (after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximizes the AUC for these two thresholds and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
+For the P+T method, we report three different scores of prediction: one including all the SNPs (remaining after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximizes the AUC for these two thresholds and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
 Note that, normally, the optimal threshold would have to be learned on the training set only.
@@
 
-All the code used in this paper along with results, such as figures, are available as HTML R notebooks in the Supplementary Materials.
+All the code used in this paper along with results such as figures and tables, are available as HTML R notebooks in the Supplementary Materials.
 
@@
 
-[REPLICATE INDIVIDUALS TO SHOW IF CAN BE USED ON LARGER DATA]
+\subsection{Simulations varying training size and number of SNPs}
 

### 2017-11-28 tmp-results.Rmd
save

@@
 # Put all results in a single tibble
-results <- list.files("results1", full.names = TRUE) %>%
+results1 <- list.files("results1", full.names = TRUE) %>%
   map_dfr(~readRDS(.x)) %>%
@@
 ```{r}
-results %>%
-  filter(method %in% c("T-Trees", "logit-simple")) %>%
+results1 %>%
   group_by_at(c(vars(starts_with("par")), "method")) %>%
@@
 **D.** Percentage of cases in the 10% largest scores."}
-ttrees_vs_logit <- filter(results, method %in% c("T-Trees", "logit-simple"))
+ttrees_vs_logit <- filter(results1, method %in% c("T-Trees", "logit-simple"))
 
@@
 
-```{r, fig.cap="AUC as function of M. For all chromosomes."}
-results2 %>%
-  filter(
-    # method %in% c("logit-simple", "PRS-max"),
-    grepl("all", par.causal)
-  ) %>%
-  mutate(M = readr::parse_number(par.causal)) %>%
-  group_by(M, par.dist, par.h2, par.model, method) %>%
-  summarise(AUC = mean(AUC)) %>%
-  # filter(par.h2 == 0.8, par.dist == "gaussian", par.model == "simple") %>%
-  filter(par.h2 == 0.8) %>%
-  myggplot(aes(M, AUC, color = method)) +
-  geom_line() + 
-  geom_point(size = 2) +
-  facet_grid(par.dist ~ par.model) +
-  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10) +
-  scale_x_sqrt(breaks = c(30, 300, 1200, 3000)) +
-  labs(x = "Number of causal SNPs (sqrt-scale)")
-```
-
-```{r, fig.cap="AUC as function of M. For chromosome 6."}
-results3 %>%
-  filter(
-    # method %in% c("logit-simple", "PRS-max"),
-    grepl("all", par.causal)
-  ) %>%
-  mutate(M = readr::parse_number(par.causal)) %>%
-  group_by(M, par.dist, par.h2, par.model, method) %>%
-  summarise(AUC = mean(AUC)) %>%
-  # filter(par.h2 == 0.8, par.dist == "gaussian", par.model == "simple") %>%
-  filter(par.h2 == 0.8) %>%
-  myggplot(aes(M, AUC, color = method)) +
-  geom_line() + 
-  geom_point(size = 2) +
-  facet_grid(par.dist ~ par.model) +
-  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10) +
-  scale_x_sqrt(breaks = c(30, 300, 1200, 3000)) +
-  labs(x = "Number of causal SNPs (sqrt-scale)")
-```
-
 ```{r, fig.cap="AUCs for h2=0.8, dist=gaussian, comparing all chromosomes and chromosome 6."}

### 2017-11-28 results2.R

@@ -6,7 +6,7 @@ results_logit <- results2 %>%
 results_logit 
 rle(results_logit$N)
 
-myggplot(results_logit, aes(par.causal, AUC_mean, fill = method)) +
+myggplot(results_logit, aes(par.causal, AUC_mean, fill = method, color = method)) +
   geom_hline(yintercept = 0.5, linetype = 2) +
   geom_bar(stat = "identity", alpha = 0.5, position = "dodge") +
   geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
@@ -18,7 +18,8 @@ myggplot(results_logit, aes(par.causal, AUC_mean, fill = method)) +
   scale_y_continuous(limits = c(0.5, 1), minor_breaks = 0:20 / 20,
                      oob = scales::rescale_none) +
   scale_fill_brewer(type = "qual", palette = 2) +
+  scale_color_brewer(type = "qual", palette = 2) +
   labs(x = "Causal SNPs (number and location)", y = "Mean of 20 AUCs",
-       fill = "Method")
+       fill = "Method", color = "Method")
 
 ggsave("figures/logit-triple.pdf", scale = 1/90, width = 600, height = 640)

### 2017-11-28 results3-2.R

@@ -4,15 +4,50 @@ results3 <- list.files("results3", full.names = TRUE) %>%
   mutate(
     par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
                         levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
-    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2])),
-    percCases10 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[top10], 2])),
-    percCases20 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[top20], 2]))
+    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2]))
   ) %>%
-  bind_rows(filter(results2, method != "logit-triple"))
+  bind_rows(filter(results2, method != "logit-triple", par.model != "fancy"))
 
 
 results3 %>%
   group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(N = n()) %>%
-  pull(N)
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, 1e4, mean), N = n()) %>%
+  print(n = Inf)
 
+results3 %>%
+  filter(par.h2 == 0.8, method %in% c("logit-simple", "PRS-max")) %>%
+  group_by_at(c(vars(starts_with("par")), "method")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, 1e4, mean)) %>%
+  myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
+  geom_hline(yintercept = 0.5, linetype = 2) +
+  geom_hline(yintercept = 0.94, linetype = 3, color = "blue") +
+  geom_bar(stat = "identity", alpha = 0.3, position=position_dodge()) +
+  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
+                position=position_dodge(width=0.9), color = "black", width = 0.2) +
+  scale_y_continuous(limits = c(0.5, NA), minor_breaks = 0:20 / 20,
+                     oob = scales::rescale_none) +
+  scale_fill_brewer(type = "qual", palette = 2) +
+  scale_color_brewer(type = "qual", palette = 2) +
+  labs(x = "Causal SNPs (number and location)", y = "Mean of 100 AUCs",
[…]

### 2017-11-28 results4.R

@@ -1,0 +1,44 @@
+results4 <- list.files("results4", full.names = TRUE) %>%
+  map_dfr(~readRDS(.x)) %>%
+  mutate(
+    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
+                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
+    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2]))
+  ) %>%
+  cbind(simu = "chr6") %>%
+  bind_rows(cbind(select(results3, -percCases10, -percCases20), simu = "all")) %>%
+  as_tibble()
+
+results4 %>%
+  filter(par.h2 == 0.8) %>%
+  myggplot() +
+  geom_boxplot(aes(method, AUC, fill = simu, color = simu), alpha = 0.3) + 
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  facet_grid(par.model ~ par.causal) +
+  theme(strip.text.x = element_text(size = rel(2)),
+        strip.text.y = element_text(size = rel(2))) +
+  scale_colour_brewer(palette = "Set1") +
+  scale_fill_brewer(palette = "Set1") +
+  geom_hline(yintercept = 0.94, color = "blue", linetype = 2) +
+  facet_grid(par.dist ~ par.model + par.h2 + par.causal)
+
+
+results4 %>%
+  filter(par.h2 == 0.8, method %in% c("logit-simple", "PRS-max"), simu == "chr6") %>%
+  group_by_at(c(vars(starts_with("par")), "method")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, 1e4, mean)) %>%
+  myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
+  geom_hline(yintercept = 0.5, linetype = 2) +
+  geom_hline(yintercept = 0.94, linetype = 3, color = "blue") +
+  geom_bar(stat = "identity", alpha = 0.3, position=position_dodge()) +
+  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
+                position=position_dodge(width=0.9), color = "black", width = 0.2) +
+  scale_y_continuous(limits = c(0.5, NA), minor_breaks = 0:20 / 20,
+                     oob = scales::rescale_none) +
[…]

### 2017-11-30 auc-max-h2.R
max AUC

@@ -1,0 +1,24 @@
+nsimu <- 1e7
+K <- 0.3
+
+replicate(10, {
+  sapply(c(0.5, 0.8), function(h2) {
+    S <- rnorm(nsimu, mean = 0, sd = sqrt(h2))
+    E <- rnorm(nsimu, mean = 0, sd = sqrt(1 - h2))
+    Y <- (S + E) > qnorm(1 - K)
+    
+    bigstatsr::AUC(S, Y)
+  })
+})
+
+# Estimation from Wray 2010
+sapply(c(0.5, 0.8), function(h2) {
+  T0 <- qnorm(1 - K)
+  z <- dnorm(T0)
+  i <- z / K
+  v <- -i * K / (1 - K)
+  num <- (i - v) * h2
+  deno.part <- 1 - h2 * i * (i - T0) + (1 - h2 * v * (v - T0))
+  pnorm(num / sqrt(h2 * deno.part))
+})
+

### 2017-12-01 paper/draft.tex
save

@@
 
-Polygenic Risk Scores (PRSs) combine the information contained in many single-nucleotide polymorphisms (SNPs) in a score that aims at reflecting the risk of developing a disease for which this score has been created.
+Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
 PRSs have proven to be useful for different applications, such as finding a common genetic contribution between two diseases \cite[]{purcell2009common}.
@@
 Package bigstatsr includes efficient functions for computing penalized linear and logistic regressions on huge datasets.
-In this paper, we present a comprehensive comparative study of the P+T method, our penalized logistic regression and finally the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in \cite{botta2014exploiting}.
+In this paper, we present a comprehensive comparative study of the P+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in \cite{botta2014exploiting}.
 Note that the SVM model is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp} and therefore isn't added to the comparison.

### 2017-12-04 3-oracle.Rmd
save

---
title: "Oracle"
author: "Florian Privé"
date: "November 17, 2017"
output: 
  html_document:
    code_folding: show
bibliography: paper/refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
                      fig.asp = 0.7, dev = "svg")
options(width = 110)
```

## Data and useful functions

```{r, message=FALSE}
library(tidyverse)
library(bigsnpr)
library(foreach)
celiac2 <- snp_attach("backingfiles/celiacQC_sub1.rds")
CHR <- celiac2$map$chromosome
POS <- celiac2$map$physical.pos
(G <- celiac2$genotypes)
# (G2 <- big_attach("backingfiles/celiacQC_sub1_tripled1.rds"))
# 
# covar.all <- readRDS("backingfiles/PCA2.rds")$u
# 
# n <- nrow(G)
# n.train <- 6e3
ind.HLA <- snp_indLRLDR(CHR, POS, subset(LD.wiki34, ID == "hild12"))
NCORES <- nb_cores()
```

```{r}
myggplot <- function(..., coeff = 1) {
  bigstatsr:::MY_THEME(ggplot(...), coeff = coeff)
} 
```

```{r}
boot <- function(x, n = 1e5, f = mean) {
  sd(replicate(n, f(sample(x, replace = TRUE))))
}
```

## Get maximum AUCs

```{r}
K <- 0.3
```

### Estimation from Wray 2010

Computation of the maximum AUC from equation (3) of @wray2010genetic. Approximation is less accurate for high heritabilities.

```{r}
sapply(c(0.5, 0.8), function(h2) {
  T0 <- qnorm(1 - K)
  z <- dnorm(T0)
  i <- z / K
  v <- -i * K / (1 - K)
  num <- (i - v) * h2
  deno.part <- 1 - h2 * i * (i - T0) + (1 - h2 * v * (v - T0))
  pnorm(num / sqrt(h2 * deno.part))
})
```

### Common estimation for all simulations

Assuming that $y \sim N(0, h^2)$ and knowing that $\epsilon \sim N(0, 1-h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the heritability $h^2$ and a liability threshold model on $y + \epsilon$ with a prevalence of 30%.

```{r}
nsimu <- 1e7

doParallel::registerDoParallel(NCORES)
aucs <- foreach(i = seq_len(100), .combine = "cbind") %:%
  foreach(h2 = c(0.5, 0.8), .combine = "rbind") %dopar% {
    S <- rnorm(nsimu, mean = 0, sd = sqrt(h2))
    E <- rnorm(nsimu, mean = 0, sd = sqrt(1 - h2))
    Y <- (S + E) > qnorm(1 - K)

    AUC(S, Y)
  }
doParallel::stopImplicitCluster()
apply(aucs, 1, summary)
```

### Different estimations for all simulations

```{r}
get_oracle <- function(
  G,                                        ## matrix of genotypes
  h2,                                       ## heritability 
  M,                                        ## nbs of causal variants
  ind.possible = cols_along(G),             ## indices of possible causal variants
  effects.dist = c("gaussian", "laplace"),  ## distribution of effects 
  model = c("simple", "fancy"),             ## model for simulation
  K = 0.3                                   ## prevalence
) {

  effects.dist  <- match.arg(effects.dist)
  model <- match.arg(model)

  set <- sample(ind.possible, size = M)
  effects <- `if`(effects.dist == "gaussian", 
                  rnorm(M, sd = sqrt(h2 / M)),
                  rmutil::rlaplace(M, s = sqrt(h2 / (2*M))))

  if (model == "simple") {
    # only linear
    y.simu <- scale(G[, set]) %*% effects
  } else {
    sets <- split(1:M, sample(rep_len(1:3, M)))
    # linear
    ind1 <- sets[[1]]
    y.simu <- scale(G[, set[ind1]]) %*% effects[ind1]
    # recessive / dominant
    ind2 <- sets[[2]]
    y.simu <- y.simu + scale(G[, set[ind2]] > 0.5) %*% effects[ind2]
    # interactions
    ind3 <- matrix(sets[[3]], ncol = 2)
    y.simu <- y.simu + scale(G[, set[ind3[, 1]]] * G[, set[ind3[, 2]]]) %*% 
      effects[ind3[, 1]] * sqrt(2)
  }

  y.simu <- y.simu / sd(y.simu) * sqrt(h2)
  stopifnot(all.equal(drop(var(y.simu)), h2))
  S <- y.simu
  y.simu <- S + rnorm(nrow(G), sd = sqrt(1 - h2))
  pheno <- as.numeric(y.simu > qnorm(1 - K))
  AUC(S, pheno)
}
```

```{r}
params.grid2 <- expand.grid(
  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
  par.dist   = c("gaussian", "laplace"), 
  par.h2     = c(0.5, 0.8), 
  par.model  = c("simple", "fancy"),
  num.simu   = 1:100,
  stringsAsFactors = FALSE
) 
```

```{r}
AUC_oracle <- function(G, params, ind.HLA) {

  par.causal <- params[["par.causal"]][[1]]

  # Simulate phenotypes
  AUC <- get_oracle(
    G,    
    h2 = params[["par.h2"]], 
    M = as.integer(par.causal[1]), 
    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
    effects.dist = params[["par.dist"]], 
    model = params[["par.model"]], 
    K = 0.3                 
  )

  params$AUC <- AUC
  params
}
```

#### On all 22 chromosomes

```{r}
doParallel::registerDoParallel(NCORES)
oracles <- foreach(i = rows_along(params.grid2), .combine = "rbind") %dopar% {
  AUC_oracle(G, params.grid2[i, ], ind.HLA)
}
doParallel::stopImplicitCluster()
```

```{r}
oracles %>% 
  mutate(
    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all")))) %>%
  group_by(par.causal, par.dist, par.h2, par.model) %>%
  summarise(AUC_mean = round(mean(AUC), 3),
            AUC_mean_boot = round(boot(AUC, 1e4, mean), 4)) %>%
  print(n = Inf)
```

#### On only the chromosome 6

```{r}
G6 <- big_copy(G, ind.col = which(CHR == 6))
ind.HLA6 <- snp_indLRLDR(CHR[CHR == 6], POS[CHR == 6], 
                         subset(LD.wiki34, ID == "hild12"))

doParallel::registerDoParallel(NCORES)
oracles_chr6 <- foreach(i = rows_along(params.grid2), .combine = "rbind") %dopar% {
  AUC_oracle(G6, params.grid2[i, ], ind.HLA6)
}
doParallel::stopImplicitCluster()
```

```{r}
tmp <- oracles_chr6 %>% 
  mutate(
    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all")))) %>%
  group_by(par.causal, par.dist, par.h2, par.model) %>%
  summarise(AUC_mean = round(mean(AUC), 4),
            AUC_mean_boot = round(boot(AUC, 1e4, mean), 4)) %>%
  print(n = Inf)
```

## References

### 2017-12-04 paper/draft.tex
save

@@
 
-\subsection{Predictive performance measures}
+\subsection{Predictive performance measures}\label{sec:auc}
 
@@
 
+\subsection{Maximum AUCs}
+
+[MOVE IN SUPMAT AND METHODS]
+
+Assuming that $y \sim N(0, h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the heritability $h^2$ (see ..SupMat..).
+We report AUCs of 0.8410 and 0.9406 for respectively an heritability of 50\% and 80\%.
+% wray2010genetic
+
 \subsection{Overview and naming}
@@
 
+\subsection{Scenario \textnumero1}
+
+In scenario \textnumero1, we compared our penalized logistic regression (``logit-simple'') with the P+T method (``PRS'') on the whole genome (Table \ref{tab:simus}). 
+When there are only 30 causal SNPs, our penalized logistic regression almost reach the maximum achievable AUC (Figures \ref{fig:main-AUC-logit} \& \ref{fig:supp-AUC-logit}).
+Still in terms of AUC, our penalized regression consistently performs better or equally to the ``PRS-max''. 
+In the case where there are correlation between predictors (i.e.\ when we simulate 30 SNPs in the HLA region), the ``logit-simple'' method particularly performs better than the ``PRS-max'' one, with a mean AUC of 92.5\% versus 83\%, for a maximum achievable AUC of 94\% (Figure \ref{fig:main-AUC-logit}).
+[NUL] Moreover, results of PRS can dramatically drop given the threshold of inclusion of SNPs in the model (Figures \ref{fig:main-AUC-PRS} \& \ref{fig:supp-AUC-PRS}).
+
 \subsection{Main simulations}
-
-[Also return p-values?]\\
 
@@
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{AUC}}
-\caption{AUC...}\label{fig:AUC}
+\centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
+\caption{Mean of AUC over 100 simulations for our penalized logistic regression (``logit-simple'') and the maximum AUC reported with the P+T method (``PRS-max'').
+Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+The blue dotted line represents the maximum achievable AUC.
+[MODEL AND H2]}\label{fig:main-AUC-logit}
 \end{figure}
@@
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{AUC-rel}}
-\caption{Relative AUC...}\label{fig:AUC-rel}
+\centerline{\includegraphics[width=\textwidth]{main-AUC-PRS}}
+\caption{Mean of AUC over 100 simulations for ..the three reported PRSs...
+Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+The blue dotted line represents the maximum achievable AUC.
+[MODEL AND H2]}\label{fig:main-AUC-PRS}
 \end{figure}
@@
 
+\newpage
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{ttrees}}
-\caption{Results of T-Trees vs penalized logistic regression.
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-logit}}
+\caption{[COPY FOR H2=0.5]}\label{fig:supp-AUC-logit}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS}}
+\caption{[COPY FOR H2=0.5]}\label{fig:supp-AUC-PRS}
+\end{figure}
+
+%\begin{figure}[h]
+%\centerline{\includegraphics[width=\textwidth]{ttrees}}
+%\caption{Results of T-Trees vs penalized logistic regression.
 \textbf{A.} Timing (in seconds).
@@
 \textbf{D.} Percentage of cases in the 10\% largest scores.}\label{fig:ttrees}
-\end{figure}
+%\end{figure}
 
@@
 
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{AUC-corr}}
-\caption{Percentage of cases in the 2 highest deciles of PRSs as a function of AUC.}\label{fig:AUC-corr}
-\end{figure}
+%\begin{figure}[h]
+%\centerline{\includegraphics[width=\textwidth]{AUC-corr}}
+%\caption{Percentage of cases in the 2 highest deciles of PRSs as a function of AUC.}\label{fig:AUC-corr}
+%\end{figure}
 
@@
 
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{meme}}
-\caption{Meme to use on Twitter when promoting the preprint?}\label{fig:meme}
-\end{figure}
-
 \end{document}

### 2017-12-04 results1.R

@@ -1,84 +1,56 @@
-
 library(tidyverse)
 
-myggplot <- function(..., coeff = 1) {
-  bigstatsr:::MY_THEME(ggplot(...), coeff = coeff)
-} 
-
-top10 <- 1:110
-top20 <- 1:220
-
-results <- list.files("../thesis/paper2-PRS/results1/", full.names = TRUE) %>%
+results1 <- list.files("results1", full.names = TRUE) %>%
   map_dfr(~readRDS(.x)) %>%
-  as_tibble() %>%
-  mutate(
-    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
-                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
-    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2])),
-    percCases10 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[top10], 2])),
-    percCases20 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[top20], 2]))
-  )
+  format_results()
 
-print(results, n = Inf)
-
-results %>%
-  filter(method == "T-Trees") %>%
-  pull(timing)
-
-results %>%
+results_ttrees <- results1 %>%
+  filter(method %in% c("logit-simple", "T-Trees")) %>%
   group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise_at(c("timing", "nb.preds", "AUC", "percCases10"), mean)
+  summarise(
+    AUC_mean = mean(AUC), 
+    AUC_boot = boot(AUC, 1e5, mean), 
+    N = n(),
+    nb_preds_mean = mean(nb.preds),
+    timing_mean = mean(timing)
+  ) %>%
+  print(n = Inf)
 
-plot_results <- function(results, y, ylab = y) {
+p1 <- barplot_causal(results_ttrees, h2 = 0.8) +
+  facet_grid(par.model ~ par.dist) +
+  labs(y = "Mean of 5 AUCs")
+
+plot_ttrees <- function(results, y, ylab = y) {
   
-  dist <- "Distribution\nof effects"
-  
-  myggplot(results) +
-    geom_boxplot(aes_string("method", y, color = "par.dist", 
-                            fill = "par.dist"), alpha = 0.3) + 
+  results %>%
+    filter(method %in% c("logit-simple", "T-Trees")) %>%
+    myggplot() +
[…]

### 2017-12-04 results3.R

@@ -1,14 +1,107 @@
-# Put all results in a single tibble
-results3 <- list.files("results3", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  as_tibble() %>%
-  mutate(
-    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
-                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
-    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2]))
-  )
+library(tidyverse)
 
-results3 %>%
-  filter(par.h2 == 0.8) %>%
-  plot_results("AUC") +
-  scale_y_continuous(breaks = 0:10 / 10, minor_breaks = c(0:9 + 0.5) / 10)
+format_results <- function(results) {
+  results %>%
+    as_tibble() %>%
+    mutate(
+      par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
+                          levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
+      AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2]))
+    )
+}
+
+results3 <- list.files(paste0("results", 2:3), full.names = TRUE) %>%
+  map_dfr(~readRDS(.x)) %>%
+  filter(method != "logit-triple", par.model != "fancy") %>%
+  format_results()
+
+
+results3.summary <- results3 %>%
+  group_by_at(c(vars(starts_with("par")), "method")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
+  print(n = Inf)
+
+## Make a function of this
+
+barplot_causal <- function(results, h2) {
+  
+  auc_max <- `if`(h2 == 0.8, 0.94, 0.84)
+  
+  results %>%
+    filter(par.h2 == h2) %>%
+    myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
+    geom_hline(yintercept = 0.5, linetype = 2) +
+    geom_hline(yintercept = auc_max, linetype = 3, color = "blue") +
+    geom_bar(stat = "identity", alpha = 0.3, position = position_dodge()) +
+    geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
[…]

### 2017-12-05 tmp-results.Rmd
save

@@
 myggplot <- function(..., coeff = 1) {
-  bigstatsr:::MY_THEME(ggplot(...), coeff = coeff)
+  bigstatsr:::MY_THEME(ggplot(...), coeff = coeff) +
+  theme(strip.text.x = element_text(size = rel(2)),
+        strip.text.y = element_text(size = rel(2)))
 } 
@@
 ```{r}
-boot <- function(x, n = 1e5, f = mean) {
+boot <- function(x, n = 1e4, f = mean) {
   sd(replicate(n, f(sample(x, replace = TRUE))))

### 2017-12-05 results1-2.R

@@ -1,32 +1,0 @@
-results_ttrees <- results1 %>%
-  filter(method %in% c("logit-simple", "T-Trees")) %>%
-  group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(
-    AUC_mean = mean(AUC), 
-    AUC_boot = boot(AUC, 1e4, mean), 
-    N = n(),
-    nb_preds_mean = mean(nb.preds),
-    timing_mean = mean(timing)
-  ) %>%
-  print(n = Inf)
-
-
-results_ttrees 
-rle(results_ttrees$N)
-
-myggplot(results_ttrees, aes(par.causal, AUC_mean, fill = method)) +
-  geom_hline(yintercept = 0.5, linetype = 2) +
-  geom_bar(stat = "identity", alpha = 0.5, position = "dodge") +
-  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
-                position=position_dodge(width = 0.9), color = "black", width = 0.2) +
-  facet_grid(par.model + par.h2 ~ par.dist) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1),
-        strip.text.x = element_text(size = rel(2)),
-        strip.text.y = element_text(size = rel(2))) +
-  scale_y_continuous(limits = c(0.5, 1), minor_breaks = 0:20 / 20,
-                     # https://stackoverflow.com/a/10365218/6103040
-                     oob = scales::rescale_none) + 
-  scale_fill_brewer(type = "qual", palette = 2) +
-  labs(x = "Causal SNPs (number and location)", y = "Mean of 5 AUCs", fill = "Method")
-
-ggsave("figures/ttrees.pdf", scale = 1/90, width = 560, height = 480)

### 2017-12-05 results1.R

@@ -7,50 +7,64 @@ results1 <- list.files("results1", full.names = TRUE) %>%
 results_ttrees <- results1 %>%
   filter(method %in% c("logit-simple", "T-Trees")) %>%
   group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(
-    AUC_mean = mean(AUC), 
-    AUC_boot = boot(AUC, 1e5, mean), 
-    N = n(),
-    nb_preds_mean = mean(nb.preds),
-    timing_mean = mean(timing)
-  ) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
   print(n = Inf)
 
-p1 <- barplot_causal(results_ttrees, h2 = 0.8) +
-  facet_grid(par.model ~ par.dist) +
-  labs(y = "Mean of 5 AUCs")
 
-plot_ttrees <- function(results, y, ylab = y) {
+
+compare_logit <- function(results, with_method) {
   
-  results %>%
-    filter(method %in% c("logit-simple", "T-Trees")) %>%
-    myggplot() +
-    geom_boxplot(aes_string("par.causal", y, color = "method", 
-                            fill = "method"), alpha = 0.3) + 
-    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  results <- filter(
+    results, method %in% c("logit-simple", with_method), par.h2 == 0.8)
+  
+  results.summary <- results %>%
+    group_by_at(c(vars(starts_with("par")), "method")) %>%
+    summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
+    print(n = Inf)
+  
+  nsimu <- results.summary$N[[1]]
+  stopifnot(all(results.summary$N == nsimu))
+  
+  plot_extra <- function(y, ylab = y) {
+    
+    results %>%
+      myggplot() +
+      geom_boxplot(aes_string("par.causal", y, color = "method", 
+                              fill = "method"), alpha = 0.3) + 
+      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+      facet_grid(par.model ~ par.dist) +
+      labs(x = "Causal SNPs (number and location)", y = ylab, 
+           fill = "Method", color = "Method") +
+      theme(legend.position = "none") +
+      scale_fill_manual(values = methods.color) +
[…]

### 2017-12-05 results2.R

@@ -1,25 +1,45 @@
-results_logit <- results2 %>%
-  filter(grepl("logit", method)) %>%
+results2 <- list.files("results2", full.names = TRUE) %>%
+  map_dfr(~readRDS(.x)) %>%
+  format_results()
+
+results2.summary <- results2 %>%
   group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, 1e4, mean), N = n()) 
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
+  print(n = Inf) 
 
-results_logit 
-rle(results_logit$N)
+barplot_causal_all <- function(results) {
+  
+  results %>%
+    barplot_causal() +
+    geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
+               data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
+    facet_grid(par.dist ~ par.h2)
+}
 
-myggplot(results_logit, aes(par.causal, AUC_mean, fill = method, color = method)) +
-  geom_hline(yintercept = 0.5, linetype = 2) +
-  geom_bar(stat = "identity", alpha = 0.5, position = "dodge") +
-  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
-                position=position_dodge(width = 0.9), color = "black", width = 0.2) +
-  facet_grid(par.model + par.h2 ~ par.dist) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1),
-        strip.text.x = element_text(size = rel(2)),
-        strip.text.y = element_text(size = rel(2))) +
-  scale_y_continuous(limits = c(0.5, 1), minor_breaks = 0:20 / 20,
-                     oob = scales::rescale_none) +
-  scale_fill_brewer(type = "qual", palette = 2) +
-  scale_color_brewer(type = "qual", palette = 2) +
-  labs(x = "Causal SNPs (number and location)", y = "Mean of 20 AUCs",
-       fill = "Method", color = "Method")
+results2.summary %>%
+  filter(par.model == "fancy", method %in% c("logit-simple", "PRS-max")) %>%
+  barplot_causal_all() +
+  labs(y = "Mean of 20 AUCs")
+  
[…]

### 2017-12-05 results3-2.R

@@ -1,53 +1,0 @@
-results3 <- list.files("results3", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  as_tibble() %>%
-  mutate(
-    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
-                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
-    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2]))
-  ) %>%
-  bind_rows(filter(results2, method != "logit-triple", par.model != "fancy"))
-
-
-results3 %>%
-  group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, 1e4, mean), N = n()) %>%
-  print(n = Inf)
-
-results3 %>%
-  filter(par.h2 == 0.8, method %in% c("logit-simple", "PRS-max")) %>%
-  group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, 1e4, mean)) %>%
-  myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
-  geom_hline(yintercept = 0.5, linetype = 2) +
-  geom_hline(yintercept = 0.94, linetype = 3, color = "blue") +
-  geom_bar(stat = "identity", alpha = 0.3, position=position_dodge()) +
-  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
-                position=position_dodge(width=0.9), color = "black", width = 0.2) +
-  scale_y_continuous(limits = c(0.5, NA), minor_breaks = 0:20 / 20,
-                     oob = scales::rescale_none) +
-  scale_fill_brewer(type = "qual", palette = 2) +
-  scale_color_brewer(type = "qual", palette = 2) +
-  labs(x = "Causal SNPs (number and location)", y = "Mean of 100 AUCs",
-       fill = "Method", color = "Method") +
-  facet_grid(par.dist ~ par.model + par.h2) + 
-  theme(strip.text.x = element_text(size = rel(2)),
-        strip.text.y = element_text(size = rel(2)))
-
-results3 %>%
-  filter(par.h2 == 0.8, grepl("PRS", method)) %>%
[…]

### 2017-12-05 results3.R

@@ -1,3 +1,7 @@
+methods.color <- setNames(scales::hue_pal()(6), 
+                          c("logit-simple", "PRS-max", "PRS-stringent",
+                            "T-Trees", "logit-triple", "PRS-all")) 
+
 library(tidyverse)
 
 format_results <- function(results) {
@@ -25,50 +29,52 @@ results3.summary <- results3 %>%
 
 barplot_causal <- function(results, h2) {
   
+  results %>%
+    myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
+    geom_hline(yintercept = 0.5, linetype = 2) +
+    geom_bar(stat = "identity", alpha = 0.3, position = position_dodge()) +
+    geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
+                  position = position_dodge(width = 0.9), color = "black", width = 0.2) + 
+    scale_y_continuous(limits = c(0.5, NA), minor_breaks = 0:20 / 20,
+                       oob = scales::rescale_none) +
+    labs(x = "Causal SNPs (number and location)", y = "Mean of 100 AUCs",
+         fill = "Method", color = "Method") +
+    scale_fill_manual(values = methods.color) +
+    scale_color_manual(values = methods.color)
+}
+
+barplot_causal_one <- function(results, h2) {
+  
   auc_max <- `if`(h2 == 0.8, 0.94, 0.84)
   
   results %>%
     filter(par.h2 == h2) %>%
-    myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
-    geom_hline(yintercept = 0.5, linetype = 2) +
+    barplot_causal() +
     geom_hline(yintercept = auc_max, linetype = 3, color = "blue") +
-    geom_bar(stat = "identity", alpha = 0.3, position = position_dodge()) +
-    geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
-                  position=position_dodge(width=0.9), color = "black", width = 0.2) +
-    facet_grid(par.dist ~ .) + 
-    scale_y_continuous(limits = c(0.5, NA), minor_breaks = 0:20 / 20,
-                       oob = scales::rescale_none) +
[…]

### 2017-12-05 results4.R

@@ -1,44 +1,50 @@
 results4 <- list.files("results4", full.names = TRUE) %>%
   map_dfr(~readRDS(.x)) %>%
-  mutate(
-    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
-                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
-    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2]))
-  ) %>%
-  cbind(simu = "chr6") %>%
-  bind_rows(cbind(select(results3, -percCases10, -percCases20), simu = "all")) %>%
-  as_tibble()
+  format_results()
 
 results4 %>%
-  filter(par.h2 == 0.8) %>%
+  group_by_at(c(vars(starts_with("par")), "method")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
+  print(n = Inf)
+
+results4 <- list.files("results4", full.names = TRUE) %>%
+  map_dfr(~readRDS(.x)) %>%
+  format_results() %>%
+  cbind(simu = "figures/supp-AUC-chr6") %>%
+  bind_rows() %>%
+  as_tibble()
+
+results34 <- rbind(
+  cbind(simu = "all", results3),
+  cbind(simu = "chr6", results4)
+) %>%
+  as_tibble()
+
+results34 %>%
   myggplot() +
+  geom_hline(yintercept = 0.5, linetype = 2) +
+  geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
+             data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
   geom_boxplot(aes(method, AUC, fill = simu, color = simu), alpha = 0.3) + 
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   facet_grid(par.model ~ par.causal) +
-  theme(strip.text.x = element_text(size = rel(2)),
-        strip.text.y = element_text(size = rel(2))) +
-  scale_colour_brewer(palette = "Set1") +
-  scale_fill_brewer(palette = "Set1") +
-  geom_hline(yintercept = 0.94, color = "blue", linetype = 2) +
-  facet_grid(par.dist ~ par.model + par.h2 + par.causal)
-
-
-results4 %>%
-  filter(par.h2 == 0.8, method %in% c("logit-simple", "PRS-max"), simu == "chr6") %>%
[…]

### 2017-12-12 4-results-simus.Rmd
results simus

---
title: "Results"
author: "Florian Privé"
date: "November 17, 2017"
output: 
  html_document:
    code_folding: hide
params:
  ggsave: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", fig.width = 10,
                      fig.asp = 0.75, dev = "svg")
options(width = 110)
```

## Useful functions

```{r, message=FALSE}
library(tidyverse)

methods.color <- setNames(scales::hue_pal()(6), 
                          c("logit-simple", "PRS-max", "PRS-stringent",
                            "T-Trees", "logit-triple", "PRS-all")) 
```

```{r}
format_results <- function(results) {
  results %>%
    as_tibble() %>%
    mutate(
      par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
                          levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
      AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2]))
    )
}
```

```{r}
boot <- function(x, n = 1e4, f = mean) {
  sd(replicate(n, f(sample(x, replace = TRUE))))
}
```

```{r}
myggplot <- function(..., coeff = 1) {
  bigstatsr:::MY_THEME(ggplot(...), coeff = coeff) +
  theme(strip.text.x = element_text(size = rel(2)),
        strip.text.y = element_text(size = rel(2)))
} 
```

```{r}
barplot_causal <- function(results, h2) {

  results %>%
    myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
    geom_hline(yintercept = 0.5, linetype = 2) +
    geom_bar(stat = "identity", alpha = 0.3, position = position_dodge()) +
    geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
                  position = position_dodge(width = 0.9), color = "black", width = 0.2) + 
    scale_y_continuous(limits = c(0.5, NA), minor_breaks = 0:20 / 20,
                       oob = scales::rescale_none) +
    labs(x = "Causal SNPs (number and location)", y = "Mean of 100 AUCs",
         fill = "Method", color = "Method") +
    scale_fill_manual(values = methods.color) +
    scale_color_manual(values = methods.color)
}

barplot_causal_one <- function(results, h2) {

  auc_max <- `if`(h2 == 0.8, 0.94, 0.84)

  results %>%
    filter(par.h2 == h2) %>%
    barplot_causal() +
    geom_hline(yintercept = auc_max, linetype = 3, color = "blue") +
    facet_grid(par.dist ~ .)
}

barplot_causal_all <- function(results) {

  results %>%
    barplot_causal() +
    geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
               data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
    facet_grid(par.dist ~ par.h2)
}
```

```{r}
compare_logit <- function(results, with_method) {

  results <- filter(
    results, method %in% c("logit-simple", with_method), par.h2 == 0.8)

  results.summary <- results %>%
    group_by_at(c(vars(starts_with("par")), "method")) %>%
    summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
    print(n = Inf)

  nsimu <- results.summary$N[[1]]
  stopifnot(all(results.summary$N == nsimu))

  plot_extra <- function(y, ylab = y) {

    results %>%
      myggplot() +
      geom_boxplot(aes_string("par.causal", y, color = "method", 
                              fill = "method"), alpha = 0.3) + 
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      facet_grid(par.model ~ par.dist) +
      labs(x = "Causal SNPs (number and location)", y = ylab, 
           fill = "Method", color = "Method") +
      theme(legend.position = "none") +
      scale_fill_manual(values = methods.color) +
      scale_color_manual(values = methods.color) +
      scale_y_continuous(limits = c(0, NA))
  }

  p1 <- results.summary %>%
    barplot_causal_one(h2 = 0.8) +
    facet_grid(par.model ~ par.dist) +
    labs(y = sprintf("Mean of %d AUCs", nsimu))

  p2 <- cowplot::plot_grid(
    plot_extra("nb.preds", "Number of predictors"),
    plot_extra("timing", "Execution time (in seconds)"),
    ncol = 1, align = "hv", scale = 0.95, 
    labels = LETTERS[2:3], label_size = 15
  )

  p12 <- cowplot::plot_grid(
    p1 + theme(legend.position = "none"), p2, 
    ncol = 2, scale = 0.95, rel_widths = c(6, 4), 
    labels = c("A", ""), label_size = 15
  )

  cowplot::plot_grid(
    cowplot::get_legend(p1 + theme(legend.direction = "horizontal")), p12, 
    rel_heights = c(0.1, 1), ncol = 1
  )
}
```

## Results with T-Trees

```{r}
results1 <- list.files("results1", full.names = TRUE) %>%
  map_dfr(~readRDS(.x)) %>%
  format_results()
```

```{r, out.width="90%", fig.width = 14}
compare_logit(results1, with_method = "T-Trees")

ggsave("figures/supp-ttrees.pdf", scale = 1/90, width = 1400, height = 1060)
```

## Results with logit-triple & model fancy

```{r}
results2 <- list.files("results2", full.names = TRUE) %>%
  map_dfr(~readRDS(.x)) %>%
  format_results()

results2.summary <- results2 %>%
  group_by_at(c(vars(starts_with("par")), "method")) %>%
  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
  print(n = Inf) 
```

### Results with logit-triple

```{r, out.width="90%", fig.width = 14}
compare_logit(bind_rows(results1, results2), with_method = "logit-triple")

ggsave("figures/supp-triple.pdf", scale = 1/90, width = 1400, height = 1060)
```

```{r}
results2.summary %>%
  filter(grepl("logit", method)) %>%
  barplot_causal_one(h2 = 0.5) +
  facet_grid(par.dist ~ par.model) +
  labs(y = "Mean of 20 AUCs")

ggsave("figures/supp-AUC-triple.pdf", scale = 1/90, width = 1242, height = 951)
```

### Results with model fancy

```{r}
results2.summary %>%
  filter(par.model == "fancy", method %in% c("logit-simple", "PRS-max")) %>%
  barplot_causal_all() +
  labs(y = "Mean of 20 AUCs")

ggsave("figures/supp-AUC-logit-fancy.pdf", scale = 1/90, width = 1242, height = 951)
```

```{r}
results2.summary %>%
  filter(par.model == "fancy", grepl("PRS", method)) %>%
  barplot_causal_all() +
  labs(y = "Mean of 20 AUCs")

ggsave("figures/supp-AUC-PRS-fancy.pdf", scale = 1/90, width = 1242, height = 951)
```

## Results with only logit-simple & PRSs

### Correlation between predictive measures

```{r}
results3 <- list.files(paste0("results", 2:3), full.names = TRUE) %>%
  map_dfr(~readRDS(.x)) %>%
  filter(method != "logit-triple", par.model != "fancy") %>%
  format_results()

results3_corr <- results3 %>%
  filter(method %in% c("logit-simple", "PRS-max")) %>%
  mutate(
    percCases10 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[1:110], 2])),
    percCases20 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[1:220], 2])),
    Parameters  = interaction(paste("h2 =", par.h2), par.dist, method, sep = " | ")
  )
```

```{r, fig.asp=0.5, out.width="90%", fig.width=15}
p1 <- results3_corr %>%
  myggplot(aes(AUC, percCases10, color = Parameters)) + 
  geom_point(size = 0.6, alpha = 0.5) +
  geom_smooth(aes(linetype = Parameters), method = "lm", se = FALSE, size = 2) +
  theme(legend.key.width = unit(3.5, "line"), 
        legend.direction = "horizontal") + 
  labs(y = "Proportion of cases among top 10% scores")
p2 <- results3_corr %>%
  myggplot(aes(AUC, percCases20, color = Parameters)) + 
  geom_point(size = 0.6, alpha = 0.5) +
  geom_smooth(aes(linetype = Parameters), method = "lm", se = FALSE, size = 2) + 
  labs(y = "Proportion of cases among top 20% scores") +
  theme(legend.position = "none")

cowplot::plot_grid(p1 + theme(legend.position = "none"), p2 , 
                   ncol = 2, align = "hv", scale = 0.95) %>%
  cowplot::plot_grid(cowplot::get_legend(p1), ., rel_heights = c(0.1, 1), ncol = 1)

ggsave("figures/supp-AUC-corr.pdf", scale = 1/90, width = 1558, height = 737)

with(results3_corr, cor(AUC, percCases10))
with(results3_corr, cor(AUC, percCases20))
```

### 6000 in training & all chromosomes

```{r}
results3.summary <- results3 %>%
  group_by_at(c(vars(starts_with("par")), "method")) %>%
  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
  print(n = Inf)
```

```{r, fig.asp = 1, out.width = "60%", fig.width=8}
results3.summary %>%
  filter(method %in% c("logit-simple", "PRS-max")) %>%
  barplot_causal_one(h2 = 0.8)

ggsave("figures/main-AUC-logit.pdf", scale = 1/90, width = 844, height = 872)
```

```{r, fig.asp = 1, out.width = "60%", fig.width=8}
results3.summary %>%
  filter(grepl("PRS", method)) %>%
  barplot_causal_one(h2 = 0.8)

ggsave("figures/main-AUC-PRS.pdf", scale = 1/90, width = 844, height = 872)
```

```{r, fig.asp = 1, out.width = "60%", fig.width=8}
results3.summary %>%
  filter(method %in% c("logit-simple", "PRS-max")) %>%
  barplot_causal_one(h2 = 0.5)

ggsave("figures/supp-AUC-logit.pdf", scale = 1/90, width = 844, height = 872)
```

```{r, fig.asp = 1, out.width = "60%", fig.width=8}
results3.summary %>%
  filter(grepl("PRS", method)) %>%
  barplot_causal_one(h2 = 0.5)

ggsave("figures/supp-AUC-PRS.pdf", scale = 1/90, width = 844, height = 872)
```

### 6000 in training & only chromosome 6

```{r}
results4 <- list.files("results4", full.names = TRUE) %>%
  map_dfr(~readRDS(.x)) %>%
  format_results()

results4 %>%
  group_by_at(c(vars(starts_with("par")), "method")) %>%
  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
  print(n = Inf)

results34 <- rbind(
  cbind(simu = "all", results3),
  cbind(simu = "chr6", results4)
) %>%
  as_tibble()
```

```{r, fig.asp = 1, out.width = "90%"}
results34 %>%
  myggplot() +
  geom_hline(yintercept = 0.5, linetype = 2) +
  geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
             data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
  geom_boxplot(aes(method, AUC, fill = simu, color = simu), alpha = 0.3) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(par.model ~ par.causal) +
  facet_grid(par.h2 + par.dist ~ par.causal) +
  scale_fill_brewer(type = "qual", palette = 2) +
  scale_color_brewer(type = "qual", palette = 2) + 
  labs(x = "Method")

ggsave("figures/supp-AUC-chr6.pdf", scale = 1/90, width = 1153, height = 908)
```

### Varying training size & all shromosomes

```{r}
results5 <- list.files("results5", full.names = TRUE) %>%
  map_dfr(~readRDS(.x)) %>%
  format_results()

results5.summary <- results5 %>%
  group_by_at(c(vars(starts_with("par")), "method", "n.train")) %>%
  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
  print(n = Inf)

results35.summary <- results3.summary %>%
  filter(par.causal == "300 in all") %>%
  cbind.data.frame(n.train = 6000) %>%
  bind_rows(results5.summary)
```

```{r}
myggplot(results35.summary, aes(n.train, AUC_mean, color = method, linetype = method)) +
  geom_hline(yintercept = 0.5, linetype = 2) +
  geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
             data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
  geom_line(size = 1.5) +
  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, 
                    ymax = AUC_mean + 2 * AUC_boot), 
                size = 1.5, width = 0) +
  facet_grid(par.h2 ~ par.dist) +
  scale_x_continuous(breaks = 1:6 * 1000, minor_breaks = NULL) + 
  scale_y_continuous(minor_breaks = 0:20 / 20) +
  scale_color_manual(values = methods.color) +
  labs(x = "Size of the training set", y = "Mean of 100 AUCs", 
       color = "Method", linetype = "Method") +
  theme(legend.key.width = unit(3, "line"))

ggsave("figures/main-AUC-ntrain.pdf", scale = 1/90, width = 1050, height = 800)
```

### 2017-12-12 results1.R

@@ -1,70 +1,0 @@
-library(tidyverse)
-
-results1 <- list.files("results1", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  format_results()
-
-results_ttrees <- results1 %>%
-  filter(method %in% c("logit-simple", "T-Trees")) %>%
-  group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
-  print(n = Inf)
-
-
-
-compare_logit <- function(results, with_method) {
-  
-  results <- filter(
-    results, method %in% c("logit-simple", with_method), par.h2 == 0.8)
-  
-  results.summary <- results %>%
-    group_by_at(c(vars(starts_with("par")), "method")) %>%
-    summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
-    print(n = Inf)
-  
-  nsimu <- results.summary$N[[1]]
-  stopifnot(all(results.summary$N == nsimu))
-  
-  plot_extra <- function(y, ylab = y) {
-    
-    results %>%
-      myggplot() +
-      geom_boxplot(aes_string("par.causal", y, color = "method", 
-                              fill = "method"), alpha = 0.3) + 
-      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-      facet_grid(par.model ~ par.dist) +
-      labs(x = "Causal SNPs (number and location)", y = ylab, 
-           fill = "Method", color = "Method") +
-      theme(legend.position = "none") +
-      scale_fill_manual(values = methods.color) +
-      scale_color_manual(values = methods.color) +
-      scale_y_continuous(limits = c(0, NA))
-  }
-  
-  p1 <- results.summary %>%
-    barplot_causal_one(h2 = 0.8) +
-    facet_grid(par.model ~ par.dist) +
-    labs(y = sprintf("Mean of %d AUCs", nsimu))
-  
-  p2 <- cowplot::plot_grid(
-    plot_extra("nb.preds", "Number of predictors"),
-    plot_extra("timing", "Execution time (in seconds)"),
-    ncol = 1, align = "hv", scale = 0.95, 
-    labels = LETTERS[2:3], label_size = 15
-  )
-  
-  p12 <- cowplot::plot_grid(
[…]

### 2017-12-12 results2.R

@@ -1,45 +1,0 @@
-results2 <- list.files("results2", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  format_results()
-
-results2.summary <- results2 %>%
-  group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
-  print(n = Inf) 
-
-barplot_causal_all <- function(results) {
-  
-  results %>%
-    barplot_causal() +
-    geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
-               data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
-    facet_grid(par.dist ~ par.h2)
-}
-
-results2.summary %>%
-  filter(par.model == "fancy", method %in% c("logit-simple", "PRS-max")) %>%
-  barplot_causal_all() +
-  labs(y = "Mean of 20 AUCs")
-  
-ggsave("figures/supp-AUC-logit-fancy.pdf", scale = 1/90, width = 1242, height = 951)
-
-results2.summary %>%
-  filter(par.model == "fancy", grepl("PRS", method)) %>%
-  barplot_causal_all() +
-  labs(y = "Mean of 20 AUCs")
-
-ggsave("figures/supp-AUC-PRS-fancy.pdf", scale = 1/90, width = 1242, height = 951)
-
-## Logistic regressions
-
-compare_logit(bind_rows(results1, results2), with_method = "logit-triple")
-
-ggsave("figures/supp-triple.pdf", scale = 1/90, width = 1400, height = 1060)
-
-results2.summary %>%
-  filter(grepl("logit", method)) %>%
-  barplot_causal_one(h2 = 0.5) +
-  facet_grid(par.dist ~ par.model) +
-  labs(y = "Mean of 20 AUCs")
-
-ggsave("figures/supp-AUC-triple.pdf", scale = 1/90, width = 1242, height = 951)

### 2017-12-12 results3.R

@@ -1,113 +1,0 @@
-methods.color <- setNames(scales::hue_pal()(6), 
-                          c("logit-simple", "PRS-max", "PRS-stringent",
-                            "T-Trees", "logit-triple", "PRS-all")) 
-
-library(tidyverse)
-
-format_results <- function(results) {
-  results %>%
-    as_tibble() %>%
-    mutate(
-      par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
-                          levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
-      AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2]))
-    )
-}
-
-results3 <- list.files(paste0("results", 2:3), full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  filter(method != "logit-triple", par.model != "fancy") %>%
-  format_results()
-
-
-results3.summary <- results3 %>%
-  group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
-  print(n = Inf)
-
-## Make a function of this
-
-barplot_causal <- function(results, h2) {
-  
-  results %>%
-    myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
-    geom_hline(yintercept = 0.5, linetype = 2) +
-    geom_bar(stat = "identity", alpha = 0.3, position = position_dodge()) +
-    geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
-                  position = position_dodge(width = 0.9), color = "black", width = 0.2) + 
-    scale_y_continuous(limits = c(0.5, NA), minor_breaks = 0:20 / 20,
-                       oob = scales::rescale_none) +
-    labs(x = "Causal SNPs (number and location)", y = "Mean of 100 AUCs",
-         fill = "Method", color = "Method") +
-    scale_fill_manual(values = methods.color) +
-    scale_color_manual(values = methods.color)
-}
-
-barplot_causal_one <- function(results, h2) {
-  
-  auc_max <- `if`(h2 == 0.8, 0.94, 0.84)
-  
-  results %>%
[…]

### 2017-12-12 results4.R

@@ -1,50 +1,0 @@
-results4 <- list.files("results4", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  format_results()
-
-results4 %>%
-  group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
-  print(n = Inf)
-
-results4 <- list.files("results4", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  format_results() %>%
-  cbind(simu = "figures/supp-AUC-chr6") %>%
-  bind_rows() %>%
-  as_tibble()
-
-results34 <- rbind(
-  cbind(simu = "all", results3),
-  cbind(simu = "chr6", results4)
-) %>%
-  as_tibble()
-
-results34 %>%
-  myggplot() +
-  geom_hline(yintercept = 0.5, linetype = 2) +
-  geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
-             data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
-  geom_boxplot(aes(method, AUC, fill = simu, color = simu), alpha = 0.3) + 
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  facet_grid(par.model ~ par.causal) +
-  facet_grid(par.h2 + par.dist ~ par.causal) +
-  scale_fill_brewer(type = "qual", palette = 2) +
-  scale_color_brewer(type = "qual", palette = 2) + 
-  labs(x = "Method")
-
-ggsave("figures/supp-AUC-chr6.pdf", scale = 1/90, width = 1153, height = 908)
-
-# results4.summary %>%
-#   filter(par.h2 == 0.8, method %in% c("logit-simple", "PRS-max")) %>%
-#   myggplot(aes(par.causal, AUC_mean, fill = method, color = method)) +
-#   geom_hline(yintercept = 0.5, linetype = 2) +
-#   geom_hline(yintercept = 0.94, linetype = 3, color = "blue") +
-#   geom_bar(stat = "identity", alpha = 0.3, position=position_dodge()) +
-#   geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
-#                 position=position_dodge(width=0.9), color = "black", width = 0.2) +
[…]

### 2017-12-12 results5.R

@@ -1,40 +1,0 @@
-results5 <- list.files("results5", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  as_tibble() %>%
-  mutate(
-    par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
-                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
-    AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2]))
-  )
-
-pryr::object_size(results5)
-
-results5_3 <- results3 %>%
-  select(-eval) %>%
-  filter(par.model == "simple", par.causal == "300 in all") %>%
-  cbind(n.train = 6000, .) %>%
-  bind_rows(select(results5, -eval)) %>%
-  as_tibble()
-
-results5_3 %>%
-  group_by_at(c(vars(starts_with("par")), "method", "n.train")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, 1e4, mean), N = n()) %>%
-  print(n = Inf)
-
-tmp <- .Last.value
-
-myggplot(tmp, aes(n.train, AUC_mean, color = method)) +
-  geom_hline(yintercept = 0.5, linetype = 2) +
-  geom_point(size = 3) +
-  geom_line(size = 1.5) +
-  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, 
-                    ymax = AUC_mean + 2 * AUC_boot), 
-                size = 1.5, width = 0) +
-  facet_grid(par.model + par.h2 ~ par.causal + par.dist) +
-  theme(strip.text.x = element_text(size = rel(2)),
-        strip.text.y = element_text(size = rel(2))) +
-  scale_x_continuous(breaks = 1:6 * 1000, minor_breaks = NULL) + 
-  scale_y_continuous(minor_breaks = 0:20 / 20) +
-  scale_fill_brewer(type = "qual", palette = 2) +
-  labs(x = "Causal SNPs (number and location)", y = "Mean of 100 AUCs",
-       fill = "Method")

### 2017-12-13 5-results-celiac.Rmd
save

--- 2017-11-04 5-regularization-paths.Rmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
-                      fig.asp = 0.7, error = TRUE)
-options(width = 85)
+knitr::opts_chunk$set(fig.align = "center", out.width = "70%", fig.width = 10,
+                      fig.asp = 0.7, dev = "svg")
+options(width = 95)
 ```
@@
 # Read file, see "Preprocessing" notebook
+library(bigsnpr)
 celiac <- snp_attach("backingfiles/celiacQC.rds")
@@
 
-```{r}
+```{r, dev="png"}
 labels <- 1:22; labels[c(11, 13, 15, 17, 18, 20, 21)] <- ""
@@
   snp_manhattan(gwas.train.gc, CHR, POS, labels = labels) +
-    coord_cartesian(ylim = c(0, 30)) + 
+    coord_cartesian(ylim = c(0, 25)) + 
     geom_hline(yintercept = -log10(5e-8), color = "red", linetype = 3),
@@
 ```{r}
-ggsave("figures/celiac-man.png", scale = 1/90, width = 1400, height = 1000)
+ggsave("figures/supp-celiac-man.png", scale = 1/90, width = 1300, height = 1000)
 ```
@@
 
-Cross-Model Selection and Averaging:
+Cross-Model Selection and Averaging (CMSA):
 
@@
 ```{r}
-ggsave("figures/celiac-regpath.pdf", scale = 1/90, width = 1000, height = 700)
+ggsave("figures/main-celiac-regpath.pdf", scale = 1/90, width = 1000, height = 700)
 ```

### 2017-12-13 celiac-effects.R

@@ -1,0 +1,73 @@
+# Effects GWAS vs logistic regression
+qplot(cmsa.logit[cols_along(G)], gwas.train.gc$estim,
+      color = predict(gwas.train.gc, log10 = FALSE) < 5e-8) %>%
+  bigstatsr:::MY_THEME() +
+  labs(color = "Bonf") + 
+  geom_abline(slope = 1) + 
+  geom_smooth(method = "lm", color = "green")
+
+
+preds2
+
+
+obj.svd3 <- obj.svd
+obj.svd3$u[ind.train, ] <- NA
+plot(obj.svd3, type = "scores")
+
+
+preds.train <- predict(cmsa.logit, X = G, ind.row = ind.train, 
+                       covar.row = obj.svd$u[ind.train, ])
+myglm <- glm(y[ind.train] ~ preds.train, family = "binomial")
+summary(myglm)
+preds2.prob <- predict(myglm, newdata = data.frame(preds.train = preds2),
+                       type = "response")
+err <- rep(NA, nrow(G))
+err[ind.test] <- (y[ind.test] != (preds2.prob > 0.5)) # (y[ind.test] - preds2.prob)^2
+
+plot(obj.svd3, type = "scores") + 
+  aes(color = as.factor(err), alpha = I(rep(0.6, nrow(G))))
+
+boot1000 <- function(x) {
+  sd(replicate(1000, mean(sample(x, replace = TRUE))))
+}
+round4 <- function(x) round(x, 4)
+data.frame(pop = celiac$fam$family.ID[ind.test],
+           pred = preds2.prob,
+           error = err[ind.test]) %>%
+  filter(pred > quantile(pred, 0.8)) %>%
+  group_by(pop) %>%
+  summarise_at("error", funs(mean, boot1000)) %>%
+  mutate_if(is.numeric, funs(round4))
+
+
+head(ord.eff <- order(abs(cmsa.logit[cols_along(G)]), decreasing = TRUE), n = 10)
+lapply(ord.eff[1:10], function(j) table(G[, j], y))
+# Manhattan plots with greatest effects
+cowplot::plot_grid(
+  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels, 
+                ind.highlight = head(ord.eff, 50)),
+  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels, 
+                ind.highlight = head(ord.eff, 50)) +
[…]

### 2017-12-13 celiac-top-cases.R

@@ -1,0 +1,59 @@
+theone <- which.min(predict(gwas.train.gc))
+table(G[ind.test, theone], y[ind.test])
+
+counts.case <- big_counts(G, ind.row = which(y == 1))
+counts.control <- big_counts(G, ind.row = which(y == 0))
+
+counts.case[, theone]
+counts.control[, theone]
+
+props <- sapply(cols_along(G), function(j) {
+  ccoj <- counts.control[, j]
+  mut <- `if`(ccoj[1] < ccoj[3], 1, 3)
+  `if`(counts.case[mut, j] > 20, counts.case[mut, j] / ccoj[mut], 0)
+})
+hist(props)
+hist(props[props > 1])
+ord.prop <- order(props, decreasing = TRUE)
+head(ord.prop, 10)
+lapply(ord.prop[1:10], function(j) table(G[, j], y))
+
+
+# one geno
+ind0 <- which(G[ind.test, ord.prop[1]] == 0)
+mean(y[ind.test[ind0]])    ## 78.2%
+# prs max
+(M <- length(ind0))
+indPRS <- order(prs[, which.max(aucs)], decreasing = TRUE)[1:M]
+mean(y[ind.test[indPRS]])  ## 75.2%
+# logit-simple (CMSA)
+indLS <- order(preds2, decreasing = TRUE)[1:M]
+mean(y[ind.test[indLS]])   ## 79.4 %
+# logit-triple (CMSA)
+indLT <- order(preds4, decreasing = TRUE)[1:M]
+mean(y[ind.test[indLT]])   ## 82.4 %
+
+plot(pROC::roc(y[ind.test], prs[, which.max(aucs)]))
+plot(pROC::roc(y[ind.test], preds2), add = TRUE, col = "blue")
+plot(pROC::roc(y[ind.test], preds4), add = TRUE, col = "red")
+
+library(tidyverse)
+library(plotROC)
+scores_tidy <- tibble(
+  d = y[ind.test], 
+  PRS_max =  prs[, which.max(aucs)],
+  LS = preds2,
+  LT = preds4
+) %>%
+  gather(key = "Method", value = "Score", -d)
+scores_tidy
+  
+bigstatsr:::MY_THEME(
+  ggplot(scores_tidy, aes(d = d, m = Score, color = Method, linetype = Method)) +
+    style_roc(xlab = "1 - Specificity", ylab = "Sensitivity")
+) +
[…]

### 2017-12-13 auc-max-h2.R
save

@@ -1,7 +1,8 @@
-nsimu <- 1e7
+
 K <- 0.3
 
-replicate(10, {
+nsimu <- 1e8
+replicate(5, {
   sapply(c(0.5, 0.8), function(h2) {
     S <- rnorm(nsimu, mean = 0, sd = sqrt(h2))
     E <- rnorm(nsimu, mean = 0, sd = sqrt(1 - h2))
@@ -10,6 +11,9 @@ replicate(10, {
     bigstatsr::AUC(S, Y)
   })
 })
+#           [,1]      [,2]      [,3]      [,4]      [,5]
+# [1,] 0.8410003 0.8410194 0.8410863 0.8410095 0.8410279
+# [2,] 0.9406166 0.9406530 0.9406010 0.9405916 0.9406338
 
 # Estimation from Wray 2010
 sapply(c(0.5, 0.8), function(h2) {

### 2017-12-13 oracle.R

@@ -47,10 +47,10 @@ get_oracle(G, 0.8, 3000, effects.dist = "gaussian", model = "simple")
 
 params.grid2 <- expand.grid(
   par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
-  par.dist   = "gaussian", 
-  par.h2     = 0.8, 
+  par.dist   = c("gaussian", "laplace"), 
+  par.h2     = c(0.5, 0.8), 
   par.model  = c("simple", "fancy"),
-  num.simu   = 1:50,
+  num.simu   = 1:100,
   stringsAsFactors = FALSE
 ) 
 
@@ -81,4 +81,5 @@ oracles2 %>%
     par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
                         levels = c("30 in HLA", paste(3 * 10^(1:3), "in all")))) %>%
   group_by(par.causal, par.dist, par.h2, par.model) %>%
-  summarise(AUC = mean(AUC))
+  summarise(AUC_mean = round(mean(AUC), 3),
+            AUC_mean_boot = boot(AUC, 1e4, mean))

### 2017-12-13 3-oracle.Rmd
save

@@
 (G <- celiac2$genotypes)
-# (G2 <- big_attach("backingfiles/celiacQC_sub1_tripled1.rds"))
-# 
-# covar.all <- readRDS("backingfiles/PCA2.rds")$u
-# 
-# n <- nrow(G)
-# n.train <- 6e3
 ind.HLA <- snp_indLRLDR(CHR, POS, subset(LD.wiki34, ID == "hild12"))
@@
 
-    AUC(S, Y)
+    bigstatsr::AUC(S, Y)
   }
@@
   group_by(par.causal, par.dist, par.h2, par.model) %>%
-  summarise(AUC_mean = round(mean(AUC), 3),
-            AUC_mean_boot = round(boot(AUC, 1e4, mean), 4)) %>%
+  summarise(AUC_mean = round(mean(AUC), 4),
+            AUC_mean_boot = round(boot(AUC), 4)) %>%
   print(n = Inf)
@@
   summarise(AUC_mean = round(mean(AUC), 4),
-            AUC_mean_boot = round(boot(AUC, 1e4, mean), 4)) %>%
+            AUC_mean_boot = round(boot(AUC), 4)) %>%
   print(n = Inf)

### 2017-12-13 4-results-simus.Rmd
save

@@
 
-```{r, fig.asp=0.5, out.width="90%", fig.width=15}
+```{r, fig.asp=0.5, out.width="95%", fig.width=16}
 p1 <- results3_corr %>%
@@
 ggsave("figures/supp-AUC-corr.pdf", scale = 1/90, width = 1558, height = 737)
+```
 
+```{r}
 with(results3_corr, cor(AUC, percCases10))

### 2017-12-13 paper/draft.tex
save

@@
 
-We make 5 different simulation procedures, summarized in table \ref{tab:simus}. 
-..Simulation..
-\textnumero1 uses the whole dataset (the 22 chromosomes), a training set of size 6000 and an heritability of 80\%.
+We make 5 different simulation scenarios, summarized in table \ref{tab:simus}. 
+Scenario \textnumero1 uses the whole dataset (the 22 chromosomes), a training set of size 6000 and an heritability of 80\%.
 It compares all methods (see section \ref{sec:methods}) on all other combination of parameters in 5 simulations.
-The goal of the first simulation is to quickly discard the T-Trees method.
-For simulation \textnumero2, we make also vary the heritability (50\% or 80\%) but we don't run the T-Trees method anymore.
-The goal of the second simulation is to compare all remaining methods, especially to see how well the ``logit-triple'' method works.
+The goal of the first scenario is to quickly discard the T-Trees method.
+For scenario \textnumero2, we make also vary the heritability (50\% or 80\%) but we don't run the T-Trees method anymore.
+The goal of the second scenario is to compare all remaining methods on all models (including the ``fancy'' model), especially to see how well the ``logit-triple'' method works.
 We run 20 simulations for each combination of parameters.
-For simulation \textnumero3, we continue with 80 more simulations (to reach 100 simulations, i.e.\ 5 times more than in \cite{zhou2013polygenic}), but discarding the ``logit-triple'' method and the ``fancy'' model.
-Simulation \textnumero4 consists in 100 simulations of remaining methods and combinations of parameters on a dataset composed of only the chromosome 6.
+For scenario \textnumero3, we continue with 80 more simulations (to reach 100 simulations, i.e.\ 5 times more than in \cite{zhou2013polygenic}), but discarding the ``logit-triple'' method and the ``fancy'' model.
+Scenario \textnumero4 consists in 100 simulations of remaining methods and combinations of parameters on a dataset composed of only the chromosome 6.
 Reducing the number of SNPs aims at increasing the ..polygenicity..
@@
 DUDBRIDGE?].
-Finally, simulation \textnumero5 reuses the whole dataset but varying the size of the training set in order to assess how the number of samples affects predictive performance of the PRS and the ``logit-simple'' methods.
+Finally, scenario \textnumero5 reuses the whole dataset but varying the size of the training set in order to assess how the number of samples affects predictive performance of the PRS and the ``logit-simple'' methods.
 100 simulations are run using 300 causal SNPs anywhere on the genome.
@@
 This measure indicates how well we can identify high-risk individuals from a higher PRS in screening campaigns.
-[COMMENT ON PREVALENCE?]
 
@@
 
-\subsection{Maximum AUCs}
+\subsection{Overview and naming}
 
-[MOVE IN SUPMAT AND METHODS]
-
-Assuming that $y \sim N(0, h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the heritability $h^2$ (see ..SupMat..).
-We report AUCs of 0.8410 and 0.9406 for respectively an heritability of 50\% and 80\%.
-% wray2010genetic
-
-\subsection{Overview and naming}
+In results, we use estimates of maximum achievable AUC values for the simulations: 84\% (respectively 94\%) for heritabilities of 50\% (respectively 80\%).
 
@@
 
+Main simulations corresponds to scenarios 3 to 5 (Table \ref{tab:simus}).
+They mainly compare our ``simple'' logistic regression to the standard P+T methods in case of a ``simple'' model (only linear effects).
+
+Using the penalized logistic regression, we report AUCs greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, which is not obtained by the upper bound of the P+T method in any scenario (Figure \ref{fig:main-AUC-logit}).
+Our penalized regression methods consistently provides better predictive performance than the P+T method for all disease architectures tested, excepted sometimes when we simulate 3000 causals SNPs anywhere on the genome, a case where all methods rarely exceeds 65\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
+
+Our method is parallelized and requires only less than 10 minutes to run (with 6 physical cores) [TABLE ?FIGURE?].
+Moreover, it results in relatively sparse models with no more than 14,000 predictors entering the models even for the linear model of 3000 causal SNPs [TABLE ?FIGURE?].
+
 First, we show that AUC is highly correlated ($r^2 > 95\%$) with the percentage of cases in either the 10\% or 20\% largest PRSs (Figure \ref{fig:AUC-corr}).
@@
 
-Using the penalized logistic regressions, we report AUCs greater than 95\%, which is not obtained by the upper bound of the P+T method in any scenario (Figure \ref{fig:AUC}).
-Our penalized regression methods consistently provides better predictive performance than the P+T method for all disease architectures tested [ou pas, FIGURE].
-Our method is parallelized and requires only less than 10 minutes to run (with 6 physical cores) [TABLE ?FIGURE?].
-Moreover, it results in relatively sparse models with no more than 14,000 predictors entering the models even for the linear model of 3000 causal SNPs [TABLE ?FIGURE?].
 P+T runs in only a few seconds because we don't consider the GWAS computation as part of the execution time of the method because this is generally executed by others (e.g.\ consortia) and then only the resulting summary statistics are used to construct a PRS.
@@
 Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & Number of & \multirow{2}{*}{Methods} \\ 
-simulation & & training set & (number and location) & of effects & & model & simulations & \\
+scenario & & training set & (number and location) & of effects & & model & simulations & \\
 \hline
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
-\caption{Mean of AUC over 100 simulations for our penalized logistic regression (``logit-simple'') and the maximum AUC reported with the P+T method (``PRS-max'').
+\caption{Scenario \textnumero3.
+Mean of AUC over 100 simulations for our penalized logistic regression (``logit-simple'') and the maximum AUC reported with the P+T method (``PRS-max'').
 Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-PRS}}
-\caption{Mean of AUC over 100 simulations for ..the three reported PRSs...
+\caption{Scenario \textnumero3.
+Mean of AUC over 100 simulations for ..the three reported PRSs...
 Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
@@
 [MODEL AND H2]}\label{fig:main-AUC-PRS}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
+\caption{Scenario \textnumero5.
+Mean of AUC over 100 simulations for ..the three reported PRSs...
+and the simple logistic regression as a function of the training size.
+This results are for an additive model with 300 causal SNPs sampled anywhere on the genome.
+Upper (lower) panels are presenting results for an heritability of 0.5 (0.8) and left (right) panels are presenting results for effets following a Gaussian (Laplace) distribution.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+The blue dotted line represents the maximum achievable AUC.
+[MODEL AND H2]}\label{fig:main-AUC-ntrain}
 \end{figure}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \newpage
-\section*{Supplementary Data}
+\section*{Supplementary Materials}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection*{Maximum AUCs} \label{sec:auc-max}
+
+We used three different ways to estimate the maximum achievable AUC for our simulations.
+First, we used the estimation from equation (3) of \cite{wray2010genetic}.
+For a prevalence fixed at 30\% and an heritability of 50\% (respectively 80\%), the approximated theoretical values of AUC are 84.1\% (respectively 93.0\%).
+Note that this approximation is reported to be less accurate for high heritabilities.
+Secondly, if we assume that the genetic part of the liabilities $y \sim N(0, h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the heritability $h^2$ through Monte Carlo simulations.
+We report AUCs of 84.1\% and 94.1\% for respectively an heritability of 50\% and 80\%.
+[HOW TO ADD ERROR?
+I HAVE SUMMARY()]
+Thirdly, we reproduce the exact same procedure of simulations and, for each combination of parameters (Table \ref{tab:simus}), we estimate the AUC of the ``oracle'', i.e.\ the true simulated genetic part of the liabilities though 100 replicates.
+For every combination of parameters, AUC of oracles are comprised between 83.2\% and 84.2\% for an heritability of 50\% and between 93.2\% and 94.1\% for an heritability of 80\%.
+Given all these estimates of the maximal achievable AUC and for the sake of simplicity, we report maximum AUCs of 84\% (respectively 94\%) for heritabilities of 50\% (respectively 80\%) whatever are the parameters of the simulation.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newpage
 
@@
 
-%\begin{figure}[h]
-%\centerline{\includegraphics[width=\textwidth]{ttrees}}
-%\caption{Results of T-Trees vs penalized logistic regression.
-\textbf{A.} Timing (in seconds).
-\textbf{B.} Number of predictors of the model.
-\textbf{C.} AUC.
-\textbf{D.} Percentage of cases in the 10\% largest scores.}\label{fig:ttrees}
-%\end{figure}
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-logit-fancy}}
+\caption{[COPY FOR H2=0.5 \& MODEL FANCY]}\label{fig:supp-AUC-logit-fancy}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS-fancy}}
+\caption{[COPY FOR H2=0.5 \& MODEL FANCY]}\label{fig:supp-AUC-PRS-fancy}
+\end{figure}
 
@@
 
-%\begin{figure}[h]
-%\centerline{\includegraphics[width=\textwidth]{AUC-corr}}
-%\caption{Percentage of cases in the 2 highest deciles of PRSs as a function of AUC.}\label{fig:AUC-corr}
-%\end{figure}
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-ttrees}}
+\caption{[T-TREES 3 PLOTS - H2=0.8]}\label{fig:supp-ttrees}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-triple}}
+\caption{[TRIPLE 3 PLOTS - H2=0.8]}\label{fig:supp-triple}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-triple}}
+\caption{[TRIPLE ONLY AUC - H2=0.5]}\label{fig:supp-AUC-triple}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6}}
+\caption{Scenario \textnumero4. Boxplots of AUC over 100 simulations for ..the three reported PRSs... and the simple logistic regression.
+Vertical panels represents different number of causal SNPs and their location (``all'': anywhere on the genome; ``HLA'': only in the HLA region).
+Horizontal panels are presenting results for effets following a Gaussian or Laplace distribution and an heritability of 0.5 or 0.8.
+The blue dotted line represents the maximum achievable AUC.
+[MODEL AND H2]}\label{fig:supp-AUC-chr6}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-corr}}
+\caption{Scenario \textnumero3. Comparison of different predictive measures. Correlation is greater than 98\%.[ADD]}\label{fig:supp-AUC-corr}
+\end{figure}
 

### 2017-12-14 paper/draft.tex
save

@@
 
-[MONTRER UN MANHATTAN PLOT POUR JUSTIFIER 30 IN HLA]
-
-[MONTRER UN GRAPH EN FCT D'INCLUSION]
-
 \subsection{Overview and naming}
@@
 In results, we use estimates of maximum achievable AUC values for the simulations: 84\% (respectively 94\%) for heritabilities of 50\% (respectively 80\%).
-
-\subsection{T-Trees}
-
-First, we ran only 5 simulations for each combination of parameters), excepted for the heritability that we fixed at $0.8$ (see section \ref{sec:simus}).
-The goal was to quickly discard the T-Trees method as being competitive with the other methods. 
-For example, compared to our penalized logistic regression, T-Trees perform worse for both predictive measures, the AUC and the percentage of cases in highest scores (Figure \ref{fig:ttrees}).
-Moreover, T-Trees takes longer to run and makes more complex predictive models because it uses more SNPs in the models and has non-linear effects (Figure \ref{fig:ttrees}).
-So, we decided to discard this method for the rest of the simulations.
 
@@
 
-Main simulations corresponds to scenarios 3 to 5 (Table \ref{tab:simus}).
+Main simulations corresponds to scenario \textnumero3 (Table \ref{tab:simus}).
 They mainly compare our ``simple'' logistic regression to the standard P+T methods in case of a ``simple'' model (only linear effects).
 
-Using the penalized logistic regression, we report AUCs greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, which is not obtained by the upper bound of the P+T method in any scenario (Figure \ref{fig:main-AUC-logit}).
+Using the penalized logistic regression, we report AUCs greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, which is not obtained by the upper bound of the P+T method in any scenario (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
 Our penalized regression methods consistently provides better predictive performance than the P+T method for all disease architectures tested, excepted sometimes when we simulate 3000 causals SNPs anywhere on the genome, a case where all methods rarely exceeds 65\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
+We also report that the choice of the threshold of inclusion of SNPs in the P+T method can greatly impact its performance, for example dropping from an AUC of 90\% to less than 60\% when including all SNPs passing clumping in a model with only 30 causal SNPs, or dropping from 60\% to a random predictor (50\% of AUC) when including only GWAS-significant SNPs in a model with 3000 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
 
-Our method is parallelized and requires only less than 10 minutes to run (with 6 physical cores) [TABLE ?FIGURE?].
-Moreover, it results in relatively sparse models with no more than 14,000 predictors entering the models even for the linear model of 3000 causal SNPs [TABLE ?FIGURE?].
+\subsection{Simulations varying number of SNPs and training size}
 
-First, we show that AUC is highly correlated ($r^2 > 95\%$) with the percentage of cases in either the 10\% or 20\% largest PRSs (Figure \ref{fig:AUC-corr}).
-Therefore, AUC can be a good measure at determining whether a score can identify high-risk individuals or not, so we report only predictive performance in terms of AUC in the following results.
+These simulations corresponds to scenarios \textnumero4 and \textnumero5.
+First, when reproducing simulations of scenario \textnumero3 on only the chromosome 6, methods globally increase or ..stagnate..
+in predictive performance (Figure \ref{fig:supp-AUC-chr6}).
+There is particularly high increase in the case when simulating 3000 causal SNPs anywhere on the dataset, for example, the ``logit-simple'' can increase from less than 60\% of AUC to nearly 80\%.
+On the contrary, when simulating only 30 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' seems to slightly decrease.
+Secondly, when varying the training size, we report an increase of AUC whith training size, with the AUC of the ``logit-simple'' method increasing faster than the ``PRS-max'' (Figure \ref{fig:main-AUC-ntrain}).
 
-P+T runs in only a few seconds because we don't consider the GWAS computation as part of the execution time of the method because this is generally executed by others (e.g.\ consortia) and then only the resulting summary statistics are used to construct a PRS.
+\subsection{``Fancy'' methods and model}
 
-As for the comparison between our two penalized logistic regression, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects (Figure \ref{fig:AUC-rel}).
+First, in scenario \textnumero1 we also tested the T-Trees method. 
+For example, compared to our penalized logistic regression, T-Trees perform worse in term predictive measures, while taking longer to run and making more complex predictive models because it uses more SNPs in the models and has non-linear effects (Figure \ref{fig:supp-ttrees}).
+Even when simulating a fancy model in which there is dominant and interaction effects that decision trees based methods should handle, AUC is still worse than the ``logit-simple''.
+Secondly, as for the comparison between our two penalized logistic regression in scenario \textnumero2, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
 Yet, the ``triple'' solution takes approximately 3 more times to run.
+Thirdly, for other simpler methods, we report the exact same results that is to say that the ``logit-simple'' give better AUCs than the ``PRS-max'', excepted in the case of 3000 causal SNP anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}).
+Moreover, the choice of the threshold inclusion of SNPs still greatly affect the performance of the P+T method (Figure \ref{fig:supp-AUC-PRS-fancy}).
 
-As for general results, we report a decrease of AUC when the number of causal variants $M$ is increasing [BATEAU? quoi d'autre?].
+\subsection{Other measures of performance}
 
-[TROUVER DES AUTRES: \% of decrease for h2 0.8 -> 0.5? Pas bon d'utliser tous les SNPs]
+Our method is parallelized and requires only approximately 500 seconds to run (with 6 physical cores) for a training size of 6000 individuals.
+It takes 1000 seconds to train on 12,000 individuals.
+Moreover, when there are 3 times more variables, it takes approximately 3 more times to run.
+We can confidently say that the ``logit-simple'' is of linear complexity in both dimensions.
+Moreover, the ``logit-simple'' results in relatively sparse models with no more than 14,000 predictors entering the models even for the linear model of 3000 causal SNPs (Figure \ref{fig:supp-triple}).
+The ``logit-triple'' is comparable in terms of number of predictors.
+Finally, for other predictive performance measures, the percentage of cases in either the 10\% or 20\% largest PRSs, we report that their values are correlated to the values of AUCs at more than 98\%, which is why we don't speak about them more in the results (Figure \ref{fig:supp-AUC-corr}).
 
-\subsection{Simulations varying training size and number of SNPs}
+\subsection{Real data results}
+
+[MONTRER UN MANHATTAN PLOT POUR JUSTIFIER 30 IN HLA]
+
+[MONTRER UN GRAPH EN FCT D'INCLUSION]
+
+We also the whole Celiac dataset with the original real phenotypes.
+For one run, splitting the dataset in 12,000 indiduals in the training set and the remaining 3155 individuals in the test set, we compare the ``regularization paths'' of all methods (T-Trees excepted), that is to say the predictive performance as a function of the number of predictors corresponding to different inclusion thresholds of the methods.
 
@@
 
-In this comparative study, we proposed other methods as a replacement to the widely used P+T method. 
+In this comparative study, we propose other methods as a replacement to the widely used P+T method. 
 We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really good performance, as reported in a previous study \cite[]{abraham2013performance}, higher than with the P+T method.
@@
 
+[DISCUSSION]As for the P+T method itself, we recall that we report the maximum AUCs of 102 thresholds, a threshold that should normally be learned on the training set only.
+The choice of this threshold is very important as it can greatly impact the
+
 [PARLER DU PROBLEME DE POP? + future paper]
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 The blue dotted line represents the maximum achievable AUC.
-[MODEL AND H2]}\label{fig:main-AUC-logit}
+[MODEL AND H2]}
+\label{fig:main-AUC-logit}
 \end{figure}
@@
 The blue dotted line represents the maximum achievable AUC.
-[MODEL AND H2]}\label{fig:main-AUC-PRS}
+[MODEL AND H2]}
+\label{fig:main-AUC-PRS}
 \end{figure}
@@
 The blue dotted line represents the maximum achievable AUC.
-[MODEL AND H2]}\label{fig:main-AUC-ntrain}
+[MODEL AND H2]}
+\label{fig:main-AUC-ntrain}
 \end{figure}
@@
 \begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6}}
+\caption{Scenario \textnumero4. Boxplots of AUC over 100 simulations for ..the three reported PRSs... and the simple logistic regression.
+Vertical panels represents different number of causal SNPs and their location (``all'': anywhere on the genome; ``HLA'': only in the HLA region).
+Horizontal panels are presenting results for effets following a Gaussian or Laplace distribution and an heritability of 0.5 or 0.8.
+The blue dotted line represents the maximum achievable AUC.
+[MODEL AND H2]}
+\label{fig:supp-AUC-chr6}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newpage
+\begin{figure}[h]
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-logit}}
-\caption{[COPY FOR H2=0.5]}\label{fig:supp-AUC-logit}
+\caption{[COPY FOR H2=0.5]}
+\label{fig:supp-AUC-logit}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS}}
-\caption{[COPY FOR H2=0.5]}\label{fig:supp-AUC-PRS}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-logit-fancy}}
-\caption{[COPY FOR H2=0.5 \& MODEL FANCY]}\label{fig:supp-AUC-logit-fancy}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS-fancy}}
-\caption{[COPY FOR H2=0.5 \& MODEL FANCY]}\label{fig:supp-AUC-PRS-fancy}
+\caption{[COPY FOR H2=0.5]}
+\label{fig:supp-AUC-PRS}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-ttrees}}
-\caption{[T-TREES 3 PLOTS - H2=0.8]}\label{fig:supp-ttrees}
+\caption{[T-TREES 3 PLOTS - H2=0.8]}
+\label{fig:supp-ttrees}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-triple}}
-\caption{[TRIPLE 3 PLOTS - H2=0.8]}\label{fig:supp-triple}
+\caption{[TRIPLE 3 PLOTS - H2=0.8]}
+\label{fig:supp-triple}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-triple}}
-\caption{[TRIPLE ONLY AUC - H2=0.5]}\label{fig:supp-AUC-triple}
+\caption{[TRIPLE ONLY AUC - H2=0.5]}
+\label{fig:supp-AUC-triple}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-logit-fancy}}
+\caption{[COPY FOR H2=0.5 \& MODEL FANCY]}
+\label{fig:supp-AUC-logit-fancy}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS-fancy}}
+\caption{[COPY FOR H2=0.5 \& MODEL FANCY]}
+\label{fig:supp-AUC-PRS-fancy}
 \end{figure}
@@
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6}}
-\caption{Scenario \textnumero4. Boxplots of AUC over 100 simulations for ..the three reported PRSs... and the simple logistic regression.
-Vertical panels represents different number of causal SNPs and their location (``all'': anywhere on the genome; ``HLA'': only in the HLA region).
-Horizontal panels are presenting results for effets following a Gaussian or Laplace distribution and an heritability of 0.5 or 0.8.
-The blue dotted line represents the maximum achievable AUC.
-[MODEL AND H2]}\label{fig:supp-AUC-chr6}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-corr}}
-\caption{Scenario \textnumero3. Comparison of different predictive measures. Correlation is greater than 98\%.[ADD]}\label{fig:supp-AUC-corr}
+\caption{Scenario \textnumero3. Comparison of different predictive measures. Correlation is greater than 98\%.[ADD]}
+\label{fig:supp-AUC-corr}
 \end{figure}

### 2017-12-15 5-results-celiac.Rmd
save

@@
 ```{r}
-library(ggplot2)
+library(tidyverse)
 myggplot <- function(..., coeff = 1) {
@@
 )
+snp_qq(gwas.train) +
+  coord_cartesian(ylim = c(0, 25))
 gwas.train.gc <- snp_gc(gwas.train)
@@
 )
-```
-
-```{r}
-ggsave("figures/supp-celiac-man.png", scale = 1/90, width = 1300, height = 1000)
 ```
@@
 ```
+
+### ROC Curves
+
+```{r, fig.asp=1, fig.width=8}
+library(plotROC)
+scores_tidy <- tibble(
+  d = y[ind.test], 
+  "PRS-max" =  prs[, which.max(aucs)],
+  "logit-simple" = preds2,
+  "logit-triple" = preds4
+) %>%
+  gather(key = "Method", value = "Score", -d)
+
+bigstatsr:::MY_THEME(
+  ggplot(scores_tidy, aes(d = d, m = Score, color = Method, linetype = Method)) +
+    style_roc(xlab = "1 - Specificity", ylab = "Sensitivity")
+) +
+  geom_roc(n.cuts = 0, size = 2) +
+  theme(legend.position = c(0.7, 0.3), legend.key.width = unit(4, "line")) +
+  coord_equal()
+```
+
+```{r}
+ggsave("figures/supp-celiac-roc.pdf", scale = 1/90, width = 700, height = 700)
+```
+
+## Top scores
+
+First, we get the SNP with the highest ratio of cases over controls for homozygous-recessive individuals.
+
+```{r}
+props <- sapply(cols_along(G), function(j) {
+  ccoj <- counts.control[, j]
+  mut <- `if`(ccoj[1] < ccoj[3], 1, 3)
+  `if`(counts.case[mut, j] > 100, counts.case[mut, j] / ccoj[mut], 0)
+})
+theone <- which.max(props)
+table(Geno = G[, theone], forcats::fct_recode(as.factor(y), Control = "0", Case = "1"))
+```
+
+```{r}
+# one geno
+ind0 <- which(G[ind.test, theone] == 0)
+(M <- length(ind0))
+mean(y[ind.test[ind0]]) 
+# prs max
+indPRS <- order(prs[, which.max(aucs)], decreasing = TRUE)[1:M]
+mean(y[ind.test[indPRS]])
+# logit-simple (CMSA)
+indLS <- order(preds2, decreasing = TRUE)[1:M]
+mean(y[ind.test[indLS]])
+# logit-triple (CMSA)
+indLT <- order(preds4, decreasing = TRUE)[1:M]
+mean(y[ind.test[indLT]])
+```
+
+## Stability of selection of the "logit-simple"
+
+Train on 100 bootstrap samples of the Celiac dataset.
+
+```{r, eval=FALSE}
+if (!dir.exists("results6")) dir.create("results6")
+
+for (i in seq_len(100)) {
+
+  res.file <- paste0("results6/simu_", i, ".rds")
+  if (file.exists(res.file)) next
+
+  ind.train <- sort(sample(nrow(G)))
+
+  cmsa.logit <- big_CMSA(FUN = big_spLogReg, feval = AUC,
+                         X = G, y.train = y[ind.train], 
+                         ind.train = ind.train, 
+                         covar.train = obj.svd$u[ind.train, ],
+                         alpha = 0.5, dfmax = 20e3, 
+                         ncores = NCORES)
+
+  saveRDS(cmsa.logit[cols_along(G)], file = res.file)
+}
+```
+
+```{r}
+results6 <- list.files("results6", full.names = TRUE) %>%
+  map(~readRDS(.x)) %>%
+  do.call("cbind", .)
+```
+
+```{r}
+n_enter <- rowSums(results6 != 0)
+mean(n_enter == 0)                ## Proportion of SNPs never entering
+n_enter_once <- n_enter[n_enter != 0]
+mean(n_enter_once)                ## Mean number of times a SNPs enter 
+                                  ## if entering at least once
+bigstatsr:::MY_THEME(
+  qplot(n_enter_once, geom = "bar") +
+    labs(x = "Number of times")
+)
+```
+
+```{r}
+ggsave("figures/supp-celiac-stabilitybar.pdf", scale = 1/90, width = 1000, height = 700)
+```
+
+```{r}
+length(always_enter <- which(n_enter == ncol(results6)))
+cowplot::plot_grid(
+  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels, 
+                ind.highlight = always_enter),
+  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels, 
+                ind.highlight = always_enter) +
+    coord_cartesian(ylim = c(0, 25)) + 
+    geom_hline(yintercept = -log10(5e-8), color = "red", linetype = 3),
+  align = "hv", ncol = 1, labels = LETTERS[1:2], label_size = 25, scale = 0.95
+)
+```
+
+```{r}
+ggsave("figures/supp-celiac-man-col.png", scale = 1/90, width = 1300, height = 1000)
+```
+
+```{r}
+sq_effects_always <- colSums((results6[always_enter, ])^2)
+sq_effects_others <- colSums((results6[-always_enter, ])^2)
+summary(100 * sq_effects_always / (sq_effects_always + sq_effects_others))
+```

### 2017-12-15 paper/draft.tex
save

@@
 
-[MONTRER UN GRAPH EN FCT D'INCLUSION]
+We also analyze the whole Celiac dataset with the original real phenotypes.
+For one run, splitting the dataset in 12,000 indiduals in the training set and the remaining 3155 individuals in the test set, we compare the ``regularization paths'' of all methods (T-Trees excepted), that is to say the predictive performance as a function of the number of predictors corresponding to different inclusion thresholds of the methods.
+We show that the regularization paths of both logistic regressions are both very smooth, reaching a maximum AUC or more than 89\% at approximately 1000 predictors (Figure \ref{fig:main-celiac-regpath}).
+Moreover, the CMSA procedure give AUC very close to the maximum AUC of the regularization path.
+For the P+T method, on the contrary, the regularization path is not smooth at all and the AUC never exceed the AUC given by the most stringent threshold [ADD VALUE OF AUC] ($p < 10^{-100}$).
+We also show some associated ROC curves [QUOI DIRE DE CES ROC?] (Figure \ref{fig:supp-celiac-roc}).
 
-We also the whole Celiac dataset with the original real phenotypes.
-For one run, splitting the dataset in 12,000 indiduals in the training set and the remaining 3155 individuals in the test set, we compare the ``regularization paths'' of all methods (T-Trees excepted), that is to say the predictive performance as a function of the number of predictors corresponding to different inclusion thresholds of the methods.
+Then, we investigate the stability of SNP selection of the ``logit-simple''.
+For that, we ..replicate..
+100 times the following procedure: we sample with replacement all individuals and train the ``logit-simple'' and report the effects of each SNPs (for which a lot are 0).
+Over these 100 replicates, [VAL]\% of the SNPs never enter the predictions and [VAL] enter all predictions (Figure \ref{fig:supp-celiac-stabilitybar}).
+If we compute the sum of squared scores of the SNPs that enter every model divided by the sum of all squared scores [HAS A NAME?], we end up with a ratio of more than 99\% for all predictions, showing that most of the predictive performance is in these SNPs always entering the models.
+Moreover, these selected SNPs correspond to peaks of the GWAS, but also to less significant SNPs (Figure \ref{fig:supp-celiac-man-col}).
 
@@
 \label{fig:main-AUC-ntrain}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{main-celiac-regpath}}
+\caption{``Regularization paths'' for the P+T, ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.
+For the P+T method, the dotted line represents the maximum AUC while for the logistic regressions, the dotted lines represent the value using the CMSA procedure to be free of the choice of the threshold.}
+\label{fig:main-celiac-regpath}
 \end{figure}
@@
 
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-celiac-roc}}
+\caption{ROC Curves for the P+T, ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.}
+\label{fig:supp-celiac-roc}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-celiac-stabilitybar}}
+\caption{Barplot showing how many times SNPs enter the predictions among predictions for 100 samples with replacement of the Celiac dataset. The bar for $0$ is not displayed for sake of visibility.}
+\label{fig:supp-celiac-stabilitybar}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-celiac-man-col}}
+\caption{Manhattan plot of the Celiac dataset. Red points are presenting SNPs that are always selected in 100 replicates of [...].}
+\label{fig:supp-celiac-man-col}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \end{document}

### 2017-12-15 celiac-top-cases.R

@@ -1,59 +1,0 @@
-theone <- which.min(predict(gwas.train.gc))
-table(G[ind.test, theone], y[ind.test])
-
-counts.case <- big_counts(G, ind.row = which(y == 1))
-counts.control <- big_counts(G, ind.row = which(y == 0))
-
-counts.case[, theone]
-counts.control[, theone]
-
-props <- sapply(cols_along(G), function(j) {
-  ccoj <- counts.control[, j]
-  mut <- `if`(ccoj[1] < ccoj[3], 1, 3)
-  `if`(counts.case[mut, j] > 20, counts.case[mut, j] / ccoj[mut], 0)
-})
-hist(props)
-hist(props[props > 1])
-ord.prop <- order(props, decreasing = TRUE)
-head(ord.prop, 10)
-lapply(ord.prop[1:10], function(j) table(G[, j], y))
-
-
-# one geno
-ind0 <- which(G[ind.test, ord.prop[1]] == 0)
-mean(y[ind.test[ind0]])    ## 78.2%
-# prs max
-(M <- length(ind0))
-indPRS <- order(prs[, which.max(aucs)], decreasing = TRUE)[1:M]
-mean(y[ind.test[indPRS]])  ## 75.2%
-# logit-simple (CMSA)
-indLS <- order(preds2, decreasing = TRUE)[1:M]
-mean(y[ind.test[indLS]])   ## 79.4 %
-# logit-triple (CMSA)
-indLT <- order(preds4, decreasing = TRUE)[1:M]
-mean(y[ind.test[indLT]])   ## 82.4 %
-
-plot(pROC::roc(y[ind.test], prs[, which.max(aucs)]))
-plot(pROC::roc(y[ind.test], preds2), add = TRUE, col = "blue")
-plot(pROC::roc(y[ind.test], preds4), add = TRUE, col = "red")
-
-library(tidyverse)
-library(plotROC)
-scores_tidy <- tibble(
-  d = y[ind.test], 
-  PRS_max =  prs[, which.max(aucs)],
-  LS = preds2,
-  LT = preds4
-) %>%
-  gather(key = "Method", value = "Score", -d)
-scores_tidy
-  
-bigstatsr:::MY_THEME(
-  ggplot(scores_tidy, aes(d = d, m = Score, color = Method, linetype = Method)) +
-    style_roc(xlab = "1 - Specificity", ylab = "Sensitivity")
-) +
[…]

### 2017-12-15 get-results.R

@@ -1,19 +1,0 @@
-tmp1 <- PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test)
-map_dbl(tmp1$eval, ~AUC(.x[, 1], .x[, 2]))
-
-tmp2 <- logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple")
-
-
-tmp3 <- logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple")
-
-
-tmp4 <- ttrees(TTree, "backingfiles/ttrees", pheno.all, ind.train, ind.test, n.trees = 10)
-
-tmp <- bind_rows(tmp1, tmp2, tmp3, tmp4)
-tmp$AUC <- map_dbl(tmp$eval, ~AUC(.x[, 1], .x[, 2]))
-tmp
-
-param1 <- params.grid[1, ]
-param1$res <- list(tmp)
-test <- unnest(param1, res)
-bind_cols(param1[rep(1, 6), ], tmp)

### 2017-12-15 oracle.R

@@ -1,85 +1,0 @@
-get_oracle <- function(
-  G,                                        ## matrix of genotypes
-  h2,                                       ## heritability 
-  M,                                        ## nbs of causal variants
-  ind.possible = cols_along(G),             ## indices of possible causal variants
-  effects.dist = c("gaussian", "laplace"),  ## distribution of effects 
-  model = c("simple", "fancy"),             ## model for simulation
-  K = 0.3                                   ## prevalence
-) {
-  
-  effects.dist  <- match.arg(effects.dist)
-  model <- match.arg(model)
-  
-  set <- sample(ind.possible, size = M)
-  effects <- `if`(effects.dist == "gaussian", 
-                  rnorm(M, sd = sqrt(h2 / M)),
-                  rmutil::rlaplace(M, s = sqrt(h2 / (2*M))))
-  
-  if (model == "simple") {
-    # only linear
-    y.simu <- scale(G[, set]) %*% effects
-  } else {
-    sets <- split(1:M, sample(rep_len(1:3, M)))
-    # linear
-    ind1 <- sets[[1]]
-    y.simu <- scale(G[, set[ind1]]) %*% effects[ind1]
-    # recessive / dominant
-    ind2 <- sets[[2]]
-    y.simu <- y.simu + scale(G[, set[ind2]] > 0.5) %*% effects[ind2]
-    # interactions
-    ind3 <- matrix(sets[[3]], ncol = 2)
-    y.simu <- y.simu + scale(G[, set[ind3[, 1]]] * G[, set[ind3[, 2]]]) %*% 
-      effects[ind3[, 1]] * sqrt(2)
-  }
-  
-  y.simu <- y.simu / sd(y.simu) * sqrt(h2)
-  stopifnot(all.equal(drop(var(y.simu)), h2))
-  S <- y.simu
-  y.simu <- S + rnorm(nrow(G), sd = sqrt(1 - h2))
-  pheno <- as.numeric(y.simu > qnorm(1 - K))
-  AUC(S, pheno)
-}
-
-get_oracle(G, 0.8, 30, effects.dist = "gaussian", model = "simple")
-get_oracle(G, 0.8, 3000, effects.dist = "gaussian", model = "simple")
-
-
-params.grid2 <- expand.grid(
-  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
[…]

### 2017-12-15 reg-paths.R

@@ -1,57 +1,0 @@
-# Simulate some phenotypes
-pheno.all <- get_pheno(G, h2 = 0.8, M = 30,
-                       ind.possible = ind.HLA,
-                       effects.dist = "gaussian", 
-                       model = "simple")
-
-# Split in training/test sets
-ind.train <- sort(sample(n, size = n.train))
-ind.test <- setdiff(1:n, ind.train)
-
-# GWAS
-gwas.train <- big_univLogReg(
-  G, pheno.all[ind.train], ind.train = ind.train, 
-  covar.train = covar.all[ind.train, , drop = FALSE], 
-  ncores = nb_cores()
-)
-gwas.train.gc <- snp_gc(gwas.train)
-
-# Clumping on the test set
-ind.keep <- snp_clumping(G, infos.chr = CHR,
-                         ind.row = ind.test,
-                         thr.r2 = 0.2, 
-                         S = abs(gwas.train.gc$score),
-                         size = 500,
-                         is.size.in.bp = TRUE,
-                         infos.pos = POS,
-                         ncores = nb_cores())
-
-# PRS
-thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
-lpS <- -predict(gwas.train.gc)
-nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
-prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
-               ind.test = ind.test,
-               ind.keep = ind.keep,
-               lpS.keep = lpS[ind.keep], 
-               thr.list = thrs)
-
-aucs <- apply(prs, 2, AUC, target = pheno.all[ind.test])
-
-cmsa.logit <- big_spLogReg(X = G, y01.train = pheno.all[ind.train], 
-                           ind.train = ind.train, 
-                           covar.train = covar.all[ind.train, , drop = FALSE],
-                           alpha = 0.5, dfmax = 20e3)
-preds <- predict(cmsa.logit, X = G, ind.row = ind.test, 
-                 covar.row = covar.all[ind.test, , drop = FALSE])
-aucs2 <- apply(preds, 2, AUC, target = pheno.all[ind.test])
-
-library(Matrix)
-bind_rows(
-  data.frame(nb = nb.pred, AUC = aucs, method = "PRS"),
[…]

### 2017-12-15 save-scripts/celiac-effects.R

@@ -10,22 +10,22 @@ qplot(cmsa.logit[cols_along(G)], gwas.train.gc$estim,
 preds2
 
 
-obj.svd3 <- obj.svd
-obj.svd3$u[ind.train, ] <- NA
-plot(obj.svd3, type = "scores")
-
-
-preds.train <- predict(cmsa.logit, X = G, ind.row = ind.train, 
-                       covar.row = obj.svd$u[ind.train, ])
-myglm <- glm(y[ind.train] ~ preds.train, family = "binomial")
-summary(myglm)
-preds2.prob <- predict(myglm, newdata = data.frame(preds.train = preds2),
-                       type = "response")
-err <- rep(NA, nrow(G))
-err[ind.test] <- (y[ind.test] != (preds2.prob > 0.5)) # (y[ind.test] - preds2.prob)^2
-
-plot(obj.svd3, type = "scores") + 
-  aes(color = as.factor(err), alpha = I(rep(0.6, nrow(G))))
+# obj.svd3 <- obj.svd
+# obj.svd3$u[ind.train, ] <- NA
+# plot(obj.svd3, type = "scores")
+# 
+# 
+# preds.train <- predict(cmsa.logit, X = G, ind.row = ind.train, 
+#                        covar.row = obj.svd$u[ind.train, ])
+# myglm <- glm(y[ind.train] ~ preds.train, family = "binomial")
+# summary(myglm)
+# preds2.prob <- predict(myglm, newdata = data.frame(preds.train = preds2),
+#                        type = "response")
+# err <- rep(NA, nrow(G))
+# err[ind.test] <- (y[ind.test] != (preds2.prob > 0.5)) # (y[ind.test] - preds2.prob)^2
+# 
+# plot(obj.svd3, type = "scores") + 
+#   aes(color = as.factor(err), alpha = I(rep(0.6, nrow(G))))
 
 boot1000 <- function(x) {
   sd(replicate(1000, mean(sample(x, replace = TRUE))))
@@ -69,5 +69,3 @@ data.frame(pop = celiac$fam$family.ID[ind.test],
   labs(fill = "Pheno")
 
 
-qplot(preds2, obj.svd$u[ind.test, 1]) +
-  geom_smooth(method = "lm")

### 2018-01-11 LIFE/pres.Rmd
pres LIFE

---
title: "Predicting complex diseases:<br>performance and robustness"
subtitle: "Florian Privé (UGA)"
author: "supervised by</br>M. Blum (UGA) & H. Aschard (Institut Pasteur)"
date: "January 12, 2018"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, comment = "",
                      fig.align = "center", dev = "svg")
```

## Personalized genetic medicine:

```{r, out.width="75%"}
knitr::include_graphics("https://www.frontiersin.org/files/Articles/137943/fmicb-06-00411-HTML/image_m/fmicb-06-00411-g001.jpg")
```

.footnote[
Source: Oliveira-Coelho, Ana, et al. "Paving the way for predictive diagnostics and personalized treatment of invasive aspergillosis." Frontiers in microbiology 6 (2015).
]

---

## Data

0, 1 or 2 mutations per individual and per locus (position on the genome):

```{r}
library(bigsnpr)
bigsnp <- snp_attachExtdata()
G <- bigsnp$genotypes
G[1:12, 1:11]
```

</br>

Size of current datasets (UK Biobank): 500,000 individuals and (at least) 800,000 loci.

---

## Goal

$$\boxed{\Large{\text{Disease} \sim \text{DNA mutations}} + \cdots}$$

</br>

## Methods

- statistical learning methods

- clever implementations for handling large datasets

- two R packages (Privé et al., 2017)

---

## Assessing predictive performance

AUC (Area Under the ROC Curve) is often used.

```{r, out.width="45%", fig.cap="Example of ROC curve."}
knitr::include_graphics("https://i.stack.imgur.com/5x3Xj.png")
```

</br>

$$\text{AUC} = P(S_\text{case} > S_\text{control})$$

---

## Results (simulating different disease architectures)

```{r}
knitr::include_graphics("main-AUC-logit.svg")
```

---

class: center, middle, inverse

# Thanks!

<br>

Twitter and GitHub: [@privefl](https://twitter.com/privefl)

.footnote[Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).]

### 2018-02-12 3-simus.Rmd
htmls
save

@@
 library(tidyverse)
+library(glue)
 ```
@@
 
+  timing <- system.time({
+
     # GWAS
@@
     gwas.train.gc <- snp_gc(gwas.train)
-
-  timing <- system.time({
 
@@
 
-    cmsa.logit <- big_CMSA(FUN = big_spLogReg, feval = AUC,
-                           X = G, y.train = pheno.all[ind.train], 
+    cmsa.logit <- big_spLogReg(X = G, y01.train = pheno.all[ind.train], 
                                ind.train = ind.train, 
                                covar.train = covar.all[ind.train, , drop = FALSE],
-                           alpha = 0.5, dfmax = 20e3, 
                                ncores = nb_cores())
 
-    preds <- predict(cmsa.logit, X = G, ind.row = ind.test, 
+    preds <- rowMeans(
+      predict(cmsa.logit, X = G, ind.row = ind.test, 
                      covar.row = covar.all[ind.test, , drop = FALSE])
+    )
   })[3]
@@
     timing   = timing,
-    nb.preds = sum(cmsa.logit != 0)
+    nb.preds = sum(rowSums(sapply(cmsa.logit, function(x) x$beta.X) != 0) > 0)
   )
@@
 n <- nrow(G)
-n.train <- 6e3
 ind.HLA <- snp_indLRLDR(CHR, POS, subset(LD.wiki34, ID == "hild12"))
@@
 
-### Simulations n°1 (with T-Trees)
+### Scenario n°1 (with T-Trees)
 
@@
 params.grid1 <- expand.grid(
+  n.train    = 6000,
   par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
@@
 
-```{r, eval=FALSE}
+```{r, eval=TRUE}
 if (!dir.exists("results1")) dir.create("results1")
@@
   # Split in training/test sets
-  ind.train <- sort(sample(n, size = n.train))
+  ind.train <- sort(sample(n, size = params[["n.train"]]))
   ind.test <- setdiff(1:n, ind.train)
@@
 
-### Simulations n°2 (without T-Trees, with logit-triple)
+### Scenario n°1 (without T-Trees)
 
@@
 params.grid2 <- expand.grid(
+  n.train = 6000,
   par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
@@
   par.model  = c("simple", "fancy"),
-  num.simu   = 1:20,
+  num.simu   = 1:100,
   stringsAsFactors = FALSE
@@
 
-```{r, eval=FALSE}
+```{r, eval=TRUE}
 if (!dir.exists("results2")) dir.create("results2")
@@
 
-### Simulations n°3 (without logit-triple)
+### Scenario n°2 (dataset with only chromosome 6)
 
 ```{r}
-params.grid3 <- expand.grid(
+params.grid4 <- expand.grid(
+  n.train    = 6000,
   par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
@@
   par.h2     = c(0.5, 0.8), 
-  par.model  = c("simple"),
-  num.simu   = 21:100,
-  stringsAsFactors = FALSE
-) 
-```
-
-```{r, eval=FALSE}
-if (!dir.exists("results3")) dir.create("results3")
-
-for (i in rows_along(params.grid3)) {
-
-  res.file <- paste0("results3/simu_", i, ".rds")
-  if (file.exists(res.file)) next
-
-  params <- params.grid3[i, ]
-  par.causal <- params[["par.causal"]][[1]]
-
-  # Simulate phenotypes
-  pheno.all <- get_pheno(
-    G,    
-    h2 = params[["par.h2"]], 
-    M = as.integer(par.causal[1]), 
-    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
-    effects.dist = params[["par.dist"]], 
-    model = params[["par.model"]], 
-    K = 0.3                 
-  )
-
-  # Split in training/test sets
-  ind.train <- sort(sample(n, size = n.train))
-  ind.test <- setdiff(1:n, ind.train)
-
-  # Get results from all methods
-  res <- bind_rows(
-    PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test),
-    logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple")
-  )
-  params[["res"]] <- list(res)
-  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
-}
-```
-
-### Simulations n°4 (dataset with only chromosome 6)
-
-```{r}
-params.grid4 <- expand.grid(
-  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
-  par.dist   = c("gaussian", "laplace"), 
-  par.h2     = c(0.5, 0.8), 
-  par.model  = c("simple"),
+  par.model  = "simple",
   num.simu   = 1:100,
@@
 
-### Simulations n°5 (varying training size)
+### Scenario n°3 (varying training size)
 
@@
   par.h2     = c(0.5, 0.8), 
-  par.model  = c("simple"),
+  par.model  = "simple",
   num.simu   = 1:100,

### 2018-02-12 4-results-simus.Rmd
save

--- 2017-12-12 4-results-simus.Rmd
@@
 ```{r}
+pAUC <- function(pred, target, p = 0.1) {
+  val.min <- min(target)
+  q <- quantile(pred[target == val.min], probs = 1 - p)
+  ind <- (target != val.min) | (pred > q)
+  tryCatch(bigstatsr::AUC(pred[ind], target[ind]) * p, 
+           error = function(e) p^2 / 2)  # all same prediction
+}
+
 format_results <- function(results) {
@@
                           levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
-      AUC = map_dbl(eval, ~bigstatsr::AUC(.x[, 1], .x[, 2]))
+      AUC = map_dbl(eval, ~ bigstatsr::AUC(.x[, 1], .x[, 2])),
+      pAUC = map_dbl(eval, ~ pAUC(.x[, 1], .x[, 2]))
     )
@@
     geom_hline(yintercept = 0.5, linetype = 2) +
-    geom_bar(stat = "identity", alpha = 0.3, position = position_dodge()) +
+    geom_col(position = position_dodge(), alpha = 0.3) +
     geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
@@
   nsimu <- results.summary$N[[1]]
-  stopifnot(all(results.summary$N == nsimu))
+  # stopifnot(all(results.summary$N == nsimu))
 
@@
 ```{r, out.width="90%", fig.width = 14}
-compare_logit(bind_rows(results1, results2), with_method = "logit-triple")
+compare_logit(results2, with_method = "logit-triple")
 
@@
   barplot_causal_one(h2 = 0.5) +
-  facet_grid(par.dist ~ par.model) +
-  labs(y = "Mean of 20 AUCs")
+  facet_grid(par.model ~ par.dist)
 
@@
   filter(par.model == "fancy", method %in% c("logit-simple", "PRS-max")) %>%
-  barplot_causal_all() +
-  labs(y = "Mean of 20 AUCs")
+  barplot_causal_all()
 
@@
   filter(par.model == "fancy", grepl("PRS", method)) %>%
-  barplot_causal_all() +
-  labs(y = "Mean of 20 AUCs")
+  barplot_causal_all()
 
@@
 
-### Correlation between predictive measures
+### Correlation between predictive measures [MOVE TOP]
 
 ```{r}
-results3 <- list.files(paste0("results", 2:3), full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  filter(method != "logit-triple", par.model != "fancy") %>%
-  format_results()
-
-results3_corr <- results3 %>%
-  filter(method %in% c("logit-simple", "PRS-max")) %>%
-  mutate(
-    percCases10 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[1:110], 2])),
-    percCases20 = map_dbl(eval, ~mean(.x[order(.x[, 1], decreasing = TRUE)[1:220], 2])),
-    Parameters  = interaction(paste("h2 =", par.h2), par.dist, method, sep = " | ")
-  )
+results3 <- results2 %>%
+  filter(par.model == "simple", method != "logit-triple")
 ```
 
-```{r, fig.asp=0.5, out.width="90%", fig.width=15}
-p1 <- results3_corr %>%
-  myggplot(aes(AUC, percCases10, color = Parameters)) + 
+```{r, fig.asp=0.5, out.width="95%", fig.width=16}
+results3 %>%
+  filter(method %in% c("logit-simple", "PRS-max")) %>%
+  mutate(Parameters  = interaction(par.h2, par.dist, method, sep = " | ")) %>%
+  myggplot(aes(AUC, pAUC, color = Parameters)) + 
   geom_point(size = 0.6, alpha = 0.5) +
-  geom_smooth(aes(linetype = Parameters), method = "lm", se = FALSE, size = 2) +
+  geom_smooth(aes(linetype = Parameters), method = "loess", se = FALSE, size = 2) +
   theme(legend.key.width = unit(3.5, "line"), 
         legend.direction = "horizontal") + 
-  labs(y = "Proportion of cases among top 10% scores")
-p2 <- results3_corr %>%
-  myggplot(aes(AUC, percCases20, color = Parameters)) + 
-  geom_point(size = 0.6, alpha = 0.5) +
-  geom_smooth(aes(linetype = Parameters), method = "lm", se = FALSE, size = 2) + 
-  labs(y = "Proportion of cases among top 20% scores") +
-  theme(legend.position = "none")
+  labs(y = "Partial AUC (specificity: [0.9, 1])") +
+  theme(legend.position = "top")
 
-cowplot::plot_grid(p1 + theme(legend.position = "none"), p2 , 
-                   ncol = 2, align = "hv", scale = 0.95) %>%
-  cowplot::plot_grid(cowplot::get_legend(p1), ., rel_heights = c(0.1, 1), ncol = 1)
+ggsave("figures/supp-AUC-corr.pdf", scale = 1/100, width = 1525, height = 983)
+```
 
-ggsave("figures/supp-AUC-corr.pdf", scale = 1/90, width = 1558, height = 737)
-
-with(results3_corr, cor(AUC, percCases10))
-with(results3_corr, cor(AUC, percCases20))
+```{r}
+cor(results3$AUC, results3$pAUC, method = "spearman")
 ```
@@
 results3.summary %>%
-  filter(method %in% c("logit-simple", "PRS-max")) %>%
+  filter(method %in% c("logit-simple", "PRS-max"), par.model == "simple") %>%
   barplot_causal_one(h2 = 0.8)

### 2018-02-12 paper/draft.tex
save

@@
 \vspace*{0.5em}
-\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|c|c|}
 \hline
-Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & Number of & \multirow{2}{*}{Methods} \\ 
-scenario & & training set & (number and location) & of effects & & model & simulations & \\
+Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
+scenario & & training set & (number and location) & of effects & & model & \\
 \hline
 \hline
-\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{4}{*}{0.8} & \multirow{2.5}{*}{simple} & \multirow{4}{*}{5} & PRS \\
-& & & 30 in all & & & & & logit-simple \\
-& & & 300 in all & \multirow{1.5}{*}{Laplace} & & \multirow{1.5}{*}{fancy} & & logit-triple \\
-& & & 3000 in all & & & & & T-Trees \\
+\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{simple} & PRS \\
+& & & 30 in all & & & & logit-simple \\
+& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{fancy} & logit-triple \\
+& & & 3000 in all & & & & (T-Trees) \\
 \hline
-\multirow{3}{*}{2} & \multirow{3}{*}{-} & \multirow{3}{*}{-} & \multirow{3}{*}{-} & \multirow{3}{*}{-} & \multirow{2}{*}{0.5} & \multirow{3}{*}{-} & \multirow{3}{*}{20} & PRS \\ 
-& & & & & \multirow{2.2}{*}{0.8} & & & logit-simple \\
-& & & & & & & & logit-triple \\
+\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{simple} & PRS \\ 
+& & & & & & & logit-simple\\
 \hline
-\multirow{2}{*}{3} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{simple} & \multirow{2}{*}{80} & PRS \\ 
-& & & & & & & & logit-simple \\
-\hline
-\multirow{2}{*}{4} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{100} & \multirow{2}{*}{-} \\
-& & & & & & & & \\
-\hline
-\multirow{5}{*}{5} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{100} & \multirow{5}{*}{-} \\
-& & 2000 & & & & & & \\
-& & 3000 & & & & & & \\
-& & 4000 & & & & & & \\
-& & 5000 & & & & & & \\
+\multirow{5}{*}{3} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
+& & 2000 & & & & & \\
+& & 3000 & & & & & \\
+& & 4000 & & & & & \\
+& & 5000 & & & & & \\
 \hline

### 2018-02-12 LIFE/HTML2PDF.R

@@ -1,0 +1,23 @@
+auc_max <- 0.94
+
+results3.summary %>%
+  filter(method %in% c("logit-simple", "PRS-max")) %>%
+  mutate(method = forcats::fct_recode(method, "Our method" = "logit-simple",
+                                      "State-of-the-art" = "PRS-max")) %>%
+  filter(par.h2 == 0.8, par.dist == "laplace") %>%
+  barplot_causal() +
+  geom_hline(yintercept = auc_max, linetype = 3, color = "blue") +
+  scale_fill_discrete() +
+  scale_color_discrete()
+
+ggsave("LIFE/main-AUC-logit.svg", scale = 1/90, width = 844, height = 572)
+
+prefix <- "LIFE/pres"
+path <- "../../useR-2017/decktape-1.0.0/"
+system(glue::glue(
+  "{path}phantomjs",
+  " {path}decktape.js", 
+  " {prefix}.html {prefix}.pdf", 
+  " --load-pause 1000", 
+  " -s 1504x1129"
+))

### 2018-02-26 paper/draft.tex
modif draft -- need results

@@
 Personalized medicine is another major application of PRSs.
-Personalized medicine will use PRSs in screening campaigns in order to identify high-risk individuals for a given disease.
+Personalized medicine ..will..
+use PRSs in screening campaigns in order to identify high-risk individuals for a given disease.
 As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}.
@@
 Yet, PRSs would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases. 
-For example, for respectively screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than respectively 75\% and 99\% \cite[]{janssens2007impact}.
+For example, for screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than respectively 75\% and 99\% \cite[]{janssens2007impact}.
 
 Many methods have been developed in order to maximize the predictive power of genotypes for diseases, or more generally for phenotypes.
-A commonly used technique, called P+T (which stands for ``Pruning and Thresholding'') -- or genetic profiling ..or even just PRS..
+A commonly used technique, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or genetic profiling ..or even just PRS..
 -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
-This technique only use summary statistics which makes it very ..suitable..
-and also very fast.
+This technique only use summary statistics which makes it very suitable and also very fast.
 Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}.
-Yet, these models are not ..constructed..
-for predicting a binary trait such as a disease status and have proven to ..fail..
-at such task in another comparative study \cite[]{abraham2013performance}.
-Moreover, these methods and their ..derivations..
-are often computationally demanding, both in terms of memory and time ..requirements.., which makes them unlikely to be used for prediction on large cohorts \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
-Statistical learning methods have also been used to derive PRSs for complex human disease by jointly estimating SNP effects.
+Yet, these models are not designed for predicting a binary trait such as a disease status and have proven to not be optimal at such task in another comparative study \cite[]{abraham2013performance}.
+Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time requirements, which makes them unlikely to be used for prediction on large cohorts \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
+Finally, statistical learning methods have also been used to derive PRSs for complex human disease by jointly estimating SNP effects.
 Such methods include logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2013walk,botta2014exploiting,wei2009disease}.
 
-We recently developed two R packages, bigstatsr and bigsnpr, for efficient management and analysis of large-scale genome-wide data \cite[]{prive2017efficient}.
-Package bigstatsr includes efficient functions for computing penalized linear and logistic regressions on huge datasets.
-In this paper, we present a comprehensive comparative study of the P+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in \cite{botta2014exploiting}.
-Note that the SVM model is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp} and therefore isn't added to the comparison.
-For the P+T model, we compare different thresholds of inclusion of SNPs.
+We recently developed two R packages, bigstatsr and bigsnpr, for efficient analysis of large-scale genome-wide data \cite[]{prive2017efficient}.
+Package bigstatsr includes efficient functions for computing sparse linear and logistic regressions on huge datasets.
+In this paper, we present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in \cite{botta2014exploiting}.
+Note that in this comparison we don't include any LMM method for the reasons mentioned before and don't include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
+For the C+T model, we compare different thresholds of inclusion of SNPs.
 For the logistic regression, we include two novel approaches. 
-First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for choosing the amount of regularization used, which directly affects the number of SNPs included in the model.
+First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for automatically choosing the amount of regularization used (one hyper-parameter of the penalized logistic regression), which directly affects the number of SNPs included in the model.
 We also show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.
@@
 There have already been comparative studies about predicting human binary phenotypes based on genotype data. 
-\citep{abraham2013performance} showed that penalized logistic regression and penalized SVM performed better than the standard P+T method and LMMs, for a wide-range of human diseases.
+\citep{abraham2013performance} showed that penalized logistic regression and penalized SVM performed better than the standard C+T method and LMMs, for a wide-range of human diseases.
 \cite{zhou2013polygenic} used simulations on real genotypes of Australian individuals to show that a hybrid of linear mixed models and sparse regression models performed better in a range of sparse to polygenic disease architectures.
 \cite{spiliopoulou2015genomic} investigated the effect of relatedness in prediction performance.
-\cite{ware2017heterogeneity} evaluated best practices for using the P+T method and found out, for example, that the optimal inclusion threshold is trait-dependent. 
+\cite{ware2017heterogeneity} evaluated best practices for using the C+T method and found out that the optimal inclusion threshold is trait-dependent. 
 In this study, we particularly extend the work of \cite{abraham2013performance} by using a simulation framework similar but ..richer..
 than the one used in \cite{zhou2013polygenic}.
-In order to make our comparison as comprehensive as possible, we compare different architectures of disease (number, size and location of causal effects and heritability) with different model of generating liability ..scores.., one with only linear effects, and one which combines linear, dominant and interaction effects.
+In order to make our comparison as comprehensive as possible, we compare different architectures of disease (number, size and location of causal effects and heritability) with different model for generating liability ..scores.., one with only linear effects, and one which combines linear, dominant and interaction effects.
 First, we quickly discard T-Trees as being the best method for predicting disease status based on genotypes.
-Then, we show that penalized logistic regression consistently performs better than the P+T method whereas predictive performance of the P+T method is very sensitive to the threshold of inclusion of SNPs, depending of the architecture of disease, as shown in \cite{ware2017heterogeneity}.
+Then, we show that penalized logistic regression consistently performs better than the C+T method whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending of the architecture of disease, as shown in \cite{ware2017heterogeneity}.
 
@@
 Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
-For simulations, in order to remove the structure induced by the celiac disease status, we first keep only controls from this cohort.
-Secondly, in order to remove population structure, we keep only people from the UK.
-Thirdly, we further subset these British people based on deviation of robust Mahalanobis distance on principal components.
-This 3-step filtering results in 7102 individuals with minimal population structure (see ..SupMat..).
+For simulations, in order to remove the structure induced by the celiac disease status, we ..first keep only..
+controls from this cohort.
+Secondly, in order to remove population structure, we ..keep only..
+people from the UK.
+Thirdly, we further subset these British people based on deviation from the robust Mahalanobis distance on principal components.
+This 3-step filtering results in 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing'').
 We decided to get rid of population structure because it can affect the predictive performance of methods, which we will investigate in another study.
@@
 
-To simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\%.
-We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only the chromosome 6), the size of the training set (from 1000 to 6000 by 1000, and the corresponding test set is composed of the remaining individuals), the number of causal variants and their location (30, 300 or 3000 anywhere on the dataset or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part..
+To simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% [CITATION?].
+We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only the chromosome 6), the size of the training set (..from 1000 to 6000 by 1000.., and the corresponding test set is composed of the remaining individuals), the number of causal variants and their location (30, 300 or 3000 anywhere on the dataset or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part..
 of the phenotypes (one ``simple model'' with only linear effects, and one ``fancy model'' which combines linear, dominant and interaction effects).
@@
 Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still display a marginal effect.
-For both models, the $y_i$ are then standardized such that they have a variance equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$ where $\epsilon \sim N(0, 1 - h^2)$.
+For both models, the $y_i$ are then standardized to make sure they have a variance exactly equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$ where $\epsilon \sim N(0, 1 - h^2)$.
 
-We make 5 different simulation scenarios, summarized in table \ref{tab:simus}. 
-Scenario \textnumero1 uses the whole dataset (the 22 chromosomes), a training set of size 6000 and an heritability of 80\%.
-It compares all methods (see section \ref{sec:methods}) on all other combination of parameters in 5 simulations.
-The goal of the first scenario is to quickly discard the T-Trees method.
-For scenario \textnumero2, we make also vary the heritability (50\% or 80\%) but we don't run the T-Trees method anymore.
-The goal of the second scenario is to compare all remaining methods on all models (including the ``fancy'' model), especially to see how well the ``logit-triple'' method works.
-We run 20 simulations for each combination of parameters.
-For scenario \textnumero3, we continue with 80 more simulations (to reach 100 simulations, i.e.\ 5 times more than in \cite{zhou2013polygenic}), but discarding the ``logit-triple'' method and the ``fancy'' model.
-Scenario \textnumero4 consists in 100 simulations of remaining methods and combinations of parameters on a dataset composed of only the chromosome 6.
+We make 3 different simulation scenarios, summarized in table \ref{tab:simus}. 
+Scenario \textnumero1 uses the whole dataset (the 22 chromosomes) and a training set of size 6000.
+It compares all methods (see section \ref{sec:methods}) on all other combination of parameters in 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}).
+Note that the T-Trees method is compared on only 5 simulations and an heritability of 80\%, which aims at quickly discarding this method that can require more that one hour of computation for one simulation.
+Scenario \textnumero2 consists in 100 simulations comparing the ``logit-simple'' and the C+T methods on a dataset composed of only the chromosome 6.
 Reducing the number of SNPs aims at increasing the ..polygenicity..
@@
 DUDBRIDGE?].
-Finally, scenario \textnumero5 reuses the whole dataset but varying the size of the training set in order to assess how the number of samples affects predictive performance of the PRS and the ``logit-simple'' methods.
+For this scenario, we use only the additive (``simple'') model, but continue to compare all previous variations of the other parameters.
+Finally, scenario \textnumero3 reuses the whole dataset but varying the size of the training set in order to assess how the number of samples affects predictive performance of the PRS and the ``logit-simple'' methods.
 100 simulations are run using 300 causal SNPs anywhere on the genome.
@@
 This measure indicates how well we can distinguish between cases and controls using PRSs.
-As a second measure, we report the percentage of cases in the 10\% and 20\% largest PRSs.
-This measure indicates how well we can identify high-risk individuals from a higher PRS in screening campaigns.
+As a second measure, we report the percentage of the partial AUC for a specificity between 90\% and 100\% \cite[]{dodd2003partial,mcclish1989analyzing}.
+This measure is similar to the AUC, but focuses on high specificities (i.e.\ discarding most of the healthy population as being diseased) which is often the most useful part of the ROC curve in clinical settings.
 
-Note that we also report the timing of the main computations and the number of SNPs used in the predictions.
+%Note that we also report the timing of the main computations and the number of SNPs used in the predictions.
 
@@
 
-In this study, we compare three different types of methods: the P+T method, the T-Trees method and a penalized logistic regression.
+In this study, we compare three different types of methods: the C+T method, the T-Trees method and a penalized logistic regression.
 
-The P+T (Pruning + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWASs), called summary statistics.
-For the P+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
-The SNPs are first clumped (P) so that there remain only loci that are weakly correlated with each other. 
+The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWASs), named summary statistics.
+For the C+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
+The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with each other. 
 Then, thresholding (T) consists in removing SNPs that are within a certain level of significance (p-value threshold to be determined). 
-Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$S_i(T) = \sum_{j \in E_\text{clumping}} \mathds{1}\{p_j < T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS and $T$ is a threshold on p-values to be determined.
+Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$S_i(T, E_\text{clumping}) = \sum_{j \in E_\text{clumping}} \mathds{1}\{p_j < T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS and $T$ is a threshold on p-values to be determined.
 In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.
 
-For the P+T method, we report three different scores of prediction: one including all the SNPs (remaining after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximizes the AUC for these two thresholds and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
-Note that, normally, the optimal threshold would have to be learned on the training set only.
-We consider this reported maximum AUC as an upper bound of the AUC for the P+T method.
+For the C+T method, we report three different scores of prediction: one including all the SNPs (remaining after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximizes the AUC for these two thresholds and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
+Note that, normally, this optimal threshold would have to be learned on the training set only in order to prevent from over-fitting the data.
+We consider this reported maximum AUC as an upper bound of the AUC for the C+T method.
 We call these three reported scores respectively ``PRS-all'', ``PRS-stringent'' and ``PRS-max'' in the results.
@@
 T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}.
-Since this method is significantly different from the other tested here and achieved surprisingly results in the past, we decided to include T-Trees in this comparison in order to assess whether or not it could give superior predictive performance.
+Since this method is significantly different from the other tested here and achieved surprisingly ..good..
+results in the past, we decided to include T-Trees in this comparison in order to assess whether or not it could give superior predictive performance.
 We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}.
-Yet, we use only 100 trees (instead of 1000) because it showed only a very subtle increase of AUC while taking way too much time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown).
+Yet, we use only 100 trees (instead of 1000) because it showed only a very subtle increase of AUC while ..taking way too much time..
+(e.g.\ AUC of 81.5\% instead of 81\%, data not shown).
 %We call this method ``T-Trees'' in the results.
@@
 The last method we compare is penalized logistic regression.
-We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that need to be determined.
-Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (ridge, \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation),
- the L1-regularization (lasso, \cite{tibshirani1996regression}) forces some of the coefficients to be exactly zero and can be used as a means of variable selection, leading to sparse (and therefore more interpretable) models  (corresponds to $\alpha = 1$),
-the L1- and L2-regularization (elastic-net, \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation where there are many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
+We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that need to be determined.
+Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
+the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be exactly zero and can be used as a means of variable selection, leading to sparse (and therefore more interpretable) models  (corresponds to $\alpha = 1$);
+the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation where there are many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
 In this study, we always use the elastic-net regularization with $\alpha = 0.5$, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.
 
-To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient}, also implemented in our package bigstatsr \cite[]{prive2017efficient}.
+To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in our package bigstatsr \cite[]{prive2017efficient}.
 This type of algorithm builds predictions for many values of $\lambda$ (typically a ``regularization path'' of 100 values).
@@
 First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds).
-Secondly, in turn, each fold is considered as an inner validation set and the others ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes a criterion (e.g.\ AUC) is determined and the vector of coefficients corresponding to the previous vector of scores is chosen.
-Finally, the $K$ resulting vectors of coefficients are combined into one vector (e.g.\ using the geometric median).
+Secondly, in turn, each fold is considered as an inner validation set and the others ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes the likelihood on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is chosen.
+Finally, the $K$ resulting vectors of coefficients are reported and can be combined into one vector (e.g.\ using the geometric median) or resulting scores can be combined (e.g.\ averaging the $K$ probabilities for each individual).
+We also use this procedure to derive an early stopping criterion so that the algorithm doesn't need to evaluate the whole regularization path, making the whole procedure much faster.
 Because of L1-regularization, this vector of coefficients is typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts.
@@
 In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct another dataset with, for each SNP variable, two more variables coding for recessive and dominant effects.
-This results in a dataset with 3 times more variables than the initial one, on which we can apply the standard logistic regression with the CMSA procedure, described previously.
+This results in a dataset with 3 times more variables than the initial one, on which we can apply the penalized logistic regression with the CMSA procedure, as described previously.
 We call this method ``logit-triple'' in the results.
@@
 
-In results, we use estimates of maximum achievable AUC values for the simulations: 84\% (respectively 94\%) for heritabilities of 50\% (respectively 80\%).
+In results, we use estimates of maximum achievable AUC values for the simulations: 84\% (respectively 94\%) for heritabilities of 50\% (respectively 80\%) based on three estimations (see Supplementary Materials).
 
-\subsection{Scenario \textnumero1}
+\subsection{Scenario \textnumero1 -- main methods}
 
-In scenario \textnumero1, we compared our penalized logistic regression (``logit-simple'') with the P+T method (``PRS'') on the whole genome (Table \ref{tab:simus}). 
-When there are only 30 causal SNPs, our penalized logistic regression almost reach the maximum achievable AUC (Figures \ref{fig:main-AUC-logit} \& \ref{fig:supp-AUC-logit}).
-Still in terms of AUC, our penalized regression consistently performs better or equally to the ``PRS-max''. 
+In scenario \textnumero1, we compared our penalized logistic regression (``logit-simple'') with the C+T method (``PRS'') on the whole genome (Table \ref{tab:simus}). 
+Using the penalized logistic regression, we report AUCs greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, which is not obtained by the upper bound of the C+T method in any scenario (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
+Our penalized regression methods consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted sometimes when we simulate 3000 causals SNPs anywhere on the genome, a case where all methods rarely exceeds 65\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
 In the case where there are correlation between predictors (i.e.\ when we simulate 30 SNPs in the HLA region), the ``logit-simple'' method particularly performs better than the ``PRS-max'' one, with a mean AUC of 92.5\% versus 83\%, for a maximum achievable AUC of 94\% (Figure \ref{fig:main-AUC-logit}).
-[NUL] Moreover, results of PRS can dramatically drop given the threshold of inclusion of SNPs in the model (Figures \ref{fig:main-AUC-PRS} \& \ref{fig:supp-AUC-PRS}).
-
-\subsection{Main simulations}
-
-Main simulations corresponds to scenario \textnumero3 (Table \ref{tab:simus}).
-They mainly compare our ``simple'' logistic regression to the standard P+T methods in case of a ``simple'' model (only linear effects).
-
-Using the penalized logistic regression, we report AUCs greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, which is not obtained by the upper bound of the P+T method in any scenario (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
-Our penalized regression methods consistently provides better predictive performance than the P+T method for all disease architectures tested, excepted sometimes when we simulate 3000 causals SNPs anywhere on the genome, a case where all methods rarely exceeds 65\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
-We also report that the choice of the threshold of inclusion of SNPs in the P+T method can greatly impact its performance, for example dropping from an AUC of 90\% to less than 60\% when including all SNPs passing clumping in a model with only 30 causal SNPs, or dropping from 60\% to a random predictor (50\% of AUC) when including only GWAS-significant SNPs in a model with 3000 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
-
-\subsection{Simulations varying number of SNPs and training size}
-
-These simulations corresponds to scenarios \textnumero4 and \textnumero5.
-First, when reproducing simulations of scenario \textnumero3 on only the chromosome 6, methods globally increase or ..stagnate..
-in predictive performance (Figure \ref{fig:supp-AUC-chr6}).
-There is particularly high increase in the case when simulating 3000 causal SNPs anywhere on the dataset, for example, the ``logit-simple'' can increase from less than 60\% of AUC to nearly 80\%.
-On the contrary, when simulating only 30 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' seems to slightly decrease.
-Secondly, when varying the training size, we report an increase of AUC whith training size, with the AUC of the ``logit-simple'' method increasing faster than the ``PRS-max'' (Figure \ref{fig:main-AUC-ntrain}).
+We also report that the choice of the threshold of inclusion of SNPs in the C+T method can greatly impact its performance, for example dropping from an AUC of 90\% to less than 60\% when including all SNPs passing clumping in a model with only 30 causal SNPs, or dropping from 60\% to a random predictor (50\% of AUC) when including only GWAS-significant SNPs in a model with 3000 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
 
@@
 First, in scenario \textnumero1 we also tested the T-Trees method. 
-For example, compared to our penalized logistic regression, T-Trees perform worse in term predictive measures, while taking longer to run and making more complex predictive models because it uses more SNPs in the models and has non-linear effects (Figure \ref{fig:supp-ttrees}).
-Even when simulating a fancy model in which there is dominant and interaction effects that decision trees based methods should handle, AUC is still worse than the ``logit-simple''.
-Secondly, as for the comparison between our two penalized logistic regression in scenario \textnumero2, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
-Yet, the ``triple'' solution takes approximately 3 more times to run.
-Thirdly, for other simpler methods, we report the exact same results that is to say that the ``logit-simple'' give better AUCs than the ``PRS-max'', excepted in the case of 3000 causal SNP anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}).
-Moreover, the choice of the threshold inclusion of SNPs still greatly affect the performance of the P+T method (Figure \ref{fig:supp-AUC-PRS-fancy}).
+For example, compared to our penalized logistic regression, T-Trees perform worse in term predictive measures, while taking much longer to run and making more complex predictive models because it uses more SNPs in the models and has non-linear effects (Figure \ref{fig:supp-ttrees}).
+Even when simulating a fancy model in which there is dominant and interaction effects that decision trees based methods should handle, AUC is still worse with T-Trees as compared with the ``logit-simple'' method.
 
-\subsection{Other measures of performance}
+Secondly, as for the comparison between our two penalized logistic regression in scenario \textnumero1, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
+Yet, the ``triple'' solution takes 2-3 more times to run and 3 more disk storage.
+
+Thirdly, for the ``fancy'' simulation model, we report the exact same results as in the previous subsection that is to say that the ``logit-simple'' give greater AUCs than the ``PRS-max'', excepted in the case of 3000 causal SNP anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}).
+Moreover, the choice of the threshold inclusion of SNPs also greatly affect the performance of the C+T method (Figure \ref{fig:supp-AUC-PRS-fancy}).
+
+\subsection{Simulations varying number of SNPs and training size}
+
+These simulations corresponds to scenarios \textnumero2 and \textnumero3 (Table \ref{tab:simus}).
+First, when reproducing simulations of scenario \textnumero1 on only the chromosome 6, methods globally increase or ..stagnate..
+in predictive performance (Figure \ref{fig:supp-AUC-chr6}).
+There is particularly a high increase in the case when simulating 3000 causal SNPs anywhere on the dataset: for example, AUC from the ``logit-simple'' can increase from less than 60\% to nearly 80\%.
+On the contrary, when simulating only 30 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' seems to slightly decrease.
+Secondly, when varying the training size, we report an increase of AUC when increasing the training size, with the AUC of the ``logit-simple'' method increasing faster than the ``PRS-max'' method (Figure \ref{fig:main-AUC-ntrain}).
+
+\subsection{[Other measures of performance]}
 
@@
 Moreover, the CMSA procedure give AUC very close to the maximum AUC of the regularization path.
-For the P+T method, on the contrary, the regularization path is not smooth at all and the AUC never exceed the AUC given by the most stringent threshold [ADD VALUE OF AUC] ($p < 10^{-100}$).
+For the C+T method, on the contrary, the regularization path is not smooth at all and the AUC never exceed the AUC given by the most stringent threshold [ADD VALUE OF AUC] ($p < 10^{-100}$).
 We also show some associated ROC curves [QUOI DIRE DE CES ROC?] (Figure \ref{fig:supp-celiac-roc}).
-
-Then, we investigate the stability of SNP selection of the ``logit-simple''.
-For that, we ..replicate..
-100 times the following procedure: we sample with replacement all individuals and train the ``logit-simple'' and report the effects of each SNPs (for which a lot are 0).
-Over these 100 replicates, [VAL]\% of the SNPs never enter the predictions and [VAL] enter all predictions (Figure \ref{fig:supp-celiac-stabilitybar}).
-If we compute the sum of squared scores of the SNPs that enter every model divided by the sum of all squared scores [HAS A NAME?], we end up with a ratio of more than 99\% for all predictions, showing that most of the predictive performance is in these SNPs always entering the models.
-Moreover, these selected SNPs correspond to peaks of the GWAS, but also to less significant SNPs (Figure \ref{fig:supp-celiac-man-col}).
 
@@
 
-In this comparative study, we propose other methods as a replacement to the widely used P+T method. 
-We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really good performance, as reported in a previous study \cite[]{abraham2013performance}, higher than with the P+T method.
-We are well aware that the P+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions.
-Yet, in the case of prediction of disease to be used for clinical interventions, best possible methods should be used, even if it has to be used by large consortia only [EVEN LARGE CONSORTIA DON'T HAVE ALL DATA?].
+In this comparative study, we propose other methods as a replacement to the widely used C+T method. 
+We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really high predictive performance, as reported in a previous study \cite[]{abraham2013performance}, which is not the case with the C+T method.
+We are well aware that the C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions yet.
+However, in disease prediction for clinical interventions, best possible methods should be used, even if it has to be used by large consortia only [EVEN LARGE CONSORTIA DON'T HAVE ALL DATA?].
 
-One does not simply identify high-risk individuals (Figure \ref{fig:meme}).
+%One does not simply identify high-risk individuals (Figure \ref{fig:meme}).
 
-[DISCUSSION]As for the P+T method itself, we recall that we report the maximum AUCs of 102 thresholds, a threshold that should normally be learned on the training set only.
-The choice of this threshold is very important as it can greatly impact the
+As for the C+T method itself, we recall that we report the maximum AUCs of 102 thresholds, a threshold that should normally be learned on the training set only.
+The choice of this threshold is very important as it can greatly impact the predictive performance of the method.
+In the penalized logistic regression we present, we remove the choice of the regularization parameter with a procedure that we call Cross-Model Selection and Averaging (CMSA) while keeping the same predictive performance.
 
-[PARLER DU PROBLEME DE POP? + future paper]
+[PARLER DU PROBLEME DE POP? + future paper + possible car fast and memory efficient]
 
@@
 \begin{table}[htbp]
-\caption{Summary of all simulations. When there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.}
+\caption{Summary of all simulations. When there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.\label{tab:simus}}
 \vspace*{0.5em}
@@
 \end{tabular}
-\label{tab:simus}
 \end{table}
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
-\caption{Scenario \textnumero3.
-Mean of AUC over 100 simulations for our penalized logistic regression (``logit-simple'') and the maximum AUC reported with the P+T method (``PRS-max'').
+\caption{Scenario \textnumero1.
+Mean of AUC over 100 simulations for our penalized logistic regression (``logit-simple'') and the maximum AUC reported with the C+T method (``PRS-max'').
 Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-PRS}}
-\caption{Scenario \textnumero3.
+\caption{Scenario \textnumero1.
 Mean of AUC over 100 simulations for ..the three reported PRSs...
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
-\caption{Scenario \textnumero5.
+\caption{Scenario \textnumero3.
 Mean of AUC over 100 simulations for ..the three reported PRSs...
@@
 \centerline{\includegraphics[width=\textwidth]{main-celiac-regpath}}
-\caption{``Regularization paths'' for the P+T, ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.
-For the P+T method, the dotted line represents the maximum AUC while for the logistic regressions, the dotted lines represent the value using the CMSA procedure to be free of the choice of the threshold.}
+\caption{``Regularization paths'' for the C+T, ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.
+For the C+T method, the dotted line represents the maximum AUC while for the logistic regressions, the dotted lines represent the value using the CMSA procedure to be free of the choice of the threshold.}
 \label{fig:main-celiac-regpath}
@@
 
-\newpage
+\clearpage
 \section*{Acknowledgements}
@@
 For every combination of parameters, AUC of oracles are comprised between 83.2\% and 84.2\% for an heritability of 50\% and between 93.2\% and 94.1\% for an heritability of 80\%.
-Given all these estimates of the maximal achievable AUC and for the sake of simplicity, we report maximum AUCs of 84\% (respectively 94\%) for heritabilities of 50\% (respectively 80\%) whatever are the parameters of the simulation.
+Given all these estimates of the maximal achievable AUC and for the sake of simplicity, we report maximum AUCs of 84\% (94\%) for heritabilities of 50\% (80\%) whatever are the parameters of the simulation.
 
@@
 \end{tabular}
-\caption{Number of individuals by population and disease status for the celiac disease cohort (after quality control, genotyped on 281,122 ..mutual.. SNPs).}
-\label{tab:celiac-data}
+\caption{Number of individuals by population and disease status for the celiac disease cohort (after quality control, genotyped on 281,122 ..mutual.. SNPs).\label{tab:celiac-data}}
 \end{table}
@@
 \end{tabular}
-\caption{The 102 thresholds used in the P+T method for this study.}
-\label{tab:thr}
+\caption{The 102 thresholds used in the C+T method for this study.\label{tab:thr}}
 \end{table}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6}}
-\caption{Scenario \textnumero4. Boxplots of AUC over 100 simulations for ..the three reported PRSs... and the simple logistic regression.
+\caption{Scenario \textnumero2. Boxplots of AUC over 100 simulations for ..the three reported PRSs... and the simple logistic regression.
 Vertical panels represents different number of causal SNPs and their location (``all'': anywhere on the genome; ``HLA'': only in the HLA region).
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-corr}}
-\caption{Scenario \textnumero3. Comparison of different predictive measures. Correlation is greater than 98\%.[ADD]}
+\caption{Scenario \textnumero1. Comparison of different predictive measures. Correlation is greater than 98\%.[ADD]}
 \label{fig:supp-AUC-corr}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-celiac-roc}}
-\caption{ROC Curves for the P+T, ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.}
+\caption{ROC Curves for the C+T, ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.}
 \label{fig:supp-celiac-roc}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-celiac-stabilitybar}}
-\caption{Barplot showing how many times SNPs enter the predictions among predictions for 100 samples with replacement of the Celiac dataset. The bar for $0$ is not displayed for sake of visibility.}
-\label{fig:supp-celiac-stabilitybar}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-celiac-man-col}}
-\caption{Manhattan plot of the Celiac dataset. Red points are presenting SNPs that are always selected in 100 replicates of [...].}
-\label{fig:supp-celiac-man-col}
 \end{figure}

### 2018-02-27 3-simus.Rmd
simus finished

@@
 ```{r}
-paste(sprintf("%.1e", 10^(-c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100))))), collapse = ", ")
+10^(-c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))) %>%
+  sprintf("%.1e", .) %>%
+  paste(collapse = ", ")
+```
+
+```{r}
 PRS <- function(G, infos.chr, infos.pos, 

### 2018-02-27 4-results-simus.Rmd
simus finished

@@
 ```{r}
-boot <- function(x, n = 1e4, f = mean) {
+boot <- function(x, n = 1e5, f = mean) {
   sd(replicate(n, f(sample(x, replace = TRUE))))
@@
     ncol = 1, align = "hv", scale = 0.95, 
-    labels = LETTERS[2:3], label_size = 15
+    labels = LETTERS[2:3], label_size = 20
   )
@@
     ncol = 2, scale = 0.95, rel_widths = c(6, 4), 
-    labels = c("A", ""), label_size = 15
+    labels = c("A", ""), label_size = 20
   )
@@
 
-ggsave("figures/supp-ttrees.pdf", scale = 1/90, width = 1400, height = 1060)
+ggsave("figures/supp-ttrees.pdf", scale = 1/100, width = 1400, height = 1060)
 ```
@@
 
-ggsave("figures/supp-triple.pdf", scale = 1/90, width = 1400, height = 1060)
+ggsave("figures/supp-triple.pdf", scale = 1/100, width = 1400, height = 1060)
 ```
@@
 
-ggsave("figures/supp-AUC-triple.pdf", scale = 1/90, width = 1242, height = 951)
+ggsave("figures/supp-AUC-triple.pdf", scale = 1/110, width = 1242, height = 951)
 ```
@@
 
-ggsave("figures/supp-AUC-logit-fancy.pdf", scale = 1/90, width = 1242, height = 951)
+ggsave("figures/supp-AUC-logit-fancy.pdf", scale = 1/110, width = 1242, height = 951)
 ```
@@
 
-ggsave("figures/supp-AUC-PRS-fancy.pdf", scale = 1/90, width = 1242, height = 951)
+ggsave("figures/supp-AUC-PRS-fancy.pdf", scale = 1/110, width = 1242, height = 951)
 ```
@@
 
-### Correlation between predictive measures [MOVE TOP]
+### Correlation between predictive measures
 
@@
 
-```{r, fig.asp=0.5, out.width="95%", fig.width=16}
+```{r, fig.asp=0.7, out.width="95%", fig.width=16}
 results3 %>%
@@
   labs(y = "Partial AUC (specificity: [0.9, 1])") +
-  theme(legend.position = "top")
+  theme(legend.position = "top", axis.title = element_text(size = rel(2.2))) +
 
@@
 
-ggsave("figures/supp-AUC-chr6.pdf", scale = 1/90, width = 1153, height = 908)
+ggsave("figures/supp-AUC-chr6.pdf", scale = 1/110, width = 1153, height = 908)
 ```
@@
 
-ggsave("figures/main-AUC-ntrain.pdf", scale = 1/90, width = 1050, height = 800)
+ggsave("figures/main-AUC-ntrain.pdf", scale = 1/100, width = 1050, height = 800)
 ```

### 2018-02-27 5-results-celiac.Rmd
need to finish real data

@@
 knitr::opts_chunk$set(fig.align = "center", out.width = "70%", fig.width = 10,
-                      fig.asp = 0.7, dev = "svg")
+                      fig.asp = 0.7, dev = "svg", cache = TRUE)
 options(width = 95)
@@
 # GWAS
+system.time(
   gwas.train <- big_univLogReg(
@@
   )
+)
+```
+
+```{r}
+# Q-Q plot
 snp_qq(gwas.train) +
-  coord_cartesian(ylim = c(0, 25))
+  coord_cartesian(ylim = c(0, 20))
 gwas.train.gc <- snp_gc(gwas.train)
@@
 ```{r, dev="png"}
+# Manhattan plot
 labels <- 1:22; labels[c(11, 13, 15, 17, 18, 20, 21)] <- ""
@@
   snp_manhattan(gwas.train.gc, CHR, POS, labels = labels) +
-    coord_cartesian(ylim = c(0, 25)) + 
+    coord_cartesian(ylim = c(0, 22)) + 
     geom_hline(yintercept = -log10(5e-8), color = "red", linetype = 3),
@@
 aucs <- apply(prs, 2, AUC, target = y[ind.test])
+summary(aucs)
+```
+
+```{r}
+pAUC <- function(pred, target, p = 0.1) {
+  val.min <- min(target)
+  q <- quantile(pred[target == val.min], probs = 1 - p)
+  ind <- (target != val.min) | (pred > q)
+  tryCatch(bigstatsr::AUC(pred[ind], target[ind]) * p, 
+           error = function(e) p^2 / 2)  # all same prediction
+}
+plot_density_AUC <- function(pred, true = y[ind.test]) {
+  data.frame(pred = pred, pheno = as.factor(true)) %>%
+  myggplot() +
+  geom_density(aes(pred, fill = pheno), alpha = 0.4) +
+  ggtitle(sprintf("AUC = %s%% | pAUC = %s", 
+                  round(100 * AUC(pred, true), 2),
+                  round(pAUC(pred, true), 4)))
+}
+plot_density_AUC(prs[, which.max(aucs)])
 ```
@@
 
+Cross-Model Selection and Averaging (CMSA):
+
 ```{r}
+# Train logit-simple
+system.time(
   logit <- big_spLogReg(X = G, y01.train = y[ind.train], 
@@
                         covar.train = obj.svd$u[ind.train, ],
-                      alpha = 0.5, dfmax = 20e3)
+                        ncores = NCORES)
+)
+# Get K = 10 predictions/individuals
 preds1 <- predict(logit, X = G, ind.row = ind.test, 
                   covar.row = obj.svd$u[ind.test, ])
-library(Matrix)
-nb.pred1 <- colSums(logit$beta != 0)
-aucs1 <- apply(preds1, 2, AUC, target = y[ind.test])
+(nb.pred1 <- sum(rowSums(sapply(logit, function(l) l$beta.X != 0)) > 0))
+(aucs1 <- apply(preds1, 2, AUC, target = y[ind.test]))
+# Combining all K predictions
+preds1.comb <- rowMeans(preds1)
+AUC(preds1.comb, y[ind.test])
 ```
 
-Cross-Model Selection and Averaging (CMSA):
-
 ```{r}
-system.time({
-  cmsa.logit <- big_CMSA(FUN = big_spLogReg, feval = AUC,
-                         X = G, y.train = y[ind.train], 
-                         ind.train = ind.train, 
-                         covar.train = obj.svd$u[ind.train, ],
-                         alpha = 0.5, dfmax = 20e3, 
-                         ncores = NCORES)
-
-  preds2 <- predict(cmsa.logit, X = G, ind.row = ind.test, 
-                    covar.row = obj.svd$u[ind.test, ])
-})
-(auc.cmsa <- AUC(preds2, y[ind.test]))
+plot_density_AUC(preds1.comb)
 ```
@@
 G2 <- big_attach("backingfiles/celiacQC_tripled1.rds")
+system.time(
   logit2 <- big_spLogReg(X = G2, y01.train = y[ind.train], 
@@
                          covar.train = obj.svd$u[ind.train, ],
-                       alpha = 0.5, dfmax = 20e3)
-preds3 <- predict(logit2, X = G2, ind.row = ind.test, 
+                         ncores = NCORES)
+)
+
+preds2 <- predict(logit2, X = G2, ind.row = ind.test, 
                   covar.row = obj.svd$u[ind.test, ])
-nb.pred2 <- colSums(logit2$beta != 0)
-aucs2 <- apply(preds3, 2, AUC, target = y[ind.test])
+(nb.pred2 <- sum(rowSums(sapply(logit2, function(l) l$beta.X != 0)) > 0))
+(aucs2 <- apply(preds2, 2, AUC, target = y[ind.test]))
+# Combining all K predictions
+preds2.comb <- rowMeans(preds2)
+AUC(preds2.comb, y[ind.test])
 ```
 
-Cross-Model Selection and Averaging (CMSA):
-
 ```{r}
-system.time({
-  cmsa.logit2 <- big_CMSA(FUN = big_spLogReg, feval = AUC,
-                          X = G2, y.train = y[ind.train], 
-                          ind.train = ind.train, 
-                          covar.train = obj.svd$u[ind.train, ],
-                          alpha = 0.5, dfmax = 20e3, 
-                          ncores = NCORES)
-
-  preds4 <- predict(cmsa.logit2, X = G2, ind.row = ind.test, 
-                    covar.row = obj.svd$u[ind.test, ])
-})
-(auc.cmsa2 <- AUC(preds4, y[ind.test]))
+plot_density_AUC(preds2.comb)
 ```
@@
 
-First, we get the SNP with the highest ratio of cases over controls for homozygous-recessive individuals.
+First, we get the most significant SNP.
 
 ```{r}
-props <- sapply(cols_along(G), function(j) {
-  ccoj <- counts.control[, j]
-  mut <- `if`(ccoj[1] < ccoj[3], 1, 3)
-  `if`(counts.case[mut, j] > 100, counts.case[mut, j] / ccoj[mut], 0)
-})
-theone <- which.max(props)
+theone <- which.max(abs(gwas.train.gc$score))
+ind0 <- which(G[ind.test, theone] == 0)
+(M <- length(ind0))
 table(Geno = G[, theone], forcats::fct_recode(as.factor(y), Control = "0", Case = "1"))
@@
 # one geno
-ind0 <- which(G[ind.test, theone] == 0)
-(M <- length(ind0))
 mean(y[ind.test[ind0]]) 
@@
 
-  ind.train <- sort(sample(nrow(G)))
+  ind.train <- sort(sample(nrow(G), replace = TRUE))
 
@@
 
-  saveRDS(cmsa.logit[cols_along(G)], file = res.file)
+  saveRDS(list(betas = cmsa.logit[cols_along(G)],
+               ind.test = setdiff(rows_along(G), ind.train)),
+          file = res.file)
 }

### 2018-02-27 paper/draft.tex
need to finish real data

@@
 \begin{table}[htbp]
-\caption{Summary of all simulations. When there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.\label{tab:simus}}
+\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.\label{tab:simus}}
 \vspace*{0.5em}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-celiac-roc}}
-\caption{ROC Curves for the C+T, ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.}
+\caption{ROC Curves for the ``C+T'', ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.}
 \label{fig:supp-celiac-roc}

### 2018-03-01 5-results-celiac.Rmd
add celiac results

@@
 author: "Florian Privé"
-date: "November 3, 2017"
+date: "March 1, 2018"
 output: html_document
@@
 
-```{r}
+```{r, cache=FALSE}
 # Read file, see "Preprocessing" notebook
@@
 dim(G)
+set.seed(1)
 ind.train <- sort(sample(nrow(G), size = 12e3))
@@
 
-```{r}
+```{r, dev='png', fig.asp=0.8}
 # Q-Q plot
 snp_qq(gwas.train) +
-  coord_cartesian(ylim = c(0, 20))
+  coord_cartesian(ylim = c(0, 15))
 gwas.train.gc <- snp_gc(gwas.train)
@@
 
-```{r, dev="png"}
+```{r, dev="png", fig.asp=0.9, out.width="90%"}
 # Manhattan plot
@@
 summary(aucs)
+nb.pred[which.max(aucs)]
 ```
@@
                   round(100 * AUC(pred, true), 2),
-                  round(pAUC(pred, true), 4)))
+                  round(pAUC(pred, true), 4))) + 
+  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
 }
-plot_density_AUC(prs[, which.max(aucs)])
+(p1 <- plot_density_AUC(prs[, which.max(aucs)]))
 ```
@@
 ```{r}
-plot_density_AUC(preds1.comb)
+(p2 <- plot_density_AUC(preds1.comb))
 ```
@@
 ```{r}
-plot_density_AUC(preds2.comb)
+(p3 <- plot_density_AUC(preds2.comb))
 ```
 
-### Regularization paths
-
-```{r}
-h <- round(c(max(aucs), auc.cmsa, auc.cmsa2), 4)
-p <- myggplot(rbind(
-  cbind.data.frame(nb = nb.pred,      AUC = aucs,      Method = "PRS"),
-  cbind.data.frame(nb = nb.pred1[-1], AUC = aucs1[-1], Method = "logit-simple"),
-  cbind.data.frame(nb = nb.pred2[-1], AUC = aucs2[-1], Method = "logit-triple")
-)) +
-  geom_point(aes(nb, AUC, color = Method)) +
-  scale_x_log10(breaks = 10^(0:6)) +
-  labs(x = "Number of predictors") +
-  geom_hline(yintercept = h, color = scales::hue_pal()(3), linetype = 2) +
-  theme(legend.position = c(0.85, 0.5))
-p + 
-  scale_y_continuous(breaks = sort(c(ggplot_build(p)$layout$panel_ranges[[1]]$y.major_source, h)))
-```
-
-```{r}
-ggsave("figures/main-celiac-regpath.pdf", scale = 1/90, width = 1000, height = 700)
+```{r, out.width="70%", fig.asp=900/780}
+list(p1, p2, p3) %>%
+  lapply(function(p) p + theme(legend.position = "none")) %>%
+  cowplot::plot_grid(plotlist = ., ncol = 1, labels = LETTERS[1:3], label_size = 20, scale = 0.95) %>%
+  cowplot::plot_grid(., cowplot::get_legend(p1), rel_widths = c(1, 0.15), scale = 0.95)
+ggsave("figures/supp-score-densities.pdf", scale = 1/100, width = 780, height = 900)
 ```
@@
   "PRS-max"      = prs[, which.max(aucs)],
-  "logit-simple" = preds2,
-  "logit-triple" = preds4
+  "logit-simple" = preds1.comb,
+  "logit-triple" = preds2.comb
 ) %>%
@@
 ```{r}
-ggsave("figures/supp-celiac-roc.pdf", scale = 1/90, width = 700, height = 700)
+ggsave("figures/celiac-roc.pdf", scale = 1/90, width = 700, height = 700)
 ```
@@
 ```{r}
-theone <- which.max(abs(gwas.train.gc$score))
+# Most significant SNP
+theone <- which.max(-predict(gwas.train.gc))
 ind0 <- which(G[ind.test, theone] == 0)
@@
 # logit-simple (CMSA)
-indLS <- order(preds2, decreasing = TRUE)[1:M]
+indLS <- order(preds1.comb, decreasing = TRUE)[1:M]
 mean(y[ind.test[indLS]])
 # logit-triple (CMSA)
-indLT <- order(preds4, decreasing = TRUE)[1:M]
+indLT <- order(preds2.comb, decreasing = TRUE)[1:M]
 mean(y[ind.test[indLT]])
 ```
-
-## Stability of selection of the "logit-simple"
-
-Train on 100 bootstrap samples of the Celiac dataset.
-
-```{r, eval=FALSE}
-if (!dir.exists("results6")) dir.create("results6")
-
-for (i in seq_len(100)) {
-
-  res.file <- paste0("results6/simu_", i, ".rds")
-  if (file.exists(res.file)) next
-
-  ind.train <- sort(sample(nrow(G), replace = TRUE))
-
-  cmsa.logit <- big_CMSA(FUN = big_spLogReg, feval = AUC,
-                         X = G, y.train = y[ind.train], 
-                         ind.train = ind.train, 
-                         covar.train = obj.svd$u[ind.train, ],
-                         alpha = 0.5, dfmax = 20e3, 
-                         ncores = NCORES)
-
-  saveRDS(list(betas = cmsa.logit[cols_along(G)],
-               ind.test = setdiff(rows_along(G), ind.train)),
-          file = res.file)
-}
-```
-
-```{r}
-results6 <- list.files("results6", full.names = TRUE) %>%
-  map(~readRDS(.x)) %>%
-  do.call("cbind", .)
-```
-
-```{r}
-n_enter <- rowSums(results6 != 0)
-mean(n_enter == 0)                ## Proportion of SNPs never entering
-n_enter_once <- n_enter[n_enter != 0]
-mean(n_enter_once)                ## Mean number of times a SNPs enter 
-                                  ## if entering at least once
-bigstatsr:::MY_THEME(
-  qplot(n_enter_once, geom = "bar") +
-    labs(x = "Number of times")
-)
-```
-
-```{r}
-ggsave("figures/supp-celiac-stabilitybar.pdf", scale = 1/90, width = 1000, height = 700)
-```
-
-```{r}
-length(always_enter <- which(n_enter == ncol(results6)))
-cowplot::plot_grid(
-  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels, 
-                ind.highlight = always_enter),
-  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels, 
-                ind.highlight = always_enter) +
-    coord_cartesian(ylim = c(0, 25)) + 
-    geom_hline(yintercept = -log10(5e-8), color = "red", linetype = 3),
-  align = "hv", ncol = 1, labels = LETTERS[1:2], label_size = 25, scale = 0.95
-)
-```
-
-```{r}
-ggsave("figures/supp-celiac-man-col.png", scale = 1/90, width = 1300, height = 1000)
-```
-
-```{r}
-sq_effects_always <- colSums((results6[always_enter, ])^2)
-sq_effects_others <- colSums((results6[-always_enter, ])^2)
-summary(100 * sq_effects_always / (sq_effects_always + sq_effects_others))
-```

### 2018-03-01 paper/draft.tex
add celiac results

@@
 
-\title{Predicting complex diseases: performance, robustness and ..utility..(impact?)}
-\author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
+\title{Predicting complex diseases: performance and robustness [OF NEW METHOD?]}
+\author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
 
@@
 \noindent$^{\text{\sf 1}}$Universit\'e Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
-\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France \\
-\noindent$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.
+\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France.
 
@@
 In this paper, we present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in \cite{botta2014exploiting}.
-Note that in this comparison we don't include any LMM method for the reasons mentioned before and don't include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
+Note that in this comparison we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
 For the C+T model, we compare different thresholds of inclusion of SNPs.
@@
 Finally, the $K$ resulting vectors of coefficients are reported and can be combined into one vector (e.g.\ using the geometric median) or resulting scores can be combined (e.g.\ averaging the $K$ probabilities for each individual).
-We also use this procedure to derive an early stopping criterion so that the algorithm doesn't need to evaluate the whole regularization path, making the whole procedure much faster.
+We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization path, making the whole procedure much faster.
 Because of L1-regularization, this vector of coefficients is typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts.
@@
 
-\subsection{Overview and naming}
+\subsection{Maximum AUCs}
 
 In results, we use estimates of maximum achievable AUC values for the simulations: 84\% (respectively 94\%) for heritabilities of 50\% (respectively 80\%) based on three estimations (see Supplementary Materials).
+Note that we do not report the simulation results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).
 
@@
 
-\subsection{[Other measures of performance]}
-
-Our method is parallelized and requires only approximately 500 seconds to run (with 6 physical cores) for a training size of 6000 individuals.
-It takes 1000 seconds to train on 12,000 individuals.
-Moreover, when there are 3 times more variables, it takes approximately 3 more times to run.
-We can confidently say that the ``logit-simple'' is of linear complexity in both dimensions.
-Moreover, the ``logit-simple'' results in relatively sparse models with no more than 14,000 predictors entering the models even for the linear model of 3000 causal SNPs (Figure \ref{fig:supp-triple}).
-The ``logit-triple'' is comparable in terms of number of predictors.
-Finally, for other predictive performance measures, the percentage of cases in either the 10\% or 20\% largest PRSs, we report that their values are correlated to the values of AUCs at more than 98\%, which is why we don't speak about them more in the results (Figure \ref{fig:supp-AUC-corr}).
-
 \subsection{Real data results}
 
-[MONTRER UN MANHATTAN PLOT POUR JUSTIFIER 30 IN HLA]
+We also analyze the whole Celiac dataset with the ..original/real..
+phenotypes.
+For one run [ENOUGH FOR REVIEWERS?], splitting the dataset in 12,000 indiduals in the training set and the remaining 3155 individuals in the test set, we compare the predictive performance of the ``C+T'', ``logit-simple'' and ``logit-triple'' methods, but also computation times and number of predictors used in the models (Table \ref{tab:results-celiac}, figure \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
+We ..see..
+that the joint logistic regressions provide higher AUC with 88.71\% with the ``logit-simple'' and 89.26 with the ``logit-triple as compared to 82.46 with the ``C+T'' method, and much higher partial AUC (specificity 0.9-1), respectively 0.0411, 0.0440 and 0.0283, while using less predictors, respectively 2198, 4617 and 13277.
+Note that for the ``C+T'' method, we still report the best result among 102 p-values thresholds.
+In terms of computation time, we report only the GWAS computation for the ``C+T'' method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once, is faster than the fast ``C+T'' method, and the ``logit-triple'' takes twice as long as the ``logit-simple''.
+Finally, note that the scores you get with the logistic regressions are probabilities, which is not the case with the ``C+T'' method (Figure \ref{fig:supp-score-densities}).[QUE RAPPORTER SUR LES DENSITY PLOTS DU SUPMAT?]
 
-We also analyze the whole Celiac dataset with the original real phenotypes.
-For one run, splitting the dataset in 12,000 indiduals in the training set and the remaining 3155 individuals in the test set, we compare the ``regularization paths'' of all methods (T-Trees excepted), that is to say the predictive performance as a function of the number of predictors corresponding to different inclusion thresholds of the methods.
-We show that the regularization paths of both logistic regressions are both very smooth, reaching a maximum AUC or more than 89\% at approximately 1000 predictors (Figure \ref{fig:main-celiac-regpath}).
+\begin{table}[h]
+\caption{Results for the real Celiac dataset.\label{tab:results-celiac}}
+\vspace*{0.5em}
+\centering
+\begin{tabular}{|l|c|c|c|c|}
+\hline
+Method & AUC (\%) & pAUC & \# predictors & Execution time (s) \\ \hline
+PRS-max & 82.46 & 0.0283 & 13277 & 149 \\ 
+logit-simple & 88.71 & 0.0411 & 2198 & 78 \\
+logit-triple & 89.26 & 0.0440 & 4617 & 161 \\ \hline
+\end{tabular}
+\end{table}
+
+%We show that the regularization paths of both logistic regressions are both very smooth, reaching a maximum AUC or more than 89\% at approximately 1000 predictors (Figure \ref{fig:main-celiac-regpath}).
 Moreover, the CMSA procedure give AUC very close to the maximum AUC of the regularization path.
 For the C+T method, on the contrary, the regularization path is not smooth at all and the AUC never exceed the AUC given by the most stringent threshold [ADD VALUE OF AUC] ($p < 10^{-100}$).
-We also show some associated ROC curves [QUOI DIRE DE CES ROC?] (Figure \ref{fig:supp-celiac-roc}).
+We also show some associated ROC curves [QUOI DIRE DE CES ROC?] (Figure \ref{fig:celiac-roc}).
 
@@
 In this comparative study, we propose other methods as a replacement to the widely used C+T method. 
-We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really high predictive performance, as reported in a previous study \cite[]{abraham2013performance}, which is not the case with the C+T method.
+We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really high predictive performance, as reported in a previous study \cite[]{abraham2013performance}, which is not the case with the C+T method that uses SNP effects learned independently.
 We are well aware that the C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions yet.
 However, in disease prediction for clinical interventions, best possible methods should be used, even if it has to be used by large consortia only [EVEN LARGE CONSORTIA DON'T HAVE ALL DATA?].
+The logistic regression we present here could be used on data as large as biobank datasets because it is very efficient in both memory and computation time (..even faster than doing a GWAS!..).
 
@@
 In the penalized logistic regression we present, we remove the choice of the regularization parameter with a procedure that we call Cross-Model Selection and Averaging (CMSA) while keeping the same predictive performance.
+
+We also introduce a simple feature engineering technique to detect and learn dominant and recessive effects..
 
@@
 \vspace*{0.5em}
+\centering
 \begin{tabular}{|c|c|c|c|c|c|c|c|}
@@
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{main-celiac-regpath}}
-\caption{``Regularization paths'' for the C+T, ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.
-For the C+T method, the dotted line represents the maximum AUC while for the logistic regressions, the dotted lines represent the value using the CMSA procedure to be free of the choice of the threshold.}
-\label{fig:main-celiac-regpath}
+\centerline{\includegraphics[width=\textwidth]{celiac-roc}}
+\caption{ROC Curves for the ``C+T'', ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.
+Plotted using R package plotROC \cite[]{sachs2017plotroc}.\label{fig:celiac-roc}}
 \end{figure}
@@
 Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).
-We are also grateful to F\'elix Balazard for useful discussions about T-Trees.
+We are also grateful to F\'elix Balazard for useful discussions about T-Trees, and to Yaohui Zeng for useful discussions about R package biglasso.
 
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-corr}}
-\caption{Scenario \textnumero1. Comparison of different predictive measures. Correlation is greater than 98\%.[ADD]}
-\label{fig:supp-AUC-corr}
+\caption{Scenario \textnumero1. Comparison of different predictive measures. ..There is a Spearman correlation.. of 98\% between values of AUC and partial AUC.[ADD]\label{fig:supp-AUC-corr}}
 \end{figure}
@@
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-celiac-roc}}
-\caption{ROC Curves for the ``C+T'', ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.}
-\label{fig:supp-celiac-roc}
+\centerline{\includegraphics[width=\textwidth]{supp-score-densities}}
+\caption{[ADD]\label{fig:supp-score-densities}}
 \end{figure}

### 2018-03-01 stability-CMSA.Rmd
add celiac results

## Stability of selection of the "logit-simple"

Train on 100 bootstrap samples of the Celiac dataset.

```{r, eval=FALSE}
if (!dir.exists("results6")) dir.create("results6")

for (i in seq_len(100)) {

  res.file <- paste0("results6/simu_", i, ".rds")
  if (file.exists(res.file)) next

  ind.train <- sort(sample(nrow(G), replace = TRUE))

  cmsa.logit <- big_CMSA(FUN = big_spLogReg, feval = AUC,
                         X = G, y.train = y[ind.train], 
                         ind.train = ind.train, 
                         covar.train = obj.svd$u[ind.train, ],
                         alpha = 0.5, dfmax = 20e3, 
                         ncores = NCORES)

  saveRDS(list(betas = cmsa.logit[cols_along(G)],
               ind.test = setdiff(rows_along(G), ind.train)),
          file = res.file)
}
```

```{r}
results6 <- list.files("results6", full.names = TRUE) %>%
  map(~readRDS(.x)) %>%
  do.call("cbind", .)
```

```{r}
n_enter <- rowSums(results6 != 0)
mean(n_enter == 0)                ## Proportion of SNPs never entering
n_enter_once <- n_enter[n_enter != 0]
mean(n_enter_once)                ## Mean number of times a SNPs enter 
                                  ## if entering at least once
bigstatsr:::MY_THEME(
  qplot(n_enter_once, geom = "bar") +
    labs(x = "Number of times")
)
```

```{r}
ggsave("figures/supp-celiac-stabilitybar.pdf", scale = 1/90, width = 1000, height = 700)
```

```{r}
length(always_enter <- which(n_enter == ncol(results6)))
cowplot::plot_grid(
  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels, 
                ind.highlight = always_enter),
  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels, 
                ind.highlight = always_enter) +
    coord_cartesian(ylim = c(0, 25)) + 
    geom_hline(yintercept = -log10(5e-8), color = "red", linetype = 3),
  align = "hv", ncol = 1, labels = LETTERS[1:2], label_size = 25, scale = 0.95
)
```

```{r}
ggsave("figures/supp-celiac-man-col.png", scale = 1/90, width = 1300, height = 1000)
```

```{r}
sq_effects_always <- colSums((results6[always_enter, ])^2)
sq_effects_others <- colSums((results6[-always_enter, ])^2)
summary(100 * sq_effects_always / (sq_effects_always + sq_effects_others))
```

### 2018-03-01 test-farmcpu.R

@@ -1,0 +1,44 @@
+# devtools::install_github("amkusmec/FarmCPUpp")
+library(FarmCPUpp)
+library(bigmemory)
+library(bigsnpr)
+
+celiac <- snp_attach("backingfiles/celiacQC.rds")
+G <- celiac$genotypes
+# ind.chr6 <- which(celiac$map$chromosome == 6)
+# 
+# G2 <- filebacked.big.matrix(nrow(G), ncol(G),
+#                             backingfile = "test_farmcpu")
+# 
+# big_apply(G, a.FUN = function(X, ind) {
+#   G2[, ind] <- X[, ind]
+#   NULL
+# }, a.combine = 'c')
+# G2[, ncol(G2)]
+
+G2 <- attach.big.matrix("test_farmcpu.desc")
+
+system.time(
+  myResults2 <- farmcpu(
+    Y = celiac$fam[c(1, 6)], 
+    GD = G2, 
+    GM = celiac$map[c(2, 1, 4)],
+    CV = readRDS("backingfiles/PCA.rds")$u,
+    ncores.glm = nb_cores(),
+    ncores.reml = nb_cores()
+  )
+) 
+# 69 / 42 for chr 6
+# 1880 / 414 for all chrs
+# With PCs: 2255 / 479
+
+plot(-log10(myResults$affection$GWAS$p.value),
+     -log10(myResults2$affection$GWAS$p.value),
+     pch = 20, cex = 0.5)
+
+str(myResults2)
+plot(-log10(myResults2$affection$GWAS$p.value), pch = 20, cex = 0.5, ylim = c(0, 30))
+plot(-log10(ppoints(length(myResults2$affection$GWAS$p.value))),
+     sort(-log10(myResults2$affection$GWAS$p.value), decreasing = TRUE),
+     pch = 20, cex = 0.5, ylim = c(0, 30))
+abline(0, 1, col = "red")

### 2018-03-02 test-new-auc.R
auc

@@ -1,0 +1,39 @@
+# plot(c(0, 1), c(0, 1))
+# bigstatsr::pasteLoc(8)
+
+pts <- list(x = c(0, 0.007, 0.027, 0.034, 0.101, 0.196, 0.351, 0.601, 0.839, 1), 
+            y = c(0, 0.134, 0.347, 0.545, 0.701, 0.789, 0.841, 0.895, 0.942, 1))
+plot(pts, type = "b")
+pts2 <- list(x = c(0, 0.02, 0.051, 0.073, 0.113, 0.152, 0.222, 0.404, 0.674, 1), 
+             y = c(0, 0.097, 0.238, 0.369, 0.539, 0.698, 0.823, 0.916, 0.962, 1))
+points(pts2, col = "red", type = "b")
+
+f1 <- function(pts) {
+  w <- diff(pts$x)
+  h <- pts$y[-length(pts$y)] + diff(pts$y) / 2 
+  drop(crossprod(h, w))
+}
+f1(pts)
+f1(pts2)
+
+
+f2 <- function(pts) {
+  x <- pts$x
+  y <- pts$y
+  ind <- seq_len(length(x) - 1)
+  parts <- sapply(ind, function(i) {
+    d <- 1 - x[[i]]
+    c <- 1 - x[[i + 1]]
+    a <- y[[i]]
+    b <- y[[i + 1]]
+    (b - a) * (d^2 + c * d + c^2) / 3 + (a * d - b * c) * (d + c) / 2
+  })
+  sum(parts) * 2
+}
+
+f2(pts)
+f2(pts2)
+
+plot(1 - pts$x, pts$y)
+curve(x + 0)
+curve(1 / exp(10*(1-x)))

### 2018-03-02 test-pauc.R

@@ -1,0 +1,29 @@
+N <- 1000
+x1 <- rnorm(N)
+x2 <- rnorm(N, mean = 1)
+x <- c(x1, x2)
+y <- rep(0:1, each = N)
+plot(print(pROC::roc(y ~ x)))
+bigstatsr::AUC(x, y)
+plot(print(pROC::roc(y[ind] ~ x[ind])))
+bigstatsr::AUC(x[x > q], y[x > q])
+plot(print(pROC::roc(y[x > q] ~ x[x > q])))
+
+
+test <- rocdemo.sca(y, x)
+ROC::AUC(test)
+plot(test)
+sapply(0:10 / 10, function(p) ROC::pAUC(test, t0 = p))
+plot(test)
+
+# Partial AUC based on AUC
+pAUC <- function(pred, target, p = 0.1) {
+  q <- quantile(pred[target == 0], probs = 1 - p)
+  ind <- (target > 0.5) | (pred > q)
+  bigstatsr::AUC(pred[ind], target[ind]) * p
+}
+sapply(1:9 / 10, function(p) pAUC(x, y, p))
+
+plot(pROC::roc(y ~ x))
+abline(v = 0.9, col = "red")
+abline(h = mean(x2 > quantile(x1, 0.9)), col = "blue")

### 2018-03-02 5-results-celiac.Rmd
need to add multiple results

@@
 
-## Predictive models
+## Methods' functions
+
+Each method's function returns a tibble (data frame) with 4 columns:
+
+1. the name of the method,
+2. the predictive scores and true phenotypes for the test set, as a [list-column](http://r4ds.had.co.nz/many-models.html),
+3. the timing of the main computations (in seconds),
+4. the number of SNPs used for the prediction.
+
+### PRS
+
+```{r}
+PRS <- function(G, infos.chr, infos.pos, 
+                pheno.all, covar.all, 
+                ind.train, ind.test) {
+
+  timing <- system.time({
+
+    # GWAS
+    gwas.train <- big_univLogReg(
+      G, pheno.all[ind.train], ind.train = ind.train, 
+      covar.train = covar.all[ind.train, , drop = FALSE], 
+      ncores = nb_cores()
+    )
+    gwas.train.gc <- snp_gc(gwas.train)
+
+    # Clumping on the test set
+    ind.keep <- snp_clumping(G, infos.chr = infos.chr,
+                             ind.row = ind.test,
+                             thr.r2 = 0.2, 
+                             S = abs(gwas.train.gc$score),
+                             size = 500,
+                             is.size.in.bp = TRUE,
+                             infos.pos = infos.pos,
+                             ncores = nb_cores())
+
+    # PRS
+    thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
+    lpS <- -predict(gwas.train.gc)
+    nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
+    prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
+                   ind.test = ind.test,
+                   ind.keep = ind.keep,
+                   lpS.keep = lpS[ind.keep], 
+                   thr.list = thrs)
+  })[3]
+
+  ind.best <- which.max(apply(prs, 2, AUC, target = pheno.all[ind.test]))
+
+  tibble(
+    method = "PRS-max",
+    eval = list(cbind(prs[, ind.best], pheno.all[ind.test])),
+    timing = timing,
+    nb.preds = nb.pred[ind.best]
+  )
+}
+```
+
+### Logistic regression
+
+```{r}
+logit.CMSA <- function(G, pheno.all, covar.all, ind.train, ind.test, method, print_all = FALSE) {
+
+  timing <- system.time({
+
+    cmsa.logit <- big_spLogReg(X = G, y01.train = pheno.all[ind.train], 
+                               ind.train = ind.train, 
+                               covar.train = covar.all[ind.train, , drop = FALSE],
+                               ncores = nb_cores())
+
+    preds.all <- predict(cmsa.logit, X = G, ind.row = ind.test, 
+                         covar.row = covar.all[ind.test, , drop = FALSE])
+    if (print_all) print(preds.all)
+    preds <- rowMeans(preds.all)
+  })[3]
+
+  tibble(
+    method   = method, 
+    eval     = list(cbind(preds, pheno.all[ind.test])),
+    timing   = timing,
+    nb.preds = sum(rowSums(sapply(cmsa.logit, function(x) x$beta.X) != 0) > 0)
+  )
+}
+```
+
+## Predictive models and plots
 
@@
 ```{r}
+res.PRS <- PRS(
+  G = G,
+  infos.chr = CHR,
+  infos.pos = POS,
+  pheno.all = y,
+  covar.all = obj.svd$u,
+  ind.train = ind.train,
+  ind.test = ind.test
+)
 # GWAS
@@
 
+## Predictive methods (multiple times)
+
+```{r}
+# Divide in training/test sets
+ind.train <- sort(sample(nrow(G), size = 12e3))
+ind.test <- setdiff(rows_along(G), ind.train)
+```
+
+### PRS
+
+```{r}
+# PRS
+PRS <- function() {
+
+  system.time(
+  gwas.train <- big_univLogReg(
+    G, y[ind.train], ind.train = ind.train, 
+    covar.train = obj.svd$u[ind.train, , drop = FALSE], 
+    ncores = NCORES
+  )
+)
+}
+
+```
+
+### Logistic regression
+
+Cross-Model Selection and Averaging (CMSA):
+
+```{r}
+# Train logit-simple
+system.time(
+  logit <- big_spLogReg(X = G, y01.train = y[ind.train], 
+                        ind.train = ind.train, 
+                        covar.train = obj.svd$u[ind.train, ],
+                        ncores = NCORES)
+)
+# Get K = 10 predictions/individuals
+preds1 <- predict(logit, X = G, ind.row = ind.test, 
+                  covar.row = obj.svd$u[ind.test, ])
+(nb.pred1 <- sum(rowSums(sapply(logit, function(l) l$beta.X != 0)) > 0))
+(aucs1 <- apply(preds1, 2, AUC, target = y[ind.test]))
+# Combining all K predictions
+preds1.comb <- rowMeans(preds1)
+AUC(preds1.comb, y[ind.test])
+```
+
+```{r}
+(p2 <- plot_density_AUC(preds1.comb))
+```
+
+### Logistic regression with feature engineering
+
+```{r}
+G2 <- big_attach("backingfiles/celiacQC_tripled1.rds")
+system.time(
+  logit2 <- big_spLogReg(X = G2, y01.train = y[ind.train], 
+                         ind.train = ind.train, 
+                         covar.train = obj.svd$u[ind.train, ],
+                         ncores = NCORES)
+)
+
+preds2 <- predict(logit2, X = G2, ind.row = ind.test, 
+                  covar.row = obj.svd$u[ind.test, ])
+(nb.pred2 <- sum(rowSums(sapply(logit2, function(l) l$beta.X != 0)) > 0))
+(aucs2 <- apply(preds2, 2, AUC, target = y[ind.test]))
+# Combining all K predictions
+preds2.comb <- rowMeans(preds2)
+AUC(preds2.comb, y[ind.test])
+```
+
+```{r}
+(p3 <- plot_density_AUC(preds2.comb))
+```
+
+```{r, out.width="70%", fig.asp=900/780}
+list(p1, p2, p3) %>%
+  lapply(function(p) p + theme(legend.position = "none")) %>%
+  cowplot::plot_grid(plotlist = ., ncol = 1, labels = LETTERS[1:3], label_size = 20, scale = 0.95) %>%
+  cowplot::plot_grid(., cowplot::get_legend(p1), rel_widths = c(1, 0.15), scale = 0.95)
+ggsave("figures/supp-score-densities.pdf", scale = 1/100, width = 780, height = 900)
+```
+
 ## Top scores

### 2018-03-02 paper/draft.tex
abstract + discussions + typos

@@
 \newpage
+[ABSTRACT TROP LONG?]
 \abstract{
-\textbf{Motivation:} \\
-\textbf{Results:} \\
-\textbf{Availability:} \\
+Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
+PRSs can have a major public health impact, allowing for screening campaigns to identify high-genetic risk individuals for a given disease.
+Yet, current PRSs are far from reaching this goal, and would have to show a high discriminative power between cases and controls in order to be of any use.
+
+The predictive power of PRS strongly relies on the underlying technique used to include SNPs in the PRS and to estimate their relative contribution.
+The most commonly used method to derive PRSs is called ``Clumping+Thresholding'' (C+T).
+It uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it very suitable and also very fast.
+We recently developed two R packages, bigstatsr and bigsnpr, for the analysis of large-scale genome-wide data.
+Package bigstatsr includes efficient functions for computing penalized linear and logistic regressions on large datasets.
+
+We present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in a previous study.
+We quickly discard the T-Trees method and show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method are very sensitive to the threshold of inclusion of SNPs, depending on the architecture of disease, as shown in previous studies.\\
 \textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
@@
 
-Many methods have been developed in order to maximize the predictive power of genotypes for diseases, or more generally for phenotypes.
-A commonly used technique, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or genetic profiling ..or even just PRS..
+Many methods have been developed to predict disease status, or more generally any phenotype, based on SNP information.
+A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- also sometimes called genetic profiling ..or even just PRS..
 -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
@@
 First, we quickly discard T-Trees as being the best method for predicting disease status based on genotypes.
-Then, we show that penalized logistic regression consistently performs better than the C+T method whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending of the architecture of disease, as shown in \cite{ware2017heterogeneity}.
+Then, we show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method are very sensitive to the threshold of inclusion of SNPs, depending on the architecture of disease, as shown in \cite{ware2017heterogeneity}.
 
@@
 To simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% [CITATION?].
-We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only the chromosome 6), the size of the training set (..from 1000 to 6000 by 1000.., and the corresponding test set is composed of the remaining individuals), the number of causal variants and their location (30, 300 or 3000 anywhere on the dataset or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part..
+We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only the chromosome 6), the size of the training set (..from 1000 to 6000 by 1000..), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part..
 of the phenotypes (one ``simple model'' with only linear effects, and one ``fancy model'' which combines linear, dominant and interaction effects).
@@
 We then compute $$y_i = \underbrace{\sum_{j\in E_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in E_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|E_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$ where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $E_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|E_\text{causal}^{(q)}\right|\right\}\right\}$.
-Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still display a marginal effect.
-For both models, the $y_i$ are then standardized to make sure they have a variance exactly equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$ where $\epsilon \sim N(0, 1 - h^2)$.
+Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still ..display..
+a marginal effect.
+For both models, the $y_i$ are then standardized to make sure they have a variance exactly equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$, where $\epsilon \sim N(0, 1 - h^2)$.
 
@@
 It compares all methods (see section \ref{sec:methods}) on all other combination of parameters in 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}).
-Note that the T-Trees method is compared on only 5 simulations and an heritability of 80\%, which aims at quickly discarding this method that can require more that one hour of computation for one simulation.
+Note that the T-Trees method is compared on only 5 simulations and an heritability of 80\%, which aims at quickly discarding this method that can require more than one hour of computation for only one simulation.
 Scenario \textnumero2 consists in 100 simulations comparing the ``logit-simple'' and the C+T methods on a dataset composed of only the chromosome 6.
@@
 For this scenario, we use only the additive (``simple'') model, but continue to compare all previous variations of the other parameters.
-Finally, scenario \textnumero3 reuses the whole dataset but varying the size of the training set in order to assess how the number of samples affects predictive performance of the PRS and the ``logit-simple'' methods.
+Finally, scenario \textnumero3 reuses the whole dataset but varying the size of the training set in order to assess how the sample size affects predictive performance of the PRS and the ``logit-simple'' methods.
 100 simulations are run using 300 causal SNPs anywhere on the genome.
@@
 This measure indicates how well we can distinguish between cases and controls using PRSs.
-As a second measure, we report the percentage of the partial AUC for a specificity between 90\% and 100\% \cite[]{dodd2003partial,mcclish1989analyzing}.
-This measure is similar to the AUC, but focuses on high specificities (i.e.\ discarding most of the healthy population as being diseased) which is often the most useful part of the ROC curve in clinical settings.
+As a second measure, the partial AUC for specificities between 90\% and 100\% is also reported \cite[]{dodd2003partial,mcclish1989analyzing}.
+This measure is similar to the AUC, but focuses on high specificities (i.e.\ discarding most of the healthy population as being diseased), which is often the most useful part of the ROC curve in clinical settings.
 
@@
 
-The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWASs), named summary statistics.
+The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Studies (GWAS), named summary statistics.
 For the C+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
@@
 T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}.
-Since this method is significantly different from the other tested here and achieved surprisingly ..good..
-results in the past, we decided to include T-Trees in this comparison in order to assess whether or not it could give superior predictive performance.
+Since this method is significantly different from the other presented methods and achieved surprisingly high predictive performance in the past, we decided to include T-Trees in this comparison in order to assess whether or not it could give superior predictive performance.
 We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}.
@@
 
-The last method we compare is penalized logistic regression.
+The last method we compare is a penalized logistic regression.
 We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that need to be determined.
@@
 
-To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in our package bigstatsr \cite[]{prive2017efficient}.
+To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in the R package bigstatsr \cite[]{prive2017efficient}.
 This type of algorithm builds predictions for many values of $\lambda$ (typically a ``regularization path'' of 100 values).
@@
 Secondly, in turn, each fold is considered as an inner validation set and the others ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes the likelihood on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is chosen.
-Finally, the $K$ resulting vectors of coefficients are reported and can be combined into one vector (e.g.\ using the geometric median) or resulting scores can be combined (e.g.\ averaging the $K$ probabilities for each individual).
-We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization path, making the whole procedure much faster.
+Finally, the $K$ resulting vectors of coefficients are reported and can be combined into one vector (e.g.\ by using the geometric median) or resulting scores can be combined (e.g.\ by averaging the $K$ probabilities for each individual).
+We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making the whole procedure much faster.
 Because of L1-regularization, this vector of coefficients is typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts.
@@
 
-In results, we use estimates of maximum achievable AUC values for the simulations: 84\% (respectively 94\%) for heritabilities of 50\% (respectively 80\%) based on three estimations (see Supplementary Materials).
+In ..the..
+results, we use estimates of maximum achievable AUC values for the simulations: 84\% (94\%) for heritabilities of 50\% (80\%) based on three estimations (see Supplementary Materials).
 Note that we do not report the simulation results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).
@@
 Using the penalized logistic regression, we report AUCs greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, which is not obtained by the upper bound of the C+T method in any scenario (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
-Our penalized regression methods consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted sometimes when we simulate 3000 causals SNPs anywhere on the genome, a case where all methods rarely exceeds 65\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
+The penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted sometimes when we simulate 3000 causals SNPs anywhere on the genome, a case where all methods rarely exceeds 65\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
 In the case where there are correlation between predictors (i.e.\ when we simulate 30 SNPs in the HLA region), the ``logit-simple'' method particularly performs better than the ``PRS-max'' one, with a mean AUC of 92.5\% versus 83\%, for a maximum achievable AUC of 94\% (Figure \ref{fig:main-AUC-logit}).
-We also report that the choice of the threshold of inclusion of SNPs in the C+T method can greatly impact its performance, for example dropping from an AUC of 90\% to less than 60\% when including all SNPs passing clumping in a model with only 30 causal SNPs, or dropping from 60\% to a random predictor (50\% of AUC) when including only GWAS-significant SNPs in a model with 3000 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
+We also report that the choice of the threshold of inclusion of SNPs in the C+T method can greatly impact its performance, for example dropping from an AUC of 90\% to less than 60\% when including all SNPs passing clumping in a model with only 30 causal SNPs, or dropping from 60\% to a random predictor (AUC of 50\%) when including only GWAS-significant SNPs in a model with 3000 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
 
@@
 
-First, in scenario \textnumero1 we also tested the T-Trees method. 
-For example, compared to our penalized logistic regression, T-Trees perform worse in term predictive measures, while taking much longer to run and making more complex predictive models because it uses more SNPs in the models and has non-linear effects (Figure \ref{fig:supp-ttrees}).
-Even when simulating a fancy model in which there is dominant and interaction effects that decision trees based methods should handle, AUC is still worse with T-Trees as compared with the ``logit-simple'' method.
+First, in scenario \textnumero1, we also tested the T-Trees method. 
+For example, as compared to the ``logit-simple'' method, T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because it uses more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}).
+Even when simulating a fancy model in which there are dominant and interaction effects that decision trees based methods should handle, AUC is still lower with T-Trees as compared with the ``logit-simple'' method.
 
-Secondly, as for the comparison between our two penalized logistic regression in scenario \textnumero1, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
+Secondly, as for the comparison between the two penalized logistic regressions in scenario \textnumero1, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
 Yet, the ``triple'' solution takes 2-3 more times to run and 3 more disk storage.
 
-Thirdly, for the ``fancy'' simulation model, we report the exact same results as in the previous subsection that is to say that the ``logit-simple'' give greater AUCs than the ``PRS-max'', excepted in the case of 3000 causal SNP anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}).
-Moreover, the choice of the threshold inclusion of SNPs also greatly affect the performance of the C+T method (Figure \ref{fig:supp-AUC-PRS-fancy}).
+Thirdly, for the ``fancy'' simulation model, we report the exact same results as in the previous subsection that is to say that the ``logit-simple'' achieves greater AUCs than the ``PRS-max'', excepted in the case of 3000 causal SNP anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}).
+Moreover, the choice of the threshold inclusion of SNPs also greatly ..affects/impacts..
+the performance of the C+T method (Figure \ref{fig:supp-AUC-PRS-fancy}).
 
@@
 
-These simulations corresponds to scenarios \textnumero2 and \textnumero3 (Table \ref{tab:simus}).
+These simulations correspond to scenarios \textnumero2 and \textnumero3 (Table \ref{tab:simus}).
 First, when reproducing simulations of scenario \textnumero1 on only the chromosome 6, methods globally increase or ..stagnate..
@@
 On the contrary, when simulating only 30 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' seems to slightly decrease.
-Secondly, when varying the training size, we report an increase of AUC when increasing the training size, with the AUC of the ``logit-simple'' method increasing faster than the ``PRS-max'' method (Figure \ref{fig:main-AUC-ntrain}).
+Secondly, when varying the training size, we report an increase of AUC when increasing the training size, with the AUCs provided by the ``logit-simple'' method increasing faster than for the ``PRS-max'' method (Figure \ref{fig:main-AUC-ntrain}).
 
@@
 phenotypes.
-For one run [ENOUGH FOR REVIEWERS?], splitting the dataset in 12,000 indiduals in the training set and the remaining 3155 individuals in the test set, we compare the predictive performance of the ``C+T'', ``logit-simple'' and ``logit-triple'' methods, but also computation times and number of predictors used in the models (Table \ref{tab:results-celiac}, figure \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
+For one run [ENOUGH FOR REVIEWERS?], splitting the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals, we compare the predictive performance of the ``C+T'', ``logit-simple'' and ``logit-triple'' methods, but also computation times and number of predictors used in the models (Table \ref{tab:results-celiac}, figure \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
 We ..see..
-that the joint logistic regressions provide higher AUC with 88.71\% with the ``logit-simple'' and 89.26 with the ``logit-triple as compared to 82.46 with the ``C+T'' method, and much higher partial AUC (specificity 0.9-1), respectively 0.0411, 0.0440 and 0.0283, while using less predictors, respectively 2198, 4617 and 13277.
-Note that for the ``C+T'' method, we still report the best result among 102 p-values thresholds.
-In terms of computation time, we report only the GWAS computation for the ``C+T'' method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once, is faster than the fast ``C+T'' method, and the ``logit-triple'' takes twice as long as the ``logit-simple''.
+that the joint logistic regressions provide higher AUC with 88.71\% for the ``logit-simple'' and 89.26 for the ``logit-triple as compared to 82.46 for the ``C+T'' method, and much higher partial AUC (specificity 0.9-1), respectively 0.0411, 0.0440 and 0.0283, while using less predictors, respectively 2198, 4617 and 13277.
+Note that for the ``C+T'' method, we still report the best result among 102 p-value thresholds. 
+In terms of computation time, we report only the GWAS computation for the ``C+T'' method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once, is faster than the ``C+T'' method, and the ``logit-triple'' takes twice as long as the ``logit-simple''.
 Finally, note that the scores you get with the logistic regressions are probabilities, which is not the case with the ``C+T'' method (Figure \ref{fig:supp-score-densities}).[QUE RAPPORTER SUR LES DENSITY PLOTS DU SUPMAT?]
@@
 In this comparative study, we propose other methods as a replacement to the widely used C+T method. 
-We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really high predictive performance, as reported in a previous study \cite[]{abraham2013performance}, which is not the case with the C+T method that uses SNP effects learned independently.
+We show that jointly estimating the SNP effects using efficient penalized logistic regressions enables to achieve really high predictive performance, as reported in a previous study \cite[]{abraham2013performance}, which is not the case with the C+T method that uses SNP effects learned independently.
 We are well aware that the C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions yet.
-However, in disease prediction for clinical interventions, best possible methods should be used, even if it has to be used by large consortia only [EVEN LARGE CONSORTIA DON'T HAVE ALL DATA?].
-The logistic regression we present here could be used on data as large as biobank datasets because it is very efficient in both memory and computation time (..even faster than doing a GWAS!..).
+However, in order to accurately predict disease status for clinical interventions, best possible methods should be used.
+The logistic regression we present in this paper could be used on data as large as biobank datasets because it is very efficient in both memory and computation time (..even faster than doing a GWAS!..).
 
-%One does not simply identify high-risk individuals (Figure \ref{fig:meme}).
+As for the C+T method itself, we recall that we report the maximum AUC of 102 thresholds, a threshold that should normally be learned on the training set only.
+The choice of this threshold is very important as it can greatly impact the predictive performance of the method.
+On the contrary, in the penalized logistic regression we present, we remove the choice of the regularization parameter with a procedure that we call Cross-Model Selection and Averaging (CMSA) while keeping high predictive performance.
 
-As for the C+T method itself, we recall that we report the maximum AUCs of 102 thresholds, a threshold that should normally be learned on the training set only.
-The choice of this threshold is very important as it can greatly impact the predictive performance of the method.
-In the penalized logistic regression we present, we remove the choice of the regularization parameter with a procedure that we call Cross-Model Selection and Averaging (CMSA) while keeping the same predictive performance.
+We also introduce a simple feature engineering technique which enables the logistic regression to detect and learn not only additive effects, but also dominant and recessive ones.
+This technique improves predictive performance of the method when there are some non-linear effects in the simulations, while providing the same predictive performance when there are only linear effects.
+Moreover, it also slightly improves predictive performance for the real disease we ..also..
+study in this paper.
+If this technique proves to be useful for predicting other diseases, it would be useful to re-implement another algorithm that could work directly on the initial dataset (e.g.
+by coding a new accessor).
 
-We also introduce a simple feature engineering technique to detect and learn dominant and recessive effects..
-
-[PARLER DU PROBLEME DE POP? + future paper + possible car fast and memory efficient]
+Finally, in this comparative study, we did not consider the problem of population structure and also did not consider non-genetic data.
+In another study, we will assess how can we use ..training..
+in one population to improve training and prediction in another population.
+We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}.
+In a ..final..
+study, we will also assess how can we combine the information provided by genetic data with clinical and environmental data, possibly in a non-linear way.
 

### 2018-03-08 5-results-celiac.Rmd
multiple runs

@@
 
-## Methods' functions
-
-Each method's function returns a tibble (data frame) with 4 columns:
-
-1. the name of the method,
-2. the predictive scores and true phenotypes for the test set, as a [list-column](http://r4ds.had.co.nz/many-models.html),
-3. the timing of the main computations (in seconds),
-4. the number of SNPs used for the prediction.
-
-### PRS
-
-```{r}
-PRS <- function(G, infos.chr, infos.pos, 
-                pheno.all, covar.all, 
-                ind.train, ind.test) {
-
-  timing <- system.time({
-
-    # GWAS
-    gwas.train <- big_univLogReg(
-      G, pheno.all[ind.train], ind.train = ind.train, 
-      covar.train = covar.all[ind.train, , drop = FALSE], 
-      ncores = nb_cores()
-    )
-    gwas.train.gc <- snp_gc(gwas.train)
-
-    # Clumping on the test set
-    ind.keep <- snp_clumping(G, infos.chr = infos.chr,
-                             ind.row = ind.test,
-                             thr.r2 = 0.2, 
-                             S = abs(gwas.train.gc$score),
-                             size = 500,
-                             is.size.in.bp = TRUE,
-                             infos.pos = infos.pos,
-                             ncores = nb_cores())
-
-    # PRS
-    thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
-    lpS <- -predict(gwas.train.gc)
-    nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
-    prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
-                   ind.test = ind.test,
-                   ind.keep = ind.keep,
-                   lpS.keep = lpS[ind.keep], 
-                   thr.list = thrs)
-  })[3]
-
-  ind.best <- which.max(apply(prs, 2, AUC, target = pheno.all[ind.test]))
-
-  tibble(
-    method = "PRS-max",
-    eval = list(cbind(prs[, ind.best], pheno.all[ind.test])),
-    timing = timing,
-    nb.preds = nb.pred[ind.best]
-  )
-}
-```
-
-### Logistic regression
-
-```{r}
-logit.CMSA <- function(G, pheno.all, covar.all, ind.train, ind.test, method, print_all = FALSE) {
-
-  timing <- system.time({
-
-    cmsa.logit <- big_spLogReg(X = G, y01.train = pheno.all[ind.train], 
-                               ind.train = ind.train, 
-                               covar.train = covar.all[ind.train, , drop = FALSE],
-                               ncores = nb_cores())
-
-    preds.all <- predict(cmsa.logit, X = G, ind.row = ind.test, 
-                         covar.row = covar.all[ind.test, , drop = FALSE])
-    if (print_all) print(preds.all)
-    preds <- rowMeans(preds.all)
-  })[3]
-
-  tibble(
-    method   = method, 
-    eval     = list(cbind(preds, pheno.all[ind.test])),
-    timing   = timing,
-    nb.preds = sum(rowSums(sapply(cmsa.logit, function(x) x$beta.X) != 0) > 0)
-  )
-}
-```
-
 ## Predictive models and plots
@@
 ```{r}
-res.PRS <- PRS(
-  G = G,
-  infos.chr = CHR,
-  infos.pos = POS,
-  pheno.all = y,
-  covar.all = obj.svd$u,
-  ind.train = ind.train,
-  ind.test = ind.test
-)
 # GWAS
@@
 
-## Predictive methods (multiple times)
-
-```{r}
-# Divide in training/test sets
-ind.train <- sort(sample(nrow(G), size = 12e3))
-ind.test <- setdiff(rows_along(G), ind.train)
-```
-
-### PRS
-
-```{r}
-# PRS
-PRS <- function() {
-
-  system.time(
-  gwas.train <- big_univLogReg(
-    G, y[ind.train], ind.train = ind.train, 
-    covar.train = obj.svd$u[ind.train, , drop = FALSE], 
-    ncores = NCORES
-  )
-)
-}
-
-```
-
-### Logistic regression
-
-Cross-Model Selection and Averaging (CMSA):
-
-```{r}
-# Train logit-simple
-system.time(
-  logit <- big_spLogReg(X = G, y01.train = y[ind.train], 
-                        ind.train = ind.train, 
-                        covar.train = obj.svd$u[ind.train, ],
-                        ncores = NCORES)
-)
-# Get K = 10 predictions/individuals
-preds1 <- predict(logit, X = G, ind.row = ind.test, 
-                  covar.row = obj.svd$u[ind.test, ])
-(nb.pred1 <- sum(rowSums(sapply(logit, function(l) l$beta.X != 0)) > 0))
-(aucs1 <- apply(preds1, 2, AUC, target = y[ind.test]))
-# Combining all K predictions
-preds1.comb <- rowMeans(preds1)
-AUC(preds1.comb, y[ind.test])
-```
-
-```{r}
-(p2 <- plot_density_AUC(preds1.comb))
-```
-
-### Logistic regression with feature engineering
-
-```{r}
-G2 <- big_attach("backingfiles/celiacQC_tripled1.rds")
-system.time(
-  logit2 <- big_spLogReg(X = G2, y01.train = y[ind.train], 
-                         ind.train = ind.train, 
-                         covar.train = obj.svd$u[ind.train, ],
-                         ncores = NCORES)
-)
-
-preds2 <- predict(logit2, X = G2, ind.row = ind.test, 
-                  covar.row = obj.svd$u[ind.test, ])
-(nb.pred2 <- sum(rowSums(sapply(logit2, function(l) l$beta.X != 0)) > 0))
-(aucs2 <- apply(preds2, 2, AUC, target = y[ind.test]))
-# Combining all K predictions
-preds2.comb <- rowMeans(preds2)
-AUC(preds2.comb, y[ind.test])
-```
-
-```{r}
-(p3 <- plot_density_AUC(preds2.comb))
-```
-
-```{r, out.width="70%", fig.asp=900/780}
-list(p1, p2, p3) %>%
-  lapply(function(p) p + theme(legend.position = "none")) %>%
-  cowplot::plot_grid(plotlist = ., ncol = 1, labels = LETTERS[1:3], label_size = 20, scale = 0.95) %>%
-  cowplot::plot_grid(., cowplot::get_legend(p1), rel_widths = c(1, 0.15), scale = 0.95)
-ggsave("figures/supp-score-densities.pdf", scale = 1/100, width = 780, height = 900)
-```
-
 ## Top scores
@@
 ```
+
+## Predictive methods (multiple times)
+
+### Methods' functions
+
+Each method's function returns a tibble (data frame) with 4 columns:
+
+1. the name of the method,
+2. the predictive scores and true phenotypes for the test set, as a [list-column](http://r4ds.had.co.nz/many-models.html),
+3. the timing of the main computations (in seconds),
+4. the number of SNPs used for the prediction.
+
+#### PRS
+
+```{r}
+PRS <- function(G, infos.chr, infos.pos, 
+                pheno.all, covar.all, 
+                ind.train, ind.test) {
+
+  timing <- system.time({
+
+    # GWAS
+    gwas.train <- big_univLogReg(
+      G, pheno.all[ind.train], ind.train = ind.train, 
+      covar.train = covar.all[ind.train, , drop = FALSE], 
+      ncores = nb_cores()
+    )
+    gwas.train.gc <- snp_gc(gwas.train)
+
+    # Clumping on the test set
+    ind.keep <- snp_clumping(G, infos.chr = infos.chr,
+                             ind.row = ind.test,
+                             thr.r2 = 0.2, 
+                             S = abs(gwas.train.gc$score),
+                             size = 500,
+                             is.size.in.bp = TRUE,
+                             infos.pos = infos.pos,
+                             ncores = nb_cores())
+
+    # PRS
+    thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
+    lpS <- -predict(gwas.train.gc)
+    nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
+    prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
+                   ind.test = ind.test,
+                   ind.keep = ind.keep,
+                   lpS.keep = lpS[ind.keep], 
+                   thr.list = thrs)
+  })[3]
+
+  ind.best <- which.max(apply(prs, 2, AUC, target = pheno.all[ind.test]))
+
+  tibble(
+    method = "PRS-max",
+    eval = list(cbind(prs[, ind.best], pheno.all[ind.test])),
+    timing = timing,
+    nb.preds = nb.pred[ind.best]
+  )
+}
+```
+
+#### Logistic regression
+
+```{r}
+logit.CMSA <- function(G, pheno.all, covar.all, ind.train, ind.test, method, print_all = FALSE) {
+
+  timing <- system.time({
+
+    cmsa.logit <- big_spLogReg(X = G, y01.train = pheno.all[ind.train], 
+                               ind.train = ind.train, 
+                               covar.train = covar.all[ind.train, , drop = FALSE],
+                               ncores = nb_cores())
+
+    preds.all <- predict(cmsa.logit, X = G, ind.row = ind.test, 
+                         covar.row = covar.all[ind.test, , drop = FALSE])
+    if (print_all) print(preds.all)
+    preds <- rowMeans(preds.all)
+  })[3]
+
+  tibble(
+    method   = method, 
+    eval     = list(cbind(preds, pheno.all[ind.test])),
+    timing   = timing,
+    nb.preds = sum(rowSums(sapply(cmsa.logit, function(x) x$beta.X) != 0) > 0)
+  )
+}
+```
+
+### Run multiple times
+
+```{r, eval=TRUE}
+if (!dir.exists("results6")) dir.create("results6")
+pheno.all <- y
+covar.all <- obj.svd$u
+
+for (i in 1:100) {
+
+  res.file <- paste0("results6/simu_", i, ".rds")
+  if (file.exists(res.file)) next
+
+  # Split in training/test sets
+  ind.train <- sort(sample(nrow(G), size = 12e3))
+  ind.test <- setdiff(rows_along(G), ind.train)
+
+  # Get results from all methods
+  res <- bind_rows(
+    PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test),
+    logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple"),
+    logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple")
+  )
+  saveRDS(res, file = res.file)
+}
+```
+```{r}
+pAUC <- function(pred, target, p = 0.1) {
+  val.min <- min(target)
+  q <- quantile(pred[target == val.min], probs = 1 - p)
+  ind <- (target != val.min) | (pred > q)
+  tryCatch(bigstatsr::AUC(pred[ind], target[ind]) * p, 
+           error = function(e) p^2 / 2)  # all same prediction
+}
+
+format_results <- function(results) {
+  results %>%
+    as_tibble() %>%
+    mutate(
+      AUC = map_dbl(eval, ~ bigstatsr::AUC(.x[, 1], .x[, 2])),
+      pAUC = map_dbl(eval, ~ pAUC(.x[, 1], .x[, 2]))
+    )
+}
+
+boot <- function(x, n = 1e5, f = mean) {
+  sd(replicate(n, f(sample(x, replace = TRUE))))
+}
+
+list.files(path = "results6/", full.names = TRUE) %>%
+  map_df(readRDS) %>%
+  format_results() %>%
+  group_by(method) %>%
+  summarize_at(-1, function(x) {
+    glue::glue("{signif(mean(x), 3)} ({signif(boot(x), 3)})")
+  }) %>%
+  print() %>%
+  xtable::xtable()
+```

### 2018-03-08 paper/draft.tex
multiple runs

@@
 \begin{table}[h]
-\caption{Results for the real Celiac dataset.\label{tab:results-celiac}}
+\caption{Results for the real Celiac dataset.
+The results are averaged over 100 runs where the training step is randomly composed of 12,000 individuals.
+In the parentheses is reported the standard deviation of $10^5$ bootstrap samples of the mean of the corresponding variable.
+Results are reported with 3 significant digits.\label{tab:results-celiac}}
 \vspace*{0.5em}
@@
   \hline
-Method & AUC (\%) & pAUC & \# predictors & Execution time (s) \\ \hline
-PRS-max & 82.46 & 0.0283 & 13277 & 149 \\ 
-logit-simple & 88.71 & 0.0411 & 2198 & 78 \\
-logit-triple & 89.26 & 0.0440 & 4617 & 161 \\ \hline
+Method & AUC & pAUC & \# predictors & Execution time (s) \\ 
+  \hline
+PRS-max & 0.824 (0.000704) & 0.0286 (0.00016) & 9850 (781) & 148 (0.414) \\ 
+logit-simple & 0.888 (0.000468) & 0.0414 (0.000164) & 3220 (62) & 83.8 (1.27) \\ 
+logit-triple & 0.892 (0.000488) & 0.0429 (0.000174) & 4470 (80.6) & 141 (1.85) \\ 
+   \hline
 \end{tabular}

### 2018-03-12 paper/draft.tex
small English changes

@@
 We present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in a previous study.
-We quickly discard the T-Trees method and show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method are very sensitive to the threshold of inclusion of SNPs, depending on the architecture of disease, as shown in previous studies.\\
+We quickly discard the T-Trees method and show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture, as shown in previous studies.\\
 \textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
@@
 As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}.
-Moreover, screening based on sequencing seems to make individuals positively change their health behavior while not causing patient anxiety nor depression \cite[]{vassy2017impact}.
+Moreover, screening based on sequencing seems to make individuals positively change their health behavior, while neither causing patient anxiety nor depression \cite[]{vassy2017impact}.
 Yet, PRSs would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases. 
-For example, for screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than respectively 75\% and 99\% \cite[]{janssens2007impact}.
+For example, for screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.
 
@@
 -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
-This technique only use summary statistics which makes it very suitable and also very fast.
+This technique only ..uses/resorts to.. summary statistics, which makes it very ..suitable.. and also very fast.
 Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}.
-Yet, these models are not designed for predicting a binary trait such as a disease status and have proven to not be optimal at such task in another comparative study \cite[]{abraham2013performance}.
+Yet, these models are not designed for predicting a binary trait such as a disease status and have proven not to be optimal at such ..a..
+task in another comparative study \cite[]{abraham2013performance}.
 Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time requirements, which makes them unlikely to be used for prediction on large cohorts \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
@@
 
-There have already been comparative studies about predicting human binary phenotypes based on genotype data. 
+There have already been comparative studies on predicting human binary phenotypes based on genotype data. 
 \citep{abraham2013performance} showed that penalized logistic regression and penalized SVM performed better than the standard C+T method and LMMs, for a wide-range of human diseases.
@@
 \cite{ware2017heterogeneity} evaluated best practices for using the C+T method and found out that the optimal inclusion threshold is trait-dependent. 
-In this study, we particularly extend the work of \cite{abraham2013performance} by using a simulation framework similar but ..richer..
+In this study, we particularly extend the work of \cite{abraham2013performance} by using a simulation framework similar but ..richer/more extenstive..
 than the one used in \cite{zhou2013polygenic}.
-In order to make our comparison as comprehensive as possible, we compare different architectures of disease (number, size and location of causal effects and heritability) with different model for generating liability ..scores.., one with only linear effects, and one which combines linear, dominant and interaction effects.
-First, we quickly discard T-Trees as being the best method for predicting disease status based on genotypes.
-Then, we show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method are very sensitive to the threshold of inclusion of SNPs, depending on the architecture of disease, as shown in \cite{ware2017heterogeneity}.
+In order to make our comparison as comprehensive as possible, we compare different disease architectures (number, size and location of causal effects and heritability) with different model for generating liability ..scores.., one with only linear effects, and one which combines linear, dominant and ..interaction/interaction-type..
+effects.
+First, we quickly discard T-Trees as being the ..best..
+method for predicting disease status based on genotypes.
+Then, we show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture, as shown in \cite{ware2017heterogeneity}.
 
@@
 
-We use real genotypes of European individuals from a case/control celiac disease cohort \cite[]{dubois2010multiple}. Composition of this dataset is presented in table \ref{tab:celiac-data}.
+We use real genotypes of European individuals from a case/control celiac disease cohort \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}.
 Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
-For simulations, in order to remove the structure induced by the celiac disease status, we ..first keep only..
-controls from this cohort.
-Secondly, in order to remove population structure, we ..keep only..
+As for simulations presented here, in order to remove the structure induced by the celiac disease status, we ..first keep only/select/restrict/retain.. controls from this cohort. 
+Secondly, in order to ..alleviate..
+population structure, we ..keep only..
 people from the UK.
-Thirdly, we further subset these British people based on deviation from the robust Mahalanobis distance on principal components.
-This 3-step filtering results in 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing'').
-We decided to get rid of population structure because it can affect the predictive performance of methods, which we will investigate in another study.
+Thirdly, we further subset these British people based on deviation from the robust Mahalanobis distance ..on/for..
+principal components.
+This 3-step filtering process results in a sample of 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing'').
+Note that we decided to ..alleviate.. population structure because it can affect the predictive performance of methods, which we will investigate in another study.
 
@@
 
-To simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% [CITATION?].
-We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only the chromosome 6), the size of the training set (..from 1000 to 6000 by 1000..), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part..
-of the phenotypes (one ``simple model'' with only linear effects, and one ``fancy model'' which combines linear, dominant and interaction effects).
-For the ``simple model'', we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights (generated with a Gaussian or a Laplace distribution), $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs).
-For the ``fancy model'', we separate the causal SNPs in three equal sets $E_\text{causal}^{(1)}$, $E_\text{causal}^{(2)}$ 
+In order to simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% [CITATION?].
+We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only chromosome 6), the size of the training set ($\{1000, 2000, \dots, 6000\}$), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part/component/term..
+of the phenotypes (a ``simple'' model with only linear effects, and a ``fancy'' model which combines linear, dominant and ..interaction/interaction-type..
+effects).
+For the ``simple'' model, we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights (generated with a Gaussian or a Laplace distribution), $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs).
+For the ``fancy'' model, we separate the causal SNPs in three equal sets $E_\text{causal}^{(1)}$, $E_\text{causal}^{(2)}$ 
 and $E_\text{causal}^{(3)}$ ($E_\text{causal}^{(3)}$ is further separated in two equal sets, $E_\text{causal}^{(3.1)}$ and $E_\text{causal}^{(3.2)}$). 
-We then compute $$y_i = \underbrace{\sum_{j\in E_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in E_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|E_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$ where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $E_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|E_\text{causal}^{(q)}\right|\right\}\right\}$.
-Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still ..display..
-a marginal effect.
-For both models, the $y_i$ are then standardized to make sure they have a variance exactly equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$, where $\epsilon \sim N(0, 1 - h^2)$.
+We then compute $$y_i = \underbrace{\sum_{j\in E_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in E_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|E_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$
+where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $E_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|E_\text{causal}^{(q)}\right|\right\}\right\}$. 
+Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still ..display/have.. a marginal effect. 
+For both models, the ..$y_i$/values of $y_i$..
+are then standardized to make sure they have a variance exactly equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$, where $\epsilon \sim N(0, 1 - h^2)$.
 
-We make 3 different simulation scenarios, summarized in table \ref{tab:simus}. 
-Scenario \textnumero1 uses the whole dataset (the 22 chromosomes) and a training set of size 6000.
-It compares all methods (see section \ref{sec:methods}) on all other combination of parameters in 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}).
-Note that the T-Trees method is compared on only 5 simulations and an heritability of 80\%, which aims at quickly discarding this method that can require more than one hour of computation for only one simulation.
-Scenario \textnumero2 consists in 100 simulations comparing the ``logit-simple'' and the C+T methods on a dataset composed of only the chromosome 6.
-Reducing the number of SNPs aims at increasing the ..polygenicity..
-of our simulated models and at ..virtually..
+We ..make/construct/design/implement/work with.. 3 different simulation scenarios, summarized in table \ref{tab:simus}. 
+Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes) and a training set of size 6000.
+It compares all methods described in section \ref{sec:methods} ..on/for/along/in relation to..
+all other combinations of parameters in 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}).
+Note that the T-Trees method is compared on only 5 simulations and an heritability of 80\%, which aims at quickly discarding this method that can require more than one hour of computation for a single simulation only.
+Scenario \textnumero2 consists of 100 simulations comparing the ``logit-simple'' and the C+T methods on a dataset composed of chromosome 6 only.
+Reducing the number of SNPs aims at increasing the polygenicity of the simulated models and at ..virtually/artificially..
 increasing the sample size [REF?
@@
 For this scenario, we use only the additive (``simple'') model, but continue to compare all previous variations of the other parameters.
-Finally, scenario \textnumero3 reuses the whole dataset but varying the size of the training set in order to assess how the sample size affects predictive performance of the PRS and the ``logit-simple'' methods.
+Finally, scenario \textnumero3 reuses the whole dataset but this time varying the size of the training set in order to assess how the sample size affects predictive performance of the PRS and the ``logit-simple'' methods.
 100 simulations are run using 300 causal SNPs anywhere on the genome.
@@
 In the case of our study, the AUC is the probability that a PRS of a case is greater than the PRS of a control.
-This measure indicates how well we can distinguish between cases and controls using PRSs.
+This measure indicates the extent to which we can distinguish between cases and controls using PRSs.
 As a second measure, the partial AUC for specificities between 90\% and 100\% is also reported \cite[]{dodd2003partial,mcclish1989analyzing}.
@@
 
-\subsection{Compared methods} \label{sec:methods}
+\subsection{Methods compared} \label{sec:methods}
 
@@
 
-The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Studies (GWAS), named summary statistics.
+The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Studies (GWAS), referred to as summary statistics.
 For the C+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
-The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with each other. 
+The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another. 
 Then, thresholding (T) consists in removing SNPs that are within a certain level of significance (p-value threshold to be determined). 
@@
 For the C+T method, we report three different scores of prediction: one including all the SNPs (remaining after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximizes the AUC for these two thresholds and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
-Note that, normally, this optimal threshold would have to be learned on the training set only in order to prevent from over-fitting the data.
+Note that, normally, this optimal threshold would have to be learned based on the training set only in order to prevent from over-fitting the data.
 We consider this reported maximum AUC as an upper bound of the AUC for the C+T method.
@@
 T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}.
-Since this method is significantly different from the other presented methods and achieved surprisingly high predictive performance in the past, we decided to include T-Trees in this comparison in order to assess whether or not it could give superior predictive performance.
-We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}.
-Yet, we use only 100 trees (instead of 1000) because it showed only a very subtle increase of AUC while ..taking way too much time..
-(e.g.\ AUC of 81.5\% instead of 81\%, data not shown).
+Since this method is ..significantly..
+different from the other ones presented here and give that it achieved surprisingly high predictive performance in the past, we decided to include T-Trees in this comparison.
+We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees only (instead of 1000) because ..it..
+showed only a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown).
 %We call this method ``T-Trees'' in the results.
 
-The last method we compare is a penalized logistic regression.
-We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that need to be determined.
+The last method ..in/of.. this comparison is a penalized logistic regression.
+We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that needs to be determined.
 Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
-the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be exactly zero and can be used as a means of variable selection, leading to sparse (and therefore more interpretable) models  (corresponds to $\alpha = 1$);
-the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation where there are many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
+the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be exactly ..zero/equal to zero..
+and can be used as a means of variable selection, leading to sparse (and therefore more interpretable) models  (corresponds to $\alpha = 1$);
+the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
 In this study, we always use the elastic-net regularization with $\alpha = 0.5$, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.
@@
 This type of algorithm builds predictions for many values of $\lambda$ (typically a ``regularization path'' of 100 values).
-To get an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA).
+To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA).
 First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds).
-Secondly, in turn, each fold is considered as an inner validation set and the others ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes the likelihood on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is chosen.
+Secondly, in turn, each fold is considered as an inner validation set and the other ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes the likelihood on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is selected.
 Finally, the $K$ resulting vectors of coefficients are reported and can be combined into one vector (e.g.\ by using the geometric median) or resulting scores can be combined (e.g.\ by averaging the $K$ probabilities for each individual).
-We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making the whole procedure much faster.
-Because of L1-regularization, this vector of coefficients is typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts.
-We call this method ``logit-simple'' in the results.
+We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making this procedure much faster.
+Because of the L1-regularization, the resulting vectors of coefficients are typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts.
+We refer to this method as ``logit-simple'' in the results section.
 
-In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct another dataset with, for each SNP variable, two more variables coding for recessive and dominant effects.
-This results in a dataset with 3 times more variables than the initial one, on which we can apply the penalized logistic regression with the CMSA procedure, as described previously.
+In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects.
+This results in a dataset with 3 times as many variables as the initial one, on which we can apply the penalized logistic regression with the CMSA procedure, as described previously.
 We call this method ``logit-triple'' in the results.
@@
 
-In ..the..
-results, we use estimates of maximum achievable AUC values for the simulations: 84\% (94\%) for heritabilities of 50\% (80\%) based on three estimations (see Supplementary Materials).
-Note that we do not report the simulation results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).
+When reporting results, we use estimates of maximum achievable AUC values for the simulations: 84\% (94\%) for heritabilities of 50\% (80\%) based on three estimations (see Supplementary Materials).
+Note that for the simulations we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).
 
@@
 In scenario \textnumero1, we compared our penalized logistic regression (``logit-simple'') with the C+T method (``PRS'') on the whole genome (Table \ref{tab:simus}). 
-Using the penalized logistic regression, we report AUCs greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, which is not obtained by the upper bound of the C+T method in any scenario (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
+Using the penalized logistic regression, we report AUCs greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, which is not obtained by the upper bound of the C+T method in any of the scenarios (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
 The penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted sometimes when we simulate 3000 causals SNPs anywhere on the genome, a case where all methods rarely exceeds 65\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
-In the case where there are correlation between predictors (i.e.\ when we simulate 30 SNPs in the HLA region), the ``logit-simple'' method particularly performs better than the ``PRS-max'' one, with a mean AUC of 92.5\% versus 83\%, for a maximum achievable AUC of 94\% (Figure \ref{fig:main-AUC-logit}).
-We also report that the choice of the threshold of inclusion of SNPs in the C+T method can greatly impact its performance, for example dropping from an AUC of 90\% to less than 60\% when including all SNPs passing clumping in a model with only 30 causal SNPs, or dropping from 60\% to a random predictor (AUC of 50\%) when including only GWAS-significant SNPs in a model with 3000 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
+In the case where there are correlations between predictors (i.e.\ when we simulate 30 SNPs in the HLA region), the ``logit-simple'' method performs particularly ..better..
+than the ``PRS-max'' one, with a mean AUC of 92.5\% versus 83\%, for a maximum achievable AUC of 94\% (Figure \ref{fig:main-AUC-logit}).
+We also report that the choice of the threshold of inclusion of SNPs in the C+T method can impact its performance greatly, for example dropping from an AUC of 90\% to less than 60\% when including all SNPs passing the clumping phase in a model with only 30 causal SNPs, or dropping from 60\% to a random predictor (AUC of 50\%) when including only GWAS-significant SNPs in a model with 3000 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
 
@@
 First, in scenario \textnumero1, we also tested the T-Trees method. 
-For example, as compared to the ``logit-simple'' method, T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because it uses more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}).
-Even when simulating a fancy model in which there are dominant and interaction effects that decision trees based methods should handle, AUC is still lower with T-Trees as compared with the ``logit-simple'' method.
+For example, as compared to the ``logit-simple'' method, T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}).
+Even when simulating a fancy model in which there are dominant and ..interaction/interaction-type..
+effects that decision trees based methods should be able to handle, AUC is still lower with T-Trees as compared with the ``logit-simple'' method.
 
 Secondly, as for the comparison between the two penalized logistic regressions in scenario \textnumero1, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
-Yet, the ``triple'' solution takes 2-3 more times to run and 3 more disk storage.
+Yet, the ``triple'' solution takes 2-3 times as much time to run and requires 3 times as much disk storage as the ``logit-simple''.
 
@@
 These simulations correspond to scenarios \textnumero2 and \textnumero3 (Table \ref{tab:simus}).
-First, when reproducing simulations of scenario \textnumero1 on only the chromosome 6, methods globally increase or ..stagnate..
+First, when reproducing simulations of scenario \textnumero1 on only chromosome 6, methods globally increase or ..stagnate..
 in predictive performance (Figure \ref{fig:supp-AUC-chr6}).
@@
 phenotypes.
-For one run [ENOUGH FOR REVIEWERS?], splitting the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals, we compare the predictive performance of the ``C+T'', ``logit-simple'' and ``logit-triple'' methods, but also computation times and number of predictors used in the models (Table \ref{tab:results-celiac}, figure \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
+For 100 runs, randomly splitting the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals, we compare the predictive performance of the ``C+T'', ``logit-simple'' and ``logit-triple'' methods, but also computation times and number of predictors used in the models (Table \ref{tab:results-celiac}, figure \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
 We ..see..
-that the joint logistic regressions provide higher AUC with 88.71\% for the ``logit-simple'' and 89.26 for the ``logit-triple as compared to 82.46 for the ``C+T'' method, and much higher partial AUC (specificity 0.9-1), respectively 0.0411, 0.0440 and 0.0283, while using less predictors, respectively 2198, 4617 and 13277.
+that the joint logistic regressions provide higher AUC with 88.8\% for the ``logit-simple'' and 89.2\% for the ``logit-triple'' as compared to 82.4\% for the ``C+T'' method, and much higher partial AUC (specificity 0.9-1), respectively 0.0414, 0.0429 and 0.0286, while using less predictors, respectively 3320, 4470 and 9850.
 Note that for the ``C+T'' method, we still report the best result among 102 p-value thresholds. 
-In terms of computation time, we report only the GWAS computation for the ``C+T'' method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once, is faster than the ``C+T'' method, and the ``logit-triple'' takes twice as long as the ``logit-simple''.
-Finally, note that the scores you get with the logistic regressions are probabilities, which is not the case with the ``C+T'' method (Figure \ref{fig:supp-score-densities}).[QUE RAPPORTER SUR LES DENSITY PLOTS DU SUPMAT?]
+In terms of computation time, we report only the GWAS computation for the ``C+T'' method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once, is faster than the ``C+T'' method (84 / 148 seconds), and the ``logit-triple'' takes less than twice as long as the ``logit-simple'' (141 / 84 seconds).
+Finally, note that the scores obtained with the logistic regressions are probabilities, which is not the case with the ``C+T'' method (Figure \ref{fig:supp-score-densities}).
 
@@
 
-In this comparative study, we propose other methods as a replacement to the widely used C+T method. 
-We show that jointly estimating the SNP effects using efficient penalized logistic regressions enables to achieve really high predictive performance, as reported in a previous study \cite[]{abraham2013performance}, which is not the case with the C+T method that uses SNP effects learned independently.
+In this comparative study, we propose other methods as a replacement for the widely used C+T method. 
+We show that jointly estimating the SNP effects using efficient penalized logistic regressions enables to achieve ..really..
+high predictive performance, as reported in a previous study \cite[]{abraham2013performance}.
+This is not the case with the C+T method that uses SNP effects learned independently.
 We are well aware that the C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions yet.
@@
 The choice of this threshold is very important as it can greatly impact the predictive performance of the method.
-On the contrary, in the penalized logistic regression we present, we remove the choice of the regularization parameter with a procedure that we call Cross-Model Selection and Averaging (CMSA) while keeping high predictive performance.
+On the contrary, in the penalized logistic regression presented here, we remove the choice of the regularization parameter with a procedure that we call Cross-Model Selection and Averaging (CMSA) while keeping high predictive performance.
 
 We also introduce a simple feature engineering technique which enables the logistic regression to detect and learn not only additive effects, but also dominant and recessive ones. 
-This technique improves predictive performance of the method when there are some non-linear effects in the simulations, while providing the same predictive performance when there are only linear effects.
+This technique improves the predictive performance of the method when there are some non-linear effects in the simulations, while providing the same predictive performance when there are only linear effects.
 Moreover, it also slightly improves predictive performance for the real disease we ..also..

### 2018-03-14 5-results-celiac.Rmd
change colors

@@
 
+methods.color <- setNames(scales::hue_pal()(6), 
+                          c("logit-simple", "PRS-max", "PRS-stringent",
+                            "T-Trees", "logit-triple", "PRS-all"))
+
 bigstatsr:::MY_THEME(
@@
   theme(legend.position = c(0.7, 0.3), legend.key.width = unit(4, "line")) +
-  coord_equal()
+  coord_equal() +
+  scale_color_manual(values = methods.color)
 ```

### 2018-03-15 5-results-celiac.Rmd
change cache

@@
 knitr::opts_chunk$set(fig.align = "center", out.width = "70%", fig.width = 10,
-                      fig.asp = 0.7, dev = "svg", cache = TRUE)
+                      fig.asp = 0.7, dev = "svg", cache = FALSE)
 options(width = 95)
@@
 
-```{r}
+```{r, cache=TRUE}
 # Compute PCA
@@
 
-```{r}
+```{r, cache=TRUE}
 # Divide in training/test sets
@@
 
-```{r}
+```{r, cache=TRUE}
 # GWAS
@@
 
-```{r, dev='png', fig.asp=0.8}
+```{r, dev='png', fig.asp=0.8, cache=TRUE}
 # Q-Q plot
@@
 
-```{r, dev="png", fig.asp=0.9, out.width="90%"}
+```{r, dev="png", fig.asp=0.9, out.width="90%", cache=TRUE}
 # Manhattan plot
@@
 
-```{r}
+```{r, cache=TRUE}
 # Clumping on the test set
@@
 
-```{r}
+```{r, cache=TRUE}
 # PRS
@@
 plot_density_AUC <- function(pred, true = y[ind.test]) {
-  data.frame(pred = pred, pheno = as.factor(true)) %>%
+  true_fct <- forcats::fct_recode(as.factor(true), "Control" = "0", "Case" = "1")
+  data.frame(Score = pred, Status = true_fct) %>%
   myggplot() +
-  geom_density(aes(pred, fill = pheno), alpha = 0.4) +
+  geom_density(aes(Score, fill = Status), alpha = 0.4) +
   ggtitle(sprintf("AUC = %s%% | pAUC = %s", 
@@
 
-```{r}
+```{r, cache=TRUE}
 # Train logit-simple
@@
 
-```{r}
+```{r, cache=TRUE}
 G2 <- big_attach("backingfiles/celiacQC_tripled1.rds")

### 2018-03-19 paper/draft.tex
add one ref
add ref dudbridge

@@
 
-In order to simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% [CITATION?].
+In order to simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}.
 We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only chromosome 6), the size of the training set ($\{1000, 2000, \dots, 6000\}$), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part/component/term..
@@
 Reducing the number of SNPs aims at increasing the polygenicity of the simulated models and at ..virtually/artificially..
-increasing the sample size [REF?
-DUDBRIDGE?].
+increasing the sample size \cite[]{dudbridge2013power}.
 For this scenario, we use only the additive (``simple'') model, but continue to compare all previous variations of the other parameters.

### 2018-03-19 plotPCs.R
save figures

@@ -1,0 +1,25 @@
+library(bigsnpr)
+obj.svd <- readRDS("backingfiles/PCA.rds")
+plot(obj.svd, type = "scores")
+# Get population
+celiac <- snp_attach("backingfiles/celiacQC.rds")
+
+library(ggplot2)
+p1 <- plot(obj.svd, type = "scores", scores = 1:2, coeff = 1.2) + 
+  aes(color = celiac$fam$family.ID) +
+  labs(color = "Population") +
+  guides(colour = guide_legend(override.aes = list(size = 4)))
+p2 <- plot(obj.svd, type = "scores", scores = 3:4, coeff = 1.2) + 
+  aes(color = celiac$fam$family.ID) +
+  labs(color = "Population")
+
+library(magrittr)
+library(cowplot)
+plot_grid(
+  plotlist = lapply(list(p1, p2), function(p) {
+    p + theme(legend.position = "none") + ggtitle(NULL)
+  })
+) %>%
+  plot_grid(get_legend(p1), rel_widths = c(1, 0.15), scale = 0.95)
+  
+ggsave("../thesis-docs/figures/PC1-4.png", scale = 1/90, width = 1270, height = 610)

### 2018-03-19 paper/draft.tex
add svg figures for pres

@@
 The last method ..in/of.. this comparison is a penalized logistic regression.
-We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that needs to be determined.
+We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda, \alpha)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that needs to be determined.
 Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);

### 2018-03-19 results-simus-pres.Rmd
figures for pres

--- 2018-02-27 4-results-simus.Rmd
@@
 author: "Florian Privé"
-date: "November 17, 2017"
+date: "March 19, 2018"
 output: 
@@
     code_folding: hide
-params:
-  ggsave: false
 ---
@@
     as_tibble() %>%
+    filter(par.h2 == 0.8, par.dist == "gaussian") %>%
     mutate(
@@
 
-barplot_causal_one <- function(results, h2) {
-
-  auc_max <- `if`(h2 == 0.8, 0.94, 0.84)
-
-  results %>%
-    filter(par.h2 == h2) %>%
-    barplot_causal() +
-    geom_hline(yintercept = auc_max, linetype = 3, color = "blue") +
-    facet_grid(par.dist ~ .)
-}
-
-barplot_causal_all <- function(results) {
+barplot_causal_one <- function(results) {
 
@@
     barplot_causal() +
-    geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
-               data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
-    facet_grid(par.dist ~ par.h2)
+    geom_hline(yintercept = 0.94, linetype = 3, color = "blue") +
+    facet_grid(par.dist ~ .)
 }
@@
 
-  results <- filter(
-    results, method %in% c("logit-simple", with_method), par.h2 == 0.8)
+  results <- filter(results, method %in% c("logit-simple", with_method))
 
@@
     group_by_at(c(vars(starts_with("par")), "method")) %>%
-    summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
-    print(n = Inf)
+    summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n())
 
@@
                               fill = "method"), alpha = 0.3) + 
-      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-      facet_grid(par.model ~ par.dist) +
+      theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
+      # facet_grid(par.model ~ par.dist) +
       labs(x = "Causal SNPs (number and location)", y = ylab, 
@@
   p1 <- results.summary %>%
-    barplot_causal_one(h2 = 0.8) +
+    barplot_causal_one() +
     facet_grid(par.model ~ par.dist) +
@@
     plot_extra("timing", "Execution time (in seconds)"),
-    ncol = 1, align = "hv", scale = 0.95, 
+    ncol = 1, align = "hv", scale = 0.90, 
     labels = LETTERS[2:3], label_size = 20
@@
 
-ggsave("figures/supp-ttrees.pdf", scale = 1/100, width = 1400, height = 1060)
+ggsave("figures/pres-ttrees.svg", scale = 1/100, width = 1300, height = 1060)
 ```
@@
   group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
-  print(n = Inf) 
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n())
 ```
@@
 
-ggsave("figures/supp-triple.pdf", scale = 1/100, width = 1400, height = 1060)
-```
-
-```{r}
-results2.summary %>%
-  filter(grepl("logit", method)) %>%
-  barplot_causal_one(h2 = 0.5) +
-  facet_grid(par.model ~ par.dist)
-
-ggsave("figures/supp-AUC-triple.pdf", scale = 1/110, width = 1242, height = 951)
-```
-
-### Results with model fancy
-
-```{r}
-results2.summary %>%
-  filter(par.model == "fancy", method %in% c("logit-simple", "PRS-max")) %>%
-  barplot_causal_all()
-
-ggsave("figures/supp-AUC-logit-fancy.pdf", scale = 1/110, width = 1242, height = 951)
-```
-
-```{r}
-results2.summary %>%
-  filter(par.model == "fancy", grepl("PRS", method)) %>%
-  barplot_causal_all()
-
-ggsave("figures/supp-AUC-PRS-fancy.pdf", scale = 1/110, width = 1242, height = 951)
+ggsave("figures/pres-triple.svg", scale = 1/100, width = 1300, height = 1060)
 ```
@@
 
-### Correlation between predictive measures
-
 ```{r}
@@
   filter(par.model == "simple", method != "logit-triple")
-```
-
-```{r, fig.asp=0.7, out.width="95%", fig.width=16}
-results3 %>%
-  filter(method %in% c("logit-simple", "PRS-max")) %>%
-  mutate(Parameters  = interaction(par.h2, par.dist, method, sep = " | ")) %>%
-  myggplot(aes(AUC, pAUC, color = Parameters)) + 
-  geom_point(size = 0.6, alpha = 0.5) +
-  geom_smooth(aes(linetype = Parameters), method = "loess", se = FALSE, size = 2) +
-  theme(legend.key.width = unit(3.5, "line"), 
-        legend.direction = "horizontal") + 
-  labs(y = "Partial AUC (specificity: [0.9, 1])") +
-  theme(legend.position = "top", axis.title = element_text(size = rel(2.2))) +
-
-ggsave("figures/supp-AUC-corr.pdf", scale = 1/100, width = 1525, height = 983)
-```
-
-```{r}
-cor(results3$AUC, results3$pAUC, method = "spearman")
 ```
@@
   group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
-  print(n = Inf)
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n())
 ```
@@
 results3.summary %>%
-  filter(method %in% c("logit-simple", "PRS-max"), par.model == "simple") %>%
-  barplot_causal_one(h2 = 0.8)
+  filter(method %in% c("logit-simple", "PRS-max")) %>%
+  barplot_causal_one()
 
-ggsave("figures/main-AUC-logit.pdf", scale = 1/90, width = 844, height = 872)
+ggsave("figures/pres-AUC-logit.svg", scale = 1/110, width = 1045, height = 770)
 ```
@@
   filter(grepl("PRS", method)) %>%
-  barplot_causal_one(h2 = 0.8)
+  barplot_causal_one()
 
-ggsave("figures/main-AUC-PRS.pdf", scale = 1/90, width = 844, height = 872)
-```
-
-```{r, fig.asp = 1, out.width = "60%", fig.width=8}
-results3.summary %>%
-  filter(method %in% c("logit-simple", "PRS-max")) %>%
-  barplot_causal_one(h2 = 0.5)
-
-ggsave("figures/supp-AUC-logit.pdf", scale = 1/90, width = 844, height = 872)
-```
-
-```{r, fig.asp = 1, out.width = "60%", fig.width=8}
-results3.summary %>%
-  filter(grepl("PRS", method)) %>%
-  barplot_causal_one(h2 = 0.5)
-
-ggsave("figures/supp-AUC-PRS.pdf", scale = 1/90, width = 844, height = 872)
+ggsave("figures/pres-AUC-PRS.svg", scale = 1/110, width = 1045, height = 770)
 ```
@@
   group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
-  print(n = Inf)
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n())
 
@@
 results34 %>%
+  filter(par.h2 == 0.8, par.dist == "gaussian", method %in% c("logit-simple", "PRS-max")) %>%
   myggplot() +
   geom_hline(yintercept = 0.5, linetype = 2) +
-  geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
-             data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
+  geom_hline(aes(yintercept = 0.94), linetype = 3, color = "blue") +
   geom_boxplot(aes(method, AUC, fill = simu, color = simu), alpha = 0.3) + 
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  facet_grid(par.model ~ par.causal) +
-  facet_grid(par.h2 + par.dist ~ par.causal) +
+  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
+  facet_grid(par.dist ~ par.causal) +
   scale_fill_brewer(type = "qual", palette = 2) +
@@
 
-ggsave("figures/supp-AUC-chr6.pdf", scale = 1/110, width = 1153, height = 908)
+ggsave("figures/pres-AUC-chr6.svg", scale = 1/110, width = 990, height = 700)
 ```
@@
   group_by_at(c(vars(starts_with("par")), "method", "n.train")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
-  print(n = Inf)
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n())
 
@@
 ```{r}
-myggplot(results35.summary, aes(n.train, AUC_mean, color = method, linetype = method)) +
+results35.summary %>%
+  filter(method %in% c("logit-simple", "PRS-max")) %>%
+  myggplot(aes(n.train, AUC_mean, color = method, linetype = method)) +
   geom_hline(yintercept = 0.5, linetype = 2) +
-  geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
-             data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
+  geom_hline(aes(yintercept = 0.94), linetype = 3, color = "blue") +
   geom_line(size = 1.5) +
@@
                 size = 1.5, width = 0) +
-  facet_grid(par.h2 ~ par.dist) +
+  facet_grid(. ~ par.dist) +
   scale_x_continuous(breaks = 1:6 * 1000, minor_breaks = NULL) + 
@@
 
-ggsave("figures/main-AUC-ntrain.pdf", scale = 1/100, width = 1050, height = 800)
+ggsave("figures/pres-AUC-ntrain.svg", scale = 1/100, width = 870, height = 635)
 ```

### 2018-03-21 paper/draft.tex
first changes Mike

@@
 
-\title{Predicting complex diseases: performance and robustness [OF NEW METHOD?]}
+\title{Joint estimation of SNP effects improves the prediction of complex diseases // Predicting complex diseases: performance and robustness [OF NEW METHOD?]}
 \author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
@@
 Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
-PRSs have proven to be useful for different applications, such as finding a common genetic contribution between two diseases \cite[]{purcell2009common}.
+PRSs are useful for different applications, such as testing polygenicity of a disease and finding a common genetic contribution between two diseases \cite[]{purcell2009common}.
 Personalized medicine is another major application of PRSs.
-Personalized medicine ..will..
-use PRSs in screening campaigns in order to identify high-risk individuals for a given disease.
+Personalized medicine envisions to use PRSs in screening campaigns in order to identify high-risk individuals for a given disease [CITE CHATTERJEE 2016].
 As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}.
@@
 Many methods have been developed to predict disease status, or more generally any phenotype, based on SNP information.
-A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- also sometimes called genetic profiling ..or even just PRS..
--- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
-This technique only ..uses/resorts to.. summary statistics, which makes it very ..suitable.. and also very fast.
+A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or even just PRS -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
+This technique uses GWAS summary statistics only, which makes it very convenient to compute and also very fast.
 Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}.
-Yet, these models are not designed for predicting a binary trait such as a disease status and have proven not to be optimal at such ..a..
-task in another comparative study \cite[]{abraham2013performance}.
+Yet, the predictions resulting from LMM, known e.g.\ as gBLUP, are not designed for binary traits such as a disease status and are not optimal in this setting \cite[]{abraham2013performance}. 
 Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time requirements, which makes them unlikely to be used for prediction on large cohorts \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
-Finally, statistical learning methods have also been used to derive PRSs for complex human disease by jointly estimating SNP effects.
+Finally, statistical learning methods have also been used to derive PRSs for complex human diseases by jointly estimating SNP effects.
 Such methods include logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2013walk,botta2014exploiting,wei2009disease}.
 
+[REFORMULER EN COMMENCANT PAR LA REGRESSION LOGISTIQUE] -> apport principal du papier (+ la meme dans l'abstract)
 We recently developed two R packages, bigstatsr and bigsnpr, for efficient analysis of large-scale genome-wide data \cite[]{prive2017efficient}.
 Package bigstatsr includes efficient functions for computing sparse linear and logistic regressions on huge datasets.
-In this paper, we present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in \cite{botta2014exploiting}.
+In this paper, we present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
 Note that in this comparison we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
@@
 For the logistic regression, we include two novel approaches. 
-First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for automatically choosing the amount of regularization used (one hyper-parameter of the penalized logistic regression), which directly affects the number of SNPs included in the model.
+First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for automatically choosing the regularization parameter of the penalized logistic regression, which directly affects the number of SNPs included in the model.
 We also show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.
@@
 There have already been comparative studies on predicting human binary phenotypes based on genotype data. 
-\citep{abraham2013performance} showed that penalized logistic regression and penalized SVM performed better than the standard C+T method and LMMs, for a wide-range of human diseases.
+\cite{abraham2013performance} showed that penalized logistic regression and penalized SVM performed better than the standard C+T method and LMMs, for a wide-range of human diseases.
 \cite{zhou2013polygenic} used simulations on real genotypes of Australian individuals to show that a hybrid of linear mixed models and sparse regression models performed better in a range of sparse to polygenic disease architectures.
@@
 \cite{ware2017heterogeneity} evaluated best practices for using the C+T method and found out that the optimal inclusion threshold is trait-dependent. 
-In this study, we particularly extend the work of \cite{abraham2013performance} by using a simulation framework similar but ..richer/more extenstive..
-than the one used in \cite{zhou2013polygenic}.
-In order to make our comparison as comprehensive as possible, we compare different disease architectures (number, size and location of causal effects and heritability) with different model for generating liability ..scores.., one with only linear effects, and one which combines linear, dominant and ..interaction/interaction-type..
-effects.
-First, we quickly discard T-Trees as being the ..best..
-method for predicting disease status based on genotypes.
+In this study, we particularly extend the work of \cite{abraham2013performance} [RELIRE ET PRECISER] while using a simulation framework similar but more extenstive than the one used in \cite{zhou2013polygenic}.
+In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the heritability.
+We also compare different models for generating liability scores, one with only linear effects, and one which combines linear, dominant and interaction-type effects.
+
+First, we quickly discard T-Trees because they show lower predictive performance as compared to other methods, while being computationally demanding.
 Then, we show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture, as shown in \cite{ware2017heterogeneity}.
@@
 Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
-As for simulations presented here, in order to remove the structure induced by the celiac disease status, we ..first keep only/select/restrict/retain.. controls from this cohort. 
-Secondly, in order to ..alleviate..
-population structure, we ..keep only..
-people from the UK.
-Thirdly, we further subset these British people based on deviation from the robust Mahalanobis distance ..on/for..
-principal components.
+As for simulations presented here, in order to remove the genetic structure induced by the celiac disease status, we first restrict our simulations to controls. 
+Then, we decided to remove population structure because it can affect the predictive performance of methods [REF, MARTINS?].
+In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on the robust Mahalanobis distances computed using the first 10 principal components of remaining individuals.
 This 3-step filtering process results in a sample of 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing'').
-Note that we decided to ..alleviate.. population structure because it can affect the predictive performance of methods, which we will investigate in another study.
 
@@
 In order to simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}.
-We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only chromosome 6), the size of the training set ($\{1000, 2000, \dots, 6000\}$), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part/component/term..
-of the phenotypes (a ``simple'' model with only linear effects, and a ``fancy'' model which combines linear, dominant and ..interaction/interaction-type..
-effects).
-For the ``simple'' model, we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights (generated with a Gaussian or a Laplace distribution), $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs).
+We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only chromosome 6), the size of the training set ($\{1000, 2000, \dots, 6000\}$), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace).
+We also use two different models to compute liability scores: a ``simple'' model with linear effects only, and a ``fancy'' model which combines linear, dominant and interaction-type effects.
+For the ``simple'' model, we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights generated with a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs).
 For the ``fancy'' model, we separate the causal SNPs in three equal sets $E_\text{causal}^{(1)}$, $E_\text{causal}^{(2)}$ 
@@
 where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $E_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|E_\text{causal}^{(q)}\right|\right\}\right\}$. 
-Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still ..display/have.. a marginal effect. 
-For both models, the ..$y_i$/values of $y_i$..
-are then standardized to make sure they have a variance exactly equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$, where $\epsilon \sim N(0, 1 - h^2)$.
+Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still have a marginal effect. 
+For both models, the values of $y_i$ are then standardized so that they have a variance equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$, where $\epsilon \sim N(0, 1 - h^2)$.
 
-We ..make/construct/design/implement/work with.. 3 different simulation scenarios, summarized in table \ref{tab:simus}. 
+We implement 3 different simulation scenarios, summarized in table \ref{tab:simus}. 
 Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes) and a training set of size 6000. 
-It compares all methods described in section \ref{sec:methods} ..on/for/along/in relation to..
-all other combinations of parameters in 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}).
-Note that the T-Trees method is compared on only 5 simulations and an heritability of 80\%, which aims at quickly discarding this method that can require more than one hour of computation for a single simulation only.
-Scenario \textnumero2 consists of 100 simulations comparing the ``logit-simple'' and the C+T methods on a dataset composed of chromosome 6 only.
-Reducing the number of SNPs aims at increasing the polygenicity of the simulated models and at ..virtually/artificially..
-increasing the sample size \cite[]{dudbridge2013power}.
+It compares all methods described in section \ref{sec:methods}; for each combination of the remaining parameters, results are based on 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}).
+However, comparison results for T-Trees relies on 5 simulations only (and an heritability of 80\%) because this method can require several hours of computation for a single simulation only. 
+Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only.
+Reducing the number of SNPs aims at increasing the polygenicity (the proportion of causal SNPs) of the simulated models and at virtually increasing the sample size \cite[]{dudbridge2013power}.
 For this scenario, we use only the additive (``simple'') model, but continue to compare all previous variations of the other parameters.
-Finally, scenario \textnumero3 reuses the whole dataset but this time varying the size of the training set in order to assess how the sample size affects predictive performance of the PRS and the ``logit-simple'' methods.
-100 simulations are run using 300 causal SNPs anywhere on the genome.
+Finally, scenario \textnumero3 reuses the whole dataset but this time varying the size of the training set in order to assess how the sample size affects predictive performance of methods.
+A total of 100 simulations are run using 300 causal SNPs anywhere on the genome.
 
@@
 As a second measure, the partial AUC for specificities between 90\% and 100\% is also reported \cite[]{dodd2003partial,mcclish1989analyzing}.
-This measure is similar to the AUC, but focuses on high specificities (i.e.\ discarding most of the healthy population as being diseased), which is often the most useful part of the ROC curve in clinical settings.
+This measure is similar to the AUC, but focuses on high specificities, which is the most useful part of the ROC curve in clinical settings.
 
@@
 
-In this study, we compare three different types of methods: the C+T method, the T-Trees method and a penalized logistic regression.
+In this study, we compare three different types of methods: the C+T method, T-Trees and penalized logistic regression.
 

### 2018-03-22 paper/draft.tex
Mike's changes

@@
 \newpage
-[ABSTRACT TROP LONG?]
+[ABSTRACT TROP LONG?]Yet, current PRSs are far from reaching this goal, and would have to show a high discriminative power between cases and controls in order to be of any use.
+The predictive power of PRS strongly relies on the underlying technique used to include SNPs in the PRS and to estimate their relative contribution.
+
 \abstract{
@@
 PRSs can have a major public health impact, allowing for screening campaigns to identify high-genetic risk individuals for a given disease.
-Yet, current PRSs are far from reaching this goal, and would have to show a high discriminative power between cases and controls in order to be of any use.
+The most commonly used method to derive PRSs is called ``Clumping+Thresholding'' (C+T).
+It uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it very suitable and fast.
+However, PRS rely on non-optimal heuristics to choose various numerical thresholds.
 
-The predictive power of PRS strongly relies on the underlying technique used to include SNPs in the PRS and to estimate their relative contribution.
-The most commonly used method to derive PRSs is called ``Clumping+Thresholding'' (C+T).
-It uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it very suitable and also very fast.
-We recently developed two R packages, bigstatsr and bigsnpr, for the analysis of large-scale genome-wide data.
-Package bigstatsr includes efficient functions for computing penalized linear and logistic regressions on large datasets.
+We implement a logistic regression model that is suited for large scale genomic data and which provides a joint estimation of SNP effects.
+The R Package bigstatsr includes efficient functions to train penalized logistic regression on large datasets.
 
-We present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in a previous study.
-We quickly discard the T-Trees method and show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture, as shown in previous studies.\\
+We present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests.
+Penalized logistic regression consistently achieves higher predictive performance than T-Trees while being as fast as the C+T method.
+Predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs which values depends on disease architecture.
+Penalized logistic regression consistently achieves superior performance than the C+T method and this improvement was more pronounced when there are few effects located in nearby genomic regions such as HLA.
+When analysing SNP data from a case-control study,  penalized logistic regression provides an AUC of $89\%$ while the standard C+T method provides a smaller $AUC$ of $82\%$.
+
+Instead of computing polygenic risk scores based on GWAS univariate summary statistics, we recommend to use penalized logistic regression as implemented in the R package bigsnpr to obtain more accurate risk scores.
+
+\\
 \textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
@@
 We also use two different models to compute liability scores: a ``simple'' model with linear effects only, and a ``fancy'' model which combines linear, dominant and interaction-type effects.
-For the ``simple'' model, we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights generated with a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs).
-For the ``fancy'' model, we separate the causal SNPs in three equal sets $E_\text{causal}^{(1)}$, $E_\text{causal}^{(2)}$ 
-and $E_\text{causal}^{(3)}$ ($E_\text{causal}^{(3)}$ is further separated in two equal sets, $E_\text{causal}^{(3.1)}$ and $E_\text{causal}^{(3.2)}$). 
-We then compute $$y_i = \underbrace{\sum_{j\in E_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in E_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|E_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$
-where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $E_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|E_\text{causal}^{(q)}\right|\right\}\right\}$. 
+For the ``simple'' model, we compute $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights generated with a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs).
+For the ``fancy'' model, we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$ 
+and $S_\text{causal}^{(3)}$ ($S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$). 
+We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$
+where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$. 
 Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still have a marginal effect. 
@@
 For the C+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
-The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another. 
-Then, thresholding (T) consists in removing SNPs that are within a certain level of significance (p-value threshold to be determined). 
-Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$S_i(T, E_\text{clumping}) = \sum_{j \in E_\text{clumping}} \mathds{1}\{p_j < T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS and $T$ is a threshold on p-values to be determined.
+The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another, which form a set of SNPs denoted $S_\text{clumping}$. 
+Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $t$ to be determined. 
+Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$\rm{PRS}_i = \sum_{j \in S_\text{clumping}} \mathds{1}\{p_j < t\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS.
 In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.
@@
 For the C+T method, we report three different scores of prediction: one including all the SNPs (remaining after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximizes the AUC for these two thresholds and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
-Note that, normally, this optimal threshold would have to be learned based on the training set only in order to prevent from over-fitting the data.
-We consider this reported maximum AUC as an upper bound of the AUC for the C+T method.
+As we report the optimal threshold based on the test set, the corresponding AUC value is an upper bound of the AUC for the C+T method.
 We call these three reported scores respectively ``PRS-all'', ``PRS-stringent'' and ``PRS-max'' in the results.
@@
 T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}.
-Since this method is ..significantly..
-different from the other ones presented here and give that it achieved surprisingly high predictive performance in the past, we decided to include T-Trees in this comparison.
-We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees only (instead of 1000) because ..it..
-showed only a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown).
+We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees instead of 1000 because because using 1000 trees provides a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown).
 %We call this method ``T-Trees'' in the results.
 
-The last method ..in/of.. this comparison is a penalized logistic regression.
-We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda, \alpha)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that needs to be determined.
+Finally, for the penalized logistic regression, we solve: $$\argmin_{\beta_0, \beta}(\lambda, \alpha)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that needs to be determined.
 Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
-the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be exactly ..zero/equal to zero..
-and can be used as a means of variable selection, leading to sparse (and therefore more interpretable) models  (corresponds to $\alpha = 1$);
+the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse and therefore more interpretable models (corresponds to $\alpha = 1$);
 the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
-In this study, we always use the elastic-net regularization with $\alpha = 0.5$, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.
+In this study, we use elastic-net regularization with $\alpha = 0.5$.%, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.
 
@@
 First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds).
-Secondly, in turn, each fold is considered as an inner validation set and the other ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes the likelihood on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is selected.
-Finally, the $K$ resulting vectors of coefficients are reported and can be combined into one vector (e.g.\ by using the geometric median) or resulting scores can be combined (e.g.\ by averaging the $K$ probabilities for each individual).
+Secondly, in turn, each fold is considered as an inner validation set and the other ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which minimizes the log loss on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is selected.
+Finally, the $K$ resulting probabilities for each individual are averaged.
 We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making this procedure much faster.
@@
 In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects.
-This results in a dataset with 3 times as many variables as the initial one, on which we can apply the penalized logistic regression with the CMSA procedure, as described previously.
+This results in a dataset with 3 times as many variables as the initial one, on which we can apply penalized logistic regression with the CMSA procedure, as described previously.
 We call this method ``logit-triple'' in the results.

### 2018-03-23 paper/draft.tex
new changes

@@
 
+[REFAIRE OU METTRE AUTRE PART]
 When reporting results, we use estimates of maximum achievable AUC values for the simulations: 84\% (94\%) for heritabilities of 50\% (80\%) based on three estimations (see Supplementary Materials).
@@
 
-\subsection{Scenario \textnumero1 -- main methods}
+\subsection{Comparison between C+T and ``logit-simple'' methods}
 
-In scenario \textnumero1, we compared our penalized logistic regression (``logit-simple'') with the C+T method (``PRS'') on the whole genome (Table \ref{tab:simus}). 
-Using the penalized logistic regression, we report AUCs greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, which is not obtained by the upper bound of the C+T method in any of the scenarios (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
-The penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted sometimes when we simulate 3000 causals SNPs anywhere on the genome, a case where all methods rarely exceeds 65\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
-In the case where there are correlations between predictors (i.e.\ when we simulate 30 SNPs in the HLA region), the ``logit-simple'' method performs particularly ..better..
-than the ``PRS-max'' one, with a mean AUC of 92.5\% versus 83\%, for a maximum achievable AUC of 94\% (Figure \ref{fig:main-AUC-logit}).
-We also report that the choice of the threshold of inclusion of SNPs in the C+T method can impact its performance greatly, for example dropping from an AUC of 90\% to less than 60\% when including all SNPs passing the clumping phase in a model with only 30 causal SNPs, or dropping from 60\% to a random predictor (AUC of 50\%) when including only GWAS-significant SNPs in a model with 3000 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
+We compared the penalized logistic regression (``logit-simple'') with the C+T method (``PRS'') using whole-genome simulations of scenario \textnumero1 (Table \ref{tab:simus}).
+Penalized logistic regression provides AUC greater than 93\% for some disease architectures, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method are always lower than 90\% (Figures \ref{fig:main-AUC-logit}).
+This situation occurs when simulating 30 causal SNPs in a linear fashion with an heritability of 80\%.
+Moreover, the penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted in simulations with 3000 causal SNPs anywhere.
+Yet, these are cases where all methods rarely exceeds 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
+
+Method ``logit-simple'' provides particularly higher predictive performance than ``PRS-max'' where there are correlations between predictors, i.e.\ when we simulate 30 SNPs in the HLA region.
+In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' and 83\% with the ``PRS-max'', for a maximum achievable AUC of 94\% (Figure \ref{fig:main-AUC-logit}).
+
+In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can impact its predictive performance greatly.
+For example, AUC ranges from less than 60\% to 90\% in a model with only 30 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
 

### 2018-03-26 paper/draft.tex
other modifs

@@
 It uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it very suitable and fast.
-However, PRS rely on non-optimal heuristics to choose various numerical thresholds.
+However, PRS rely on non-optimal heuristics to account for correlation and reduce noise in the predictor.
 
-We implement a logistic regression model that is suited for large scale genomic data and which provides a joint estimation of SNP effects.
-The R Package bigstatsr includes efficient functions to train penalized logistic regression on large datasets.
+In R package bigstatsr, we implemented a logistic regression model that provides a joint estimation of SNP effects.
+We used an efficient algorithm with new implementation tricks in order for this model to be used on huge datasets as large as the UK Biobank.
 
-We present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests.
-Penalized logistic regression consistently achieves higher predictive performance than T-Trees while being as fast as the C+T method.
-Predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs which values depends on disease architecture.
-Penalized logistic regression consistently achieves superior performance than the C+T method and this improvement was more pronounced when there are few effects located in nearby genomic regions such as HLA.
-When analysing SNP data from a case-control study,  penalized logistic regression provides an AUC of $89\%$ while the standard C+T method provides a smaller $AUC$ of $82\%$.
+In this paper, we present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests.
+The penalized logistic regression consistently achieves higher predictive performance than the two other methods while being faster than the C+T method.
+Moreover, predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, whose optimal value depends on disease architecture. 
+The improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs. 
+When analyzing real SNP data from a case-control study,  penalized logistic regression provides an AUC of 89\% while the standard C+T method provides a smaller AUC of 82\%.
 
-Instead of computing polygenic risk scores based on GWAS univariate summary statistics, we recommend to use penalized logistic regression as implemented in the R package bigsnpr to obtain more accurate risk scores.
+Instead of computing polygenic risk scores based on GWAS univariate summary statistics, we recommend to use penalized logistic regression as implemented in the R package bigstatsr to obtain more accurate polygenic risk scores.\\
 
-\\
 \textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
@@
 Personalized medicine is another major application of PRSs.
-Personalized medicine envisions to use PRSs in screening campaigns in order to identify high-risk individuals for a given disease [CITE CHATTERJEE 2016].
+Personalized medicine envisions to use PRSs in screening campaigns in order to identify high-risk individuals for a given disease \cite[]{chatterjee2016developing}.
 As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}.
@@
 
-[REFORMULER EN COMMENCANT PAR LA REGRESSION LOGISTIQUE] -> apport principal du papier (+ la meme dans l'abstract)
 We recently developed two R packages, bigstatsr and bigsnpr, for efficient analysis of large-scale genome-wide data \cite[]{prive2017efficient}.
-Package bigstatsr includes efficient functions for computing sparse linear and logistic regressions on huge datasets.
-In this paper, we present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
-Note that in this comparison we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
+Package bigstatsr includes an efficient algorithm with new implementation tricks for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
+In this paper, we present a comprehensive comparative study of the C+T method, our implementation of penalized logistic regression and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
+Note that in this comparison we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp} [<- IN METHODS?
+MAIS BIEN AUSSI DE DIRE PQ TOT..].
 For the C+T model, we compare different thresholds of inclusion of SNPs.
@@
 There have already been comparative studies on predicting human binary phenotypes based on genotype data. 
-\cite{abraham2013performance} showed that penalized logistic regression and penalized SVM performed better than the standard C+T method and LMMs, for a wide-range of human diseases.
+\cite{abraham2013performance} showed that sparse penalized methods performed better than the standard C+T method and LMMs, for a wide-range of human diseases.
 \cite{zhou2013polygenic} used simulations on real genotypes of Australian individuals to show that a hybrid of linear mixed models and sparse regression models performed better in a range of sparse to polygenic disease architectures.
@@
 \cite{ware2017heterogeneity} evaluated best practices for using the C+T method and found out that the optimal inclusion threshold is trait-dependent. 
-In this study, we particularly extend the work of \cite{abraham2013performance} [RELIRE ET PRECISER] while using a simulation framework similar but more extenstive than the one used in \cite{zhou2013polygenic}.
+In this study, we particularly extend the work of \cite{abraham2013performance} [RELIRE ET PRECISER] by using a simulation framework similar but more extenstive than the one used in \cite{zhou2013polygenic}.
 In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the heritability.
@@
 As for simulations presented here, in order to remove the genetic structure induced by the celiac disease status, we first restrict our simulations to controls. 
-Then, we decided to remove population structure because it can affect the predictive performance of methods [REF, MARTINS?].
+Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.
 In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on the robust Mahalanobis distances computed using the first 10 principal components of remaining individuals.
@@
 In order to simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}.
-We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only chromosome 6), the size of the training set ($\{1000, 2000, \dots, 6000\}$), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace).
+We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only chromosome 6), the size of the training set (from 1000 to 6000), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace).
 We also use two different models to compute liability scores: a ``simple'' model with linear effects only, and a ``fancy'' model which combines linear, dominant and interaction-type effects.
@@
 The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another, which form a set of SNPs denoted $S_\text{clumping}$. 
-Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $t$ to be determined. 
-Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$\rm{PRS}_i = \sum_{j \in S_\text{clumping}} \mathds{1}\{p_j < t\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS.
+Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $p_T$ to be determined. 
+Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$\rm{PRS}_i = \sum_{j \in S_\text{clumping}} \mathds{1}\{p_j < p_T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS.
 In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.
@@
 
+\subsection{Evaluating predictive performance for Celiac data}
+
+We use Monte Carlo cross-validation to compute AUC, partial AUC, number of predictors used and execution time for the original Celiac dataset with real phenotypes.
+We randomly split 100 times the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals.
+
 \subsection{Reproduciblity}
@@
 We compared the penalized logistic regression (``logit-simple'') with the C+T method (``PRS'') using whole-genome simulations of scenario \textnumero1 (Table \ref{tab:simus}).
-Penalized logistic regression provides AUC greater than 93\% for some disease architectures, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method are always lower than 90\% (Figures \ref{fig:main-AUC-logit}).
-This situation occurs when simulating 30 causal SNPs in a linear fashion with an heritability of 80\%.
-Moreover, the penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted in simulations with 3000 causal SNPs anywhere.
-Yet, these are cases where all methods rarely exceeds 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
+
+When there are 30 causal SNPs and 80\% of heritability, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}).
+Moreover, the penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted in simulations with 3000 causal SNPs.
+Yet, these are cases where all methods hardly exceed 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
 
@@
 
-In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can impact its predictive performance greatly.
+In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can impact its predictive performance.
 For example, AUC ranges from less than 60\% to 90\% in a model with only 30 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
 
-\subsection{``Fancy'' methods and model}
+\subsection{Other methods and ``fancy'' model}
 
-First, in scenario \textnumero1, we also tested the T-Trees method. 
-For example, as compared to the ``logit-simple'' method, T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}).
-Even when simulating a fancy model in which there are dominant and ..interaction/interaction-type..
-effects that decision trees based methods should be able to handle, AUC is still lower with T-Trees as compared with the ``logit-simple'' method.
+We tested the T-Trees method in scenario \textnumero1. 
+As compared to the ``logit-simple'' method, T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}).
+Even when simulating a fancy model in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower with T-Trees as compared with the ``logit-simple'' method.
 
-Secondly, as for the comparison between the two penalized logistic regressions in scenario \textnumero1, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
+Secondly, we compare the two penalized logistic regressions in scenario \textnumero1, ``logit-simple'' and ``logit-triple'' that uses additional features coding for recessive and dominant effects. 
+We report that ``logit-triple'' is nearly as good as ``logit-simple'' when there are only linear effects (differences of AUC smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
+For the ``fancy model'', ``logit-triple'' provides AUC values at least 3.5\% higher than ``logit-simple'', except when there are 3000 causal SNPs.
 Yet, the ``triple'' solution takes 2-3 times as much time to run and requires 3 times as much disk storage as the ``logit-simple''.
 
-Thirdly, for the ``fancy'' simulation model, we report the exact same results as in the previous subsection that is to say that the ``logit-simple'' achieves greater AUCs than the ``PRS-max'', excepted in the case of 3000 causal SNP anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}).
-Moreover, the choice of the threshold inclusion of SNPs also greatly ..affects/impacts..
-the performance of the C+T method (Figure \ref{fig:supp-AUC-PRS-fancy}).
+Thirdly, for the ``fancy'' simulation model, we report similar results as in the previous subsection.
+As before, ``logit-simple'' achieves greater AUCs than the ``PRS-max'', excepted in the case of 3000 causal SNP anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}).
+Moreover, the choice of the threshold inclusion of SNPs also greatly impacts the performance of the C+T method (Figure \ref{fig:supp-AUC-PRS-fancy}).
 
@@
 These simulations correspond to scenarios \textnumero2 and \textnumero3 (Table \ref{tab:simus}). 
-First, when reproducing simulations of scenario \textnumero1 on only chromosome 6, methods globally increase or ..stagnate..
-in predictive performance (Figure \ref{fig:supp-AUC-chr6}).
-There is particularly a high increase in the case when simulating 3000 causal SNPs anywhere on the dataset: for example, AUC from the ``logit-simple'' can increase from less than 60\% to nearly 80\%.
-On the contrary, when simulating only 30 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' seems to slightly decrease.
-Secondly, when varying the training size, we report an increase of AUC when increasing the training size, with the AUCs provided by the ``logit-simple'' method increasing faster than for the ``PRS-max'' method (Figure \ref{fig:main-AUC-ntrain}).
+First, when reproducing simulations of scenario \textnumero1 using only chromosome 6, the predictive performance of all methods globally increase (Figure \ref{fig:supp-AUC-chr6}). 
+There is particularly a large increase when simulating 3000 causal SNPs anywhere in the dataset: AUC from the ``logit-simple'' increases from less than 60\% to nearly 80\%. 
+On the contrary, when simulating only 30 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' does not increase. 
+Secondly, when varying the training size, we report an increase of AUC when increasing the training size, with a faster increase of AUC provided by ``logit-simple'' as compared to ``PRS-max'' (Figure \ref{fig:main-AUC-ntrain}).
 
-\subsection{Real data results}
+\subsection{Polygenic scores for the celiac disease}
 
-We also analyze the whole Celiac dataset with the ..original/real..
-phenotypes.
-For 100 runs, randomly splitting the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals, we compare the predictive performance of the ``C+T'', ``logit-simple'' and ``logit-triple'' methods, but also computation times and number of predictors used in the models (Table \ref{tab:results-celiac}, figure \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
-We ..see..
-that the joint logistic regressions provide higher AUC with 88.8\% for the ``logit-simple'' and 89.2\% for the ``logit-triple'' as compared to 82.4\% for the ``C+T'' method, and much higher partial AUC (specificity 0.9-1), respectively 0.0414, 0.0429 and 0.0286, while using less predictors, respectively 3320, 4470 and 9850.
-Note that for the ``C+T'' method, we still report the best result among 102 p-value thresholds. 
-In terms of computation time, we report only the GWAS computation for the ``C+T'' method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once, is faster than the ``C+T'' method (84 / 148 seconds), and the ``logit-triple'' takes less than twice as long as the ``logit-simple'' (141 / 84 seconds).
-Finally, note that the scores obtained with the logistic regressions are probabilities, which is not the case with the ``C+T'' method (Figure \ref{fig:supp-score-densities}).
+Joint logistic regressions provide higher AUC for the Celiac data: 88.8\% with ``logit-simple'' and 89.2\% with ``logit-triple'' as compared to 82.4\% with the C+T method. 
+The relative differences in partial AUC, for specificities larger than 90\%, are even larger with values of 0.0414, 0.0429 and 0.0286 obtained with ``logit-simple'', ``logit-triple'' and the C+T method, respectively.
+Moreover, logistic regressions use less predictors, respectively 3320, 4470 and 9850 (Table \ref{tab:results-celiac}, figures \ref{fig:celiac-roc} and \ref{fig:supp-score-densities} and supplementary notebook ``results-celiac'').
+Note that for the C+T method, we still report the best result among 102 p-value thresholds. 
+In terms of computation time, we report only the GWAS computation for the C+T method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once, is faster than the C+T method (84 vs 148 seconds), and the ``logit-triple'' takes less than twice as long as the ``logit-simple'' (141 vs 84 seconds).
 
@@
 
-%We show that the regularization paths of both logistic regressions are both very smooth, reaching a maximum AUC or more than 89\% at approximately 1000 predictors (Figure \ref{fig:main-celiac-regpath}).
-Moreover, the CMSA procedure give AUC very close to the maximum AUC of the regularization path.
-For the C+T method, on the contrary, the regularization path is not smooth at all and the AUC never exceed the AUC given by the most stringent threshold [ADD VALUE OF AUC] ($p < 10^{-100}$).
-We also show some associated ROC curves [QUOI DIRE DE CES ROC?] (Figure \ref{fig:celiac-roc}).
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-In this comparative study, we propose other methods as a replacement for the widely used C+T method. 
-We show that jointly estimating the SNP effects using efficient penalized logistic regressions enables to achieve ..really..
-high predictive performance, as reported in a previous study \cite[]{abraham2013performance}.
-This is not the case with the C+T method that uses SNP effects learned independently.
-We are well aware that the C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions yet.
+In this comparative study, we propose penalized logistic regressions as a replacement for the widely used C+T method for computing polygenic risk scores. 
+We show that jointly estimating SNP effects using efficient penalized logistic regressions enables to achieve higher predictive performance, as reported in a previous study \cite[]{abraham2013performance}.
+Indeed, when disease architecture and sample size allow for high predictive performance (AUC greater than 85\%), ``logit-simple'' provides higher predictive performance than the widely-used C+T method, nearly reaching optimal AUC values, which is encouraging for using polygenic scores in clinical settings.
+The C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions yet. 
 However, in order to accurately predict disease status for clinical interventions, best possible methods should be used.
-The logistic regression we present in this paper could be used on data as large as biobank datasets because it is very efficient in both memory and computation time (..even faster than doing a GWAS!..).
+The logistic regression we present in this paper could be used on data as large as biobank datasets because it is very efficient in both memory and computation time.
 
 As for the C+T method itself, we recall that we report the maximum AUC of 102 thresholds, a threshold that should normally be learned on the training set only.
-The choice of this threshold is very important as it can greatly impact the predictive performance of the method.
+The choice of this threshold is very important as it can greatly impact the predictive performance of the method \cite[]{ware2017heterogeneity}.
 On the contrary, in the penalized logistic regression presented here, we remove the choice of the regularization parameter with a procedure that we call Cross-Model Selection and Averaging (CMSA) while keeping high predictive performance.
 
-We also introduce a simple feature engineering technique which enables the logistic regression to detect and learn not only additive effects, but also dominant and recessive ones. 
+We also introduce a simple feature engineering technique which enables the logistic regression to detect and learn not only additive effects, but also dominant and recessive ones [DEJA ETE FAIT?
+CITATION?].
 This technique improves the predictive performance of the method when there are some non-linear effects in the simulations, while providing the same predictive performance when there are only linear effects.
-Moreover, it also slightly improves predictive performance for the real disease we ..also..
-study in this paper.
-If this technique proves to be useful for predicting other diseases, it would be useful to re-implement another algorithm that could work directly on the initial dataset (e.g.
-by coding a new accessor).
+Moreover, it also improves predictive performance for the celiac disease.
 
-Finally, in this comparative study, we did not consider the problem of population structure and also did not consider non-genetic data.
+Finally, in this comparative study, we did not consider the problem of population structure \cite[]{martin2017human} and also did not consider non-genetic data such as environmental and clinical data [CITATION].
 In another study, we will assess how can we use ..training..

### 2018-03-29 paper/draft.tex
draft for Hugues

@@
 \newpage
-[ABSTRACT TROP LONG?]Yet, current PRSs are far from reaching this goal, and would have to show a high discriminative power between cases and controls in order to be of any use.
-The predictive power of PRS strongly relies on the underlying technique used to include SNPs in the PRS and to estimate their relative contribution.
 
@@
 \cite{ware2017heterogeneity} evaluated best practices for using the C+T method and found out that the optimal inclusion threshold is trait-dependent. 
-In this study, we particularly extend the work of \cite{abraham2013performance} [RELIRE ET PRECISER] by using a simulation framework similar but more extenstive than the one used in \cite{zhou2013polygenic}.
+In this study, we particularly extend the work of \cite{abraham2013performance} by using a simulation framework similar but more extenstive than the one used in \cite{zhou2013polygenic}. 
 In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the heritability.
@@
 It compares all methods described in section \ref{sec:methods}; for each combination of the remaining parameters, results are based on 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}).
-However, comparison results for T-Trees relies on 5 simulations only (and an heritability of 80\%) because this method can require several hours of computation for a single simulation only. 
+However, comparison results for T-Trees relies on 5 simulations only (and an heritability of 80\%) because this method can require several hours of computation for a single simulation. 
 Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only.
@@
 This measure is similar to the AUC, but focuses on high specificities, which is the most useful part of the ROC curve in clinical settings.
+
+When reporting results of AUC, we use estimates of maximum achievable AUC values for the simulations: 84\% and 94\% for heritabilities of respectively 50\% and 80\% based on three different estimations (see Supplementary Materials).
 
@@
 the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
-In this study, we use elastic-net regularization with $\alpha = 0.5$.%, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.
+In this study, we use the elastic-net regularization with $\alpha = 0.5$.%, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.
 
@@
 
-\subsection{Maximum AUCs}
-
-[REFAIRE OU METTRE AUTRE PART]
-When reporting results, we use estimates of maximum achievable AUC values for the simulations: 84\% (94\%) for heritabilities of 50\% (80\%) based on three estimations (see Supplementary Materials).
-Note that for the simulations we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).
-
 \subsection{Comparison between C+T and ``logit-simple'' methods}
@@
 For example, AUC ranges from less than 60\% to 90\% in a model with only 30 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
+
+Note that for the simulations we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).
 
@@
 The C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions yet. 
-However, in order to accurately predict disease status for clinical interventions, best possible methods should be used.
+However, in order to accurately predict disease status for clinical interventions, best possible predictive methods should be used.
 The logistic regression we present in this paper could be used on data as large as biobank datasets because it is very efficient in both memory and computation time.
@@
 
-Finally, in this comparative study, we did not consider the problem of population structure \cite[]{martin2017human} and also did not consider non-genetic data such as environmental and clinical data [CITATION].
-In another study, we will assess how can we use ..training..
-in one population to improve training and prediction in another population.
+Finally, in this comparative study, we did not consider the problem of population structure \cite[]{martin2017human} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}.
+In another study, we will assess how can we use models and effects learned in one population to improve learning and prediction in another population.
 We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}.
-In a ..final..
-study, we will also assess how can we combine the information provided by genetic data with clinical and environmental data, possibly in a non-linear way.
+In another study, we wish also to assess how can we combine the information provided by genetic data with clinical and environmental data, possibly in a non-linear way.
 
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
-\caption{Scenario \textnumero1.
-Mean of AUC over 100 simulations for our penalized logistic regression (``logit-simple'') and the maximum AUC reported with the C+T method (``PRS-max'').
+\caption{Main comparison of the C+T method and ``logit-simple'' in scenario \textnumero1.
+Mean of AUC over 100 simulations for ``logit-simple'' and the maximum AUC reported with the C+T method (``PRS-max'').
 Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
@@
 The blue dotted line represents the maximum achievable AUC.
-[MODEL AND H2]}
+These results corresponds to the ``simple'' model and an heritability of 80\%.}
 \label{fig:main-AUC-logit}
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-PRS}}
-\caption{Scenario \textnumero1.
-Mean of AUC over 100 simulations for ..the three reported PRSs...
+\caption{Comparison of three different threshold used in the C+T method in scenario \textnumero1.
+Mean of AUC over 100 simulations.
 Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
@@
 The blue dotted line represents the maximum achievable AUC.
-[MODEL AND H2]}
+These results corresponds to the ``simple'' model and an heritability of 80\%.}
 \label{fig:main-AUC-PRS}
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
-\caption{Scenario \textnumero3.
+\caption{Comparison of models when varying sample size in scenario \textnumero3.
 Mean of AUC over 100 simulations for ..the three reported PRSs...
-and the simple logistic regression as a function of the training size.
-This results are for an additive model with 300 causal SNPs sampled anywhere on the genome.
+and the ``logit-simple'' as a function of the training size.
+These results are for the ``simple'' model with 300 causal SNPs sampled anywhere on the genome.
 Upper (lower) panels are presenting results for an heritability of 0.5 (0.8) and left (right) panels are presenting results for effets following a Gaussian (Laplace) distribution.
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.
-[MODEL AND H2]}
+The blue dotted line represents the maximum achievable AUC.}
 \label{fig:main-AUC-ntrain}
@@
 We report AUCs of 84.1\% and 94.1\% for respectively an heritability of 50\% and 80\%.
-[HOW TO ADD ERROR?
-I HAVE SUMMARY()]
+[HOW TO ADD ERROR OF ESTIMATION?
+I HAVE SUMMARY() IN R]
 Thirdly, we reproduce the exact same procedure of simulations and, for each combination of parameters (Table \ref{tab:simus}), we estimate the AUC of the ``oracle'', i.e.\ the true simulated genetic part of the liabilities though 100 replicates.
@@
 \end{tabular}
-\caption{Number of individuals by population and disease status for the celiac disease cohort (after quality control, genotyped on 281,122 ..mutual.. SNPs).\label{tab:celiac-data}}
+\caption{Number of individuals by population and disease status in the celiac disease case-control study (after quality control, genotyped on 281,122 SNPs).\label{tab:celiac-data}}
 \end{table}
@@
 \end{tabular}
-\caption{The 102 thresholds used in the C+T method for this study.\label{tab:thr}}
+\caption{The 102 thresholds used for the C+T method for this study.\label{tab:thr}}
 \end{table}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6}}
-\caption{Scenario \textnumero2. Boxplots of AUC over 100 simulations for ..the three reported PRSs... and the simple logistic regression.
+\caption{Comparison of models when using only chromosome 6 in scenario \textnumero2. Boxplots of AUC over 100 simulations for ..the three reported PRSs... and the ``logit-simple''.
 Vertical panels represents different number of causal SNPs and their location (``all'': anywhere on the genome; ``HLA'': only in the HLA region).
@@
 The blue dotted line represents the maximum achievable AUC.
-[MODEL AND H2]}
+The ``simple'' model was used to simulate phenotypes.}
 \label{fig:supp-AUC-chr6}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-ttrees}}
-\caption{[T-TREES 3 PLOTS - H2=0.8]}
+\caption{Comparison of T-Trees and ``logit-simple'' in scenario \textnumero1.
+Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
+Horizontal panels are presenting results for the ``simple'' and ``fancy'' models for simulating phenotypes.
+Simulations use an heritability of 80\%.
+\textbf{A:} Mean of AUC over 5 simulations.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+The blue dotted line represents the maximum achievable AUC.
+\textbf{B:} Boxplots of numbers of predictors used by the methods for 5 simulations.
+\textbf{C:} Boxplots of execution times for 5 simulations.}
 \label{fig:supp-ttrees}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-triple}}
-\caption{[TRIPLE 3 PLOTS - H2=0.8]}
+\caption{Comparison of ``logit-triple'' and ``logit-simple'' in scenario \textnumero1.
+Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
+Horizontal panels are presenting results for the ``simple'' and ``fancy'' models for simulating phenotypes.
+Simulations use an heritability of 80\%.
+\textbf{A:} Mean of AUC over 100 simulations.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+The blue dotted line represents the maximum achievable AUC.
+\textbf{B:} Boxplots of numbers of predictors used by the methods for 100 simulations.
+\textbf{C:} Boxplots of execution times for 100 simulations.}
 \label{fig:supp-triple}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-corr}}
-\caption{Scenario \textnumero1. Comparison of different predictive measures. ..There is a Spearman correlation.. of 98\% between values of AUC and partial AUC.[ADD]\label{fig:supp-AUC-corr}}
+\caption{Correlation between AUC and partial AUC values in scenario \textnumero1.
+There is a Spearman correlation of 98\% between values of AUC and partial AUC.
+The relation between the two values are the same whatever are the heritability, distribution of effects and method used.\label{fig:supp-AUC-corr}}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-score-densities}}
-\caption{[ADD]\label{fig:supp-score-densities}}
+\caption{Density of polygenic risk scores for the ``C+T'' (A), ``logit-simple'' (B) and ``logit-triple'' (C) methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.\label{fig:supp-score-densities}}
 \end{figure}

### 2018-04-02 paper/draft.tex
typo

@@
 I HAVE SUMMARY() IN R]
-Thirdly, we reproduce the exact same procedure of simulations and, for each combination of parameters (Table \ref{tab:simus}), we estimate the AUC of the ``oracle'', i.e.\ the true simulated genetic part of the liabilities though 100 replicates.
+Thirdly, we reproduce the exact same procedure of simulations and, for each combination of parameters (Table \ref{tab:simus}), we estimate the AUC of the ``oracle'', i.e.\ the true simulated genetic part of the liabilities through 100 replicates.
 For every combination of parameters, AUC of oracles are comprised between 83.2\% and 84.2\% for an heritability of 50\% and between 93.2\% and 94.1\% for an heritability of 80\%.

### 2018-04-16 results-simus-pres.Rmd
update figures

@@
     barplot_causal() +
-    geom_hline(yintercept = 0.94, linetype = 3, color = "blue") +
-    facet_grid(par.dist ~ .)
+    geom_hline(yintercept = 0.94, linetype = 3, color = "blue")
 }
@@
       theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
-      # facet_grid(par.model ~ par.dist) +
       labs(x = "Causal SNPs (number and location)", y = ylab, 
@@
     barplot_causal_one() +
-    facet_grid(par.model ~ par.dist) +
+    facet_grid(par.model ~ .) +
     labs(y = sprintf("Mean of %d AUCs", nsimu))
@@
   theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
-  facet_grid(par.dist ~ par.causal) +
+  facet_grid(. ~ par.causal) +
   scale_fill_brewer(type = "qual", palette = 2) +
@@
                 size = 1.5, width = 0) +
-  facet_grid(. ~ par.dist) +
   scale_x_continuous(breaks = 1:6 * 1000, minor_breaks = NULL) + 

### 2018-04-16 CMSA-graphic.R

@@ -25,18 +25,22 @@ plot_remains(5)
 plot_remains(5, size = 30)
 
 
-p1 <- plot_remains(5) +
-  geom_text(position = position_stack(vjust = 0.5))
+p1 <- plot_remains(5, size = 25) +
+  geom_text(position = position_stack(vjust = 0.5), size = rel(4)) +
+  theme(legend.key.height = unit(1.5, "line"), 
+        legend.key.width  = unit(1.5, "line"),
+        legend.position = "left")
 print(p1)
 ggsave("../thesis-docs/figures/CMSA-explained1.svg", p1,
        scale = 1/90, width = 740, height = 580)
 
 source("CMSA-inside.R")
 ggsave("../thesis-docs/figures/CMSA-explained2.svg", p2,
-       scale = 1/90, width = 800, height = 620)
+       scale = 1/100, width = 900, height = 650)
 
-cowplot::plot_grid(p1, p2, align = "hv", ncol = 1,
+cowplot::plot_grid(p1, p2, nrow = 1,
                    labels = LETTERS[1:2], 
                    label_size = 25,
-                   scale = 0.95)
-ggsave("figures/CMSA-explained.pdf", scale = 1/90, width = 760, height = 1000)
+                   scale = 0.95, rel_widths = c(4, 7))
+ggsave("figures/CMSA-explained.pdf", scale = 1/100, width = 1300, height = 600)
+ggsave("../thesis-docs/figures/CMSA-explained.svg", scale = 1/100, width = 1300, height = 600)

### 2018-04-16 CMSA-inside.R

@@ -17,13 +17,14 @@ myggplot <- function(..., coeff = 1) {
 # saveRDS(tmp, "one-logit-CMSA.rds")
 tmp <- readRDS("one-logit-CMSA.rds")
 ind.min <- which.min(tmp$loss.val)
+ind <- seq(2, length(tmp$lambda), by = 2)
 
 library(tidyverse)
 
 p2 <- bind_rows(
-  data_frame(lambda = tmp$lambda, loss = tmp$loss.val, 
+  data_frame(lambda = tmp$lambda[ind], loss = tmp$loss.val[ind], 
              Dataset = "Remaining fold"),
-  data_frame(lambda = tmp$lambda, loss = tmp$loss,
+  data_frame(lambda = tmp$lambda[ind], loss = tmp$loss[ind],
              Dataset = "Training folds")
 ) %>%
   myggplot(coeff = 1.2) +

### 2018-04-18 check-cmsa.Rmd
need to check this

--- 2018-02-12 3-simus.Rmd
@@
 
-### PRS
-
-```{r}
-paste(sprintf("%.1e", 10^(-c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100))))), collapse = ", ")
-PRS <- function(G, infos.chr, infos.pos, 
-                pheno.all, covar.all, 
-                ind.train, ind.test) {
-
-  timing <- system.time({
-
-    # GWAS
-    gwas.train <- big_univLogReg(
-      G, pheno.all[ind.train], ind.train = ind.train, 
-      covar.train = covar.all[ind.train, , drop = FALSE], 
-      ncores = nb_cores()
-    )
-    gwas.train.gc <- snp_gc(gwas.train)
-
-    # Clumping on the test set
-    ind.keep <- snp_clumping(G, infos.chr = infos.chr,
-                             ind.row = ind.test,
-                             thr.r2 = 0.2, 
-                             S = abs(gwas.train.gc$score),
-                             size = 500,
-                             is.size.in.bp = TRUE,
-                             infos.pos = infos.pos,
-                             ncores = nb_cores())
-
-    # PRS
-    thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
-    lpS <- -predict(gwas.train.gc)
-    nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
-    prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
-                   ind.test = ind.test,
-                   ind.keep = ind.keep,
-                   lpS.keep = lpS[ind.keep], 
-                   thr.list = thrs)
-  })[3]
-
-  ind.best <- which.max(apply(prs, 2, AUC, target = pheno.all[ind.test]))
-
-  methods <- c("PRS-all", "PRS-stringent", "PRS-max")
-  indices <- c(1:2, ind.best)
-
-  lapply(1:3, function(i) {
-    k <- indices[i]
-    tibble(
-      method = methods[i],
-      eval = list(cbind(prs[, k], pheno.all[ind.test])),
-      timing = timing,
-      nb.preds = nb.pred[k]
-    )
-  }) %>%
-    bind_rows()
-}
-```
-
 ### Logistic regression
@@
     nb.preds = sum(rowSums(sapply(cmsa.logit, function(x) x$beta.X) != 0) > 0)
-  )
-}
-```
-
-### T-Trees
-
-```{r}
-TTree <- "../../TTree-source/TTree"
-
-ttrees <- function(TTree, file.base, pheno.all, ind.train, ind.test,
-                   n.trees = 100) {
-
-  file0.jdb  <- paste0(file.base, ".jdb")
-  file0.bloc <- paste0(file.base, ".bloc")
-
-  # Write jdb file with new pheno
-  tmpfile <- tempfile()
-  write(c(rep("", 5), pheno.all), ncolumns = 1, tmpfile)
-  # https://stackoverflow.com/a/7846550/6103040
-  system(sprintf("awk 'FNR==NR{a[NR]=$1;next}{$2=a[FNR]}1' %s %s > %s",
-                 tmpfile, file0.jdb, file.jdb <- paste0(tmpfile, ".jdb")))
-  # Write indices of learning and validation sets
-  file.learn <- paste0(tmpfile, "_learn.txt")
-  file.val   <- paste0(tmpfile, "_val.txt")
-  cat(ind.train - 1, file = file.learn, sep = "\t")
-  cat(ind.test - 1,  file = file.val,   sep = "\t")
-
-  timing <- system.time(
-    system(glue::glue(
-      "{TTree}",
-      " -j {file.jdb}",
-      " -m 3",
-      " -b {file0.bloc}",
-      " -l {file.learn}",
-      " -v {file.val}", 
-      " -t {n.trees} -k 1000 -c 5 -n 2000",
-      " -x -s"
-    ))
-  )[3]
-
-  file.roc <- sprintf("%s_k1000_m3_t%d_ic5_nmin2000_0000.roc",
-                      file.jdb, n.trees)
-  file.vim <- sub("\\.roc$", ".vim", file.roc)
-  preds <- read.table(file.roc, header = FALSE)
-
-  tibble(
-    method   = "T-Trees",
-    eval     = list(preds[match(ind.test - 1, preds[[1]]), 2:3]),
-    timing   = timing,
-    nb.preds = sum(read.table(file.vim)$V2 != 0)
   )

### 2018-04-18 check-cmsa.Rmd
save

@@
 library(bigsnpr)
-celiac2 <- snp_attach("backingfiles/celiacQC_sub1.rds")
+celiac2 <- snp_attach("backingfiles/celiacQC.rds")
 CHR <- celiac2$map$chromosome
 POS <- celiac2$map$physical.pos
-(G <- celiac2$genotypes)
-(G2 <- big_attach("backingfiles/celiacQC_sub1_tripled1.rds"))
+G <- celiac2$genotypes
 
-covar.all <- readRDS("backingfiles/PCA2.rds")$u
+covar.all <- readRDS("backingfiles/PCA.rds")$u
 
 n <- nrow(G)
-ind.HLA <- snp_indLRLDR(CHR, POS, subset(LD.wiki34, ID == "hild12"))
 ```
@@
   pheno <- as.numeric(y.simu > qnorm(1 - K))
-}
-```
-
-### Scenario n°1 (with T-Trees)
-
-```{r}
-params.grid1 <- expand.grid(
-  n.train    = 6000,
-  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
-  par.dist   = c("gaussian", "laplace"), 
-  par.h2     = 0.8, 
-  par.model  = c("simple", "fancy"),
-  num.simu   = 1:5,
-  stringsAsFactors = FALSE
-) 
-```
-
-```{r, eval=TRUE}
-if (!dir.exists("results1")) dir.create("results1")
-
-for (i in rows_along(params.grid1)) {
-
-  res.file <- paste0("results1/simu_", i, ".rds")
-  if (file.exists(res.file)) next
-
-  params <- params.grid1[i, ]
-  par.causal <- params[["par.causal"]][[1]]
-
-  # Simulate phenotypes
-  pheno.all <- get_pheno(
-    G,    
-    h2 = params[["par.h2"]], 
-    M = as.integer(par.causal[1]), 
-    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
-    effects.dist = params[["par.dist"]], 
-    model = params[["par.model"]], 
-    K = 0.3                 
-  )
-
-  # Split in training/test sets
-  ind.train <- sort(sample(n, size = params[["n.train"]]))
-  ind.test <- setdiff(1:n, ind.train)
-
-  # Get results from all methods
-  res <- bind_rows(
-    PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test),
-    logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple"),
-    logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple"),
-    ttrees(TTree, "backingfiles/ttrees", pheno.all, ind.train, ind.test, n.trees = 100)
-  )
-  params[["res"]] <- list(res)
-  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
 }

### 2018-04-19 paper/draft.tex
modifs

@@
 It uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it very suitable and fast.
-However, PRS rely on non-optimal heuristics to account for correlation and reduce noise in the predictor.
+However, PRS rely on non-optimal heuristics to account for correlation and to reduce noise in the predictor.
 
@@
 Moreover, predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, whose optimal value depends on disease architecture. 
-The improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs. 
+The improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs, or when sample size is increasing. 
 When analyzing real SNP data from a case-control study,  penalized logistic regression provides an AUC of 89\% while the standard C+T method provides a smaller AUC of 82\%.
 
-Instead of computing polygenic risk scores based on GWAS univariate summary statistics, we recommend to use penalized logistic regression as implemented in the R package bigstatsr to obtain more accurate polygenic risk scores.\\
+Instead of computing polygenic risk scores based on GWAS univariate summary statistics, we recommend to use penalized logistic regression as implemented in the R package bigstatsr to obtain more accurate polygenic risk scores when individual genotype data are available.\\
 
@@
 Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}.
-Yet, the predictions resulting from LMM, known e.g.\ as gBLUP, are not designed for binary traits such as a disease status and are not optimal in this setting \cite[]{abraham2013performance}. 
-Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time requirements, which makes them unlikely to be used for prediction on large cohorts \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
+Yet, the predictions resulting from LMM, known e.g.\ as ``gBLUP'', are not designed for binary traits such as a disease status and are not optimal in this setting \cite[]{abraham2013performance}. 
+Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time requirements, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
 Finally, statistical learning methods have also been used to derive PRSs for complex human diseases by jointly estimating SNP effects.
-Such methods include logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2013walk,botta2014exploiting,wei2009disease}.
+Such methods include logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,wei2009disease}.
 
-We recently developed two R packages, bigstatsr and bigsnpr, for efficient analysis of large-scale genome-wide data \cite[]{prive2017efficient}.
+We recently developed two R packages, bigstatsr and bigsnpr, for efficiently analyzing large-scale genome-wide data \cite[]{prive2017efficient}.
 Package bigstatsr includes an efficient algorithm with new implementation tricks for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
@@
 
-First, we quickly discard T-Trees because they show lower predictive performance as compared to other methods, while being computationally demanding.
-Then, we show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture, as shown in \cite{ware2017heterogeneity}.
+First, we quickly discard T-Trees because this method shows lower predictive performance as compared to other methods, while being computationally demanding.
+Then, we show that the penalized logistic regression consistently achieves higher predictive performance while being faster than the C+T method, whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture, as shown in \cite{ware2017heterogeneity}.
 
@@
 
-We use real genotypes of European individuals from a case/control celiac disease cohort \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}.
+We use real genotypes of European individuals from a case-control celiac disease study \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}.
 Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
@@
 In order to simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}.
-We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only chromosome 6), the size of the training set (from 1000 to 6000), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace).
+We vary different parameters: the number of chromosomes used (either all 22 chromosomes or chromosome 6 only), the size of the training set (from 1000 to 6000), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace).
 We also use two different models to compute liability scores: a ``simple'' model with linear effects only, and a ``fancy'' model which combines linear, dominant and interaction-type effects.
@@
 For the ``fancy'' model, we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$ 
-and $S_\text{causal}^{(3)}$ ($S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$). 
+and $S_\text{causal}^{(3)}$; $S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$. 
 We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$
@@
 Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only.
-Reducing the number of SNPs aims at increasing the polygenicity (the proportion of causal SNPs) of the simulated models and at virtually increasing the sample size \cite[]{dudbridge2013power}.
-For this scenario, we use only the additive (``simple'') model, but continue to compare all previous variations of the other parameters.
+Reducing the number of SNPs aims at increasing the polygenicity (the proportion of causal SNPs) of the simulated models and at virtually increasing the sample size \cite[]{dudbridge2013power,marquez2017multiethnic,vilhjalmsson2015modeling}.
+For this scenario, we use only the additive (``simple'') model, but continue to compare all previous different values of the other parameters.
 Finally, scenario \textnumero3 reuses the whole dataset but this time varying the size of the training set in order to assess how the sample size affects predictive performance of methods.
-A total of 100 simulations are run using 300 causal SNPs anywhere on the genome.
+A total of 100 simulations per combination of parameters are run using 300 causal SNPs anywhere on the genome.
 
@@
 
-The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Studies (GWAS), referred to as summary statistics.
-For the C+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
+The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Study (GWAS), referred to as summary statistics.
+In the GWAS, a coefficient of regression (estimated effect size) is learned independently for each SNP along with a corresponding p-value. 
 The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another, which form a set of SNPs denoted $S_\text{clumping}$. 
 Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $p_T$ to be determined. 
-Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$\rm{PRS}_i = \sum_{j \in S_\text{clumping}} \mathds{1}\{p_j < p_T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS.
+Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS.
 In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.
@@
 
-Finally, for the penalized logistic regression, we solve: $$\argmin_{\beta_0, \beta}(\lambda, \alpha)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that needs to be determined.
+Finally, for the penalized logistic regression, we solve: $$\arg\!\min_{\beta_0,~\beta}(\lambda, \alpha)\left\{  \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization}  \right\}$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that needs to be determined.
 Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
-the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse and therefore more interpretable models (corresponds to $\alpha = 1$);
+the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
 the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
@@
 To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in the R package bigstatsr \cite[]{prive2017efficient}.
-This type of algorithm builds predictions for many values of $\lambda$ (typically a ``regularization path'' of 100 values).
-To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA).
-First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds).
-Secondly, in turn, each fold is considered as an inner validation set and the other ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which minimizes the log loss on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is selected.
-Finally, the $K$ resulting probabilities for each individual are averaged.
-We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making this procedure much faster.
+This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''.
+To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
 Because of the L1-regularization, the resulting vectors of coefficients are typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts.
@@
 In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects.
-This results in a dataset with 3 times as many variables as the initial one, on which we can apply penalized logistic regression with the CMSA procedure, as described previously.
+This results in a dataset with 3 times as many variables as the initial one, on which we can apply the penalized logistic regression with the CMSA procedure (Figure \ref{fig:CMSA}).
 We call this method ``logit-triple'' in the results.
@@
 
-We use Monte Carlo cross-validation to compute AUC, partial AUC, number of predictors used and execution time for the original Celiac dataset with real phenotypes.
-We randomly split 100 times the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals.
+We use Monte Carlo cross-validation to compute AUC, partial AUC, number of predictors used and execution time for the original Celiac dataset with real phenotypes: we randomly split 100 times the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals.
 
@@
 When there are 30 causal SNPs and 80\% of heritability, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}).
-Moreover, the penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted in simulations with 3000 causal SNPs.
-Yet, these are cases where all methods hardly exceed 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
+Moreover, the penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted in simulations with 3000 causal SNPs, which are cases where all methods hardly exceed 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
 
@@
 
-\subsection{Other methods and ``fancy'' model}
+\subsection{The other methods and the ``fancy'' model}
 
@@
 
-In this comparative study, we propose penalized logistic regressions as a replacement for the widely used C+T method for computing polygenic risk scores. 
+In this comparative study, we propose penalized logistic regressions as a replacement for the widely used C+T method for computing polygenic risk scores when individual genotype data are available. 
 We show that jointly estimating SNP effects using efficient penalized logistic regressions enables to achieve higher predictive performance, as reported in a previous study \cite[]{abraham2013performance}.
-Indeed, when disease architecture and sample size allow for high predictive performance (AUC greater than 85\%), ``logit-simple'' provides higher predictive performance than the widely-used C+T method, nearly reaching optimal AUC values, which is encouraging for using polygenic scores in clinical settings.
-The C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions yet. 
+Indeed, when disease architecture and sample size allow for high predictive performance (say AUC greater than 85\%), ``logit-simple'' provides higher predictive performance than the widely-used C+T method, nearly reaching optimal AUC values, which is encouraging for using polygenic risk scores in clinical settings.
+The C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not yet the case for our penalized logistic regressions. 
 However, in order to accurately predict disease status for clinical interventions, best possible predictive methods should be used.
@@
 The choice of this threshold is very important as it can greatly impact the predictive performance of the method \cite[]{ware2017heterogeneity}.
-On the contrary, in the penalized logistic regression presented here, we remove the choice of the regularization parameter with a procedure that we call Cross-Model Selection and Averaging (CMSA) while keeping high predictive performance.
+On the contrary, in the penalized logistic regression presented here, we remove the choice of the regularization parameter with a procedure that we call Cross-Model Selection and Averaging (CMSA) while keeping high predictive performance [TODO: COMPARAISON AVEC LE MAX DE BIGLASSO].
+Usually, cross-validation is used to assess the best value for the hyper-parameter and then the model is train again with this particular hyper-parameter value \cite[]{Hastie2008,wei2013large}.
+Yet, as the value of $\lambda$ depends on the sample size and doing cross-validation and retraining is computationally demanding, we prefer to use CMSA when there is a model to choose from a regularization path, with the possibility of adding an early stopping criterion to fasten the whole procedure.
 
@@
 
-Finally, in this comparative study, we did not consider the problem of population structure \cite[]{martin2017human} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}.
+Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}.
 In another study, we will assess how can we use models and effects learned in one population to improve learning and prediction in another population.
@@
 \begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{simple-CMSA}}
+\caption{Illustration of one turn of the Cross-Model Selection and Averaging (CMSA) procedure. First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds). 
+Secondly, in turn, each fold is considered as an inner validation set (red) and the other ($K - 1$) folds form an inner training set (blue).
+A ``regularization path'' of models is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed.
+The model that minimizes the loss on the inner validation set is selected.
+Finally, the $K$ resulting models are averaged.
+We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making this procedure much faster.}
+\label{fig:CMSA}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newpage
+\begin{figure}[h]
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6}}

### 2018-04-19 check-cmsa.R

@@ -1,0 +1,50 @@
+params.grid2 <- expand.grid(
+  n.train = 3000,
+  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
+  par.dist   = c("gaussian", "laplace"), 
+  par.h2     = c(0.5, 0.8), 
+  par.model  = c("simple", "fancy"),
+  num.simu   = 1:100,
+  stringsAsFactors = FALSE
+) 
+
+i <- 1
+params <- params.grid2[i, ]
+par.causal <- params[["par.causal"]][[1]]
+
+# Simulate phenotypes
+pheno.all <- get_pheno(
+  G,    
+  h2 = params[["par.h2"]], 
+  M = as.integer(par.causal[1]), 
+  ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
+  effects.dist = params[["par.dist"]], 
+  model = params[["par.model"]], 
+  K = 0.3                 
+)
+
+# Split in training/test sets
+ind.train <- sort(sample(n, size = params[["n.train"]]))
+ind.test <- setdiff(1:n, ind.train)
+
+system.time(
+  test <- logit.CMSA(G, pheno.all, covar.all, 
+                     ind.train, ind.test, "logit-simple")
+)
+AUC(test$eval[[1]][, 1], test$eval[[1]][, 2])
+
+library(biglasso)
+# G2 <- big.matrix(nrow(G), ncol(G) + ncol(covar.all), 
+#                 backingfile = "G-PC", backingpath = "backingfiles")
+# dim(G2)
+# big_apply(G, function(X, ind) {
+#   G2[, ind] <- X[, ind]
+#   NULL
+# }, a.combine = 'c')
+# G2[, ncol(G) + cols_along(covar.all)] <- covar.all
+G2 <- attach.big.matrix("backingfiles/G-PC.desc")
+system.time(
+  test2 <- biglasso(G2, pheno.all, ind.train, ncores = nb_cores(),
+                    penalty = "enet", family = "binomial", alpha = 0.5)
+)
+

### 2018-04-24 test-pruning-pool.R
test pcadapt

@@ -1,0 +1,69 @@
+library(bigsnpr)
+
+celiac <- snp_attach("backingfiles/celiacQC.rds")
+G <- celiac$genotypes
+CHR <- celiac$map$chromosome
+POS <- celiac$map$physical.pos
+obj.svd <- readRDS("backingfiles/PCA.rds")
+# plot(obj.svd, type = "loadings", loadings = 1:10, coeff = 0.5)
+
+files <- list.files("backingfiles/others/", "^cluster",
+           full.names = TRUE)
+pop <- snp_getSampleInfos(celiac, files)[[1]]
+pop[pop == 4] <- 3
+library(ggplot2)
+plot(obj.svd, type = "scores", scores = 3:4) +
+  aes(color = as.factor(pop))
+
+ind.pop <- split(rows_along(G), pop)
+
+afs <- sapply(seq_along(ind.pop), function(i) {
+  big_scale()(G, ind.row = ind.pop[[i]])$center / 2
+})
+
+# bigsnpr:::write.table2(t(afs), "backingfiles/celiac-pool.txt")
+
+library(pcadapt)
+pool_mat <- read.pcadapt(t(afs), type = "pool")
+dim(pool_mat)
+class(pool_mat)
+
+obj.pool <- pcadapt(pool_mat)
+plot(obj.pool)
+scree_plot(obj.pool, 4)
+plot(obj.pool$loadings[, 1], pch = 20)
+plot(obj.pool$loadings[, 2], pch = 20)
+plot(obj.pool$loadings[, 3], pch = 20)
+plot(obj.pool$loadings[, 4], pch = 20)
+
+plink <- download_plink()
+file.base <- "../Dubois2010_data/FinnuncorrNLITUK1UK3hap300_QC_norel"
+tmp <- tempfile()
+system(glue::glue("{plink} --bfile {file.base} --out {tmp}", 
+                  " --indep-pairwise 500 1 0.05"))
+to_keep <- read.table(paste0(tmp, ".prune.in"), stringsAsFactors = FALSE)[[1]]
+ind.keep <- match(to_keep, celiac$map$marker.ID)
+
+pool_mat.sub <- read.pcadapt(t(afs[ind.keep, ]), type = "pool")
+pcadapt.sub <- pcadapt(pool_mat.sub)
+plot(pcadapt.sub, chr.info = celiac$map$chromosome[ind.keep])
[…]

### 2018-04-25 3-simus.Rmd
add biglasso to simulations

@@
     nb.preds = sum(read.table(file.vim)$V2 != 0)
+  )
+}
+```
+
+### Logistic regression with biglasso
+
+```{r}
+logit.biglasso <- function(G3, pheno.all, covar.all, ind.train, ind.test) {
+
+  timing <- system.time({
+
+    biglasso <- biglasso(G3, pheno.all, ind.train, ncores = nb_cores(),
+                         penalty = "enet", family = "binomial", alpha = 0.5)
+
+    preds <- 1 / (1 + exp(-predict(biglasso, G3, ind.test)))
+  })[3]
+
+  aucs <- apply(preds, 2, bigstatsr::AUC, target = pheno.all[ind.test])
+  ind.max <- which.max(aucs)
+
+  tibble(
+    method   = "biglasso", 
+    eval     = list(cbind(preds[, ind.max], pheno.all[ind.test])),
+    timing   = timing,
+    nb.preds = sum(biglasso$beta[, ind.max] != 0)
   )
@@
 ```
+
+### Scenario n°1 (comparison with biglasso)
+
+```{r}
+library(biglasso)
+library(Matrix)
+```
+
+```{r, eval=FALSE}
+G3 <- big.matrix(nrow(G), ncol(G) + ncol(covar.all),
+                 backingfile = "G-PC", 
+                 backingpath = "backingfiles")
+big_apply(G, function(X, ind) {
+  G3[, ind] <- X[, ind]
+  NULL
+}, a.combine = 'c')
+G3[, ncol(G) + cols_along(covar.all)] <- covar.all
+```
+
+```{r}
+G3 <- attach.big.matrix("backingfiles/G-PC.desc")
+```
+
+```{r}
+params.grid7 <- expand.grid(
+  n.train = 6000,
+  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
+  par.dist   = c("gaussian", "laplace"), 
+  par.h2     = c(0.8), 
+  par.model  = c("simple"),
+  num.simu   = 1:100,
+  stringsAsFactors = FALSE
+) 
+```
+
+```{r, eval=TRUE}
+if (!dir.exists("results7")) dir.create("results7")
+
+for (i in rows_along(params.grid7)) {
+
+  res.file <- paste0("results7/simu_", i, ".rds")
+  if (file.exists(res.file)) next
+
+  params <- params.grid7[i, ]
+  par.causal <- params[["par.causal"]][[1]]
+
+  # Simulate phenotypes
+  pheno.all <- get_pheno(
+    G,    
+    h2 = params[["par.h2"]], 
+    M = as.integer(par.causal[1]), 
+    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
+    effects.dist = params[["par.dist"]], 
+    model = params[["par.model"]], 
+    K = 0.3                 
+  )
+
+  # Split in training/test sets
+  ind.train <- sort(sample(n, size = params[["n.train"]]))
+  ind.test <- setdiff(1:n, ind.train)
+
+  # Get results from all methods
+  res <- bind_rows(
+    logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test, "logit-simple"),
+    logit.biglasso(G3, pheno.all, covar.all, ind.train, ind.test)
+  )
+  params[["res"]] <- list(res)
+  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
+}
+```

### 2018-04-25 4-results-simus.Rmd
save figures
add biglasso to simulations

@@
 
-methods.color <- setNames(scales::hue_pal()(6), 
+methods.color <- setNames(c(scales::hue_pal()(6), "black"), 
                           c("logit-simple", "PRS-max", "PRS-stringent",
-                            "T-Trees", "logit-triple", "PRS-all")) 
+                            "T-Trees", "logit-triple", "PRS-all", "biglasso")) 
 ```
@@
 ```
+
+### Comparison with biglasso
+
+```{r}
+results7 <- list.files("results7", full.names = TRUE) %>%
+  map_dfr(~readRDS(.x)) %>%
+  format_results()
+```
+
+```{r}
+compare_logit(results7, "biglasso")
+
+ggsave("figures/supp-biglasso.pdf", scale = 1/100, width = 1600, height = 1000)
+```

### 2018-04-25 check-cmsa.R

@@ -1,50 +1,0 @@
-params.grid2 <- expand.grid(
-  n.train = 3000,
-  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
-  par.dist   = c("gaussian", "laplace"), 
-  par.h2     = c(0.5, 0.8), 
-  par.model  = c("simple", "fancy"),
-  num.simu   = 1:100,
-  stringsAsFactors = FALSE
-) 
-
-i <- 1
-params <- params.grid2[i, ]
-par.causal <- params[["par.causal"]][[1]]
-
-# Simulate phenotypes
-pheno.all <- get_pheno(
-  G,    
-  h2 = params[["par.h2"]], 
-  M = as.integer(par.causal[1]), 
-  ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
-  effects.dist = params[["par.dist"]], 
-  model = params[["par.model"]], 
-  K = 0.3                 
-)
-
-# Split in training/test sets
-ind.train <- sort(sample(n, size = params[["n.train"]]))
-ind.test <- setdiff(1:n, ind.train)
-
-system.time(
-  test <- logit.CMSA(G, pheno.all, covar.all, 
-                     ind.train, ind.test, "logit-simple")
-)
-AUC(test$eval[[1]][, 1], test$eval[[1]][, 2])
-
-library(biglasso)
-# G2 <- big.matrix(nrow(G), ncol(G) + ncol(covar.all), 
-#                 backingfile = "G-PC", backingpath = "backingfiles")
-# dim(G2)
-# big_apply(G, function(X, ind) {
-#   G2[, ind] <- X[, ind]
-#   NULL
-# }, a.combine = 'c')
-# G2[, ncol(G) + cols_along(covar.all)] <- covar.all
-G2 <- attach.big.matrix("backingfiles/G-PC.desc")
-system.time(
-  test2 <- biglasso(G2, pheno.all, ind.train, ncores = nb_cores(),
-                    penalty = "enet", family = "binomial", alpha = 0.5)
-)
-

### 2018-04-26 3-simus.Rmd
add biglasso comparison

@@
 library(glue)
+NCORES <- 6
 ```
@@
       covar.train = covar.all[ind.train, , drop = FALSE], 
-      ncores = nb_cores()
+      ncores = NCORES
     )
@@
                              infos.pos = infos.pos,
-                             ncores = nb_cores())
+                             ncores = NCORES)
 
@@
                                covar.train = covar.all[ind.train, , drop = FALSE],
-                               ncores = nb_cores())
+                               ncores = NCORES)
 
@@
 
-    biglasso <- biglasso(G3, pheno.all, ind.train, ncores = nb_cores(),
+    biglasso <- biglasso(G3, pheno.all, ind.train, ncores = NCORES,
                          penalty = "enet", family = "binomial", alpha = 0.5)

### 2018-04-26 3-simus.Rmd
review H

@@
 ```{r}
-logit.CMSA <- function(G, pheno.all, covar.all, ind.train, ind.test, method) {
+logit.CMSA <- function(G, pheno.all, covar.all, ind.train, ind.test,
+                       method = "logit-simple", alpha = 0.5) {
 
@@
                                covar.train = covar.all[ind.train, , drop = FALSE],
-                               ncores = NCORES)
+                               alpha = alpha, ncores = NCORES)
 
@@
   res <- bind_rows(
-    logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test, "logit-simple"),
+    logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test),
     logit.biglasso(G3, pheno.all, covar.all, ind.train, ind.test)
@@
 ```
+
+### Scenario n°1 (comparison of different $\alpha$)
+
+```{r, eval=TRUE}
+if (!dir.exists("results8")) dir.create("results8")
+
+for (i in rows_along(params.grid7)) {
+
+  res.file <- paste0("results8/simu_", i, ".rds")
+  if (file.exists(res.file)) next
+
+  params <- params.grid7[i, ]
+  par.causal <- params[["par.causal"]][[1]]
+
+  # Simulate phenotypes
+  pheno.all <- get_pheno(
+    G,    
+    h2 = params[["par.h2"]], 
+    M = as.integer(par.causal[1]), 
+    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
+    effects.dist = params[["par.dist"]], 
+    model = params[["par.model"]], 
+    K = 0.3                 
+  )
+
+  # Split in training/test sets
+  ind.train <- sort(sample(n, size = params[["n.train"]]))
+  ind.test <- setdiff(1:n, ind.train)
+
+  # Get results from all methods
+  res <- bind_rows(
+    logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test, 
+               method = "logit-simple-enet-ridge", alpha = 0.001),
+    logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test, 
+               method = "logit-simple",            alpha = 0.5),
+    logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test, 
+               method = "logit-simple-lasso",      alpha = 1)
+  )
+  params[["res"]] <- list(res)
+  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
+}
+```

### 2018-04-26 4-results-simus.Rmd
add biglasso comparison
review H

@@
 
-```{r}
+```{r, fig.asp=0.65, out.width="95%", fig.width=16}
 compare_logit(results7, "biglasso")
@@
 ```
+
+### Comparison of different alpha values
+
+```{r}
+results8 <- list.files("results8", full.names = TRUE) %>%
+  map_dfr(~readRDS(.x)) %>%
+  format_results()
+```
+
+```{r, fig.asp=0.65, out.width="95%", fig.width=16}
+compare_logit(results8, c("logit-simple-enet-ridge", "logit-simple-lasso"))
+
+ggsave("figures/supp-alpha.pdf", scale = 1/100, width = 1600, height = 1000)
+```

### 2018-04-26 paper/draft.tex
review H

@@
 
-\title{Joint estimation of SNP effects improves the prediction of complex diseases // Predicting complex diseases: performance and robustness [OF NEW METHOD?]}
+\title{Highly efficient implementation of penalized models and their application to genetic risk prediction}
 \author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
@@
 Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. 
-PRSs can have a major public health impact, allowing for screening campaigns to identify high-genetic risk individuals for a given disease.
+PRSs might have a major public health impact, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.
 The most commonly used method to derive PRSs is called ``Clumping+Thresholding'' (C+T).
 It uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it very suitable and fast.
-However, PRS rely on non-optimal heuristics to account for correlation and to reduce noise in the predictor.
+However, these PRS rely on non-optimal heuristics.
 
-In R package bigstatsr, we implemented a logistic regression model that provides a joint estimation of SNP effects.
-We used an efficient algorithm with new implementation tricks in order for this model to be used on huge datasets as large as the UK Biobank.
+In R package bigstatsr, we implemented linear and logistic regression models that can be used to jointly estimate SNP effects for computing PRS.
+We use an efficient algorithm with a new implementation in order for this model to be used on huge datasets.
 
-In this paper, we present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests.
+In this paper, we present a comprehensive comparative study of the C+T method, our penalized logistic regression and a derivation of random forests.
 The penalized logistic regression consistently achieves higher predictive performance than the two other methods while being faster than the C+T method.
 Moreover, predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, whose optimal value depends on disease architecture. 
-The improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs, or when sample size is increasing. 
-When analyzing real SNP data from a case-control study,  penalized logistic regression provides an AUC of 89\% while the standard C+T method provides a smaller AUC of 82\%.
+The improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs, and with an increasing sample size. 
+When analyzing real SNP data from a case-control study for celiac disease, penalized logistic regression provides an AUC of 89\% while the standard C+T method provides a smaller AUC of 82\%.
 
-Instead of computing polygenic risk scores based on GWAS univariate summary statistics, we recommend to use penalized logistic regression as implemented in the R package bigstatsr to obtain more accurate polygenic risk scores when individual genotype data are available.\\
+[TOMOD] Instead of computing polygenic risk scores based on GWAS univariate summary statistics, we recommend to use penalized logistic regression as implemented in the R package bigstatsr to obtain more accurate polygenic risk scores when individual genotype data are available.\\
 
@@
 As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}.
-Moreover, screening based on sequencing seems to make individuals positively change their health behavior, while neither causing patient anxiety nor depression \cite[]{vassy2017impact}.
 Yet, PRSs would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases. 
@@
 Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}.
-Yet, the predictions resulting from LMM, known e.g.\ as ``gBLUP'', are not designed for binary traits such as a disease status and are not optimal in this setting \cite[]{abraham2013performance}. 
+Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes \cite[]{abraham2013performance}. 
 Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time requirements, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
 Finally, statistical learning methods have also been used to derive PRSs for complex human diseases by jointly estimating SNP effects.
-Such methods include logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,wei2009disease}.
+Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,wei2009disease}.
 
 We recently developed two R packages, bigstatsr and bigsnpr, for efficiently analyzing large-scale genome-wide data \cite[]{prive2017efficient}.
-Package bigstatsr includes an efficient algorithm with new implementation tricks for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
+Package bigstatsr includes an efficient algorithm with a new implementation for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
 In this paper, we present a comprehensive comparative study of the C+T method, our implementation of penalized logistic regression and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
-Note that in this comparison we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp} [<- IN METHODS?
-MAIS BIEN AUSSI DE DIRE PQ TOT..].
+Note that in this comparison we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
 For the C+T model, we compare different thresholds of inclusion of SNPs.
-For the logistic regression, we include two novel approaches. 
+For the penalized logistic regression, we include two practical improvements. 
 First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for automatically choosing the regularization parameter of the penalized logistic regression, which directly affects the number of SNPs included in the model.
-We also show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.
+Secondly, we also show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.
 
@@
 \cite{ware2017heterogeneity} evaluated best practices for using the C+T method and found out that the optimal inclusion threshold is trait-dependent. 
-In this study, we particularly extend the work of \cite{abraham2013performance} by using a simulation framework similar but more extenstive than the one used in \cite{zhou2013polygenic}. 
+[TOMOD] In this study, we particularly extend the work of \cite{abraham2013performance} by using a simulation framework similar but more extenstive than the one used in \cite{zhou2013polygenic}. 
 In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the heritability.
@@
 
-First, we quickly discard T-Trees because this method shows lower predictive performance as compared to other methods, while being computationally demanding.
+[TOMOD] First, we quickly discard T-Trees because this method shows lower predictive performance as compared to other methods, while being computationally demanding.
 Then, we show that the penalized logistic regression consistently achieves higher predictive performance while being faster than the C+T method, whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture, as shown in \cite{ware2017heterogeneity}.
@@
 Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
-As for simulations presented here, in order to remove the genetic structure induced by the celiac disease status, we first restrict our simulations to controls. 
+For simulations presented later, we first restrict the dataset to controls in order to remove the genetic structure induced by the celiac disease status. 
 Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.
@@
 
-The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Study (GWAS), referred to as summary statistics.
-In the GWAS, a coefficient of regression (estimated effect size) is learned independently for each SNP along with a corresponding p-value. 
+The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Study (GWAS).
+In the GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$. 
 The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another, which form a set of SNPs denoted $S_\text{clumping}$. 
@@
 
-For the C+T method, we report three different scores of prediction: one including all the SNPs (remaining after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximizes the AUC for these two thresholds and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
+For the C+T method, we report three different scores of prediction: one including all the SNPs remaining after clumping (``PRS-all''), one including only SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``PRS-stringent''), and one that maximizes the AUC (``PRS-max'') for these two thresholds ($0$ and $5 \cdot 10^{-8}$) and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
 As we report the optimal threshold based on the test set, the corresponding AUC value is an upper bound of the AUC for the C+T method.
-We call these three reported scores respectively ``PRS-all'', ``PRS-stringent'' and ``PRS-max'' in the results.
 
@@
 
-Finally, for the penalized logistic regression, we solve: $$\arg\!\min_{\beta_0,~\beta}(\lambda, \alpha)\left\{  \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization}  \right\}$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that needs to be determined.
+Finally, for the penalized logistic regression, we solve: $$\arg\!\min_{\beta_0,~\beta}(\lambda, \alpha)\left\{  \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization}  \right\}$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict, $\lambda$ and $alpha$ are two regularization parameters that need to be chosen.
 Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
@@
 When there are 30 causal SNPs and 80\% of heritability, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}).
-Moreover, the penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted in simulations with 3000 causal SNPs, which are cases where all methods hardly exceed 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
+Moreover, the penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted in some simulations with 3000 causal SNPs, which are cases where all methods hardly exceed 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
 
@@
 In this comparative study, we propose penalized logistic regressions as a replacement for the widely used C+T method for computing polygenic risk scores when individual genotype data are available. 
-We show that jointly estimating SNP effects using efficient penalized logistic regressions enables to achieve higher predictive performance, as reported in a previous study \cite[]{abraham2013performance}.
+We show that jointly estimating SNP effects using efficient penalized logistic regressions can achieve higher predictive performance, as reported in \cite{abraham2013performance}.
 Indeed, when disease architecture and sample size allow for high predictive performance (say AUC greater than 85\%), ``logit-simple'' provides higher predictive performance than the widely-used C+T method, nearly reaching optimal AUC values, which is encouraging for using polygenic risk scores in clinical settings.
@@
 However, in order to accurately predict disease status for clinical interventions, best possible predictive methods should be used. 
-The logistic regression we present in this paper could be used on data as large as biobank datasets because it is very efficient in both memory and computation time.
+The penalized logistic regression we present in this paper could be used on data as large as biobank datasets because it is both very memory- and computation-efficient.
 

### 2018-05-07 4-results-simus.Rmd
final figures

@@
 
-methods.color <- setNames(c(scales::hue_pal()(6), "black"), 
-                          c("logit-simple", "PRS-max", "PRS-stringent",
-                            "T-Trees", "logit-triple", "PRS-all", "biglasso")) 
+methods.color <- setNames(
+  c(scales::hue_pal()(6), c("black", "blue", "green")), 
+  c("logit-simple", "PRS-max", "PRS-stringent", "T-Trees", "logit-triple", "PRS-all", 
+    "biglasso", "logit-simple-enet-ridge", "logit-simple-lasso")
+) 
 ```

### 2018-05-08 some-tests/test-capture-interaction.R
some tests

@@ -1,0 +1,35 @@
+N <- 1000
+
+test <- replicate(1000, {
+  f1 <- runif(1)
+  f2 <- runif(1)
+  g1 <- rbinom(N, size = 2, prob = f1)
+  g1.1 <- g1 > 0.5
+  g1.2 <- g1 > 1.5
+  g2 <- rbinom(N, size = 2, prob = f2)
+  g2.1 <- g2 > 0.5
+  g2.2 <- g2 > 1.5
+  
+  g12 <- scale(cbind(g1, g1.1, g1.2, g2, g2.1, g2.2))
+  
+  g <- scale(g1) * scale(g2)
+  
+  g12.sub <- g12[, attr(g12, "scaled:scale") > 1e-4, drop = FALSE]
+  mod <- glmnet::glmnet(g12.sub, g, alpha = 0.01)
+  # coef(mod)
+  
+  pred <- predict(mod, g12.sub)
+  corr <- apply(pred, 2, cor, g)
+  # plot(pred[, which.max(corr)], g)
+  
+  c(f1, f2, max(corr[-1]))
+})
+
+plot(test[1, ], test[3, ], pch = 20)
+plot(abs(test[1, ] - test[2, ])^2, test[3, ], pch = 20)
+rgl::plot3d(test[1, ], test[2, ], test[3, ])
+
+library(ggplot2)
+qplot(test[1, ], test[2, ], col = test[3, ]) + 
+  viridis::scale_color_viridis() + 
+  coord_equal()

### 2018-05-08 some-tests/test-pruning-pool.R

@@ -34,7 +34,6 @@ scree_plot(obj.pool, 4)
 plot(obj.pool$loadings[, 1], pch = 20)
 plot(obj.pool$loadings[, 2], pch = 20)
 plot(obj.pool$loadings[, 3], pch = 20)
-plot(obj.pool$loadings[, 4], pch = 20)
 
 plink <- download_plink()
 file.base <- "../Dubois2010_data/FinnuncorrNLITUK1UK3hap300_QC_norel"
@@ -47,23 +46,32 @@ ind.keep <- match(to_keep, celiac$map$marker.ID)
 pool_mat.sub <- read.pcadapt(t(afs[ind.keep, ]), type = "pool")
 pcadapt.sub <- pcadapt(pool_mat.sub)
 plot(pcadapt.sub, chr.info = celiac$map$chromosome[ind.keep])
-scree_plot(pcadapt.sub, 5)
+scree_plot(pcadapt.sub, 4)
 plot(pcadapt.sub$loadings[, 1], pch = 20)
 plot(pcadapt.sub$loadings[, 2], pch = 20)
 plot(pcadapt.sub$loadings[, 3], pch = 20)
 
-obj.pcadapt <- snp_pcadapt(G, obj.svd$u)
-snp_manhattan(obj.pcadapt, CHR, POS, npoints = 50e3)
-
-tmat <- scale(t(afs))
-svd <- svd(tmat[, ind.keep])
+tmat <- scale(t(afs), center = TRUE, scale = FALSE)
+ind.keep2 <- ind.keep[seq(1, length(ind.keep), by = 5)]
+svd <- svd(tmat[, ind.keep2])
 w <- crossprod(tmat, sweep(svd$u, 2, svd$d, '/')[, 1:3])
-plot(w[ind.keep, ], svd$v[, 1:3], pch = 20)
+
+plot(w[ind.keep2, ], svd$v[, 1:3], pch = 20)
 abline(0, 1, col = "red")
+
+plot(w[, 1], pch = 20)
+plot(w[, 2], pch = 20)
+plot(w[, 3], pch = 20)
 
 res <- pcadapt:::get_statistics(w, method = "mahalanobis", 
                                 pass = rows_along(w))
 plot(-log10(res$pvalues), pch = 20)
 
-obj.pcadapt.sub <- snp_pcadapt(G, obj.svd$u, ind.col = ind.keep)
[…]

### 2018-05-08 some-tests/test-varimax.R

@@ -1,0 +1,198 @@
+
+library(bigsnpr)
+
+celiac <- snp_attach("backingfiles/celiacQC.rds")
+G <- celiac$genotypes
+CHR <- celiac$map$chromosome
+POS <- celiac$map$physical.pos
+obj.svd <- readRDS("backingfiles/PCA.rds")
+plot(obj.svd) + 
+  ggplot2::scale_y_log10(breaks = c(500, 1000, 2000))
+
+files <- list.files("backingfiles/others/", "^cluster",
+                    full.names = TRUE)
+pop <- snp_getSampleInfos(celiac, files)[[1]]
+pop[pop == 4] <- 3
+library(ggplot2)
+plot(obj.svd, type = "scores", scores = 3:4) +
+  aes(color = as.factor(pop))
+
+
+test <- varimax(tmp$u)
+plot(test$loadings)
+plot(test$loadings[, 3:4])
+
+
+tmp2 <- prcomp(iris[1:4], center = TRUE, scale. = TRUE)
+plot(tmp2$x, pch = 20, col = iris$Species)
+rot_var <- varimax(tmp2$rotation[, 1:2])
+rot_var
+plot(rot_var$loadings)
+plot(tmp2$x[, 1:2] %*% rot_var$rotmat, pch = 20, col = iris$Species)
+
+# test on PCs of Celiac
+test <- varimax(tmp$v)
+test2 <- varimax(tmp$v[, 1:3])
+test2$rotmat
+test$rotmat[1:3, 1:3]
+str(test)     
+test$loadings
+plot(test$rotmat)
+test$rotmat
+plot(tmp$u, pch = 20)
+u_rot <- tmp$u %*% test$rotmat
+u_rot3 <- tmp$u[, 1:3] %*% test2$rotmat
+plot(u_rot, col = scales::alpha(pop, 0.4), pch = 20)
+plot(u_rot3, col = scales::alpha(pop, 0.4), pch = 20)
+plot(u_rot[, 3:4], col = scales::alpha(pop, 0.4), pch = 20)
+plot(u_rot[, 5:6], col = scales::alpha(pop, 0.4), pch = 20)
+u <- obj.svd$u
+plot(u[, 5:6], col = scales::alpha(pop, 0.4), pch = 20)
+
+
+test3 <- varimax(obj.svd$u, normalize = FALSE)
+u_rot2 <- u %*% test3$rotmat
+plot(u, col = scales::alpha(pop, 0.4), pch = 20)
[…]

### 2018-05-08 some-tests/test-varimax2.R

@@ -1,0 +1,55 @@
+library(bigsnpr)
+
+# snp_readBed("../2017_project_5/data/hapmap3_qc.bed")
+hapmap3 <- snp_attach("../2017_project_5/data/hapmap3_qc.rds")
+G <- hapmap3$genotypes
+CHR <- hapmap3$map$chromosome
+POS <- hapmap3$map$physical.pos
+
+counts <- big_counts(G)
+counts[, 1:4]
+dim(G)
+
+snp_fastImpute(G, CHR, ncores = nb_cores())
+
+rds <- "../2017_project_5/data/hapmap3_qc-infos-impute.rds"
+
+tmp <- big_attach(rds)
+ind.col <- which(!is.na(tmp[1, ]))
+
+G <- G$copy(code = bigsnpr:::CODE_IMPUTE_PRED)
+obj.svd <- snp_autoSVD(G, CHR, POS, ind.col = ind.col, k = 20,
+                       ncores = nb_cores())
+plot(obj.svd) + 
+  ggplot2::scale_y_log10()
+plot(obj.svd, type = "loadings", loadings = 1:6, coeff = 0.5)
+plot(obj.svd, type = "scores", scores = 9:10)
+
+test <- varimax(obj.svd$v) 
+u <- obj.svd$u
+u_rot <- u %*% test$rotmat
+
+# obj.pcadapt <- snp_pcadapt(G, u, ind.col = ind.col)
+# snp_qq(obj.pcadapt) + ggplot2::xlim(1, NA)
+# snp_manhattan(snp_gc(obj.pcadapt), 
+#               infos.chr = CHR[ind.col],
+#               infos.pos = POS[ind.col],
+#               npoints = 50e3) + 
+#   ggplot2::geom_hline(yintercept = -log10(5e-8), color = "red")
+
+infos <- data.table::fread("ftp://ftp.ncbi.nlm.nih.gov/hapmap/genotypes/hapmap3/plink_format/draft_1/relationships_w_pops_051208.txt", 
+                  data.table = FALSE)
+
+pop <- as.factor(infos$population[match(hapmap3$fam$sample.ID, infos$IID)])
+
+plot(u[, 1:2], col = pop, pch = 20)
+plot(u_rot[, 1:2], col = pop, pch = 20)
+plot(u_rot[, 3:4], col = pop, pch = 20)
+plot(u_rot[, 5:6], col = pop, pch = 20)
[…]

### 2018-05-08 some-tests/test-varimax3.R

@@ -1,0 +1,111 @@
+library(bigsnpr)
+
+celiac <- snp_attach("backingfiles/celiacQC.rds")
+hapmap3 <- snp_attach("../2017_project_5/data/hapmap3_qc.rds")
+
+# imputation
+rds <- "../2017_project_5/data/hapmap3_qc-infos-impute.rds"
+tmp <- big_attach(rds)
+
+hasNA <- is.na(tmp[1, ])
+ind <- match(hapmap3$map$marker.ID, celiac$map$marker.ID)
+obj.svd0 <- snp_autoSVD(
+  G = hapmap3$genotypes$copy(bigsnpr:::CODE_IMPUTE_PRED),
+  infos.chr = hapmap3$map$chromosome,
+  infos.pos = hapmap3$map$physical.pos,
+  ind.col = which(!is.na(ind) & !hasNA), 
+  k = 20, 
+  ncores = nb_cores()
+)
+
+plot(obj.svd0) + 
+  ggplot2::scale_y_log10()
+plot(obj.svd0, type = "loadings", loadings = 1:6, coeff = 0.5)
+plot(obj.svd0, type = "scores", scores = 9:10)
+
+test <- varimax(obj.svd0$v)
+
+G <- celiac$genotypes
+CHR <- celiac$map$chromosome
+POS <- celiac$map$physical.pos
+obj.svd <- readRDS("backingfiles/PCA.rds")
+u <- obj.svd$u
+
+u0 <- predict(obj.svd0, G, ind.col = ind[attr(obj.svd0, "subset")])
+u0_rot <- u0 %*% test$rotmat
+
+files <- list.files("backingfiles/others/", "^cluster",
+                    full.names = TRUE)
+pop <- snp_getSampleInfos(celiac, files)[[1]]
+pop[pop == 4] <- 3
+
+plot(u0, col = pop, pch = 20)
+plot(u0_rot, col = pop, pch = 20)
+plot(u0[, 3:4], col = pop, pch = 20)
+plot(u0_rot[, 3:4], col = pop, pch = 20)
+
+
+# PRS
+y <- celiac$fam$affection - 1
+ind.train <- sort(sample(nrow(G), 10e3))
+ind.test <- setdiff(rows_along(G), ind.train)
+system.time(
+  prs0 <- big_spLogReg(G, y01.train = y[ind.train], 
+                       ind.train = ind.train,
+                       covar.train = u[ind.train, ], 
[…]

### 2018-05-08 paper/draft.tex
some modifs by H

@@
 
-\title{Highly efficient implementation of penalized models and their application to genetic risk prediction}
+\title{Efficient implementation of penalized models and their application to genetic risk prediction}
 \author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
@@
 \abstract{
-Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. 
+Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
 PRSs might have a major public health impact, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.
-The most commonly used method to derive PRSs is called ``Clumping+Thresholding'' (C+T).
-It uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it very suitable and fast.
-However, these PRS rely on non-optimal heuristics.
+The ``Clumping+Thresholding'' (C+T) approach, which is is the most common method to derive PRS, uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it suitable and fast.
+However, previous work showed that jointly estimating SNP effects for computing PRS have the potential to significantly improve the predictive performance of PRS as compared to C+T.
 
-In R package bigstatsr, we implemented linear and logistic regression models that can be used to jointly estimate SNP effects for computing PRS.
-We use an efficient algorithm with a new implementation in order for this model to be used on huge datasets.
+In this paper, we present an efficient implementation of such penalized joint estimation, allowing for practical application of penalized logistic regression on modern datasets including hundreds of thousands of individuals.
+Our implementation of penalized logistic regression also includes nearly-optimal choices for hyper-parameters.
+The choice of hyper-parameters for a method is very important as it can dramatically impact the predictive performance of the method (e.g.
+the p-value threshold in C+T).
 
-In this paper, we present a comprehensive comparative study of the C+T method, our penalized logistic regression and a derivation of random forests.
-The penalized logistic regression consistently achieves higher predictive performance than the two other methods while being faster than the C+T method.
-Moreover, predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, whose optimal value depends on disease architecture. 
-The improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs, and with an increasing sample size. 
-When analyzing real SNP data from a case-control study for celiac disease, penalized logistic regression provides an AUC of 89\% while the standard C+T method provides a smaller AUC of 82\%.
+We compare the performance of penalized logistic regression to the C+T method and to a derivation of random forests.
+The penalized logistic regression consistently achieves higher predictive performance than the two other methods while being faster.
+We also found that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; AUC values increase from 83\% with C+T to 92.5\% with penalized logistic regression.
+We confirmed these results in a data analysis of a case-control study for celiac disease where penalized logistic regression and the standard C+T method achieve AUC of 89\% and of 82\%.
 
-[TOMOD] Instead of computing polygenic risk scores based on GWAS univariate summary statistics, we recommend to use penalized logistic regression as implemented in the R package bigstatsr to obtain more accurate polygenic risk scores when individual genotype data are available.\\
+In conclusion, our study demonstrates that penalized logistic regression are applicable to large-scale individual-level data and can achieve more discriminative polygenic risk scores.
+Our implementation is publicly available in the R package bigstatsr.\\
 
@@
 A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or even just PRS -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
-This technique uses GWAS summary statistics only, which makes it very convenient to compute and also very fast.
+This technique uses GWAS summary statistics only, allowing for trivial and fast implementation.
+However, the approach also has limitations.
+In particular, previous studies showed that predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite{ware2017heterogeneity}.
 Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}.
 Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes \cite[]{abraham2013performance}.
-Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time requirements, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
+Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
 Finally, statistical learning methods have also been used to derive PRSs for complex human diseases by jointly estimating SNP effects.
@@
 Package bigstatsr includes an efficient algorithm with a new implementation for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
-In this paper, we present a comprehensive comparative study of the C+T method, our implementation of penalized logistic regression and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
+In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression against the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
 Note that in this comparison we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
 For the C+T model, we compare different thresholds of inclusion of SNPs.
-For the penalized logistic regression, we include two practical improvements. 
-First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for automatically choosing the regularization parameter of the penalized logistic regression, which directly affects the number of SNPs included in the model.
+For the penalized logistic regression, we also introduce two improvements.
+First, we present a procedure that we call Cross-Model Selection and Averaging (CMSA) for automatically choosing the regularization parameter of the penalized logistic regression, which directly affects the number of SNPs included in the model.
 Secondly, we also show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.
 
-There have already been comparative studies on predicting human binary phenotypes based on genotype data. 
-\cite{abraham2013performance} showed that sparse penalized methods performed better than the standard C+T method and LMMs, for a wide-range of human diseases.
-\cite{zhou2013polygenic} used simulations on real genotypes of Australian individuals to show that a hybrid of linear mixed models and sparse regression models performed better in a range of sparse to polygenic disease architectures.
-\cite{spiliopoulou2015genomic} investigated the effect of relatedness in prediction performance.
-\cite{ware2017heterogeneity} evaluated best practices for using the C+T method and found out that the optimal inclusion threshold is trait-dependent. 
-[TOMOD] In this study, we particularly extend the work of \cite{abraham2013performance} by using a simulation framework similar but more extenstive than the one used in \cite{zhou2013polygenic}. 
+For the simulations of this paper, we use real genotype data and simulate new phenotypes \cite{zhou2013polygenic}. 
 In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the heritability.
-We also compare different models for generating liability scores, one with only linear effects, and one which combines linear, dominant and interaction-type effects.
-
-[TOMOD] First, we quickly discard T-Trees because this method shows lower predictive performance as compared to other methods, while being computationally demanding.
-Then, we show that the penalized logistic regression consistently achieves higher predictive performance while being faster than the C+T method, whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture, as shown in \cite{ware2017heterogeneity}.
+We also compare different models for simulating new phenotypes, one with only linear effects, and one which combines linear, dominant and interaction-type effects.
+Overall, we found that the penalized logistic regression consistently achieves higher predictive performance while being faster than the C+T and T-Trees methods.
+This demonstrates the feasibility and relevance of this approach in PRS modeling in future studies.
 
@@
 
-We use real genotypes of European individuals from a case-control celiac disease study \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}.
+We use real genotypes of European individuals from a case-control celiac disease study \cite[]{dubois2010multiple}.
+The composition of this dataset is presented in table \ref{tab:celiac-data}.
 Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
 For simulations presented later, we first restrict the dataset to controls in order to remove the genetic structure induced by the celiac disease status.
-Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.
-In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on the robust Mahalanobis distances computed using the first 10 principal components of remaining individuals.
+Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on the robust Mahalanobis distances computed using the first 10 principal components of remaining individuals.
 This 3-step filtering process results in a sample of 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing'').
+We also use this dataset for a real data application, in this case keeping all individuals.
 
@@
 
-In order to simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}.
-We vary different parameters: the number of chromosomes used (either all 22 chromosomes or chromosome 6 only), the size of the training set (from 1000 to 6000), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace).
-We also use two different models to compute liability scores: a ``simple'' model with linear effects only, and a ``fancy'' model which combines linear, dominant and interaction-type effects.
-For the ``simple'' model, we compute $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights generated with a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs).
+We simulate binary phenotypes using a Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}.
+We vary parameters of the simulations in order to match a range of genetic architecture, from low to high polygenicity.
+For practical reason, this was achieved by varying the number of chromosomes used (either all 22 autosomal chromosomes or chromosome 6 only), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), and the heritability (50\% or 80\%).
+We also considered deviation form the standard normal additive model, drawing effects of causal SNPs from either Normal or Laplace distribution, and computing liability scores either from a ``simple'' model with linear effects only or a ``fancy'' model which combines linear, dominant and interaction-type effects.
+For the ``simple'' model, we compute $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i$$ where $w_j$ are weights generated with a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs) and $\epsilon \sim N(0, 1 - h^2)$.
 For the ``fancy'' model, we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$
 and $S_\text{causal}^{(3)}$; $S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$.
-We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$
+We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction} ~~+~~ \epsilon_i$$
 where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$.
-Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still have a marginal effect. 
-For both models, the values of $y_i$ are then standardized so that they have a variance equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$, where $\epsilon \sim N(0, 1 - h^2)$.
+Note that, because we used the raw genotype counts when coding the interaction terms, the corresponding SNPs also display a marginal effect.
 
@@
 Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes) and a training set of size 6000.
-It compares all methods described in section \ref{sec:methods}; for each combination of the remaining parameters, results are based on 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}).
-However, comparison results for T-Trees relies on 5 simulations only (and an heritability of 80\%) because this method can require several hours of computation for a single simulation. 
+It compares all methods described in section \ref{sec:methods}; for each combination of the remaining parameters, results are based on 100 simulations except  for T-Trees, which relies on 5 simulations only (and an heritability of 80\%) because of a much higher computational burden (several hours of computation for a single simulation) as compared to other approaches.
 Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only.
@@
 In this study, we use two different measures of predictive accuracy.
-First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction,hanley1982meaning,lusted1971signal}.
+First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction,lusted1971signal}.
 In the case of our study, the AUC is the probability that a PRS of a case is greater than the PRS of a control.
@@
 This measure is similar to the AUC, but focuses on high specificities, which is the most useful part of the ROC curve in clinical settings.
-
 When reporting results of AUC, we use estimates of maximum achievable AUC values for the simulations: 84\% and 94\% for heritabilities of respectively 50\% and 80\% based on three different estimations (see Supplementary Materials).
-
 %Note that we also report the timing of the main computations and the number of SNPs used in the predictions.
@@
 When there are 30 causal SNPs and 80\% of heritability, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}).
-Moreover, the penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted in some simulations with 3000 causal SNPs, which are cases where all methods hardly exceed 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
+Moreover, the penalized logistic regression consistently provides better predictive performance than the C+T method across all scenarios we considered, except in the case of high polygenicity and small sample size, in which case all methods hardly exceed 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
 
@@
 Yet, as the value of $\lambda$ depends on the sample size and doing cross-validation and retraining is computationally demanding, we prefer to use CMSA when there is a model to choose from a regularization path, with the possibility of adding an early stopping criterion to fasten the whole procedure.
-
-We also introduce a simple feature engineering technique which enables the logistic regression to detect and learn not only additive effects, but also dominant and recessive ones [DEJA ETE FAIT?
-CITATION?].
+We also introduce a simple feature engineering technique which enables the logistic regression to detect and learn not only additive effects, but also dominant and recessive ones.
 This technique improves the predictive performance of the method when there are some non-linear effects in the simulations, while providing the same predictive performance when there are only linear effects.
-Moreover, it also improves predictive performance for the celiac disease.
-
-Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}.
+Moreover, it also improves predictive performance for the celiac disease.Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}.
 In another study, we will assess how can we use models and effects learned in one population to improve learning and prediction in another population.
@@
 \begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-corr}}
+\caption{Correlation between AUC and partial AUC values in scenario \textnumero1.
+There is a Spearman correlation of 98\% between values of AUC and partial AUC.
+The relation between the two values are the same whatever are the heritability, distribution of effects and method used.\label{fig:supp-AUC-corr}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newpage
+\begin{figure}[h]
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6}}
@@
 \label{fig:supp-AUC-PRS}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-triple}}
+\caption{[TRIPLE ONLY AUC - H2=0.5]}
+\label{fig:supp-AUC-triple}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-logit-fancy}}
+\caption{[COPY FOR H2=0.5 \& MODEL FANCY]}
+\label{fig:supp-AUC-logit-fancy}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS-fancy}}
+\caption{[COPY FOR H2=0.5 \& MODEL FANCY]}
+\label{fig:supp-AUC-PRS-fancy}
 \end{figure}
@@
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-triple}}
-\caption{[TRIPLE ONLY AUC - H2=0.5]}
-\label{fig:supp-AUC-triple}
+\centerline{\includegraphics[width=\textwidth]{supp-biglasso}}
+\caption{Comparison of ``logit-triple'' and the best prediction for ``biglasso'' (another implementation of penalized logistic regression without procedure for choosing the $\lambda$ hyper-parameter) in scenario \textnumero1.
+Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
+Horizontal panels are presenting results for the ``simple'' model for simulating phenotypes.
+Simulations use an heritability of 80\%.
+\textbf{A:} Mean of AUC over 100 simulations.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+The blue dotted line represents the maximum achievable AUC.
+\textbf{B:} Boxplots of numbers of predictors used by the methods for 100 simulations.
+\textbf{C:} Boxplots of execution times for 100 simulations.}
+\label{fig:supp-biglasso}
 \end{figure}
@@
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-logit-fancy}}
-\caption{[COPY FOR H2=0.5 \& MODEL FANCY]}
-\label{fig:supp-AUC-logit-fancy}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS-fancy}}
-\caption{[COPY FOR H2=0.5 \& MODEL FANCY]}
-\label{fig:supp-AUC-PRS-fancy}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-corr}}
-\caption{Correlation between AUC and partial AUC values in scenario \textnumero1.
-There is a Spearman correlation of 98\% between values of AUC and partial AUC.
-The relation between the two values are the same whatever are the heritability, distribution of effects and method used.\label{fig:supp-AUC-corr}}
+\centerline{\includegraphics[width=\textwidth]{supp-alpha}}
+\caption{Comparison of different values of the $\alpha$ hyper-parameter of the ``logit-simple'' method in scenario \textnumero1.
+``logit-simple'', ``logit-simple-enet-ridge'' and ``logit-simple-lasso'' corresponds to values of 0.5, 0.001 and 1, respectively.
+Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
+Horizontal panels are presenting results for the ``simple'' model for simulating phenotypes.
+Simulations use an heritability of 80\%.
+\textbf{A:} Mean of AUC over 100 simulations.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+The blue dotted line represents the maximum achievable AUC.
+\textbf{B:} Boxplots of numbers of predictors used by the methods for 100 simulations.
+\textbf{C:} Boxplots of execution times for 100 simulations.}
+\label{fig:supp-alpha}
 \end{figure}

### 2018-05-10 paper/draft.tex
modif discussion

@@
 
-In this comparative study, we propose penalized logistic regressions as a replacement for the widely used C+T method for computing polygenic risk scores when individual genotype data are available.
-We show that jointly estimating SNP effects using efficient penalized logistic regressions can achieve higher predictive performance, as reported in \cite{abraham2013performance}.
-Indeed, when disease architecture and sample size allow for high predictive performance (say AUC greater than 85\%), ``logit-simple'' provides higher predictive performance than the widely-used C+T method, nearly reaching optimal AUC values, which is encouraging for using polygenic risk scores in clinical settings.
-The C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not yet the case for our penalized logistic regressions.
-However, in order to accurately predict disease status for clinical interventions, best possible predictive methods should be used.
-The penalized logistic regression we present in this paper could be used on data as large as biobank datasets because it is both very memory- and computation-efficient.
+\subsection{Joint estimation improves predictive performance}
 
-As for the C+T method itself, we recall that we report the maximum AUC of 102 thresholds, a threshold that should normally be learned on the training set only.
-The choice of this threshold is very important as it can greatly impact the predictive performance of the method \cite[]{ware2017heterogeneity}.
-On the contrary, in the penalized logistic regression presented here, we remove the choice of the regularization parameter with a procedure that we call Cross-Model Selection and Averaging (CMSA) while keeping high predictive performance [TODO: COMPARAISON AVEC LE MAX DE BIGLASSO].
-Usually, cross-validation is used to assess the best value for the hyper-parameter and then the model is train again with this particular hyper-parameter value \cite[]{Hastie2008,wei2013large}.
-Yet, as the value of $\lambda$ depends on the sample size and doing cross-validation and retraining is computationally demanding, we prefer to use CMSA when there is a model to choose from a regularization path, with the possibility of adding an early stopping criterion to fasten the whole procedure.
-We also introduce a simple feature engineering technique which enables the logistic regression to detect and learn not only additive effects, but also dominant and recessive ones.
+In this comparative study, we present a computationally efficient implementation of penalized logistic regression.
+This model can be used to build polygenic risk scores based on very large SNP datasets such as the UK biobank \cite[]{bycroft2017genome}.
+In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to dramatically improve predictive performance as compared to the standard C+T approach in which these effects are learned independently.
+Penalized logistic regression consistently outperform the C+T method, excepted in the case where there are only small effects and sample size is limited, in which case all methods perform poorly.
+Note that in this particular case, penalized logistic regression could outperform the C+T method if choosing a different value for hyper-parameter $\alpha$.
+
+\subsection{Importance of hyper-parameters}
+
+Indeed, the choice of hyper-parameters' values is very important as it can greatly impact the performance of the methods.
+In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps, respectively.
+The choice of the $r^2$ threshold of the clumping step is important.
+Indeed, on the one hand, choosing a low value for this threshold may discard independently predictive SNPs that are in LD; yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would bias SNP effects.
+Moreover, in this paper, we reported the maximum AUC of 102 $p_T$ thresholds, a threshold that should normally be learned on the training set only.
+The choice of this threshold is very important as it can greatly impact the predictive performance of the C+T method \cite[]{ware2017heterogeneity}, which we confirm in this study.
+
+On the contrary, in the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing an optimal hyper-parameter $\lambda$, accouting for the amount of regularization used in the model.
+Not only this procedure provides near-optimal results, but it also accelerates the training of the model thanks to the development of an early stopping criterion.
+Usually, cross-validation is used to assess the best value for this hyper-parameter and then the model is train again with this particular hyper-parameter value \cite[]{Hastie2008,wei2013large}.
+Yet, as the value of $\lambda$ depends on the sample size and doing cross-validation and retraining is computationally demanding, we think using CMSA is more appropriate in this setting.
+Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we show that an hybrid approach ($\alpha = 0.5$) gives similar results as a model using only L1 regularization ($\alpha = 1$).
+Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model.
+Yet, when testing with $\alpha = 0.001$, we show that this setting consistently provides lower predictive performance, excepted in the case where there are many small effects.
+In the light of this result, we now integrate an optional grid-search procedure in our implementation in order to determine an optimal value for hyper-parameter $\alpha$.
+
+\subsection{Non-linear effects}
+
+In this paper, we also explored how to capture non-linear effects.
+For this, we introduced a simple feature engineering technique that enables the logistic regression to detect and learn not only additive effects, but also dominant and recessive effects.
 This technique improves the predictive performance of the method when there are some non-linear effects in the simulations, while providing the same predictive performance when there are only linear effects.
-Moreover, it also improves predictive performance for the celiac disease.Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}.
+Moreover, it also improves predictive performance for the celiac disease.
+
+Yet, this approach is not able to detect interaction-type effects. 
+In order to capture interaction-type effects, we tested T-Trees, a method that is able to exploit SNP correlations thanks to special decision trees \cite[]{botta2014exploiting}. 
+However, predictive performance of T-Trees were consistently lower than penalized logistic regression, even when simulating a fancy model in which there are dominant and interaction-type effects that T-Trees should be able to handle.
+
+\subsection{Limitations}
+
+Our approach has one major limitations: the main advantage of the C+T method is that it is applicable directly on summary statistics, allowing to leverage the largest GWAS sample size to date, even when individual cohort data cannot be merged because of practical and ethical reasons (e.g.
+consortium data including hundreds of cohorts).
+As of today, the proposed penalized logistic regression does not allow for the analysis of summary data, but represents an important future direction of our work.
+The current version is of particular interest for the analysis of modern SNP dataset including hundreds of thousands of individuals.
+
+Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}.
 In another study, we will assess how can we use models and effects learned in one population to improve learning and prediction in another population.

### 2018-05-28 paper/draft.tex
add some captions
some modifs

@@
 
-\subsection{Comparison between C+T and ``logit-simple'' methods}
+\subsection{Joint estimation improves predictive performance}
 
@@
 
-When there are 30 causal SNPs and 80\% of heritability, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}).
+When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}).
 Moreover, the penalized logistic regression consistently provides better predictive performance than the C+T method across all scenarios we considered, except in the case of high polygenicity and small sample size, in which case all methods hardly exceed 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
 
-Method ``logit-simple'' provides particularly higher predictive performance than ``PRS-max'' where there are correlations between predictors, i.e.\ when we simulate 30 SNPs in the HLA region.
+Method ``logit-simple'' provides particularly higher predictive performance than ``PRS-max'' where there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region.
 In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' and 83\% with the ``PRS-max'', for a maximum achievable AUC of 94\% (Figure \ref{fig:main-AUC-logit}).
+
+Note that for the simulations we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).
+
+\subsection{Importance of hyper-parameters}
 
@@
 
-Note that for the simulations we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).
+Method ``logit-simple'', that automatically chooses hyper-parameter $\lambda$, provides similar predictive performance than the best predictive performance of the biglasso implementation \cite[]{zeng2017efficient}, only slightly better for biglasso but likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).
 
-\subsection{The other methods and the ``fancy'' model}
+When comparing different values of hyper-parameter $\alpha$ for ``logit-simple'', we report similar predictive performance for $\alpha = 0.5$ and $\alpha = 1$.
+The case of $\alpha = 0.001$ consistently provides lower predictive performance, except in the case where there are 3000 causal SNPs whose effects follow a normal distribution (Figure \ref{fig:supp-alpha}).
+Note that this is also the only case where ``PRS-all'' outperforms logit-simple.
+
+\subsection{Non-linear effects}
 
 We tested the T-Trees method in scenario \textnumero1.
-As compared to the ``logit-simple'' method, T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}).
-Even when simulating a fancy model in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower with T-Trees as compared with the ``logit-simple'' method.
+As compared to ``logit-simple'', T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}).
+Even when simulating a fancy model in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower with T-Trees as compared with ``logit-simple''.
 
-Secondly, we compare the two penalized logistic regressions in scenario \textnumero1, ``logit-simple'' and ``logit-triple'' that uses additional features coding for recessive and dominant effects.
-We report that ``logit-triple'' is nearly as good as ``logit-simple'' when there are only linear effects (differences of AUC smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
+Secondly, we compare the two penalized logistic regressions in scenario \textnumero1, ``logit-simple'' and ``logit-triple'' that uses additional features (variables) coding for recessive and dominant effects.
+Predictive performance of ``logit-triple'' are nearly as good as ``logit-simple'' when there are only linear effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
 For the ``fancy model'', ``logit-triple'' provides AUC values at least 3.5\% higher than ``logit-simple'', except when there are 3000 causal SNPs.
@@
 Thirdly, for the ``fancy'' simulation model, we report similar results as in the previous subsection.
-As before, ``logit-simple'' achieves greater AUCs than the ``PRS-max'', excepted in the case of 3000 causal SNP anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}).
+As before, ``logit-simple'' achieves greater AUCs than the ``PRS-max'', except in the case of 3000 causal SNP anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}).
 Moreover, the choice of the threshold inclusion of SNPs also greatly impacts the performance of the C+T method (Figure \ref{fig:supp-AUC-PRS-fancy}).
@@
 
-These simulations correspond to scenarios \textnumero2 and \textnumero3 (Table \ref{tab:simus}).
-First, when reproducing simulations of scenario \textnumero1 using only chromosome 6, the predictive performance of all methods globally increase (Figure \ref{fig:supp-AUC-chr6}).
+First, when reproducing simulations of scenario \textnumero1 using only chromosome 6 (scenario \textnumero2), the predictive performance of all methods globally increase (Figure \ref{fig:supp-AUC-chr6}).
 There is particularly a large increase when simulating 3000 causal SNPs anywhere in the dataset: AUC from the ``logit-simple'' increases from less than 60\% to nearly 80\%. 
 On the contrary, when simulating only 30 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' does not increase. 
-Secondly, when varying the training size, we report an increase of AUC when increasing the training size, with a faster increase of AUC provided by ``logit-simple'' as compared to ``PRS-max'' (Figure  \ref{fig:main-AUC-ntrain}).
+Secondly, when varying the training size (scenario \textnumero3), we report an increase of AUC when increasing the training size, with a faster increase of AUC provided by ``logit-simple'' as compared to ``PRS-max'' (Figure  \ref{fig:main-AUC-ntrain}).
 
@@
 Joint logistic regressions provide higher AUC for the Celiac data: 88.8\% with ``logit-simple'' and 89.2\% with ``logit-triple'' as compared to 82.4\% with the C+T method.
-The relative differences in partial AUC, for specificities larger than 90\%, are even larger with values of 0.0414, 0.0429 and 0.0286 obtained with ``logit-simple'', ``logit-triple'' and the C+T method, respectively.
+The relative increase in partial AUC, for specificities larger than 90\%, is even larger (45\% and 50\%) with partial AUC values of 0.0414, 0.0429 and 0.0286 obtained with ``logit-simple'', ``logit-triple'' and the C+T method, respectively.
 Moreover, logistic regressions use less predictors, respectively 3320, 4470 and 9850 (Table \ref{tab:results-celiac}, figures \ref{fig:celiac-roc} and \ref{fig:supp-score-densities} and supplementary notebook ``results-celiac'').
@@
 In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to dramatically improve predictive performance as compared to the standard C+T approach in which these effects are learned independently.
-Penalized logistic regression consistently outperform the C+T method, excepted in the case where there are only small effects and sample size is limited, in which case all methods perform poorly.
+Penalized logistic regression consistently outperform the C+T method, except in the case where there are only small effects and sample size is limited, in which case all methods perform poorly.
 Note that in this particular case, penalized logistic regression could outperform the C+T method if choosing a different value for hyper-parameter $\alpha$.
@@
 
-On the contrary, in the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing an optimal hyper-parameter $\lambda$, accouting for the amount of regularization used in the model.
+On the contrary, in the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$ that accounts for the amount of regularization used in the model.
 Not only this procedure provides near-optimal results, but it also accelerates the training of the model thanks to the development of an early stopping criterion.
@@
 Yet, as the value of $\lambda$ depends on the sample size and doing cross-validation and retraining is computationally demanding, we think using CMSA is more appropriate in this setting.
-Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we show that an hybrid approach ($\alpha = 0.5$) gives similar results as a model using only L1 regularization ($\alpha = 1$).
+Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we show that an hybrid approach ($\alpha = 0.5$) gives similar results as a model using only L1 regularization ($\alpha = 1$) in all simulated models.
 Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model.
-Yet, when testing with $\alpha = 0.001$, we show that this setting consistently provides lower predictive performance, excepted in the case where there are many small effects.
-In the light of this result, we now integrate an optional grid-search procedure in our implementation in order to determine an optimal value for hyper-parameter $\alpha$.
+Yet, when testing with $\alpha = 0.001$, we show that this setting consistently provides lower predictive performance, except in the case where there are many small effects and small sample size, which is a case where all methods perform poorly.
+%In light of this result, we now integrate an optional grid-search procedure in our implementation in order to determine an optimal value for hyper-parameter $\alpha$.
 
@@
 
-Our approach has one major limitations: the main advantage of the C+T method is that it is applicable directly on summary statistics, allowing to leverage the largest GWAS sample size to date, even when individual cohort data cannot be merged because of practical and ethical reasons (e.g.
+Our approach has one major limitation: the main advantage of the C+T method is that it is applicable directly to summary statistics, allowing to leverage the largest GWAS sample size to date, even when individual cohort data cannot be merged because of practical and ethical reasons (e.g.
 consortium data including hundreds of cohorts).
@@
 We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}.
-In another study, we wish also to assess how can we combine the information provided by genetic data with clinical and environmental data, possibly in a non-linear way.
+Finally, we also wish to assess how can we combine the information provided by genetic data with clinical and environmental data, possibly in a non-linear way.
 
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-logit}}
-\caption{[COPY FOR H2=0.5]}
+\caption{Comparison of the C+T method and ``logit-simple'' in scenario \textnumero1.
+Mean of AUC over 100 simulations for ``logit-simple'' and the maximum AUC reported with the C+T method (``PRS-max'').
+Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+The blue dotted line represents the maximum achievable AUC.
+These results corresponds to the ``simple'' model and an heritability of 50\%.}
 \label{fig:supp-AUC-logit}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS}}
-\caption{[COPY FOR H2=0.5]}
+\caption{Comparison of three different threshold used in the C+T method in scenario \textnumero1.
+Mean of AUC over 100 simulations.
+Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+The blue dotted line represents the maximum achievable AUC.
+These results corresponds to the ``simple'' model and an heritability of 50\%.}
 \label{fig:supp-AUC-PRS}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-triple}}
-\caption{[TRIPLE ONLY AUC - H2=0.5]}
+\caption{Comparison of ``logit-triple'' and ``logit-simple'' in scenario \textnumero1.
+Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
+Horizontal panels are presenting results for the ``simple'' and ``fancy'' models for simulating phenotypes.
+Simulations use an heritability of 50\%.
+\textbf{A:} Mean of AUC over 100 simulations.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+The blue dotted line represents the maximum achievable AUC.}
 \label{fig:supp-AUC-triple}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-logit-fancy}}
-\caption{[COPY FOR H2=0.5 \& MODEL FANCY]}
+\caption{Comparison of the C+T method and ``logit-simple'' in scenario \textnumero1.
+Mean of AUC over 100 simulations for ``logit-simple'' and the maximum AUC reported with the C+T method (``PRS-max'').
+Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
+Left (right) panel is presenting results for an heritability of 50\% (80\%).
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+The blue dotted line represents the maximum achievable AUC.
+These results corresponds to the ``fancy'' model.}
 \label{fig:supp-AUC-logit-fancy}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS-fancy}}
-\caption{[COPY FOR H2=0.5 \& MODEL FANCY]}
+\caption{Comparison of three different threshold used in the C+T method in scenario \textnumero1.
+Mean of AUC over 100 simulations for ``logit-simple'' and the maximum AUC reported with the C+T method (``PRS-max'').
+Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
+Left (right) panel is presenting results for an heritability of 50\% (80\%).
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+The blue dotted line represents the maximum achievable AUC.
+These results corresponds to the ``fancy'' model.}
 \label{fig:supp-AUC-PRS-fancy}

### 2018-06-07 3-simus.Rmd
some modifs by H
update

@@
 library(glue)
-NCORES <- 6
+library(bigsnpr)
+# https://stackoverflow.com/q/9486952/6103040
+options(bigstatsr.cluster.type = "PSOCK")
 ```
@@
 logit.CMSA <- function(G, pheno.all, covar.all, ind.train, ind.test,
-                       method = "logit-simple", alpha = 0.5) {
+                       method = "logit-simple", alphas = 0.5) {
 
@@
                                covar.train = covar.all[ind.train, , drop = FALSE],
-                               alpha = alpha, ncores = NCORES)
+                               alphas = alphas, ncores = NCORES)
 
@@
 ```{r}
-library(bigsnpr)
 celiac2 <- snp_attach("backingfiles/celiacQC_sub1.rds")
@@
 ind.HLA <- snp_indLRLDR(CHR, POS, subset(LD.wiki34, ID == "hild12"))
+
+NCORES <- nb_cores()
 ```
@@
 ```{r, eval=FALSE}
-G3 <- big.matrix(nrow(G), ncol(G) + ncol(covar.all),
+G3 <- bigmemory::big.matrix(nrow(G), ncol(G) + ncol(covar.all),
                             backingfile = "G-PC",                 
@@
 ```{r}
-G3 <- attach.big.matrix("backingfiles/G-PC.desc")
+G3 <- bigmemory::attach.big.matrix("backingfiles/G-PC.desc")
 ```
@@
     logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test, 
-               method = "logit-simple-enet-ridge", alpha = 0.001),
+               method = "logit-simple-enet-ridge", alphas = 0.001),
     logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test, 
-               method = "logit-simple",            alpha = 0.5),
+               method = "logit-simple",            alphas = 0.5),
     logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test, 
-               method = "logit-simple-lasso",      alpha = 1)
+               method = "logit-simple-lasso",      alphas = 1)
   )
@@
 ```
+
+```{r, include=FALSE, error=TRUE}
+library(gmailr)
+send_message(
+  mime(To = "florian.prive.21@gmail.com",
+       From = "florian.prive.21@gmail.com",
+       Subject = sprintf("Knitting of %s has finished",
+                         knitr::current_input()))
+)
+```

### 2018-06-07 5-results-celiac.Rmd
update

@@
 myggplot <- function(..., coeff = 1) {
-  bigstatsr:::MY_THEME(ggplot(...), coeff = coeff)
+  ggplot(...) + bigstatsr::theme_bigstatsr(size.rel = coeff)
 } 
@@
 y <- celiac$fam$affection - 1
+
+G2 <- big_attach("backingfiles/celiacQC_tripled1.rds")
 ```
@@
                         covar.train = obj.svd$u[ind.train, ],
-                        ncores = NCORES)
+                        ncores = NCORES, alpha = 0.5)
 )
@@
 ```{r, cache=TRUE}
-G2 <- big_attach("backingfiles/celiacQC_tripled1.rds")
 system.time(
@@
                          covar.train = obj.svd$u[ind.train, ],
-                         ncores = NCORES)
+                         ncores = NCORES, alphas = 0.5)
 )
@@
                                covar.train = covar.all[ind.train, , drop = FALSE],
-                               ncores = nb_cores())
+                               ncores = NCORES, alphas = 0.5)
 
@@
 ```
+
 ```{r}
@@
   group_by(method) %>%
-  summarize_at(-1, function(x) {
+  summarize_at(c("AUC", "pAUC", "nb.preds", "timing"), function(x) {
     glue::glue("{signif(mean(x), 3)} ({signif(boot(x), 3)})")

### 2018-06-07 big-data.R

@@ -1,0 +1,22 @@
+library(bigsnpr)
+
+celiac <- snp_attach("backingfiles/celiacQC.rds")
+G <- celiac$genotypes
+
+# G2 <- big_copy(G, ind.row = rep(rows_along(G), 10), 
+#                backingfile = "backingfiles/celiacQC_rep10", save = TRUE)
+G2 <- big_attach("backingfiles/celiacQC_rep10.rds")
+
+n <- nrow(G)
+ind_train <- sort(sample(n, 12e3))
+ind_test <- setdiff(ind_unique, ind_train)
+
+ind_train10 <- as.vector(outer(ind_train, 0:9 * n, '+'))
+ind_test10  <- as.vector(outer(ind_test,  0:9 * n, '+'))
+y_10 <- rep(celiac$fam$affection - 1, 10)
+
+system.time(
+  cmsa.logit <- big_spLogReg(X = G2, y01.train = y_10[ind_train10], 
+                             ind.train = ind_train10, 
+                             ncores = 6, alphas = 0.5)
+)

### 2018-06-12 paper/draft.tex
minor text changes

@@
 
-In this paper, we present an efficient implementation of such penalized joint estimation, allowing for practical application of penalized logistic regression on modern datasets including hundreds of thousands of individuals.
+In this paper, we present an efficient method to jointly estimate SNP effects, allowing for practical application of penalized logistic regression on modern datasets including hundreds of thousands of individuals.
 Our implementation of penalized logistic regression also includes nearly-optimal choices for hyper-parameters.
-The choice of hyper-parameters for a method is very important as it can dramatically impact the predictive performance of the method (e.g.
-the p-value threshold in C+T).
+The choice of hyper-parameters for a method is very important as it can dramatically impact the predictive performance of the method.
+As an example, AUC values range from less than 60\% to 90\% in a model with 30 causal SNPs, depending on the p-value threshold in C+T.
 
 We compare the performance of penalized logistic regression to the C+T method and to a derivation of random forests. 
-The penalized logistic regression consistently achieves higher predictive performance than the two other methods while being faster.
+Penalized logistic regression consistently achieves higher predictive performance than the two other methods while being faster.
 We also found that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; AUC values increase from 83\% with C+T to 92.5\% with penalized logistic regression.
@@
 A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or even just PRS -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
-This technique uses GWAS summary statistics only, allowing for trivial and fast implementation.
-However, the approach also has limitations.
-In particular, previous studies showed that predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite{ware2017heterogeneity}.
+This technique uses GWAS summary statistics only, allowing for fast implementation.
+However, the ``C+T'' approach also has several limitations.
+Previous studies showed that predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
 Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}.
@@
 In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression against the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
-Note that in this comparison we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
+In this comparison, we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
 For the C+T model, we compare different thresholds of inclusion of SNPs.
@@
 First, we present a procedure that we call Cross-Model Selection and Averaging (CMSA) for automatically choosing the regularization parameter of the penalized logistic regression, which directly affects the number of SNPs included in the model.
-Secondly, we also show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.
+Secondly, we show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.
 
-For the simulations of this paper, we use real genotype data and simulate new phenotypes \cite{zhou2013polygenic}. 
+For the simulations of this paper, we use real genotype data and simulate new phenotypes \cite[]{zhou2013polygenic}. 
 In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the heritability.
@@
 Overall, we found that the penalized logistic regression consistently achieves higher predictive performance while being faster than the C+T and T-Trees methods.
-This demonstrates the feasibility and relevance of this approach in PRS modeling in future studies.
+This demonstrates the feasibility and relevance of this approach for PRS computation.
 
@@
 For simulations presented later, we first restrict the dataset to controls in order to remove the genetic structure induced by the celiac disease status.
-Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on the robust Mahalanobis distances computed using the first 10 principal components of remaining individuals.
+Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.
+In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on the robust Mahalanobis distances computed using the first 10 principal components of the remaining individuals.
 This 3-step filtering process results in a sample of 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing'').
-We also use this dataset for a real data application, in this case keeping all individuals.
+We also use this dataset for real data application, in this case keeping all 15,155 individuals.
 
@@
 We vary parameters of the simulations in order to match a range of genetic architecture, from low to high polygenicity.
-For practical reason, this was achieved by varying the number of chromosomes used (either all 22 autosomal chromosomes or chromosome 6 only), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), and the heritability (50\% or 80\%).
-We also considered deviation form the standard normal additive model, drawing effects of causal SNPs from either Normal or Laplace distribution, and computing liability scores either from a ``simple'' model with linear effects only or a ``fancy'' model which combines linear, dominant and interaction-type effects.
-For the ``simple'' model, we compute $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i$$ where $w_j$ are weights generated with a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs) and $\epsilon \sim N(0, 1 - h^2)$.
+This was achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere or 30 in the HLA region only), and the heritability (50\% or 80\%). 
+Then, to further increase the proportion of causal variants, instead of using all 22 automosal chromosomes, we restricted the dataset to chromosome 6 only.
+We also considered deviation form the standard normal additive model, drawing effects of causal SNPs from either a Normal or from a Laplace distribution, and computing liability scores either from a ``simple'' model with linear effects only or a ``fancy'' model which combines linear, dominant and interaction-type effects.
+For the ``simple'' model, we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated with a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs) and $\epsilon \sim N(0, 1 - h^2)$.
 For the ``fancy'' model, we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$
 and $S_\text{causal}^{(3)}$; $S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$.
-We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction} ~~+~~ \epsilon_i$$
+We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction} ~~+~~ \epsilon_i~,$$
 where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$.
@@
 Finally, scenario \textnumero3 reuses the whole dataset but this time varying the size of the training set in order to assess how the sample size affects predictive performance of methods.
-A total of 100 simulations per combination of parameters are run using 300 causal SNPs anywhere on the genome.
+A total of 100 simulations per combination of parameters are run using 300 causal SNPs randomly chosen anywhere on the genome.
 
@@
 This measure is similar to the AUC, but focuses on high specificities, which is the most useful part of the ROC curve in clinical settings.
-When reporting results of AUC, we use estimates of maximum achievable AUC values for the simulations: 84\% and 94\% for heritabilities of respectively 50\% and 80\% based on three different estimations (see Supplementary Materials).
+When reporting AUC results of simulations, we use estimates of maximum achievable AUC of 84\% and 94\% for heritabilities of respectively 50\% and 80\%, based on three different yet consistent estimations (see Supplementary Materials).
 %Note that we also report the timing of the main computations and the number of SNPs used in the predictions.
@@
 The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Study (GWAS).
-In the GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$.
+In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$.
 The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another, which form a set of SNPs denoted $S_\text{clumping}$.
 Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $p_T$ to be determined.
-Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS.
+Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS.
 In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.
@@
 
-Finally, for the penalized logistic regression, we solve: $$\arg\!\min_{\beta_0,~\beta}(\lambda, \alpha)\left\{  \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization}  \right\}$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict, $\lambda$ and $alpha$ are two regularization parameters that need to be chosen.
+Finally, for the penalized logistic regression, we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function: $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict, $\lambda$ and $\alpha$ are two regularization parameters that need to be chosen.
 Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
@@
 the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
-In this study, we use the elastic-net regularization with $\alpha = 0.5$.%, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.
+In this study, we use elastic-net regularization with $\alpha = 0.5$.%, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.
 
@@
 To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
-Because of the L1-regularization, the resulting vectors of coefficients are typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts.
+Because of L1-regularization, the resulting vectors of coefficients are typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts.
 We refer to this method as ``logit-simple'' in the results section.
@@
 In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects.
-This results in a dataset with 3 times as many variables as the initial one, on which we can apply the penalized logistic regression with the CMSA procedure (Figure \ref{fig:CMSA}).
+This results in a dataset with 3 times as many variables as the initial one, on which we can apply penalized logistic regression with the CMSA procedure (Figure \ref{fig:CMSA}).
 We call this method ``logit-triple'' in the results.
@@
 
-We use Monte Carlo cross-validation to compute AUC, partial AUC, number of predictors used and execution time for the original Celiac dataset with real phenotypes: we randomly split 100 times the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals.
+We use Monte Carlo cross-validation to compute AUC, partial AUC, the number of predictors and execution time for the original Celiac dataset with real phenotypes: we randomly split 100 times the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals.
 
@@
 
-We compared the penalized logistic regression (``logit-simple'') with the C+T method (``PRS'') using whole-genome simulations of scenario \textnumero1 (Table \ref{tab:simus}).
+We compared penalized logistic regression (``logit-simple'') with the C+T method (``PRS'') using whole-genome simulations of scenario \textnumero1 (Table \ref{tab:simus}).
 
 When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}).
-Moreover, the penalized logistic regression consistently provides better predictive performance than the C+T method across all scenarios we considered, except in the case of high polygenicity and small sample size, in which case all methods hardly exceed 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
+Moreover, penalized logistic regression consistently provides better predictive performance than the C+T method across all scenarios we considered, except in the case of high polygenicity and small sample size, in which case all methods hardly exceed 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
 
 Method ``logit-simple'' provides particularly higher predictive performance than ``PRS-max'' where there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region.
-In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' and 83\% with the ``PRS-max'', for a maximum achievable AUC of 94\% (Figure \ref{fig:main-AUC-logit}).
+In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' approach and 83\% with ``PRS-max'', and the maximum achievable AUC is of 94\% (Figure \ref{fig:main-AUC-logit}).
 
@@
 
-Method ``logit-simple'', that automatically chooses hyper-parameter $\lambda$, provides similar predictive performance than the best predictive performance of the biglasso implementation \cite[]{zeng2017efficient}, only slightly better for biglasso but likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).
+Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of the biglasso implementation \cite[]{zeng2017efficient}, only slightly better for biglasso but likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).
 
@@
 As compared to ``logit-simple'', T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}).
-Even when simulating a fancy model in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower with T-Trees as compared with ``logit-simple''.
+Even when simulating a fancy model in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower with T-Trees as compared with ``logit-simple'' (Figure \ref{fig:supp-ttrees}).
 
 Secondly, we compare the two penalized logistic regressions in scenario \textnumero1, ``logit-simple'' and ``logit-triple'' that uses additional features (variables) coding for recessive and dominant effects.
-Predictive performance of ``logit-triple'' are nearly as good as ``logit-simple'' when there are only linear effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
+Predictive performance of ``logit-triple'' are nearly as good as ``logit-simple'' when there are only linear effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-AUC-triple} and \ref{fig:supp-triple}).
 For the ``fancy model'', ``logit-triple'' provides AUC values at least 3.5\% higher than ``logit-simple'', except when there are 3000 causal SNPs.
@@
 Thirdly, for the ``fancy'' simulation model, we report similar results as in the previous subsection.
-As before, ``logit-simple'' achieves greater AUCs than the ``PRS-max'', except in the case of 3000 causal SNP anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}).
+As before, ``logit-simple'' achieves greater AUCs than the ``PRS-max'', except in the case of 3000 causal SNPs randomly chosen anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}).
 Moreover, the choice of the threshold inclusion of SNPs also greatly impacts the performance of the C+T method (Figure \ref{fig:supp-AUC-PRS-fancy}).
@@
 First, when reproducing simulations of scenario \textnumero1 using only chromosome 6 (scenario \textnumero2), the predictive performance of all methods globally increase (Figure \ref{fig:supp-AUC-chr6}).
-There is particularly a large increase when simulating 3000 causal SNPs anywhere in the dataset: AUC from the ``logit-simple'' increases from less than 60\% to nearly 80\%. 
+There is particularly a large increase when simulating 3000 causal SNPs: AUC from the ``logit-simple'' increases from less than 60\% to nearly 80\%. 
 On the contrary, when simulating only 30 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' does not increase. 
@@
 
-Joint logistic regressions provide higher AUC for the Celiac data: 88.8\% with ``logit-simple'' and 89.2\% with ``logit-triple'' as compared to 82.4\% with the C+T method.
+Joint logistic regressions provide higher AUC values for the Celiac data: 88.8\% with ``logit-simple'' and 89.2\% with ``logit-triple'' as compared to 82.4\% with the C+T method.
 The relative increase in partial AUC, for specificities larger than 90\%, is even larger (45\% and 50\%) with partial AUC values of 0.0414, 0.0429 and 0.0286 obtained with ``logit-simple'', ``logit-triple'' and the C+T method, respectively.
@@
 In this comparative study, we present a computationally efficient implementation of penalized logistic regression. 
-This model can be used to build polygenic risk scores based on very large SNP datasets such as the UK biobank \cite[]{bycroft2017genome}.
-In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to dramatically improve predictive performance as compared to the standard C+T approach in which these effects are learned independently.
+%This model can be used to build polygenic risk scores based on very large SNP datasets such as the UK biobank \cite[]{bycroft2017genome}. 
+In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently.
 Penalized logistic regression consistently outperform the C+T method, except in the case where there are only small effects and sample size is limited, in which case all methods perform poorly.
-Note that in this particular case, penalized logistic regression could outperform the C+T method if choosing a different value for hyper-parameter $\alpha$.
+Note that in this particular case, penalized logistic regression could outperform the C+T method if choosing $\alpha = 10^{-3}$, which gives more importance to ridge regularization and accept more predictors in the model.
 
@@
 
-Indeed, the choice of hyper-parameters' values is very important as it can greatly impact the performance of the methods.
+The choice of hyper-parameters' values is very important as it can greatly impact method performance.
 In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps, respectively.
@@
 Moreover, in this paper, we reported the maximum AUC of 102 $p_T$ thresholds, a threshold that should normally be learned on the training set only.
-The choice of this threshold is very important as it can greatly impact the predictive performance of the C+T method \cite[]{ware2017heterogeneity}, which we confirm in this study.
+The choice of this threshold is very important as it can greatly impact the predictive performance of the C+T method, which we confirm in this study \cite[]{ware2017heterogeneity}.
 
@@
 Not only this procedure provides near-optimal results, but it also accelerates the training of the model thanks to the development of an early stopping criterion.
-Usually, cross-validation is used to assess the best value for this hyper-parameter and then the model is train again with this particular hyper-parameter value \cite[]{Hastie2008,wei2013large}.
-Yet, as the value of $\lambda$ depends on the sample size and doing cross-validation and retraining is computationally demanding, we think using CMSA is more appropriate in this setting.
+Usually, cross-validation is used to choose hyper-parameter value and then the model is trained again with this particular hyper-parameter value \cite[]{Hastie2008,wei2013large}.
+Yet, performing cross-validation and retraining is computationally demanding, CMSA offers a less burdensome alternative.
 Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we show that an hybrid approach ($\alpha = 0.5$) gives similar results as a model using only L1 regularization ($\alpha = 1$) in all simulated models.
@@
 In this paper, we also explored how to capture non-linear effects.
-For this, we introduced a simple feature engineering technique that enables the logistic regression to detect and learn not only additive effects, but also dominant and recessive effects.
-This technique improves the predictive performance of the method when there are some non-linear effects in the simulations, while providing the same predictive performance when there are only linear effects.
+For this, we introduced a simple feature engineering technique that enables logistic regression to detect and learn not only additive effects, but also dominant and recessive effects.
+This technique improves the predictive performance of logistic regression when there are some non-linear effects in the simulations, while providing the same predictive performance when there are only linear effects.
 Moreover, it also improves predictive performance for the celiac disease.
@@
 In another study, we will assess how can we use models and effects learned in one population to improve learning and prediction in another population.
-We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}.
+%We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}.
 Finally, we also wish to assess how can we combine the information provided by genetic data with clinical and environmental data, possibly in a non-linear way.
@@
 \caption{Comparison of models when varying sample size in scenario \textnumero3.
-Mean of AUC over 100 simulations for ..the three reported PRSs...
-and the ``logit-simple'' as a function of the training size.
+Mean of AUC over 100 simulations for the three reported PRSs and the ``logit-simple'' as a function of the training size.
 These results are for the ``simple'' model with 300 causal SNPs sampled anywhere on the genome.
@@
 \centerline{\includegraphics[width=\textwidth]{celiac-roc}}
-\caption{ROC Curves for the ``C+T'', ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.
-Plotted using R package plotROC \cite[]{sachs2017plotroc}.\label{fig:celiac-roc}}
+\caption{ROC Curves for the ``C+T'', ``logit-simple'' and ``logit-triple'' methods for Celiac disease dataset.
+Models were trained using 12,000 individuals.
+These are results projecting these models on the remaining 3155 individuals.
+The figure is plotted using R package plotROC \cite[]{sachs2017plotroc}.\label{fig:celiac-roc}}
 \end{figure}
@@
 We report AUCs of 84.1\% and 94.1\% for respectively an heritability of 50\% and 80\%.
-[HOW TO ADD ERROR OF ESTIMATION?
-I HAVE SUMMARY() IN R]
 Thirdly, we reproduce the exact same procedure of simulations and, for each combination of parameters (Table \ref{tab:simus}), we estimate the AUC of the ``oracle'', i.e.\ the true simulated genetic part of the liabilities through 100 replicates.
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6}}
-\caption{Comparison of models when using only chromosome 6 in scenario \textnumero2. Boxplots of AUC over 100 simulations for ..the three reported PRSs... and the ``logit-simple''.
+\caption{Comparison of models when using only chromosome 6 in scenario \textnumero2. Boxplots of AUC over 100 simulations for the three reported PRSs and the ``logit-simple''.
 Vertical panels represents different number of causal SNPs and their location (``all'': anywhere on the genome; ``HLA'': only in the HLA region).
@@
 The blue dotted line represents the maximum achievable AUC.
-These results corresponds to the ``simple'' model and an heritability of 50\%.}
+These results correspond to the ``simple'' model and an heritability of 50\%.}
 \label{fig:supp-AUC-logit}
@@
 The blue dotted line represents the maximum achievable AUC.
-These results corresponds to the ``simple'' model and an heritability of 50\%.}
+These results correspond to the ``simple'' model and an heritability of 50\%.}
 \label{fig:supp-AUC-PRS}
@@
 The blue dotted line represents the maximum achievable AUC.
-These results corresponds to the ``fancy'' model.}
+These results correspond to the ``fancy'' model.}
 \label{fig:supp-AUC-logit-fancy}
@@
 The blue dotted line represents the maximum achievable AUC.
-These results corresponds to the ``fancy'' model.}
+These results correspond to the ``fancy'' model.}
 \label{fig:supp-AUC-PRS-fancy}

### 2018-07-12 1-preprocessing.Rmd
save

@@
 ```{r}
-file.copy(paste0("../paper1-packages/backingfiles/celiacQC", c(".bk", ".rds")),
-          paste0("backingfiles/celiacQC", c(".bk", ".rds")))
-
 library(bigsnpr)
@@
 # Get population
-pop.files <- list.files(path = "../thesis-celiac/Dubois2010_data/", 
+pop.files <- list.files(path = "backingfiles", 
                         pattern = "cluster_*", full.names = TRUE)
@@
                        ncores = nb_cores())
+# Save this PCA
+saveRDS(obj.svd, "backingfiles/PCA.rds")
 
@@
                         ncores = nb_cores())
+# Save this new PCA
+saveRDS(obj.svd2, "backingfiles/PCA2.rds")
+
 plot(obj.svd2)
 plot(obj.svd2, type = "scores") 
-# Save this new PCA
-saveRDS(obj.svd2, "backingfiles/PCA2.rds")
 ```

### 2018-07-12 3-simus.Rmd
save

@@
 library(bigsnpr)
-# https://stackoverflow.com/q/9486952/6103040
-options(bigstatsr.cluster.type = "PSOCK")
 ```
@@
 
-    # Clumping on the test set
+  })[3]
+
+  lapply(c(0.05, 0.2, 0.8), function(thr.r2) {
+
+    # Clumping
     ind.keep <- snp_clumping(G, infos.chr = infos.chr,
                              ind.row = ind.test,
-                             thr.r2 = 0.2, 
+                             thr.r2 = thr.r2, 
                              S = abs(gwas.train.gc$score),
@@
     lpS <- -predict(gwas.train.gc)
-    nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
     prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
@@
                    thr.list = thrs)
-  })[3]
 
@@
         method = methods[i],
-      eval = list(cbind(prs[, k], pheno.all[ind.test])),
-      timing = timing,
-      nb.preds = nb.pred[k]
+        pred = list(prs[, k]),
+        thr.r2 = thr.r2,
+        set = list(intersect(ind.keep, which(lpS > thrs[k]))),
+        timing = timing
       )
+    }) %>%
+      bind_rows()
   }) %>%
@@
 ```{r}
-logit.CMSA <- function(G, pheno.all, covar.all, ind.train, ind.test,
-                       method = "logit-simple", alphas = 0.5) {
+logit.CMSA <- function(G, pheno.all, covar.all, ind.train, ind.test, method,
+                       alphas = c(1, 0.5, 0.05, 0.001)) {
 
@@
 
-    preds <- rowMeans(
-      predict(cmsa.logit, X = G, ind.row = ind.test, 
+    preds <- predict(cmsa.logit, X = G, ind.row = ind.test, 
                      covar.row = covar.all[ind.test, , drop = FALSE])
-    )
   })[3]
@@
     method   = method, 
-    eval     = list(cbind(preds, pheno.all[ind.test])),
+    pred     = list(preds),
     timing   = timing,
-    nb.preds = sum(rowSums(sapply(cmsa.logit, function(x) x$beta.X) != 0) > 0)
+    alpha = attr(cmsa.logit, "alpha"),
+    set = list(which(rowSums(sapply(cmsa.logit, function(x) x$beta.X) != 0) > 0))
   )
 }
+
 ```
@@
 ```{r}
-TTree <- "../../TTree-source/TTree"
+bigstatsr:::assert_exist(TTree <- "../TTree-source/TTree")
 
@@
     method   = "T-Trees",
-    eval     = list(preds[match(ind.test - 1, preds[[1]]), 2:3]),
+    pred     = list(preds[match(ind.test - 1, preds[[1]]), 2]),
     timing   = timing,
-    nb.preds = sum(read.table(file.vim)$V2 != 0)
-  )
-}
-```
-
-### Logistic regression with biglasso
-
-```{r}
-logit.biglasso <- function(G3, pheno.all, covar.all, ind.train, ind.test) {
-
-  timing <- system.time({
-
-    biglasso <- biglasso(G3, pheno.all, ind.train, ncores = NCORES,
-                         penalty = "enet", family = "binomial", alpha = 0.5)
-
-    preds <- 1 / (1 + exp(-predict(biglasso, G3, ind.test)))
-  })[3]
-
-  aucs <- apply(preds, 2, bigstatsr::AUC, target = pheno.all[ind.test])
-  ind.max <- which.max(aucs)
-
-  tibble(
-    method   = "biglasso", 
-    eval     = list(cbind(preds[, ind.max], pheno.all[ind.test])),
-    timing   = timing,
-    nb.preds = sum(biglasso$beta[, ind.max] != 0)
+    set = list(which(read.table(file.vim)$V2 != 0))
   )
@@
   y.simu <- y.simu + rnorm(nrow(G), sd = sqrt(1 - h2))
-  pheno <- as.numeric(y.simu > qnorm(1 - K))
+  list(pheno = as.integer(y.simu > qnorm(1 - K)), set = set, effects = effects)
 }
@@
   # Simulate phenotypes
-  pheno.all <- get_pheno(
+  simu_pheno <- get_pheno(
     G,    
@@
     effects.dist = params[["par.dist"]], 
-    model = params[["par.model"]], 
-    K = 0.3                 
+    model = params[["par.model"]]
   )
+  pheno.all <- simu_pheno$pheno
+  params[["set"]] <- list(simu_pheno$set)
 
@@
   ind.test <- setdiff(1:n, ind.train)
+  params[["pheno"]] <- list(pheno.all[ind.test])
 
@@
   res <- bind_rows(
-    PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test),
+    # PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test),
     logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple"),
@@
   # Simulate phenotypes
-  pheno.all <- get_pheno(
-    G6,    
+  simu_pheno <- get_pheno(
+    G,    
     h2 = params[["par.h2"]], 
     M = as.integer(par.causal[1]), 
-    ind.possible = `if`(par.causal[2] == "all", cols_along(G6), ind.HLA6),
+    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
     effects.dist = params[["par.dist"]], 
-    model = params[["par.model"]], 
-    K = 0.3                 
+    model = params[["par.model"]]
   )
+  pheno.all <- simu_pheno$pheno
+  params[["set"]] <- list(simu_pheno$set)
 
@@
 
-### Scenario n°1 (comparison with biglasso)
-
-```{r}
-library(biglasso)
-library(Matrix)
-```
-
-```{r, eval=FALSE}
-G3 <- bigmemory::big.matrix(nrow(G), ncol(G) + ncol(covar.all),
-                            backingfile = "G-PC",                 
-                            backingpath = "backingfiles")
-big_apply(G, function(X, ind) {
-  G3[, ind] <- X[, ind]
-  NULL
-}, a.combine = 'c')
-G3[, ncol(G) + cols_along(covar.all)] <- covar.all
-```
-
-```{r}
-G3 <- bigmemory::attach.big.matrix("backingfiles/G-PC.desc")
-```
-
-```{r}
-params.grid7 <- expand.grid(
-  n.train = 6000,
-  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
-  par.dist   = c("gaussian", "laplace"), 
-  par.h2     = c(0.8), 
-  par.model  = c("simple"),
-  num.simu   = 1:100,
-  stringsAsFactors = FALSE
-) 
-```
-
-```{r, eval=TRUE}
-if (!dir.exists("results7")) dir.create("results7")
-
-for (i in rows_along(params.grid7)) {
-
-  res.file <- paste0("results7/simu_", i, ".rds")
-  if (file.exists(res.file)) next
-
-  params <- params.grid7[i, ]
-  par.causal <- params[["par.causal"]][[1]]
-
-  # Simulate phenotypes
-  pheno.all <- get_pheno(
-    G,    
-    h2 = params[["par.h2"]], 
-    M = as.integer(par.causal[1]), 
-    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
-    effects.dist = params[["par.dist"]], 
-    model = params[["par.model"]], 
-    K = 0.3                 
-  )
-
-  # Split in training/test sets
-  ind.train <- sort(sample(n, size = params[["n.train"]]))
-  ind.test <- setdiff(1:n, ind.train)
-
-  # Get results from all methods
-  res <- bind_rows(
-    logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test),
-    logit.biglasso(G3, pheno.all, covar.all, ind.train, ind.test)
-  )
-  params[["res"]] <- list(res)
-  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
-}
-```
-
-### Scenario n°1 (comparison of different $\alpha$)
-
-```{r, eval=TRUE}
-if (!dir.exists("results8")) dir.create("results8")
-
-for (i in rows_along(params.grid7)) {
-
-  res.file <- paste0("results8/simu_", i, ".rds")
-  if (file.exists(res.file)) next
-
-  params <- params.grid7[i, ]
-  par.causal <- params[["par.causal"]][[1]]
-
-  # Simulate phenotypes
-  pheno.all <- get_pheno(
-    G,    
-    h2 = params[["par.h2"]], 
-    M = as.integer(par.causal[1]), 
-    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
-    effects.dist = params[["par.dist"]], 
-    model = params[["par.model"]], 
-    K = 0.3                 
-  )
-
-  # Split in training/test sets
-  ind.train <- sort(sample(n, size = params[["n.train"]]))
-  ind.test <- setdiff(1:n, ind.train)
-
-  # Get results from all methods
-  res <- bind_rows(
-    logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test, 
-               method = "logit-simple-enet-ridge", alphas = 0.001),
-    logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test, 
-               method = "logit-simple",            alphas = 0.5),
-    logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test, 
-               method = "logit-simple-lasso",      alphas = 1)
-  )
-  params[["res"]] <- list(res)
-  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
-}
-```
-
 ```{r, include=FALSE, error=TRUE}
@@
        From = "florian.prive.21@gmail.com",
-       Subject = sprintf("Knitting of %s has finished",
-                         knitr::current_input()))
+       Subject = sprintf("Knitting of %s has finished", knitr::current_input()))
 )

### 2018-07-12 save-scripts/new-analysis-of-results.R

@@ -1,0 +1,85 @@
+add_AUC <- function(x, pAUC) {
+  x %>% 
+    mutate(AUC  = map2_dbl(pred, pheno, bigstatsr::AUC),
+           pAUC = map2_dbl(pred, pheno, pAUC))
+}
+
+add_sens_FDP <- function(x, corr) {
+  
+  max2 <- function(x) {
+    x <- na.omit(x)
+    `if`(length(x) == 0, 0, max(x))
+  }
+  
+  alt <- map2(x$true_set, x$set, ~ {
+    
+    corr_set <- abs(corr[.x, .y, drop = FALSE])
+    
+    AltSens <- mean(apply(corr_set, 1, max2))
+    AltFDP <- 1 - mean(apply(corr_set, 2, max2))
+    c(AltSens, AltFDP)
+  }) %>%
+    transpose()
+  
+  x %>%
+    mutate(AltSens = unlist(alt[[1]]),
+           AltFDP  = unlist(alt[[2]]))
+}
+
+results4 <- list.files("results4", full.names = TRUE) %>%
+  big_parallelize(., p.FUN = function(files, ind, add_AUC, pAUC, add_sens_FDP, corr) {
+    
+    library(tidyverse)
+    library(Matrix)
+    
+    files[ind] %>%
+      map_dfr(~readRDS(.x)) %>%
+      as_tibble() %>%
+      mutate(
+        par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
+                            levels = c("30 in HLA", paste(3 * 10^(1:3), "in all")))) %>%
+      add_AUC(pAUC) %>%
+      add_sens_FDP(corr)
+  }, p.combine = bind_rows, ind = seq_along(.), ncores = NCORES, 
+  add_AUC = add_AUC, pAUC = pAUC, add_sens_FDP = add_sens_FDP, corr = corr)
+
+print(results4, width = Inf)
+
+results4 %>%
+  filter(method %in% c("PRS-max", "logit-simple")) %>%
+  group_by_at(c(vars(starts_with("par")), "method", "thr.r2")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
+  print(n = Inf) -> results4.summary
+
+
+c_method_r2 <- function(method, r2) {
+  ifelse(is.na(r2), method, paste0(method, "-", r2))
+}
+
+
+barplot_causal <- function(results) {
+  
+  results %>%
[…]

### 2018-07-13 pkg.paper.PRS/R/RcppExports.R
make a pkg

@@ -1,0 +1,7 @@
+# Generated by using Rcpp::compileAttributes() -> do not edit by hand
+# Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
+
+which_cond <- function(i, j, x, y) {
+    .Call(`_pkg_paper_PRS_which_cond`, i, j, x, y)
+}
+

### 2018-07-13 pkg.paper.PRS/R/pkg.paper.PRS-package.R

@@ -1,0 +1,5 @@
+#' @useDynLib pkg.paper.PRS, .registration = TRUE
+#' @importFrom Rcpp sourceCpp
+#' @keywords internal
+#' @importFrom magrittr %>%
+"_PACKAGE"

### 2018-07-13 pkg.paper.PRS/R/read-format-results.R

@@ -1,0 +1,117 @@
+################################################################################
+
+utils::globalVariables(
+  c("par.causal", "num.simu", "true_set", "set", "pheno", "pred")
+)
+
+################################################################################
+
+#' Alternative sensitivity & False Discovery Proportion
+#'
+#' Compute alternative sensitivity & False Discovery Proportion from
+#' https://doi.org/10.1289/EHP172.
+#'
+#' @param x Data frame containing variables "true_set" and "set".
+#' @param corr Sparse correlation matrix of the genotype matrix.
+#'
+#' @return `x` with 2 new variables "AltSens" and "AltFDP".
+#' @import Matrix
+#'
+add_sens_FDP <- function(x, corr) {
+
+  stopifnot(class(corr) == "dsCMatrix")
+  corr <- methods::as(abs(corr), "dgTMatrix")
+
+  alt <- purrr::map2(x$true_set, x$set, ~ {
+    if (length(.y) == 0) return(c(0, 0))
+    bool.x <- bool.y <- logical(nrow(corr))
+    bool.x[.x] <- TRUE
+    bool.y[.y] <- TRUE
+    ind <- which_cond(corr@i, corr@j, bool.x, bool.y)
+    AltSens <-     sum(tapply(corr@x[ind], corr@i[ind], max)) / length(.x)
+    AltFDP  <- 1 - sum(tapply(corr@x[ind], corr@j[ind], max)) / length(.y)
+    c(AltSens, AltFDP)
+  }) %>%
+    purrr::transpose()
+
+  x %>%
+    dplyr::mutate(AltSens = unlist(alt[[1]]),
+                  AltFDP  = unlist(alt[[2]]))
+}
+
+################################################################################
+
+pAUC <- function(pred, target, p = 0.1) {
+  val.min <- min(target)
+  q <- stats::quantile(pred[target == val.min], probs = 1 - p)
+  ind <- (target != val.min) | (pred > q)
+  tryCatch(bigstatsr::AUC(pred[ind], target[ind]) * p,
+           error = function(e) p^2 / 2)  # all same prediction
+}
+
+add_AUC <- function(x, pAUC) {
+  x %>%
+    dplyr::mutate(AUC  = purrr::map2_dbl(pred, pheno, bigstatsr::AUC),
+                  pAUC = purrr::map2_dbl(pred, pheno, pAUC))
+}
+
+################################################################################
+
+#' Read & format results
+#'
+#' @param files Files to read from.
[…]

### 2018-07-13 pkg.paper.PRS/src/RcppExports.cpp

@@ -1,0 +1,31 @@
+// Generated by using Rcpp::compileAttributes() -> do not edit by hand
+// Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
+
+#include <Rcpp.h>
+
+using namespace Rcpp;
+
+// which_cond
+IntegerVector which_cond(IntegerVector i, IntegerVector j, LogicalVector x, LogicalVector y);
+RcppExport SEXP _pkg_paper_PRS_which_cond(SEXP iSEXP, SEXP jSEXP, SEXP xSEXP, SEXP ySEXP) {
+BEGIN_RCPP
+    Rcpp::RObject rcpp_result_gen;
+    Rcpp::RNGScope rcpp_rngScope_gen;
+    Rcpp::traits::input_parameter< IntegerVector >::type i(iSEXP);
+    Rcpp::traits::input_parameter< IntegerVector >::type j(jSEXP);
+    Rcpp::traits::input_parameter< LogicalVector >::type x(xSEXP);
+    Rcpp::traits::input_parameter< LogicalVector >::type y(ySEXP);
+    rcpp_result_gen = Rcpp::wrap(which_cond(i, j, x, y));
+    return rcpp_result_gen;
+END_RCPP
+}
+
+static const R_CallMethodDef CallEntries[] = {
+    {"_pkg_paper_PRS_which_cond", (DL_FUNC) &_pkg_paper_PRS_which_cond, 4},
+    {NULL, NULL, 0}
+};
+
+RcppExport void R_init_pkg_paper_PRS(DllInfo *dll) {
+    R_registerRoutines(dll, NULL, CallEntries, NULL, NULL);
+    R_useDynamicSymbols(dll, FALSE);
+}

### 2018-07-13 pkg.paper.PRS/src/which-cond.cpp

@@ -1,0 +1,17 @@
+#include <Rcpp.h>
+using namespace Rcpp;
+
+// [[Rcpp::export]]
+IntegerVector which_cond(IntegerVector i, IntegerVector j,
+                         LogicalVector x, LogicalVector y) {
+
+  std::vector<int> res;
+  int K = i.size();
+  for (int k = 0; k < K; k++) {
+    if (x[i[k]] && y[j[k]]) res.push_back(k + 1);
+  }
+
+  IntegerVector res2(res.size());
+  std::copy(res.begin(), res.end(), res2.begin());
+  return res2;
+}

### 2018-07-13 3-simus.Rmd
use a package

@@
 library(tidyverse)
-library(glue)
+# devtools::install_github("privefl/bigsnpr")
 library(bigsnpr)
+# devtools::install_github("privefl/paper2-PRS/pkg.paper.PRS")
+library(pkg.paper.PRS)  ## See code source for functions' code.
 ```
@@
 
-### PRS
-
 ```{r}
@@
   paste(collapse = ", ")
-```
-
-```{r}
-PRS <- function(G, infos.chr, infos.pos, 
-                pheno.all, covar.all, 
-                ind.train, ind.test) {
-
-  timing <- system.time({
-
-    # GWAS
-    gwas.train <- big_univLogReg(
-      G, pheno.all[ind.train], ind.train = ind.train, 
-      covar.train = covar.all[ind.train, , drop = FALSE], 
-      ncores = NCORES
-    )
-    gwas.train.gc <- snp_gc(gwas.train)
-
-  })[3]
-
-  lapply(c(0.05, 0.2, 0.8), function(thr.r2) {
-
-    # Clumping
-    ind.keep <- snp_clumping(G, infos.chr = infos.chr,
-                             ind.row = ind.test,
-                             thr.r2 = thr.r2, 
-                             S = abs(gwas.train.gc$score),
-                             size = 500,
-                             is.size.in.bp = TRUE,
-                             infos.pos = infos.pos,
-                             ncores = NCORES)
-
-    # PRS
-    thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
-    lpS <- -predict(gwas.train.gc)
-    prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
-                   ind.test = ind.test,
-                   ind.keep = ind.keep,
-                   lpS.keep = lpS[ind.keep], 
-                   thr.list = thrs)
-
-    ind.best <- which.max(apply(prs, 2, AUC, target = pheno.all[ind.test]))
-
-    methods <- c("PRS-all", "PRS-stringent", "PRS-max")
-    indices <- c(1:2, ind.best)
-
-    lapply(1:3, function(i) {
-      k <- indices[i]
-      tibble(
-        method = methods[i],
-        pred = list(prs[, k]),
-        thr.r2 = thr.r2,
-        set = list(intersect(ind.keep, which(lpS > thrs[k]))),
-        timing = timing
-      )
-    }) %>%
-      bind_rows()
-  }) %>%
-    bind_rows()
-}
-```
-
-### Logistic regression
-
-```{r}
-logit.CMSA <- function(G, pheno.all, covar.all, ind.train, ind.test, method,
-                       alphas = c(1, 0.5, 0.05, 0.001)) {
-
-  timing <- system.time({
-
-    cmsa.logit <- big_spLogReg(X = G, y01.train = pheno.all[ind.train], 
-                               ind.train = ind.train, 
-                               covar.train = covar.all[ind.train, , drop = FALSE],
-                               alphas = alphas, ncores = NCORES)
-
-    preds <- predict(cmsa.logit, X = G, ind.row = ind.test, 
-                     covar.row = covar.all[ind.test, , drop = FALSE])
-  })[3]
-
-  tibble(
-    method   = method, 
-    pred     = list(preds),
-    timing   = timing,
-    alpha = attr(cmsa.logit, "alpha"),
-    set = list(which(rowSums(sapply(cmsa.logit, function(x) x$beta.X) != 0) > 0))
-  )
-}
-
-```
-
-### T-Trees
-
-```{r}
-bigstatsr:::assert_exist(TTree <- "../TTree-source/TTree")
-
-ttrees <- function(TTree, file.base, pheno.all, ind.train, ind.test,
-                   n.trees = 100) {
-
-  file0.jdb  <- paste0(file.base, ".jdb")
-  file0.bloc <- paste0(file.base, ".bloc")
-
-  # Write jdb file with new pheno
-  tmpfile <- tempfile()
-  write(c(rep("", 5), pheno.all), ncolumns = 1, tmpfile)
-  # https://stackoverflow.com/a/7846550/6103040
-  system(sprintf("awk 'FNR==NR{a[NR]=$1;next}{$2=a[FNR]}1' %s %s > %s",
-                 tmpfile, file0.jdb, file.jdb <- paste0(tmpfile, ".jdb")))
-  # Write indices of learning and validation sets
-  file.learn <- paste0(tmpfile, "_learn.txt")
-  file.val   <- paste0(tmpfile, "_val.txt")
-  cat(ind.train - 1, file = file.learn, sep = "\t")
-  cat(ind.test - 1,  file = file.val,   sep = "\t")
-
-  timing <- system.time(
-    system(glue::glue(
-      "{TTree}",
-      " -j {file.jdb}",
-      " -m 3",
-      " -b {file0.bloc}",
-      " -l {file.learn}",
-      " -v {file.val}", 
-      " -t {n.trees} -k 1000 -c 5 -n 2000",
-      " -x -s"
-    ))
-  )[3]
-
-  file.roc <- sprintf("%s_k1000_m3_t%d_ic5_nmin2000_0000.roc",
-                      file.jdb, n.trees)
-  file.vim <- sub("\\.roc$", ".vim", file.roc)
-  preds <- read.table(file.roc, header = FALSE)
-
-  tibble(
-    method   = "T-Trees",
-    pred     = list(preds[match(ind.test - 1, preds[[1]]), 2]),
-    timing   = timing,
-    set = list(which(read.table(file.vim)$V2 != 0))
-  )
-}
 ```
@@
 
+TTree <- "../TTree-source/TTree"
+
 NCORES <- nb_cores()
-```
-
-### Simulation of phenotypes
-
-```{r}
-get_pheno <- function(
-  G,                                        ## matrix of genotypes
-  h2,                                       ## heritability 
-  M,                                        ## nbs of causal variants
-  ind.possible = cols_along(G),             ## indices of possible causal variants
-  effects.dist = c("gaussian", "laplace"),  ## distribution of effects 
-  model = c("simple", "fancy"),             ## model for simulation
-  K = 0.3                                   ## prevalence
-) {
-
-  effects.dist  <- match.arg(effects.dist)
-  model <- match.arg(model)
-
-  set <- sample(ind.possible, size = M)
-  effects <- `if`(effects.dist == "gaussian", 
-                  rnorm(M, sd = sqrt(h2 / M)),
-                  rmutil::rlaplace(M, s = sqrt(h2 / (2*M))))
-
-  if (model == "simple") {
-    # only linear
-    y.simu <- scale(G[, set]) %*% effects
-  } else {
-    sets <- split(1:M, sample(rep_len(1:3, M)))
-    # linear
-    ind1 <- sets[[1]]
-    y.simu <- scale(G[, set[ind1]]) %*% effects[ind1]
-    # recessive / dominant
-    ind2 <- sets[[2]]
-    y.simu <- y.simu + scale(G[, set[ind2]] > 0.5) %*% effects[ind2]
-    # interactions
-    ind3 <- matrix(sets[[3]], ncol = 2)
-    y.simu <- y.simu + scale(G[, set[ind3[, 1]]] * G[, set[ind3[, 2]]]) %*% 
-      effects[ind3[, 1]] * sqrt(2)
-  }
-
-  y.simu <- y.simu / sd(y.simu) * sqrt(h2)
-  stopifnot(all.equal(drop(var(y.simu)), h2))
-  y.simu <- y.simu + rnorm(nrow(G), sd = sqrt(1 - h2))
-  list(pheno = as.integer(y.simu > qnorm(1 - K)), set = set, effects = effects)
-}
 ```
@@
   pheno.all <- simu_pheno$pheno
-  params[["set"]] <- list(simu_pheno$set)
+  params[["true_set"]] <- list(simu_pheno$set)
 
@@
     logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple"),
-    ttrees(TTree, "backingfiles/ttrees", pheno.all, ind.train, ind.test, n.trees = 100)
+    ttrees("../TTree-source/TTree", "backingfiles/ttrees", 
+           pheno.all, ind.train, ind.test, n.trees = 100)
   )
@@
   simu_pheno <- get_pheno(
-    G,    
+    G6,    
     h2 = params[["par.h2"]], 
     M = as.integer(par.causal[1]), 
-    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
+    ind.possible = `if`(par.causal[2] == "all", cols_along(G6), ind.HLA6),
     effects.dist = params[["par.dist"]], 

### 2018-07-13 4-results-simus.Rmd
save
use a package

@@
 ```{r}
-pAUC <- function(pred, target, p = 0.1) {
-  val.min <- min(target)
-  q <- quantile(pred[target == val.min], probs = 1 - p)
-  ind <- (target != val.min) | (pred > q)
-  tryCatch(bigstatsr::AUC(pred[ind], target[ind]) * p, 
-           error = function(e) p^2 / 2)  # all same prediction
-}
-
-format_results <- function(results) {
-  results %>%
-    as_tibble() %>%
-    mutate(
-      par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
-                          levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
-      AUC = map_dbl(eval, ~ bigstatsr::AUC(.x[, 1], .x[, 2])),
-      pAUC = map_dbl(eval, ~ pAUC(.x[, 1], .x[, 2]))
-    )
-}
-```
-
-```{r}
-boot <- function(x, n = 1e5, f = mean) {
-  sd(replicate(n, f(sample(x, replace = TRUE))))
-}
-```
-
-```{r}
 myggplot <- function(..., coeff = 1) {
@@
 ```{r}
-barplot_causal <- function(results, h2) {
+barplot_causal <- function(results) {
 

### 2018-07-13 pkg.paper.PRS/R/models.R

@@ -1,0 +1,182 @@
+################################################################################
+
+
+#' PRS (Clumping + Thresholding)
+#'
+#' @inheritParams bigsnpr::snp_clumping
+#' @param pheno.all All phenotypes.
+#' @param covar.all All covariables.
+#' @param ind.train Indices corresponding to the training set.
+#' @param ind.test Indices corresponding to the test set.
+#' @param ncores Number of cores to use. Default is `bigstatsr::nb_cores()`.
+#'
+#' @return A data frame with 5 variables:
+#'   - "method": the name of the method,
+#'   - "pred": the predictions for the test set,
+#'   - "thr.r2": the threshold used for clumping,
+#'   - "set": the set of non-zero coefficients,
+#'   - "timing": the execution time of the GWAS step only.
+#' @export
+#' @import bigsnpr
+#'
+PRS <- function(G, infos.chr, infos.pos,
+                pheno.all, covar.all, ind.train, ind.test, ncores = nb_cores()) {
+
+  timing <- system.time({
+
+    # GWAS
+    gwas.train <- bigstatsr::big_univLogReg(
+      G, pheno.all[ind.train], ind.train = ind.train,
+      covar.train = covar.all[ind.train, , drop = FALSE],
+      ncores = ncores
+    )
+    gwas.train.gc <- snp_gc(gwas.train)
+
+  })[3]
+
+  lapply(c(0.05, 0.2, 0.8), function(thr.r2) {
+
+    # Clumping
+    ind.keep <- snp_clumping(G, infos.chr = infos.chr,
+                             ind.row = ind.test,
+                             thr.r2 = thr.r2,
+                             S = abs(gwas.train.gc$score),
+                             size = 500,
+                             is.size.in.bp = TRUE,
+                             infos.pos = infos.pos,
+                             ncores = ncores)
+
+    # PRS
+    thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
+    lpS <- -stats::predict(gwas.train.gc)
+    prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
+                   ind.test = ind.test,
+                   ind.keep = ind.keep,
+                   lpS.keep = lpS[ind.keep],
+                   thr.list = thrs)
+
[…]

### 2018-07-13 pkg.paper.PRS/R/pheno.R

@@ -1,0 +1,64 @@
+################################################################################
+
+#' Simulate phenotypes
+#'
+#' @param G Matrix of genotypes.
+#' @param h2 Heritability.
+#' @param M Number of causal variants. Should be multiple of 6.
+#' @param ind.possible Indices of possible causal variants.
+#' @param effects.dist Distribution of effects.
+#'   Either `"gaussian"` (the default) or `"laplace"`.
+#' @param model Model for simulation.
+#'   Either `"simple"` (the default) with linear effets only,
+#'   or `"fancy"` with linear, domination and interaction-type effetcs.
+#' @param K Prevalence. Default is `0.3`.
+#'
+#' @return A list with 3 elements:
+#'   - "pheno": vector of binary phenotypes,
+#'   - "set": indices of causal variants,
+#'   - "effects": Size of effects corresponding to `set`.
+#' @export
+#' @importFrom stats qnorm rnorm sd var
+#'
+get_pheno <- function(
+  G,                                        ## matrix of genotypes
+  h2,                                       ## heritability
+  M,                                        ## nbs of causal variants
+  ind.possible = bigstatsr::cols_along(G),  ## indices of possible causal variants
+  effects.dist = c("gaussian", "laplace"),  ## distribution of effects
+  model = c("simple", "fancy"),             ## model for simulation
+  K = 0.3                                   ## prevalence
+) {
+
+  effects.dist  <- match.arg(effects.dist)
+  model <- match.arg(model)
+
+  set <- sample(ind.possible, size = M)
+  effects <- `if`(effects.dist == "gaussian",
+                  rnorm(M, sd = sqrt(h2 / M)),
+                  rmutil::rlaplace(M, s = sqrt(h2 / (2*M))))
+
+  if (model == "simple") {
+    # only linear
+    y.simu <- scale(G[, set]) %*% effects
+  } else {
+    sets <- split(1:M, sample(rep_len(1:3, M)))
+    # linear
+    ind1 <- sets[[1]]
+    y.simu <- scale(G[, set[ind1]]) %*% effects[ind1]
+    # recessive / dominant
+    ind2 <- sets[[2]]
+    y.simu <- y.simu + scale(G[, set[ind2]] > 0.5) %*% effects[ind2]
+    # interactions
+    ind3 <- matrix(sets[[3]], ncol = 2)
[…]

### 2018-07-13 pkg.paper.PRS/R/pkg.paper.PRS-package.R

@@ -2,4 +2,7 @@
 #' @importFrom Rcpp sourceCpp
 #' @keywords internal
 #' @importFrom magrittr %>%
+#' @importFrom dplyr bind_rows
+#' @importFrom tibble tibble
+#' @importFrom bigstatsr AUC nb_cores
 "_PACKAGE"

### 2018-07-13 pkg.paper.PRS/R/read-format-results.R

@@ -45,13 +45,13 @@ pAUC <- function(pred, target, p = 0.1) {
   val.min <- min(target)
   q <- stats::quantile(pred[target == val.min], probs = 1 - p)
   ind <- (target != val.min) | (pred > q)
-  tryCatch(bigstatsr::AUC(pred[ind], target[ind]) * p,
+  tryCatch(AUC(pred[ind], target[ind]) * p,
            error = function(e) p^2 / 2)  # all same prediction
 }
 
 add_AUC <- function(x, pAUC) {
   x %>%
-    dplyr::mutate(AUC  = purrr::map2_dbl(pred, pheno, bigstatsr::AUC),
+    dplyr::mutate(AUC  = purrr::map2_dbl(pred, pheno, AUC),
                   pAUC = purrr::map2_dbl(pred, pheno, pAUC))
 }
 
@@ -66,7 +66,7 @@ add_AUC <- function(x, pAUC) {
 #' @return A data frame of formated and aggregated results.
 #' @export
 #'
-read_format_results <- function(files, corr, ncores = bigstatsr::nb_cores()) {
+read_format_results <- function(files, corr, ncores = nb_cores()) {
 
   bigstatsr::big_parallelize(files, p.FUN = function(files, ind, corr) {
 
@@ -80,7 +80,7 @@ read_format_results <- function(files, corr, ncores = bigstatsr::nb_cores()) {
       add_sens_FDP(corr) %>%
       dplyr::select(-c(num.simu, true_set, set, pheno, pred))
 
-  }, p.combine = dplyr::bind_rows, ind = seq_along(files), ncores = ncores, corr = corr)
+  }, p.combine = bind_rows, ind = seq_along(files), ncores = ncores, corr = corr)
 }
 
 ################################################################################

### 2018-07-13 save-scripts/compute-alt-metrics.cpp

@@ -1,0 +1,41 @@
+#include <Rcpp.h>
+using namespace Rcpp;
+
+// [[Rcpp::export]]
+IntegerVector which_cond(IntegerVector i, IntegerVector j,
+                         LogicalVector x, LogicalVector y) {
+  
+  std::vector<int> res;
+  int K = i.size();
+  for (int k = 0; k < K; k++) {
+    if (x[i[k]] && y[j[k]]) res.push_back(k + 1);
+  }
+  
+  IntegerVector res2(res.size());
+  std::copy(res.begin(), res.end(), res2.begin());
+  return res2;
+}
+
+/***R
+add_sens_FDP3 <- function(x, corr, which_cond) {
+  
+  bigstatsr:::assert_class(corr, "dsCMatrix")
+  corr <- as(abs(corr), "dgTMatrix")
+  
+  alt <- map2(x$true_set, x$set, ~ {
+    if (length(.y) == 0) return(c(0, 0))
+    bool.x <- bool.y <- logical(nrow(corr))
+    bool.x[.x] <- TRUE
+    bool.y[.y] <- TRUE
+    ind <- which_cond(corr@i, corr@j, bool.x, bool.y)
+    AltSens <-     sum(tapply(corr@x[ind], corr@i[ind], max)) / length(.x)
+    AltFDP  <- 1 - sum(tapply(corr@x[ind], corr@j[ind], max)) / length(.y)
+    c(AltSens, AltFDP)
+  }) %>%
+    transpose()
+  
+  x %>%
+    mutate(AltSens = unlist(alt[[1]]),
+           AltFDP  = unlist(alt[[2]]))
+}
+*/

### 2018-07-13 save-scripts/new-analysis-of-results.R

@@ -6,17 +6,14 @@ add_AUC <- function(x, pAUC) {
 
 add_sens_FDP <- function(x, corr) {
   
-  max2 <- function(x) {
-    x <- na.omit(x)
-    `if`(length(x) == 0, 0, max(x))
-  }
-  
   alt <- map2(x$true_set, x$set, ~ {
+    
+    if (length(.y) == 0) return(c(0, 0))
     
     corr_set <- abs(corr[.x, .y, drop = FALSE])
     
-    AltSens <- mean(apply(corr_set, 1, max2))
-    AltFDP <- 1 - mean(apply(corr_set, 2, max2))
+    AltSens <- mean(apply(corr_set, 1, max, na.rm = TRUE))
+    AltFDP <- 1 - mean(apply(corr_set, 2, max, na.rm = TRUE))
     c(AltSens, AltFDP)
   }) %>%
     transpose()
@@ -39,11 +36,15 @@ results4 <- list.files("results4", full.names = TRUE) %>%
         par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
                             levels = c("30 in HLA", paste(3 * 10^(1:3), "in all")))) %>%
       add_AUC(pAUC) %>%
-      add_sens_FDP(corr)
+      add_sens_FDP(corr) %>%
+      select(-c(num.simu, true_set, set, pheno, pred))
+    
   }, p.combine = bind_rows, ind = seq_along(.), ncores = NCORES, 
   add_AUC = add_AUC, pAUC = pAUC, add_sens_FDP = add_sens_FDP, corr = corr)
 
 print(results4, width = Inf)
+
+####
 
 results4 %>%
   filter(method %in% c("PRS-max", "logit-simple")) %>%
@@ -73,13 +74,39 @@ barplot_causal <- function(results) {
 }
 
 barplot_causal_all(results4.summary)
-ggsave("new-results.pdf", scale = 1/90, width = 1235, height = 917)
+# ggsave("new-results.pdf", scale = 1/90, width = 1235, height = 917)
 
-results4.summary %>%
-  barplot_causal_one(h2 = 0.5) +
-  facet_grid(par.model ~ par.dist)
 
-results4 %>%
+####
+
+results4.summary2 <- results4 %>%
   filter(method %in% c("PRS-max", "logit-simple")) %>%
   group_by_at(c(vars(starts_with("par")), "method", "thr.r2")) %>%
[…]

### 2018-07-13 save-scripts/new-analysis-of-results2.R

@@ -1,0 +1,36 @@
+library(tidyverse)
+library(bigsnpr)
+library(pkg.paper.PRS)
+
+corr <- readRDS("backingfiles/corr2.rds")
+system.time(
+  results5 <- list.files("results5", full.names = TRUE) %>%
+    read_format_results(corr)
+)  # 62 sec for 647 of ~1Mb
+
+results5 %>%
+  filter(method %in% c("PRS-max", "logit-simple")) %>%
+  group_by_at(c(vars("n.train", starts_with("par")), "method", "thr.r2")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, n = 1e4), N = n()) %>%
+  print(n = Inf) -> results5.summary
+
+results5.summary %>%
+  mutate(method2 = c_method_r2(method, thr.r2)) %>%
+  ggplot(aes(n.train, AUC_mean, color = method2, linetype = method2)) +
+  geom_hline(yintercept = 0.5, linetype = 2) +
+  geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
+             data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
+  geom_line(size = 1.5) +
+  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, 
+                    ymax = AUC_mean + 2 * AUC_boot), 
+                size = 1.5, width = 0) +
+  facet_grid(par.h2 ~ par.dist) +
+  scale_x_continuous(breaks = 1:6 * 1000, minor_breaks = NULL) + 
+  scale_y_continuous(minor_breaks = 0:20 / 20) +
+  # scale_color_manual(values = methods.color) +
+  labs(x = "Size of the training set", y = "Mean of 100 AUCs", 
+       color = "Method", linetype = "Method") +
+  theme(legend.key.width = unit(3, "line")) +
+  theme_bigstatsr()
+
+

### 2018-07-13 some-tests/optimize-sens-FDP.R

@@ -1,0 +1,74 @@
+add_sens_FDP2 <- function(x, corr) {
+  
+  bigstatsr:::assert_class(corr, "dsCMatrix")
+  corr <- as(abs(corr), "dgTMatrix")
+  
+  alt <- map2(x$true_set, x$set, ~ {
+    if (length(.y) == 0) return(c(0, 0))
+    corr_set <- corr[.x, .y, drop = FALSE]
+    AltSens <-     sum(tapply(corr_set@x, corr_set@i, max)) / nrow(corr_set)
+    AltFDP  <- 1 - sum(tapply(corr_set@x, corr_set@j, max)) / ncol(corr_set)
+    c(AltSens, AltFDP)
+  }) %>%
+    transpose()
+  
+  x %>%
+    mutate(AltSens = unlist(alt[[1]]),
+           AltFDP  = unlist(alt[[2]]))
+}
+
+add_sens_FDP3 <- function(x, corr) {
+  
+  bigstatsr:::assert_class(corr, "dsCMatrix")
+  corr <- as(abs(corr), "dgTMatrix")
+  
+  alt <- map2(x$true_set, x$set, ~ {
+    if (length(.y) == 0) return(c(0, 0))
+    bool.x <- bool.y <- logical(nrow(corr))
+    bool.x[.x] <- TRUE
+    bool.y[.y] <- TRUE
+    ind <- which_cond(corr@i, corr@j, bool.x, bool.y)
+    AltSens <-     sum(tapply(corr@x[ind], corr@i[ind], max)) / length(.x)
+    AltFDP  <- 1 - sum(tapply(corr@x[ind], corr@j[ind], max)) / length(.y)
+    c(AltSens, AltFDP)
+  }) %>%
+    transpose()
+  
+  x %>%
+    mutate(AltSens = unlist(alt[[1]]),
+           AltFDP  = unlist(alt[[2]]))
+}
+
+x <- list.files("results5", full.names = TRUE)[1:20] %>%
+  map_dfr(~readRDS(.x)) %>%
+  select(true_set, set) %>%
+  as_tibble()
+  
+corr <- readRDS("backingfiles/corr2.rds")
+
+# system.time(print(add_sens_FDP(x, corr)))
+# system.time(print(add_sens_FDP2(x, corr)))
+Rcpp::sourceCpp('save-scripts/compute-alt-metrics.cpp')
[…]

### 2018-07-13 4-results-simus.Rmd
save

@@
 myggplot <- function(..., coeff = 1) {
-  bigstatsr:::MY_THEME(ggplot(...), coeff = coeff) +
+  ggplot(...) + 
+    bigstatsr::theme_bigstatsr(size.rel = coeff) +
     theme(strip.text.x = element_text(size = rel(2)),

### 2018-07-13 save-scripts/new-analysis-of-results.R

@@ -1,46 +1,12 @@
-add_AUC <- function(x, pAUC) {
-  x %>% 
-    mutate(AUC  = map2_dbl(pred, pheno, bigstatsr::AUC),
-           pAUC = map2_dbl(pred, pheno, pAUC))
-}
+library(tidyverse)
+library(bigsnpr)
+library(pkg.paper.PRS)
 
-add_sens_FDP <- function(x, corr) {
-  
-  alt <- map2(x$true_set, x$set, ~ {
-    
-    if (length(.y) == 0) return(c(0, 0))
-    
-    corr_set <- abs(corr[.x, .y, drop = FALSE])
-    
-    AltSens <- mean(apply(corr_set, 1, max, na.rm = TRUE))
-    AltFDP <- 1 - mean(apply(corr_set, 2, max, na.rm = TRUE))
-    c(AltSens, AltFDP)
-  }) %>%
-    transpose()
-  
-  x %>%
-    mutate(AltSens = unlist(alt[[1]]),
-           AltFDP  = unlist(alt[[2]]))
-}
-
-results4 <- list.files("results4", full.names = TRUE) %>%
-  big_parallelize(., p.FUN = function(files, ind, add_AUC, pAUC, add_sens_FDP, corr) {
-    
-    library(tidyverse)
-    library(Matrix)
-    
-    files[ind] %>%
-      map_dfr(~readRDS(.x)) %>%
-      as_tibble() %>%
-      mutate(
-        par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
-                            levels = c("30 in HLA", paste(3 * 10^(1:3), "in all")))) %>%
-      add_AUC(pAUC) %>%
-      add_sens_FDP(corr) %>%
-      select(-c(num.simu, true_set, set, pheno, pred))
-    
-  }, p.combine = bind_rows, ind = seq_along(.), ncores = NCORES, 
-  add_AUC = add_AUC, pAUC = pAUC, add_sens_FDP = add_sens_FDP, corr = corr)
+corr <- readRDS("backingfiles/corr2.rds")
+system.time(
+  results4 <- list.files("results4", full.names = TRUE) %>%
+    read_format_results(corr)
+)  # 78 sec for 932 of ~150KB
 
 print(results4, width = Inf)
 
@@ -49,14 +15,8 @@ print(results4, width = Inf)
 results4 %>%
   filter(method %in% c("PRS-max", "logit-simple")) %>%
[…]

### 2018-07-13 save-scripts/new-analysis-of-results2.R

@@ -6,7 +6,7 @@ corr <- readRDS("backingfiles/corr2.rds")
 system.time(
   results5 <- list.files("results5", full.names = TRUE) %>%
     read_format_results(corr)
-)  # 62 sec for 647 of ~1Mb
+)  # 62 sec for 647 of ~1MB
 
 results5 %>%
   filter(method %in% c("PRS-max", "logit-simple")) %>%

### 2018-07-27 3-simus.Rmd
save

@@
 ind.HLA <- snp_indLRLDR(CHR, POS, subset(LD.wiki34, ID == "hild12"))
-
-TTree <- "../TTree-source/TTree"
-
-NCORES <- nb_cores()
 ```
@@
     logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple"),
-    logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple"),
+    # logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple"),
     ttrees("../TTree-source/TTree", "backingfiles/ttrees", 

### 2018-07-27 save-scripts/new-analysis-of-results3.R

@@ -1,0 +1,101 @@
+library(tidyverse)
+library(bigsnpr)
+library(pkg.paper.PRS)
+
+corr <- readRDS("backingfiles/corr2.rds")
+system.time(
+  results2 <- list.files("results2", full.names = TRUE) %>%
+    read_format_results(corr)
+)  # 198 sec for 1636 of ~1MB
+
+print(results2, width = Inf)
+
+# saveRDS(results2, "results2_all.rds")
+
+####
+
+results2 %>%
+  filter(method %in% c("PRS-max", "logit-simple")) %>%
+  group_by_at(c(vars(starts_with("par")), "method", "thr.r2")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, n = 1e4), N = n()) %>%
+  print(n = Inf) -> results2.summary
+
+barplot_causal <- function(results) {
+  
+  results %>%
+    mutate(method2 = c_method_r2(method, thr.r2)) %>%
+    myggplot(aes(par.causal, AUC_mean, fill = method2, color = method2)) +
+    geom_hline(yintercept = 0.5, linetype = 2) +
+    geom_col(position = position_dodge(), alpha = 0.3) +
+    geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
+                  position = position_dodge(width = 0.9), color = "black", width = 0.2) + 
+    scale_y_continuous(limits = c(0.5, NA), minor_breaks = 0:20 / 20,
+                       oob = scales::rescale_none) +
+    labs(x = "Causal SNPs (number and location)", y = "Mean of 100 AUCs",
+         fill = "Method", color = "Method")
+}
+
+barplot_causal_all <- function(results) {
+  
+  results %>%
+    barplot_causal() +
+    geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
+               data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
+    facet_grid(par.dist ~ par.h2)
+}
+
+results2.summary %>%
+  filter(par.model == "simple") %>%
+  barplot_causal_all()
+# ggsave("new-results-all.pdf", scale = 1/90, width = 1235, height = 917)
+
+
+####
+
+results2.summary2 <- results2 %>%
+  filter(method %in% c("PRS-max", "logit-simple")) %>%
[…]

### 2018-08-01 pkg.paper.PRS/R/read-format-results.R
add nb.preds

@@ -14,7 +14,7 @@ utils::globalVariables(
 #' @param x Data frame containing variables "true_set" and "set".
 #' @param corr Sparse correlation matrix of the genotype matrix.
 #'
-#' @return `x` with 2 new variables "AltSens" and "AltFDP".
+#' @return `x` with 3 new variables: "AltSens", "AltFDP" and "nb.preds".
 #' @import Matrix
 #'
 add_sens_FDP <- function(x, corr) {
@@ -36,7 +36,8 @@ add_sens_FDP <- function(x, corr) {
 
   x %>%
     dplyr::mutate(AltSens = unlist(alt[[1]]),
-                  AltFDP  = unlist(alt[[2]]))
+                  AltFDP  = unlist(alt[[2]]),
+                  nb.preds = purrr::map_int(set, length))
 }
 
 ################################################################################

### 2018-08-01 pkg.paper.PRS/R/read-format-results.R
nboot as global option

@@ -111,7 +111,7 @@ c_method_r2 <- function(method, r2) {
 #' @return Standard deviation of bootstrap results.
 #' @export
 #'
-boot <- function(x, n = 1e5, f = mean) {
+boot <- function(x, n = getOption("nboot"), f = mean) {
   stats::sd(replicate(n, f(sample(x, replace = TRUE))))
 }
 

### 2018-08-01 pkg.paper.PRS/R/zzz.R

@@ -1,0 +1,17 @@
+################################################################################
+
+.onLoad <- function(libname, pkgname) {
+  options(
+    nboot = 1e5
+  )
+}
+
+################################################################################
+
+.onUnload <- function(libpath) {
+  options(
+    nboot = NULL
+  )
+}
+
+################################################################################

### 2018-08-01 4-results-simus.Rmd
analyze new results

@@
 
+```{r}
+library(tidyverse)
+# devtools::install_github("privefl/bigsnpr")
+library(bigsnpr)
+# devtools::install_github("privefl/paper2-PRS/pkg.paper.PRS")
+library(pkg.paper.PRS)  ## See code source for functions' code.
+options(nboot = 1e4)
+```
+
 ## Useful functions
@@
 ```{r, message=FALSE}
-library(tidyverse)
-
+COLORS <- scales::hue_pal()(6)
 methods.color <- setNames(
-  c(scales::hue_pal()(6), c("black", "blue", "green")), 
+  c(COLORS, COLORS[c(4, 2, 5)]), 
   c("logit-simple", "PRS-max", "PRS-stringent", "T-Trees", "logit-triple", "PRS-all",
-    "biglasso", "logit-simple-enet-ridge", "logit-simple-lasso")
+    "PRS-max-0.05", "PRS-max-0.2", "PRS-max-0.8")
 ) 
@@
 
+```{r, eval=FALSE}
+G <- snp_attach("backingfiles/celiacQC_sub1.rds")$genotypes
+corr <- snp_cor(G, size = 100)
+saveRDS(corr, "backingfiles/corr2.rds")
+```
+
+```{r}
+corr <- readRDS("backingfiles/corr2.rds")
+```
+
 ## Results with T-Trees
@@
 results1 <- list.files("results1", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  format_results()
+  read_format_results(corr)
 ```
@@
 results2 <- list.files("results2", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  format_results()
-
-results2.summary <- results2 %>%
-  group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
-  print(n = Inf) 
+  read_format_results(corr)
 ```
@@
 ```{r}
-results2.summary %>%
+results2 %>%
   filter(grepl("logit", method)) %>%
+  group_by_at(c(vars(starts_with("par")), "method")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
   barplot_causal_one(h2 = 0.5) +
@@
 ```{r}
-results2.summary %>%
+results2 %>%
   filter(par.model == "fancy", method %in% c("logit-simple", "PRS-max")) %>%
+  filter(thr.r2 == 0.2 | is.na(thr.r2)) %>%
+  group_by_at(c(vars(starts_with("par")), "method")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
   barplot_causal_all()
@@
 ```{r}
-results2.summary %>%
+results2 %>%
   filter(par.model == "fancy", grepl("PRS", method)) %>%
+  filter(thr.r2 == 0.2 | is.na(thr.r2)) %>%
+  group_by_at(c(vars(starts_with("par")), "method")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
   barplot_causal_all()
@@
 
+### Different r2 thresholds
+
+```{r}
+results2 %>%
+  filter(par.model == "simple", method %in% c("logit-simple", "PRS-max")) %>%
+  mutate(method = c_method_r2(method, thr.r2)) %>%
+  group_by_at(c(vars(starts_with("par")), "method")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
+  barplot_causal_all()
+
+ggsave("figures/main-AUC-all-r2.pdf", scale = 1/110, width = 1242, height = 951)
+```
+
+```{r}
+results2 %>%
+  filter(par.model == "simple", method %in% c("logit-simple", "PRS-max")) %>%
+  mutate(method = c_method_r2(method, thr.r2)) %>%
+  group_by_at(c(vars(starts_with("par")), "method")) %>%
+  summarise(AltSens_mean = mean(AltSens), AltSens_boot = boot(AltSens), N = n()) %>%
+  myggplot(aes(par.causal, AltSens_mean, fill = method, color = method)) + 
+  geom_col(position = position_dodge(), alpha = 0.3) +
+  geom_errorbar(aes(ymin = AltSens_mean - 2 * AltSens_boot,
+                    ymax = AltSens_mean + 2 * AltSens_boot),
+                position = position_dodge(width = 0.9), color = "black", width = 0.2) + 
+  scale_y_continuous(limits = c(0, 1)) +
+  labs(x = "Causal SNPs (number and location)", y = "Mean of 100 AltSens",
+       fill = "Method", color = "Method") +
+  scale_fill_manual(values = methods.color) +
+  scale_color_manual(values = methods.color) +
+  facet_grid(par.dist ~ par.h2)
+
+ggsave("figures/supp-AltSens-all-r2.pdf", scale = 1/110, width = 1242, height = 951)
+```
+
 ### Correlation between predictive measures
@@
 results3 <- results2 %>%
+  filter(thr.r2 == 0.2 | is.na(thr.r2)) %>%
   filter(par.model == "simple", method != "logit-triple")
@@
 results4 <- list.files("results4", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  format_results()
-
-results4 %>%
-  group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
-  print(n = Inf)
-
-results34 <- rbind(
-  cbind(simu = "all", results3),
-  cbind(simu = "chr6", results4)
-) %>%
-  as_tibble()
+  read_format_results(corr)
 ```
@@
 ```{r, fig.asp = 1, out.width = "90%"}
-results34 %>%
-  myggplot() +
-  geom_hline(yintercept = 0.5, linetype = 2) +
-  geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
-             data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
-  geom_boxplot(aes(method, AUC, fill = simu, color = simu), alpha = 0.3) + 
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  facet_grid(par.model ~ par.causal) +
-  facet_grid(par.h2 + par.dist ~ par.causal) +
-  scale_fill_brewer(type = "qual", palette = 2) +
-  scale_color_brewer(type = "qual", palette = 2) + 
-  labs(x = "Method")
+results4 %>%
+  filter(method %in% c("logit-simple", "PRS-max")) %>%
+  mutate(method = c_method_r2(method, thr.r2)) %>%
+  group_by_at(c(vars(starts_with("par")), "method")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
+  barplot_causal_all()
 
-ggsave("figures/supp-AUC-chr6.pdf", scale = 1/110, width = 1153, height = 908)
+ggsave("figures/supp-AUC-chr6-all-r2.pdf", scale = 1/110, width = 1242, height = 951)
 ```
@@
 results5 <- list.files("results5", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  format_results()
+  read_format_results(corr)
 
-results5.summary <- results5 %>%
-  group_by_at(c(vars(starts_with("par")), "method", "n.train")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
-  print(n = Inf)
-
-results35.summary <- results3.summary %>%
-  filter(par.causal == "300 in all") %>%
-  cbind.data.frame(n.train = 6000) %>%
-  bind_rows(results5.summary)
+results6 <- results2 %>%
+  filter(par.model == "simple", par.causal == "300 in all") %>%
+  bind_rows(results5) %>%
+  filter(method %in% c("logit-simple", "PRS-max")) %>%
+  mutate(method = c_method_r2(method, thr.r2))
 ```
@@
 ```{r}
-myggplot(results35.summary, aes(n.train, AUC_mean, color = method, linetype = method)) +
+results6 %>%
+  group_by_at(c(vars(starts_with("par")), "method", "n.train")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
+  myggplot(aes(n.train, AUC_mean, color = method, linetype = method)) +
   geom_hline(yintercept = 0.5, linetype = 2) +
@@
                 size = 1.5, width = 0) +
-  facet_grid(par.h2 ~ par.dist) +
+  facet_grid(par.dist ~ par.h2) +
   scale_x_continuous(breaks = 1:6 * 1000, minor_breaks = NULL) + 
@@
 ```
-
-### Comparison with biglasso
-
-```{r}
-results7 <- list.files("results7", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  format_results()
-```
-
-```{r, fig.asp=0.65, out.width="95%", fig.width=16}
-compare_logit(results7, "biglasso")
-
-ggsave("figures/supp-biglasso.pdf", scale = 1/100, width = 1600, height = 1000)
-```
-
-### Comparison of different alpha values
-
-```{r}
-results8 <- list.files("results8", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  format_results()
-```
-
-```{r, fig.asp=0.65, out.width="95%", fig.width=16}
-compare_logit(results8, c("logit-simple-enet-ridge", "logit-simple-lasso"))
-
-ggsave("figures/supp-alpha.pdf", scale = 1/100, width = 1600, height = 1000)
-```

### 2018-08-01 save-scripts/new-analysis-of-results.R

@@ -1,81 +1,0 @@
-library(tidyverse)
-library(bigsnpr)
-library(pkg.paper.PRS)
-
-corr <- readRDS("backingfiles/corr2.rds")
-system.time(
-  results4 <- list.files("results4", full.names = TRUE) %>%
-    read_format_results(corr)
-)  # 78 sec for 932 of ~150KB
-
-print(results4, width = Inf)
-
-####
-
-results4 %>%
-  filter(method %in% c("PRS-max", "logit-simple")) %>%
-  group_by_at(c(vars(starts_with("par")), "method", "thr.r2")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, n = 1e4), N = n()) %>%
-  print(n = Inf) -> results4.summary
-
-barplot_causal <- function(results) {
-  
-  results %>%
-    mutate(method2 = c_method_r2(method, thr.r2)) %>%
-    myggplot(aes(par.causal, AUC_mean, fill = method2, color = method2)) +
-    geom_hline(yintercept = 0.5, linetype = 2) +
-    geom_col(position = position_dodge(), alpha = 0.3) +
-    geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
-                  position = position_dodge(width = 0.9), color = "black", width = 0.2) + 
-    scale_y_continuous(limits = c(0.5, NA), minor_breaks = 0:20 / 20,
-                       oob = scales::rescale_none) +
-    labs(x = "Causal SNPs (number and location)", y = "Mean of 100 AUCs",
-         fill = "Method", color = "Method")
-}
-
-barplot_causal_all <- function(results) {
-  
-  results %>%
-    barplot_causal() +
-    geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
-               data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
-    facet_grid(par.dist ~ par.h2)
-}
-
-barplot_causal_all(results4.summary)
-# ggsave("new-results.pdf", scale = 1/90, width = 1235, height = 917)
-
-
-####
-
-results4.summary2 <- results4 %>%
-  filter(method %in% c("PRS-max", "logit-simple")) %>%
-  group_by_at(c(vars(starts_with("par")), "method", "thr.r2")) %>%
[…]

### 2018-08-01 save-scripts/new-analysis-of-results2.R

@@ -1,36 +1,0 @@
-library(tidyverse)
-library(bigsnpr)
-library(pkg.paper.PRS)
-
-corr <- readRDS("backingfiles/corr2.rds")
-system.time(
-  results5 <- list.files("results5", full.names = TRUE) %>%
-    read_format_results(corr)
-)  # 62 sec for 647 of ~1MB
-
-results5 %>%
-  filter(method %in% c("PRS-max", "logit-simple")) %>%
-  group_by_at(c(vars("n.train", starts_with("par")), "method", "thr.r2")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, n = 1e4), N = n()) %>%
-  print(n = Inf) -> results5.summary
-
-results5.summary %>%
-  mutate(method2 = c_method_r2(method, thr.r2)) %>%
-  ggplot(aes(n.train, AUC_mean, color = method2, linetype = method2)) +
-  geom_hline(yintercept = 0.5, linetype = 2) +
-  geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
-             data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
-  geom_line(size = 1.5) +
-  geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, 
-                    ymax = AUC_mean + 2 * AUC_boot), 
-                size = 1.5, width = 0) +
-  facet_grid(par.h2 ~ par.dist) +
-  scale_x_continuous(breaks = 1:6 * 1000, minor_breaks = NULL) + 
-  scale_y_continuous(minor_breaks = 0:20 / 20) +
-  # scale_color_manual(values = methods.color) +
-  labs(x = "Size of the training set", y = "Mean of 100 AUCs", 
-       color = "Method", linetype = "Method") +
-  theme(legend.key.width = unit(3, "line")) +
-  theme_bigstatsr()
-
-

### 2018-08-01 save-scripts/new-analysis-of-results3.R

@@ -1,101 +1,0 @@
-library(tidyverse)
-library(bigsnpr)
-library(pkg.paper.PRS)
-
-corr <- readRDS("backingfiles/corr2.rds")
-system.time(
-  results2 <- list.files("results2", full.names = TRUE) %>%
-    read_format_results(corr)
-)  # 198 sec for 1636 of ~1MB
-
-print(results2, width = Inf)
-
-# saveRDS(results2, "results2_all.rds")
-
-####
-
-results2 %>%
-  filter(method %in% c("PRS-max", "logit-simple")) %>%
-  group_by_at(c(vars(starts_with("par")), "method", "thr.r2")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC, n = 1e4), N = n()) %>%
-  print(n = Inf) -> results2.summary
-
-barplot_causal <- function(results) {
-  
-  results %>%
-    mutate(method2 = c_method_r2(method, thr.r2)) %>%
-    myggplot(aes(par.causal, AUC_mean, fill = method2, color = method2)) +
-    geom_hline(yintercept = 0.5, linetype = 2) +
-    geom_col(position = position_dodge(), alpha = 0.3) +
-    geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
-                  position = position_dodge(width = 0.9), color = "black", width = 0.2) + 
-    scale_y_continuous(limits = c(0.5, NA), minor_breaks = 0:20 / 20,
-                       oob = scales::rescale_none) +
-    labs(x = "Causal SNPs (number and location)", y = "Mean of 100 AUCs",
-         fill = "Method", color = "Method")
-}
-
-barplot_causal_all <- function(results) {
-  
-  results %>%
-    barplot_causal() +
-    geom_hline(aes(yintercept = auc_max), linetype = 3, color = "blue",
-               data = data.frame(par.h2 = c(0.5, 0.8), auc_max = c(0.84, 0.94))) +
-    facet_grid(par.dist ~ par.h2)
-}
-
-results2.summary %>%
-  filter(par.model == "simple") %>%
-  barplot_causal_all()
-# ggsave("new-results-all.pdf", scale = 1/90, width = 1235, height = 917)
-
-
-####
-
-results2.summary2 <- results2 %>%
-  filter(method %in% c("PRS-max", "logit-simple")) %>%
[…]

### 2018-08-01 4-results-simus.Rmd
new figures

@@
 
-```{r, fig.asp = 1, out.width = "90%"}
+```{r}
 results4 %>%

### 2018-08-02 paper/draft.tex
save

@@
 PRSs might have a major public health impact, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.
-The ``Clumping+Thresholding'' (C+T) approach, which is is the most common method to derive PRS, uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it suitable and fast.
+The ``Clumping+Thresholding'' (C+T) approach, which is the most common method to derive PRS, uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it easy to use and fast.
 However, previous work showed that jointly estimating SNP effects for computing PRS have the potential to significantly improve the predictive performance of PRS as compared to C+T.
@@
 In this paper, we present an efficient method to jointly estimate SNP effects, allowing for practical application of penalized logistic regression on modern datasets including hundreds of thousands of individuals.
-Our implementation of penalized logistic regression also includes nearly-optimal choices for hyper-parameters.
-The choice of hyper-parameters for a method is very important as it can dramatically impact the predictive performance of the method.
+Moreover, our implementation of penalized logistic regression directly includes automatic choices for hyper-parameters.
+The choice of hyper-parameters for a predictive model is very important since it can dramatically impact its predictive performance.
 As an example, AUC values range from less than 60\% to 90\% in a model with 30 causal SNPs, depending on the p-value threshold in C+T.
@@
 We compare the performance of penalized logistic regression to the C+T method and to a derivation of random forests. 
-Penalized logistic regression consistently achieves higher predictive performance than the two other methods while being faster.
-We also found that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; AUC values increase from 83\% with C+T to 92.5\% with penalized logistic regression.
-We confirmed these results in a data analysis of a case-control study for celiac disease where penalized logistic regression and the standard C+T method achieve AUC of 89\% and of 82\%.
+Penalized logistic regression consistently achieves higher predictive performance than the two other methods while being very fast.
+We find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; AUC values increase from 83\% with the best prediction of C+T to 92.5\% with penalized logistic regression.
+We confirm these results in a data analysis of a case-control study for celiac disease where penalized logistic regression and the standard C+T method achieve AUC of 89\% and of 82\%.
 
-In conclusion, our study demonstrates that penalized logistic regression are applicable to large-scale individual-level data and can achieve more discriminative polygenic risk scores.
+In conclusion, our study demonstrates that penalized logistic regression is applicable to large-scale individual-level data and can achieve more discriminative polygenic risk scores.
 Our implementation is publicly available in the R package bigstatsr.\\
@@
 Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
-PRSs are useful for different applications, such as testing polygenicity of a disease and finding a common genetic contribution between two diseases \cite[]{purcell2009common}.
+PRSs are useful for different applications, such as testing the polygenicity of one disease and finding a common genetic contribution between two diseases \cite[]{purcell2009common}.
 Personalized medicine is another major application of PRSs.
@@
 
-Many methods have been developed to predict disease status, or more generally any phenotype, based on SNP information.
+Several methods have been developed to predict disease status, or more generally any phenotype, based on SNP information.
 A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or even just PRS -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
@@
 We recently developed two R packages, bigstatsr and bigsnpr, for efficiently analyzing large-scale genome-wide data \cite[]{prive2017efficient}.
-Package bigstatsr includes an efficient algorithm with a new implementation for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
+Package bigstatsr now includes an efficient algorithm with a new implementation for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
 In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression against the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
 In this comparison, we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
-For the C+T model, we compare different thresholds of inclusion of SNPs.
-For the penalized logistic regression, we also introduce two improvements.
-First, we present a procedure that we call Cross-Model Selection and Averaging (CMSA) for automatically choosing the regularization parameter of the penalized logistic regression, which directly affects the number of SNPs included in the model.
-Secondly, we show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.
+For the C+T model, we report results for a large grid of hyper-parameters. 
+For the penalized logistic regression, the choice of hyper-parameters is left to the algorithm so that we report only one model for each simulation.
+We also use this penalized logistic regression on a feature-augmented dataset in order to capture not only linear effects, but also recessive and dominant effects.
 
@@
 In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the heritability.
-We also compare different models for simulating new phenotypes, one with only linear effects, and one which combines linear, dominant and interaction-type effects.
-Overall, we found that the penalized logistic regression consistently achieves higher predictive performance while being faster than the C+T and T-Trees methods.
-This demonstrates the feasibility and relevance of this approach for PRS computation.
+We also compare different models for simulating new phenotypes, one with only linear effects, and one that combines linear, dominant and interaction-type effects.
+Overall, we found that the penalized logistic regression consistently achieves higher predictive performance than the C+T and T-Trees methods while being very fast.
+This demonstrates the feasibility and relevance of this approach for PRS computation on large modern datasets.
 
@@
 Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
-For simulations presented later, we first restrict the dataset to controls in order to remove the genetic structure induced by the celiac disease status.
+For simulations presented later, we first restrict this dataset to controls in order to remove the genetic structure induced by the celiac disease status.
 Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.
@@
 We also use this dataset for real data application, in this case keeping all 15,155 individuals.
+Both datasets contain 281,122 SNPs.
 
@@
 We simulate binary phenotypes using a Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}.
-We vary parameters of the simulations in order to match a range of genetic architecture, from low to high polygenicity.
-This was achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere or 30 in the HLA region only), and the heritability (50\% or 80\%). 
-Then, to further increase the proportion of causal variants, instead of using all 22 automosal chromosomes, we restricted the dataset to chromosome 6 only.
-We also considered deviation form the standard normal additive model, drawing effects of causal SNPs from either a Normal or from a Laplace distribution, and computing liability scores either from a ``simple'' model with linear effects only or a ``fancy'' model which combines linear, dominant and interaction-type effects.
-For the ``simple'' model, we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated with a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs) and $\epsilon \sim N(0, 1 - h^2)$.
+We vary parameters of the simulations in order to match a range of genetic architecture from low to high polygenicity.
+This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 chromosomes or 30 in the HLA region of chromosome 6), and the heritability (50\% or 80\%).
+In a second phase, in order to further increase the proportion of causal variants, we restrict the dataset to chromosome 6 only (18,941 SNPs) instead of using all 22 automosal chromosomes (281,122 SNPs).
+We also consider deviation from the standard normal additive model, drawing effects of causal SNPs from either a Normal or from a Laplace distribution, and computing liability scores either from a ``simple'' model with linear effects only or a ``fancy'' model that combines linear, dominant and interaction-type effects.
+For the ``simple'' model, we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs) and $\epsilon \sim N(0, 1 - h^2)$.
 For the ``fancy'' model, we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$
@@
 where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$.
-Note that, because we used the raw genotype counts when coding the interaction terms, the corresponding SNPs also display a marginal effect.
+%Note that, because we use the raw genotype counts when coding the interaction terms, the corresponding SNPs also display a marginal effect.
 
@@
 Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes) and a training set of size 6000.
-It compares all methods described in section \ref{sec:methods}; for each combination of the remaining parameters, results are based on 100 simulations except  for T-Trees, which relies on 5 simulations only (and an heritability of 80\%) because of a much higher computational burden (several hours of computation for a single simulation) as compared to other approaches.
+It compares all methods described in section \ref{sec:methods}.
+For each combination of the remaining parameters, results are based on 100 simulations excepted in the first simulation comparing penalized logistic regression with T-Trees; this simulation relies on 5 simulations only because of a much higher computational burden (several hours of computation for a single simulation) of T-Trees as compared to other approaches.
 Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only.
 Reducing the number of SNPs aims at increasing the polygenicity (the proportion of causal SNPs) of the simulated models and at virtually increasing the sample size \cite[]{dudbridge2013power,marquez2017multiethnic,vilhjalmsson2015modeling}.
-For this scenario, we use only the additive (``simple'') model, but continue to compare all previous different values of the other parameters.
+For this scenario, we use the additive (``simple'') model only, but continue to compare all previous different values of the other parameters.
 Finally, scenario \textnumero3 reuses the whole dataset but this time varying the size of the training set in order to assess how the sample size affects predictive performance of methods.
@@
 First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction,lusted1971signal}.
-In the case of our study, the AUC is the probability that a PRS of a case is greater than the PRS of a control.
+In the case of our study, the AUC is the probability that the PRS of a case is greater than the PRS of a control.
 This measure indicates the extent to which we can distinguish between cases and controls using PRSs.
@@
 
-The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Study (GWAS).
+The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWAS).
 In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$.
-The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another, which form a set of SNPs denoted $S_\text{clumping}$.
+The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another (this set of SNPs is denoted $S_\text{clumping}$).
 Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $p_T$ to be determined.
 Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS.
-In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.
+In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$.
 
-For the C+T method, we report three different scores of prediction: one including all the SNPs remaining after clumping (``PRS-all''), one including only SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``PRS-stringent''), and one that maximizes the AUC (``PRS-max'') for these two thresholds ($0$ and $5 \cdot 10^{-8}$) and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
+For the C+T method, we report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``PRS-all''), one including only SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``PRS-stringent''), and one that maximizes the AUC (``PRS-max'') for these two thresholds ($0$ and $5 \cdot 10^{-8}$) and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
 As we report the optimal threshold based on the test set, the corresponding AUC value is an upper bound of the AUC for the C+T method.
@@
 T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}.
-We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees instead of 1000 because because using 1000 trees provides a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown).
+We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees instead of 1000 because using 1000 trees provides a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown).
 %We call this method ``T-Trees'' in the results.
@@
 the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
-In this study, we use elastic-net regularization with $\alpha = 0.5$.%, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.
+In this study, we use elastic-net regularization with $\alpha \in \{1, 0.5, 0.05, 0.001\}$ (determined thanks to an embedded grid search).
 
-To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in the R package bigstatsr \cite[]{prive2017efficient}.
+To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in the R package bigstatsr.
 This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''.
@@
 In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects.
-This results in a dataset with 3 times as many variables as the initial one, on which we can apply penalized logistic regression with the CMSA procedure (Figure \ref{fig:CMSA}).
-We call this method ``logit-triple'' in the results.
+This results in a dataset with 3 times as many variables as the initial one, on which we can apply penalized logistic regression.
+We refer to this method ``logit-triple'' in the results.
 
@@
 When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}).
-Moreover, penalized logistic regression consistently provides better predictive performance than the C+T method across all scenarios we considered, except in the case of high polygenicity and small sample size, in which case all methods hardly exceed 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).
+Moreover, penalized logistic regression consistently provides better predictive performance than the C+T method across all scenarios we considered, except in some case of high polygenicity or small sample size, in which cases all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).
 
-Method ``logit-simple'' provides particularly higher predictive performance than ``PRS-max'' where there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region.
-In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' approach and 83\% with ``PRS-max'', and the maximum achievable AUC is of 94\% (Figure \ref{fig:main-AUC-logit}).
+Method ``logit-simple'' provides particularly higher predictive performance than ``PRS-max'' when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region.
+In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' approach and 83\% with ``PRS-max'', while the maximum achievable AUC is 94\% (Figure \ref{fig:main-AUC-logit}).
 
@@
 
-In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can impact its predictive performance.
+In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can dramatically impact the predictive performance of C+T.
 For example, AUC ranges from less than 60\% to 90\% in a model with only 30 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
 
-Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of the biglasso implementation \cite[]{zeng2017efficient}, only slightly better for biglasso but likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).
+Concerning the $r^2$ threshold of the clumping step in C+T, we mostly used the commonly used value of $0.2$.
+Yet, using a more stringent value of $0.05$ seems to globally provide higher predictive performance than using $0.2$ (Figures \ref{fig:supp-AUC-all-r2}, \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-chr6-all-r2})
 
-When comparing different values of hyper-parameter $\alpha$ for ``logit-simple'', we report similar predictive performance for $\alpha = 0.5$ and $\alpha = 1$.
-The case of $\alpha = 0.001$ consistently provides lower predictive performance, except in the case where there are 3000 causal SNPs whose effects follow a normal distribution (Figure \ref{fig:supp-alpha}).
-Note that this is also the only case where ``PRS-all'' outperforms logit-simple.
+Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of the biglasso implementation \cite[]{zeng2017efficient}, only slightly better for biglasso but likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}) [GARDER?
+-> SI OUI, A RELANCER].
 
@@
 As compared to ``logit-simple'', T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}).
-Even when simulating a fancy model in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower with T-Trees as compared with ``logit-simple'' (Figure \ref{fig:supp-ttrees}).
+Even when simulating a fancy model in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower when using T-Trees than when using ``logit-simple'' (Figure \ref{fig:supp-ttrees}).
 
-Secondly, we compare the two penalized logistic regressions in scenario \textnumero1, ``logit-simple'' and ``logit-triple'' that uses additional features (variables) coding for recessive and dominant effects.
-Predictive performance of ``logit-triple'' are nearly as good as ``logit-simple'' when there are only linear effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-AUC-triple} and \ref{fig:supp-triple}).
+We also compared the two penalized logistic regressions in scenario \textnumero1, ``logit-simple'' and ``logit-triple'' that uses additional features (variables) coding for recessive and dominant effects.
+Predictive performance of ``logit-triple'' are nearly as good as ``logit-simple'' when there are only linear effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
 For the ``fancy model'', ``logit-triple'' provides AUC values at least 3.5\% higher than ``logit-simple'', except when there are 3000 causal SNPs.
@@
 
-Thirdly, for the ``fancy'' simulation model, we report similar results as in the previous subsection.
-As before, ``logit-simple'' achieves greater AUCs than the ``PRS-max'', except in the case of 3000 causal SNPs randomly chosen anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}).
-Moreover, the choice of the threshold inclusion of SNPs also greatly impacts the performance of the C+T method (Figure \ref{fig:supp-AUC-PRS-fancy}).
+\subsection{Simulations varying number of SNPs and training size} %%%%
 
-\subsection{Simulations varying number of SNPs and training size}
-
-First, when reproducing simulations of scenario \textnumero1 using only chromosome 6 (scenario \textnumero2), the predictive performance of all methods globally increase (Figure \ref{fig:supp-AUC-chr6}).
+First, when reproducing simulations of scenario \textnumero1 using chromosome 6 only (scenario \textnumero2), the predictive performance of all methods globally increase (Figures \ref{fig:supp-AUC-all-r2} and \ref{fig:supp-AUC-chr6-all-r2}).
 There is particularly a large increase when simulating 3000 causal SNPs: AUC from the ``logit-simple'' increases from less than 60\% to nearly 80\%. 
@@
 The relative increase in partial AUC, for specificities larger than 90\%, is even larger (45\% and 50\%) with partial AUC values of 0.0414, 0.0429 and 0.0286 obtained with ``logit-simple'', ``logit-triple'' and the C+T method, respectively.
-Moreover, logistic regressions use less predictors, respectively 3320, 4470 and 9850 (Table \ref{tab:results-celiac}, figures \ref{fig:celiac-roc} and \ref{fig:supp-score-densities} and supplementary notebook ``results-celiac'').
+Moreover, logistic regressions use less predictors, respectively 3320, 4470 and 9850 (Table \ref{tab:results-celiac}, figures \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
 Note that for the C+T method, we still report the best result among 102 p-value thresholds.
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
-\caption{Comparison of models when varying sample size in scenario \textnumero3.
+\caption{[CHANGE] Comparison of models when varying sample size in scenario \textnumero3.
 Mean of AUC over 100 simulations for the three reported PRSs and the ``logit-simple'' as a function of the training size.
@@
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6}}
-\caption{Comparison of models when using only chromosome 6 in scenario \textnumero2. Boxplots of AUC over 100 simulations for the three reported PRSs and the ``logit-simple''.
-Vertical panels represents different number of causal SNPs and their location (``all'': anywhere on the genome; ``HLA'': only in the HLA region).
-Horizontal panels are presenting results for effets following a Gaussian or Laplace distribution and an heritability of 0.5 or 0.8.
-The blue dotted line represents the maximum achievable AUC.
-The ``simple'' model was used to simulate phenotypes.}
-\label{fig:supp-AUC-chr6}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-\begin{figure}[h]
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-logit}}
@@
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-triple}}
-\caption{Comparison of ``logit-triple'' and ``logit-simple'' in scenario \textnumero1.
-Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
-Horizontal panels are presenting results for the ``simple'' and ``fancy'' models for simulating phenotypes.
-Simulations use an heritability of 50\%.
-\textbf{A:} Mean of AUC over 100 simulations.
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-all-r2}}
+\caption{[CHANGE] Comparison of three different threshold used in the C+T method in scenario \textnumero1.
+Mean of AUC over 100 simulations.
+Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.}
-\label{fig:supp-AUC-triple}
+The blue dotted line represents the maximum achievable AUC.
+These results correspond to the ``simple'' model and an heritability of 50\%.}
+\label{fig:supp-AUC-all-r2}
 \end{figure}
@@
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-logit-fancy}}
-\caption{Comparison of the C+T method and ``logit-simple'' in scenario \textnumero1.
-Mean of AUC over 100 simulations for ``logit-simple'' and the maximum AUC reported with the C+T method (``PRS-max'').
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
+\caption{[CHANGE] Comparison of three different threshold used in the C+T method in scenario \textnumero1.
+Mean of AUC over 100 simulations.
 Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
-Left (right) panel is presenting results for an heritability of 50\% (80\%).
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
 The blue dotted line represents the maximum achievable AUC.
-These results correspond to the ``fancy'' model.}
-\label{fig:supp-AUC-logit-fancy}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS-fancy}}
-\caption{Comparison of three different threshold used in the C+T method in scenario \textnumero1.
-Mean of AUC over 100 simulations for ``logit-simple'' and the maximum AUC reported with the C+T method (``PRS-max'').
-Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
-Left (right) panel is presenting results for an heritability of 50\% (80\%).
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.
-These results correspond to the ``fancy'' model.}
-\label{fig:supp-AUC-PRS-fancy}
+These results correspond to the ``simple'' model and an heritability of 50\%.}
+\label{fig:supp-AUC-chr6-all-r2}
 \end{figure}
@@
 \begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-triple}}
+\caption{Comparison of ``logit-triple'' and ``logit-simple'' in scenario \textnumero1.
+Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
+Horizontal panels are presenting results for the ``simple'' and ``fancy'' models for simulating phenotypes.
+Simulations use an heritability of 50\%.
+\textbf{A:} Mean of AUC over 100 simulations.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+The blue dotted line represents the maximum achievable AUC.}
+\label{fig:supp-AUC-triple}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newpage
+\begin{figure}[h]
 \centerline{\includegraphics[width=\textwidth]{supp-biglasso}}
@@
 
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-alpha}}
-\caption{Comparison of different values of the $\alpha$ hyper-parameter of the ``logit-simple'' method in scenario \textnumero1.
-``logit-simple'', ``logit-simple-enet-ridge'' and ``logit-simple-lasso'' corresponds to values of 0.5, 0.001 and 1, respectively.
-Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
-Horizontal panels are presenting results for the ``simple'' model for simulating phenotypes.
-Simulations use an heritability of 80\%.
-\textbf{A:} Mean of AUC over 100 simulations.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.
-\textbf{B:} Boxplots of numbers of predictors used by the methods for 100 simulations.
-\textbf{C:} Boxplots of execution times for 100 simulations.}
-\label{fig:supp-alpha}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-score-densities}}
-\caption{Density of polygenic risk scores for the ``C+T'' (A), ``logit-simple'' (B) and ``logit-triple'' (C) methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.\label{fig:supp-score-densities}}
-\end{figure}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-08-05 4-results-simus.Rmd
figs with black color

@@
     geom_hline(yintercept = 0.5, linetype = 2) +
-    geom_col(position = position_dodge(), alpha = 0.3) +
+    geom_col(position = position_dodge(), alpha = 0.5, color = "black", size = 1) +
     geom_errorbar(aes(ymin = AUC_mean - 2 * AUC_boot, ymax = AUC_mean + 2 * AUC_boot),
-                  position = position_dodge(width = 0.9), color = "black", width = 0.2) + 
+                  position = position_dodge(width = 0.9), color = "black", width = 0.2, size = 1) + 
     scale_y_continuous(limits = c(0.5, NA), minor_breaks = 0:20 / 20,
@@
 ```{r}
-results2 %>%
+results2.all.r2 <- results2 %>%
   filter(par.model == "simple", method %in% c("logit-simple", "PRS-max")) %>%
@@
   group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
-  barplot_causal_all()
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n())
 
-ggsave("figures/main-AUC-all-r2.pdf", scale = 1/110, width = 1242, height = 951)
+barplot_causal_all(results2.all.r2)
+
+ggsave("figures/supp-AUC-all-r2.pdf", scale = 1/110, width = 1242, height = 951)
 ```
@@
   summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n()) %>%
-  barplot_causal_all()
+  barplot_causal_all() + 
+  geom_col(aes(par.causal, AUC_mean), data = results2.all.r2, 
+           position = position_dodge(), color = "black", alpha = 0)
 

### 2018-08-05 paper/draft.tex
save

@@
 
-First, when reproducing simulations of scenario \textnumero1 using chromosome 6 only (scenario \textnumero2), the predictive performance of all methods globally increase (Figures \ref{fig:supp-AUC-all-r2} and \ref{fig:supp-AUC-chr6-all-r2}).
-There is particularly a large increase when simulating 3000 causal SNPs: AUC from the ``logit-simple'' increases from less than 60\% to nearly 80\%. 
-On the contrary, when simulating only 30 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' does not increase. 
+First, when reproducing simulations of scenario \textnumero1 using chromosome 6 only (scenario \textnumero2), the predictive performance of ``logit-simple'' always increase (Figure \ref{fig:supp-AUC-chr6-all-r2}).
+There is particularly a large increase when simulating 3000 causal SNPs: AUC from the ``logit-simple'' increases from 60\% to nearly 80\% for Gaussian effects and a heritability of 80\%.
+On the contrary, when simulating only 30 or 300 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' does not increase, and even decreases for an heritability of 80\% (Figure \ref{fig:supp-AUC-chr6-all-r2}).
 Secondly, when varying the training size (scenario \textnumero3), we report an increase of AUC when increasing the training size, with a faster increase of AUC provided by ``logit-simple'' as compared to ``PRS-max'' (Figure  \ref{fig:main-AUC-ntrain}).
@@
 In terms of computation time, we report only the GWAS computation for the C+T method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once, is faster than the C+T method (84 vs 148 seconds), and the ``logit-triple'' takes less than twice as long as the ``logit-simple'' (141 vs 84 seconds).
+[CHANGE: A RELANCER AVEC GRID-SEARCH]
 
@@
 
-In this comparative study, we present a computationally efficient implementation of penalized logistic regression. 
-%This model can be used to build polygenic risk scores based on very large SNP datasets such as the UK biobank \cite[]{bycroft2017genome}. 
+[REFAIRE] In this comparative study, we present a computationally efficient implementation of penalized logistic regression. 
+This model can be used to build polygenic risk scores based on very large SNP datasets such as the UK biobank \cite[]{bycroft2017genome}. 
 In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently.
 Penalized logistic regression consistently outperform the C+T method, except in the case where there are only small effects and sample size is limited, in which case all methods perform poorly.
-Note that in this particular case, penalized logistic regression could outperform the C+T method if choosing $\alpha = 10^{-3}$, which gives more importance to ridge regularization and accept more predictors in the model.
 
@@
 Indeed, on the one hand, choosing a low value for this threshold may discard independently predictive SNPs that are in LD; yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would bias SNP effects.
+Yet, based on the simulations of this paper, it seems that using stringent threshold ($r^2 = 0.05$) leads to better predictive performance, even in the case where causal SNPs are correlated.
+It means that accurately estimating SNP effects is more important than including all causal SNPs.
 Moreover, in this paper, we reported the maximum AUC of 102 $p_T$ thresholds, a threshold that should normally be learned on the training set only.
@@
 Yet, performing cross-validation and retraining is computationally demanding, CMSA offers a less burdensome alternative.
-Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we show that an hybrid approach ($\alpha = 0.5$) gives similar results as a model using only L1 regularization ($\alpha = 1$) in all simulated models.
+Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we use a grid  search directly embedded in the CMSA procedure. 
 Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model. 
-Yet, when testing with $\alpha = 0.001$, we show that this setting consistently provides lower predictive performance, except in the case where there are many small effects and small sample size, which is a case where all methods perform poorly.
-%In light of this result, we now integrate an optional grid-search procedure in our implementation in order to determine an optimal value for hyper-parameter $\alpha$.
+Yet, if using e.g. $\alpha = 0.001$, it would lead to a similar model.
 
@@
 For this, we introduced a simple feature engineering technique that enables logistic regression to detect and learn not only additive effects, but also dominant and recessive effects.
-This technique improves the predictive performance of logistic regression when there are some non-linear effects in the simulations, while providing the same predictive performance when there are only linear effects.
+This technique improves the predictive performance of logistic regression when there are some non-linear effects in the simulations, while providing nearly the same predictive performance when there are only linear effects.
 Moreover, it also improves predictive performance for the celiac disease.
@@
 consortium data including hundreds of cohorts).
-As of today, the proposed penalized logistic regression does not allow for the analysis of summary data, but represents an important future direction of our work.
+As of today, the proposed penalized logistic regression does not allow for the analysis of summary data, but this represents an important future direction of our work.
 The current version is of particular interest for the analysis of modern SNP dataset including hundreds of thousands of individuals.
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
-\caption{[CHANGE] Comparison of models when varying sample size in scenario \textnumero3.
-Mean of AUC over 100 simulations for the three reported PRSs and the ``logit-simple'' as a function of the training size.
+\caption{Comparison of models when varying sample size in scenario \textnumero3.
+Mean of AUC over 100 simulations for the maximum values of PRSs for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the training size.
 These results are for the ``simple'' model with 300 causal SNPs sampled anywhere on the genome.
-Upper (lower) panels are presenting results for an heritability of 0.5 (0.8) and left (right) panels are presenting results for effets following a Gaussian (Laplace) distribution.
+Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
@@
 
-Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).
+Authors acknowledge LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).
+Authors also acknowledge the Grenoble Alpes Data Institute that is supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02).
 We are also grateful to F\'elix Balazard for useful discussions about T-Trees, and to Yaohui Zeng for useful discussions about R package biglasso.
@@
 Secondly, if we assume that the genetic part of the liabilities $y \sim N(0, h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the heritability $h^2$ through Monte Carlo simulations.
-We report AUCs of 84.1\% and 94.1\% for respectively an heritability of 50\% and 80\%.
+We report AUCs of 84.1\% and 94.1\% for an heritability of 50\% and 80\%, respectively.
 Thirdly, we reproduce the exact same procedure of simulations and, for each combination of parameters (Table \ref{tab:simus}), we estimate the AUC of the ``oracle'', i.e.\ the true simulated genetic part of the liabilities through 100 replicates.
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-all-r2}}
-\caption{[CHANGE] Comparison of three different threshold used in the C+T method in scenario \textnumero1.
-Mean of AUC over 100 simulations.
-Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
+\caption{Comparison of models in scenario \textnumero1.
+Mean of AUC over 100 simulations for the maximum values of PRSs for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the number and location of causal SNPs.
+Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
 The blue dotted line represents the maximum achievable AUC.
-These results correspond to the ``simple'' model and an heritability of 50\%.}
+These results correspond to the ``simple'' model.}
 \label{fig:supp-AUC-all-r2}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
-\caption{[CHANGE] Comparison of three different threshold used in the C+T method in scenario \textnumero1.
-Mean of AUC over 100 simulations.
-Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
+\caption{Comparison of models in scenario \textnumero2 (using chromosome 6 only).
+Thinner lines represents results in scenario \textnumero1 (Figure \ref{fig:supp-AUC-all-r2}).
+Mean of AUC over 100 simulations for the maximum values of PRSs for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the number and location of causal SNPs.
+Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
 The blue dotted line represents the maximum achievable AUC.
-These results correspond to the ``simple'' model and an heritability of 50\%.}
+These results correspond to the ``simple'' model.
+}
 \label{fig:supp-AUC-chr6-all-r2}

### 2018-08-06 paper/draft.tex
now wait review H.

@@
 
-[REFAIRE] In this comparative study, we present a computationally efficient implementation of penalized logistic regression. 
+In this comparative study, we present a computationally efficient implementation of penalized logistic regression. 
 This model can be used to build polygenic risk scores based on very large SNP datasets such as the UK biobank \cite[]{bycroft2017genome}. 
 In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently.
-Penalized logistic regression consistently outperform the C+T method, except in the case where there are only small effects and sample size is limited, in which case all methods perform poorly.
+Penalized logistic regression nearly always outperform the C+T method, and the benefits of using it are stronger with an increasing sample size or when causal SNPs are correlated with one another.
 

### 2018-08-21 5-results-celiac.Rmd
celiac results with grid search

@@
 library(tidyverse)
-myggplot <- function(..., coeff = 1) {
-  ggplot(...) + bigstatsr::theme_bigstatsr(size.rel = coeff)
-} 
+library(bigsnpr)
+library(pkg.paper.PRS)
 ```
@@
 # Read file, see "Preprocessing" notebook
-library(bigsnpr)
 celiac <- snp_attach("backingfiles/celiacQC.rds")
@@
 y <- celiac$fam$affection - 1
+obj.svd <- readRDS("backingfiles/PCA.rds")
 
 G2 <- big_attach("backingfiles/celiacQC_tripled1.rds")
-```
-
-```{r, cache=TRUE}
-# Compute PCA
-obj.svd <- snp_autoSVD(G, CHR, POS, ncores = NCORES)
 ```
@@
 ```{r}
-pAUC <- function(pred, target, p = 0.1) {
-  val.min <- min(target)
-  q <- quantile(pred[target == val.min], probs = 1 - p)
-  ind <- (target != val.min) | (pred > q)
-  tryCatch(bigstatsr::AUC(pred[ind], target[ind]) * p, 
-           error = function(e) p^2 / 2)  # all same prediction
-}
+pAUC <- pkg.paper.PRS:::pAUC
+
 plot_density_AUC <- function(pred, true = y[ind.test]) {
@@
   data.frame(Score = pred, Status = true_fct) %>%
-  myggplot() +
+  ggplot() +
   geom_density(aes(Score, fill = Status), alpha = 0.4) +
@@
                   round(pAUC(pred, true), 4))) + 
+    theme_bigstatsr() +
     theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
@@
                         covar.train = obj.svd$u[ind.train, ],
-                        ncores = NCORES, alpha = 0.5)
+                        ncores = NCORES, alphas = 1)
 )
@@
 (nb.pred1 <- sum(rowSums(sapply(logit, function(l) l$beta.X != 0)) > 0))
-(aucs1 <- apply(preds1, 2, AUC, target = y[ind.test]))
-# Combining all K predictions
-preds1.comb <- rowMeans(preds1)
-AUC(preds1.comb, y[ind.test])
+(aucs1 <- AUC(preds1, target = y[ind.test]))
 ```
@@
 ```{r}
-(p2 <- plot_density_AUC(preds1.comb))
+(p2 <- plot_density_AUC(preds1))
 ```
@@
                          covar.train = obj.svd$u[ind.train, ],
-                         ncores = NCORES, alphas = 0.5)
+                         ncores = NCORES, alphas = 1)
 )
@@
 (nb.pred2 <- sum(rowSums(sapply(logit2, function(l) l$beta.X != 0)) > 0))
-(aucs2 <- apply(preds2, 2, AUC, target = y[ind.test]))
-# Combining all K predictions
-preds2.comb <- rowMeans(preds2)
-AUC(preds2.comb, y[ind.test])
+(aucs2 <- AUC(preds2, target = y[ind.test]))
 ```
@@
 ```{r}
-(p3 <- plot_density_AUC(preds2.comb))
+(p3 <- plot_density_AUC(preds2))
 ```
@@
   "PRS-max"      = prs[, which.max(aucs)],
-  "logit-simple" = preds1.comb,
-  "logit-triple" = preds2.comb
+  "logit-simple" = preds1,
+  "logit-triple" = preds2
 ) %>%
@@
 
-bigstatsr:::MY_THEME(
 ggplot(scores_tidy, aes(d = d, m = Score, color = Method, linetype = Method)) +
-    style_roc(xlab = "1 - Specificity", ylab = "Sensitivity")
-) +
+  style_roc(xlab = "1 - Specificity", ylab = "Sensitivity") +
+  theme_bigstatsr() +
   geom_roc(n.cuts = 0, size = 2) +
@@
 ggsave("figures/celiac-roc.pdf", scale = 1/90, width = 700, height = 700)
-```
-
-## Top scores
-
-First, we get the most significant SNP.
-
-```{r}
-# Most significant SNP
-theone <- which.max(-predict(gwas.train.gc))
-ind0 <- which(G[ind.test, theone] == 0)
-(M <- length(ind0))
-table(Geno = G[, theone], forcats::fct_recode(as.factor(y), Control = "0", Case = "1"))
-```
-
-```{r}
-# one geno
-mean(y[ind.test[ind0]]) 
-# prs max
-indPRS <- order(prs[, which.max(aucs)], decreasing = TRUE)[1:M]
-mean(y[ind.test[indPRS]])
-# logit-simple (CMSA)
-indLS <- order(preds1.comb, decreasing = TRUE)[1:M]
-mean(y[ind.test[indLS]])
-# logit-triple (CMSA)
-indLT <- order(preds2.comb, decreasing = TRUE)[1:M]
-mean(y[ind.test[indLT]])
 ```
@@
 4. the number of SNPs used for the prediction.
-
-#### PRS
-
-```{r}
-PRS <- function(G, infos.chr, infos.pos, 
-                pheno.all, covar.all, 
-                ind.train, ind.test) {
-
-  timing <- system.time({
-
-    # GWAS
-    gwas.train <- big_univLogReg(
-      G, pheno.all[ind.train], ind.train = ind.train, 
-      covar.train = covar.all[ind.train, , drop = FALSE], 
-      ncores = nb_cores()
-    )
-    gwas.train.gc <- snp_gc(gwas.train)
-
-    # Clumping on the test set
-    ind.keep <- snp_clumping(G, infos.chr = infos.chr,
-                             ind.row = ind.test,
-                             thr.r2 = 0.2, 
-                             S = abs(gwas.train.gc$score),
-                             size = 500,
-                             is.size.in.bp = TRUE,
-                             infos.pos = infos.pos,
-                             ncores = nb_cores())
-
-    # PRS
-    thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
-    lpS <- -predict(gwas.train.gc)
-    nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
-    prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
-                   ind.test = ind.test,
-                   ind.keep = ind.keep,
-                   lpS.keep = lpS[ind.keep], 
-                   thr.list = thrs)
-  })[3]
-
-  ind.best <- which.max(apply(prs, 2, AUC, target = pheno.all[ind.test]))
-
-  tibble(
-    method = "PRS-max",
-    eval = list(cbind(prs[, ind.best], pheno.all[ind.test])),
-    timing = timing,
-    nb.preds = nb.pred[ind.best]
-  )
-}
-```
-
-#### Logistic regression
-
-```{r}
-logit.CMSA <- function(G, pheno.all, covar.all, ind.train, ind.test, method, print_all = FALSE) {
-
-  timing <- system.time({
-
-    cmsa.logit <- big_spLogReg(X = G, y01.train = pheno.all[ind.train], 
-                               ind.train = ind.train, 
-                               covar.train = covar.all[ind.train, , drop = FALSE],
-                               ncores = NCORES, alphas = 0.5)
-
-    preds.all <- predict(cmsa.logit, X = G, ind.row = ind.test, 
-                         covar.row = covar.all[ind.test, , drop = FALSE])
-    if (print_all) print(preds.all)
-    preds <- rowMeans(preds.all)
-  })[3]
-
-  tibble(
-    method   = method, 
-    eval     = list(cbind(preds, pheno.all[ind.test])),
-    timing   = timing,
-    nb.preds = sum(rowSums(sapply(cmsa.logit, function(x) x$beta.X) != 0) > 0)
-  )
-}
-```
 
@@
   res <- bind_rows(
-    PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test),
+    filter(PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test), thr.r2 == 0.2),
     logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple"),
@@
   )
+  res$pheno <- list(pheno.all[ind.test])
+
   saveRDS(res, file = res.file)
@@
 ```{r}
-pAUC <- function(pred, target, p = 0.1) {
-  val.min <- min(target)
-  q <- quantile(pred[target == val.min], probs = 1 - p)
-  ind <- (target != val.min) | (pred > q)
-  tryCatch(bigstatsr::AUC(pred[ind], target[ind]) * p, 
-           error = function(e) p^2 / 2)  # all same prediction
-}
-
-format_results <- function(results) {
-  results %>%
+list.files(path = "results6/", full.names = TRUE) %>%
+  map_df(readRDS) %>%
   as_tibble() %>%
   mutate(
-      AUC = map_dbl(eval, ~ bigstatsr::AUC(.x[, 1], .x[, 2])),
-      pAUC = map_dbl(eval, ~ pAUC(.x[, 1], .x[, 2]))
-    )
-}
-
-boot <- function(x, n = 1e5, f = mean) {
-  sd(replicate(n, f(sample(x, replace = TRUE))))
-}
-
-list.files(path = "results6/", full.names = TRUE) %>%
-  map_df(readRDS) %>%
-  format_results() %>%
+    AUC = map2_dbl(pred, pheno, ~bigstatsr::AUC(.x, .y)),
+    pAUC = map2_dbl(pred, pheno, ~pAUC(.x, .y)),
+    nb.preds = map_int(set, length)
+  ) %>%
   group_by(method) %>%

### 2018-08-22 3-simus.Rmd
save

@@
 
+### Scenario n°1 (comparison with biglasso)
+
+```{r}
+library(biglasso)
+library(Matrix)
+```
+
+```{r, eval=FALSE}
+G3 <- bigmemory::big.matrix(nrow(G), ncol(G) + ncol(covar.all),
+                            backingfile = "G-PC",                 
+                            backingpath = "backingfiles")
+big_apply(G, function(X, ind) { G3[, ind] <- X[, ind]; NULL }, a.combine = 'c')
+G3[, ncol(G) + cols_along(covar.all)] <- covar.all
+```
+
+```{r}
+G3 <- bigmemory::attach.big.matrix("backingfiles/G-PC.desc")
+```
+
+```{r}
+logit.biglasso <- function(G3, pheno.all, covar.all, ind.train, ind.test) {
+
+  timing <- system.time({
+
+    biglasso <- biglasso(G3, pheno.all, ind.train, ncores = nb_cores(),
+                         penalty = "lasso", alpha = 1, family = "binomial")
+
+    preds <- 1 / (1 + exp(-predict(biglasso, G3, ind.test)))
+  })[3]
+
+  aucs <- apply(preds, 2, bigstatsr::AUC, target = pheno.all[ind.test])
+  ind.max <- which.max(aucs)
+
+  tibble(
+    method   = "biglasso", 
+    pred     = list(preds[, ind.max]),
+    timing   = timing,
+    alpha    = 1,
+    set      = list(which(head(biglasso$beta[, ind.max], -ncol(covar.all)) != 0))
+  )
+}
+```
+
+```{r}
+params.grid7 <- expand.grid(
+  n.train = 6000,
+  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
+  par.dist   = c("gaussian", "laplace"), 
+  par.h2     = c(0.8), 
+  par.model  = c("simple"),
+  num.simu   = 1:100,
+  stringsAsFactors = FALSE
+) 
+```
+
+```{r, eval=TRUE}
+if (!dir.exists("results7")) dir.create("results7")
+
+for (i in rows_along(params.grid7)) {
+
+  res.file <- paste0("results7/simu_", i, ".rds")
+  if (file.exists(res.file)) next
+
+  params <- params.grid7[i, ]
+  par.causal <- params[["par.causal"]][[1]]
+
+  # Simulate phenotypes
+  simu_pheno <- get_pheno(
+    G,    
+    h2 = params[["par.h2"]], 
+    M = as.integer(par.causal[1]), 
+    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
+    effects.dist = params[["par.dist"]], 
+    model = params[["par.model"]]
+  )
+  pheno.all <- simu_pheno$pheno
+  params[["true_set"]] <- list(simu_pheno$set)
+
+  # Split in training/test sets
+  ind.train <- sort(sample(n, size = params[["n.train"]]))
+  ind.test <- setdiff(1:n, ind.train)
+  params[["pheno"]] <- list(pheno.all[ind.test])
+
+  # Get results from all methods
+  res <- bind_rows(
+    logit.CMSA(G, pheno.all, covar.all, ind.train, ind.test, "logit-simple", alphas = 1),
+    logit.biglasso(G3, pheno.all, covar.all, ind.train, ind.test)
+  )
+  params[["res"]] <- list(res)
+  saveRDS(unnest(params, res, .drop = FALSE), file = res.file)
+}
+```
+
 ```{r, include=FALSE, error=TRUE}

### 2018-08-22 4-results-simus.Rmd
save

@@
 methods.color <- setNames(
-  c(COLORS, COLORS[c(4, 2, 5)]), 
+  c(COLORS, COLORS[c(4, 2, 5)], "black"), 
   c("logit-simple", "PRS-max", "PRS-stringent", "T-Trees", "logit-triple", "PRS-all",
-    "PRS-max-0.05", "PRS-max-0.2", "PRS-max-0.8")
+    "PRS-max-0.05", "PRS-max-0.2", "PRS-max-0.8", "biglasso")
 ) 
@@
 ```
+
+### Comparison with biglasso
+
+```{r, fig.asp=0.65, out.width="95%", fig.width=16}
+list.files("results7", full.names = TRUE) %>%
+  read_format_results(corr) %>%
+  compare_logit("biglasso")
+
+ggsave("figures/supp-biglasso.pdf", scale = 1/100, width = 1600, height = 1000)
+```

### 2018-08-22 paper/draft.tex
save

@@
 
-\title{Efficient implementation of penalized models and their application to genetic risk prediction}
+\title{Efficient implementation of penalized regression for genetic risk prediction}
 \author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
@@
 Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
-PRSs are useful for different applications, such as testing the polygenicity of one disease and finding a common genetic contribution between two diseases \cite[]{purcell2009common}.
+PRSs are useful for genetic epidemiology when testing the polygenicity of one disease and finding a common genetic contribution between two diseases \cite[]{purcell2009common}.
 Personalized medicine is another major application of PRSs.
@@
 For the C+T model, we report results for a large grid of hyper-parameters. 
-For the penalized logistic regression, the choice of hyper-parameters is left to the algorithm so that we report only one model for each simulation.
+For the penalized logistic regression, the choice of hyper-parameters is included in the algorithm so that we report only one model for each simulation.
 We also use this penalized logistic regression on a feature-augmented dataset in order to capture not only linear effects, but also recessive and dominant effects.
 
-For the simulations of this paper, we use real genotype data and simulate new phenotypes \cite[]{zhou2013polygenic}. 
+To perform simulations, we use real genotype data and simulate new phenotypes \cite[]{zhou2013polygenic}. 
 In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the heritability.
-We also compare different models for simulating new phenotypes, one with only linear effects, and one that combines linear, dominant and interaction-type effects.
+We also compare different models for simulating phenotypes, one with only linear effects, and one that combines linear, dominant and interaction-type effects.
 Overall, we found that the penalized logistic regression consistently achieves higher predictive performance than the C+T and T-Trees methods while being very fast.
@@
 Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.
-In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on the robust Mahalanobis distances computed using the first 10 principal components of the remaining individuals.
+In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on a robust Mahalanobis distances computed using the first 10 principal components of the remaining individuals.
 This 3-step filtering process results in a sample of 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing'').
-We also use this dataset for real data application, in this case keeping all 15,155 individuals.
+We also use this dataset for real data application, in this case keeping all 15,155 individuals (4496 cases and 10,659 controls).
 Both datasets contain 281,122 SNPs.
@@
 We also consider deviation from the standard normal additive model, drawing effects of causal SNPs from either a Normal or from a Laplace distribution, and computing liability scores either from a ``simple'' model with linear effects only or a ``fancy'' model that combines linear, dominant and interaction-type effects.
-For the ``simple'' model, we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs) and $\epsilon \sim N(0, 1 - h^2)$.
-For the ``fancy'' model, we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$
-and $S_\text{causal}^{(3)}$; $S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$.
-We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction} ~~+~~ \epsilon_i~,$$
-where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$.
-%Note that, because we use the raw genotype counts when coding the interaction terms, the corresponding SNPs also display a marginal effect.
+For the ``simple'' model, we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$ and $S_\text{causal}$ is the set of causal SNPs.
+For the ``fancy'' model, we simulate phenotypes using linear, dominant and interaction-type effects (see Supplementary Materials).
 
@@
 Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $p_T$ to be determined.
-Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS.
+Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS.
 In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$.
@@
 
-Finally, for the penalized logistic regression, we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function: $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict, $\lambda$ and $\alpha$ are two regularization parameters that need to be chosen.
+Finally, for the penalized logistic regression, we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization parameters that need to be chosen.
 Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
@@
 
-In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects.
+In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects: one variable is coded 1 if homozygous variant and 0 otherwise, and the other is coded 0 for homozygous referent and 1 otherwise.
 This results in a dataset with 3 times as many variables as the initial one, on which we can apply penalized logistic regression.
@@
 When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}).
-Moreover, penalized logistic regression consistently provides better predictive performance than the C+T method across all scenarios we considered, except in some case of high polygenicity or small sample size, in which cases all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).
+Moreover, penalized logistic regression consistently provides better predictive performance than the C+T method across all scenarios we considered, except in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).
 
@@
 In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can dramatically impact the predictive performance of C+T.
-For example, AUC ranges from less than 60\% to 90\% in a model with only 30 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
+For example, in a model with only 30 causal SNPs, AUC ranges from less than 60\% when using all SNPs passing clumping to 90\% if choosing the optimal p-value threshold (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
 
 Concerning the $r^2$ threshold of the clumping step in C+T, we mostly used the commonly used value of $0.2$.
-Yet, using a more stringent value of $0.05$ seems to globally provide higher predictive performance than using $0.2$ (Figures \ref{fig:supp-AUC-all-r2}, \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-chr6-all-r2})
+Yet, using a more stringent value of $0.05$ provides higher predictive performance than using $0.2$ in most of the cases tested in this paper (Figures \ref{fig:supp-AUC-all-r2}, \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-chr6-all-r2})
 
-Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of the biglasso implementation \cite[]{zeng2017efficient}, only slightly better for biglasso but likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}) [GARDER?
--> SI OUI, A RELANCER].
+Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of the biglasso implementation \cite[]{zeng2017efficient}, only slightly better for biglasso but likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).
 
@@
 
-Joint logistic regressions provide higher AUC values for the Celiac data: 88.8\% with ``logit-simple'' and 89.2\% with ``logit-triple'' as compared to 82.4\% with the C+T method.
-The relative increase in partial AUC, for specificities larger than 90\%, is even larger (45\% and 50\%) with partial AUC values of 0.0414, 0.0429 and 0.0286 obtained with ``logit-simple'', ``logit-triple'' and the C+T method, respectively.
-Moreover, logistic regressions use less predictors, respectively 3320, 4470 and 9850 (Table \ref{tab:results-celiac}, figures \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
+Joint logistic regressions provide higher AUC values for the Celiac data: 88.7\% with ``logit-simple'' and 89.1\% with ``logit-triple'' as compared to 82.5\% with the C+T method.
+The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with ``logit-simple'', ``logit-triple'' and the C+T method, respectively.
+Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360 (Table \ref{tab:results-celiac}, figures \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
 Note that for the C+T method, we still report the best result among 102 p-value thresholds.
-In terms of computation time, we report only the GWAS computation for the C+T method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once, is faster than the C+T method (84 vs 148 seconds), and the ``logit-triple'' takes less than twice as long as the ``logit-simple'' (141 vs 84 seconds).
-[CHANGE: A RELANCER AVEC GRID-SEARCH]
+In terms of computation time, we report only the GWAS computation for the C+T method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once and testing different hyper-parameter values, is almost as fast as the C+T method (190 vs 130 seconds), and the ``logit-triple'' takes less than twice as long as the ``logit-simple'' (296 vs 190 seconds).
 
@@
   \hline
-PRS-max & 0.824 (0.000704) & 0.0286 (0.00016) & 9850 (781) & 148 (0.414) \\
-logit-simple & 0.888 (0.000468) & 0.0414 (0.000164) & 3220 (62) & 83.8 (1.27) \\
-logit-triple & 0.892 (0.000488) & 0.0429 (0.000174) & 4470 (80.6) & 141 (1.85) \\
+PRS-max & 0.825 (0.000664) & 0.0289 (0.000187) & 8360 (744) & 130 (0.143) \\ 
+logit-simple & 0.887 (0.00061) & 0.0411 (0.000224) & 1570 (46.4) & 190 (1.21) \\ 
+logit-triple & 0.891 (0.000628) & 0.0426 (0.000219) & 2260 (56.1) & 296 (2.03) \\ 
    \hline
@@
 
-The choice of hyper-parameters' values is very important as it can greatly impact method performance.
+The choice of hyper-parameter values is very important as it can greatly impact method performance.
 In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps, respectively.
@@
 Indeed, on the one hand, choosing a low value for this threshold may discard independently predictive SNPs that are in LD; yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would bias SNP effects.
-Yet, based on the simulations of this paper, it seems that using stringent threshold ($r^2 = 0.05$) leads to better predictive performance, even in the case where causal SNPs are correlated.
+Based on the simulations, we find that using stringent threshold ($r^2 = 0.05$) leads to better predictive performance, even when causal SNPs are correlated.
 It means that accurately estimating SNP effects is more important than including all causal SNPs.
-Moreover, in this paper, we reported the maximum AUC of 102 $p_T$ thresholds, a threshold that should normally be learned on the training set only.
+Moreover, in this paper, we reported the maximum AUC of 102 different p-value thresholds, a threshold that should normally be learned on the training set only.
 The choice of this threshold is very important as it can greatly impact the predictive performance of the C+T method, which we confirm in this study \cite[]{ware2017heterogeneity}.
@@
 Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we use a grid  search directly embedded in the CMSA procedure. 
-Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model. 
-Yet, if using e.g. $\alpha = 0.001$, it would lead to a similar model.
+%Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model.
+Yet, if using e.g.
+$\alpha = 0.001$, it would lead to a similar model.
 
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
-\caption{Main comparison of the C+T method and ``logit-simple'' in scenario \textnumero1.
+\caption{Main comparison of the C+T and ``logit-simple'' methods in scenario \textnumero1 for the ``simple'' model and an heritability of 80\%.
 Mean of AUC over 100 simulations for ``logit-simple'' and the maximum AUC reported with the C+T method (``PRS-max'').
@@
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.
-These results corresponds to the ``simple'' model and an heritability of 80\%.}
+The blue dotted line represents the maximum achievable AUC.}
 \label{fig:main-AUC-logit}
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-PRS}}
-\caption{Comparison of three different threshold used in the C+T method in scenario \textnumero1.
+\caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for the ``simple'' model and an heritability of 80\%.
 Mean of AUC over 100 simulations.
@@
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.
-These results corresponds to the ``simple'' model and an heritability of 80\%.}
+The blue dotted line represents the maximum achievable AUC.}
 \label{fig:main-AUC-PRS}
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
-\caption{Comparison of models when varying sample size in scenario \textnumero3.
+\caption{Comparison of models when varying sample size in scenario \textnumero3 for the ``simple'' model with 300 causal SNPs sampled anywhere on the genome.
 Mean of AUC over 100 simulations for the maximum values of PRSs for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the training size.
-These results are for the ``simple'' model with 300 causal SNPs sampled anywhere on the genome.
 Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
@@
 
+\subsection*{``Fancy'' model}
+
+For the ``fancy'' model, we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$
+and $S_\text{causal}^{(3)}$; $S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$.
+We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction} ~~+~~ \epsilon_i~,$$
+where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$.
+%Note that, because we use the raw genotype counts when coding the interaction terms, the corresponding SNPs also display a marginal effect.
+
 \subsection*{Maximum AUCs} \label{sec:auc-max}
@@
 Note that this approximation is reported to be less accurate for high heritabilities.
-Secondly, if we assume that the genetic part of the liabilities $y \sim N(0, h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the heritability $h^2$ through Monte Carlo simulations.
+Secondly, if we assume that the genetic part of the liabilities follows a Gaussian distribution $N(0, h^2)$ and that the environmental part follows a Gaussian distribution $N(0, 1 - h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the heritability $h^2$ through Monte Carlo simulations.
 We report AUCs of 84.1\% and 94.1\% for an heritability of 50\% and 80\%, respectively.
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-logit}}
-\caption{Comparison of the C+T method and ``logit-simple'' in scenario \textnumero1.
+\caption{Comparison of the C+T and ``logit-simple'' methods in scenario \textnumero1 for the ``simple'' model and an heritability of 50\%.
 Mean of AUC over 100 simulations for ``logit-simple'' and the maximum AUC reported with the C+T method (``PRS-max'').
@@
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.
-These results correspond to the ``simple'' model and an heritability of 50\%.}
+The blue dotted line represents the maximum achievable AUC.}
 \label{fig:supp-AUC-logit}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS}}
-\caption{Comparison of three different threshold used in the C+T method in scenario \textnumero1.
+\caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for the ``simple'' model and an heritability of 50\%.
 Mean of AUC over 100 simulations.
@@
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.
-These results correspond to the ``simple'' model and an heritability of 50\%.}
+The blue dotted line represents the maximum achievable AUC.}
 \label{fig:supp-AUC-PRS}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-all-r2}}
-\caption{Comparison of models in scenario \textnumero1.
+\caption{Comparison of models in scenario \textnumero1 for the ``simple'' model.
 Mean of AUC over 100 simulations for the maximum values of PRSs for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the number and location of causal SNPs.
@@
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.
-These results correspond to the ``simple'' model.}
+The blue dotted line represents the maximum achievable AUC.}
 \label{fig:supp-AUC-all-r2}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
-\caption{Comparison of models in scenario \textnumero2 (using chromosome 6 only).
+\caption{Comparison of models in scenario \textnumero2 (using chromosome 6 only) for the ``simple'' model.
 Thinner lines represents results in scenario \textnumero1 (Figure \ref{fig:supp-AUC-all-r2}).
@@
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.
-These results correspond to the ``simple'' model.
-}
+The blue dotted line represents the maximum achievable AUC.}
 \label{fig:supp-AUC-chr6-all-r2}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-ttrees}}
-\caption{Comparison of T-Trees and ``logit-simple'' in scenario \textnumero1.
+\caption{Comparison of T-Trees and ``logit-simple'' in scenario \textnumero1 for an heritability of 80\%.
 Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
 Horizontal panels are presenting results for the ``simple'' and ``fancy'' models for simulating phenotypes.
-Simulations use an heritability of 80\%.
 \textbf{A:} Mean of AUC over 5 simulations.
@@
 \centerline{\includegraphics[width=\textwidth]{supp-triple}}
-\caption{Comparison of ``logit-triple'' and ``logit-simple'' in scenario \textnumero1.
+\caption{Comparison of ``logit-triple'' and ``logit-simple'' in scenario \textnumero1 for an heritability of 80\%.
 Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
 Horizontal panels are presenting results for the ``simple'' and ``fancy'' models for simulating phenotypes.
-Simulations use an heritability of 80\%.
 \textbf{A:} Mean of AUC over 100 simulations.
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-triple}}
-\caption{Comparison of ``logit-triple'' and ``logit-simple'' in scenario \textnumero1.
+\caption{Comparison of ``logit-triple'' and ``logit-simple'' in scenario \textnumero1 for an heritability of 50\%.
 Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
 Horizontal panels are presenting results for the ``simple'' and ``fancy'' models for simulating phenotypes.
-Simulations use an heritability of 50\%.
 \textbf{A:} Mean of AUC over 100 simulations.
@@
 \centerline{\includegraphics[width=\textwidth]{supp-biglasso}}
-\caption{Comparison of ``logit-triple'' and the best prediction for ``biglasso'' (another implementation of penalized logistic regression without procedure for choosing the $\lambda$ hyper-parameter) in scenario \textnumero1.
+\caption{Comparison of ``logit-triple'' and the best prediction (among 100 tested $\lambda$ values) for ``biglasso'' (another implementation of penalized logistic regression) in scenario \textnumero1.
+Simulations use the ``simple'' model, an heritability of 80\% and $\alpha = 1$.
 Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
-Horizontal panels are presenting results for the ``simple'' model for simulating phenotypes.
-Simulations use an heritability of 80\%.
 \textbf{A:} Mean of AUC over 100 simulations.

### 2018-08-28 2-newfiles.Rmd
update
polish SupMat

@@
 author: "Florian Privé"
-date: "20 octobre 2017"
+date: "October 20, 2017"
 output: html_document
@@
 ```{r}
+celiac <- snp_attach("backingfiles/celiacQC.rds")
 snp_triple(celiac)

### 2018-08-28 3-oracle.Rmd
polish SupMat

@@
 library(bigsnpr)
-library(foreach)
+library(doParallel)
 celiac2 <- snp_attach("backingfiles/celiacQC_sub1.rds")
@@
 POS <- celiac2$map$physical.pos
-(G <- celiac2$genotypes)
+G <- celiac2$genotypes
 ind.HLA <- snp_indLRLDR(CHR, POS, subset(LD.wiki34, ID == "hild12"))
-NCORES <- nb_cores()
-```
-
-```{r}
-myggplot <- function(..., coeff = 1) {
-  bigstatsr:::MY_THEME(ggplot(...), coeff = coeff)
-} 
-```
-
-```{r}
-boot <- function(x, n = 1e5, f = mean) {
-  sd(replicate(n, f(sample(x, replace = TRUE))))
-}
 ```
@@
 
-doParallel::registerDoParallel(NCORES)
+registerDoParallel(cl <- makeCluster(nb_cores()))
 aucs <- foreach(i = seq_len(100), .combine = "cbind") %:%
   foreach(h2 = c(0.5, 0.8), .combine = "rbind") %dopar% {
+
     S <- rnorm(nsimu, mean = 0, sd = sqrt(h2))
@@
   }
-doParallel::stopImplicitCluster()
+stopCluster(cl)
+
 apply(aucs, 1, summary)
@@
   pheno <- as.numeric(y.simu > qnorm(1 - K))
-  AUC(S, pheno)
+  bigstatsr::AUC(S, pheno)
 }
@@
     M = as.integer(par.causal[1]), 
-    ind.possible = `if`(par.causal[2] == "all", cols_along(G), ind.HLA),
+    ind.possible = `if`(par.causal[2] == "all", bigstatsr::cols_along(G), ind.HLA),
     effects.dist = params[["par.dist"]], 
@@
 
-#### On all 22 chromosomes
+#### With all 22 chromosomes
 
 ```{r}
-doParallel::registerDoParallel(NCORES)
+registerDoParallel(cl <- makeCluster(nb_cores()))
 oracles <- foreach(i = rows_along(params.grid2), .combine = "rbind") %dopar% {
@@
 }
-doParallel::stopImplicitCluster()
+stopCluster(cl)
+```
+
+```{r}
+boot <- pkg.paper.PRS::boot
 ```
@@
                         levels = c("30 in HLA", paste(3 * 10^(1:3), "in all")))) %>%
-  group_by(par.causal, par.dist, par.h2, par.model) %>%
+  group_by_at(vars(starts_with("par"))) %>%
   summarise(AUC_mean = round(mean(AUC), 4),
             AUC_mean_boot = round(boot(AUC), 4)) %>%
+  arrange(AUC_mean) %>%
   print(n = Inf)
@@
 
-#### On only the chromosome 6
+#### With chromosome 6 only
 
@@
 
-doParallel::registerDoParallel(NCORES)
+registerDoParallel(cl <- makeCluster(nb_cores()))
 oracles_chr6 <- foreach(i = rows_along(params.grid2), .combine = "rbind") %dopar% {
@@
 }
-doParallel::stopImplicitCluster()
+stopCluster(cl)
 ```
@@
 ```{r}
-tmp <- oracles_chr6 %>% 
+oracles_chr6 %>% 
   mutate(

### 2018-08-29 paper/draft.tex
save before Biorxiv

@@
 
-\title{Efficient implementation of penalized regression for genetic risk prediction}
+\title{Efficient implementation of penalized regression\\for genetic risk prediction}
 \author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
@@
 Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
-PRSs might have a major public health impact, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.
-The ``Clumping+Thresholding'' (C+T) approach, which is the most common method to derive PRS, uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it easy to use and fast.
-However, previous work showed that jointly estimating SNP effects for computing PRS have the potential to significantly improve the predictive performance of PRS as compared to C+T.
+PRS might have a major public health impact, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.
+The ``Clumping+Thresholding'' (C+T) approach, which is the most common method to derive PRS, uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it fast and easy to use.
+However, previous work showed that jointly estimating SNP effects for computing PRS has the potential to significantly improve the predictive performance of PRS as compared to C+T.
 
@@
 We find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; AUC values increase from 83\% with the best prediction of C+T to 92.5\% with penalized logistic regression.
-We confirm these results in a data analysis of a case-control study for celiac disease where penalized logistic regression and the standard C+T method achieve AUC of 89\% and of 82\%.
+We confirm these results in a data analysis of a case-control study for celiac disease where penalized logistic regression and the standard C+T method achieve AUC of 89\% and of 82.5\%.
 
 In conclusion, our study demonstrates that penalized logistic regression is applicable to large-scale individual-level data and can achieve more discriminative polygenic risk scores.
-Our implementation is publicly available in the R package bigstatsr.\\
+Our implementation is publicly available in R package bigstatsr.\\
 
@@
 
-Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
-PRSs are useful for genetic epidemiology when testing the polygenicity of one disease and finding a common genetic contribution between two diseases \cite[]{purcell2009common}.
-Personalized medicine is another major application of PRSs.
-Personalized medicine envisions to use PRSs in screening campaigns in order to identify high-risk individuals for a given disease \cite[]{chatterjee2016developing}.
+Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
+PRS are useful for genetic epidemiology when testing the polygenicity of one disease and finding a common genetic contribution between two diseases \cite[]{purcell2009common}.
+Personalized medicine is another major application of PRS.
+Personalized medicine envisions to use PRS in screening campaigns in order to identify high-risk individuals for a given disease \cite[]{chatterjee2016developing}.
 As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}.
-Yet, PRSs would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases.
-For example, for screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.
+Yet, PRS would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases.
+For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.
 
 Several methods have been developed to predict disease status, or more generally any phenotype, based on SNP information.
-A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or even just PRS -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
+A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or even just PRS -- is used to derive PRS from results of Genome-Wide Association Studies (GWAS) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
 This technique uses GWAS summary statistics only, allowing for fast implementation.
 However, the ``C+T'' approach also has several limitations.
-Previous studies showed that predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
+Previous studies have shown that predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
 Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}.
@@
 Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
-Finally, statistical learning methods have also been used to derive PRSs for complex human diseases by jointly estimating SNP effects.
+Finally, statistical learning methods have also been used to derive PRS for complex human diseases by jointly estimating SNP effects.
 Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,wei2009disease}.
@@
 We also compare different models for simulating phenotypes, one with only linear effects, and one that combines linear, dominant and interaction-type effects.
-Overall, we found that the penalized logistic regression consistently achieves higher predictive performance than the C+T and T-Trees methods while being very fast.
+Overall, we find that the penalized logistic regression consistently achieves higher predictive performance than the C+T and T-Trees methods while being very fast.
 This demonstrates the feasibility and relevance of this approach for PRS computation on large modern datasets.
@@
 In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on a robust Mahalanobis distances computed using the first 10 principal components of the remaining individuals.
-This 3-step filtering process results in a sample of 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing'').
+This 3-step filtering process results in a sample of 7100 individuals with minimal population structure (see supplementary notebook ``preprocessing'').
 We also use this dataset for real data application, in this case keeping all 15,155 individuals (4496 cases and 10,659 controls).
@@
 In the case of our study, the AUC is the probability that the PRS of a case is greater than the PRS of a control.
-This measure indicates the extent to which we can distinguish between cases and controls using PRSs.
-As a second measure, the partial AUC for specificities between 90\% and 100\% is also reported \cite[]{dodd2003partial,mcclish1989analyzing}.
+This measure indicates the extent to which we can distinguish between cases and controls using PRS. 
+As a second measure, we also report the partial AUC for specificities between 90\% and 100\% \cite[]{dodd2003partial,mcclish1989analyzing}.
 This measure is similar to the AUC, but focuses on high specificities, which is the most useful part of the ROC curve in clinical settings.
-When reporting AUC results of simulations, we use estimates of maximum achievable AUC of 84\% and 94\% for heritabilities of respectively 50\% and 80\%, based on three different yet consistent estimations (see Supplementary Materials).
+When reporting AUC results of simulations, we use estimates of maximum achievable AUC of 84\% and 94\% for heritabilities of respectively 50\% and 80\%.
+These estimates are based on three different yet consistent estimations (see Supplementary Materials).
 %Note that we also report the timing of the main computations and the number of SNPs used in the predictions.
@@
 In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$.
-
-For the C+T method, we report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``PRS-all''), one including only SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``PRS-stringent''), and one that maximizes the AUC (``PRS-max'') for these two thresholds ($0$ and $5 \cdot 10^{-8}$) and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
-As we report the optimal threshold based on the test set, the corresponding AUC value is an upper bound of the AUC for the C+T method.
+We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``PRS-all''), one including only SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``PRS-stringent''), and one that maximizes the AUC (``PRS-max'') for these two thresholds ($0$ and $5 \cdot 10^{-8}$) and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
+As we report the optimal threshold based on the test set, the AUC for ``PRS-max'' is an upper bound of the AUC for the C+T method.
 
@@
 
-Finally, for the penalized logistic regression, we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization parameters that need to be chosen.
+Finally, for the penalized logistic regression, we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
 Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
@@
 the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
-In this study, we use elastic-net regularization with $\alpha \in \{1, 0.5, 0.05, 0.001\}$ (determined thanks to an embedded grid search).
+In this study, we use an embedded grid search over $\alpha \in \{1, 0.5, 0.05, 0.001\}$.
 
-To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in the R package bigstatsr.
+To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in R package bigstatsr.
 This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''.
 To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
-Because of L1-regularization, the resulting vectors of coefficients are typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts.
+Because of L1-regularization, the resulting vectors of coefficients are sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts.
 We refer to this method as ``logit-simple'' in the results section.
@@
 
-When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}).
-Moreover, penalized logistic regression consistently provides better predictive performance than the C+T method across all scenarios we considered, except in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).
+When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-PRS}).
+Moreover, penalized logistic regression consistently provides higher predictive performance than the C+T method across all scenarios we considered, excepted in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).
 
 Method ``logit-simple'' provides particularly higher predictive performance than ``PRS-max'' when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region.
-In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' approach and 83\% with ``PRS-max'', while the maximum achievable AUC is 94\% (Figure \ref{fig:main-AUC-logit}).
+In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' approach and 84\% with ``PRS-max'', while the maximum achievable AUC is 94\% (Figure \ref{fig:main-AUC-logit}).
 
@@
 
-Concerning the $r^2$ threshold of the clumping step in C+T, we mostly used the commonly used value of $0.2$.
+Concerning the $r^2$ threshold of the clumping step in C+T, we mostly used the common value of $0.2$.
 Yet, using a more stringent value of $0.05$ provides higher predictive performance than using $0.2$ in most of the cases tested in this paper (Figures \ref{fig:supp-AUC-all-r2}, \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-chr6-all-r2})
 
-Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of the biglasso implementation \cite[]{zeng2017efficient}, only slightly better for biglasso but likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).
+Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of the implementation of R package biglasso \cite[]{zeng2017efficient}, only slightly better for biglasso, which is likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).
 
@@
 As compared to ``logit-simple'', T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}).
-Even when simulating a fancy model in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower when using T-Trees than when using ``logit-simple'' (Figure \ref{fig:supp-ttrees}).
+Even when simulating a ``fancy'' model in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower when using T-Trees than when using ``logit-simple'' (Figure \ref{fig:supp-ttrees}).
 
@@
 Predictive performance of ``logit-triple'' are nearly as good as ``logit-simple'' when there are only linear effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
-For the ``fancy model'', ``logit-triple'' provides AUC values at least 3.5\% higher than ``logit-simple'', except when there are 3000 causal SNPs.
-Yet, the ``triple'' solution takes 2-3 times as much time to run and requires 3 times as much disk storage as the ``logit-simple''.
+For the ``fancy model'', ``logit-triple'' provides AUC values at least 3.5\% higher than ``logit-simple'', excepted when there are 3000 causal SNPs.
+Yet, the ``triple'' solution takes 2-3 times as much time to run and requires 3 times as much disk storage as the ``simple'' solution.
 
@@
 
-Joint logistic regressions provide higher AUC values for the Celiac data: 88.7\% with ``logit-simple'' and 89.1\% with ``logit-triple'' as compared to 82.5\% with the C+T method.
+Joint logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with ``logit-simple'' and 89.1\% with ``logit-triple'' as compared to 82.5\% with the C+T method.
 The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with ``logit-simple'', ``logit-triple'' and the C+T method, respectively.
@@
 In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently.
-Penalized logistic regression nearly always outperform the C+T method, and the benefits of using it are stronger with an increasing sample size or when causal SNPs are correlated with one another.
+Penalized logistic regression nearly always outperform the C+T method, and the benefits of using it are more pronounced with an increasing sample size or when causal SNPs are correlated with one another.
 
@@
 
-The choice of hyper-parameter values is very important as it can greatly impact method performance.
+The choice of hyper-parameter values is very important since it can greatly impact method performance.
 In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps, respectively.
 The choice of the $r^2$ threshold of the clumping step is important.
-Indeed, on the one hand, choosing a low value for this threshold may discard independently predictive SNPs that are in LD; yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would bias SNP effects.
-Based on the simulations, we find that using stringent threshold ($r^2 = 0.05$) leads to better predictive performance, even when causal SNPs are correlated.
+Indeed, on the one hand, choosing a low value for this threshold may discard independently predictive SNPs that are in Linkage Desiquilibrium; yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would bias SNP effects.
+Based on the simulations, we find that using a stringent threshold ($r^2 = 0.05$) leads to higher predictive performance, even when causal SNPs are correlated.
 It means that accurately estimating SNP effects is more important than including all causal SNPs.
@@
 On the contrary, in the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$ that accounts for the amount of regularization used in the model.
-Not only this procedure provides near-optimal results, but it also accelerates the training of the model thanks to the development of an early stopping criterion.
-Usually, cross-validation is used to choose hyper-parameter value and then the model is trained again with this particular hyper-parameter value \cite[]{Hastie2008,wei2013large}.
-Yet, performing cross-validation and retraining is computationally demanding, CMSA offers a less burdensome alternative.
+Not only this procedure provides near-optimal results (as compared to the best prediction when using R package biglasso), but it also accelerates the training of the model thanks to the development of an early stopping criterion.
+Usually, cross-validation is used to choose hyper-parameter values and then the model is trained again with these particular hyper-parameter values \cite[]{Hastie2008,wei2013large}.
+Yet, performing cross-validation and retraining the model is computationally demanding; CMSA offers a less burdensome alternative.
 Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we use a grid  search directly embedded in the CMSA procedure. 
@@
 In order to capture interaction-type effects, we tested T-Trees, a method that is able to exploit SNP correlations thanks to special decision trees \cite[]{botta2014exploiting}. 
-However, predictive performance of T-Trees were consistently lower than penalized logistic regression, even when simulating a fancy model in which there are dominant and interaction-type effects that T-Trees should be able to handle.
+However, predictive performance of T-Trees were consistently lower than with penalized logistic regression, even when simulating a model with dominant and interaction-type effects that T-Trees should be able to handle.
 
@@
 Our approach has one major limitation: the main advantage of the C+T method is that it is applicable directly to summary statistics, allowing to leverage the largest GWAS sample size to date, even when individual cohort data cannot be merged because of practical and ethical reasons (e.g.
-consortium data including hundreds of cohorts).
+consortium data including many cohorts).
 As of today, the proposed penalized logistic regression does not allow for the analysis of summary data, but this represents an important future direction of our work.
@@
 Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}.
-In another study, we will assess how can we use models and effects learned in one population to improve learning and prediction in another population.
+In next study, we will assess how can we use models and effects learned in one population to improve learning and prediction in another population.
 %We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}.
@@
 \caption{Comparison of models when varying sample size in scenario \textnumero3 for the ``simple'' model with 300 causal SNPs sampled anywhere on the genome.
-Mean of AUC over 100 simulations for the maximum values of PRSs for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the training size.
+Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the training size.
 Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
@@
 For every combination of parameters, AUC of oracles are comprised between 83.2\% and 84.2\% for an heritability of 50\% and between 93.2\% and 94.1\% for an heritability of 80\%.
-Given all these estimates of the maximal achievable AUC and for the sake of simplicity, we report maximum AUCs of 84\% (94\%) for heritabilities of 50\% (80\%) whatever are the parameters of the simulation.
+Given all these estimates of the maximal achievable AUC and for the sake of simplicity, we report maximum AUCs of 84\% (94\%) for heritabilities of 50\% (80\%) whatever are the parameters of the simulations.
 
@@
 \caption{Comparison of models in scenario \textnumero1 for the ``simple'' model.
-Mean of AUC over 100 simulations for the maximum values of PRSs for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the number and location of causal SNPs.
+Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the number and location of causal SNPs.
 Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
@@
 Thinner lines represents results in scenario \textnumero1 (Figure \ref{fig:supp-AUC-all-r2}).
-Mean of AUC over 100 simulations for the maximum values of PRSs for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the number and location of causal SNPs.
+Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the number and location of causal SNPs.
 Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).

### 2018-09-18 paper/draft.tex
save

@@
 
-In this paper, we present an efficient method to jointly estimate SNP effects, allowing for practical application of penalized logistic regression on modern datasets including hundreds of thousands of individuals.
-Moreover, our implementation of penalized logistic regression directly includes automatic choices for hyper-parameters.
+In this paper, we present an efficient method to jointly estimate SNP effects, allowing for practical application of penalized logistic regression (PLR) on modern datasets including hundreds of thousands of individuals.
+Moreover, our implementation of PLR directly includes automatic choices for hyper-parameters.
 The choice of hyper-parameters for a predictive model is very important since it can dramatically impact its predictive performance.
@@
 
-We compare the performance of penalized logistic regression to the C+T method and to a derivation of random forests. 
-Penalized logistic regression consistently achieves higher predictive performance than the two other methods while being very fast.
-We find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; AUC values increase from 83\% with the best prediction of C+T to 92.5\% with penalized logistic regression.
-We confirm these results in a data analysis of a case-control study for celiac disease where penalized logistic regression and the standard C+T method achieve AUC of 89\% and of 82.5\%.
+We compare the performance of PLR, C+T and a derivation of random forests using both real and simulated data. 
+PLR consistently achieves higher predictive performance than the two other methods while being as fast as C+T. 
+We find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; for instance, AUC values increase from 83\% with the best prediction of C+T to 92.5\% with PLR.
+We confirm these results in a data analysis of a case-control study for celiac disease where PLR and the standard C+T method achieve AUC of 89\% and of 82.5\%.
 
@@
 Several methods have been developed to predict disease status, or more generally any phenotype, based on SNP information.
-A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or even just PRS -- is used to derive PRS from results of Genome-Wide Association Studies (GWAS) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
+A commonly used method often called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') is used to derive PRS from results of Genome-Wide Association Studies (GWAS) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
 This technique uses GWAS summary statistics only, allowing for fast implementation.
 However, the ``C+T'' approach also has several limitations.
-Previous studies have shown that predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
+Previous studies have shown that predictive performance of C+T is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
 Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}.
@@
 Package bigstatsr now includes an efficient algorithm with a new implementation for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
-In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression against the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
+In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression (PLR) against the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
 In this comparison, we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
-For the C+T model, we report results for a large grid of hyper-parameters. 
-For the penalized logistic regression, the choice of hyper-parameters is included in the algorithm so that we report only one model for each simulation.
-We also use this penalized logistic regression on a feature-augmented dataset in order to capture not only linear effects, but also recessive and dominant effects.
+For C+T, we report results for a large grid of hyper-parameters. 
+For PLR, the choice of hyper-parameters is included in the algorithm so that we report only one model for each simulation.
+We also use a modified version of PLR in order to capture not only linear effects, but also recessive and dominant effects.
 
-To perform simulations, we use real genotype data and simulate new phenotypes \cite[]{zhou2013polygenic}. 
-In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the heritability.
+To perform simulations, we use real genotype data and simulate new phenotypes. 
+In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the disease heritability.
 We also compare different models for simulating phenotypes, one with only linear effects, and one that combines linear, dominant and interaction-type effects.
-Overall, we find that the penalized logistic regression consistently achieves higher predictive performance than the C+T and T-Trees methods while being very fast.
+Overall, we find that the penalized logistic regression consistently achieves higher predictive performance than the C+T and T-Trees methods while being as fast as C+T.
 This demonstrates the feasibility and relevance of this approach for PRS computation on large modern datasets.
@@
 Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
-For simulations presented later, we first restrict this dataset to controls in order to remove the genetic structure induced by the celiac disease status.
-Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.
-In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on a robust Mahalanobis distances computed using the first 10 principal components of the remaining individuals.
-This 3-step filtering process results in a sample of 7100 individuals with minimal population structure (see supplementary notebook ``preprocessing'').
-We also use this dataset for real data application, in this case keeping all 15,155 individuals (4496 cases and 10,659 controls).
-Both datasets contain 281,122 SNPs.
+For simulations presented later, we first restrict this dataset to controls from UK in order to remove the genetic structure induced by the celiac disease status and population structure.
+This filtering process results in a sample of 7100 individuals (see supplementary notebook ``preprocessing'').
+We also use this dataset for real data application, in this case keeping all 15,155 individuals (4496 cases and 10,659 controls). Both datasets contain 281,122 SNPs.
 
@@
 We vary parameters of the simulations in order to match a range of genetic architecture from low to high polygenicity.
-This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 chromosomes or 30 in the HLA region of chromosome 6), and the heritability (50\% or 80\%).
-In a second phase, in order to further increase the proportion of causal variants, we restrict the dataset to chromosome 6 only (18,941 SNPs) instead of using all 22 automosal chromosomes (281,122 SNPs).
-We also consider deviation from the standard normal additive model, drawing effects of causal SNPs from either a Normal or from a Laplace distribution, and computing liability scores either from a ``simple'' model with linear effects only or a ``fancy'' model that combines linear, dominant and interaction-type effects.
-For the ``simple'' model, we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$ and $S_\text{causal}$ is the set of causal SNPs.
-For the ``fancy'' model, we simulate phenotypes using linear, dominant and interaction-type effects (see Supplementary Materials).
+This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 chromosomes or 30 in the HLA region of chromosome 6), and the disease heritability (50\% or 80\%).
+%SNP effects are drawn from either a Gaussian or from a Laplace distribution.
+Liability scores are computed either from an additive model with linear effects only (``ADD'') or a more complex model that combines linear, dominant and interaction-type effects (``COMP'').
+For model ``ADD'', we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$ and $S_\text{causal}$ is the set of causal SNPs.
+For model ``COMP'', we simulate liability scores using linear, dominant and interaction-type effects (see Supplementary Materials).
 
 We implement 3 different simulation scenarios, summarized in table \ref{tab:simus}.
-Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes) and a training set of size 6000.
+Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes -- 281,122 SNPs) and a training set of size 6000.
 It compares all methods described in section \ref{sec:methods}.
-For each combination of the remaining parameters, results are based on 100 simulations excepted in the first simulation comparing penalized logistic regression with T-Trees; this simulation relies on 5 simulations only because of a much higher computational burden (several hours of computation for a single simulation) of T-Trees as compared to other approaches.
-Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only.
-Reducing the number of SNPs aims at increasing the polygenicity (the proportion of causal SNPs) of the simulated models and at virtually increasing the sample size \cite[]{dudbridge2013power,marquez2017multiethnic,vilhjalmsson2015modeling}.
-For this scenario, we use the additive (``simple'') model only, but continue to compare all previous different values of the other parameters.
-Finally, scenario \textnumero3 reuses the whole dataset but this time varying the size of the training set in order to assess how the sample size affects predictive performance of methods.
-A total of 100 simulations per combination of parameters are run using 300 causal SNPs randomly chosen anywhere on the genome.
+For each combination of the remaining parameters, results are based on 100 simulations excepted when comparing PLR with T-Trees, which relies on 5 simulations only because of a much higher computational burden of T-Trees as compared to other methods.
+Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only (18,941 SNPs).
+Reducing the number of SNPs increases the polygenicity (i.e.
+the proportion of causal SNPs) of the simulated models.
+Reducing the number of SNPs ($p$) is also equivalent to increasing the sample size ($n$) as predictive power is dependent on $n/p$ \cite[]{dudbridge2013power,vilhjalmsson2015modeling}.
+For this scenario, we use the additive model only, but continue to vary all other simulation parameters.
+Finally, scenario \textnumero3 uses the whole dataset as in scenario \textnumero1 while varying the size of the training set in order to assess how the sample size affects predictive performance of methods.
+A total of 100 simulations per combination of parameters are run using 300 causal SNPs randomly chosen on the genome.
 
@@
 This measure is similar to the AUC, but focuses on high specificities, which is the most useful part of the ROC curve in clinical settings.
-When reporting AUC results of simulations, we use estimates of maximum achievable AUC of 84\% and 94\% for heritabilities of respectively 50\% and 80\%.
+When reporting AUC results of simulations, we also report maximum achievable AUC values of 84\% and 94\% for heritabilities of respectively 50\% and 80\%.
 These estimates are based on three different yet consistent estimations (see Supplementary Materials).
@@
 
-To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in R package bigstatsr.
+To fit this penalized logistic regression, we use an efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in R package bigstatsr.
 This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''.
@@
 
-In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects: one variable is coded 1 if homozygous variant and 0 otherwise, and the other is coded 0 for homozygous referent and 1 otherwise.
-This results in a dataset with 3 times as many variables as the initial one, on which we can apply penalized logistic regression.
-We refer to this method ``logit-triple'' in the results.
+To capture recessive and dominant effects on top of additive effects in PLR, we use simple feature engineering: we construct a separate dataset with 3 times as many variables as the initial one. 
+For each SNP variable, we add two more variables coding for recessive and dominant effects: one variable is coded 1 if homozygous variant and 0 otherwise, and the other is coded 0 for homozygous referent and 1 otherwise.
+We refer to this method as ``PLR3'' in the results.
 
@@
 
-We use Monte Carlo cross-validation to compute AUC, partial AUC, the number of predictors and execution time for the original Celiac dataset with real phenotypes: we randomly split 100 times the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals.
+We use Monte Carlo cross-validation to compute AUC, partial AUC, the number of predictors and execution time for the original Celiac dataset with the observed case-control status: we randomly split 100 times the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals.
 
@@
 All the code used in this paper along with results such as figures and tables, are available as HTML R notebooks in the Supplementary Materials.
+TODO: déplacer dans code availbility après ackn + ajouter les results + le tuto
 
@@
 
-When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-PRS}).
+When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-PRS}).
 Moreover, penalized logistic regression consistently provides higher predictive performance than the C+T method across all scenarios we considered, excepted in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).
@@
 Method ``logit-simple'' provides particularly higher predictive performance than ``PRS-max'' when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region.
-In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' approach and 84\% with ``PRS-max'', while the maximum achievable AUC is 94\% (Figure \ref{fig:main-AUC-logit}).
+In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' approach and 84\% with ``PRS-max'' (Figure \ref{fig:main-AUC-logit}).
 
@@
 Concerning the $r^2$ threshold of the clumping step in C+T, we mostly used the common value of $0.2$.
-Yet, using a more stringent value of $0.05$ provides higher predictive performance than using $0.2$ in most of the cases tested in this paper (Figures \ref{fig:supp-AUC-all-r2}, \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-chr6-all-r2})
+Yet, using a more stringent value of $0.05$ provides higher predictive performance than using $0.2$ in most of the cases we considered (Figures \ref{fig:supp-AUC-all-r2}, \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-chr6-all-r2})
 
-Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of the implementation of R package biglasso \cite[]{zeng2017efficient}, only slightly better for biglasso, which is likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).
+Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance achieved ...
+of the implementation of R package biglasso \cite[]{zeng2017efficient}, only slightly better for biglasso, which is likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).
+TODO: juste dire qu'on est quasiment le max, sans mentionner biglasso. SIMILAR
 
@@
 First, when reproducing simulations of scenario \textnumero1 using chromosome 6 only (scenario \textnumero2), the predictive performance of ``logit-simple'' always increase (Figure \ref{fig:supp-AUC-chr6-all-r2}).
-There is particularly a large increase when simulating 3000 causal SNPs: AUC from the ``logit-simple'' increases from 60\% to nearly 80\% for Gaussian effects and a heritability of 80\%.
+There is particularly a large increase when simulating 3000 causal SNPs: AUC from the ``logit-simple'' increases from 60\% to nearly 80\% for Gaussian effects and a disease heritability of 80\%.
 On the contrary, when simulating only 30 or 300 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' does not increase, and even decreases for an heritability of 80\% (Figure \ref{fig:supp-AUC-chr6-all-r2}).
@@
 Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360 (Table \ref{tab:results-celiac}, figures \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
-Note that for the C+T method, we still report the best result among 102 p-value thresholds.
-In terms of computation time, we report only the GWAS computation for the C+T method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once and testing different hyper-parameter values, is almost as fast as the C+T method (190 vs 130 seconds), and the ``logit-triple'' takes less than twice as long as the ``logit-simple'' (296 vs 190 seconds).
+%Note that for the C+T method, we still report the best result among 102 p-value thresholds.
+In terms of computation time, we show that the ``logit-simple'' method, while learning jointly on all SNPs at once and testing different hyper-parameter values, is almost as fast as the C+T method (190 vs 130 seconds), and the ``logit-triple'' takes less than twice as long as the ``logit-simple'' (296 vs 190 seconds).
 
@@
 In this comparative study, we present a computationally efficient implementation of penalized logistic regression. 
-This model can be used to build polygenic risk scores based on very large SNP datasets such as the UK biobank \cite[]{bycroft2017genome}. 
+This model can be used to build polygenic risk scores based on very large individual-level SNP datasets such as the UK biobank \cite[]{bycroft2017genome}. 
 In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently.
@@
 The choice of the $r^2$ threshold of the clumping step is important. 
-Indeed, on the one hand, choosing a low value for this threshold may discard independently predictive SNPs that are in Linkage Desiquilibrium; yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would bias SNP effects.
+Indeed, on the one hand, choosing a low value for this threshold may discard informative SNPs that are correlated.
+Yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would add some noise to the PRS.
 Based on the simulations, we find that using a stringent threshold ($r^2 = 0.05$) leads to higher predictive performance, even when causal SNPs are correlated.
-It means that accurately estimating SNP effects is more important than including all causal SNPs.
-Moreover, in this paper, we reported the maximum AUC of 102 different p-value thresholds, a threshold that should normally be learned on the training set only.
-The choice of this threshold is very important as it can greatly impact the predictive performance of the C+T method, which we confirm in this study \cite[]{ware2017heterogeneity}.
+It means that, in most cases, avoiding redundant information is more important than including all causal SNPs.
+The choice of the $p_T$ threshold is also very important as it can greatly impact the predictive performance of the C+T method, which we confirm in this study \cite[]{ware2017heterogeneity}. 
+In this paper, we reported the maximum AUC of 102 different p-value thresholds, a threshold that should normally be learned on the training set only.
+To our knowledge, there are no clear standard on how to choose these two critical hyper-parameters.
 
@@
 Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}.
-In next study, we will assess how can we use models and effects learned in one population to improve learning and prediction in another population.
+In future work, we aim at extending our method to overcome those issues.
 %We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}.
 Finally, we also wish to assess how can we combine the information provided by genetic data with clinical and environmental data, possibly in a non-linear way.
+
+\subsection{Conclusion}
+
+TODO
 
@@
 Note that this approximation is reported to be less accurate for high heritabilities.
-Secondly, if we assume that the genetic part of the liabilities follows a Gaussian distribution $N(0, h^2)$ and that the environmental part follows a Gaussian distribution $N(0, 1 - h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the heritability $h^2$ through Monte Carlo simulations.
+Secondly, if we assume that the genetic part of the liabilities follows a Gaussian distribution $N(0, h^2)$ and that the environmental part follows a Gaussian distribution $N(0, 1 - h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the disease heritability $h^2$ through Monte Carlo simulations.
 We report AUCs of 84.1\% and 94.1\% for an heritability of 50\% and 80\%, respectively.
@@
 There is a Spearman correlation of 98\% between values of AUC and partial AUC.
-The relation between the two values are the same whatever are the heritability, distribution of effects and method used.\label{fig:supp-AUC-corr}}
+The relation between the two values are the same whatever are the disease heritability, distribution of effects and method used.\label{fig:supp-AUC-corr}}
 \end{figure}

### 2018-09-21 4-results-simus.Rmd
polish SupMat
begin renaming

@@
 library(pkg.paper.PRS)  ## See code source for functions' code.
-options(nboot = 1e4)
+options(nboot = 1e3)
 ```
@@
   c(COLORS, COLORS[c(4, 2, 5)], "black"), 
-  c("logit-simple", "PRS-max", "PRS-stringent", "T-Trees", "logit-triple", "PRS-all",
-    "PRS-max-0.05", "PRS-max-0.2", "PRS-max-0.8", "biglasso")
+  c("PLR", "C+T-max", "C+T-stringent", "T-Trees", "PLR3", "C+T-all",
+    "C+T-max-0.05", "C+T-max-0.2", "C+T-max-0.8", "biglasso")
 ) 

### 2018-09-21 pkg.paper.PRS/R/read-format-results.R

@@ -58,6 +58,30 @@ add_AUC <- function(x, pAUC) {
 
 ################################################################################
 
+sub_list <- function(x, from, to) {
+  for (i in seq_along(from)) {
+    x <- sub(from[[i]], to[[i]], x)
+  }
+  x
+}
+
+rename_lvl <- function(x) {
+
+  methods  <- c("PRS-all", "PRS-stringent", "PRS-max", "logit-simple", "logit-triple")
+  methods2 <- c("C+T-all", "C+T-stringent", "C+T-max", "PLR",          "PLR3")
+
+  dplyr::mutate(
+    x,
+    par.causal = factor(purrr::map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
+                        levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
+    par.dist = stringr::str_to_title(par.dist),
+    par.model = sub_list(par.model, c("simple", "fancy"), c("ADD", "COMP")),
+    method = sub_list(method, methods, methods2)
+  )
+}
+
+################################################################################
+
 #' Read & format results
 #'
 #' @param files Files to read from.
@@ -74,9 +98,7 @@ read_format_results <- function(files, corr, ncores = nb_cores()) {
     files[ind] %>%
       purrr::map_dfr(~readRDS(.x)) %>%
       tibble::as_tibble() %>%
-      dplyr::mutate(
-        par.causal = factor(purrr::map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
-                            levels = c("30 in HLA", paste(3 * 10^(1:3), "in all")))) %>%
+      rename_lvl() %>%
       add_AUC(pAUC) %>%
       add_sens_FDP(corr) %>%
       dplyr::select(-c(num.simu, true_set, set, pheno, pred))

### 2018-09-21 paper/draft.tex

@@
 Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
-PRS might have a major public health impact, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.
+PRS might have a major impact on public health, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.
 The ``Clumping+Thresholding'' (C+T) approach, which is the most common method to derive PRS, uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it fast and easy to use.
@@
 
-In conclusion, our study demonstrates that penalized logistic regression is applicable to large-scale individual-level data and can achieve more discriminative polygenic risk scores.
-Our implementation is publicly available in R package bigstatsr.\\
+In conclusion, our study demonstrates that penalized logistic regression can achieve more discriminative polygenic risk scores, while being applicable to large-scale individual-level data thanks to the implementation we provide in the R package bigstatsr.\\
 
-\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
-\textbf{Supplementary information:}
+\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}
 }
@@
 \newpage
+
 \section{Introduction}
@@
 A commonly used method often called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') is used to derive PRS from results of Genome-Wide Association Studies (GWAS) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
-This technique uses GWAS summary statistics only, allowing for fast implementation.
-However, the ``C+T'' approach also has several limitations.
-Previous studies have shown that predictive performance of C+T is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
+This technique uses GWAS summary statistics only, allowing for a fast implementation of C+T.
+However, C+T also has several limitations; for instance, previous studies have shown that predictive performance of C+T is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
 Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}.
 Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes \cite[]{abraham2013performance}.
-Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
+Moreover, these methods and their derivatives are often computationally very demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
 Finally, statistical learning methods have also been used to derive PRS for complex human diseases by jointly estimating SNP effects.
@@
 In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the disease heritability.
-We also compare different models for simulating phenotypes, one with only linear effects, and one that combines linear, dominant and interaction-type effects.
-Overall, we find that the penalized logistic regression consistently achieves higher predictive performance than the C+T and T-Trees methods while being as fast as C+T.
+We also compare two different models for simulating phenotypes, one with additive effects only, and one that combines additive, dominant and interaction-type effects.
+Overall, we find that PLR consistently achieves higher predictive performance than the C+T and T-Trees methods while being as fast as C+T.
 This demonstrates the feasibility and relevance of this approach for PRS computation on large modern datasets.
@@
 
-We use real genotypes of European individuals from a case-control celiac disease study \cite[]{dubois2010multiple}.
+We use real genotypes of European individuals from a case-control study for celiac disease \cite[]{dubois2010multiple}.
 The composition of this dataset is presented in table \ref{tab:celiac-data}.
@@
 We vary parameters of the simulations in order to match a range of genetic architecture from low to high polygenicity.
-This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 chromosomes or 30 in the HLA region of chromosome 6), and the disease heritability (50\% or 80\%).
+This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 autosomal chromosomes or 30 in the HLA region of chromosome 6), and the disease heritability (50\% or 80\%).
 %SNP effects are drawn from either a Gaussian or from a Laplace distribution.
-Liability scores are computed either from an additive model with linear effects only (``ADD'') or a more complex model that combines linear, dominant and interaction-type effects (``COMP'').
+Liability scores are computed either from an additive model with linear effects only (``ADD'') or a more complex model that combines additive, dominant and interaction-type effects (``COMP'').
 For model ``ADD'', we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$ and $S_\text{causal}$ is the set of causal SNPs.
-For model ``COMP'', we simulate liability scores using linear, dominant and interaction-type effects (see Supplementary Materials).
+For model ``COMP'', we simulate liability scores using additive, dominant and interaction-type effects (see Supplementary Materials).
 
@@
 
-In this study, we compare three different types of methods: the C+T method, T-Trees and penalized logistic regression.
+In this study, we compare three different types of methods: the C+T method, T-Trees and penalized logistic regression (PLR).
 
@@
 The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another (this set of SNPs is denoted $S_\text{clumping}$).
-Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $p_T$ to be determined.
-Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS.
+Then, thresholding (T) consists in removing SNPs with p-values larger than a user-defined threshold $p_T$.
+Finally, the PRS for individual $i$ is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS.
 In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$.
-We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``PRS-all''), one including only SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``PRS-stringent''), and one that maximizes the AUC (``PRS-max'') for these two thresholds ($0$ and $5 \cdot 10^{-8}$) and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
-As we report the optimal threshold based on the test set, the AUC for ``PRS-max'' is an upper bound of the AUC for the C+T method.
+We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``C+T-all''), one including only SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``C+T-stringent''), and one that maximizes the AUC (``C+T-max'') for 102 thresholds between $1$ and $10^{-100}$ (Table \ref{tab:thr}).
+As we report the optimal threshold based on the test set, the AUC for ``C+T-max'' is an upper bound of the AUC for the C+T method.
 
 T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}.
-We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees instead of 1000 because using 1000 trees provides a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown).
+We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees instead of 1000.
+Using 1000 trees provides a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown).
 %We call this method ``T-Trees'' in the results.
 
-Finally, for the penalized logistic regression, we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
+Finally, for the penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
 Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
 the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
-the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
+the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $p \gg n$ situation ($p$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
 In this study, we use an embedded grid search over $\alpha \in \{1, 0.5, 0.05, 0.001\}$.
@@
 To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
-Because of L1-regularization, the resulting vectors of coefficients are sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts.
-We refer to this method as ``logit-simple'' in the results section.
+Because of L1-regularization, the resulting vectors of coefficients are sparse and can be used to make a PRS based on a \emph{linear} combination of allele counts.
+We refer to this method as ``PLR'' in the results section.
 
@@
 
-\subsection{Reproduciblity}
-
-All the code used in this paper along with results such as figures and tables, are available as HTML R notebooks in the Supplementary Materials.
-TODO: déplacer dans code availbility après ackn + ajouter les results + le tuto
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We compared penalized logistic regression (``logit-simple'') with the C+T method (``PRS'') using whole-genome simulations of scenario \textnumero1 (Table \ref{tab:simus}).
+We compared penalized logistic regression (PLR) with the C+T method using whole-genome simulations of scenario \textnumero1 (Table \ref{tab:simus}).
 
-When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-PRS}).
-Moreover, penalized logistic regression consistently provides higher predictive performance than the C+T method across all scenarios we considered, excepted in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).
+When simulating a model with 30 causal SNPs and an heritability of 80\%, PLR provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T range between 83\% and 90\% (Figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-PRS}).
+Moreover, PLR consistently provides higher predictive performance than C+T across all scenarios we considered, excepted in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).
 
-Method ``logit-simple'' provides particularly higher predictive performance than ``PRS-max'' when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region.
-In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' approach and 84\% with ``PRS-max'' (Figure \ref{fig:main-AUC-logit}).
+PLR provides particularly higher predictive performance than C+T when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region.
+In this situation, the mean AUC reaches 92.5\% for PLR and 84\% for ``C+T-max'' (Figure \ref{fig:main-AUC-logit}).
 
-Note that for the simulations we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).
+Note that, for the simulations, we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).
 
@@
 In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can dramatically impact the predictive performance of C+T.
-For example, in a model with only 30 causal SNPs, AUC ranges from less than 60\% when using all SNPs passing clumping to 90\% if choosing the optimal p-value threshold (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
+For example, in a model with only 30 causal SNPs, AUC ranges from less than 60\% when using all SNPs passing clumping to 90\% \emph{if} choosing the optimal p-value threshold (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
 
@@
 
-Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance achieved ...
-of the implementation of R package biglasso \cite[]{zeng2017efficient}, only slightly better for biglasso, which is likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).
-TODO: juste dire qu'on est quasiment le max, sans mentionner biglasso. SIMILAR
+Our implementation of PLR that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of 100 models corresponding to different values of $\lambda$ (Figure \ref{fig:supp-biglasso}).
 
@@
 We tested the T-Trees method in scenario \textnumero1.
-As compared to ``logit-simple'', T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}).
-Even when simulating a ``fancy'' model in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower when using T-Trees than when using ``logit-simple'' (Figure \ref{fig:supp-ttrees}).
+As compared to PLR, T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}).
+Even when simulating with model ``COMP'' in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower when using T-Trees than when using PLR (Figure \ref{fig:supp-ttrees}).
 
-We also compared the two penalized logistic regressions in scenario \textnumero1, ``logit-simple'' and ``logit-triple'' that uses additional features (variables) coding for recessive and dominant effects.
-Predictive performance of ``logit-triple'' are nearly as good as ``logit-simple'' when there are only linear effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
-For the ``fancy model'', ``logit-triple'' provides AUC values at least 3.5\% higher than ``logit-simple'', excepted when there are 3000 causal SNPs.
-Yet, the ``triple'' solution takes 2-3 times as much time to run and requires 3 times as much disk storage as the ``simple'' solution.
+We also compared the two penalized logistic regressions in scenario \textnumero1, PLR and PLR3 that uses additional features (variables) coding for recessive and dominant effects.
+Predictive performance of PLR3 are nearly as good as PLR when there are only addtive effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
+For model ``COMP'', PLR3 provides AUC values at least 3.5\% higher than PLR, excepted when there are 3000 causal SNPs.
+Yet, PLR3 takes 2-3 times as much time to run and requires 3 times as much disk storage as PLR.
 
@@
 
-First, when reproducing simulations of scenario \textnumero1 using chromosome 6 only (scenario \textnumero2), the predictive performance of ``logit-simple'' always increase (Figure \ref{fig:supp-AUC-chr6-all-r2}).
-There is particularly a large increase when simulating 3000 causal SNPs: AUC from the ``logit-simple'' increases from 60\% to nearly 80\% for Gaussian effects and a disease heritability of 80\%.
-On the contrary, when simulating only 30 or 300 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' does not increase, and even decreases for an heritability of 80\% (Figure \ref{fig:supp-AUC-chr6-all-r2}).
-Secondly, when varying the training size (scenario \textnumero3), we report an increase of AUC when increasing the training size, with a faster increase of AUC provided by ``logit-simple'' as compared to ``PRS-max'' (Figure  \ref{fig:main-AUC-ntrain}).
+First, when reproducing simulations of scenario \textnumero1 using chromosome 6 only (scenario \textnumero2), the predictive performance of PLR always increase (Figure \ref{fig:supp-AUC-chr6-all-r2}).
+There is particularly a large increase when simulating 3000 causal SNPs: AUC from PLR increases from 60\% to nearly 80\% for Gaussian effects and a disease heritability of 80\%.
+On the contrary, when simulating only 30 or 300 causal SNPs with the corresponding dataset, AUC of ``C+T-max'' does not increase, and even decreases for an heritability of 80\% (Figure \ref{fig:supp-AUC-chr6-all-r2}).
+Secondly, when varying the training size (scenario \textnumero3), we report an increase of AUC with a larger training size, with a faster increase of AUC provided with PLR as compared to ``C+T-max'' (Figure  \ref{fig:main-AUC-ntrain}).
 
@@
 
-Joint logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with ``logit-simple'' and 89.1\% with ``logit-triple'' as compared to 82.5\% with the C+T method.
-The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with ``logit-simple'', ``logit-triple'' and the C+T method, respectively.
-Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360 (Table \ref{tab:results-celiac}, figures \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
-%Note that for the C+T method, we still report the best result among 102 p-value thresholds.
-In terms of computation time, we show that the ``logit-simple'' method, while learning jointly on all SNPs at once and testing different hyper-parameter values, is almost as fast as the C+T method (190 vs 130 seconds), and the ``logit-triple'' takes less than twice as long as the ``logit-simple'' (296 vs 190 seconds).
+Joint logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with PLR and 89.1\% with PLR3 as compared to 82.5\% with ``C+T-max''.
+The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with PLR, PLR3 and ``C+T-max'', respectively.
+Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360 (Table \ref{tab:results-celiac}, figure \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
+In terms of computation time, we show that PLR, while learning jointly on all SNPs at once and testing four different hyper-parameter values, is almost as fast as the C+T method (190 vs 130 seconds), and PLR3 takes less than twice as long as PLR (296 vs 190 seconds).
 
@@
   \hline
-PRS-max & 0.825 (0.000664) & 0.0289 (0.000187) & 8360 (744) & 130 (0.143) \\ 
-logit-simple & 0.887 (0.00061) & 0.0411 (0.000224) & 1570 (46.4) & 190 (1.21) \\ 
-logit-triple & 0.891 (0.000628) & 0.0426 (0.000219) & 2260 (56.1) & 296 (2.03) \\ 
+C+T-max & 0.825 (0.000664) & 0.0289 (0.000187) & 8360 (744) & 130 (0.143) \\ 
+PLR & 0.887 (0.00061) & 0.0411 (0.000224) & 1570 (46.4) & 190 (1.21) \\ 
+PLR3 & 0.891 (0.000628) & 0.0426 (0.000219) & 2260 (56.1) & 296 (2.03) \\ 
    \hline
@@
 
-In this comparative study, we present a computationally efficient implementation of penalized logistic regression. 
+In this comparative study, we present a computationally efficient implementation of penalized logistic regression (PLR). 
 This model can be used to build polygenic risk scores based on very large individual-level SNP datasets such as the UK biobank \cite[]{bycroft2017genome}. 
 In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently.
-Penalized logistic regression nearly always outperform the C+T method, and the benefits of using it are more pronounced with an increasing sample size or when causal SNPs are correlated with one another.
+PLR nearly always outperform the C+T method, and the benefits of using PLR are more pronounced with an increasing sample size or when causal SNPs are correlated with one another.
 
@@
 In this paper, we reported the maximum AUC of 102 different p-value thresholds, a threshold that should normally be learned on the training set only.
-To our knowledge, there are no clear standard on how to choose these two critical hyper-parameters.
+To our knowledge, there is no clear standard on how to choose these two critical hyper-parameters.
 
 On the contrary, in the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$ that accounts for the amount of regularization used in the model.
-Not only this procedure provides near-optimal results (as compared to the best prediction when using R package biglasso), but it also accelerates the training of the model thanks to the development of an early stopping criterion.
+Not only this procedure provides near-optimal results, but it also accelerates the model training thanks to the development of an early stopping criterion.
 Usually, cross-validation is used to choose hyper-parameter values and then the model is trained again with these particular hyper-parameter values \cite[]{Hastie2008,wei2013large}.
@@
 %Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model.
-Yet, if using e.g.
-$\alpha = 0.001$, it would lead to a similar model.
+Yet, if using e.g.\ $\alpha = 0.001$, it would lead to a similar model.
 
@@
 
-In this paper, we also explored how to capture non-linear effects.
-For this, we introduced a simple feature engineering technique that enables logistic regression to detect and learn not only additive effects, but also dominant and recessive effects.
-This technique improves the predictive performance of logistic regression when there are some non-linear effects in the simulations, while providing nearly the same predictive performance when there are only linear effects.
+We also explored how to capture non-linear effects.
+For this, we introduced a simple feature engineering technique that enables PLR to detect and learn not only additive effects, but also dominant and recessive effects.
+This technique improves the predictive performance of PLR when there are some non-linear effects in the simulations, while providing nearly the same predictive performance when there are only additive effects.
 Moreover, it also improves predictive performance for the celiac disease.
@@
 
-Our approach has one major limitation: the main advantage of the C+T method is that it is applicable directly to summary statistics, allowing to leverage the largest GWAS sample size to date, even when individual cohort data cannot be merged because of practical and ethical reasons (e.g.
-consortium data including many cohorts).
+Our approach has one major limitation: the main advantage of the C+T method is its direct applicability to summary statistics, allowing to leverage the largest GWAS results to date, even when individual cohort data cannot be merged because of practical or ethical reasons (e.g.\ consortium data including many cohorts).
 As of today, the proposed penalized logistic regression does not allow for the analysis of summary data, but this represents an important future direction of our work.
-The current version is of particular interest for the analysis of modern SNP dataset including hundreds of thousands of individuals.
+The current version is of particular interest for the analysis of modern SNP datasets including hundreds of thousands of individuals.
 
 Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}.
-In future work, we aim at extending our method to overcome those issues.
+%In future work, we aim at exploring how we can overcome those issues.
 %We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}.
@@
 
-TODO
+In this comparative study, we have presented a computationally efficient implementation of penalized logistic regression that can be used to predict disease status based on genotypes.
+Note that a similar penalized linear regression is also available in our software.
+Usually, joint statistical learning methods are so computationally demanding that they cannot be used on large-scale datasets such as the UK biobank \cite[]{bycroft2017genome}. 
+A partial solution to this problem is to pre-filter data, but it usually leads to a decrease in predictive performance as some information is removed \cite[]{accuheight2018}.
+This pre-filtering step is not necessary anymore thanks to our implementation.
+Moreover, the model we present in this paper can flexibly adapt to traits with different architectures thanks to two hyper-parameters that are automatically learned with the model.
+Thus, we strongly believe that our implementation of penalized regressions would be appropriate for predicting autoimmune diseases with a few strong effects (e.g.\ celiac disease) as well as highly polygenic traits (e.g.\ standing height).
+These models could also be used to predict phenotypes based on other omics data since the implementation is not specific to genotype data.
 
@@
 \hline
-\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{simple} & PRS \\
-& & & 30 in all & & & & logit-simple \\
-& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{fancy} & logit-triple \\
+\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{ADD} & PRS \\
+& & & 30 in all & & & & PLR \\
+& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{COMP} & PLR3 \\
 & & & 3000 in all & & & & (T-Trees) \\
 \hline
-\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{simple} & PRS \\ 
-& & & & & & & logit-simple\\
+\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{ADD} & PRS \\ 
+& & & & & & & PLR\\
 \hline
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
-\caption{Main comparison of the C+T and ``logit-simple'' methods in scenario \textnumero1 for the ``simple'' model and an heritability of 80\%.
-Mean of AUC over 100 simulations for ``logit-simple'' and the maximum AUC reported with the C+T method (``PRS-max'').
+\caption{Main comparison of C+T and PLR in scenario \textnumero1 for model ``ADD'' and an heritability of 80\%.
+Mean of AUC over 100 simulations for PLR and the maximum AUC reported with ``C+T-max''.
 Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-PRS}}
-\caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for the ``simple'' model and an heritability of 80\%.
+\caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for model ``ADD'' and an heritability of 80\%.
 Mean of AUC over 100 simulations.
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
-\caption{Comparison of models when varying sample size in scenario \textnumero3 for the ``simple'' model with 300 causal SNPs sampled anywhere on the genome.
-Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the training size.
+\caption{Comparison of models when varying sample size in scenario \textnumero3 for model ``ADD'' with 300 causal SNPs sampled anywhere on the genome.
+Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the training size.
 Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
@@
 \centerline{\includegraphics[width=\textwidth]{celiac-roc}}
-\caption{ROC Curves for the ``C+T'', ``logit-simple'' and ``logit-triple'' methods for Celiac disease dataset.
+\caption{ROC Curves for C+T, PLR and PLR3 for the celiac disease dataset.
 Models were trained using 12,000 individuals.
@@
 We are also grateful to F\'elix Balazard for useful discussions about T-Trees, and to Yaohui Zeng for useful discussions about R package biglasso.
+
+\section*{Reproducibility}
+
+All the code used in this paper along with results such as figures and tables, are available as HTML R notebooks in the Supplementary Materials.
+Results are available at [TODO]. A tutorial on how to start with R packages bigstatsr and bigsnpr is available at \url{https://privefl.github.io/bigsnpr/articles/demo.html}.
 
@@
 
-For the ``fancy'' model, we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$
+For model ``COMP'', we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$
 and $S_\text{causal}^{(3)}$; $S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$.
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-logit}}
-\caption{Comparison of the C+T and ``logit-simple'' methods in scenario \textnumero1 for the ``simple'' model and an heritability of 50\%.
-Mean of AUC over 100 simulations for ``logit-simple'' and the maximum AUC reported with the C+T method (``PRS-max'').
+\caption{Comparison of C+T and PLR in scenario \textnumero1 for model ``ADD'' and an heritability of 50\%.
+Mean of AUC over 100 simulations for PLR and the maximum AUC reported with the C+T method (``C+T-max'').
 Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS}}
-\caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for the ``simple'' model and an heritability of 50\%.
+\caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for model ``ADD'' and an heritability of 50\%.
 Mean of AUC over 100 simulations.
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-all-r2}}
-\caption{Comparison of models in scenario \textnumero1 for the ``simple'' model.
-Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the number and location of causal SNPs.
+\caption{Comparison of models in scenario \textnumero1 for model ``ADD''.
+Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the number and location of causal SNPs.
 Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
-\caption{Comparison of models in scenario \textnumero2 (using chromosome 6 only) for the ``simple'' model.
+\caption{Comparison of models in scenario \textnumero2 (using chromosome 6 only) for model ``ADD''.
 Thinner lines represents results in scenario \textnumero1 (Figure \ref{fig:supp-AUC-all-r2}).
-Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the number and location of causal SNPs.
+Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the number and location of causal SNPs.
 Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
@@
 \centerline{\includegraphics[width=\textwidth]{supp-ttrees}}
-\caption{Comparison of T-Trees and ``logit-simple'' in scenario \textnumero1 for an heritability of 80\%.
+\caption{Comparison of T-Trees and PLR in scenario \textnumero1 for an heritability of 80\%.
 Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
@@
 \centerline{\includegraphics[width=\textwidth]{supp-triple}}
-\caption{Comparison of ``logit-triple'' and ``logit-simple'' in scenario \textnumero1 for an heritability of 80\%.
+\caption{Comparison of PLR3 and PLR in scenario \textnumero1 for an heritability of 80\%.
 Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-triple}}
-\caption{Comparison of ``logit-triple'' and ``logit-simple'' in scenario \textnumero1 for an heritability of 50\%.
+\caption{Comparison of PLR3 and PLR in scenario \textnumero1 for an heritability of 50\%.
 Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
@@
 \centerline{\includegraphics[width=\textwidth]{supp-biglasso}}
-\caption{Comparison of ``logit-triple'' and the best prediction (among 100 tested $\lambda$ values) for ``biglasso'' (another implementation of penalized logistic regression) in scenario \textnumero1.
-Simulations use the ``simple'' model, an heritability of 80\% and $\alpha = 1$.
+\caption{Comparison of PLR and the best prediction (among 100 tested $\lambda$ values) for ``biglasso'' (another implementation of penalized logistic regression -- \cite{zeng2017efficient}) in scenario \textnumero1.
+Simulations use model ``ADD'', an heritability of 80\% and $\alpha = 1$.
 Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.

### 2018-09-21 4-results-simus.Rmd
change names

@@
 author: "Florian Privé"
-date: "November 17, 2017"
+date: "September 21, 2018"
 output: 
@@
 library(pkg.paper.PRS)  ## See code source for functions' code.
-options(nboot = 1e3)
+options(nboot = 1e5)
 ```
@@
 ```{r}
-compare_logit <- function(results, with_method) {
+compare_PLR <- function(results, with_method) {
 
   results <- filter(
-    results, method %in% c("logit-simple", with_method), par.h2 == 0.8)
+    results, method %in% c("PLR", with_method), par.h2 == 0.8)
 
@@
 ```{r, out.width="90%", fig.width = 14}
-compare_logit(results1, with_method = "T-Trees")
+compare_PLR(results1, with_method = "T-Trees")
 
@@
 
-## Results with logit-triple & model fancy
+## Results with PLR3 & model COMP
 
@@
 
-### Results with logit-triple
+### Results with PLR3
 
 ```{r, out.width="90%", fig.width = 14}
-compare_logit(results2, with_method = "logit-triple")
+compare_PLR(results2, with_method = "PLR3")
 
@@
 results2 %>%
-  filter(grepl("logit", method)) %>%
+  filter(grepl("PLR", method)) %>%
   group_by_at(c(vars(starts_with("par")), "method")) %>%
@@
 
-### Results with model fancy
+### Results with model COMP
 
@@
 results2 %>%
-  filter(par.model == "fancy", method %in% c("logit-simple", "PRS-max")) %>%
+  filter(par.model == "COMP", method %in% c("PLR", "C+T-max")) %>%
   filter(thr.r2 == 0.2 | is.na(thr.r2)) %>%
@@
 results2 %>%
-  filter(par.model == "fancy", grepl("PRS", method)) %>%
+  filter(par.model == "COMP", grepl("C\\+T", method)) %>%
   filter(thr.r2 == 0.2 | is.na(thr.r2)) %>%
@@
 
-## Results with only logit-simple & PRSs
+## Results with only PLR & C+T
 
@@
 results2.all.r2 <- results2 %>%
-  filter(par.model == "simple", method %in% c("logit-simple", "PRS-max")) %>%
+  filter(par.model == "ADD", method %in% c("PLR", "C+T-max")) %>%
   mutate(method = c_method_r2(method, thr.r2)) %>%
@@
 results2 %>%
-  filter(par.model == "simple", method %in% c("logit-simple", "PRS-max")) %>%
+  filter(par.model == "ADD", method %in% c("PLR", "C+T-max")) %>%
   mutate(method = c_method_r2(method, thr.r2)) %>%
@@
   filter(thr.r2 == 0.2 | is.na(thr.r2)) %>%
-  filter(par.model == "simple", method != "logit-triple")
+  filter(par.model == "ADD", method != "PLR3")
 ```
@@
 results3 %>%
-  filter(method %in% c("logit-simple", "PRS-max")) %>%
+  filter(method %in% c("PLR", "C+T-max")) %>%
   mutate(Parameters  = interaction(par.h2, par.dist, method, sep = " | ")) %>%
@@
 results3.summary %>%
-  filter(method %in% c("logit-simple", "PRS-max"), par.model == "simple") %>%
+  filter(method %in% c("PLR", "C+T-max"), par.model == "ADD") %>%
   barplot_causal_one(h2 = 0.8)
@@
 results3.summary %>%
-  filter(grepl("PRS", method)) %>%
+  filter(grepl("C\\+T", method)) %>%
   barplot_causal_one(h2 = 0.8)
@@
 results3.summary %>%
-  filter(method %in% c("logit-simple", "PRS-max")) %>%
+  filter(method %in% c("PLR", "C+T-max")) %>%
   barplot_causal_one(h2 = 0.5)
@@
 results3.summary %>%
-  filter(grepl("PRS", method)) %>%
+  filter(grepl("C\\+T", method)) %>%
   barplot_causal_one(h2 = 0.5)
@@
 results4 %>%
-  filter(method %in% c("logit-simple", "PRS-max")) %>%
+  filter(method %in% c("PLR", "C+T-max")) %>%
   mutate(method = c_method_r2(method, thr.r2)) %>%
@@
 results6 <- results2 %>%
-  filter(par.model == "simple", par.causal == "300 in all") %>%
+  filter(par.model == "ADD", par.causal == "300 in all") %>%
   bind_rows(results5) %>%
-  filter(method %in% c("logit-simple", "PRS-max")) %>%
-  mutate(method = c_method_r2(method, thr.r2))
+  filter(method %in% c("PLR", "C+T-max")) %>%
+  mutate(method = c_method_r2(method, thr.r2),
+         par.causal = as.character(par.causal))
 ```
@@
   read_format_results(corr) %>%
-  compare_logit("biglasso")
+  compare_PLR("biglasso")
 

### 2018-09-21 5-results-celiac.Rmd
change names

@@
 library(pkg.paper.PRS)
+pAUC <- pkg.paper.PRS:::pAUC
 ```
@@
 
-### PRS
+### C+T
 
@@
 ```{r, cache=TRUE}
-# PRS
+# C+T
 thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
@@
 ```{r}
-pAUC <- pkg.paper.PRS:::pAUC
-
 plot_density_AUC <- function(pred, true = y[ind.test]) {
-  true_fct <- forcats::fct_recode(as.factor(true), "Control" = "0", "Case" = "1")
+  true_fct <- forcats::fct_recode(as.factor(true), Control = "0", Case = "1")
   data.frame(Score = pred, Status = true_fct) %>%
@@
 ```{r, cache=TRUE}
-# Train logit-simple
+# Train PLR
 system.time(
@@
 
-### Logistic regression with feature engineering
+### Logistic regression with feature engineering (PLR3)
 
@@
   d = y[ind.test], 
-  "PRS-max"      = prs[, which.max(aucs)],
-  "logit-simple" = preds1,
-  "logit-triple" = preds2
+  "C+T-max" = prs[, which.max(aucs)],
+  "PLR"     = preds1,
+  "PLR3"    = preds2
 ) %>%
@@
 methods.color <- setNames(scales::hue_pal()(6), 
-                          c("logit-simple", "PRS-max", "PRS-stringent",
-                            "T-Trees", "logit-triple", "PRS-all"))
+                          c("PLR", "C+T-max", "C+T-stringent",
+                            "T-Trees", "PLR3", "C+T-all"))
 
@@
   mutate(
+    method = fct_recode(sub("PRS", "C+T", method), 
+                        PLR = "logit-simple", PLR3 = "logit-triple"),
     AUC = map2_dbl(pred, pheno, ~bigstatsr::AUC(.x, .y)),

### 2018-09-24 paper/draft.tex
need results + read again
need results.zip

@@
 PRS might have a major impact on public health, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.
-The ``Clumping+Thresholding'' (C+T) approach, which is the most common method to derive PRS, uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it fast and easy to use.
+The ``Clumping+Thresholding'' (C+T) approach is the most common method to derive PRS.
+C+T uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it fast and easy to use.
 However, previous work showed that jointly estimating SNP effects for computing PRS has the potential to significantly improve the predictive performance of PRS as compared to C+T.
@@
 
-Finally, for the penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
+Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
 Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
 the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
-the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $p \gg n$ situation ($p$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
-In this study, we use an embedded grid search over $\alpha \in \{1, 0.5, 0.05, 0.001\}$.
+the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $p \gg n$ situation ($p$ is the number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
+In this study, we use an embedded grid search over $\alpha \in \{1,~0.5,~0.05,~0.001\}$.
 
@@
 For each SNP variable, we add two more variables coding for recessive and dominant effects: one variable is coded 1 if homozygous variant and 0 otherwise, and the other is coded 0 for homozygous referent and 1 otherwise.
-We refer to this method as ``PLR3'' in the results.
+We refer to this method as ``PLR3'' in the rest of the paper.
 
@@
 
-We compared penalized logistic regression (PLR) with the C+T method using whole-genome simulations of scenario \textnumero1 (Table \ref{tab:simus}).
+We compared penalized logistic regression (PLR) with the C+T method using simulations of scenario \textnumero1 (Table \ref{tab:simus}).
 
@@
 In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can dramatically impact the predictive performance of C+T. 
-For example, in a model with only 30 causal SNPs, AUC ranges from less than 60\% when using all SNPs passing clumping to 90\% \emph{if} choosing the optimal p-value threshold (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
+For example, in a model with 30 causal SNPs only, AUC ranges from less than 60\% when using all SNPs passing clumping to 90\% \emph{if} choosing the optimal p-value threshold (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
 
@@
 We tested the T-Trees method in scenario \textnumero1.
-As compared to PLR, T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}).
-Even when simulating with model ``COMP'' in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower when using T-Trees than when using PLR (Figure \ref{fig:supp-ttrees}).
+As compared to PLR, T-Trees perform worse in terms of predictive ability, while taking much longer to run (Figure \ref{fig:supp-ttrees}). 
+Even for simulations with model ``COMP'' in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower when using T-Trees than when using PLR (Figure \ref{fig:supp-ttrees}).
 
-We also compared the two penalized logistic regressions in scenario \textnumero1, PLR and PLR3 that uses additional features (variables) coding for recessive and dominant effects.
+We also compared the two penalized logistic regressions in scenario \textnumero1: PLR versus PLR3 that uses additional features (variables) coding for recessive and dominant effects.
 Predictive performance of PLR3 are nearly as good as PLR when there are only addtive effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
@@
 First, when reproducing simulations of scenario \textnumero1 using chromosome 6 only (scenario \textnumero2), the predictive performance of PLR always increase (Figure \ref{fig:supp-AUC-chr6-all-r2}).
-There is particularly a large increase when simulating 3000 causal SNPs: AUC from PLR increases from 60\% to nearly 80\% for Gaussian effects and a disease heritability of 80\%.
+There is a particularly large increase when simulating 3000 causal SNPs: AUC from PLR increases from 60\% to nearly 80\% for Gaussian effects and a disease heritability of 80\%.
 On the contrary, when simulating only 30 or 300 causal SNPs with the corresponding dataset, AUC of ``C+T-max'' does not increase, and even decreases for an heritability of 80\% (Figure \ref{fig:supp-AUC-chr6-all-r2}).
-Secondly, when varying the training size (scenario \textnumero3), we report an increase of AUC with a larger training size, with a faster increase of AUC provided with PLR as compared to ``C+T-max'' (Figure  \ref{fig:main-AUC-ntrain}).
+Secondly, when varying the training size (scenario \textnumero3), we report an increase of AUC with a larger training size, with a faster increase of AUC for PLR as compared to ``C+T-max'' (Figure  \ref{fig:main-AUC-ntrain}).
 
@@
 In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently.
-PLR nearly always outperform the C+T method, and the benefits of using PLR are more pronounced with an increasing sample size or when causal SNPs are correlated with one another.
+PLR always outperform the C+T method, excepted in some highly underpowered cases, and the benefits of using PLR are more pronounced with an increasing sample size or when causal SNPs are correlated with one another.
 
@@
 In this paper, we reported the maximum AUC of 102 different p-value thresholds, a threshold that should normally be learned on the training set only.
-To our knowledge, there is no clear standard on how to choose these two critical hyper-parameters.
+To our knowledge, there is no clear standard on how to choose these two critical hyper-parameters for C+T.
 
-On the contrary, in the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$ that accounts for the amount of regularization used in the model.
+On the contrary, for the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$ that accounts for the amount of regularization used in the model.
 Not only this procedure provides near-optimal results, but it also accelerates the model training thanks to the development of an early stopping criterion.
@@
 For this, we introduced a simple feature engineering technique that enables PLR to detect and learn not only additive effects, but also dominant and recessive effects.
-This technique improves the predictive performance of PLR when there are some non-linear effects in the simulations, while providing nearly the same predictive performance when there are only additive effects.
+This technique improves the predictive performance of PLR when there are some non-linear effects in the simulations, while providing nearly the same predictive performance when there are additive effects only.
 Moreover, it also improves predictive performance for the celiac disease.
@@
 Yet, this approach is not able to detect interaction-type effects. 
-In order to capture interaction-type effects, we tested T-Trees, a method that is able to exploit SNP correlations thanks to special decision trees \cite[]{botta2014exploiting}. 
+In order to capture interaction-type effects, we tested T-Trees, a method that is able to exploit SNP correlations and interactions thanks to special decision trees \cite[]{botta2014exploiting}. 
 However, predictive performance of T-Trees were consistently lower than with penalized logistic regression, even when simulating a model with dominant and interaction-type effects that T-Trees should be able to handle.
@@
 As of today, the proposed penalized logistic regression does not allow for the analysis of summary data, but this represents an important future direction of our work.
-The current version is of particular interest for the analysis of modern SNP datasets including hundreds of thousands of individuals.
+The current version is of particular interest for the analysis of modern individual-level datasets including hundreds of thousands of individuals.
 
@@
 Thus, we strongly believe that our implementation of penalized regressions would be appropriate for predicting autoimmune diseases with a few strong effects (e.g.\ celiac disease) as well as highly polygenic traits (e.g.\ standing height).
-These models could also be used to predict phenotypes based on other omics data since the implementation is not specific to genotype data.
+Finally, note that these models could also be used to predict phenotypes based on other omics data since the implementation is not specific to genotype data.
 
@@
 
-\subsection*{``Fancy'' model}
+\subsection*{Model ``COMP''}
 
@@
 For every combination of parameters, AUC of oracles are comprised between 83.2\% and 84.2\% for an heritability of 50\% and between 93.2\% and 94.1\% for an heritability of 80\%.
-Given all these estimates of the maximal achievable AUC and for the sake of simplicity, we report maximum AUCs of 84\% (94\%) for heritabilities of 50\% (80\%) whatever are the parameters of the simulations.
+Given all these estimates of maximal achievable AUC and for the sake of simplicity, we report maximum AUCs of 84\% (94\%) for heritabilities of 50\% (80\%) whatever are the simulation parameters.
 
@@
 Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
-Horizontal panels are presenting results for the ``simple'' and ``fancy'' models for simulating phenotypes.
+Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes.
 \textbf{A:} Mean of AUC over 5 simulations.
@@
 Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
-Horizontal panels are presenting results for the ``simple'' and ``fancy'' models for simulating phenotypes.
+Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes.
 \textbf{A:} Mean of AUC over 100 simulations.

### 2018-09-24 paper/PRS-supp.tex
some reformating

--- 2018-09-24 paper/draft.tex
@@
 
-\linenumbers
-\doublespacing
+%\linenumbers
+%\doublespacing
 %\usepackage[authoryear]{natbib}
@@
 
-\title{Efficient implementation of penalized regression\\for genetic risk prediction}
-\author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
-
-\date{~ }
-\maketitle
-
-\noindent$^{\text{\sf 1}}$Universit\'e Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
-\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France.
-
-\noindent$^\ast$To whom correspondence should be addressed.
-
-\newpage
-
-\abstract{
-Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
-PRS might have a major impact on public health, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.
-The ``Clumping+Thresholding'' (C+T) approach is the most common method to derive PRS.
-C+T uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it fast and easy to use.
-However, previous work showed that jointly estimating SNP effects for computing PRS has the potential to significantly improve the predictive performance of PRS as compared to C+T.
-
-In this paper, we present an efficient method to jointly estimate SNP effects, allowing for practical application of penalized logistic regression (PLR) on modern datasets including hundreds of thousands of individuals.
-Moreover, our implementation of PLR directly includes automatic choices for hyper-parameters.
-The choice of hyper-parameters for a predictive model is very important since it can dramatically impact its predictive performance.
-As an example, AUC values range from less than 60\% to 90\% in a model with 30 causal SNPs, depending on the p-value threshold in C+T.
-
-We compare the performance of PLR, C+T and a derivation of random forests using both real and simulated data. 
-PLR consistently achieves higher predictive performance than the two other methods while being as fast as C+T. 
-We find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; for instance, AUC values increase from 83\% with the best prediction of C+T to 92.5\% with PLR.
-We confirm these results in a data analysis of a case-control study for celiac disease where PLR and the standard C+T method achieve AUC of 89\% and of 82.5\%.
-
-In conclusion, our study demonstrates that penalized logistic regression can achieve more discriminative polygenic risk scores, while being applicable to large-scale individual-level data thanks to the implementation we provide in the R package bigstatsr.\\
-
-\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-
-\section{Introduction}
-
-Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
-PRS are useful for genetic epidemiology when testing the polygenicity of one disease and finding a common genetic contribution between two diseases \cite[]{purcell2009common}.
-Personalized medicine is another major application of PRS.
-Personalized medicine envisions to use PRS in screening campaigns in order to identify high-risk individuals for a given disease \cite[]{chatterjee2016developing}.
-As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}.
-Yet, PRS would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases.
-For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.
-
-Several methods have been developed to predict disease status, or more generally any phenotype, based on SNP information.
-A commonly used method often called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') is used to derive PRS from results of Genome-Wide Association Studies (GWAS) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
-This technique uses GWAS summary statistics only, allowing for a fast implementation of C+T.
-However, C+T also has several limitations; for instance, previous studies have shown that predictive performance of C+T is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
-Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}.
-Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes \cite[]{abraham2013performance}.
-Moreover, these methods and their derivatives are often computationally very demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
-Finally, statistical learning methods have also been used to derive PRS for complex human diseases by jointly estimating SNP effects.
-Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,wei2009disease}.
-
-We recently developed two R packages, bigstatsr and bigsnpr, for efficiently analyzing large-scale genome-wide data \cite[]{prive2017efficient}.
-Package bigstatsr now includes an efficient algorithm with a new implementation for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
-In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression (PLR) against the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
-In this comparison, we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
-For C+T, we report results for a large grid of hyper-parameters. 
-For PLR, the choice of hyper-parameters is included in the algorithm so that we report only one model for each simulation.
-We also use a modified version of PLR in order to capture not only linear effects, but also recessive and dominant effects.
-
-To perform simulations, we use real genotype data and simulate new phenotypes. 
-In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the disease heritability.
-We also compare two different models for simulating phenotypes, one with additive effects only, and one that combines additive, dominant and interaction-type effects.
-Overall, we find that PLR consistently achieves higher predictive performance than the C+T and T-Trees methods while being as fast as C+T.
-This demonstrates the feasibility and relevance of this approach for PRS computation on large modern datasets.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Methods}
-
-\subsection{Genotype data}
-
-We use real genotypes of European individuals from a case-control study for celiac disease \cite[]{dubois2010multiple}.
-The composition of this dataset is presented in table \ref{tab:celiac-data}.
-Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
-For simulations presented later, we first restrict this dataset to controls from UK in order to remove the genetic structure induced by the celiac disease status and population structure.
-This filtering process results in a sample of 7100 individuals (see supplementary notebook ``preprocessing'').
-We also use this dataset for real data application, in this case keeping all 15,155 individuals (4496 cases and 10,659 controls). Both datasets contain 281,122 SNPs.
-
-\subsection{Simulations of phenotypes} \label{sec:simus}
-
-We simulate binary phenotypes using a Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}.
-We vary parameters of the simulations in order to match a range of genetic architecture from low to high polygenicity.
-This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 autosomal chromosomes or 30 in the HLA region of chromosome 6), and the disease heritability (50\% or 80\%).
-%SNP effects are drawn from either a Gaussian or from a Laplace distribution.
-Liability scores are computed either from an additive model with linear effects only (``ADD'') or a more complex model that combines additive, dominant and interaction-type effects (``COMP'').
-For model ``ADD'', we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$ and $S_\text{causal}$ is the set of causal SNPs.
-For model ``COMP'', we simulate liability scores using additive, dominant and interaction-type effects (see Supplementary Materials).
-
-We implement 3 different simulation scenarios, summarized in table \ref{tab:simus}.
-Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes -- 281,122 SNPs) and a training set of size 6000.
-It compares all methods described in section \ref{sec:methods}.
-For each combination of the remaining parameters, results are based on 100 simulations excepted when comparing PLR with T-Trees, which relies on 5 simulations only because of a much higher computational burden of T-Trees as compared to other methods.
-Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only (18,941 SNPs).
-Reducing the number of SNPs increases the polygenicity (i.e.
-the proportion of causal SNPs) of the simulated models.
-Reducing the number of SNPs ($p$) is also equivalent to increasing the sample size ($n$) as predictive power is dependent on $n/p$ \cite[]{dudbridge2013power,vilhjalmsson2015modeling}.
-For this scenario, we use the additive model only, but continue to vary all other simulation parameters.
-Finally, scenario \textnumero3 uses the whole dataset as in scenario \textnumero1 while varying the size of the training set in order to assess how the sample size affects predictive performance of methods.
-A total of 100 simulations per combination of parameters are run using 300 causal SNPs randomly chosen on the genome.
-
-\subsection{Predictive performance measures}\label{sec:auc}
-
-In this study, we use two different measures of predictive accuracy.
-First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction,lusted1971signal}.
-In the case of our study, the AUC is the probability that the PRS of a case is greater than the PRS of a control.
-This measure indicates the extent to which we can distinguish between cases and controls using PRS. 
-As a second measure, we also report the partial AUC for specificities between 90\% and 100\% \cite[]{dodd2003partial,mcclish1989analyzing}.
-This measure is similar to the AUC, but focuses on high specificities, which is the most useful part of the ROC curve in clinical settings.
-When reporting AUC results of simulations, we also report maximum achievable AUC values of 84\% and 94\% for heritabilities of respectively 50\% and 80\%.
-These estimates are based on three different yet consistent estimations (see Supplementary Materials).
-%Note that we also report the timing of the main computations and the number of SNPs used in the predictions.
-
-\subsection{Methods compared} \label{sec:methods}
-
-In this study, we compare three different types of methods: the C+T method, T-Trees and penalized logistic regression (PLR).
-
-The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWAS).
-In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$.
-The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another (this set of SNPs is denoted $S_\text{clumping}$).
-Then, thresholding (T) consists in removing SNPs with p-values larger than a user-defined threshold $p_T$.
-Finally, the PRS for individual $i$ is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS.
-In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$.
-We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``C+T-all''), one including only SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``C+T-stringent''), and one that maximizes the AUC (``C+T-max'') for 102 thresholds between $1$ and $10^{-100}$ (Table \ref{tab:thr}).
-As we report the optimal threshold based on the test set, the AUC for ``C+T-max'' is an upper bound of the AUC for the C+T method.
-
-T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}.
-We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees instead of 1000.
-Using 1000 trees provides a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown).
-%We call this method ``T-Trees'' in the results.
-
-Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
-Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
-the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
-the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $p \gg n$ situation ($p$ is the number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
-In this study, we use an embedded grid search over $\alpha \in \{1,~0.5,~0.05,~0.001\}$.
-
-To fit this penalized logistic regression, we use an efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in R package bigstatsr.
-This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''.
-To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
-Because of L1-regularization, the resulting vectors of coefficients are sparse and can be used to make a PRS based on a \emph{linear} combination of allele counts.
-We refer to this method as ``PLR'' in the results section.
-
-To capture recessive and dominant effects on top of additive effects in PLR, we use simple feature engineering: we construct a separate dataset with 3 times as many variables as the initial one. 
-For each SNP variable, we add two more variables coding for recessive and dominant effects: one variable is coded 1 if homozygous variant and 0 otherwise, and the other is coded 0 for homozygous referent and 1 otherwise.
-We refer to this method as ``PLR3'' in the rest of the paper.
-
-\subsection{Evaluating predictive performance for Celiac data}
-
-We use Monte Carlo cross-validation to compute AUC, partial AUC, the number of predictors and execution time for the original Celiac dataset with the observed case-control status: we randomly split 100 times the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Results}
-
-\subsection{Joint estimation improves predictive performance}
-
-We compared penalized logistic regression (PLR) with the C+T method using simulations of scenario \textnumero1 (Table \ref{tab:simus}).
-
-When simulating a model with 30 causal SNPs and an heritability of 80\%, PLR provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T range between 83\% and 90\% (Figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-PRS}).
-Moreover, PLR consistently provides higher predictive performance than C+T across all scenarios we considered, excepted in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).
-
-PLR provides particularly higher predictive performance than C+T when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region.
-In this situation, the mean AUC reaches 92.5\% for PLR and 84\% for ``C+T-max'' (Figure \ref{fig:main-AUC-logit}).
-
-Note that, for the simulations, we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).
-
-\subsection{Importance of hyper-parameters}
-
-In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can dramatically impact the predictive performance of C+T. 
-For example, in a model with 30 causal SNPs only, AUC ranges from less than 60\% when using all SNPs passing clumping to 90\% \emph{if} choosing the optimal p-value threshold (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
-
-Concerning the $r^2$ threshold of the clumping step in C+T, we mostly used the common value of $0.2$.
-Yet, using a more stringent value of $0.05$ provides higher predictive performance than using $0.2$ in most of the cases we considered (Figures \ref{fig:supp-AUC-all-r2}, \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-chr6-all-r2})
-
-Our implementation of PLR that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of 100 models corresponding to different values of $\lambda$ (Figure \ref{fig:supp-biglasso}).
-
-\subsection{Non-linear effects}
-
-We tested the T-Trees method in scenario \textnumero1.
-As compared to PLR, T-Trees perform worse in terms of predictive ability, while taking much longer to run (Figure \ref{fig:supp-ttrees}). 
-Even for simulations with model ``COMP'' in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower when using T-Trees than when using PLR (Figure \ref{fig:supp-ttrees}).
-
-We also compared the two penalized logistic regressions in scenario \textnumero1: PLR versus PLR3 that uses additional features (variables) coding for recessive and dominant effects.
-Predictive performance of PLR3 are nearly as good as PLR when there are only addtive effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
-For model ``COMP'', PLR3 provides AUC values at least 3.5\% higher than PLR, excepted when there are 3000 causal SNPs.
-Yet, PLR3 takes 2-3 times as much time to run and requires 3 times as much disk storage as PLR.
-
-\subsection{Simulations varying number of SNPs and training size} %%%%
-
-First, when reproducing simulations of scenario \textnumero1 using chromosome 6 only (scenario \textnumero2), the predictive performance of PLR always increase (Figure \ref{fig:supp-AUC-chr6-all-r2}).
-There is a particularly large increase when simulating 3000 causal SNPs: AUC from PLR increases from 60\% to nearly 80\% for Gaussian effects and a disease heritability of 80\%.
-On the contrary, when simulating only 30 or 300 causal SNPs with the corresponding dataset, AUC of ``C+T-max'' does not increase, and even decreases for an heritability of 80\% (Figure \ref{fig:supp-AUC-chr6-all-r2}).
-Secondly, when varying the training size (scenario \textnumero3), we report an increase of AUC with a larger training size, with a faster increase of AUC for PLR as compared to ``C+T-max'' (Figure  \ref{fig:main-AUC-ntrain}).
-
-\subsection{Polygenic scores for the celiac disease}
-
-Joint logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with PLR and 89.1\% with PLR3 as compared to 82.5\% with ``C+T-max''.
-The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with PLR, PLR3 and ``C+T-max'', respectively.
-Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360 (Table \ref{tab:results-celiac}, figure \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
-In terms of computation time, we show that PLR, while learning jointly on all SNPs at once and testing four different hyper-parameter values, is almost as fast as the C+T method (190 vs 130 seconds), and PLR3 takes less than twice as long as PLR (296 vs 190 seconds).
-
-\begin{table}[h]
-\caption{Results for the real Celiac dataset.
-The results are averaged over 100 runs where the training step is randomly composed of 12,000 individuals.
-In the parentheses is reported the standard deviation of $10^5$ bootstrap samples of the mean of the corresponding variable.
-Results are reported with 3 significant digits.\label{tab:results-celiac}}
-\vspace*{0.5em}
-\centering
-\begin{tabular}{|l|c|c|c|c|}
-  \hline
-Method & AUC & pAUC & \# predictors & Execution time (s) \\
-  \hline
-C+T-max & 0.825 (0.000664) & 0.0289 (0.000187) & 8360 (744) & 130 (0.143) \\ 
-PLR & 0.887 (0.00061) & 0.0411 (0.000224) & 1570 (46.4) & 190 (1.21) \\ 
-PLR3 & 0.891 (0.000628) & 0.0426 (0.000219) & 2260 (56.1) & 296 (2.03) \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Discussion}
-
-\subsection{Joint estimation improves predictive performance}
-
-In this comparative study, we present a computationally efficient implementation of penalized logistic regression (PLR). 
-This model can be used to build polygenic risk scores based on very large individual-level SNP datasets such as the UK biobank \cite[]{bycroft2017genome}. 
-In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently.
-PLR always outperform the C+T method, excepted in some highly underpowered cases, and the benefits of using PLR are more pronounced with an increasing sample size or when causal SNPs are correlated with one another.
-
-\subsection{Importance of hyper-parameters}
-
-The choice of hyper-parameter values is very important since it can greatly impact method performance.
-In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps, respectively.
-The choice of the $r^2$ threshold of the clumping step is important. 
-Indeed, on the one hand, choosing a low value for this threshold may discard informative SNPs that are correlated.
-Yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would add some noise to the PRS.
-Based on the simulations, we find that using a stringent threshold ($r^2 = 0.05$) leads to higher predictive performance, even when causal SNPs are correlated.
-It means that, in most cases, avoiding redundant information is more important than including all causal SNPs.
-The choice of the $p_T$ threshold is also very important as it can greatly impact the predictive performance of the C+T method, which we confirm in this study \cite[]{ware2017heterogeneity}. 
-In this paper, we reported the maximum AUC of 102 different p-value thresholds, a threshold that should normally be learned on the training set only.
-To our knowledge, there is no clear standard on how to choose these two critical hyper-parameters for C+T.
-
-On the contrary, for the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$ that accounts for the amount of regularization used in the model.
-Not only this procedure provides near-optimal results, but it also accelerates the model training thanks to the development of an early stopping criterion.
-Usually, cross-validation is used to choose hyper-parameter values and then the model is trained again with these particular hyper-parameter values \cite[]{Hastie2008,wei2013large}.
-Yet, performing cross-validation and retraining the model is computationally demanding; CMSA offers a less burdensome alternative.
-Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we use a grid  search directly embedded in the CMSA procedure. 
-%Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model.
-Yet, if using e.g.\ $\alpha = 0.001$, it would lead to a similar model.
-
-\subsection{Non-linear effects}
-
-We also explored how to capture non-linear effects.
-For this, we introduced a simple feature engineering technique that enables PLR to detect and learn not only additive effects, but also dominant and recessive effects.
-This technique improves the predictive performance of PLR when there are some non-linear effects in the simulations, while providing nearly the same predictive performance when there are additive effects only.
-Moreover, it also improves predictive performance for the celiac disease.
-
-Yet, this approach is not able to detect interaction-type effects. 
-In order to capture interaction-type effects, we tested T-Trees, a method that is able to exploit SNP correlations and interactions thanks to special decision trees \cite[]{botta2014exploiting}. 
-However, predictive performance of T-Trees were consistently lower than with penalized logistic regression, even when simulating a model with dominant and interaction-type effects that T-Trees should be able to handle.
-
-\subsection{Limitations}
-
-Our approach has one major limitation: the main advantage of the C+T method is its direct applicability to summary statistics, allowing to leverage the largest GWAS results to date, even when individual cohort data cannot be merged because of practical or ethical reasons (e.g.\ consortium data including many cohorts).
-As of today, the proposed penalized logistic regression does not allow for the analysis of summary data, but this represents an important future direction of our work.
-The current version is of particular interest for the analysis of modern individual-level datasets including hundreds of thousands of individuals.
-
-Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}.
-%In future work, we aim at exploring how we can overcome those issues.
-%We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}.
-Finally, we also wish to assess how can we combine the information provided by genetic data with clinical and environmental data, possibly in a non-linear way.
-
-\subsection{Conclusion}
-
-In this comparative study, we have presented a computationally efficient implementation of penalized logistic regression that can be used to predict disease status based on genotypes.
-Note that a similar penalized linear regression is also available in our software.
-Usually, joint statistical learning methods are so computationally demanding that they cannot be used on large-scale datasets such as the UK biobank \cite[]{bycroft2017genome}. 
-A partial solution to this problem is to pre-filter data, but it usually leads to a decrease in predictive performance as some information is removed \cite[]{accuheight2018}.
-This pre-filtering step is not necessary anymore thanks to our implementation.
-Moreover, the model we present in this paper can flexibly adapt to traits with different architectures thanks to two hyper-parameters that are automatically learned with the model.
-Thus, we strongly believe that our implementation of penalized regressions would be appropriate for predicting autoimmune diseases with a few strong effects (e.g.\ celiac disease) as well as highly polygenic traits (e.g.\ standing height).
-Finally, note that these models could also be used to predict phenotypes based on other omics data since the implementation is not specific to genotype data.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-\begin{table}[htbp]
-\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.\label{tab:simus}}
-\vspace*{0.5em}
-\centering
-\begin{tabular}{|c|c|c|c|c|c|c|c|}
-\hline
-Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
-scenario & & training set & (number and location) & of effects & & model & \\
-\hline
-\hline
-\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{ADD} & PRS \\
-& & & 30 in all & & & & PLR \\
-& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{COMP} & PLR3 \\
-& & & 3000 in all & & & & (T-Trees) \\
-\hline
-\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{ADD} & PRS \\ 
-& & & & & & & PLR\\
-\hline
-\multirow{5}{*}{3} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
-& & 2000 & & & & & \\
-& & 3000 & & & & & \\
-& & 4000 & & & & & \\
-& & 5000 & & & & & \\
-\hline
-\end{tabular}
-\end{table}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
-\caption{Main comparison of C+T and PLR in scenario \textnumero1 for model ``ADD'' and an heritability of 80\%.
-Mean of AUC over 100 simulations for PLR and the maximum AUC reported with ``C+T-max''.
-Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.}
-\label{fig:main-AUC-logit}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{main-AUC-PRS}}
-\caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for model ``ADD'' and an heritability of 80\%.
-Mean of AUC over 100 simulations.
-Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.}
-\label{fig:main-AUC-PRS}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
-\caption{Comparison of models when varying sample size in scenario \textnumero3 for model ``ADD'' with 300 causal SNPs sampled anywhere on the genome.
-Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the training size.
-Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.}
-\label{fig:main-AUC-ntrain}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{celiac-roc}}
-\caption{ROC Curves for C+T, PLR and PLR3 for the celiac disease dataset.
-Models were trained using 12,000 individuals.
-These are results projecting these models on the remaining 3155 individuals.
-The figure is plotted using R package plotROC \cite[]{sachs2017plotroc}.\label{fig:celiac-roc}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-\section*{Acknowledgements}
-
-Authors acknowledge LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).
-Authors also acknowledge the Grenoble Alpes Data Institute that is supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02).
-We are also grateful to F\'elix Balazard for useful discussions about T-Trees, and to Yaohui Zeng for useful discussions about R package biglasso.
-
-\section*{Reproducibility}
-
-All the code used in this paper along with results such as figures and tables, are available as HTML R notebooks in the Supplementary Materials.
-Results are available at [TODO]. A tutorial on how to start with R packages bigstatsr and bigsnpr is available at \url{https://privefl.github.io/bigsnpr/articles/demo.html}.
-
-\vspace*{-12pt}
-
-\bibliographystyle{natbib}
-\bibliography{refs}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-\section*{Supplementary Materials}
+\section*{Supplementary Material and Methods}
 
@@
 We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction} ~~+~~ \epsilon_i~,$$
-where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$.
+where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$, $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$.
 %Note that, because we use the raw genotype counts when coding the interaction terms, the corresponding SNPs also display a marginal effect.
@@
 We used three different ways to estimate the maximum achievable AUC for our simulations.
+
 First, we used the estimation from equation (3) of \cite{wray2010genetic}.
@@
 Note that this approximation is reported to be less accurate for high heritabilities.
+
 Secondly, if we assume that the genetic part of the liabilities follows a Gaussian distribution $N(0, h^2)$ and that the environmental part follows a Gaussian distribution $N(0, 1 - h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the disease heritability $h^2$ through Monte Carlo simulations.
 We report AUCs of 84.1\% and 94.1\% for an heritability of 50\% and 80\%, respectively.
-Thirdly, we reproduce the exact same procedure of simulations and, for each combination of parameters (Table \ref{tab:simus}), we estimate the AUC of the ``oracle'', i.e.\ the true simulated genetic part of the liabilities through 100 replicates.
+
+Thirdly, we reproduce the exact same procedure of simulations and, for each combination of parameters (Table 2), we estimate the AUC of the ``oracle'', i.e.\ the true simulated genetic part of the liabilities through 100 replicates.
 For every combination of parameters, AUC of oracles are comprised between 83.2\% and 84.2\% for an heritability of 50\% and between 93.2\% and 94.1\% for an heritability of 80\%.
+
 Given all these estimates of maximal achievable AUC and for the sake of simplicity, we report maximum AUCs of 84\% (94\%) for heritabilities of 50\% (80\%) whatever are the simulation parameters.
+
+\vspace*{-12pt}
+
+\bibliographystyle{natbib}
+\bibliography{refs}
 

### 2018-09-24 paper/PRS.tex
some reformating

--- 2018-09-24 paper/draft.tex
@@
 \usepackage{bbm}
+
+\usepackage{xr}
+\externaldocument{PRS-supp}
 
@@
 
-\section{Methods}
+\section{Material and Methods}
 
@@
 Liability scores are computed either from an additive model with linear effects only (``ADD'') or a more complex model that combines additive, dominant and interaction-type effects (``COMP'').
-For model ``ADD'', we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$ and $S_\text{causal}$ is the set of causal SNPs.
+For model ``ADD'', we compute the liability score of the i-th individual \[y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,\] where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$ and $S_\text{causal}$ is the set of causal SNPs.
 For model ``COMP'', we simulate liability scores using additive, dominant and interaction-type effects (see Supplementary Materials).
@@
 Then, thresholding (T) consists in removing SNPs with p-values larger than a user-defined threshold $p_T$.
-Finally, the PRS for individual $i$ is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS.
+Finally, the PRS for individual $i$ is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients \[\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,\] where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS.
 In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$.
@@
 
-Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
+Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function \[L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,\] where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
 Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
@@
 Note that a similar penalized linear regression is also available in our software.
-Usually, joint statistical learning methods are so computationally demanding that they cannot be used on large-scale datasets such as the UK biobank \cite[]{bycroft2017genome}. 
-A partial solution to this problem is to pre-filter data, but it usually leads to a decrease in predictive performance as some information is removed \cite[]{accuheight2018}.
-This pre-filtering step is not necessary anymore thanks to our implementation.
-Moreover, the model we present in this paper can flexibly adapt to traits with different architectures thanks to two hyper-parameters that are automatically learned with the model.
-Thus, we strongly believe that our implementation of penalized regressions would be appropriate for predicting autoimmune diseases with a few strong effects (e.g.\ celiac disease) as well as highly polygenic traits (e.g.\ standing height).
+Our approach solves the dramatic computational burden faced by standard implementations, thus allowing for the analysis of large-scale datasets such as the UK biobank \cite[]{bycroft2017genome}.
+
+We also demonstrated in simulations that our implementation of penalized regressions remains highly effective over a broad range of disease architectures.
+It can be appropriate for predicting autoimmune diseases with a few strong effects (e.g.\ celiac disease) as well as highly polygenic traits (e.g.\ standing height).
 Finally, note that these models could also be used to predict phenotypes based on other omics data since the implementation is not specific to genotype data.
@@
 \hline
-\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{ADD} & PRS \\
+\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{ADD} & C+T \\
 & & & 30 in all & & & & PLR \\
@@
 \hline
-\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{ADD} & PRS \\ 
+\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{ADD} & C+T \\ 
 & & & & & & & PLR\\
@@
 \caption{Comparison of models when varying sample size in scenario \textnumero3 for model ``ADD'' with 300 causal SNPs sampled anywhere on the genome.
-Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the training size.
+Mean of AUC over 100 simulations for the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the training size.
 Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
@@
 \clearpage
+
+\section*{Description of Supplemental Data}
+
+Supplemental Data include a PDF with two sections of methods, two tables and ten figures. 
+Supplemental Data also include six HTML R notebooks including all code and results used in this paper, for reproducibility purposes.
+
+\section*{Declaration of Interests}
+
+The authors declare no competing interests.
+
 \section*{Acknowledgements}
@@
 
-\section*{Reproducibility}
+\section*{Web Resources}
 
-All the code used in this paper along with results such as figures and tables, are available as HTML R notebooks in the Supplementary Materials.
-Results are available at [TODO]. A tutorial on how to start with R packages bigstatsr and bigsnpr is available at \url{https://privefl.github.io/bigsnpr/articles/demo.html}.
+Results of simulations are available at [TODO].
+A tutorial on how to start with R packages bigstatsr and bigsnpr is available at \url{https://privefl.github.io/bigsnpr/articles/demo.html}.
+The two R packages are available on GitHub.
 
-\vspace*{-12pt}
+\newpage
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-\section*{Supplementary Materials}
-
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{table}{0}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection*{Model ``COMP''}
-
-For model ``COMP'', we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$
-and $S_\text{causal}^{(3)}$; $S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$.
-We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction} ~~+~~ \epsilon_i~,$$
-where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$.
-%Note that, because we use the raw genotype counts when coding the interaction terms, the corresponding SNPs also display a marginal effect.
-
-\subsection*{Maximum AUCs} \label{sec:auc-max}
-
-We used three different ways to estimate the maximum achievable AUC for our simulations.
-First, we used the estimation from equation (3) of \cite{wray2010genetic}.
-For a prevalence fixed at 30\% and an heritability of 50\% (respectively 80\%), the approximated theoretical values of AUC are 84.1\% (respectively 93.0\%).
-Note that this approximation is reported to be less accurate for high heritabilities.
-Secondly, if we assume that the genetic part of the liabilities follows a Gaussian distribution $N(0, h^2)$ and that the environmental part follows a Gaussian distribution $N(0, 1 - h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the disease heritability $h^2$ through Monte Carlo simulations.
-We report AUCs of 84.1\% and 94.1\% for an heritability of 50\% and 80\%, respectively.
-Thirdly, we reproduce the exact same procedure of simulations and, for each combination of parameters (Table \ref{tab:simus}), we estimate the AUC of the ``oracle'', i.e.\ the true simulated genetic part of the liabilities through 100 replicates.
-For every combination of parameters, AUC of oracles are comprised between 83.2\% and 84.2\% for an heritability of 50\% and between 93.2\% and 94.1\% for an heritability of 80\%.
-Given all these estimates of maximal achievable AUC and for the sake of simplicity, we report maximum AUCs of 84\% (94\%) for heritabilities of 50\% (80\%) whatever are the simulation parameters.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{|l|cccc|r|}
-\hline
-Population & UK & Finland & Netherlands & Italy & Total\\
-\hline
-Cases & 2569 & 637 & 795 & 495 & 4496\\
-Controls & 7492 & 1799 & 828 & 540 & 10659\\
-\hline
-Total & 10061 & 2436 & 1623 & 1035 & 15155\\
-\hline
-\end{tabular}
-\caption{Number of individuals by population and disease status in the celiac disease case-control study (after quality control, genotyped on 281,122 SNPs).\label{tab:celiac-data}}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% latex table generated in R 3.4.1 by xtable 1.8-2 package
-% Thu Oct 26 16:46:26 2017
-\begin{table}[ht]
-\centering
-\begin{tabular}{ccccccccccccccc}
-  \hline
-1.00e+00 & 7.22e-01 & 5.87e-01 & 4.20e-01 & 2.43e-01 & 1.00e-01 & 2.35e-02 & 2.21e-03 & 4.69e-05 & 8.81e-08 & 3.18e-12 & 1.83e-19 & 2.89e-31 & 1.70e-50 & 7.71e-82 \\ 
-  5.00e-08 & 7.05e-01 & 5.65e-01 & 3.95e-01 & 2.20e-01 & 8.47e-02 & 1.79e-02 & 1.42e-03 & 2.28e-05 & 2.73e-08 & 4.69e-13 & 8.08e-21 & 1.80e-33 & 4.30e-54 & 1.06e-87 \\ 
-  7.94e-01 & 6.87e-01 & 5.42e-01 & 3.69e-01 & 1.97e-01 & 7.08e-02 & 1.34e-02 & 8.83e-04 & 1.05e-05 & 7.74e-09 & 6.03e-14 & 2.86e-22 & 7.73e-36 & 5.97e-58 & 5.49e-94 \\ 
-  7.81e-01 & 6.69e-01 & 5.19e-01 & 3.43e-01 & 1.75e-01 & 5.85e-02 & 9.79e-03 & 5.31e-04 & 4.61e-06 & 2.01e-09 & 6.69e-15 & 7.92e-24 & 2.24e-38 & 4.37e-62 & 1.00e-100 \\ 
-  7.67e-01 & 6.50e-01 & 4.95e-01 & 3.18e-01 & 1.54e-01 & 4.76e-02 & 7.01e-03 & 3.08e-04 & 1.90e-06 & 4.72e-10 & 6.32e-16 & 1.70e-25 & 4.26e-41 & 1.61e-66 &  \\ 
-  7.53e-01 & 6.30e-01 & 4.70e-01 & 2.93e-01 & 1.35e-01 & 3.82e-02 & 4.90e-03 & 1.72e-04 & 7.31e-07 & 1.00e-10 & 5.04e-17 & 2.75e-27 & 5.16e-44 & 2.83e-71 &  \\ 
-  7.38e-01 & 6.09e-01 & 4.46e-01 & 2.68e-01 & 1.17e-01 & 3.02e-02 & 3.33e-03 & 9.18e-05 & 2.63e-07 & 1.89e-11 & 3.35e-18 & 3.31e-29 & 3.84e-47 & 2.26e-76 &  \\ 
-   \hline
-\end{tabular}
-\caption{The 102 thresholds used for the C+T method for this study.\label{tab:thr}}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{simple-CMSA}}
-\caption{Illustration of one turn of the Cross-Model Selection and Averaging (CMSA) procedure. First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds). 
-Secondly, in turn, each fold is considered as an inner validation set (red) and the other ($K - 1$) folds form an inner training set (blue).
-A ``regularization path'' of models is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed.
-The model that minimizes the loss on the inner validation set is selected.
-Finally, the $K$ resulting models are averaged.
-We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making this procedure much faster.}
-\label{fig:CMSA}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-corr}}
-\caption{Correlation between AUC and partial AUC values in scenario \textnumero1.
-There is a Spearman correlation of 98\% between values of AUC and partial AUC.
-The relation between the two values are the same whatever are the disease heritability, distribution of effects and method used.\label{fig:supp-AUC-corr}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-logit}}
-\caption{Comparison of C+T and PLR in scenario \textnumero1 for model ``ADD'' and an heritability of 50\%.
-Mean of AUC over 100 simulations for PLR and the maximum AUC reported with the C+T method (``C+T-max'').
-Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.}
-\label{fig:supp-AUC-logit}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS}}
-\caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for model ``ADD'' and an heritability of 50\%.
-Mean of AUC over 100 simulations.
-Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.}
-\label{fig:supp-AUC-PRS}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-all-r2}}
-\caption{Comparison of models in scenario \textnumero1 for model ``ADD''.
-Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the number and location of causal SNPs.
-Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.}
-\label{fig:supp-AUC-all-r2}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
-\caption{Comparison of models in scenario \textnumero2 (using chromosome 6 only) for model ``ADD''.
-Thinner lines represents results in scenario \textnumero1 (Figure \ref{fig:supp-AUC-all-r2}).
-Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the number and location of causal SNPs.
-Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.}
-\label{fig:supp-AUC-chr6-all-r2}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-ttrees}}
-\caption{Comparison of T-Trees and PLR in scenario \textnumero1 for an heritability of 80\%.
-Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
-Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes.
-\textbf{A:} Mean of AUC over 5 simulations.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.
-\textbf{B:} Boxplots of numbers of predictors used by the methods for 5 simulations.
-\textbf{C:} Boxplots of execution times for 5 simulations.}
-\label{fig:supp-ttrees}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-triple}}
-\caption{Comparison of PLR3 and PLR in scenario \textnumero1 for an heritability of 80\%.
-Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
-Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes.
-\textbf{A:} Mean of AUC over 100 simulations.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.
-\textbf{B:} Boxplots of numbers of predictors used by the methods for 100 simulations.
-\textbf{C:} Boxplots of execution times for 100 simulations.}
-\label{fig:supp-triple}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-triple}}
-\caption{Comparison of PLR3 and PLR in scenario \textnumero1 for an heritability of 50\%.
-Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
-Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes.
-\textbf{A:} Mean of AUC over 100 simulations.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.}
-\label{fig:supp-AUC-triple}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-biglasso}}
-\caption{Comparison of PLR and the best prediction (among 100 tested $\lambda$ values) for ``biglasso'' (another implementation of penalized logistic regression -- \cite{zeng2017efficient}) in scenario \textnumero1.
-Simulations use model ``ADD'', an heritability of 80\% and $\alpha = 1$.
-Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
-\textbf{A:} Mean of AUC over 100 simulations.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
-The blue dotted line represents the maximum achievable AUC.
-\textbf{B:} Boxplots of numbers of predictors used by the methods for 100 simulations.
-\textbf{C:} Boxplots of execution times for 100 simulations.}
-\label{fig:supp-biglasso}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \end{document}

### 2018-09-27 4-results-simus.Rmd
new figures for pres

@@
 myggplot <- function(..., coeff = 1) {
-  ggplot(...) + 
-    bigstatsr::theme_bigstatsr(size.rel = coeff) +
-    theme(strip.text.x = element_text(size = rel(2)),
-          strip.text.y = element_text(size = rel(2)))
+  ggplot(...) + bigstatsr::theme_bigstatsr(size.rel = coeff)
 } 

### 2018-09-27 results-simus-pres.Rmd
new figures for pres

@@
 author: "Florian Privé"
-date: "March 19, 2018"
+date: "September 27, 2018"
 output: 
@@
 library(tidyverse)
+library(pkg.paper.PRS)
+options(nboot = 1e4)
 
-methods.color <- setNames(scales::hue_pal()(6), 
-                          c("logit-simple", "PRS-max", "PRS-stringent",
-                            "T-Trees", "logit-triple", "PRS-all")) 
+COLORS <- scales::hue_pal()(6)
+methods.color <- setNames(
+  c(COLORS, COLORS[c(4, 2, 5)], "black"), 
+  c("PLR", "C+T-max", "C+T-stringent", "T-Trees", "PLR3", "C+T-all",
+    "C+T-max-0.05", "C+T-max-0.2", "C+T-max-0.8", "biglasso")
+) 
 ```
@@
 ```{r}
-pAUC <- function(pred, target, p = 0.1) {
-  val.min <- min(target)
-  q <- quantile(pred[target == val.min], probs = 1 - p)
-  ind <- (target != val.min) | (pred > q)
-  tryCatch(bigstatsr::AUC(pred[ind], target[ind]) * p, 
-           error = function(e) p^2 / 2)  # all same prediction
-}
-
-format_results <- function(results) {
-  results %>%
-    as_tibble() %>%
-    filter(par.h2 == 0.8, par.dist == "gaussian") %>%
-    mutate(
-      par.causal = factor(map_chr(par.causal, ~paste(.x[1], .x[2], sep = " in ")),
-                          levels = c("30 in HLA", paste(3 * 10^(1:3), "in all"))),
-      AUC = map_dbl(eval, ~ bigstatsr::AUC(.x[, 1], .x[, 2])),
-      pAUC = map_dbl(eval, ~ pAUC(.x[, 1], .x[, 2]))
-    )
+myggplot <- function(..., coeff = 1) {
+  ggplot(...) + bigstatsr::theme_bigstatsr(size.rel = coeff)
 } 
@@
 ```{r}
-boot <- function(x, n = 1e5, f = mean) {
-  sd(replicate(n, f(sample(x, replace = TRUE))))
-}
-```
-
-```{r}
-myggplot <- function(..., coeff = 1) {
-  bigstatsr:::MY_THEME(ggplot(...), coeff = coeff) +
-  theme(strip.text.x = element_text(size = rel(2)),
-        strip.text.y = element_text(size = rel(2)))
-} 
-```
-
-```{r}
-barplot_causal <- function(results, h2) {
+barplot_causal <- function(results) {
 
@@
 ```{r}
-compare_logit <- function(results, with_method) {
+compare_PLR <- function(results, with_method) {
 
-  results <- filter(results, method %in% c("logit-simple", with_method))
+  results <- filter(results, method %in% c("PLR", with_method))
 
@@
                               fill = "method"), alpha = 0.3) + 
-      theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
+      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+      facet_grid(par.model ~ .) +
       labs(x = "Causal SNPs (number and location)", y = ylab, 
@@
     plot_extra("timing", "Execution time (in seconds)"),
-    ncol = 1, align = "hv", scale = 0.90, 
+    ncol = 1, align = "hv", scale = 0.95, 
     labels = LETTERS[2:3], label_size = 20
@@
 
-## Results with T-Trees
-
 ```{r}
-results1 <- list.files("results1", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  format_results()
-```
-
-```{r, out.width="90%", fig.width = 14}
-compare_logit(results1, with_method = "T-Trees")
-
-ggsave("figures/pres-ttrees.svg", scale = 1/100, width = 1300, height = 1060)
+corr <- readRDS("backingfiles/corr2.rds")
 ```
@@
 results2 <- list.files("results2", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  format_results()
+  read_format_results(corr) %>%
+  filter(par.h2 == 0.8, par.dist == "Gaussian")
+```
 
+### Results with PLR3
+
+```{r, out.width="90%", fig.width = 14}
+compare_PLR(results2, with_method = "PLR3")
+
+ggsave("../thesis-docs/figures/pres2-triple.svg", scale = 1/100, width = 1300, height = 1060)
+```
+
+## Results with only logit-simple & PRSs
+
+### 6000 in training & all chromosomes
+
+```{r}
 results2.summary <- results2 %>%
+  filter(par.model == "ADD", method != "PLR3") %>%
   group_by_at(c(vars(starts_with("par")), "method")) %>%
@@
 
-### Results with logit-triple
+```{r}
+results2.summary %>%
+  filter(method %in% c("PLR", "C+T-max")) %>%
+  barplot_causal_one()
 
-```{r, out.width="90%", fig.width = 14}
-compare_logit(results2, with_method = "logit-triple")
-
-ggsave("figures/pres-triple.svg", scale = 1/100, width = 1300, height = 1060)
+ggsave("../thesis-docs/figures/pres2-AUC-logit.svg", scale = 1/110, width = 1045, height = 770)
 ```
 
-## Results with only logit-simple & PRSs
-
 ```{r}
-results3 <- results2 %>%
-  filter(par.model == "simple", method != "logit-triple")
-```
-
-### 6000 in training & all chromosomes
-
-```{r}
-results3.summary <- results3 %>%
-  group_by_at(c(vars(starts_with("par")), "method")) %>%
-  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n())
-```
-
-```{r, fig.asp = 1, out.width = "60%", fig.width=8}
-results3.summary %>%
-  filter(method %in% c("logit-simple", "PRS-max")) %>%
+results2.summary %>%
+  filter(grepl("C\\+T", method)) %>%
   barplot_causal_one()
 
-ggsave("figures/pres-AUC-logit.svg", scale = 1/110, width = 1045, height = 770)
-```
-
-```{r, fig.asp = 1, out.width = "60%", fig.width=8}
-results3.summary %>%
-  filter(grepl("PRS", method)) %>%
-  barplot_causal_one()
-
-ggsave("figures/pres-AUC-PRS.svg", scale = 1/110, width = 1045, height = 770)
+ggsave("../thesis-docs/figures/pres2-AUC-PRS.svg", scale = 1/110, width = 1045, height = 770)
 ```
@@
 results4 <- list.files("results4", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  format_results()
+  read_format_results(corr) %>%
+  filter(par.h2 == 0.8, par.dist == "Gaussian")
 
-results4 %>%
+results4.summary <- results4 %>%
+  filter(method %in% c("PLR", "C+T-max")) %>%
+  mutate(method = c_method_r2(method, thr.r2)) %>%
   group_by_at(c(vars(starts_with("par")), "method")) %>%
@@
 
-results34 <- rbind(
-  cbind(simu = "all", results3),
-  cbind(simu = "chr6", results4)
-) %>%
-  as_tibble()
+results2.all.r2 <- results2 %>%
+  filter(par.model == "ADD", method %in% c("PLR", "C+T-max")) %>%
+  mutate(method = c_method_r2(method, thr.r2)) %>%
+  group_by_at(c(vars(starts_with("par")), "method")) %>%
+  summarise(AUC_mean = mean(AUC), AUC_boot = boot(AUC), N = n())
 ```
 
-```{r, fig.asp = 1, out.width = "90%"}
-results34 %>%
-  filter(par.h2 == 0.8, par.dist == "gaussian", method %in% c("logit-simple", "PRS-max")) %>%
-  myggplot() +
-  geom_hline(yintercept = 0.5, linetype = 2) +
-  geom_hline(aes(yintercept = 0.94), linetype = 3, color = "blue") +
-  geom_boxplot(aes(method, AUC, fill = simu, color = simu), alpha = 0.3) + 
-  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
-  facet_grid(. ~ par.causal) +
-  scale_fill_brewer(type = "qual", palette = 2) +
-  scale_color_brewer(type = "qual", palette = 2) + 
-  labs(x = "Method")
+```{r}
+results4.summary %>%
+  barplot_causal_one() + 
+  geom_col(aes(par.causal, AUC_mean), data = results2.all.r2, 
+           position = position_dodge(), color = "black", alpha = 0)
 
-ggsave("figures/pres-AUC-chr6.svg", scale = 1/110, width = 990, height = 700)
+ggsave("../thesis-docs/figures/pres2-AUC-chr6.svg", scale = 1/110, width = 990, height = 700)
 ```
@@
 results5 <- list.files("results5", full.names = TRUE) %>%
-  map_dfr(~readRDS(.x)) %>%
-  format_results()
+  read_format_results(corr) %>%
+  filter(par.h2 == 0.8, par.dist == "Gaussian")
 
@@
 
-results35.summary <- results3.summary %>%
-  filter(par.causal == "300 in all") %>%
+results35.summary <- results2.summary %>%
   cbind.data.frame(n.train = 6000) %>%
-  bind_rows(results5.summary)
+  bind_rows(results5.summary) %>%
+  filter(par.causal == "300 in all")
 ```
@@
 results35.summary %>%
-  filter(method %in% c("logit-simple", "PRS-max")) %>%
+  filter(method %in% c("PLR", "C+T-max")) %>%
   myggplot(aes(n.train, AUC_mean, color = method, linetype = method)) +
@@
 
-ggsave("figures/pres-AUC-ntrain.svg", scale = 1/100, width = 870, height = 635)
+ggsave("../thesis-docs/figures/pres2-AUC-ntrain.svg", scale = 1/100, width = 870, height = 635)
 ```

### 2018-10-01 paper/PRS.tex
last minor changes

@@
 
-\noindent$^{\text{\sf 1}}$Universit\'e Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
+\noindent$^{\text{\sf 1}}$Univ.\ Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
 \noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France.
@@
 \caption{Main comparison of C+T and PLR in scenario \textnumero1 for model ``ADD'' and an heritability of 80\%.
-Mean of AUC over 100 simulations for PLR and the maximum AUC reported with ``C+T-max''.
+Mean AUC over 100 simulations for PLR and the maximum AUC reported with ``C+T-max''.
 Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
 The blue dotted line represents the maximum achievable AUC.}
@@
 \caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for model ``ADD'' and an heritability of 80\%.
-Mean of AUC over 100 simulations.
+Mean AUC over 100 simulations.
 Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
 The blue dotted line represents the maximum achievable AUC.}
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
-\caption{Comparison of models when varying sample size in scenario \textnumero3 for model ``ADD'' with 300 causal SNPs sampled anywhere on the genome.
-Mean of AUC over 100 simulations for the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the training size.
+\caption{Comparison of methods when varying sample size in scenario \textnumero3 for model ``ADD'' with 300 causal SNPs sampled anywhere on the genome.
+Mean AUC over 100 simulations for the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the training size.
 Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
 The blue dotted line represents the maximum achievable AUC.}
@@
 
-Results of simulations are available at [TODO].
+Results of simulations are available at \url{https://figshare.com/articles/results_zip/7126964}.
 A tutorial on how to start with R packages bigstatsr and bigsnpr is available at \url{https://privefl.github.io/bigsnpr/articles/demo.html}.

### 2018-10-02 paper/PRS.tex
prepare submission

@@
 \title{Efficient implementation of penalized regression\\for genetic risk prediction}
-\author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
+\author{Florian Priv\'e,$^{\text{1,}*}$ Hugues Aschard$^{\text{2}}$ and Michael G.B. Blum$^{\text{1,}*}$}
 
@@
 
-\noindent$^{\text{\sf 1}}$Univ.\ Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
-\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France.
+\noindent$^{\text{\sf 1}}$Laboratoire TIMC-IMAG, UMR 5525, Univ.\ Grenoble Alpes, CNRS, La Tronche (38700), France, \\
+\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris (75015), France.
 
@@
 We simulate binary phenotypes using a Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}.
-We vary parameters of the simulations in order to match a range of genetic architecture from low to high polygenicity.
-This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 autosomal chromosomes or 30 in the HLA region of chromosome 6), and the disease heritability (50\% or 80\%).
+We vary simulation parameters in order to match a range of genetic architectures from low to high polygenicity.
+This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 autosomal chromosomes or 30 in the HLA region of chromosome 6), and the disease heritability $h^2$ (50\% or 80\%).
 %SNP effects are drawn from either a Gaussian or from a Laplace distribution.
-Liability scores are computed either from an additive model with linear effects only (``ADD'') or a more complex model that combines additive, dominant and interaction-type effects (``COMP'').
-For model ``ADD'', we compute the liability score of the i-th individual \[y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,\] where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$ and $S_\text{causal}$ is the set of causal SNPs.
+Liability scores are computed either from a model with additive effects only (``ADD'') or a more complex model that combines additive, dominant and interaction-type effects (``COMP'').
+For model ``ADD'', we compute the liability score of the i-th individual \[y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,\] where $S_\text{causal}$ is the set of causal SNPs, $w_j$ are weights generated from a Gaussian distribution $N(0, h^2 / \vert S_\text{causal} \vert)$ or a Laplace distribution $Laplace(0, \sqrt{h^2 / (2~\vert S_\text{causal} \vert)})$, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), and $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$.
 For model ``COMP'', we simulate liability scores using additive, dominant and interaction-type effects (see Supplementary Materials).
@@
 Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only (18,941 SNPs).
-Reducing the number of SNPs increases the polygenicity (i.e.
-the proportion of causal SNPs) of the simulated models.
+Reducing the number of SNPs increases the polygenicity (i.e.\ the proportion of causal SNPs) of the simulated models.
 Reducing the number of SNPs ($p$) is also equivalent to increasing the sample size ($n$) as predictive power is dependent on $n/p$ \cite[]{dudbridge2013power,vilhjalmsson2015modeling}.
@@
 This measure is similar to the AUC, but focuses on high specificities, which is the most useful part of the ROC curve in clinical settings.
-When reporting AUC results of simulations, we also report maximum achievable AUC values of 84\% and 94\% for heritabilities of respectively 50\% and 80\%.
+When reporting AUC results of simulations, we also report maximum achievable AUC values of 84\% and 94\% for heritabilities of 50\% and 80\% respectively.
 These estimates are based on three different yet consistent estimations (see Supplementary Materials).
@@
 
-In this study, we compare three different types of methods: the C+T method, T-Trees and penalized logistic regression (PLR).
+In this paper, we compare three different types of methods: the C+T method, T-Trees and penalized logistic regression (PLR).
 
@@
 In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$.
-We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``C+T-all''), one including only SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``C+T-stringent''), and one that maximizes the AUC (``C+T-max'') for 102 thresholds between $1$ and $10^{-100}$ (Table \ref{tab:thr}).
+We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``C+T-all''), one including only the SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``C+T-stringent''), and one that maximizes the AUC (``C+T-max'') for 102 p-value thresholds between $1$ and $10^{-100}$ (Table \ref{tab:thr}).
 As we report the optimal threshold based on the test set, the AUC for ``C+T-max'' is an upper bound of the AUC for the C+T method.
@@
 For each SNP variable, we add two more variables coding for recessive and dominant effects: one variable is coded 1 if homozygous variant and 0 otherwise, and the other is coded 0 for homozygous referent and 1 otherwise.
-We refer to this method as ``PLR3'' in the rest of the paper.
+We then apply our PLR implementation to this dataset with 3 times as many variables as the initial one; we refer to this method as ``PLR3'' in the rest of the paper.
 
@@
 We compared penalized logistic regression (PLR) with the C+T method using simulations of scenario \textnumero1 (Table \ref{tab:simus}).
-
-When simulating a model with 30 causal SNPs and an heritability of 80\%, PLR provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T range between 83\% and 90\% (Figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-PRS}).
+When simulating a model with 30 causal SNPs and an heritability of 80\%, PLR provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\% for this setting (Figure \ref{fig:main-AUC-logit}).
 Moreover, PLR consistently provides higher predictive performance than C+T across all scenarios we considered, excepted in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).
-
 PLR provides particularly higher predictive performance than C+T when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region.
 In this situation, the mean AUC reaches 92.5\% for PLR and 84\% for ``C+T-max'' (Figure \ref{fig:main-AUC-logit}).
-
 Note that, for the simulations, we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).
@@
 In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can dramatically impact the predictive performance of C+T. 
-For example, in a model with 30 causal SNPs only, AUC ranges from less than 60\% when using all SNPs passing clumping to 90\% \emph{if} choosing the optimal p-value threshold (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
+For example, in a model with 30 causal SNPs, AUC ranges from less than 60\% when using all SNPs passing clumping to 90\% \emph{if} choosing the optimal p-value threshold (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
 
@@
 We also compared the two penalized logistic regressions in scenario \textnumero1: PLR versus PLR3 that uses additional features (variables) coding for recessive and dominant effects.
-Predictive performance of PLR3 are nearly as good as PLR when there are only addtive effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
+Predictive performance of PLR3 are nearly as good as PLR when there are additive effects only (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
 For model ``COMP'', PLR3 provides AUC values at least 3.5\% higher than PLR, excepted when there are 3000 causal SNPs.
@@
 Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360 (Table \ref{tab:results-celiac}, figure \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
-In terms of computation time, we show that PLR, while learning jointly on all SNPs at once and testing four different hyper-parameter values, is almost as fast as the C+T method (190 vs 130 seconds), and PLR3 takes less than twice as long as PLR (296 vs 190 seconds).
+In terms of computation time, we show that PLR, while learning jointly on all SNPs at once and testing four different values for hyper-parameter $\alpha$, is almost as fast as the C+T method (190 vs 130 seconds), and PLR3 takes less than twice as long as PLR (296 vs 190 seconds).
 
@@
 In order to capture interaction-type effects, we tested T-Trees, a method that is able to exploit SNP correlations and interactions thanks to special decision trees \cite[]{botta2014exploiting}. 
-However, predictive performance of T-Trees were consistently lower than with penalized logistic regression, even when simulating a model with dominant and interaction-type effects that T-Trees should be able to handle.
+However, predictive performance of T-Trees are consistently lower than with penalized logistic regression, even when simulating a model with dominant and interaction-type effects that T-Trees should be able to handle.
 

### 2018-10-02 paper/PRS.tex
submitted

@@
 
-\noindent$^\ast$To whom correspondence should be addressed.
+\noindent$^\ast$To whom correspondence should be addressed.\\
+
+\noindent Contacts: 
+\begin{itemize}
+\item \url{florian.prive@univ-grenoble-alpes.fr}
+\item \url{hugues.aschard@pasteur.fr}
+\item \url{michael.blum@univ-grenoble-alpes.fr}
+\end{itemize}
 
@@
 
-In conclusion, our study demonstrates that penalized logistic regression can achieve more discriminative polygenic risk scores, while being applicable to large-scale individual-level data thanks to the implementation we provide in the R package bigstatsr.\\
-
-\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}
+In conclusion, our study demonstrates that penalized logistic regression can achieve more discriminative polygenic risk scores, while being applicable to large-scale individual-level data thanks to the implementation we provide in the R package bigstatsr.
 }

### 2018-10-09 paper/PRS.tex
notebooks in figshare

@@
 Supplemental Data include a PDF with two sections of methods, two tables and ten figures. 
-Supplemental Data also include six HTML R notebooks including all code and results used in this paper, for reproducibility purposes.
+Supplemental Data also include six HTML R notebooks including all code and results used in this paper, for reproducibility purposes, and available at \url{https://figshare.com/articles/code/7178750}.
 

### 2018-10-11 stacked-PRS.R
stacked PRS

@@ -1,0 +1,87 @@
+library(bigsnpr)
+
+celiac <- snp_attach("backingfiles/celiacQC.rds")
+G <- celiac$genotypes
+CHR <- celiac$map$chromosome
+y <- celiac$fam$affection - 1L
+
+U <- predict(readRDS("backingfiles/PCA.rds"))
+
+set.seed(1)
+ind.train <- sample(nrow(G), 5e3)
+ind.test <- setdiff(rows_along(G), ind.train)
+
+system.time(
+  gwas <- big_univLogReg(G, y[ind.train], ind.train, covar.train = U[ind.train, ], 
+                         ncores = nb_cores())
+) # 140 sec
+
+
+THR_P <- seq(0, max(-predict(gwas)))
+THR_R2 <- c(0.01, 0.02, 0.05, 0.2, 0.5)
+
+library(foreach)
+test <- foreach(thr.r2 = THR_R2, .combine = "cbind") %do% {
+  ind.keep <- snp_clumping(G, CHR, ind.row = ind.train,
+                           S = abs(gwas$score), 
+                           thr.r2 = thr.r2,
+                           ncores = nb_cores())
+  prs <- snp_PRS(G, betas.keep = gwas$estim[ind.keep],
+                 ind.keep = ind.keep,
+                 lpS.keep = -predict(gwas)[ind.keep],
+                 thr.list = THR_P)
+}
+dim(test)
+aucs <- apply(test[ind.test, ], 2, AUC, target = y[ind.test])
+plot(aucs, pch = 20)
+max(aucs)
+
+plot(rep(THR_P, 5), aucs, pch = 20, col = rep(1:5, each = length(THR)), log = "x")
+
+glm <- glmnet::glmnet(test[ind.train, ], y[ind.train], alpha = 1e-6, family = "binomial",
+                      lower.limits = 0)
+pred <- predict(glm, test[ind.test, ])
+aucs.stacked <- apply(pred, 2, AUC, target = y[ind.test])
+plot(aucs.stacked[-1], pch = 20)
+max(aucs.stacked)
+
+
+
+prs.train <- snp_PRS(G, betas.keep = gwas$estim[ind.keep],
+                     ind.test = ind.train, 
+                     ind.keep = ind.keep,
+                     lpS.keep = -predict(gwas)[ind.keep],
+                     thr.list = THR)
[…]

### 2018-10-11 paper/PRS-supp.tex

@@
 
-We used three different ways to estimate the maximum achievable AUC for our simulations.
+We use three different ways to estimate the maximum achievable AUC for simulations (see supplementary notebook ``oracle'').
 
-First, we used the estimation from equation (3) of \cite{wray2010genetic}.
+First, we use the estimation from equation (3) of \cite{wray2010genetic}.
 For a prevalence fixed at 30\% and an heritability of 50\% (respectively 80\%), the approximated theoretical values of AUC are 84.1\% (respectively 93.0\%).
@@
 
-Secondly, if we assume that the genetic part of the liabilities follows a Gaussian distribution $N(0, h^2)$ and that the environmental part follows a Gaussian distribution $N(0, 1 - h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the disease heritability $h^2$ through Monte Carlo simulations.
-We report AUCs of 84.1\% and 94.1\% for an heritability of 50\% and 80\%, respectively.
+Secondly, if we assume that the genetic part of the liabilities follows a Gaussian distribution $N(0, h^2)$ and that the environmental part follows a Gaussian distribution $N(0, 1 - h^2)$, we can estimate the theoretical value of the AUC that can be achieved given the disease heritability $h^2$ (and prevalence $K$) through Monte Carlo simulations.
+We report AUCs of 84.1\% and 94.1\% for heritabilities of 50\% and 80\%, respectively.
 
-Thirdly, we reproduce the exact same procedure of simulations and, for each combination of parameters (Table 2), we estimate the AUC of the ``oracle'', i.e.\ the true simulated genetic part of the liabilities through 100 replicates.
-For every combination of parameters, AUC of oracles are comprised between 83.2\% and 84.2\% for an heritability of 50\% and between 93.2\% and 94.1\% for an heritability of 80\%.
+Thirdly, we reproduce the exact same procedure of simulations and, for each combination of parameters (Table 2), we estimate the AUC of the ``oracle'', i.e.\ the true simulated genetic part of the liabilities, through 100 replicates.
+For every combination of parameters, AUC values of oracles vary between 83.2\% and 84.2\% for an heritability of 50\% and between 93.2\% and 94.1\% for an heritability of 80\%.
 
@@
 \caption{Comparison of C+T and PLR in scenario \textnumero1 for model ``ADD'' and an heritability of 50\%.
-Mean of AUC over 100 simulations for PLR and the maximum AUC reported with the C+T method (``C+T-max'').
+Mean AUC over 100 simulations for PLR and the maximum AUC reported with the C+T method (``C+T-max'').
 Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
 The blue dotted line represents the maximum achievable AUC.}
@@
 \caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for model ``ADD'' and an heritability of 50\%.
-Mean of AUC over 100 simulations.
+Mean AUC over 100 simulations.
 Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
 The blue dotted line represents the maximum achievable AUC.}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-all-r2}}
-\caption{Comparison of models in scenario \textnumero1 for model ``ADD''.
-Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the number and location of causal SNPs.
+\caption{Comparison of methods in scenario \textnumero1 for model ``ADD''.
+Mean AUC over 100 simulations for PLR and the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) as a function of the number and location of causal SNPs.
 Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
 The blue dotted line represents the maximum achievable AUC.}
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
-\caption{Comparison of models in scenario \textnumero2 (using chromosome 6 only) for model ``ADD''.
+\caption{Comparison of methods in scenario \textnumero2 (using chromosome 6 only) for model ``ADD''.
 Thinner lines represents results in scenario \textnumero1 (Figure \ref{fig:supp-AUC-all-r2}).
-Mean of AUC over 100 simulations for the maximum values of PRS for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the number and location of causal SNPs.
+Mean AUC over 100 simulations for PLR and the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) as a function of the number and location of causal SNPs.
 Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
 The blue dotted line represents the maximum achievable AUC.}
@@
 Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes.
-\textbf{A:} Mean of AUC over 5 simulations.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+\textbf{A:} Mean AUC over 5 simulations.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
 The blue dotted line represents the maximum achievable AUC.
@@
 Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes.
-\textbf{A:} Mean of AUC over 100 simulations.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+\textbf{A:} Mean AUC over 100 simulations.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
 The blue dotted line represents the maximum achievable AUC.
@@
 Horizontal panels are presenting results for models ``ADD'' and ``COMP'' for simulating phenotypes.
-\textbf{A:} Mean of AUC over 100 simulations.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+\textbf{A:} Mean AUC over 100 simulations.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
 The blue dotted line represents the maximum achievable AUC.}
@@
 Vertical panels are presenting results for effects following a Gaussian or Laplace distribution.
-\textbf{A:} Mean of AUC over 100 simulations.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC.
+\textbf{A:} Mean AUC over 100 simulations.
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
 The blue dotted line represents the maximum achievable AUC.

### 2018-12-04 paper/PRS.tex
typos

@@
 It compares all methods described in section \ref{sec:methods}.
-For each combination of the remaining parameters, results are based on 100 simulations excepted when comparing PLR with T-Trees, which relies on 5 simulations only because of a much higher computational burden of T-Trees as compared to other methods.
+For each combination of the remaining parameters, results are based on 100 simulations except when comparing PLR with T-Trees, which relies on 5 simulations only because of a much higher computational burden of T-Trees as compared to other methods.
 Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only (18,941 SNPs).
@@
 
-We use Monte Carlo cross-validation to compute AUC, partial AUC, the number of predictors and execution time for the original Celiac dataset with the observed case-control status: we randomly split 100 times the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals.
+We use Monte Carlo cross-validation to compute AUC, partial AUC, the number of predictors and execution time for the original Celiac dataset with the observed case-control status: we randomly split 100 times the dataset in a training set of 12,000 individuals and a test set composed of the remaining 3155 individuals.
 
@@
 When simulating a model with 30 causal SNPs and an heritability of 80\%, PLR provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\% for this setting (Figure \ref{fig:main-AUC-logit}).
-Moreover, PLR consistently provides higher predictive performance than C+T across all scenarios we considered, excepted in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).
+Moreover, PLR consistently provides higher predictive performance than C+T across all scenarios we considered, except in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).
 PLR provides particularly higher predictive performance than C+T when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region.
@@
 Predictive performance of PLR3 are nearly as good as PLR when there are additive effects only (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
-For model ``COMP'', PLR3 provides AUC values at least 3.5\% higher than PLR, excepted when there are 3000 causal SNPs.
+For model ``COMP'', PLR3 provides AUC values at least 3.5\% higher than PLR, except when there are 3000 causal SNPs.
 Yet, PLR3 takes 2-3 times as much time to run and requires 3 times as much disk storage as PLR.
@@
 In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently.
-PLR always outperform the C+T method, excepted in some highly underpowered cases, and the benefits of using PLR are more pronounced with an increasing sample size or when causal SNPs are correlated with one another.
+PLR always outperforms the C+T method, except in some highly underpowered cases, and the benefits of using PLR are more pronounced with an increasing sample size or when causal SNPs are correlated with one another.
 
@@
 
-On the contrary, for the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$ that accounts for the amount of regularization used in the model.
+In contrast, for the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$ that accounts for the amount of regularization used in the model.
 Not only this procedure provides near-optimal results, but it also accelerates the model training thanks to the development of an early stopping criterion.
@@
 Mean AUC over 100 simulations for PLR and the maximum AUC reported with ``C+T-max''.
-Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
+Upper (lower) panel is presenting results for effects following a Gaussian (Laplace) distribution.
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
@@
 Mean AUC over 100 simulations.
-Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
+Upper (lower) panel is presenting results for effects following a Gaussian (Laplace) distribution.
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
@@
 Mean AUC over 100 simulations for the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the training size.
-Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
+Upper (lower) panels are presenting results for effects following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.

### 2018-12-04 paper/PRS.tex
test sparSNP

@@
 
-In this paper, we present an efficient method to jointly estimate SNP effects, allowing for practical application of penalized logistic regression (PLR) on modern datasets including hundreds of thousands of individuals.
+In this paper, we present an efficient method to jointly estimate SNP effects using individual-level data, allowing for practical application of penalized logistic regression (PLR) on modern datasets including hundreds of thousands of individuals.
 Moreover, our implementation of PLR directly includes automatic choices for hyper-parameters.
-The choice of hyper-parameters for a predictive model is very important since it can dramatically impact its predictive performance.
-As an example, AUC values range from less than 60\% to 90\% in a model with 30 causal SNPs, depending on the p-value threshold in C+T.
 
@@
 
-Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function \[L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,\] where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
+Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function \[L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(z_i\right) + (1 - y_i) \log\left(1 - z_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,\] where $z_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
 Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);

### 2018-12-04 test-sparSNP.R

@@ -1,0 +1,26 @@
+library(bigsnpr)
+celiac <- snp_attach("backingfiles/celiacQC.rds")
+G <- celiac$genotypes
+CHR <- celiac$map$chromosome
+POS <- celiac$map$physical.pos
+NCORES <- nb_cores()
+y <- celiac$fam$affection - 1
+
+dim(G)
+set.seed(1)
+ind.train <- sort(sample(nrow(G), size = 12e3))
+ind.test <- setdiff(rows_along(G), ind.train)
+
+snp_writeBed(celiac, "backingfiles/celiac_train.bed", ind.row = ind.train)
+snp_writeBed(celiac, "backingfiles/celiac_test.bed", ind.row = ind.test)
+
+Sys.setenv(PATH = paste0("~/Bureau/SparSNP:", Sys.getenv("PATH")))
+path <- setwd("~/Bureau/SparSNP")
+unlink("discovery", recursive = TRUE)
+system.time(
+  system(sprintf("./crossval.sh %s sqrhinge",
+                 file.path(path, "backingfiles/celiac_train")))
+)
+
+
+       

### 2018-12-19 sparSNP.R
sparSNP

@@ -12,15 +12,18 @@ ind.train <- sort(sample(nrow(G), size = 12e3))
 ind.test <- setdiff(rows_along(G), ind.train)
 
 snp_writeBed(celiac, "backingfiles/celiac_train.bed", ind.row = ind.train)
-snp_writeBed(celiac, "backingfiles/celiac_test.bed", ind.row = ind.test)
+snp_writeBed(celiac, "backingfiles/celiac_test.bed",  ind.row = ind.test)
 
 Sys.setenv(PATH = paste0("~/Bureau/SparSNP:", Sys.getenv("PATH")))
 path <- setwd("~/Bureau/SparSNP")
 unlink("discovery", recursive = TRUE)
+
+# With NFOLDS=5 & NREPS=1
 system.time(
   system(sprintf("./crossval.sh %s sqrhinge",
                  file.path(path, "backingfiles/celiac_train")))
-)
+) # 3.2 h
 
-
-       
+system.time(
+  logit <- big_spLogReg(G, y[ind.train], ind.train, K = 5, ncores = 1)
+) # 38 sec  ~ 300 fold

### 2018-12-19 lassosum.R
lassosum

@@ -1,0 +1,53 @@
+# devtools::install_github("tshmak/lassosum")
+library(bigsnpr)
+celiac <- snp_attach("backingfiles/celiacQC.rds")
+G <- celiac$genotypes
+CHR <- celiac$map$chromosome
+POS <- celiac$map$physical.pos
+A1 <- celiac$map$allele1
+A2 <- celiac$map$allele2
+NCORES <- nb_cores()
+y <- celiac$fam$affection - 1
+
+dim(G)
+set.seed(1)
+ind.train <- sort(sample(nrow(G), size = 12e3))
+ind.test <- setdiff(rows_along(G), ind.train)
+ind.test.split <- split(ind.test, sample(1:2, length(ind.test), TRUE))
+
+U <- predict(readRDS("backingfiles/PCA.rds"))
+
+system.time(
+  gwas <- big_univLogReg(G, y[ind.train], ind.train,
+                         covar.train = U[ind.train, ],
+                         ncores = nb_cores())
+) # 140 sec
+hist(pval <- predict(gwas, log10 = FALSE))
+
+library(lassosum)
+snp_writeBed(celiac, "backingfiles/celiac_test1.bed",
+             ind.row = ind.test.split[[1]])
+snp_writeBed(celiac, "backingfiles/celiac_test2.bed",
+             ind.row = ind.test.split[[2]])
+cor <- p2cor(p = pval, n = 12e3, sign = gwas$estim)
+cor <- ifelse(is.na(cor), max(abs(cor), na.rm = TRUE) * sign(gwas$estim), cor)
+library(doParallel)
+registerDoParallel(cl <- makeCluster(nb_cores()))
+system.time(
+  out <- lassosum.pipeline(cor = cor, chr = CHR, pos = POS, A1 = A1, A2 = A2,
+                           test.bfile = "backingfiles/celiac_test1",
+                           LDblocks = "EUR.hg19",
+                           cluster = cl)
+) # 64 sec
+stopCluster(cl)
+
+v <- validate(out, covar = U[ind.test.split[[1]], ])
+out2 <- subset(out, s = v$best.s, lambda = v$best.lambda)
+v2 <- validate(out2, covar = U[ind.test.split[[2]], ],
+               test.bfile = "backingfiles/celiac_test2")
+
+AUC(v2$pgs[[1]], v2$pheno - 1)  ## 0.204
+AUC(v2$best.pgs, v2$pheno - 1)  ## 0.205
+
+
+

### 2018-12-19 BOLT-LMM.R
bolt lmm

@@ -1,0 +1,33 @@
+library(bigsnpr)
+celiac <- snp_attach("backingfiles/celiacQC.rds")
+G <- celiac$genotypes
+CHR <- celiac$map$chromosome
+POS <- celiac$map$physical.pos
+NCORES <- nb_cores()
+y <- celiac$fam$affection - 1
+
+dim(G)
+set.seed(1)
+ind.train <- sort(sample(nrow(G), size = 12e3))
+ind.test <- setdiff(rows_along(G), ind.train)
+
+snp_writeBed(celiac, "backingfiles/celiac_train.bed", ind.row = ind.train)
+snp_writeBed(celiac, "backingfiles/celiac_test.bed",  ind.row = ind.test)
+
+system("BOLT-LMM_v2.3.2/bolt -h")
+system(paste(
+  "BOLT-LMM_v2.3.2/bolt", 
+  "--bfile backingfiles/celiac_train", 
+  "--phenoUseFam",
+  "--lmm",
+  "--LDscoresFile=BOLT-LMM_v2.3.2/tables/LDSCORE.1000G_EUR.tab.gz",
+  "--statsFile res.txt",
+  "--numThreads=6"
+)) ## 55 min
+
+gwas <- bigreadr::fread2("res.txt")
+all(gwas$BETA != 0)  # TRUE
+hist(pval_lmm <- as.numeric(gwas$P_BOLT_LMM))
+
+preds <- big_prodVec(G, -gwas$BETA, ind.test)
+AUC(preds, y[ind.test])   # 0.813

### 2018-12-19 paper/PRS.tex
minor

@@
 However, C+T also has several limitations; for instance, previous studies have shown that predictive performance of C+T is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
-Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}.
+Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{accuheight2018,yang2010common}.
 Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes \cite[]{abraham2013performance}.
@@
 Overall, we find that PLR consistently achieves higher predictive performance than the C+T and T-Trees methods while being as fast as C+T.
-This demonstrates the feasibility and relevance of this approach for PRS computation on large modern datasets.
+This demonstrates the feasibility and relevance of this approach for PRS computation using large individual-level datasets such as the UK Biobank.
 
@@
 As we report the optimal threshold based on the test set, the AUC for ``C+T-max'' is an upper bound of the AUC for the C+T method.
+For the computation of this paper, the GWAS part is done on the training set while clumping is done on the test set (all individuals not included in the training set).
 

### 2018-12-29 lassosum.R
verif with bigstatsr

@@ -28,10 +28,10 @@ unlink("backingfiles/celiac_test*")
 celiac$fam$affection <- y
 snp_writeBed(celiac, "backingfiles/celiac_test1.bed",
              ind.row = ind.test.split[[1]])
-celiac$fam$affection[] <- 9999      ## just to be sure that this is not used
-snp_writeBed(celiac, "backingfiles/celiac_test2.bed",
-             ind.row = ind.test.split[[2]])
-cor <- p2cor(p = pval, n = 12e3, sign = gwas$estim)
+# celiac$fam$affection[] <- 9999      ## just to be sure that this is not used
+# snp_writeBed(celiac, "backingfiles/celiac_test2.bed",
+#              ind.row = ind.test.split[[2]])
+cor <- p2cor(p = pval, n = length(ind.train), sign = gwas$estim)
 cor <- ifelse(is.na(cor), max(abs(cor), na.rm = TRUE) * sign(gwas$estim), cor)
 library(doParallel)
 registerDoParallel(cl <- makeCluster(nb_cores()))
@@ -39,16 +39,18 @@ system.time(
   out <- lassosum.pipeline(cor = cor, chr = CHR, pos = POS, A1 = A2, A2 = A1,
                            test.bfile = "backingfiles/celiac_test1",
                            LDblocks = "EUR.hg19",
-                           cluster = cl)
+                           cluster = cl,
+                           exclude.ambiguous = FALSE)
 ) # 64 sec
 stopCluster(cl)
 
 v <- validate(out, covar = U[ind.test.split[[1]], ])
-out2 <- subset(out, s = v$best.s, lambda = v$best.lambda)
-v2 <- validate(out2, covar = U[ind.test.split[[2]], ],
-               test.bfile = "backingfiles/celiac_test2")
+length(ind <- which(v$best.beta != 0))
+pred <- big_prodVec(G, -v$best.beta[ind], 
+                    ind.row = ind.test.split[[2]],  
+                    ind.col = ind)
 
-AUC(v2$best.pgs, y[ind.test.split[[2]]])  ## 83.34
+AUC(pred, y[ind.test.split[[2]]])  ## 83.2 [81.1-85.3]
 
 
 

### 2018-12-29 paper/PRS.tex
some modifs (not finished)

@@
 However, C+T also has several limitations; for instance, previous studies have shown that predictive performance of C+T is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
-Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{accuheight2018,yang2010common}.
-Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes \cite[]{abraham2013performance}.
-Moreover, these methods and their derivatives are often computationally very demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
+Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{yang2010common}.
+Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes (see table 2 of \cite{abraham2013performance}), maybe because they were not primarily designed for this purpose.
+%Moreover, these methods and their derivatives are often computationally very demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
 Finally, statistical learning methods have also been used to derive PRS for complex human diseases by jointly estimating SNP effects.
@@
 We also compare two different models for simulating phenotypes, one with additive effects only, and one that combines additive, dominant and interaction-type effects.
-Overall, we find that PLR consistently achieves higher predictive performance than the C+T and T-Trees methods while being as fast as C+T.
-This demonstrates the feasibility and relevance of this approach for PRS computation using large individual-level datasets such as the UK Biobank.
+Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being as fast.
+This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some well power studies in return.
 
@@
 In terms of computation time, we show that PLR, while learning jointly on all SNPs at once and testing four different values for hyper-parameter $\alpha$, is almost as fast as the C+T method (190 vs 130 seconds), and PLR3 takes less than twice as long as PLR (296 vs 190 seconds).
+Finally, on this Celiac dataset, our implementation of PLR runs 300 times as fast as SparSNP \cite[]{abraham2012sparsnp}.
 
@@
 
+\subsection{Penalized regression for the UK Biobank}
+
+We tested our implementation on 656K genotyped SNPs of the UK Biobank, keeping only caucasian individuals and removing individuals related to someone else in the data (second individual in each pair with a kinship greater than 0.08).
+Our L1-penalized linear regression runs in less than one day for 350K individuals, achieving a correlation of 65.5\% with true height for both sexes.
+The best C+T model achieves a correlation of 55\% for women and 56\% for men, where the GWAS part takes 1 hour only.
+Our L1-penalized logistic regression on breast cancer runs in only 12 minutes for 150K women, achieving an AUC of 0.587.
+The best C+T model achieves the same AUC score, where the GWAS part takes 15 hours.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section{Discussion}
+
+\subsection{Time and memory requirements}
+
+The computation time of our PLR implementation mainly depends on the sample size and the number of active variables.
+Indeed, the algorithm is composed of two steps: first, for each variable, the algorithm computes one statistic that will be used to decide if the variable enters the model or not.
+Then, the algorithm iterates over a regularization path of decreasing values of $\lambda$, which progressively enables variables to enter the model (Figure  \ref{fig:CMSA}).
+So, in the second step, the number of variables is increasing progressively, and at some point the computation stops due to the early stopping criterion (prediction is getting worse on the corresponding validation set, see figure \ref{fig:CMSA}).
+
+So, for highly polygenic traits such as height and when using huge datasets such as the UK Biobank, the algorithm might iterate over dozens of thousands of variables, which takes time (but still less than one day with our implementation).
+On the contrary, for traits like the celiac disease or breast cancer that seems to be less polygenic, a lot less variables are entering the model and the fitting is very fast (12 minutes only for 150K women of the UK Biobank for breast cancer!).
+Using more cases (there were only 3542 for Celiac and 6374 for Breast Cancer), one might achieve better predictive performance for these two diseases, and the computation time should stay within a few hours.
+
+Regarding memory requirements, this is basically the same as for computation time.
+Indeed, active variables are accessed in memory thanks to memory-mapping when they are used \cite[]{prive2017efficient}.
+If there are too many variables accessed, the operating system (OS) has to make some space for new comping variables.
+So, there should never be any memory issue thanks to memory-mapping, yet, at some point, if there are too many active variables, the OS would regularly swap memory between disk and RAM, which is very slow.
+If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM.
+We have quickly tested processing each chromosome separately and then combining the models later but this did not seem to perform as well as fitting all SNPs at once.
 
@@
 In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently.
-PLR always outperforms the C+T method, except in some highly underpowered cases, and the benefits of using PLR are more pronounced with an increasing sample size or when causal SNPs are correlated with one another.
+PLR always outperforms the C+T method, except in some highly underpowered cases (AUC values always less than 0.6), and the benefits of using PLR are more pronounced with an increasing sample size or when causal SNPs are correlated with one another (Figures \ref{fig:supp-AUC-chr6-all-r2} and \ref{fig:main-AUC-ntrain}).
+This could be explained by the fact that PLR is too conservative in underpowered cases.
 
@@
 Based on the simulations, we find that using a stringent threshold ($r^2 = 0.05$) leads to higher predictive performance, even when causal SNPs are correlated.
-It means that, in most cases, avoiding redundant information is more important than including all causal SNPs.
+It means that, in most cases tested in this paper, avoiding redundant information is more important than including all causal SNPs.
 The choice of the $p_T$ threshold is also very important as it can greatly impact the predictive performance of the C+T method, which we confirm in this study \cite[]{ware2017heterogeneity}. 
@@
 To our knowledge, there is no clear standard on how to choose these two critical hyper-parameters for C+T.
+We did not want to choose the C+T parameters so that one could not criticize the procedure we would have used.
+Instead, for C+T, we chose to report the best AUC value on the test set, even if it leads to over-optimistic results for C+T as compared to PLR.
 
@@
 Our approach has one major limitation: the main advantage of the C+T method is its direct applicability to summary statistics, allowing to leverage the largest GWAS results to date, even when individual cohort data cannot be merged because of practical or ethical reasons (e.g.\ consortium data including many cohorts).
-As of today, the proposed penalized logistic regression does not allow for the analysis of summary data, but this represents an important future direction of our work.
+As of today, our penalized logistic regression does not allow for the analysis of summary data, but this represents an important future direction of our work.
 The current version is of particular interest for the analysis of modern individual-level datasets including hundreds of thousands of individuals.
@@
 
-We also demonstrated in simulations that our implementation of penalized regressions remains highly effective over a broad range of disease architectures.
+We also demonstrated in simulations and real datasets that our implementation of penalized regressions is highly effective over a broad range of disease architectures.
 It can be appropriate for predicting autoimmune diseases with a few strong effects (e.g.\ celiac disease) as well as highly polygenic traits (e.g.\ standing height).
@@
 Authors also acknowledge the Grenoble Alpes Data Institute that is supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02).
+We used the UK Biobank resource (application ID 25589).
 We are also grateful to F\'elix Balazard for useful discussions about T-Trees, and to Yaohui Zeng for useful discussions about R package biglasso.
+Finally, we are grateful to the two anonymous reviewers who contributed to improving this paper.
 

### 2018-12-29 paper/PRS.tex
add subsetting discu + UKB code

@@
 If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM. 
-We have quickly tested processing each chromosome separately and then combining the models later but this did not seem to perform as well as fitting all SNPs at once.
+We have quickly tested processing each chromosome separately and then combining the models later but this did not seem to perform as well as fitting all SNPs at once (data not shown). 
+Another widely used approach is to apply the penalized regression on a subset of SNPs using univariate tests (GWAS) to prioritize SNPs.
+Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm here by achieving a correlation of 65.5\% for height using 350K individuals of the UK Biobank, whereas \cite{accuheight2018} achieved only 61\% for the same sample size when using only 50K SNPs (see figure 3 of their paper).
 
@@
 Supplemental Data also include six HTML R notebooks including all code and results used in this paper, for reproducibility purposes, and available at \url{https://figshare.com/articles/code/7178750}.
+Additional analyses of the UK Biobank are available as 3 R scripts at \url{https://figshare.com/articles/code_UKB/7531559}.
 

### 2019-01-01 paper/PRS.tex
save

@@
 
+Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being as fast.
+This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some well power studies in return.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.
+Note that these AUC figures are also dependent on the prevalence of the disease.
 
@@
 Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{yang2010common}.
-Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes (see table 2 of \cite{abraham2013performance}), maybe because they were not primarily designed for this purpose.
+Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes (see table 2 of \cite{abraham2013performance}).
 %Moreover, these methods and their derivatives are often computationally very demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.

### 2019-01-09 paper/PRS.tex
save

@@
 As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}.
-Yet, PRS would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases.
+Yet, PRS would have to show a high discriminative power between cases and controls in order to be used in clinical settings.
 For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.
@@
 When simulating a model with 30 causal SNPs and an heritability of 80\%, PLR provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\% for this setting (Figure \ref{fig:main-AUC-logit}).
-Moreover, PLR consistently provides higher predictive performance than C+T across all scenarios we considered, except in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).
+Moreover, PLR consistently provides higher predictive performance than C+T across all scenarios we considered, except in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-ntrain}).
 PLR provides particularly higher predictive performance than C+T when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region.
@@
 Concerning the $r^2$ threshold of the clumping step in C+T, we mostly used the common value of $0.2$.
-Yet, using a more stringent value of $0.05$ provides higher predictive performance than using $0.2$ in most of the cases we considered (Figures \ref{fig:supp-AUC-all-r2}, \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-chr6-all-r2})
+Yet, using a more stringent value of $0.05$ provides higher predictive performance than using $0.2$ in most of the cases we considered (Figures \ref{fig:supp-AUC-all-r2} and \ref{fig:main-AUC-ntrain})
 
@@
 
-First, when reproducing simulations of scenario \textnumero1 using chromosome 6 only (scenario \textnumero2), the predictive performance of PLR always increase (Figure \ref{fig:supp-AUC-chr6-all-r2}).
+First, when reproducing simulations of scenario \textnumero1 using chromosome 6 only (scenario \textnumero2), the predictive performance of PLR always increase (Figure \ref{fig:supp-AUC-all-r2}). 
 There is a particularly large increase when simulating 3000 causal SNPs: AUC from PLR increases from 60\% to nearly 80\% for Gaussian effects and a disease heritability of 80\%.
-On the contrary, when simulating only 30 or 300 causal SNPs with the corresponding dataset, AUC of ``C+T-max'' does not increase, and even decreases for an heritability of 80\% (Figure \ref{fig:supp-AUC-chr6-all-r2}).
+On the contrary, when simulating only 30 or 300 causal SNPs with the corresponding dataset, AUC of ``C+T-max'' does not increase, and even decreases for an heritability of 80\% (Figure \ref{fig:supp-AUC-all-r2}).
 Secondly, when varying the training size (scenario \textnumero3), we report an increase of AUC with a larger training size, with a faster increase of AUC for PLR as compared to ``C+T-max'' (Figure  \ref{fig:main-AUC-ntrain}).
@@
 In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently.
-PLR always outperforms the C+T method, except in some highly underpowered cases (AUC values always less than 0.6), and the benefits of using PLR are more pronounced with an increasing sample size or when causal SNPs are correlated with one another (Figures \ref{fig:supp-AUC-chr6-all-r2} and \ref{fig:main-AUC-ntrain}).
-This could be explained by the fact that PLR is too conservative in underpowered cases.
+PLR always outperforms the C+T method, except in some highly underpowered cases (AUC values always less than 0.6), and the benefits of using PLR are more pronounced with an increasing sample size or when causal SNPs are correlated with one another.
+
+When there are many effects of small size and a small sample size, PLR performs worse than (the best result for) C+T.
+We think this is because PLR is too conservative in including effects, whereas it has been shown in the literature that in such underpowered scenarios, including (nearly) all SNPs was the best strategy \cite[]{purcell2009common}.
+In scenario \textnumero3, by varying sample size, we showed that with increasing sample size, PLR would achieve higher predictive performance than C+T at some point and that the difference between the two methods would increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}).
+Moreover, in scenario \textnumero2, we decrease the search space (p: number of SNPs) so that it could increase the polygenicity of the models and virtually increase the sample size (because power is dependent on n/p).
+In this scenario, even there are small effects and a high polygenicity (3000 causal variants out of the ~18K SNPs of chromosome 6), PLR gets a large increase in predictive performance, consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
 
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
-\caption{Main comparison of C+T and PLR in scenario \textnumero1 for model ``ADD'' and an heritability of 80\%.
+\caption{Main comparison of C+T and PLR in scenario \textnumero1 for model ``ADD''.
 Mean AUC over 100 simulations for PLR and the maximum AUC reported with ``C+T-max''.
-Upper (lower) panel is presenting results for effects following a Gaussian (Laplace) distribution.
+Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
@@
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{main-AUC-PRS}}
-\caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for model ``ADD'' and an heritability of 80\%.
-Mean AUC over 100 simulations.
-Upper (lower) panel is presenting results for effects following a Gaussian (Laplace) distribution.
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
+\caption{Comparison of methods in scenario \textnumero2 (using chromosome 6 only) for model ``ADD''.
+Thinner lines represents results in scenario \textnumero1.
+Mean AUC over 100 simulations for PLR and the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) as a function of the number and location of causal SNPs.
+Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
 The blue dotted line represents the maximum achievable AUC.}
-\label{fig:main-AUC-PRS}
+\label{fig:supp-AUC-all-r2}
 \end{figure}

### 2019-01-10 4-results-simus.Rmd
merge figures

@@
 
-```{r, fig.asp = 1, out.width = "60%", fig.width=8}
+```{r}
 results3.summary %>%
   filter(method %in% c("PLR", "C+T-max"), par.model == "ADD") %>%
-  barplot_causal_one(h2 = 0.8)
+  barplot_causal_all()
 
-ggsave("figures/main-AUC-logit.pdf", scale = 1/90, width = 844, height = 872)
+ggsave("figures/main-AUC-logit.pdf", scale = 1/110, width = 1242, height = 951)
 ```
 
-```{r, fig.asp = 1, out.width = "60%", fig.width=8}
+```{r}
 results3.summary %>%
   filter(grepl("C\\+T", method)) %>%
-  barplot_causal_one(h2 = 0.8)
+  barplot_causal_all()
 
-ggsave("figures/main-AUC-PRS.pdf", scale = 1/90, width = 844, height = 872)
-```
-
-```{r, fig.asp = 1, out.width = "60%", fig.width=8}
-results3.summary %>%
-  filter(method %in% c("PLR", "C+T-max")) %>%
-  barplot_causal_one(h2 = 0.5)
-
-ggsave("figures/supp-AUC-logit.pdf", scale = 1/90, width = 844, height = 872)
-```
-
-```{r, fig.asp = 1, out.width = "60%", fig.width=8}
-results3.summary %>%
-  filter(grepl("C\\+T", method)) %>%
-  barplot_causal_one(h2 = 0.5)
-
-ggsave("figures/supp-AUC-PRS.pdf", scale = 1/90, width = 844, height = 872)
+ggsave("figures/supp-AUC-PRS.pdf", scale = 1/110, width = 1242, height = 951)
 ```

### 2019-01-10 paper/PRS-supp.tex

@@
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-logit}}
-\caption{Comparison of C+T and PLR in scenario \textnumero1 for model ``ADD'' and an heritability of 50\%.
-Mean AUC over 100 simulations for PLR and the maximum AUC reported with the C+T method (``C+T-max'').
-Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
-The blue dotted line represents the maximum achievable AUC.}
-\label{fig:supp-AUC-logit}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-PRS}}
-\caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for model ``ADD'' and an heritability of 50\%.
+\caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for model ``ADD''.
 Mean AUC over 100 simulations.
-Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution.
+Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
 Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
@@
 \label{fig:supp-AUC-PRS}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-all-r2}}
-\caption{Comparison of methods in scenario \textnumero1 for model ``ADD''.
-Mean AUC over 100 simulations for PLR and the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) as a function of the number and location of causal SNPs.
-Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
-The blue dotted line represents the maximum achievable AUC.}
-\label{fig:supp-AUC-all-r2}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
-\caption{Comparison of methods in scenario \textnumero2 (using chromosome 6 only) for model ``ADD''.
-Thinner lines represents results in scenario \textnumero1 (Figure \ref{fig:supp-AUC-all-r2}).
-Mean AUC over 100 simulations for PLR and the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) as a function of the number and location of causal SNPs.
-Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
-The blue dotted line represents the maximum achievable AUC.}
-\label{fig:supp-AUC-chr6-all-r2}
 \end{figure}

### 2019-01-10 paper/PRS.tex

@@
 
+{\color{red}
 In this paper, we present an efficient method to jointly estimate SNP effects using individual-level data, allowing for practical application of penalized logistic regression (PLR) on modern datasets including hundreds of thousands of individuals.
 Moreover, our implementation of PLR directly includes automatic choices for hyper-parameters.
+}
 
@@
 
+{\color{red}
 Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being as fast.
 This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some well power studies in return.
+}
 
@@
 As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}.
+{\color{red}
 Yet, PRS would have to show a high discriminative power between cases and controls in order to be used in clinical settings.
@@
 Note that these AUC figures are also dependent on the prevalence of the disease.
+}
 
@@
 However, C+T also has several limitations; for instance, previous studies have shown that predictive performance of C+T is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
+{\color{red}
 Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{yang2010common}.
@@
 %Moreover, these methods and their derivatives are often computationally very demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
+}
 Finally, statistical learning methods have also been used to derive PRS for complex human diseases by jointly estimating SNP effects.
@@
 We also compare two different models for simulating phenotypes, one with additive effects only, and one that combines additive, dominant and interaction-type effects.
+{\color{red}
 Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being as fast.
 This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some well power studies in return.
+}
 
@@
 As we report the optimal threshold based on the test set, the AUC for ``C+T-max'' is an upper bound of the AUC for the C+T method.
+{\color{red}
 For the computation of this paper, the GWAS part is done on the training set while clumping is done on the test set (all individuals not included in the training set).
+}
 
@@
 In terms of computation time, we show that PLR, while learning jointly on all SNPs at once and testing four different values for hyper-parameter $\alpha$, is almost as fast as the C+T method (190 vs 130 seconds), and PLR3 takes less than twice as long as PLR (296 vs 190 seconds).
+{\color{red}
 Finally, on this Celiac dataset, our implementation of PLR runs 300 times as fast as SparSNP \cite[]{abraham2012sparsnp}.
+}
 
@@
 
+{\color{red}
 \subsection{Penalized regression for the UK Biobank}
@@
 
+}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+{\color{red}
 \subsection{Time and memory requirements}
@@
 
+}
+
 \subsection{Joint estimation improves predictive performance}
@@
 In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently.
+{\color{red}
 PLR always outperforms the C+T method, except in some highly underpowered cases (AUC values always less than 0.6), and the benefits of using PLR are more pronounced with an increasing sample size or when causal SNPs are correlated with one another.
@@
 In this scenario, even there are small effects and a high polygenicity (3000 causal variants out of the ~18K SNPs of chromosome 6), PLR gets a large increase in predictive performance, consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
+}
 
@@
 To our knowledge, there is no clear standard on how to choose these two critical hyper-parameters for C+T.
+{\color{red}
 We did not want to choose the C+T parameters so that one could not criticize the procedure we would have used.
 Instead, for C+T, we chose to report the best AUC value on the test set, even if it leads to over-optimistic results for C+T as compared to PLR.
+}
 
@@
 Our approach has one major limitation: the main advantage of the C+T method is its direct applicability to summary statistics, allowing to leverage the largest GWAS results to date, even when individual cohort data cannot be merged because of practical or ethical reasons (e.g.\ consortium data including many cohorts).
-As of today, our penalized logistic regression does not allow for the analysis of summary data, but this represents an important future direction of our work.
+As of today, our implementation of penalized logistic regression does not allow for the analysis of summary data, but this represents an important future direction of our work.
 The current version is of particular interest for the analysis of modern individual-level datasets including hundreds of thousands of individuals.

### 2019-01-10 paper/PRS.tex
save

@@
 However, C+T also has several limitations; for instance, previous studies have shown that predictive performance of C+T is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
-{\color{red}
 Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{yang2010common}.
@@
 %Moreover, these methods and their derivatives are often computationally very demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
-}
 Finally, statistical learning methods have also been used to derive PRS for complex human diseases by jointly estimating SNP effects.
@@
 Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being as fast.
-This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some well power studies in return.
+This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some high predictive performance in return.
 }
@@
 A total of 100 simulations per combination of parameters are run using 300 causal SNPs randomly chosen on the genome.
+
+\begin{table}[htbp]
+\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.\label{tab:simus}}
+\vspace*{0.5em}
+\centering
+\begin{tabular}{|c|c|c|c|c|c|c|c|}
+\hline
+Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
+scenario & & training set & (number and location) & of effects & & model & \\
+\hline
+\hline
+\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{ADD} & C+T \\
+& & & 30 in all & & & & PLR \\
+& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{COMP} & PLR3 \\
+& & & 3000 in all & & & & (T-Trees) \\
+\hline
+\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{ADD} & C+T \\ 
+& & & & & & & PLR\\
+\hline
+\multirow{5}{*}{3} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
+& & 2000 & & & & & \\
+& & 3000 & & & & & \\
+& & 4000 & & & & & \\
+& & 5000 & & & & & \\
+\hline
+\end{tabular}
+\end{table}
 
@@
 We compared penalized logistic regression (PLR) with the C+T method using simulations of scenario \textnumero1 (Table \ref{tab:simus}).
-When simulating a model with 30 causal SNPs and an heritability of 80\%, PLR provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\% for this setting (Figure \ref{fig:main-AUC-logit}).
+When simulating a model with 30 causal SNPs and a heritability of 80\%, PLR provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\% for this setting (Figure \ref{fig:main-AUC-logit}).
 Moreover, PLR consistently provides higher predictive performance than C+T across all scenarios we considered, except in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-ntrain}).
@@
 
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
+\caption{Main comparison of C+T and PLR in scenario \textnumero1 for model ``ADD''.
+Mean AUC over 100 simulations for PLR and the maximum AUC reported with ``C+T-max'' (clumping threshold at $r^2 > 0.2$).
+Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for a heritability of 0.5 (0.8).
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
+The blue dotted line represents the maximum achievable AUC.}
+\label{fig:main-AUC-logit}
+\end{figure}
+
 \subsection{Importance of hyper-parameters}
@@
 In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can dramatically impact the predictive performance of C+T. 
-For example, in a model with 30 causal SNPs, AUC ranges from less than 60\% when using all SNPs passing clumping to 90\% \emph{if} choosing the optimal p-value threshold (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).
+For example, in a model with 30 causal SNPs, AUC ranges from less than 60\% when using all SNPs passing clumping to 90\% \emph{if} choosing the optimal p-value threshold (Figure \ref{fig:supp-AUC-PRS}).
 
@@
 There is a particularly large increase when simulating 3000 causal SNPs: AUC from PLR increases from 60\% to nearly 80\% for Gaussian effects and a disease heritability of 80\%.
-On the contrary, when simulating only 30 or 300 causal SNPs with the corresponding dataset, AUC of ``C+T-max'' does not increase, and even decreases for an heritability of 80\% (Figure \ref{fig:supp-AUC-all-r2}).
+On the contrary, when simulating only 30 or 300 causal SNPs with the corresponding dataset, AUC of ``C+T-max'' does not increase, and even decreases for a heritability of 80\% (Figure \ref{fig:supp-AUC-all-r2}).
 Secondly, when varying the training size (scenario \textnumero3), we report an increase of AUC with a larger training size, with a faster increase of AUC for PLR as compared to ``C+T-max'' (Figure  \ref{fig:main-AUC-ntrain}).
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
+\caption{Comparison of methods in scenario \textnumero2 (using chromosome 6 only) for model ``ADD''.
+Thinner lines represents results in scenario \textnumero1.
+Mean AUC over 100 simulations for PLR and the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) as a function of the number and location of causal SNPs.
+Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for a heritability of 0.5 (0.8).
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
+The blue dotted line represents the maximum achievable AUC.}
+\label{fig:supp-AUC-all-r2}
+\end{figure}
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
+\caption{Comparison of methods when varying sample size in scenario \textnumero3 for model ``ADD'' with 300 causal SNPs sampled anywhere on the genome.
+Mean AUC over 100 simulations for the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the training size.
+Upper (lower) panels are presenting results for effects following a Gaussian (Laplace) distribution and left (right) panels are presenting results for a heritability of 0.5 (0.8).
+Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
+The blue dotted line represents the maximum achievable AUC.}
+\label{fig:main-AUC-ntrain}
+\end{figure}
 
@@
 
-Joint logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with PLR and 89.1\% with PLR3 as compared to 82.5\% with ``C+T-max''.
+Joint logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with PLR and 89.1\% with PLR3 as compared to 82.5\% with ``C+T-max'' (Table \ref{tab:results-celiac} and figure \ref{fig:celiac-roc}).
 The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with PLR, PLR3 and ``C+T-max'', respectively.
-Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360 (Table \ref{tab:results-celiac}, figure \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
+Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360.
 In terms of computation time, we show that PLR, while learning jointly on all SNPs at once and testing four different values for hyper-parameter $\alpha$, is almost as fast as the C+T method (190 vs 130 seconds), and PLR3 takes less than twice as long as PLR (296 vs 190 seconds).
@@
 
+[TODO: RESULTS TABLE]
+
 }
@@
 Then, the algorithm iterates over a regularization path of decreasing values of $\lambda$, which progressively enables variables to enter the model (Figure  \ref{fig:CMSA}).
-So, in the second step, the number of variables is increasing progressively, and at some point the computation stops due to the early stopping criterion (prediction is getting worse on the corresponding validation set, see figure \ref{fig:CMSA}).
+So, in the second step, the number of variables is increasing progressively, and at some point the computation stops due to the early stopping criterion (when prediction is getting worse on the corresponding validation set, see figure \ref{fig:CMSA}).
 
@@
 On the contrary, for traits like the celiac disease or breast cancer that seems to be less polygenic, a lot less variables are entering the model and the fitting is very fast (12 minutes only for 150K women of the UK Biobank for breast cancer!).
-Using more cases (there were only 3542 for Celiac and 6374 for Breast Cancer), one might achieve better predictive performance for these two diseases, and the computation time should stay within a few hours.
+Using more cases (there were only 3542 for Celiac and 6374 for Breast Cancer), one might achieve better predictive performance for these two diseases, and the computation time should remain within a few hours.
 
@@
 Indeed, active variables are accessed in memory thanks to memory-mapping when they are used \cite[]{prive2017efficient}.
-If there are too many variables accessed, the operating system (OS) has to make some space for new comping variables. 
-So, there should never be any memory issue thanks to memory-mapping, yet, at some point, if there are too many active variables, the OS would regularly swap memory between disk and RAM, which is very slow.
+If there are too many variables accessed, the operating system (OS) has to make some space for new coming variables. 
+So, there should never be any memory issue thanks to memory-mapping.
+Yet, at some point, if there are too many active variables, the OS would regularly swap memory between disk and RAM, which is very slow.
 If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM. 
@@
 Another widely used approach is to apply the penalized regression on a subset of SNPs using univariate tests (GWAS) to prioritize SNPs.
-Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm here by achieving a correlation of 65.5\% for height using 350K individuals of the UK Biobank, whereas \cite{accuheight2018} achieved only 61\% for the same sample size when using only 50K SNPs (see figure 3 of their paper).
+Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm here by achieving a correlation of 65.5\% for height using 350K individuals of the UK Biobank, whereas \cite{accuheight2018} achieved only 61\% for the same sample size when using only the 50K most significantly associated SNPs (see figure 3 of their paper).
 
@@
 When there are many effects of small size and a small sample size, PLR performs worse than (the best result for) C+T.
-We think this is because PLR is too conservative in including effects, whereas it has been shown in the literature that in such underpowered scenarios, including (nearly) all SNPs was the best strategy \cite[]{purcell2009common}.
-In scenario \textnumero3, by varying sample size, we showed that with increasing sample size, PLR would achieve higher predictive performance than C+T at some point and that the difference between the two methods would increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}).
-Moreover, in scenario \textnumero2, we decrease the search space (p: number of SNPs) so that it could increase the polygenicity of the models and virtually increase the sample size (because power is dependent on n/p).
-In this scenario, even there are small effects and a high polygenicity (3000 causal variants out of the ~18K SNPs of chromosome 6), PLR gets a large increase in predictive performance, consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
+This may be because PLR is too conservative in including effects, whereas it has been shown in the literature that, in such underpowered scenarios, including (nearly) all SNPs was the best strategy \cite[]{purcell2009common}.
+In scenario \textnumero3, by varying sample size, we showed that when increasing sample size, PLR would achieve higher predictive performance than C+T at some point and that the benefits of using PLR over C+T would increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}).
+Moreover, in scenario \textnumero2, we decrease the search space (p: number of SNPs) in order to increase the polygenicity of the models and to virtually increase the sample size \cite[]{dudbridge2013power}.
+In this scenario, even when there are small effects and a high polygenicity (3000 causal variants out of 18,941), PLR gets a large increase in predictive performance, now consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
 }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-\begin{table}[htbp]
-\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.\label{tab:simus}}
-\vspace*{0.5em}
-\centering
-\begin{tabular}{|c|c|c|c|c|c|c|c|}
-\hline
-Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
-scenario & & training set & (number and location) & of effects & & model & \\
-\hline
-\hline
-\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{ADD} & C+T \\
-& & & 30 in all & & & & PLR \\
-& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{COMP} & PLR3 \\
-& & & 3000 in all & & & & (T-Trees) \\
-\hline
-\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{ADD} & C+T \\ 
-& & & & & & & PLR\\
-\hline
-\multirow{5}{*}{3} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
-& & 2000 & & & & & \\
-& & 3000 & & & & & \\
-& & 4000 & & & & & \\
-& & 5000 & & & & & \\
-\hline
-\end{tabular}
-\end{table}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
-\caption{Main comparison of C+T and PLR in scenario \textnumero1 for model ``ADD''.
-Mean AUC over 100 simulations for PLR and the maximum AUC reported with ``C+T-max''.
-Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
-The blue dotted line represents the maximum achievable AUC.}
-\label{fig:main-AUC-logit}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
-\caption{Comparison of methods in scenario \textnumero2 (using chromosome 6 only) for model ``ADD''.
-Thinner lines represents results in scenario \textnumero1.
-Mean AUC over 100 simulations for PLR and the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) as a function of the number and location of causal SNPs.
-Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
-The blue dotted line represents the maximum achievable AUC.}
-\label{fig:supp-AUC-all-r2}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
-\caption{Comparison of methods when varying sample size in scenario \textnumero3 for model ``ADD'' with 300 causal SNPs sampled anywhere on the genome.
-Mean AUC over 100 simulations for the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the training size.
-Upper (lower) panels are presenting results for effects following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8).
-Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC.
-The blue dotted line represents the maximum achievable AUC.}
-\label{fig:main-AUC-ntrain}
-\end{figure}
-
-\newpage
-\begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{celiac-roc}}
-\caption{ROC Curves for C+T, PLR and PLR3 for the celiac disease dataset.
-Models were trained using 12,000 individuals.
-These are results projecting these models on the remaining 3155 individuals.
-The figure is plotted using R package plotROC \cite[]{sachs2017plotroc}.\label{fig:celiac-roc}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2019-01-23 5-results-celiac.Rmd
use bigstatsr 0.8

@@
 )
-# Get K = 10 predictions/individuals
-preds1 <- predict(logit, X = G, ind.row = ind.test, 
+pred1 <- predict(logit, X = G, ind.row = ind.test, 
                   covar.row = obj.svd$u[ind.test, ])
-(nb.pred1 <- sum(rowSums(sapply(logit, function(l) l$beta.X != 0)) > 0))
-(aucs1 <- AUC(preds1, target = y[ind.test]))
+(nb.pred1 <- summary(logit, best.only = TRUE)$nb_var)
+(aucs1 <- AUC(pred1, target = y[ind.test]))
 ```
@@
 ```{r}
-(p2 <- plot_density_AUC(preds1))
+(p2 <- plot_density_AUC(pred1))
 ```
@@
 
-preds2 <- predict(logit2, X = G2, ind.row = ind.test, 
+pred2 <- predict(logit2, X = G2, ind.row = ind.test, 
                   covar.row = obj.svd$u[ind.test, ])
-(nb.pred2 <- sum(rowSums(sapply(logit2, function(l) l$beta.X != 0)) > 0))
-(aucs2 <- AUC(preds2, target = y[ind.test]))
+(nb.pred2 <- summary(logit2, best.only = TRUE)$nb_var)
+(aucs2 <- AUC(pred2, target = y[ind.test]))
 ```
@@
 ```{r}
-(p3 <- plot_density_AUC(preds2))
+(p3 <- plot_density_AUC(pred2))
 ```
@@
   "C+T-max" = prs[, which.max(aucs)],
-  "PLR"     = preds1,
-  "PLR3"    = preds2
+  "PLR"     = pred1,
+  "PLR3"    = pred2
 ) %>%

### 2019-01-23 pkg.paper.PRS/R/models.R

@@ -117,13 +117,15 @@ logit.CMSA <- function(G, pheno.all, covar.all, ind.train, ind.test, method,
     preds <- stats::predict(cmsa.logit, X = G, ind.row = ind.test,
                             covar.row = covar.all[ind.test, , drop = FALSE])
   })[3]
+  
+  summ_best <- summary(cmsa.logit, best.only = TRUE)
 
   tibble(
     method   = method,
     pred     = list(preds),
     timing   = timing,
-    alpha = attr(cmsa.logit, "alpha"),
-    set = list(which(rowSums(sapply(cmsa.logit, function(x) x$beta.X) != 0) > 0))
+    alpha = summ_best$alpha,
+    set = list(which(summ_best$beta[[1]] != 0))
   )
 }
 

### 2019-01-23 LDpred.R
save

@@ -1,3 +1,6 @@
+# system("pip2 install --user --upgrade pip")
+# system("pip2 install --user ldpred")
+
 library(bigsnpr)
 celiac <- snp_attach("backingfiles/celiacQC.rds")
 G <- celiac$genotypes
@@ -21,24 +24,42 @@ system.time(
                          ncores = nb_cores())
 ) # 140 sec
 hist(pval <- predict(gwas, log10 = FALSE))
+pval[pval == 0] <- .Machine$double.xmin
 
 library(dplyr)
 celiac$map %>%
   select(hg19chrc = chromosome, snpid = marker.ID,
          a1 = allele1, a2 = allele2, bp = physical.pos) %>%
-  mutate(or = exp(gwas$estim), p = pval) %>%
+  mutate(hg19chrc = paste0("chr", hg19chrc), or = exp(gwas$estim), p = pval) %>%
   bigreadr::fwrite2("SUM_STATS_FILE.txt", sep = "\t")
-  head()
-# hg19chrc    snpid    a1    a2    bp    or    p
-# chr1    rs4951859    C    G    729679    0.97853    0.2083
-# chr1    rs142557973    T    C    731718    1.01949    0.3298
 
+readLines("SUM_STATS_FILE.txt", n = 5)
+# [1] "hg19chrc\tsnpid\ta1\ta2\tbp\tor\tp"                                 
+# [2] "chr1\trs3934834\tA\tG\t995669\t1.03012433299041\t0.449674955932202" 
+# [3] "chr1\trs3737728\tA\tG\t1011278\t1.02982046003153\t0.363400424211927"
+# [4] "chr1\trs6687776\tA\tG\t1020428\t1.08042108970176\t0.048125118630458"
+# [5] "chr1\trs9651273\tA\tG\t1021403\t1.0347803708118\t0.291137612222826"
 
+unlink(paste0("backingfiles/celiac_test1", c(".bed", ".bim", ".fam")))
+snp_writeBed(celiac, "backingfiles/celiac_test1.bed", ind.row = ind.test.split[[1]])
+
+unlink("OUT_COORD_FILE.hdf5")
 system(glue::glue(
   "coord", 
   " --gf=backingfiles/celiac_test1",
-  " --ssf=SUM_STATS_FILE",
-  " --SSF_FORMAT=BASIC",
[…]

### 2019-01-23 plot-AUC-UKB.R

@@ -1,0 +1,28 @@
+res_auc <- data.frame(AUC = c(0.587, 0.587), Method = c("C+T-max", "PLR"))
+res_cor <- data.frame(cor = c(0.555, 0.655), Method = c("C+T-max", "PLR"))
+library(ggplot2)
+p1 <- ggplot() +
+  geom_col(aes(Method, AUC, fill = Method), data = res_auc, position = "dodge",
+           width = 0.9, alpha = 0.5, color = "black", size = 1) +
+  geom_hline(yintercept = 0.5, linetype = 2) +
+  bigstatsr::theme_bigstatsr() +
+  scale_y_continuous(limits = c(0.5, 0.6), breaks = seq(0.5, 0.6, 0.02),
+                     minor_breaks = seq(0.5, 0.6, 0.01), oob = scales::rescale_none) +
+  ggtitle("PRS for Breast Cancer")
+
+
+p2 <- ggplot() +
+  geom_col(aes(Method, cor, fill = Method), data = res_cor, position = "dodge",
+           width = 0.9, alpha = 0.5, color = "black", size = 1) +
+  bigstatsr::theme_bigstatsr() +
+  scale_y_continuous(breaks = seq(0, 1, 0.05), minor_breaks = seq(0, 1, 0.01)) +
+  labs(y = "Correlation (within each sex)") +
+  ggtitle("PRS for Height")
+
+library(cowplot)
+plot_grid(
+  p1 + theme(legend.position = "none"),
+  p2 + theme(legend.position = "none"),
+  scale = 0.95
+)
+

### 2019-01-23 paper/PRS.tex
nearly last updates

@@
 Moreover, our implementation of PLR directly includes automatic choices for hyper-parameters.
+We also provide a similar implementation of penalized linear regression.
 }
@@
 We compare the performance of PLR, C+T and a derivation of random forests using both real and simulated data. 
-PLR consistently achieves higher predictive performance than the two other methods while being as fast as C+T. 
-We find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; for instance, AUC values increase from 83\% with the best prediction of C+T to 92.5\% with PLR.
+{\color{red}
+Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being scalable to biobank data.
+}
+We find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; for instance, in simulations, AUC values increase from 83\% with the best prediction of C+T to 92.5\% with PLR.
 We confirm these results in a data analysis of a case-control study for celiac disease where PLR and the standard C+T method achieve AUC of 89\% and of 82.5\%.
 
-In conclusion, our study demonstrates that penalized logistic regression can achieve more discriminative polygenic risk scores, while being applicable to large-scale individual-level data thanks to the implementation we provide in the R package bigstatsr.
+{\color{red}
+Using 350,000 individuals of the UK Biobank, we predict height with 10\% more correlation than with the best prediction of C+T.
+Although height is a highly polygenic trait with many small effects, our implementation of penalized regression proves to be scalable and, to the best of our knowledge, provides the best genetic predictor of height to date.
+Moreover, using 150,000 individuals of the UK Biobank, we are able to predict breast cancer better than C+T, fitting PLR in a few minutes only.
+This demonstrates the feasibility and relevance of using penalized regression for PRS computation when large individual-level datasets are available.
 }
 
-{\color{red}
-Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being as fast.
-This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some well power studies in return.
+In conclusion, our study demonstrates that penalized regression can achieve much more discriminative polygenic risk scores, while being applicable to large-scale individual-level data thanks to the implementation we provide in R package bigstatsr.
 }
@@
 {\color{red}
-Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being as fast.
-This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some high predictive performance in return.
+Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being scalable to biobank data.
+%This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some high predictive performance in return.
 }
@@
 {\color{red}
-For the computation of this paper, the GWAS part is done on the training set while clumping is done on the test set (all individuals not included in the training set).
+For the computation of this paper, the GWAS part uses the training set while clumping uses the test set (all individuals not included in the training set).
 }
@@
 
-We tested our implementation on 656K genotyped SNPs of the UK Biobank, keeping only caucasian individuals and removing individuals related to someone else in the data (second individual in each pair with a kinship greater than 0.08).
+We tested our implementation on 656K genotyped SNPs of the UK Biobank, keeping only caucasian individuals and removing individuals related to someone else in the data (excluding the second individual in each pair with a kinship greater than 0.08).
 Our L1-penalized linear regression runs in less than one day for 350K individuals, achieving a correlation of 65.5\% with true height for both sexes.
@@
 
-[TODO: RESULTS TABLE]
+[TODO: COMPLETER + RESULTS TABLE]
 
@@
 The computation time of our PLR implementation mainly depends on the sample size and the number of active variables.
-Indeed, the algorithm is composed of two steps: first, for each variable, the algorithm computes one statistic that will be used to decide if the variable enters the model or not.
+Indeed, the algorithm is composed of two steps: first, for each variable, the algorithm computes one statistic that will be used to decide if the variable enters the model or not (for each value of $\lambda$).
 Then, the algorithm iterates over a regularization path of decreasing values of $\lambda$, which progressively enables variables to enter the model (Figure  \ref{fig:CMSA}).
@@
 If there are too many variables accessed, the operating system (OS) has to make some space for new coming variables. 
-So, there should never be any memory issue thanks to memory-mapping.
+There should never be any memory issue thanks to memory-mapping.
 Yet, at some point, if there are too many active variables, the OS would regularly swap memory between disk and RAM, which is very slow.
 If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM. 
-We have quickly tested processing each chromosome separately and then combining the models later but this did not seem to perform as well as fitting all SNPs at once (data not shown). 
+We have quickly tested processing each chromosome separately and then combining the models later but this did not perform as well as fitting all SNPs at once (data not shown). 
 Another widely used approach is to apply the penalized regression on a subset of SNPs using univariate tests (GWAS) to prioritize SNPs.
-Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm here by achieving a correlation of 65.5\% for height using 350K individuals of the UK Biobank, whereas \cite{accuheight2018} achieved only 61\% for the same sample size when using only the 50K most significantly associated SNPs (see figure 3 of their paper).
+Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm here by achieving a correlation of 65.5\% for height using 350K individuals of the UK Biobank, whereas \cite{accuheight2018} achieved only 61\% for the same sample size when using only the 50K most significantly associated SNPs [TODO: COMPARER PLUTOT A MOI DIRECTEMENT].
 
@@
 This may be because PLR is too conservative in including effects, whereas it has been shown in the literature that, in such underpowered scenarios, including (nearly) all SNPs was the best strategy \cite[]{purcell2009common}.
-In scenario \textnumero3, by varying sample size, we showed that when increasing sample size, PLR would achieve higher predictive performance than C+T at some point and that the benefits of using PLR over C+T would increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}).
+In scenario \textnumero3, by varying sample size, we show that when increasing sample size, PLR achieves higher predictive performance than C+T at some point and that the benefits of using PLR over C+T increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}).
 Moreover, in scenario \textnumero2, we decrease the search space (p: number of SNPs) in order to increase the polygenicity of the models and to virtually increase the sample size \cite[]{dudbridge2013power}.

### 2019-01-28 paper/PRS.tex
add UKBB table

@@
 
+\begin{table}[h]
+\caption{Results for the UK Biobank.
+The training/test sets for Height (resp.
+Breast Cancer) are 350,000/24,131 (resp.
+150,000/38,628).
+For PLR, dfmax is a parameter that controls the maximum number of variables allowed to enter the model before stopping fitting (default is 50,000 for each model corresponding to each validation set).
+Third row corresponds to fitting PLR on the top 100,000 SNPs from a GWAS on the training set.
+For Height, $r$ (correlation with true height) is reported within women / men; for Breast Cancer, AUC is reported.\label{tab:results-UKBB}}
+\vspace*{0.5em}
+\centering
+\begin{tabular}{|l|l|c|c|c|}
+  \hline
+Trait & Method & $r$ or AUC & \# predictors & Execution time \\
+  \hline
+Height & PLR (dfmax = 50e3) & 0.656 / 0.657 & 115,997 & 21h \\ 
+Height & PLR (dfmax = Inf)  & 0.657 / 0.658 & 148,341 & 61h \\ 
+Height & PLR (100K top SNPs, dfmax = Inf)  & 0.634 / 0.643 & 70,162 & 48.5h \\ 
+Height & C+T-max & 0.549 / 0.561 & 45,570 & 69m \\ 
+Breast Cancer & PLR & 0.598 & 2653 & 13m \\ 
+Breast Cancer & C+T-max & 0.589 & 21 & 15h \\
+  \hline
+\end{tabular}
+\end{table}
+
 }

### 2019-01-31 paper/PRS-supp.tex
add table/figure

@@
 \begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{celiac-roc}}
+\caption{ROC Curves for C+T-max, PLR and PLR3 for the celiac disease dataset.
+Models were trained using 12,000 individuals.
+These are results projecting these models on the remaining 3155 individuals.
+The figure is plotted using R package plotROC \cite[]{sachs2017plotroc}.\label{fig:celiac-roc}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newpage
+\begin{figure}[h]
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-corr}}
@@
 
+\newpage
+\begin{figure}[h]
+\centerline{\includegraphics[width=\textwidth]{supp-lasso-path-height}}
+\caption{Regularization path for `K = 10' validation sets when fitting our L1-penalized linear regression on 350,000 individuals of the UK Biobank.
+This plot can be obtained by using the `plot()' method on the object resulting from this fitting.
+This fitting used `nlambda = 200` and `n.abort = 10' (the defaults).}
+\label{fig:supp-path}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \end{document}

### 2019-02-04 LDpred.R
save

@@ -1,6 +1,3 @@
-# system("pip2 install --user --upgrade pip")
-# system("pip2 install --user ldpred")
-
 library(bigsnpr)
 celiac <- snp_attach("backingfiles/celiacQC.rds")
 G <- celiac$genotypes
@@ -31,6 +28,7 @@ celiac$map %>%
   select(hg19chrc = chromosome, snpid = marker.ID,
          a1 = allele1, a2 = allele2, bp = physical.pos) %>%
   mutate(hg19chrc = paste0("chr", hg19chrc), or = exp(gwas$estim), p = pval) %>%
+  # filter(hg19chrc %in% c("chr1", "chr2")) %>%
   bigreadr::fwrite2("SUM_STATS_FILE.txt", sep = "\t")
 
 readLines("SUM_STATS_FILE.txt", n = 5)
@@ -43,22 +41,26 @@ readLines("SUM_STATS_FILE.txt", n = 5)
 unlink(paste0("backingfiles/celiac_test1", c(".bed", ".bim", ".fam")))
 snp_writeBed(celiac, "backingfiles/celiac_test1.bed", ind.row = ind.test.split[[1]])
 
+reticulate::use_python("/home/privef/anaconda3/bin/python3")
+reticulate::py_config()
+stopifnot(system("python3 --version", intern = TRUE) == "Python 3.7.0")
+ldpred <- "../ldpred/LDpred.py"
 unlink("OUT_COORD_FILE.hdf5")
 system(glue::glue(
-  "coord", 
-  " --gf=backingfiles/celiac_test1",
-  " --ssf=SUM_STATS_FILE.txt --ssf_format=BASIC",
-  " --N={length(ind.train)}",
-  " --out=OUT_COORD_FILE.hdf5"
+  "python3 {ldpred} coord", 
+  " --gf backingfiles/celiac_test1",
+  " --ssf SUM_STATS_FILE.txt --ssf-format BASIC",
+  " --N {length(ind.train)}",
+  " --out OUT_COORD_FILE.hdf5"
 ))
 
 system(glue::glue(
   "ldpred", 
-  " --coord=OUT_COORD_FILE.hdf5",
-  " --ld_radius={round(ncol(G) / 3000)}",
-  " --PS=0.01,0.001,0.0001",
-  " --N={length(ind.train)}",
-  " --out=OUT_COORD_FILE"
+  " --coord OUT_COORD_FILE.hdf5",
[…]

### 2019-02-04 paper/PRS.tex

--- 2019-01-23 paper/PRS.tex
@@
 
-In conclusion, our study demonstrates that penalized regression can achieve much more discriminative polygenic risk scores, while being applicable to large-scale individual-level data thanks to the implementation we provide in R package bigstatsr.
+In conclusion, our study demonstrates that penalized regression can achieve much more predictive polygenic risk scores, while being applicable to large-scale individual-level data thanks to the implementation we provide in R package bigstatsr.
 }
@@
 We tested our implementation on 656K genotyped SNPs of the UK Biobank, keeping only caucasian individuals and removing individuals related to someone else in the data (excluding the second individual in each pair with a kinship greater than 0.08).
-Our L1-penalized linear regression runs in less than one day for 350K individuals, achieving a correlation of 65.5\% with true height for both sexes.
+Results are presented in table \ref{tab:results-UKBB}.
+
+Our L1-penalized linear regression runs in less than one day for 350K individuals, achieving a correlation of more than 65.5\% with true height for both sexes.
 The best C+T model achieves a correlation of 55\% for women and 56\% for men, where the GWAS part takes 1 hour only.
-Our L1-penalized logistic regression on breast cancer runs in only 12 minutes for 150K women, achieving an AUC of 0.587.
-The best C+T model achieves the same AUC score, where the GWAS part takes 15 hours.
+If using only the top 100,000 SNPs from a GWAS on the training set to train our L1-PLR, correlation between predicted and true heights drops at 0.634 for women and 0.643 for men.
+Our L1-penalized logistic regression on breast cancer runs in only 13 minutes for 150K women, achieving an AUC of 0.598.
+The best C+T model achieves an AUC of 0.589, where the GWAS part takes 15 hours.
 
-[TODO: COMPLETER + RESULTS TABLE]
+\begin{table}[h]
+\caption{Results for the UK Biobank.
+The training/test sets for height (resp.
+breast cancer) are 350,000/24,131 (resp.
+150,000/38,628).
+For PLR, dfmax is a parameter that controls the maximum number of variables allowed to enter the model before stopping fitting (default is 50,000 for each model corresponding to each validation set).
+Third row corresponds to fitting PLR on the top 100,000 SNPs from a GWAS on the training set.
+For height, $r$ (correlation with true height) is reported within women / men; for breast cancer, AUC is reported.\label{tab:results-UKBB}}
+\vspace*{0.5em}
+\centering
+\begin{tabular}{|l|l|c|c|c|}
+  \hline
+Trait & Method & $r$ or AUC & \# predictors & Execution time \\
+  \hline
+height & PLR (dfmax = 50e3) & 0.656 / 0.657 & 115,997 & 21h \\ 
+height & PLR (dfmax = Inf)  & 0.657 / 0.658 & 148,341 & 61h \\ 
+height & PLR (100K top SNPs, dfmax = Inf)  & 0.634 / 0.643 & 70,162 & 48.5h \\ 
+height & C+T-max & 0.549 / 0.561 & 45,570 & 69m \\ 
+breast cancer & PLR & 0.598 & 2653 & 13m \\ 
+breast cancer & C+T-max & 0.589 & 21 & 15h \\
+  \hline
+\end{tabular}
+\end{table}
 
@@
 Indeed, the algorithm is composed of two steps: first, for each variable, the algorithm computes one statistic that will be used to decide if the variable enters the model or not (for each value of $\lambda$).
+This first step should be very fast.
 Then, the algorithm iterates over a regularization path of decreasing values of $\lambda$, which progressively enables variables to enter the model (Figure  \ref{fig:CMSA}).
@@
 
-So, for highly polygenic traits such as height and when using huge datasets such as the UK Biobank, the algorithm might iterate over dozens of thousands of variables, which takes time (but still less than one day with our implementation).
-On the contrary, for traits like the celiac disease or breast cancer that seems to be less polygenic, a lot less variables are entering the model and the fitting is very fast (12 minutes only for 150K women of the UK Biobank for breast cancer!).
-Using more cases (there were only 3542 for Celiac and 6374 for Breast Cancer), one might achieve better predictive performance for these two diseases, and the computation time should remain within a few hours.
+So, for highly polygenic traits such as height and when using huge datasets such as the UK Biobank, the algorithm might iterate over more than 100,000 variables, which takes time.
+On the contrary, for traits like the celiac disease or breast cancer that seems to be less polygenic, a lot less variables are entering the model and the fitting is very fast (13 minutes only for 150K women of the UK Biobank for breast cancer!).
+Using more cases (there were only 3542 for Celiac and 6768 for breast cancer), one might achieve better predictive performance for these two diseases, and the computation time should remain within a few hours.
 
@@
 If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM. 
+Another solution would be to reduce the `n.abort' parameter (default value is 10) in order to stop fitting the regularization path even sooner.
+This would avoid fitting models with a large number of variables without changing the solution since regularization paths for very large sample sizes are very smooth (Figure \ref{fig:supp-path}).
 We have quickly tested processing each chromosome separately and then combining the models later but this did not perform as well as fitting all SNPs at once (data not shown). 
-Another widely used approach is to apply the penalized regression on a subset of SNPs using univariate tests (GWAS) to prioritize SNPs.
-Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm here by achieving a correlation of 65.5\% for height using 350K individuals of the UK Biobank, whereas \cite{accuheight2018} achieved only 61\% for the same sample size when using only the 50K most significantly associated SNPs [TODO: COMPARER PLUTOT A MOI DIRECTEMENT].
+Another widely used approach is to apply the penalized regression on a subset of SNPs by prioritizing SNPs using univariate tests (GWAS computed from the same dataset).
+Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm in this paper.
+Indeed, when using only the 100K most significantly associated SNPs, this reduces correlation between predicted and true heights from 0.657/0.658 to 0.634/0.643 within women/men, while decreasing computation time by 20\% only (Table \ref{tab:results-UKBB}).
 
@@
 Moreover, in scenario \textnumero2, we decrease the search space (p: number of SNPs) in order to increase the polygenicity of the models and to virtually increase the sample size \cite[]{dudbridge2013power}.
-In this scenario, even when there are small effects and a high polygenicity (3000 causal variants out of 18,941), PLR gets a large increase in predictive performance, now consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
+In this scenario \textnumero2, even when there are small effects and a high polygenicity (3000 causal variants out of 18,941), PLR gets a large increase in predictive performance, now consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
 }

### 2019-02-08 paper/PRS.tex
save

@@
 {\color{red}
-In this paper, we present an efficient method to jointly estimate SNP effects using individual-level data, allowing for practical application of penalized logistic regression (PLR) on modern datasets including hundreds of thousands of individuals.
+In this paper, we present an efficient method for the joint estimation of SNP effects using individual-level data, allowing for practical application of penalized logistic regression (PLR) on modern datasets including hundreds of thousands of individuals.
 Moreover, our implementation of PLR directly includes automatic choices for hyper-parameters.
-We also provide a similar implementation of penalized linear regression.
+We also provide an implementation of penalized linear regression for quantitative traits.
 }
@@
 {\color{red}
-Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being scalable to biobank data.
+Overall, we find that PLR achieves equal or higher predictive performance than C+T in most scenarios considered, while being scalable to biobank data.
 }
-We find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; for instance, in simulations, AUC values increase from 83\% with the best prediction of C+T to 92.5\% with PLR.
+In particular, we find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; for instance, in simulations, AUC values increase from 83\% with the best prediction of C+T to 92.5\% with PLR.
 We confirm these results in a data analysis of a case-control study for celiac disease where PLR and the standard C+T method achieve AUC of 89\% and of 82.5\%.
@@
 {\color{red}
-Using 350,000 individuals of the UK Biobank, we predict height with 10\% more correlation than with the best prediction of C+T.
-Although height is a highly polygenic trait with many small effects, our implementation of penalized regression proves to be scalable and, to the best of our knowledge, provides the best genetic predictor of height to date.
+Applying our model to 350,000 individuals of the UK Biobank, we predict height with a larger correlation than with the best prediction of C+T [TODO CHIFFRES FROM TO], further demonstrating its scalibility and strong predictive power even for highly polygenic traits.
 Moreover, using 150,000 individuals of the UK Biobank, we are able to predict breast cancer better than C+T, fitting PLR in a few minutes only.
 This demonstrates the feasibility and relevance of using penalized regression for PRS computation when large individual-level datasets are available.
+
+In conclusion, our study demonstrates that penalized regression can provide discriminative polygenic risk scores, while being applicable to large-scale individual-level data because of an efficient implementation available in R package bigstatsr.
 }
-
-In conclusion, our study demonstrates that penalized regression can achieve much more predictive polygenic risk scores, while being applicable to large-scale individual-level data thanks to the implementation we provide in R package bigstatsr.
 }
@@
 Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
-PRS are useful for genetic epidemiology when testing the polygenicity of one disease and finding a common genetic contribution between two diseases \cite[]{purcell2009common}.
+PRS are useful for genetic epidemiology when testing polygenicity of diseases and finding a common genetic contribution between two diseases \cite[]{purcell2009common}.
 Personalized medicine is another major application of PRS.
@@
 {\color{red}
-Yet, PRS would have to show a high discriminative power between cases and controls in order to be used in clinical settings.
-For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.
-Note that these AUC figures are also dependent on the prevalence of the disease.
+However, in order to be used in clinical settings, PRS should discriminate well enough between cases and controls.
+For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact} [TODO: find prevalence].
 }
@@
 Package bigstatsr now includes an efficient algorithm with a new implementation for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
-In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression (PLR) against the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
+In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression (PLR), which we compare to the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
 In this comparison, we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
@@
 Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being scalable to biobank data.
+For example, this underpowered situation occurs when there are many causal variants (3K) to find among many typed variants (280K) and a small sample size (6K).
 %This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some high predictive performance in return.
@@
 Reducing the number of SNPs increases the polygenicity (i.e.\ the proportion of causal SNPs) of the simulated models.
-Reducing the number of SNPs ($p$) is also equivalent to increasing the sample size ($n$) as predictive power is dependent on $n/p$ \cite[]{dudbridge2013power,vilhjalmsson2015modeling}.
+Reducing the number of SNPs ($p$) is also equivalent to increasing the sample size ($n$) as predictive power increases as a function of $n/p$ \cite[]{dudbridge2013power,vilhjalmsson2015modeling}.
 For this scenario, we use the additive model only, but continue to vary all other simulation parameters.
@@
 \begin{table}[htbp]
-\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.\label{tab:simus}}
+\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box. [TODO: number of SNPs in dataset]\label{tab:simus}}
 \vspace*{0.5em}
@@
 \hline
-Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
-scenario & & training set & (number and location) & of effects & & model & \\
+Numero of & \multirow{2}{*}{Dataset} & Sample size & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
+scenario & & of training set & (number and location) & of effects & & model & \\
 \hline
@@
 {\color{red}
-For the computation of this paper, the GWAS part uses the training set while clumping uses the test set (all individuals not included in the training set).
+Here, the GWAS part uses the training set while clumping uses the test set (all individuals not included in the training set).
 }
@@
 the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $p \gg n$ situation ($p$ is the number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
-In this study, we use an embedded grid search over $\alpha \in \{1,~0.5,~0.05,~0.001\}$.
+In this study, we use a grid search over $\alpha \in \{1,~0.5,~0.05,~0.001\}$, which is embedded in our PLR implementation.
 
-To fit this penalized logistic regression, we use an efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in R package bigstatsr.
+To fit penalized logistic regression, we use an efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in R package bigstatsr.
 This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''.
-To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
-Because of L1-regularization, the resulting vectors of coefficients are sparse and can be used to make a PRS based on a \emph{linear} combination of allele counts.
+To obtain an algorithm that does not require to choose this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
+Because of L1-regularization, the resulting vector of estimated effect sizes is sparse. 
 We refer to this method as ``PLR'' in the results section.
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
-\caption{Main comparison of C+T and PLR in scenario \textnumero1 for model ``ADD''.
+\caption{Main comparison of C+T and PLR when simulating phenotypes with additive effects (scenario \textnumero1, model ``ADD'') [TODO: chnages for every plot].
 Mean AUC over 100 simulations for PLR and the maximum AUC reported with ``C+T-max'' (clumping threshold at $r^2 > 0.2$).
@@
 In terms of computation time, we show that PLR, while learning jointly on all SNPs at once and testing four different values for hyper-parameter $\alpha$, is almost as fast as the C+T method (190 vs 130 seconds), and PLR3 takes less than twice as long as PLR (296 vs 190 seconds).
-{\color{red}
-Finally, on this Celiac dataset, our implementation of PLR runs 300 times as fast as SparSNP \cite[]{abraham2012sparsnp}.
-}
+[TODO: reformuler SparSNP reponse]
 
@@
 
-We tested our implementation on 656K genotyped SNPs of the UK Biobank, keeping only caucasian individuals and removing individuals related to someone else in the data (excluding the second individual in each pair with a kinship greater than 0.08).
+We tested our implementation on 656K genotyped SNPs of the UK Biobank, keeping only caucasian individuals and removing related individuals (excluding the second individual in each pair with a kinship coefficient greater than 0.08).
 Results are presented in table \ref{tab:results-UKBB}.
 
-Our L1-penalized linear regression runs in less than one day for 350K individuals, achieving a correlation of more than 65.5\% with true height for both sexes.
-The best C+T model achieves a correlation of 55\% for women and 56\% for men, where the GWAS part takes 1 hour only.
+Our implementation of L1-penalized linear regression runs in less than one day for 350K individuals, achieving a correlation of more than 65.5\% with true height for both sexes [TODO: IN..
+ajouter test set].
+By comparison, the best C+T model achieves a correlation of 55\% for women and 56\% for men, and the GWAS part takes 1 hour.
 If using only the top 100,000 SNPs from a GWAS on the training set to train our L1-PLR, correlation between predicted and true heights drops at 0.634 for women and 0.643 for men.
-Our L1-penalized logistic regression on breast cancer runs in only 13 minutes for 150K women, achieving an AUC of 0.598.
-The best C+T model achieves an AUC of 0.589, where the GWAS part takes 15 hours.
+Our L1-penalized logistic regression on breast cancer runs in 13 minutes for 150K women, achieving an AUC of 0.598 [TODO test set].
+The best C+T model achieves an AUC of 0.589, and the GWAS part takes 15 hours.
 
 \begin{table}[h]
-\caption{Results for the UK Biobank.
+\caption{[TODO deplace dfmax supmat] Results for the UK Biobank.
 The training/test sets for height (resp.
@@
   \hline
-Trait & Method & $r$ or AUC & \# predictors & Execution time \\
+Trait & Method & $r$ (women / men) or AUC & \# predictors & Execution time \\
   \hline
@@
 
-{\color{red}
-\subsection{Time and memory requirements}
-
-The computation time of our PLR implementation mainly depends on the sample size and the number of active variables.
-Indeed, the algorithm is composed of two steps: first, for each variable, the algorithm computes one statistic that will be used to decide if the variable enters the model or not (for each value of $\lambda$).
-This first step should be very fast.
-Then, the algorithm iterates over a regularization path of decreasing values of $\lambda$, which progressively enables variables to enter the model (Figure  \ref{fig:CMSA}).
-So, in the second step, the number of variables is increasing progressively, and at some point the computation stops due to the early stopping criterion (when prediction is getting worse on the corresponding validation set, see figure \ref{fig:CMSA}).
-
-So, for highly polygenic traits such as height and when using huge datasets such as the UK Biobank, the algorithm might iterate over more than 100,000 variables, which takes time.
-On the contrary, for traits like the celiac disease or breast cancer that seems to be less polygenic, a lot less variables are entering the model and the fitting is very fast (13 minutes only for 150K women of the UK Biobank for breast cancer!).
-Using more cases (there were only 3542 for Celiac and 6768 for breast cancer), one might achieve better predictive performance for these two diseases, and the computation time should remain within a few hours.
-
-Regarding memory requirements, this is basically the same as for computation time.
-Indeed, active variables are accessed in memory thanks to memory-mapping when they are used \cite[]{prive2017efficient}.
-If there are too many variables accessed, the operating system (OS) has to make some space for new coming variables. 
-There should never be any memory issue thanks to memory-mapping.
-Yet, at some point, if there are too many active variables, the OS would regularly swap memory between disk and RAM, which is very slow.
-If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM. 
-Another solution would be to reduce the `n.abort' parameter (default value is 10) in order to stop fitting the regularization path even sooner.
-This would avoid fitting models with a large number of variables without changing the solution since regularization paths for very large sample sizes are very smooth (Figure \ref{fig:supp-path}).
-We have quickly tested processing each chromosome separately and then combining the models later but this did not perform as well as fitting all SNPs at once (data not shown). 
-Another widely used approach is to apply the penalized regression on a subset of SNPs by prioritizing SNPs using univariate tests (GWAS computed from the same dataset).
-Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm in this paper.
-Indeed, when using only the 100K most significantly associated SNPs, this reduces correlation between predicted and true heights from 0.657/0.658 to 0.634/0.643 within women/men, while decreasing computation time by 20\% only (Table \ref{tab:results-UKBB}).
-
-}
-
 \subsection{Joint estimation improves predictive performance}
@@
 
-When there are many effects of small size and a small sample size, PLR performs worse than (the best result for) C+T.
-This may be because PLR is too conservative in including effects, whereas it has been shown in the literature that, in such underpowered scenarios, including (nearly) all SNPs was the best strategy \cite[]{purcell2009common}.
-In scenario \textnumero3, by varying sample size, we show that when increasing sample size, PLR achieves higher predictive performance than C+T at some point and that the benefits of using PLR over C+T increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}).
-Moreover, in scenario \textnumero2, we decrease the search space (p: number of SNPs) in order to increase the polygenicity of the models and to virtually increase the sample size \cite[]{dudbridge2013power}.
+When there are many small effects and a small sample size, PLR performs worse than (the best result for) C+T.
+For example, this underpowered situation occurs when there are many causal variants (3K) to find among many typed variants (280K) and a small sample size (6K).
+In such underpowered scenarios, it is difficult to detect true causal variants, which makes PLR too conservative whereas the best strategy is to include nearly all SNPs \cite[]{purcell2009common}.
+When increasing sample size (scenario \textnumero3), PLR achieves higher predictive performance than C+T and the benefits of using PLR over C+T increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}).
+Moreover, when decreasing the search space (total number of candidate SNPs) in scenario \textnumero2, we increase the proportion of causal variants and we virtually increase the sample size \cite[]{dudbridge2013power}.
 In this scenario \textnumero2, even when there are small effects and a high polygenicity (3000 causal variants out of 18,941), PLR gets a large increase in predictive performance, now consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
@@
 {\color{red}
-We did not want to choose the C+T parameters so that one could not criticize the procedure we would have used.
-Instead, for C+T, we chose to report the best AUC value on the test set, even if it leads to over-optimistic results for C+T as compared to PLR.
+We did not want to choose the C+T parameters as, to our knowledge, there is no gold standard to do so.
+So, for C+T, we report the best AUC value on the test set, even if it leads to over-optimistic results for C+T as compared to PLR.
 }
@@
 
+{\color{red}
+\subsection{Time and memory requirements}
+
+The computation time of our PLR implementation mainly depends on the sample size and the number of active variables (variables that are included in the gradient descent).
+Indeed, the algorithm is composed of two steps: first, for each variable, the algorithm computes an univariate statistic that is used to decide if the variable is included in the model (for each value of $\lambda$).
+This first step is very fast.
+Then, the algorithm iterates over a regularization path of decreasing values of $\lambda$, which progressively enables variables to enter the model (Figure  \ref{fig:CMSA}).
+In the second step, the number of variables increases, and computations stop when an early stopping criterion is reached (when prediction is getting worse on the corresponding validation set, see figure \ref{fig:CMSA}).
+
+For highly polygenic traits such as height and when using huge datasets such as the UK Biobank, the algorithm might iterate over more than 100,000 variables, which is computationally demanding.
+On the contrary, for traits like celiac disease or breast cancer that are less polygenic, the number of variables included in the model is much smaller so that fitting is very fast (13 minutes only for 150K women of the UK Biobank for breast cancer).
+%Using more cases (there were only 3542 for Celiac and 6768 for breast cancer), one might achieve better predictive performance for these two diseases, and the computation time should remain within a few hours.
+
+Memory requirements are tightly linked to computation time. Indeed, active variables are accessed in memory thanks to memory-mapping when they are used \cite[]{prive2017efficient}. 
+If there are too many variables accessed, the operating system (OS) must free some memory for new coming variables. 
+There should never be any memory issue thanks to memory-mapping.
+Yet, if there are too many active variables, the OS would regularly swap memory between disk and RAM, severely slowing down computations.
+%If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM. 
+%Another solution would be to reduce the `n.abort' parameter (default value is 10) in order to stop fitting the regularization path even sooner. 
+[TODO: add doc function]
+%This would avoid fitting models with a large number of variables without changing the solution since regularization paths for very large sample sizes are very smooth (Figure \ref{fig:supp-path}).
+%We tested processing each chromosome separately and then combining the models later but this did not perform as well as fitting all SNPs at once (data not shown). 
+A possible approach to reduce computational burden is to apply penalized regression on a subset of SNPs by prioritizing SNPs using univariate tests (GWAS computed from the same dataset).
+Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm in this paper.
+Indeed, when using only the 100K most significantly associated SNPs, this reduces correlation between predicted and true heights from 0.657/0.658 to 0.634/0.643 within women/men, while decreasing computation time by 20\% only (Table \ref{tab:results-UKBB}).
+A key advantage of our implementation of PLR is that prior filtering of variables is no more required for computational feasibility.
+
+}
+
 \subsection{Limitations}
@@
 Our approach has one major limitation: the main advantage of the C+T method is its direct applicability to summary statistics, allowing to leverage the largest GWAS results to date, even when individual cohort data cannot be merged because of practical or ethical reasons (e.g.\ consortium data including many cohorts).
-As of today, our implementation of penalized logistic regression does not allow for the analysis of summary data, but this represents an important future direction of our work.
+As of today, our implementation of penalized logistic regression does not allow for the analysis of summary data, but this represents an important future direction.
 The current version is of particular interest for the analysis of modern individual-level datasets including hundreds of thousands of individuals.
@@
 In this comparative study, we have presented a computationally efficient implementation of penalized logistic regression that can be used to predict disease status based on genotypes.
-Note that a similar penalized linear regression is also available in our software.
+A similar penalized linear regression is also available in R package bigstatsr.
 Our approach solves the dramatic computational burden faced by standard implementations, thus allowing for the analysis of large-scale datasets such as the UK biobank \cite[]{bycroft2017genome}.
@@
 It can be appropriate for predicting autoimmune diseases with a few strong effects (e.g.\ celiac disease) as well as highly polygenic traits (e.g.\ standing height).
-Finally, note that these models could also be used to predict phenotypes based on other omics data since the implementation is not specific to genotype data.
+Finally, PLR as implemented in bigstatsr can also be used to predict phenotypes based on other omics data since our implementation is not specific to genotype data.
 
@@
 Authors also acknowledge the Grenoble Alpes Data Institute that is supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02).
-We used the UK Biobank resource (application ID 25589).
+This research has been conducted using the UK Biobank Resource under Application Number 25589.
 We are also grateful to F\'elix Balazard for useful discussions about T-Trees, and to Yaohui Zeng for useful discussions about R package biglasso.

### 2019-02-09 paper/PRS.tex
save

@@
 {\color{red}
-Applying our model to 350,000 individuals of the UK Biobank, we predict height with a larger correlation than with the best prediction of C+T [TODO CHIFFRES FROM TO], further demonstrating its scalibility and strong predictive power even for highly polygenic traits.
+Applying our model to 350,000 individuals of the UK Biobank, we predict height with a larger correlation than with the best prediction of C+T (\textasciitilde65\% instead of \textasciitilde55\%), further demonstrating its scalibility and strong predictive power, even for highly polygenic traits.
 Moreover, using 150,000 individuals of the UK Biobank, we are able to predict breast cancer better than C+T, fitting PLR in a few minutes only.
@@
 However, in order to be used in clinical settings, PRS should discriminate well enough between cases and controls.
-For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact} [TODO: find prevalence].
+For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that, for a 10\% disease prevalence, the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.
 }
@@
 \begin{table}[htbp]
-\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box. [TODO: number of SNPs in dataset]\label{tab:simus}}
+\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box. \label{tab:simus}}
 \vspace*{0.5em}
@@
 \hline
-Numero of & \multirow{2}{*}{Dataset} & Sample size & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
-scenario & & of training set & (number and location) & of effects & & model & \\
+Numero of & Dataset & Sample size & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
+scenario & (number of SNPs) & of training set & (number and location) & of effects & & model & \\
 \hline
 \hline
-\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{ADD} & C+T \\
+\multirow{4}{*}{1} & \multirow{3}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{ADD} & C+T \\
 & & & 30 in all & & & & PLR \\
-& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{COMP} & PLR3 \\
+& (281,122 SNPs) & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{COMP} & PLR3 \\
 & & & 3000 in all & & & & (T-Trees) \\
 \hline
-\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{ADD} & C+T \\ 
-& & & & & & & PLR\\
+\multirow{2}{*}{2} & Chromosome 6 only & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{ADD} & C+T \\ 
+& (18,941 SNPs) & & & & & & PLR\\
 \hline
-\multirow{5}{*}{3} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
-& & 2000 & & & & & \\
+\multirow{5}{*}{3} & \multirow{4}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
+& \multirow{4}{*}{(281,122 SNPs)} & 2000 & & & & & \\
 & & 3000 & & & & & \\
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
-\caption{Main comparison of C+T and PLR when simulating phenotypes with additive effects (scenario \textnumero1, model ``ADD'') [TODO: chnages for every plot].
+\caption{Main comparison of C+T and PLR when simulating phenotypes with additive effects (scenario \textnumero1, model ``ADD'').
 Mean AUC over 100 simulations for PLR and the maximum AUC reported with ``C+T-max'' (clumping threshold at $r^2 > 0.2$).
@@
 \centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
-\caption{Comparison of methods in scenario \textnumero2 (using chromosome 6 only) for model ``ADD''.
+\caption{Comparison of methods when simulating phenotypes with additive effects and using chromosome 6 only (scenario \textnumero2).
 Thinner lines represents results in scenario \textnumero1.
@@
 \centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
-\caption{Comparison of methods when varying sample size in scenario \textnumero3 for model ``ADD'' with 300 causal SNPs sampled anywhere on the genome.
+\caption{Comparison of methods when simulating 300 causal SNPs with additive effects and when varying sample size (scenario \textnumero3).
 Mean AUC over 100 simulations for the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the training size.
@@
 In terms of computation time, we show that PLR, while learning jointly on all SNPs at once and testing four different values for hyper-parameter $\alpha$, is almost as fast as the C+T method (190 vs 130 seconds), and PLR3 takes less than twice as long as PLR (296 vs 190 seconds).
-[TODO: reformuler SparSNP reponse]
 
@@
 
-Our implementation of L1-penalized linear regression runs in less than one day for 350K individuals, achieving a correlation of more than 65.5\% with true height for both sexes [TODO: IN..
-ajouter test set].
-By comparison, the best C+T model achieves a correlation of 55\% for women and 56\% for men, and the GWAS part takes 1 hour.
-If using only the top 100,000 SNPs from a GWAS on the training set to train our L1-PLR, correlation between predicted and true heights drops at 0.634 for women and 0.643 for men.
-Our L1-penalized logistic regression on breast cancer runs in 13 minutes for 150K women, achieving an AUC of 0.598 [TODO test set].
-The best C+T model achieves an AUC of 0.589, and the GWAS part takes 15 hours.
+Our implementation of L1-penalized linear regression runs in less than one day for 350K individuals (training set), achieving a correlation of more than 65.5\% with true height for both sexes in the remaining 24K individuals (test set).
+By comparison, the best C+T model achieves a correlation of 55\% for women and 56\% for men (in the test set), and the GWAS part takes one hour (for the training set).
+If using only the top 100,000 SNPs from a GWAS on the training set to fit our L1-PLR, correlation between predicted and true heights drops at 0.634 for women and 0.643 for men.
+Our L1-penalized logistic regression on breast cancer runs in 13 minutes for 150K women, achieving an AUC of 0.598 in the remaining 39K women, while the best C+T model achieves an AUC of 0.589, and the GWAS part takes 15 hours.
 
 \begin{table}[h]
-\caption{[TODO deplace dfmax supmat] Results for the UK Biobank.
-The training/test sets for height (resp.
-breast cancer) are 350,000/24,131 (resp.
-150,000/38,628).
-For PLR, dfmax is a parameter that controls the maximum number of variables allowed to enter the model before stopping fitting (default is 50,000 for each model corresponding to each validation set).
-Third row corresponds to fitting PLR on the top 100,000 SNPs from a GWAS on the training set.
-For height, $r$ (correlation with true height) is reported within women / men; for breast cancer, AUC is reported.\label{tab:results-UKBB}}
+\caption{Results for the UK Biobank.
+The sizes of training / test sets for height (resp.\ breast cancer) are 350,000 / 24,131 (resp.\ 150,000 / 38,628).
+For height, $r$ (correlation between predicted and true heights) is reported within women / men separately; for breast cancer, AUC is reported.\label{tab:results-UKBB}}
 \vspace*{0.5em}
@@
   \hline
-Trait & Method & $r$ (women / men) or AUC & \# predictors & Execution time \\
+Trait & Method & $r$ (women / men) & \# predictors & Execution time \\
   \hline
-height & PLR (dfmax = 50e3) & 0.656 / 0.657 & 115,997 & 21h \\ 
-height & PLR (dfmax = Inf)  & 0.657 / 0.658 & 148,341 & 61h \\ 
-height & PLR (100K top SNPs, dfmax = Inf)  & 0.634 / 0.643 & 70,162 & 48.5h \\ 
+height & PLR & 0.656 / 0.657 & 115,997 & 21h \\ 
 height & C+T-max & 0.549 / 0.561 & 45,570 & 69m \\ 
+  \hline
+  \hline
+Trait & Method & AUC & \# predictors & Execution time \\
+  \hline
 breast cancer & PLR & 0.598 & 2653 & 13m \\ 
@@
 When there are many small effects and a small sample size, PLR performs worse than (the best result for) C+T.
-For example, this underpowered situation occurs when there are many causal variants (3K) to find among many typed variants (280K) and a small sample size (6K).
+For example, this situation occurs when there are many causal variants (3K) to distinguish among many typed variants (280K) while using a small sample size (6K).
 In such underpowered scenarios, it is difficult to detect true causal variants, which makes PLR too conservative whereas the best strategy is to include nearly all SNPs \cite[]{purcell2009common}.
@@
 Memory requirements are tightly linked to computation time. Indeed, active variables are accessed in memory thanks to memory-mapping when they are used \cite[]{prive2017efficient}. 
-If there are too many variables accessed, the operating system (OS) must free some memory for new coming variables. 
+When there is not enough memory left, the operating system (OS) frees some memory for new coming variables. 
 There should never be any memory issue thanks to memory-mapping.
@@
 %Another solution would be to reduce the `n.abort' parameter (default value is 10) in order to stop fitting the regularization path even sooner. 
-[TODO: add doc function]
 %This would avoid fitting models with a large number of variables without changing the solution since regularization paths for very large sample sizes are very smooth (Figure \ref{fig:supp-path}).

### 2019-02-09 paper/PRS.tex
save

@@
 {\color{red}
-Applying our model to 350,000 individuals of the UK Biobank, we predict height with a larger correlation than with the best prediction of C+T (\textasciitilde65\% instead of \textasciitilde55\%), further demonstrating its scalibility and strong predictive power, even for highly polygenic traits.
+Applying penalized linear regression to 350,000 individuals of the UK Biobank, we predict height with a larger correlation than with the best prediction of C+T (\textasciitilde65\% instead of \textasciitilde55\%), further demonstrating its scalability and strong predictive power, even for highly polygenic traits.
 Moreover, using 150,000 individuals of the UK Biobank, we are able to predict breast cancer better than C+T, fitting PLR in a few minutes only.
@@
 
-In conclusion, our study demonstrates that penalized regression can provide discriminative polygenic risk scores, while being applicable to large-scale individual-level data because of an efficient implementation available in R package bigstatsr.
+In conclusion, our study demonstrates that penalized regression can provide discriminative polygenic risk scores, while being applicable to large-scale individual-level data because of an highly efficient implementation available in R package bigstatsr.
 }
@@
 A commonly used method often called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') is used to derive PRS from results of Genome-Wide Association Studies (GWAS) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
-This technique uses GWAS summary statistics only, allowing for a fast implementation of C+T.
+This technique uses GWAS summary statistics, allowing for a fast implementation of C+T.
 However, C+T also has several limitations; for instance, previous studies have shown that predictive performance of C+T is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
-Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{yang2010common}.
-Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes (see table 2 of \cite{abraham2013performance}).
-%Moreover, these methods and their derivatives are often computationally very demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
-Finally, statistical learning methods have also been used to derive PRS for complex human diseases by jointly estimating SNP effects.
+In parallel, statistical learning methods have also been used to derive PRS for complex human diseases by jointly estimating SNP effects.
 Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,wei2009disease}.
+Finally, Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly polygenic quantitative human phenotypes such as height \cite[]{yang2010common}.
+{\color{red}
+Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', have not proven as efficient as other methods for predicting several complex diseases based on genotypes (see table 2 of \cite{abraham2013performance}).
+}
 
@@
 In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression (PLR), which we compare to the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
-In this comparison, we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
+{\color{red}
+In this comparison, we do not include any LMM method, yet, L2-PLR should be very similar to those methods (see section \ref{sec:PLR}).
+}
+Moreover, we do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}. 
 For C+T, we report results for a large grid of hyper-parameters. 
@@
 Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being scalable to biobank data.
-For example, this underpowered situation occurs when there are many causal variants (3K) to find among many typed variants (280K) and a small sample size (6K).
-%This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some high predictive performance in return.
+For example, this underpowered situation occurs when there are many causal variants (3K) to distinguish among many typed variants (280K) while using a small sample size (6K).
 }
@@
 Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only (18,941 SNPs).
-Reducing the number of SNPs increases the polygenicity (i.e.\ the proportion of causal SNPs) of the simulated models.
+Reducing the number of SNPs increases the polygenicity (the proportion of causal SNPs) of the simulated models.
 Reducing the number of SNPs ($p$) is also equivalent to increasing the sample size ($n$) as predictive power increases as a function of $n/p$ \cite[]{dudbridge2013power,vilhjalmsson2015modeling}.
@@
 
-The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWAS).
+The C+T (Clumping + Thresholding) method directly derives Polygenic Risk Scores (PRS) from the results of Genome-Wide Associations Studies (GWAS).
 In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$.
@@
 the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $p \gg n$ situation ($p$ is the number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
-In this study, we use a grid search over $\alpha \in \{1,~0.5,~0.05,~0.001\}$, which is embedded in our PLR implementation.
+{\color{red}
+In this study, we use a grid search over $\alpha \in \{1,~0.5,~0.05,~0.001\}$.
+This grid-search is directly embedded in our PLR implementation for simplicity.
+Using $\alpha = 0.001$ should result in a model very similar to gBLUP.\label{sec:PLR}
+}
 
@@
 When simulating a model with 30 causal SNPs and a heritability of 80\%, PLR provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\% for this setting (Figure \ref{fig:main-AUC-logit}).
-Moreover, PLR consistently provides higher predictive performance than C+T across all scenarios we considered, except in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-ntrain}).
+Moreover, PLR consistently provides higher predictive performance than C+T across all scenarios we considered, except in some cases of high polygenicity and small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-ntrain}).
 PLR provides particularly higher predictive performance than C+T when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region.
@@
 Concerning the $r^2$ threshold of the clumping step in C+T, we mostly used the common value of $0.2$.
-Yet, using a more stringent value of $0.05$ provides higher predictive performance than using $0.2$ in most of the cases we considered (Figures \ref{fig:supp-AUC-all-r2} and \ref{fig:main-AUC-ntrain})
+Yet, using a more stringent value of $0.05$ provides equal or higher predictive performance than using $0.2$ in most of the cases we considered (Figures \ref{fig:supp-AUC-all-r2} and \ref{fig:main-AUC-ntrain})
 
@@
 
-\subsection{Simulations varying number of SNPs and training size} %%%%
+\subsection{Simulations varying number of SNPs and sample size} %%%%
 
@@
 
-\subsection{Polygenic scores for the celiac disease}
+\subsection{Polygenic scores for celiac disease}
 
-Joint logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with PLR and 89.1\% with PLR3 as compared to 82.5\% with ``C+T-max'' (Table \ref{tab:results-celiac} and figure \ref{fig:celiac-roc}).
+Joint penalized logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with PLR and 89.1\% with PLR3 as compared to 82.5\% with ``C+T-max'' (Table \ref{tab:results-celiac} and figure \ref{fig:celiac-roc}).
 The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with PLR, PLR3 and ``C+T-max'', respectively.
@@
 {\color{red}
-\subsection{Penalized regression for the UK Biobank}
+\subsection{Polygenic scores for the UK Biobank}
 
@@
 By comparison, the best C+T model achieves a correlation of 55\% for women and 56\% for men (in the test set), and the GWAS part takes one hour (for the training set).
-If using only the top 100,000 SNPs from a GWAS on the training set to fit our L1-PLR, correlation between predicted and true heights drops at 0.634 for women and 0.643 for men.
+If using only the top 100,000 SNPs from a GWAS on the training set to fit our L1-PLR, correlation between predicted and true heights drops at 63.4\% for women and 64.3\% for men.
 Our L1-penalized logistic regression on breast cancer runs in 13 minutes for 150K women, achieving an AUC of 0.598 in the remaining 39K women, while the best C+T model achieves an AUC of 0.589, and the GWAS part takes 15 hours.
@@
   \hline
-height & PLR & 0.656 / 0.657 & 115,997 & 21h \\ 
-height & C+T-max & 0.549 / 0.561 & 45,570 & 69m \\ 
+height & PLR & 0.656 / 0.657 & 115,997 & 21 hours \\ 
+height & C+T-max & 0.549 / 0.561 & 45,570 & 69 minutes \\ 
   \hline
@@
   \hline
-breast cancer & PLR & 0.598 & 2653 & 13m \\ 
-breast cancer & C+T-max & 0.589 & 21 & 15h \\
+breast cancer & PLR & 0.598 & 2653 & 13 minutes \\ 
+breast cancer & C+T-max & 0.589 & 21 & 15 hours \\
   \hline
@@
 In such underpowered scenarios, it is difficult to detect true causal variants, which makes PLR too conservative whereas the best strategy is to include nearly all SNPs \cite[]{purcell2009common}.
+
 When increasing sample size (scenario \textnumero3), PLR achieves higher predictive performance than C+T and the benefits of using PLR over C+T increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}).
@@
 
-The choice of hyper-parameter values is very important since it can greatly impact method performance.
+The choice of hyper-parameter values is very important since it can greatly impact the performance of methods.
 In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps, respectively.
@@
 Indeed, on the one hand, choosing a low value for this threshold may discard informative SNPs that are correlated.
-Yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would add some noise to the PRS.
+Yet, on the other hand, when choosing a high value for this threshold, too much redundant information is included in the model, which adds some noise to the PRS.
 Based on the simulations, we find that using a stringent threshold ($r^2 = 0.05$) leads to higher predictive performance, even when causal SNPs are correlated.
-It means that, in most cases tested in this paper, avoiding redundant information is more important than including all causal SNPs.
+It means that, in most cases tested in this paper, avoiding redundant information in C+T is more important than including all causal SNPs.
 The choice of the $p_T$ threshold is also very important as it can greatly impact the predictive performance of the C+T method, which we confirm in this study \cite[]{ware2017heterogeneity}. 
@@
 {\color{red}
-We did not want to choose the C+T parameters as, to our knowledge, there is no gold standard to do so.
 So, for C+T, we report the best AUC value on the test set, even if it leads to over-optimistic results for C+T as compared to PLR.
@@
 
-The computation time of our PLR implementation mainly depends on the sample size and the number of active variables (variables that are included in the gradient descent).
+The computation time of our PLR implementation mainly depends on the sample size and the number of candidate variables (variables that are included in the gradient descent).
 Indeed, the algorithm is composed of two steps: first, for each variable, the algorithm computes an univariate statistic that is used to decide if the variable is included in the model (for each value of $\lambda$).
@@
 Then, the algorithm iterates over a regularization path of decreasing values of $\lambda$, which progressively enables variables to enter the model (Figure  \ref{fig:CMSA}).
-In the second step, the number of variables increases, and computations stop when an early stopping criterion is reached (when prediction is getting worse on the corresponding validation set, see figure \ref{fig:CMSA}).
+In the second step, the number of variables increases and computations stop when an early stopping criterion is reached (when prediction is getting worse on the corresponding validation set, see figure \ref{fig:CMSA}).
 
@@
 
-Memory requirements are tightly linked to computation time. Indeed, active variables are accessed in memory thanks to memory-mapping when they are used \cite[]{prive2017efficient}. 
+Memory requirements are tightly linked to computation time. Indeed, variables are accessed in memory thanks to memory-mapping when they are used \cite[]{prive2017efficient}. 
 When there is not enough memory left, the operating system (OS) frees some memory for new coming variables. 
-There should never be any memory issue thanks to memory-mapping.
-Yet, if there are too many active variables, the OS would regularly swap memory between disk and RAM, severely slowing down computations.
+%There should never be any memory issue thanks to memory-mapping. 
+Yet, if too many variables are used in the gradient descent, the OS would regularly swap memory between disk and RAM, severely slowing down computations. 
 %If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM. 
@@
 Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm in this paper.
-Indeed, when using only the 100K most significantly associated SNPs, this reduces correlation between predicted and true heights from 0.657/0.658 to 0.634/0.643 within women/men, while decreasing computation time by 20\% only (Table \ref{tab:results-UKBB}).
-A key advantage of our implementation of PLR is that prior filtering of variables is no more required for computational feasibility.
+Indeed, when using only the 100K most significantly associated SNPs, this reduces correlation between predicted and true heights from 0.657/0.658 to 0.634/0.643 within women/men. 
+A key advantage of our implementation of PLR is that prior filtering of variables is no more required for computational feasibility, thanks to the use of sequential strong rules and early stopping criterions.
 
@@
 A similar penalized linear regression is also available in R package bigstatsr.
-Our approach solves the dramatic computational burden faced by standard implementations, thus allowing for the analysis of large-scale datasets such as the UK biobank \cite[]{bycroft2017genome}.
+Our approach solves the dramatic memory and computational burdens faced by standard implementations, thus allowing for the analysis of large-scale datasets such as the UK biobank \cite[]{bycroft2017genome}.
 
 We also demonstrated in simulations and real datasets that our implementation of penalized regressions is highly effective over a broad range of disease architectures.
-It can be appropriate for predicting autoimmune diseases with a few strong effects (e.g.\ celiac disease) as well as highly polygenic traits (e.g.\ standing height).
+It can be appropriate for predicting autoimmune diseases with a few strong effects (e.g.\ celiac disease), as well as highly polygenic traits (e.g.\ standing height) providing that sample size is not too small.
 Finally, PLR as implemented in bigstatsr can also be used to predict phenotypes based on other omics data since our implementation is not specific to genotype data.

### 2019-02-12 paper/PRS.tex
finished

@@
 \abstract{
-Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
+Polygenic Risk Scores (PRS) consist in combining genotype information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
 PRS might have a major impact on public health, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.
@@
 Moreover, using 150,000 individuals of the UK Biobank, we are able to predict breast cancer better than C+T, fitting PLR in a few minutes only.
-This demonstrates the feasibility and relevance of using penalized regression for PRS computation when large individual-level datasets are available.
-
-In conclusion, our study demonstrates that penalized regression can provide discriminative polygenic risk scores, while being applicable to large-scale individual-level data because of an highly efficient implementation available in R package bigstatsr.
+In conclusion, this paper demonstrates the feasibility and relevance of using penalized regression for PRS computation when large individual-level datasets are available, thanks to the efficient implementation available in our R package bigstatsr.
 }
@@
 
-Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
+Polygenic Risk Scores (PRS) consist in combining genotype information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
 PRS are useful for genetic epidemiology when testing polygenicity of diseases and finding a common genetic contribution between two diseases \cite[]{purcell2009common}.
@@
 
-Several methods have been developed to predict disease status, or more generally any phenotype, based on SNP information.
+Several methods have been developed to predict disease status, or any phenotype, based on SNP information.
 A commonly used method often called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') is used to derive PRS from results of Genome-Wide Association Studies (GWAS) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
@@
 In parallel, statistical learning methods have also been used to derive PRS for complex human diseases by jointly estimating SNP effects.
-Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,wei2009disease}.
+Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,accuheight2018,mavaddat2019polygenic,okser2014regularized,wei2009disease}.
 Finally, Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly polygenic quantitative human phenotypes such as height \cite[]{yang2010common}.
 {\color{red}
-Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', have not proven as efficient as other methods for predicting several complex diseases based on genotypes (see table 2 of \cite{abraham2013performance}).
+Yet, predictions resulting from LMM, known e.g.\ as ``gBLUP'', have not proven as efficient as other methods for predicting several complex diseases based on genotypes (see table 2 of \cite{abraham2013performance}).
 }
@@
 {\color{red}
-In this comparison, we do not include any LMM method, yet, L2-PLR should be very similar to those methods (see section \ref{sec:PLR}).
+In this comparison, we do not include any LMM method, yet, L2-penalized logistic regression should be very similar to LMM methods (see section \ref{sec:PLR}).
 }
@@
 To perform simulations, we use real genotype data and simulate new phenotypes. 
-In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the disease heritability.
+In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as disease heritability.
 We also compare two different models for simulating phenotypes, one with additive effects only, and one that combines additive, dominant and interaction-type effects.
@@
 Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being scalable to biobank data.
-For example, this underpowered situation occurs when there are many causal variants (3K) to distinguish among many typed variants (280K) while using a small sample size (6K).
 }
@@
 We use real genotypes of European individuals from a case-control study for celiac disease \cite[]{dubois2010multiple}.
-The composition of this dataset is presented in table \ref{tab:celiac-data}.
+This dataset is presented in table \ref{tab:celiac-data}.
 Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
@@
 Liability scores are computed either from a model with additive effects only (``ADD'') or a more complex model that combines additive, dominant and interaction-type effects (``COMP'').
-For model ``ADD'', we compute the liability score of the i-th individual \[y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,\] where $S_\text{causal}$ is the set of causal SNPs, $w_j$ are weights generated from a Gaussian distribution $N(0, h^2 / \vert S_\text{causal} \vert)$ or a Laplace distribution $Laplace(0, \sqrt{h^2 / (2~\vert S_\text{causal} \vert)})$, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), and $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$.
+For model ``ADD'', we compute the liability score of the i-th individual as \[y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,\] where $S_\text{causal}$ is the set of causal SNPs, $w_j$ are weights generated from a Gaussian distribution $N(0, h^2 / \vert S_\text{causal} \vert)$ or a Laplace distribution $Laplace(0, \sqrt{h^2 / (2~\vert S_\text{causal} \vert)})$, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), and $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$.
 For model ``COMP'', we simulate liability scores using additive, dominant and interaction-type effects (see Supplementary Materials).
@@
 
-T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}.
+T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium \cite[]{botta2014exploiting}.
 We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees instead of 1000.
@@
 
-Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function \[L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(z_i\right) + (1 - y_i) \log\left(1 - z_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,\] where $z_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
+Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function \[L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(z_i\right) + (1 - y_i) \log\left(1 - z_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,\] where $z_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ denotes the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
 Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
@@
 To fit penalized logistic regression, we use an efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in R package bigstatsr.
-This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''.
+This algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''.
 To obtain an algorithm that does not require to choose this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
@@
 In this situation, the mean AUC reaches 92.5\% for PLR and 84\% for ``C+T-max'' (Figure \ref{fig:main-AUC-logit}).
-Note that, for the simulations, we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).
+For the simulations, we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).
 
@@
 
-Our implementation of L1-penalized linear regression runs in less than one day for 350K individuals (training set), achieving a correlation of more than 65.5\% with true height for both sexes in the remaining 24K individuals (test set).
+Our implementation of L1-penalized linear regression runs in less than one day for 350K individuals (training set), achieving a correlation of more than 65.5\% with true height for each sex in the remaining 24K individuals (test set).
 By comparison, the best C+T model achieves a correlation of 55\% for women and 56\% for men (in the test set), and the GWAS part takes one hour (for the training set).
@@
 The choice of hyper-parameter values is very important since it can greatly impact the performance of methods.
-In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps, respectively.
-The choice of the $r^2$ threshold of the clumping step is important. 
+In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps.
+For the clumping step, appropriately choosing the $r^2$ threshold is important. 
 Indeed, on the one hand, choosing a low value for this threshold may discard informative SNPs that are correlated.
-Yet, on the other hand, when choosing a high value for this threshold, too much redundant information is included in the model, which adds some noise to the PRS.
+On the other hand, when choosing a high value for this threshold, too much redundant information is included in the model, which adds noise to the PRS.
 Based on the simulations, we find that using a stringent threshold ($r^2 = 0.05$) leads to higher predictive performance, even when causal SNPs are correlated.
@@
 
-In contrast, for the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$ that accounts for the amount of regularization used in the model.
+In contrast, for penalized logistic regression, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$.
 Not only this procedure provides near-optimal results, but it also accelerates the model training thanks to the development of an early stopping criterion.
@@
 For this, we introduced a simple feature engineering technique that enables PLR to detect and learn not only additive effects, but also dominant and recessive effects.
-This technique improves the predictive performance of PLR when there are some non-linear effects in the simulations, while providing nearly the same predictive performance when there are additive effects only.
+This technique improves the predictive performance of PLR when there are non-linear effects in the simulations, while providing nearly the same predictive performance when there are additive effects only.
 Moreover, it also improves predictive performance for the celiac disease.
@@
 Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm in this paper.
-Indeed, when using only the 100K most significantly associated SNPs, this reduces correlation between predicted and true heights from 0.657/0.658 to 0.634/0.643 within women/men. 
-A key advantage of our implementation of PLR is that prior filtering of variables is no more required for computational feasibility, thanks to the use of sequential strong rules and early stopping criterions.
+Indeed, when using only the 100K most significantly associated SNPs, correlation between predicted and true heights is reduced from 0.656/0.657 to 0.634/0.643 within women/men. 
+A key advantage of our implementation of PLR is that prior filtering of variables is no more required for computational feasibility, thanks to the use of sequential strong rules and early stopping criteria.
 
@@
 
-Our approach has one major limitation: the main advantage of the C+T method is its direct applicability to summary statistics, allowing to leverage the largest GWAS results to date, even when individual cohort data cannot be merged because of practical or ethical reasons (e.g.\ consortium data including many cohorts).
-As of today, our implementation of penalized logistic regression does not allow for the analysis of summary data, but this represents an important future direction.
+Our approach has one major limitation: the main advantage of the C+T method is its direct applicability to summary statistics, allowing to leverage the largest GWAS results to date, even when individual cohort data cannot be merged because of practical or legal reasons.
+Our implementation of penalized logistic regression does not allow yet for the analysis of summary data, but this represents an important future direction.
 The current version is of particular interest for the analysis of modern individual-level datasets including hundreds of thousands of individuals.
@@
 In this comparative study, we have presented a computationally efficient implementation of penalized logistic regression that can be used to predict disease status based on genotypes.
-A similar penalized linear regression is also available in R package bigstatsr.
+A similar penalized linear regression for quantitative traits is also available in R package bigstatsr.
 Our approach solves the dramatic memory and computational burdens faced by standard implementations, thus allowing for the analysis of large-scale datasets such as the UK biobank \cite[]{bycroft2017genome}.
@@
 We also demonstrated in simulations and real datasets that our implementation of penalized regressions is highly effective over a broad range of disease architectures.
-It can be appropriate for predicting autoimmune diseases with a few strong effects (e.g.\ celiac disease), as well as highly polygenic traits (e.g.\ standing height) providing that sample size is not too small.
+It can be appropriate for predicting autoimmune diseases with a few strong effects (e.g.\ celiac disease), as well as highly polygenic traits (e.g.\ standing height) provided that sample size is not too small.
 Finally, PLR as implemented in bigstatsr can also be used to predict phenotypes based on other omics data since our implementation is not specific to genotype data.
@@
 
-Authors acknowledge LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01).
+Authors acknowledge LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01) and ANR project FROGH (ANR-16-CE12-0033).
 Authors also acknowledge the Grenoble Alpes Data Institute that is supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02).

### 2019-02-25 paper-accepted/PRS.tex
paper accepted
rm color, linenumbers and spacing

--- 2019-02-12 paper/PRS.tex
@@
 \usepackage{graphicx}
-\graphicspath{{../figures/}}
+%\graphicspath{{../figures/}}
 \DeclareMathOperator*{\argmin}{\arg\!\min}
@@
 
-{\color{red}
 In this paper, we present an efficient method for the joint estimation of SNP effects using individual-level data, allowing for practical application of penalized logistic regression (PLR) on modern datasets including hundreds of thousands of individuals.
@@
 We also provide an implementation of penalized linear regression for quantitative traits.
-}
 
 We compare the performance of PLR, C+T and a derivation of random forests using both real and simulated data. 
-{\color{red}
 Overall, we find that PLR achieves equal or higher predictive performance than C+T in most scenarios considered, while being scalable to biobank data.
-}
 In particular, we find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; for instance, in simulations, AUC values increase from 83\% with the best prediction of C+T to 92.5\% with PLR.
@@
 
-{\color{red}
 Applying penalized linear regression to 350,000 individuals of the UK Biobank, we predict height with a larger correlation than with the best prediction of C+T (\textasciitilde65\% instead of \textasciitilde55\%), further demonstrating its scalability and strong predictive power, even for highly polygenic traits.
@@
 In conclusion, this paper demonstrates the feasibility and relevance of using penalized regression for PRS computation when large individual-level datasets are available, thanks to the efficient implementation available in our R package bigstatsr.
-}
 }
@@
 As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}.
-{\color{red}
 However, in order to be used in clinical settings, PRS should discriminate well enough between cases and controls.
 For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that, for a 10\% disease prevalence, the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.
-}
 
@@
 Finally, Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly polygenic quantitative human phenotypes such as height \cite[]{yang2010common}.
-{\color{red}
 Yet, predictions resulting from LMM, known e.g.\ as ``gBLUP'', have not proven as efficient as other methods for predicting several complex diseases based on genotypes (see table 2 of \cite{abraham2013performance}).
-}
 
@@
 In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression (PLR), which we compare to the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
-{\color{red}
 In this comparison, we do not include any LMM method, yet, L2-penalized logistic regression should be very similar to LMM methods (see section \ref{sec:PLR}).
-}
 Moreover, we do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}. 
@@
 We also compare two different models for simulating phenotypes, one with additive effects only, and one that combines additive, dominant and interaction-type effects.
-{\color{red}
 Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being scalable to biobank data.
-}
 
@@
 This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 autosomal chromosomes or 30 in the HLA region of chromosome 6), and the disease heritability $h^2$ (50\% or 80\%).
-%SNP effects are drawn from either a Gaussian or from a Laplace distribution.
 Liability scores are computed either from a model with additive effects only (``ADD'') or a more complex model that combines additive, dominant and interaction-type effects (``COMP'').
@@
 As we report the optimal threshold based on the test set, the AUC for ``C+T-max'' is an upper bound of the AUC for the C+T method.
-{\color{red}
 Here, the GWAS part uses the training set while clumping uses the test set (all individuals not included in the training set).
-}
 
@@
 Using 1000 trees provides a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown).
-%We call this method ``T-Trees'' in the results.
 
@@
 the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $p \gg n$ situation ($p$ is the number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}.
-{\color{red}
 In this study, we use a grid search over $\alpha \in \{1,~0.5,~0.05,~0.001\}$.
@@
 Using $\alpha = 0.001$ should result in a model very similar to gBLUP.\label{sec:PLR}
-}
 
@@
 
-{\color{red}
 \subsection{Polygenic scores for the UK Biobank}
@@
 
-}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently.
-{\color{red}
 PLR always outperforms the C+T method, except in some highly underpowered cases (AUC values always less than 0.6), and the benefits of using PLR are more pronounced with an increasing sample size or when causal SNPs are correlated with one another.
@@
 In this scenario \textnumero2, even when there are small effects and a high polygenicity (3000 causal variants out of 18,941), PLR gets a large increase in predictive performance, now consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
-}
 
@@
 To our knowledge, there is no clear standard on how to choose these two critical hyper-parameters for C+T.
-{\color{red}
 So, for C+T, we report the best AUC value on the test set, even if it leads to over-optimistic results for C+T as compared to PLR.
-}
 
@@
 Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we use a grid  search directly embedded in the CMSA procedure.
-%Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model.
-Yet, if using e.g.\ $\alpha = 0.001$, it would lead to a similar model.
 
@@
 
-{\color{red}
 \subsection{Time and memory requirements}
@@
 On the contrary, for traits like celiac disease or breast cancer that are less polygenic, the number of variables included in the model is much smaller so that fitting is very fast (13 minutes only for 150K women of the UK Biobank for breast cancer).
-%Using more cases (there were only 3542 for Celiac and 6768 for breast cancer), one might achieve better predictive performance for these two diseases, and the computation time should remain within a few hours.
 
@@
 When there is not enough memory left, the operating system (OS) frees some memory for new coming variables. 
-%There should never be any memory issue thanks to memory-mapping. 
 Yet, if too many variables are used in the gradient descent, the OS would regularly swap memory between disk and RAM, severely slowing down computations.
-%If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM. 
-%Another solution would be to reduce the `n.abort' parameter (default value is 10) in order to stop fitting the regularization path even sooner. 
-%This would avoid fitting models with a large number of variables without changing the solution since regularization paths for very large sample sizes are very smooth (Figure \ref{fig:supp-path}).
-%We tested processing each chromosome separately and then combining the models later but this did not perform as well as fitting all SNPs at once (data not shown). 
 A possible approach to reduce computational burden is to apply penalized regression on a subset of SNPs by prioritizing SNPs using univariate tests (GWAS computed from the same dataset).
@@
 A key advantage of our implementation of PLR is that prior filtering of variables is no more required for computational feasibility, thanks to the use of sequential strong rules and early stopping criteria.
-
-}
 
@@
 Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}.
-%In future work, we aim at exploring how we can overcome those issues.
-%We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}.
-Finally, we also wish to assess how can we combine the information provided by genetic data with clinical and environmental data, possibly in a non-linear way.
 
@@
 
-Supplemental Data include a PDF with two sections of methods, two tables and ten figures. 
+Supplemental Data include a PDF with two sections of methods, two tables and eight figures. 
 Supplemental Data also include six HTML R notebooks including all code and results used in this paper, for reproducibility purposes, and available at \url{https://figshare.com/articles/code/7178750}.

### 2019-02-26 paper-accepted/PRS.tex
use Genetics format

@@
 %% Do not edit unless you really know what you are doing.
-\documentclass[english, 12pt]{article}
+\documentclass[9pt,twocolumn,twoside]{gsajnl}
 \usepackage{times}
@@
 \usepackage{bbm}
-\usepackage[T1]{fontenc}
-\usepackage[latin1]{inputenc}
-\usepackage{geometry}
-\geometry{verbose,letterpaper,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
-\usepackage{rotating}
-\usepackage{color}
+%\usepackage[T1]{fontenc}
+%\usepackage[latin1]{inputenc}
+%\usepackage{geometry}
 \usepackage{graphicx}
@@
 
-%\linenumbers
-%\doublespacing
-%\usepackage[authoryear]{natbib}
-\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}
-
-%Pour les rajouts
-\usepackage{color}
-\definecolor{trustcolor}{rgb}{0,0,1}
-
-\usepackage{dsfont}
-\usepackage[warn]{textcomp}
 \usepackage{adjustbox}
 \usepackage{multirow}
-\usepackage{graphicx}
-%\graphicspath{{../figures/}}
-\DeclareMathOperator*{\argmin}{\arg\!\min}
 
@@
 \let\tabend\endtabular
-\renewenvironment{tabular}{\begin{adjustbox}{max width=\textwidth}\tabbeg}{\tabend\end{adjustbox}}
+\renewenvironment{tabular}{\begin{adjustbox}{max width=\columnwidth}\tabbeg}{\tabend\end{adjustbox}}
 
-\makeatletter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
-%% Bold symbol macro for standard LaTeX users
-%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}
-
-%% Because html converters don't know tabularnewline
-\providecommand{\tabularnewline}{\\}
-
-\usepackage{babel}
-\makeatother
-
-\begin{document}
+\newcommand{\ftextnumero}{{\fontfamily{txr}\selectfont \textnumero}}
 
 \title{Efficient implementation of penalized regression\\for genetic risk prediction}
-\author{Florian Priv\'e,$^{\text{1,}*}$ Hugues Aschard$^{\text{2}}$ and Michael G.B. Blum$^{\text{1,}*}$}
 
-\date{~ }
-\maketitle
+\author[$\ast$,1]{Florian Priv\'e}
+\author[2]{Hugues Aschard}
+\author[$\ast$,1]{Michael G.B. Blum}
 
-\noindent$^{\text{\sf 1}}$Laboratoire TIMC-IMAG, UMR 5525, Univ.\ Grenoble Alpes, CNRS, La Tronche (38700), France, \\
-\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris (75015), France.
+\affil[$\ast$]{To whom correspondence should be addressed.}
+\affil[$1$]{Laboratoire TIMC-IMAG, UMR 5525, Univ.\ Grenoble Alpes, CNRS, La Tronche (38700), France}
+\affil[$2$]{Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris (75015), France}
 
-\noindent$^\ast$To whom correspondence should be addressed.\\
+\keywords{Polygenic Risk Scores; SNP; LASSO}
 
-\noindent Contacts: 
-\begin{itemize}
-\item \url{florian.prive@univ-grenoble-alpes.fr}
-\item \url{hugues.aschard@pasteur.fr}
-\item \url{michael.blum@univ-grenoble-alpes.fr}
-\end{itemize}
+\runningtitle{Efficient penalized regression for PRS} % For use in the footer
 
-\newpage
+%% For the footnote.
+%% Give the last name of the first author if only one author;
+% \runningauthor{FirstAuthorLastname}
+%% last names of both authors if there are two authors;
+% \runningauthor{FirstAuthorLastname and SecondAuthorLastname}
+%% last name of the first author followed by et al, if more than two authors.
+\runningauthor{Priv\'e \textit{et al.}}
 
-\abstract{
+\begin{abstract}
 Polygenic Risk Scores (PRS) consist in combining genotype information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease.
@@
 In conclusion, this paper demonstrates the feasibility and relevance of using penalized regression for PRS computation when large individual-level datasets are available, thanks to the efficient implementation available in our R package bigstatsr.
-}
+\end{abstract}
 
@@
 
-\newpage
+\begin{document}
+
+\maketitle
+\thispagestyle{firststyle}
+\marginmark
+\firstpagefootnote
+
+\noindent Contacts: 
+\begin{itemize}
+\item \url{florian.prive@univ-grenoble-alpes.fr}
+\item \url{hugues.aschard@pasteur.fr}
+\item \url{michael.blum@univ-grenoble-alpes.fr}
+\end{itemize}
 
@@
 Several methods have been developed to predict disease status, or any phenotype, based on SNP information.
-A commonly used method often called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') is used to derive PRS from results of Genome-Wide Association Studies (GWAS) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}.
+A commonly used method often called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') is used to derive PRS from results of Genome-Wide Association Studies (GWAS) \cite[]{wray2007prediction,evans2009harnessing,purcell2009common,chatterjee2013projecting,dudbridge2013power}.
 This technique uses GWAS summary statistics, allowing for a fast implementation of C+T.
@@
 In parallel, statistical learning methods have also been used to derive PRS for complex human diseases by jointly estimating SNP effects.
-Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,accuheight2018,mavaddat2019polygenic,okser2014regularized,wei2009disease}.
+Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{wei2009disease,abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,accuheight2018,mavaddat2019polygenic}.
 Finally, Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly polygenic quantitative human phenotypes such as height \cite[]{yang2010common}.
@@
 In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression (PLR), which we compare to the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
-In this comparison, we do not include any LMM method, yet, L2-penalized logistic regression should be very similar to LMM methods (see section \ref{sec:PLR}).
+In this comparison, we do not include any LMM method, yet, L2-penalized logistic regression should be very similar to LMM methods.
 Moreover, we do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}. 
@@
 We implement 3 different simulation scenarios, summarized in table \ref{tab:simus}.
-Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes -- 281,122 SNPs) and a training set of size 6000.
-It compares all methods described in section \ref{sec:methods}.
+Scenario \ftextnumero1 uses the whole dataset (all 22 autosomal chromosomes -- 281,122 SNPs) and a training set of size 6000.
 For each combination of the remaining parameters, results are based on 100 simulations except when comparing PLR with T-Trees, which relies on 5 simulations only because of a much higher computational burden of T-Trees as compared to other methods.
-Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only (18,941 SNPs).
+Scenario \ftextnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only (18,941 SNPs).
 Reducing the number of SNPs increases the polygenicity (the proportion of causal SNPs) of the simulated models.
@@
 For this scenario, we use the additive model only, but continue to vary all other simulation parameters.
-Finally, scenario \textnumero3 uses the whole dataset as in scenario \textnumero1 while varying the size of the training set in order to assess how the sample size affects predictive performance of methods.
+Finally, scenario \ftextnumero3 uses the whole dataset as in scenario \ftextnumero1 while varying the size of the training set in order to assess how the sample size affects predictive performance of methods.
 A total of 100 simulations per combination of parameters are run using 300 causal SNPs randomly chosen on the genome.
@@
 In this study, we use two different measures of predictive accuracy.
-First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction,lusted1971signal}.
+First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{lusted1971signal,fawcett2006introduction}.
 In the case of our study, the AUC is the probability that the PRS of a case is greater than the PRS of a control.
 This measure indicates the extent to which we can distinguish between cases and controls using PRS. 
-As a second measure, we also report the partial AUC for specificities between 90\% and 100\% \cite[]{dodd2003partial,mcclish1989analyzing}.
+As a second measure, we also report the partial AUC for specificities between 90\% and 100\% \cite[]{mcclish1989analyzing,dodd2003partial}.
 This measure is similar to the AUC, but focuses on high specificities, which is the most useful part of the ROC curve in clinical settings.
@@
 
-Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function \[L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(z_i\right) + (1 - y_i) \log\left(1 - z_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,\] where $z_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ denotes the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
+Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function 
+\begin{equation}
+\begin{split}
+L(\lambda, \alpha) = & \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(z_i\right) + (1 - y_i) \log\left(1 - z_i\right) \right) }_\text{Loss function} \\
+ & +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,
+\end{split}
+\end{equation}
+where $z_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ denotes the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
 Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
@@
 This grid-search is directly embedded in our PLR implementation for simplicity.
-Using $\alpha = 0.001$ should result in a model very similar to gBLUP.\label{sec:PLR}
+Using $\alpha = 0.001$ should result in a model very similar to gBLUP.
 
@@
 
+\subsection{Data Availability Statement}
+
+Supplemental Data include a PDF with two sections of methods, two tables and eight figures. 
+Supplemental Data also include six HTML R notebooks including all code and results used in this paper, for reproducibility purposes, and available at \url{https://figshare.com/articles/code/7178750}.
+Additional analyses of the UK Biobank are available as three R scripts at \url{https://figshare.com/articles/code_UKB/7531559}.
+Results of simulations are available at \url{https://figshare.com/articles/results_zip/7126964}.
+A tutorial on how to start with R packages bigstatsr and bigsnpr is available at \url{https://privefl.github.io/bigsnpr/articles/demo.html}.
+The two R packages are available on GitHub.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We compared penalized logistic regression (PLR) with the C+T method using simulations of scenario \textnumero1 (Table \ref{tab:simus}).
+We compared penalized logistic regression (PLR) with the C+T method using simulations of scenario \ftextnumero1 (Table \ref{tab:simus}).
 When simulating a model with 30 causal SNPs and a heritability of 80\%, PLR provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\% for this setting (Figure \ref{fig:main-AUC-logit}).
@@
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
-\caption{Main comparison of C+T and PLR when simulating phenotypes with additive effects (scenario \textnumero1, model ``ADD'').
+\centerline{\includegraphics[width=\columnwidth]{main-AUC-logit}}
+\caption{Main comparison of C+T and PLR when simulating phenotypes with additive effects (scenario \ftextnumero1, model ``ADD'').
 Mean AUC over 100 simulations for PLR and the maximum AUC reported with ``C+T-max'' (clumping threshold at $r^2 > 0.2$).
@@
 
-We tested the T-Trees method in scenario \textnumero1.
+We tested the T-Trees method in scenario \ftextnumero1.
 As compared to PLR, T-Trees perform worse in terms of predictive ability, while taking much longer to run (Figure \ref{fig:supp-ttrees}). 
@@
 
-We also compared the two penalized logistic regressions in scenario \textnumero1: PLR versus PLR3 that uses additional features (variables) coding for recessive and dominant effects.
+We also compared the two penalized logistic regressions in scenario \ftextnumero1: PLR versus PLR3 that uses additional features (variables) coding for recessive and dominant effects.
 Predictive performance of PLR3 are nearly as good as PLR when there are additive effects only (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
@@
 
-First, when reproducing simulations of scenario \textnumero1 using chromosome 6 only (scenario \textnumero2), the predictive performance of PLR always increase (Figure \ref{fig:supp-AUC-all-r2}). 
+First, when reproducing simulations of scenario \ftextnumero1 using chromosome 6 only (scenario \ftextnumero2), the predictive performance of PLR always increase (Figure \ref{fig:supp-AUC-all-r2}). 
 There is a particularly large increase when simulating 3000 causal SNPs: AUC from PLR increases from 60\% to nearly 80\% for Gaussian effects and a disease heritability of 80\%.
 On the contrary, when simulating only 30 or 300 causal SNPs with the corresponding dataset, AUC of ``C+T-max'' does not increase, and even decreases for a heritability of 80\% (Figure \ref{fig:supp-AUC-all-r2}).
-Secondly, when varying the training size (scenario \textnumero3), we report an increase of AUC with a larger training size, with a faster increase of AUC for PLR as compared to ``C+T-max'' (Figure  \ref{fig:main-AUC-ntrain}).
+Secondly, when varying the training size (scenario \ftextnumero3), we report an increase of AUC with a larger training size, with a faster increase of AUC for PLR as compared to ``C+T-max'' (Figure  \ref{fig:main-AUC-ntrain}).
 
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
-\caption{Comparison of methods when simulating phenotypes with additive effects and using chromosome 6 only (scenario \textnumero2).
-Thinner lines represents results in scenario \textnumero1.
+\centerline{\includegraphics[width=\columnwidth]{supp-AUC-chr6-all-r2}}
+\caption{Comparison of methods when simulating phenotypes with additive effects and using chromosome 6 only (scenario \ftextnumero2).
+Thinner lines represents results in scenario \ftextnumero1.
 Mean AUC over 100 simulations for PLR and the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) as a function of the number and location of causal SNPs.
@@
 \begin{figure}[h]
-\centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
-\caption{Comparison of methods when simulating 300 causal SNPs with additive effects and when varying sample size (scenario \textnumero3).
+\centerline{\includegraphics[width=\columnwidth]{main-AUC-ntrain}}
+\caption{Comparison of methods when simulating 300 causal SNPs with additive effects and when varying sample size (scenario \ftextnumero3).
 Mean AUC over 100 simulations for the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the training size.
@@
 
-When increasing sample size (scenario \textnumero3), PLR achieves higher predictive performance than C+T and the benefits of using PLR over C+T increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}).
-Moreover, when decreasing the search space (total number of candidate SNPs) in scenario \textnumero2, we increase the proportion of causal variants and we virtually increase the sample size \cite[]{dudbridge2013power}.
-In this scenario \textnumero2, even when there are small effects and a high polygenicity (3000 causal variants out of 18,941), PLR gets a large increase in predictive performance, now consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
+When increasing sample size (scenario \ftextnumero3), PLR achieves higher predictive performance than C+T and the benefits of using PLR over C+T increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}).
+Moreover, when decreasing the search space (total number of candidate SNPs) in scenario \ftextnumero2, we increase the proportion of causal variants and we virtually increase the sample size \cite[]{dudbridge2013power}.
+In this scenario \ftextnumero2, even when there are small effects and a high polygenicity (3000 causal variants out of 18,941), PLR gets a large increase in predictive performance, now consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
 
@@
 
-Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}.
+Finally, in this comparative study, we did not consider the problem of population structure \cite[]{vilhjalmsson2015modeling,marquez2017multiethnic,martin2017human} and also did not consider non-genetic data such as environmental and clinical data \cite[]{van2012integration,dey2013integration}.
 
@@
 
-\clearpage
-
-\section*{Description of Supplemental Data}
-
-Supplemental Data include a PDF with two sections of methods, two tables and eight figures. 
-Supplemental Data also include six HTML R notebooks including all code and results used in this paper, for reproducibility purposes, and available at \url{https://figshare.com/articles/code/7178750}.
-Additional analyses of the UK Biobank are available as 3 R scripts at \url{https://figshare.com/articles/code_UKB/7531559}.
-
-\section*{Declaration of Interests}
-
-The authors declare no competing interests.
-
 \section*{Acknowledgements}
@@
 
-\section*{Web Resources}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-Results of simulations are available at \url{https://figshare.com/articles/results_zip/7126964}.
-A tutorial on how to start with R packages bigstatsr and bigsnpr is available at \url{https://privefl.github.io/bigsnpr/articles/demo.html}.
-The two R packages are available on GitHub.
-
-\newpage
-
-\bibliographystyle{natbib}
 \bibliography{refs}

### 2019-02-26 paper-accepted/PRS.tex
hardcode supmat

@@
 \usepackage{bbm}
-
-\usepackage{xr}
-\externaldocument{PRS-supp}
 
@@
 We use real genotypes of European individuals from a case-control study for celiac disease \cite[]{dubois2010multiple}.
-This dataset is presented in table \ref{tab:celiac-data}.
+This dataset is presented in table S1.
 Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
@@
 In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$.
-We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``C+T-all''), one including only the SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``C+T-stringent''), and one that maximizes the AUC (``C+T-max'') for 102 p-value thresholds between $1$ and $10^{-100}$ (Table \ref{tab:thr}).
+We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``C+T-all''), one including only the SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``C+T-stringent''), and one that maximizes the AUC (``C+T-max'') for 102 p-value thresholds between $1$ and $10^{-100}$ (Table S2).
 As we report the optimal threshold based on the test set, the AUC for ``C+T-max'' is an upper bound of the AUC for the C+T method.
@@
 This algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''.
-To obtain an algorithm that does not require to choose this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
+To obtain an algorithm that does not require to choose this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure S1).
 Because of L1-regularization, the resulting vector of estimated effect sizes is sparse. 
@@
 In this situation, the mean AUC reaches 92.5\% for PLR and 84\% for ``C+T-max'' (Figure \ref{fig:main-AUC-logit}).
-For the simulations, we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).
+For the simulations, we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure S3).
 
@@
 In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can dramatically impact the predictive performance of C+T. 
-For example, in a model with 30 causal SNPs, AUC ranges from less than 60\% when using all SNPs passing clumping to 90\% \emph{if} choosing the optimal p-value threshold (Figure \ref{fig:supp-AUC-PRS}).
+For example, in a model with 30 causal SNPs, AUC ranges from less than 60\% when using all SNPs passing clumping to 90\% \emph{if} choosing the optimal p-value threshold (Figure S4).
 
@@
 
-Our implementation of PLR that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of 100 models corresponding to different values of $\lambda$ (Figure \ref{fig:supp-biglasso}).
+Our implementation of PLR that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of 100 models corresponding to different values of $\lambda$ (Figure S8).
 
@@
 We tested the T-Trees method in scenario \ftextnumero1.
-As compared to PLR, T-Trees perform worse in terms of predictive ability, while taking much longer to run (Figure \ref{fig:supp-ttrees}). 
-Even for simulations with model ``COMP'' in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower when using T-Trees than when using PLR (Figure \ref{fig:supp-ttrees}).
+As compared to PLR, T-Trees perform worse in terms of predictive ability, while taking much longer to run (Figure S5). 
+Even for simulations with model ``COMP'' in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower when using T-Trees than when using PLR (Figure S5).
 
 We also compared the two penalized logistic regressions in scenario \ftextnumero1: PLR versus PLR3 that uses additional features (variables) coding for recessive and dominant effects.
-Predictive performance of PLR3 are nearly as good as PLR when there are additive effects only (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}).
+Predictive performance of PLR3 are nearly as good as PLR when there are additive effects only (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures S6 and S7).
 For model ``COMP'', PLR3 provides AUC values at least 3.5\% higher than PLR, except when there are 3000 causal SNPs.
@@
 
-Joint penalized logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with PLR and 89.1\% with PLR3 as compared to 82.5\% with ``C+T-max'' (Table \ref{tab:results-celiac} and figure \ref{fig:celiac-roc}).
+Joint penalized logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with PLR and 89.1\% with PLR3 as compared to 82.5\% with ``C+T-max'' (Table \ref{tab:results-celiac} and figure S2).
 The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with PLR, PLR3 and ``C+T-max'', respectively.
@@
 This first step is very fast.
-Then, the algorithm iterates over a regularization path of decreasing values of $\lambda$, which progressively enables variables to enter the model (Figure  \ref{fig:CMSA}).
-In the second step, the number of variables increases and computations stop when an early stopping criterion is reached (when prediction is getting worse on the corresponding validation set, see figure \ref{fig:CMSA}).
+Then, the algorithm iterates over a regularization path of decreasing values of $\lambda$, which progressively enables variables to enter the model (Figure  S1).
+In the second step, the number of variables increases and computations stop when an early stopping criterion is reached (when prediction is getting worse on the corresponding validation set, see figure S1).
 

### 2020-10-28 response-snpnet/code/assess-pred.R
export code

@@ -1,0 +1,85 @@
+library(bigsnpr)
+ukbb <- snp_attach("data/ukbb.rds")
+G <- ukbb$genotypes
+
+df0 <- readRDS("data/pheno_covar.rds")
+covar <- dplyr::select(df0, sex, age, PC1:PC16)
+ind.test <- which(df0$set == 5)
+
+# perf <- function(pred, target) {
+#   is_not_na <- which(!is.na(target))
+#   AUC(pred[is_not_na], target[is_not_na])
+# }
+
+perf <- function(pred, target) {
+  my_pcor <- function(x, y, z) {
+    df <- cbind.data.frame(z, x = x, y = y)
+    cor(lm(x ~ . - y, data = df)$residuals,
+        lm(y ~ . - x, data = df)$residuals,
+        use = "pairwise.complete.obs")
+  }
+  my_pcor(pred, target, covar[ind.test, ])
+}
+
+# name <- "asthma"
+# name <- "high_cholesterol"
+
+for (name in c("height", "BMI", "asthma", "high_cholesterol")) {
+
+  res <- readRDS(paste0("bigstatsr/", name, ".rds"))
+  y <- df0[[name]]
+
+  beta_one <- head(res$mod[[1]][[1]]$beta, -ncol(covar))
+  ind <- which(beta_one != 0)
+  pred_one <- big_prodVec(G, beta_one[ind], ind.row = ind.test,
+                          ind.col = ind, ncores = 15)
+
+  pred_all <- predict(res$mod, G, ind.row = ind.test,
+                      covar.row = matrix(0, length(ind.test), ncol(covar)),
+                      proba = FALSE, ncores = 15)
+
+  res2 <- readRDS(paste0("snpnet2/", name, ".rds"))
+  id <- which.max(res2$mod$metric.val)
+  beta_snpnet <- tail(res2$mod$beta[[id]], -ncol(covar))
+  ind <- match(names(beta_snpnet), with(ukbb$map, paste0(marker.ID, "_", allele1)))
+  pred_snpnet <- big_prodVec(G, beta_snpnet, ind.row = ind.test,
+                             ind.col = ind, ncores = 15)
+
+  res3 <- readRDS(paste0("snpnet2/", name, "2.rds"))
+  id <- length(res3$mod$metric.train)
+  beta_snpnet2 <- tail(res3$mod$beta[[id]], -ncol(covar))
[…]

### 2020-10-28 response-snpnet/code/prepare-data.R

@@ -1,0 +1,99 @@
+##### Prepare bed file (merge + QC) ####
+
+# file.symlink("~/NCRR-PRS/faststorage/UKBB/", ".")
+
+write(sapply(1:22, function(chr) {
+  paste0("UKBB/bed/",
+         c(paste0("ukb_cal_chr", chr, "_v2.bed"),
+           paste0("ukb_snp_chr", chr, "_v2.bim"),
+           paste0("ukb58024_cal_chr", chr, "_v2_s488264.fam")))
+}), tmp <- tempfile(), ncolumns = 3)
+
+library(bigsnpr)
+snp_plinkQC(
+  plink.path = download_plink("tmp-data"),
+  prefix.in = tmp,
+  file.type = "--merge-list",
+  prefix.out = "data/ukbb",
+  geno = 0.01,
+  mind = 0.1,
+  maf = 0.01,
+  autosome.only = TRUE,
+  extra.options = "--memory 100000"
+)
+# 504,139 variants and 488,371 people pass filters and QC.
+
+
+#### Prepare subset, phenotypes and covariates ####
+
+library(dplyr)
+library(bigreadr)
+csv <- "UKBB/ukb41181.csv"
+df0 <- fread2(
+  csv,
+  select = c("eid", "50-0.0", "21001-0.0", "21022-0.0", "22001-0.0", "22006-0.0",
+             "22020-0.0", paste0("22009-0.", 1:16)),
+  col.names = c("eid", "height", "BMI", "age", "sex", "white_british",
+                "used_in_pca", paste0("PC", 1:16))
+) %>%
+  filter(!is.na(eid), white_british == 1, used_in_pca == 1) %>%
+  select(-white_british, -used_in_pca)
+# 337,475 individuals remaining
+
+# Sets: #5 will be test (20%), 3 of them will be training (60%)
+#       and the last will be validation (20%)
+df0$set <- sample(rep_len(1:5, nrow(df0)))
+
+# Read from bed to bk/rds
+system.time(
+  snp_readBed2("data/ukbb.bed",
+               ind.row = match(df0$eid, fread2("data/ukbb.fam")[[2]]),
+               ncores = 16)
+) # 6 min
+
+# Impute with mean (+ save with new code.256 CODE_DOSAGE)
+ukb <- snp_attach("data/ukbb.rds")
+system.time(
[…]

### 2020-10-28 response-snpnet/code/run-bigstatsr.R

@@ -1,0 +1,45 @@
+library(bigsnpr)
+library(dplyr)
+
+ukbb <- snp_attach("data/ukbb.rds")
+G <- ukbb$genotypes
+file.size(G$bk) / 1024^3  # 158 GB
+
+df0 <- readRDS("data/pheno_covar.rds")
+covar <- select(df0, sex, age, PC1:PC16)
+pheno <- select(df0, height, BMI, asthma, high_cholesterol)
+set <- df0$set
+
+bigassertr::assert_dir("bigstatsr")
+bigassertr::assert_dir("log")
+
+library(future.batchtools)
+NCORES <- 16
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES, mem = "128g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_imap(pheno, ~ {
+
+  res_file <- paste0("bigstatsr/", .y, ".rds")
+  if (file.exists(res_file)) return(NULL)
+
+  y <- .x
+  ind.train <- which(!is.na(y) & complete.cases(covar) & set != 5)
+
+  FUN <- if (all(y[ind.train] %in% 0:1)) big_spLogReg else big_spLinReg
+
+  ## Penalized reg
+  time <- system.time(
+    mod <- FUN(G, y[ind.train], ind.train = ind.train,
+               covar.train = as.matrix(covar[ind.train, ]),
+               pf.covar = rep(0, ncol(covar)),
+               alphas = 1,         # lasso
+               power_scale = 0,    # not scaled
+               ind.sets = set[ind.train], K = 4,
+               n.abort = 2, nlam.min = 30,
+               ncores = NCORES)
+  )[3]
+
+  saveRDS(list(mod = mod, time = time), res_file)
+})

### 2020-10-28 response-snpnet/code/run-snpnet.R

@@ -1,0 +1,108 @@
+plink2 <- bigsnpr::download_plink2("tmp-data")
+
+ids <- bigsnpr::snp_attach("data/ukbb.rds")$fam[1:2]
+
+if (!file.exists("data/ukbb.pgen")) {
+  bigsnpr:::write.table2(ids, tmp <- tempfile())
+  system(glue::glue("{plink2} --bfile data/ukbb --keep {tmp}",
+                    " --make-pgen vzs --out data/ukbb",
+                    " --threads 16 --memory 200000"))
+}
+file.size("data/ukbb.pgen") / 1024^3  # 22 GB
+
+# Make sure individuals are in the sale order
+library(bigreadr)
+all.equal(fread2("data/ukbb.psam", select = 1:2), ids, check.attributes = FALSE)
+
+library(dplyr)
+df0 <- readRDS("data/pheno_covar.rds")
+covar <- select(df0, sex, age, PC1:PC16)
+pheno <- select(df0, height, BMI, asthma, high_cholesterol)
+set <- df0$set
+split <- case_when(set == 5 ~ "test", set == 1 ~ "val", TRUE ~ "train")
+table(split)
+#  test  train    val
+# 67495 202485  67495
+
+phenotype.file <- bigreadr::fwrite2(cbind.data.frame(
+  FID = ids[[1]],
+  IID = ids[[2]],
+  split,
+  split_refit = case_when(set == 5 ~ "val", TRUE ~ "train"),
+  pheno,
+  covar
+), "data/pheno_snpnet.txt")
+
+bigassertr::assert_dir("snpnet2")
+bigassertr::assert_dir("log")
+
+NCORES <- 16
+library(future.batchtools)
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES, mem = "500g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_iwalk(pheno, ~ {
+
+  res_file <- paste0("snpnet2/", .y, ".rds")
+  if (file.exists(res_file)) return(NULL)
+
+  unlink(res_dir <- paste0("tmp-data/", .y, 2), recursive = TRUE)
+
+  configs <- list(plink2.path = plink2, nCores = NCORES, niter = 100,
+                  use.glmnetPlus = TRUE, results.dir = res_dir,
+                  early.stopping = TRUE, stopping.lag = 2)
+
+  ## Penalized reg
+  library(snpnet)
+  time <- system.time(
[…]

### 2020-10-29 response-snpnet/code/run-snpnet.R
rerun BMI

@@ -39,7 +39,7 @@ bigassertr::assert_dir("log")
 NCORES <- 16
 library(future.batchtools)
 plan(batchtools_slurm(resources = list(
-  t = "12:00:00", c = NCORES, mem = "500g",
+  t = "2-00:00", c = NCORES, mem = "500g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
 furrr::future_iwalk(pheno, ~ {

### 2020-12-17 response-snpnet/supp-note-snpnet-comparison.tex
reformat as supp note

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english, 12pt]{article}
\usepackage{times}
%\usepackage{algorithm2e}
\usepackage{url}
\usepackage{bbm}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=1.5cm,rmargin=1.5cm}
\usepackage{rotating}
\usepackage{color}
\usepackage{graphicx}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{bbm}
\usepackage{makecell}

\renewcommand{\arraystretch}{1.2}

%\usepackage{xr}
%\externaldocument{paper-ldpred2-supp}

%\linenumbers
%\doublespacing
\onehalfspacing
%\usepackage[authoryear]{natbib}
\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}

%Pour les rajouts
\usepackage{color}
\definecolor{trustcolor}{rgb}{0,0,1}

\usepackage{dsfont}
\usepackage[warn]{textcomp}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{graphicx}
\graphicspath{{../figures/}}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\usepackage{algorithm} 
\usepackage{algpseudocode}

\let\tabbeg\tabular
\let\tabend\endtabular
\renewenvironment{tabular}{\begin{adjustbox}{max width=0.9\textwidth}\tabbeg}{\tabend\end{adjustbox}}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Bold symbol macro for standard LaTeX users
%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}

%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

\usepackage{babel}
\makeatother

\begin{document}

\title{Supplementary Note: Comparison between bigstatsr and snpnet\\for fitting penalized regressions on very large genetic data}

\date{~ }
\maketitle

Penalized regression with L1 penalty, also known as ``lasso'', has been widely used since it proved to be an effective method for simultaneously performing variable selection and model fitting \cite[]{tibshirani1996regression}.
R package glmnet is a popular software to fit the lasso efficiently \cite[]{friedman2010regularization}.
However, glmnet cannot handle very large datasets such as biobank-scale data that are now available in human genetics, where both the sample size and the number of variables are very large.
One strategy used to run penalized regressions on such large datasets such as the UK Biobank \cite[]{bycroft2018uk} has been to apply a variable pre-selection step before fitting the lasso \cite[]{lello2018accurate}.
Recently, authors of the glmnet package have developed a new R package, snpnet, to fit penalized regressions on the UK Biobank without having to perform any pre-filtering \cite[]{qian2020fast}.
Earlier, we developed two R packages for efficiently analyzing large-scale (genetic) data, namely bigstatsr and bigsnpr \cite[]{prive2018efficient}.
We then specifically derived a highly efficient implementation of penalized linear and logistic regressions in R package bigstatsr, and showed how these functions were useful for genetic prediction with some applications to the UK Biobank \cite[]{prive2019efficient}.
Here we benchmark bigstatsr against snpnet for fitting penalized regressions on large genetic data.
Through some theoretical expectations and empirical comparisons, we show that package bigstatsr is generally much faster than snpnet.
We also make more recommendations on how to fit penalized regressions in the context of genetic data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection*{Main motivation for snpnet}

Before we can present the main motivation behind snpnet developed by \cite{qian2020fast}, let us recall how the lasso regression is fitted.
Fitting the lasso consists in finding regression coefficients $\beta$ that minimize the following regularized loss function 
\begin{equation}
L(\lambda) = \underbrace{ \sum_{i=1}^n \left( y_i - \beta_0 - \sum_{j=1}^p X_{i,j} \beta_j \right)^2 }_\text{Loss function}   +   \underbrace{ \lambda \sum_{j=1}^p |\beta_j| }_\text{Penalization} ~,\label{eq:lasso}
\end{equation}
where $X$ denotes the matrix composed of $p$ (standardized) genotypes and possible covariates (e.g.\ sex, age and principal components) for $n$ individuals, $y$ is the (continuous) trait to predict, $\lambda$ ($> 0$) is a regularization hyper-parameter that control the strength of the penalty.
For a sequence of $\lambda$, one can find $\argmin_{\beta} L(\lambda)$ using cyclical coordinate descent \cite[]{friedman2010regularization}.
To speed up the coordinate descent, one can use sequential strong rules for discarding lots of variables, i.e.\ setting lots of $\beta_j$ to $0$, a priori \cite[]{tibshirani2012strong}.
Therefore the cyclical coordinate descent used to solve the lasso can be performed in a subset of the data only thanks to these strong rules.
However, the main drawback of these strong rules is that they require checking Karush-Kuhn-Tucker (KKT) conditions a posteriori, usually in two phases.
These KKT conditions are first checked in the ever-active set, i.e.\ the set of all variables $j$ with $\beta_j \neq 0$ for any previous $\lambda$.
Then, the cyclical coordinate descent has to be rerun while adding the new variables that do not satisfy these KKT conditions (if any).
In a second phase, the KKT conditions are also checked for all the remaining variables, i.e.\ the ones not in the ever-active set.
This last step requires to pass over the whole dataset at least once again for every $\lambda$ tested.
Then, when the available random access memory (RAM) is not large enough to cache the whole dataset, data has to be read from disk, which can be extremely time consuming.
To alleviate this particular issue, \cite{qian2020fast} have developed a clever approach called batch screening iterative lasso (BASIL) to be able to check these KKT conditions on the whole dataset only after having fitted solutions for many $\lambda$, instead of performing this operation for each $\lambda$.
Hence, for very large datasets, the BASIL procedure enables to fit the exact lasso solution faster than when checking the KKT conditions for all variables at each $\lambda$, as performed in e.g.\ R package biglasso \cite[]{zeng2017biglasso}.

\subsection*{A more pragmatic approach in bigstatsr}

Since our package bigstatsr is not using this BASIL procedure, one might think that fitting penalized regression with R package bigstatsr should be slow for very large datasets.
We do check the KKT conditions for variables in the ever-active set, i.e.\ for a (small) subset of all variables only; this first checking is therefore fast. However, KKT conditions almost always hold when $p > n$ \cite[]{tibshirani2012strong}, which is particularly the case for the remaining variables in the second phase of checking. 
Because of this, we decided in \cite{prive2019efficient} to skip this second checking when designing functions \texttt{big\_spLinReg} and \texttt{big\_spLogReg} for fitting penalized regression on very large datasets in R package bigstatsr.
Thanks to this approximation, these two functions effectively access all variables only once at the very beginning to compute the statistics used by the strong rules, and then access a subset of variables only (the ever-active set).
As we show later, this means that fitting penalized regressions using the approximation we proposed in \cite{prive2019efficient} is computationally more efficient than using the BASIL procedure proposed by \cite{qian2020fast}, and yet provides as accurate predictors.
Moreover, as bigstatsr uses memory-mapping, data that resides on disk is accessed only once from disk to memory and then stays in memory while there is no need to free memory.
Only when the ever-active set becomes very large, e.g.\ for very polygenic traits, memory can become an issue, but this extreme case would become a problem for package snpnet as well.
Please refer to the Discussion section of \cite{prive2019efficient} for more details on these matters.
In summary, bigstatsr effectively performs only one pass on the whole dataset while snpnet performs many passes, even though the number of passes in snpnet is reduced thanks to the BASIL approach.
Moreover, bigstatsr still uses a single pass even when performing CMSA (a variant of cross-validation (CV), see figure \ref{fig:CMSA}) internally, whereas performing CV with snpnet would multiply the number of passes to the data by the number of folds used in the CV.

\begin{figure}[!h]
	\centerline{\includegraphics[width=0.85\textwidth]{simple-CMSA}}
	\caption{Illustration of one turn of the Cross-Model Selection and Averaging (CMSA) procedure.
		This figure comes from \cite{prive2019efficient}; the Genetics Society of America has granted us permission to re-use it. First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds). 
		Secondly, in turn, each fold is considered as an inner validation set (red) and the other ($K - 1$) folds form an inner training set (blue). A ``regularization path'' of models is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed. The model that minimizes the loss on the inner validation set is selected. Finally, the $K$ resulting models are averaged; this is different to standard cross-validation where the model is refitted on the whole training set using the best-performing hyper-parameters. 
		We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making this procedure much faster.}
	\label{fig:CMSA}
\end{figure}

\subsection*{Benchmark}

Before, we have presented why we expect bigstatsr to be more efficient than snpnet.
To practically support this claim, we perform comparisons for the four real traits used in the UK Biobank analyses of \cite{qian2020fast}.
We compare R package snpnet (v0.3) with bigstatsr (v1.3) and bigsnpr (v1.5).
We use similar quality controls as \cite{qian2020fast} (see ``Data \& Methods'').
We also use the same splitting strategy: 20\% test, 20\% validation and 60\% training.
To use the same sets for bigstatsr as for snpnet, we use the same test set, use K=4 folds for training with bigstatsr while making sure the first split is composed of the same 20\% of the data used for validation in snpnet. 
Moreover, we use penalty factors to effectively use unscaled genotypes in bigstatsr (see ``Conclusion \& further recommendations''), as performed by default in snpnet.
This enables us to compare predictions from snpnet and bigstatsr using the exact same model and the same single validation fold.
Note that, to make the most of the training set, bigstatsr uses CMSA (Figure \ref{fig:CMSA}) while \cite{qian2020fast} propose to refit the model (on the whole training + validation) using the best $\lambda$ identified using the validation set in snpnet.
Also note that the parallelism used by snpnet and bigstatsr is different; snpnet relies on PLINK 2.0 to check KKT conditions in parallel, while bigstatsr parallelizes fitting of models from different folds and hyper-parameters.
Because bigstatsr uses memory-mapping, the data is shared across processes and therefore it can fit these models in parallel without multiplying the memory needed.
We allow for 16 cores to be used in these comparisons; bigstatsr effectively uses only 4 here (the number of folds).
We allow for 128 GB of RAM for bigstatsr, but allow for 500 GB of RAM for snpnet because we had memory issues running it with only 128 GB or 256 GB.

Table \ref{tab:results} presents the results of this benchmark.
Fitting lasso is 35 times faster using bigstatsr than using snpnet for high cholesterol, 29 times faster for asthma, 16 times faster for BMI, and 4.5 times faster for height.
When using only one validation fold for choosing the best-performing $\lambda$ and no refitting, snpnet and bigstatsr provide the same predictive performance, validating the use of the approximation in bigstatsr.
When using the whole training set, i.e.\ when refitting in snpnet and using CMSA in bigstatsr, predictive performance is much higher than when the validation set is not used for training. 
For example, partial correlation for height is of 0.6116 with CMSA (i.e.\ using the average of 4 models) compared to 0.5856 when using only one of these models, showing how important it is to make the most of the training + validation sets.
Also, CMSA can provide slightly higher predictive performance than the refitting strategy in snpnet, e.g.\ a partial correlation of 0.3324 vs 0.3221 for BMI.

\begin{table}[h]
	\caption{Benchmark of snpnet against bigstatsr in terms of predictive performance and computation time. Predictive performance is reported in terms of partial correlations between the polygenic scores and the phenotypes, residualized using the covariates. Timings are reported in minutes. Timings for snpnet report the training for 60\% of the data (using the training set only) + the refitting for 80\% of the data (using both the training and validation sets). Timings for bigstatsr report the time taken by the CMSA procedure (fitting K=4 models here). \label{tab:results}}
	\vspace*{0.5em}
	\centering
	\begin{tabular}{|l|c|c|c|c|c|c|}
		\cline{2-7}
		\multicolumn{1}{l|}{} & \multicolumn{3}{c|}{snpnet} & \multicolumn{3}{c|}{bigstatsr} \\
		\hline
		Trait & Perf. (1 fold) & Perf. (refit) & Time & Perf. (1 fold) & Perf. (CMSA) & Time \\
		\hline
		Asthma & 0.1349 & 0.1438 & 188 + 101 & 0.1348 & 0.1493 & 10 \\
		High cholesterol & 0.1254 & 0.1366 & 101 + 146 & 0.1257 & 0.1387 & 7 \\
		BMI & 0.3031 & 0.3231 & 161 + 893 & 0.3018 & 0.3324 & 65 \\
		Height & 0.5871 & 0.6106 & 409 + 715 & 0.5856 & 0.6116 & 249 \\
		\hline
	\end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection*{Conclusion \& further recommendations}

We have found the BASIL approach derived in \cite{qian2020fast} to be a clever approach that alleviates the I/O problem of other penalized regression implementations for very large datasets.
BASIL makes significant and valuable contributions to the important problem of fitting penalized regression models efficiently. 
However, we also find that the implementation of BASIL in snpnet is still an order of magnitude slower than our  package bigstatsr, which uses a simpler and more pragmatic approach \cite[]{prive2019efficient}.
Hereinafter we also come back to some statements made in \cite{qian2020fast} and provide more recommendations on how to best use penalized regression for deriving polygenic scores based on very large individual-level genetic data. This also enables us to highlight further similarities and differences between implementations from snpnet and bigstatsr.

First, in their UK Biobank applications, \cite{qian2020fast} have tried using elastic-net regularization (a combination of L1 and L2 penalties) instead of lasso (only L1), i.e.\ introducing a new hyper-parameter $\alpha$ ($0 < \alpha < 1$, with the special case of $\alpha = 1$ being the L1 regularization). 
They show that L1 regularization is very effective for very large sample sizes, and elastic-net regularization is not needed in this case, which we have also experienced.
Yet, in smaller sample sizes and for very polygenic architectures, we showed through extensive simulations that using lower values for $\alpha$ can significantly improve predictive performance \cite[]{prive2019efficient}.
In \cite{qian2020fast}, they tried $\alpha \in \{0.1, 0.5, 0.9, 1\}$; we recommend to use a grid on the log scale with smaller values (e.g.\ $1$, $0.1$, $0.01$, $0.001$, and even until $0.0001$) for smaller sample sizes.
Note that using a smaller $\alpha$ leads to a larger number of non-zero variables and therefore more time and memory required to fit the penalized regression.
In functions \texttt{big\_spLinReg} and \texttt{big\_spLogReg} of R package bigstatsr, we allow to directly test many $\alpha$ values in parallel within the CMSA procedure.
Therefore an optimal $\alpha$ value can be chosen automatically within the CMSA framework, without the need for more passes on the data.

Second, for large datasets, one should always use early-stopping. We have not found this to be emphasized enough in \cite{qian2020fast}.
Indeed, while fitting the regularization path of decreasing $\lambda$ values on the training set, one can monitor the predictive performance on the validation set, and stop early in the regularization path when the model starts to overfit (Figure \ref{fig:CMSA}). 
For large datasets, performance on the validation sets is usually very smooth and monotonic (before and after the minimum) along the regularization path, then one can safely stop very early, e.g.\ after the second iteration for which prediction becomes worse on the validation set.
This corresponds to setting \texttt{n.abort=2} in bigstatsr and \texttt{stopping.lag=2} in snpnet.
This is particularly useful because, when we move down the regularization path of $\lambda$ values, more and more variables enter the model and the cyclical coordinate descent takes more and more time and memory.
Therefore, the early-stopping criterion used in both bigstatsr and snpnet prevents from fitting very costly models, saving a lot of time and memory.

Third, \cite{qian2020fast} recommend not to use scaled genotypes when applying lasso to genetic data. 
However, using scaled genotypes is common practice in genetics, and is the assumption behind models in popular software such as GCTA and LDSC \cite[]{yang2011gcta,bulik2015ld}.
Scaling genotypes assumes that, on average, all variants explain the same amount of variance and that low-frequency variants have larger effects.
\cite{speed2012improved} argued that this assumption might not be reasonable and proposed another model: $\mathbb{E}[h^2_j] \propto \left[p_j (1 - p_j)\right]^\nu$, where $h^2_j$ is the variance explained by variant $j$ and $p_j$ is its allele frequency.
In \cite{speed2017reevaluation}, they estimated $\nu$ to be between $-0.25$ and $-0.5$ for most traits.
Note that scaling genotypes by dividing them by their standard deviations $\text{SD}_j$ as done by default in bigstatsr assumes $\nu=-1$ while not using any scaling as argued by \cite{qian2020fast} assumes $\nu=0$.
Therefore, using a trade-off between these two approaches can provide higher predictive performance and is therefore recommended \cite[]{zhang2020improved}.
In the case of L1 regularization, using a different scaling can be obtained by using different penalty factors $\lambda_j$ in equation \eqref{eq:lasso}, which is an option available in both bigstatsr and snpnet.
For example, using $\lambda_j=1/\text{SD}_j$ allows to effectively use unscaled genotypes.
Recently, we have implemented a new parameter \texttt{power\_scale} to allow for different scalings when fitting the lasso in bigstatsr. Note that a vector of values to try can be provided, and the best-performing scaling is automatically chosen within the CMSA procedure.

Fourth, \cite{qian2020fast} write that bigstatsr ``do not provide as much functionality as needed in [their] real-data application'', mainly because bigstatsr requires converting the input data and cannot handle missing values.
It is true that bigstatsr uses an intermediate format, which is a simple on-disk matrix format accessed via memory-mapping.
However, package bigsnpr provides fast parallel functions \texttt{snp\_readBed2} for converting from `.bed' files and \texttt{snp\_readBGEN} for converting from imputed `.bgen' files, the two formats used by the UK Biobank.
For example, it took 6 minutes only to read from the UK biobank `.bed' file used in this paper. We then used function \texttt{snp\_fastImputeSimple} to impute by the variant means in 5 minutes only, which is also the imputation strategy used in snpnet.
When reading imputed dosages instead, it takes less than one hour to access and convert 400K individuals over 1M variants using function \texttt{snp\_readBGEN} with 15 cores, and less than three hours for 5M variants.
When available, we recommend to directly read from `.bgen' files to get dosages from external reference imputation.
As for package snpnet, it uses the PLINK 2.0 `.pgen` format, which is still under active development (in alpha testing, see \url{https://www.cog-genomics.org/plink/2.0/formats#pgen}).
This format is not currently provided by the UK Biobank, and can therefore be considered as an intermediate format as well.

\subsection*{Data \& Methods}

As in \cite{qian2020fast}, we use the UK Biobank data \cite[]{bycroft2018uk}, which is a large cohort of half a million individuals from the UK, for which we have access to both genotypes and multiple phenotypes (\url{https://www.ukbiobank.ac.uk/}).
We apply some quality control filters to the genotyped data; we remove individuals with more than 10\% missing values, variants with more than 1\% missing values, variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants. 
We restrict individuals to the ones used for computing the principal components in the UK Biobank (Field 22020); these individuals are unrelated and have passed some quality control \cite[]{bycroft2018uk}.
We also restrict to the ``White British'' group defined by the UK Biobank (Field 22006) to get a set of genetically homogeneous individuals.
These filters result in 337,475 individuals and 504,139 genotyped variants.

We use the same four phenotypes as used in \cite{qian2020fast}, namely height, body mass index (BMI), high cholesterol and asthma. We define height using field 50, BMI using field 21001, high cholesterol using field 20002 (``Non-cancer illness code, self-reported''). Asthma is defined using field 20002 as well as fields 40001, 40002, 41202 and 41204 (ICD10 codes); please see code for further details at \url{https://github.com/privefl/paper2-PRS/tree/master/response-snpnet/code}.
For height and BMI, L1-penalized linear regressions are fitted using function \texttt{big\_spLinReg} from bigstatsr and using parameter \texttt{family=``gaussian''} in snpnet.
For high cholesterol and asthma,  L1-penalized logistic regressions are fitted using function \texttt{big\_spLogReg} from bigstatsr and using parameter \texttt{family=``binomial''} in snpnet.
We use sex (Field 22001), age (Field 21022), and the first 16 principal components (Field 22009) as unpenalized covariates when fitting the lasso models.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\clearpage

\bibliographystyle{natbib}
\bibliography{refs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2020-12-21 response-snpnet/supp-note-snpnet-comparison.tex
hug

@@
 However, glmnet cannot handle very large datasets such as biobank-scale data that are now available in human genetics, where both the sample size and the number of variables are very large.
-One strategy used to run penalized regressions on such large datasets such as the UK Biobank \cite[]{bycroft2018uk} has been to apply a variable pre-selection step before fitting the lasso \cite[]{lello2018accurate}.
+One strategy used to run penalized regressions on large datasets such as the UK Biobank \cite[]{bycroft2018uk} has been to apply a variable pre-selection step before fitting the lasso \cite[]{lello2018accurate}.
 Recently, authors of the glmnet package have developed a new R package, snpnet, to fit penalized regressions on the UK Biobank without having to perform any pre-filtering \cite[]{qian2020fast}.
@@
 Through some theoretical expectations and empirical comparisons, we show that package bigstatsr is generally much faster than snpnet.
-We also make more recommendations on how to fit penalized regressions in the context of genetic data.
+We also take that opportunity to provide more recommendations on how to fit penalized regressions in the context of genetic data.
 
@@
 
-Since our package bigstatsr is not using this BASIL procedure, one might think that fitting penalized regression with R package bigstatsr should be slow for very large datasets.
-We do check the KKT conditions for variables in the ever-active set, i.e.\ for a (small) subset of all variables only; this first checking is therefore fast.
+In our R package bigstatsr, we proposed a different strategy.
+We also check the KKT conditions for variables in the ever-active set, i.e.\ for a (small) subset of all variables only; this first checking is therefore fast.
 However, KKT conditions almost always hold when $p > n$ \cite[]{tibshirani2012strong}, which is particularly the case for the remaining variables in the second phase of checking.
@@
 Thanks to this approximation, these two functions effectively access all variables only once at the very beginning to compute the statistics used by the strong rules, and then access a subset of variables only (the ever-active set).
-As we show later, this means that fitting penalized regressions using the approximation we proposed in \cite{prive2019efficient} is computationally more efficient than using the BASIL procedure proposed by \cite{qian2020fast}, and yet provides as accurate predictors.
+As we show later, this means that fitting penalized regressions using the approximation we proposed in \cite{prive2019efficient} is computationally more efficient than using the BASIL procedure proposed by \cite{qian2020fast}, and yet provides equally accurate predictors.
 Moreover, as bigstatsr uses memory-mapping, data that resides on disk is accessed only once from disk to memory and then stays in memory while there is no need to free memory.
@@
 For example, partial correlation for height is of 0.6116 with CMSA (i.e.\ using the average of 4 models) compared to 0.5856 when using only one of these models, showing how important it is to make the most of the training + validation sets.
-Also, CMSA can provide slightly higher predictive performance than the refitting strategy in snpnet, e.g.\ a partial correlation of 0.3324 vs 0.3221 for BMI.
+Also, CMSA can provide slightly higher predictive performance than the refitting strategy in snpnet, with e.g.\ a partial correlation of 0.3324 vs 0.3221 for BMI.
 
@@
 
-Fourth, \cite{qian2020fast} write that bigstatsr ``do not provide as much functionality as needed in [their] real-data application'', mainly because bigstatsr requires converting the input data and cannot handle missing values.
+Fourth, \cite{qian2020fast} stated that bigstatsr ``do not provide as much functionality as needed in [their] real-data application'', mainly because bigstatsr requires converting the input data and cannot handle missing values.
 It is true that bigstatsr uses an intermediate format, which is a simple on-disk matrix format accessed via memory-mapping.
